### 2023-06-07 manuscript_v04.Rmd
manuscript draft 1

---
title: "Detection of eco-evolutionary dynamics in metacommunities using Joint Species Distribution Models"
author: "Jelena H. Pantel, Ruben Hermann"
date: Ecological Modelling, Faculty of Biology, University of Duisburg-Essen, Universitätsstraße 5, 45141 Essen, Germany
output:
  pdf_document: default
  bookdown::pdf_document2: default
  html_document: default
  word_document:
    reference_docx: docx_template.docx
header-includes:
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \fancyhead[L]{Pantel, JH}
- \fancyhead[R]{Detection of ecoevo dynamics}
- \usepackage{lineno}
- \linenumbers
linestretch: 2
link-citations: yes
linkcolor: blue
csl: evolution.csl
bibliography: refs.bib
---

```{r setup,echo=FALSE,warning=FALSE,message=FALSE}
library(bookdown)
library(R.matlab)
library(Hmsc)
library(vegan)
library(qgraph)
library(SNFtool)

library(ecoevor)
library(here)

library(grDevices)
library(knitr)

library(beanplot)
library(dplyr)
library(gplots)
library(formatR)
library(rmarkdown)
library(rsvg)
```
```{r echo=FALSE,eval=TRUE}
CV <- function(x){
  a <- mean(x,na.rm=T)	#Get the average of all values
  b <- (x - a)^2					#Squared deviations from mean
  c <- sqrt((sum(b,na.rm=T))/(length(x[is.na(x) == F])-1)) #sum squared deviations, divide by sample N-1, take square root
  d <- c/a  #coefficient of variation
  return(CV=d)
}

standard_error <- function(x){
	st_err <- sd(x) / sqrt(length(x))

	return(st_err)
	}
```

Abstract
===============================================================================

Biodiversity at the metacommunity scale is typically influenced by a number of environmental, spatial, biotic, and stochastic factors. At the same time, populations of individual species are evolving in ways that can be impacted by the same factors, as sites present different selection pressures and site connectivity can impact gene flow and genetic drift. Identifying the relative impacts of environmental, spatial, biotic, and other drivers of community composition across spatial and temporal scales has been greatly facilitated by joint species distribution models, but these models have not yet considered the role that evolution can play in driving changes in composition. I used Heirarchical Models of Species Communities (HMSC) to analyze data from a simulation model of an evolving metacommunity, in order to establish whether these models can sufficiently quantify the contribution of phenotypic evolution for metacommunity composition. The models successfully partitioned variance contributed by environmental, spatial, and phenotypic drivers, and also estimated site- and year-specific covariance. I also included estimates for eco-evolutionary interactions, and the model successfully estimated their role as well. The study of eco-evolutionary dynamics may require data that reflects numerous complex, interacting processes and its necessary to have flexible, generalized statistical models to analyze this data. HMSC models present one promising path for analysis of eco-evolutionary dynamics in multi-species communities.

Introduction
================================================================================

Analysis of biodiversity at large spatial scales can be complex, because numerous abiotic factors, spatial structure and connectivity, interactions between species, and also stochastic processes such as drift and priority effects operate and determine observed patterns. Statistical models in ecology have made tremendous progress in reflecting these numerous ecological processes at multiple scales, and there are now a diversity of joint species distribution models (Pollock et al. 2014 [@Pollock2014]; Pichler & Hartig 2021 [@Pichler2021]) or multi-species occupancy models (REF; Devarajan et al. 2020). These models can incorporate critical components that structure variance in biological data such as observer, measurement, and process error, error propagation, the existence of present but undetected species or life stages (REF) and other forms of imperfect detection (REF). Many of these models increasingly have implementations in R packages (e.g. spOccupancy, Doser et al. 2022), and there are intriguing extensions of these occupancy models with consideration of the error and data types particularly associated with eDNA (e.g. EDNAOCCUPANCY, Dorazio & Erickson 2018) and citizen science data (Altwegg & Nichols 2019).

In addition to ecological processes, biodiversity at different levels of ecological organization – populations, communities, and ecosystems – are partially determined by the genotypic and phenotypic properties of organisms as well. These ecological processes can impact selection on genes and phenotypes, potentially resulting in dynamics feedback loops between ecological and evolutionary processes (Lion 2018; Barbour et al. 2022). Although an increasing number of theoretical (REF) and empirical studies have evaluated the impacts of evolutionary dynamics for community assebmly and metacommunity dynamics (REF), analytical methods to consider this additional complexity lag behind the scope of data collected in eco-evolutionary studies. A series of methods papers (Hairston et al. 2005; Ellner et al. 2011; Govaert et al. 2016) has made progress in utilizing an ANOVA framework to partition variance (in phenotypes or in other ecological response variables) across contributing ecological and evolutionary components. These approaches have been applied to categorical ecological and evolutionary treatments with two levels. For example, the study of terHorst et al. (2014) evaluated the impacts of a categorical evolutionary treatment, where host plants were experimentally adapted to a dry or wet soil environment, and a categorical environmental treatment, where experiments were conducted in a dry or wet soil environment, for bacterial and fungal community richness and diversity. Hiltunen and Becks (2014) compared the relative importance of the (ecological) change in prey density and the (evolutionary) change in prey defense traits for the change in predator density from one time point to the next. However, these methods are ultimately limited by the same requirements that can limit application of ANOVA - data should follow a normal distribution, the processes that structure the target response variable must be linear and additive, and complicated structuring mechanisms must be reduced into one or a few target categories labeled as ‘ecology’ or ‘evolution’. Flexible models, which can vary depending on the mechanisms and structure of the diverse kinds of data that most studies of eco-evolutionary dynamics will collect, are needed for rigorous inference about eco-evolutionary processes. These models should consider the mechanistic basis of data structure, with realistic models of variance and uncertainty, and ideally can be used to make predictions about future states of the system.

Two existing statistical models are well suited for analysis of eco-evolutionary data at the microevolutionary and metacommunity scale. Lasky et al. (2020) developed an integrated reaction norm model, linking genetic, phenotypic, and demographic processes, and Benito Garzón et al. (2019) presented a species distribution model with local adaptation and phenotypic plasticity ($\Delta$ SDMs). The model of Lasky et al. is an excellent candidate for spatially complex population dynamics, and awaits an extension to the multispecies level. The $\Delta$ SDM models described by Benito Garzón et al. truly disentangle the role that plastic and evolutionary trait divergence can play in species distributions, but they also require common garden experimental data across sites to estimate these effects. Data typically collected in a metacommunity survey (of eukaryotes, which thus focuses on the species level) may only be time series of population size and trait values, with surveys of environmental properties and spatial connectivity as potential predictors.

One particular metacommunity statistical model, the Hierarchical Modelling of Species Communities (HMSC) framework (Ovaskainen et al. 2017; Tikhonov et al. 2020), was built explicitly to consider the multitude of processes that can structure species occupancy and abundance data across time and space and is thus well suited to consider analysis of metacommunity eco-evolutionary data (Leibold et al. 2022). This model of environmental filtering via variation and covariation in how species respond to their environment considers the impacts of traits and phylogenetic relationships, and uses latent variables to account for the numerous unobserved environmental and spatial features that are difficult to exhaustively measure in community data. The associated R package has immense flexibility and accompanying data analysis examples and code, and has been applied to study environmental responses of diverse assemblages of organisms (REF). In this study, I use a simulation of an evolving metacommunity model to generate realistic time series of population sizes and mean trait values for multiple species in a spatially and environmentally complex landscape. I then use HMSC to analyze the resulting data, to determine whether HMSC can successfully estimate the impacts of trait evolution for community composition, and to generate relative contributions of fixed environmental, spatial, and trait evolution drivers as well as random effects of spatial and temporal covariance among species. Finally, in order to determine whether HMSC is useful for hypothesis testing about the drivers of eco-evolutionary dynamics in communities in a landscape, I also conducted a few numerical experiments with the evolving metacommunity simulation - I varied the degree of environmental fluctuations in the landscape, and I varied the strength of species interactions, to determine whether these potential drivers of increased selection pressure alter the importance of evolution for metacommunity composition. For these experiments, I hypothesized that XX and YY. I found that XXXX.

Methods
================================================================================

**Simulation model**

Population growth for species in the metacommunity simulation follows a Leslie-Gower model (a discrete-time version of a Lotka-Volterra model; Beverton & Holt 1957; Leslie & Gower 1958). I consider the impact of trait evolution for growth using a discrete time quantitative genetic model of evolutionary rescue (Gomulkiewicz & Holt 1995). The model for population size is as follows:

$$ N_{i,t+1} =\frac{\hat{W}e^ \frac{-[(\frac{w+(1-h^2)P} {P+w})(E-x_t)]^2}{2(P+w)}N_{i,t}} {1 + \alpha_{ii}N_{i,t} + \sum_{j \neq i}^{S} \alpha_{ij}N_{j,t}} $$
where $N_i,t$ is the population size of species *i* at time *t*, \( \hat{W} \) is calculated as \( \hat{W}=W_{max}\sqrt(\frac{w}{P+w}) \), \(W_{max}\) is the species' maximum per-capita growth rate, *w* is the width of the Gaussian fitness function (which determines the strength of selection), *P* is is the width of the distribution of the phenotype *x*, $h^2$ is the heritability of the trait *x*, *E* is the local environmental optimum trait value, $x_{i,t}$ is the trait value of species *i* at time *t*, $\alpha_{ii}$ is the intraspecific competition coefficient (the per capita impact of species *i* on itself) and $\alpha_{ij}$ is the interspecific competition coefficient. Populations have a critical density $N_c$, below which the population is subject to extinction due to demographic stochasticity at a probability of *p* (Gomulkiewicz & Holt 1995).

This model was used in Pantel & Becks (2023) to evaluate the consequences of adaptive evolution for coexistence in a three-species system. To expand this model to a metacommunity, I consider the evolution of multiple species (*S* = 15) in a landscape of patches (*k* = 50). The patches have values of an environmental property *E* (drawn from a uniform distribution ranging from 0 to 1) that determines the local optimum phenotype *E*, i.e. where species experience their absolute fitness $W_{max}$, and patches also have spatial locations *X* and *Y* (both drawn from U(0, 1)). The patches thus have a connectivity matrix **D** (here given by their Euclidean distance), as well as a connectivity matrix **C** that is a Gaussian function of **D** and a species dispersal rate $d_i$.

To simulate phenotypic evolution as well as species growth and competition in the landscape, I used parameter values of \(W_{max}\) = 2, *P* = 1, *w* = 10, $N_c$ = 100, and *p* = 0.001. The simulated metacommunity was initialized in a way that mimics community assembly in small vernal pools that periodically flood (and thus are reset seasonally). In the simulated pool system, local communities are drawn at random from the regional species pool, and thus their initial degree of maladaptation in the pool is initially random as well. The initial species richness of each site $s_{0k}$ was drawn from a random Poisson distribution, $s_{0k} \sim Pois(0.75)$, the initial population size for all species $N_{0i} \sim Pois(10)$, and the initial degree of maladaptation \(\beta_{i0} \sim Gamma(0.75,1)\). From this \(\beta_{i0}\), the initial distance to the local patch's optimum phenotype was calculated as $d_{i0}=\sqrt{B_{i0}(w+P)}$ and the initial trait value for each species $x_{i0}$ was calculated by subtracting $d_{i0}$ from the lowest site environmental value $E_k$.

The species interaction matrix was fixed throughout the simulation, with \(\alpha_{ii} = 0.00125\) and \(\alpha_{ij} \sim U(0.0015,1)\). After draws for all random parameter values, the same resulting initial community was used for all simulations, to allow direct comparison across dispersal and heritability values (Figure \@ref(fig:figS1)). Simulations were run across a range of dispersal values (identical for all species; *d* = 0, $10^{-9},10^{-8},10^{-7},10^{-6},10^{-5},10^{-4},10^{-3},10^{-2}$, 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 1) and heritability values (identical for all species: $h^2$ = 0, 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 1). The simulations were run in MATLAB (version R2022a) for 1000 time steps and produced records of $N_{ikt}$ and $x_{ikt}$.

**Statistical model**

To determine the relative influence of environmental and spatial properties, population dynamics, and evolutionary dynamics for community structure, I applied hierarchical modelling of species communities (HMSC, Ovaskainen et al. 2017; Tikhonov et al. 2020) to model the abundance at each site and year for each species $N_{ikt}$. The abundance values were modeled as $N_{ikt} \sim N(L_{ikt},\sigma_2)$, where the linear predictors **L** are the sum of fixed and random effects **L^F^** + **L^R^** (Tikhonov et al. 2020). The fixed effects were (1) $E_k$, the value of the site’s environmental parameter, (2) the value of each species' population size the time step before $N_{ik,t-1}$, (3) the absolute value of the change in each species' trait value from the previous time step to the next $|\Delta x_{ik,t-1 \to t}|$, (4) and an interaction term for population size by change in trait value $N_{ik,t-1} \times |\Delta x_{ik,t-1 \to t}|$, which led to a total of 46 predictors. The random effects were modeled at the site-level ($\epsilon_{ik} \sim N(0, \Omega^S)$, where $\Omega^S$ is the site species-to-species covariance matrix) and at the day level ($\epsilon_{it} \sim N(0, \Omega^T)$, where $\Omega^T$ is the time species-to-species covariance matrix) using a latent variable approach. In this approach, species-to-species covariance matrices are computed as $\Omega = \lambda \intercal \lambda$, where $\lambda$ are the loadings of each species with latent variables, and latent variables are modeled as $\epsilon_{ik} = \sum_B \eta_{kB} \lambda_{Bi}$ and $\epsilon_{it} = \sum_B \eta_{tB} \lambda_{Bi}$ for *B* latent variables considered (Ovaskainen et al. 2016). The spatial latent variables represent a spatial model, where species respond to some latent spatial predictor associated with the *xy* coordinates of the site (Tikhonov et al. 2020 [@Tikhonov2020]). Analysis and was conducted in R (version 4.1.1) using packages ‘Hmsc’, 'R.matlab', and XX (Bengtsson 2022; Tikhonov et al. 2022; XX).

Results
===============================================================================

The simulation model followed population size and trait value dynamics for species in the landscape across 1000 time steps, and results depended on the dispersal level and heritability values (Figure \@ref(fig:fig1)). For dispersal, results matched predictions of existing metacommunity models demonstrating that diversity is higher for intermediate rates of dispersal (as species can increasingly reach more suitable patches at intermediate dispersal, but then experience mass effects at higher dispersal levels; Mouquet & Loreau 2003 [@Mouquet2003]; Figure \@ref(fig:fig2)), and that among-site similarity (measured as the average among-site Hellinger distance; Legendre & Gallagher 2001 [@Legendre2001]) decreases with increasing dispersal level (Figure \@ref(fig:fig3)). The inclusion of trait heritability did have a strong effect on community diversity, although there was no indication that increasing heritability beyond $h^2 = 0.1$ had any impact (Figure \@ref(fig:figS2)).

Discussion
================================================================================

I have analysed data collected by Herman Bumpus [@Bumpus1898] on the relationship between sparrow (*Passer domesticus*) total length and surival following an unusually severe storm. I found that sparrows that died in the storm were longer than sparrows that survived, which suggests that higher sparrow body length decreased survival. Of course, it is not possible to definitively conclude a causal relationship between any aspect of body size and sparrow survival<!--- BD Note: maybe explain this better --->, and even the available data collected by Bumpus would permit a more thoughtful analysis than that conducted in this study (see [Appendix Table 1](#appendix)).

<!---

Here is one way to add some more detailed comments into the manuscript itself, though this can also be done within the text (see above).

--->

Overall, this document demonstrates how high quality, professional looking documents can be written using Rmarkdown. The [underlying code](https://github.com/StirlingCodingClub/Manuscripts_in_Rmarkdown/blob/master/ms.Rmd) for this manuscript is publicly available, along with [accompanying notes](https://stirlingcodingclub.github.io/Manuscripts_in_Rmarkdown/Rmarkdown_notes.html) to understand how it was written. By using Rmarkdown to write manuscripts, authors can more easily use version control (e.g., git) throughout the writing process. The ability to easily integrate citations though BibTeX, LaTeX tools, and dynamic R code can also make writing much more efficient and more enjoyable. Further, obtaining the benefits of using Rmarkdown does not need to come with the cost of isolating colleagues who prefer to work with Word or LaTeX because Rmarkdown can easily be converted to these formats (in the case of Word, with the push of a button). By learning all of the tools used in this manuscript, readers should have all of the necessary knowledge to get started writing and collaborating in Rmarkdown.

References
===============================================================================

<div id="refs"></div>

<a name="appendix">Figures & Tables</a>
================================================================================

```{r fig1, echo = FALSE, eval = TRUE, fig.width = 5, fig.height = 5, fig.cap = "Time series of population size (*y*-axis) over time (*x*-axis) for all species in 5 of 50 total patches, for varying levels of dispersal *d* and heritability $h^2$. All patches have a distinct value for *E*, the environmental parameter (that determines the local population mean fitness; here *E* = 0.60198194, 0.45054160, 0.08382138, 0.07596669, and 0.23991615 for patches from top to bottom row). Each of the *s* = 15 species is plotted with a unique color. The left column has no dispersal (*d*=0) and no evolution ($h^2$=0). The middle column includes intermediate dispersal (*d* = $10^-3$) and no evolution, while the third column has intermediate dispersal and evoltution ($h^2$=0.1)."}
d_lev <- c("d_zero","d_minus9","d_minus8","d_minus7","d_minus6","d_minus5","d_minus4","d_minus3","d_minus2","d_01","d_02","d_03","d_04","d_05","d_06","d_07","d_08","d_09","d_10")
h_lev <- c("h_0_","h_01_","h_02_","h_03_","h_04_","h_05_","h_06_","h_07_","h_08_","h_09_","h_10_")
setwd("data/")
result <- paste(h_lev[1],d_lev[1],"_res.mat",sep="")
r1  <- R.matlab::readMat(toString(result))$N
result <- paste(h_lev[1],d_lev[8],"_res.mat",sep="")
r2  <- R.matlab::readMat(toString(result))$N
result <- paste(h_lev[2],d_lev[8],"_res.mat",sep="")
r3  <- R.matlab::readMat(toString(result))$N

## Plot of Patch 6,11,12,46,47 community dynamics over time
#cl <- grDevices::colors(distinct = TRUE)
#mycols2 <- sample(cl, 15)
mycols2 <- c("maroon4","coral3","peachpuff3","gray47","orange3","darkorange","gray33","lightgoldenrod2","gray96","cadetblue4","gray40","lightseagreen","darkorchid4","midnightblue","rosybrown1" )

par(mfcol=(c(5,3)))
par(mar=c(2,2,1.5,1))
k <- 1
for(j in c(6,11,12,46,47)){
    plot(eval(parse(text=paste("r",k,sep="")))[j,1,],col=mycols2[1],ylim=c(0,800),axes=FALSE,type="n",ylab=NA,xlab=NA)
  for(i in 1:15){
    points(eval(parse(text=paste("r",k,sep="")))[j,i,],col=mycols2[i])
  }
  if (j %in% 6){
    axis(2,col="grey40",col.axis="grey20",at = c(0,250,500,750,1000))
    title(main=expression("d = 0, " ~ "h"^2 ~ "= 0"),cex.sub=2)
  }
  if (j %in% c(11,12,46)){
    axis(2,col="grey40",col.axis="grey20",at = c(0,250,500,750,1000))
  }
  if (j %in% c(47)){
    axis(1,col="grey40",col.axis="grey20",at = c(0,200,400,600,800))
    axis(2,col="grey40",col.axis="grey20",at = c(0,250,500,750,1000))
  }
}

k <- 2
for(j in c(6,11,12,46,47)){
    plot(eval(parse(text=paste("r",k,sep="")))[j,1,],col=mycols2[1],ylim=c(0,800),axes=FALSE,type="n",ylab=NA,xlab=NA)
  for(i in 1:15){
    points(eval(parse(text=paste("r",k,sep="")))[j,i,],col=mycols2[i])
  }
  if (j %in% 6){
    title(main=expression("d = 10"^-3 ~ ", h"^2 ~ "= 0"),cex.sub=2)
  }
  if (j %in% c(47)){
    axis(1,col="grey40",col.axis="grey20",at = c(0,200,400,600,800))
    }
}

k <- 3
for(j in c(6,11,12,46,47)){
    plot(eval(parse(text=paste("r",k,sep="")))[j,1,],col=mycols2[1],ylim=c(0,800),axes=FALSE,type="n",ylab=NA,xlab=NA,xaxs="i",yaxs="i")
  for(i in 1:15){
    points(eval(parse(text=paste("r",k,sep="")))[j,i,],col=mycols2[i])
  }
  if (j %in% 6){
    title(main=expression("d = 10"^-3 ~ ", h"^2 ~ "= 0.1"),cex.sub=2)
  }
  if (j %in% c(47)){
    axis(1,col="grey40",col.axis="grey20",at = c(0,200,400,600,800))
    }
}
```

\newpage

```{r fig2, echo = FALSE, eval = TRUE, warning=FALSE, message=FALSE, fig.width = 8, fig.height = 5, fig.cap = "Inverse Simpson's diversity ($^2D$) at the local (average $\\alpha$ value across all 50 sites, with error bars at ± 1 standard error) and regional ($\\gamma$) level, and turnover between sites (average $\\beta$ value across all 50 sites, with error bars at ± 1 SE) for $h^2$=0 (solid lines) and $h^2$=0.1 (dashed lines), indicating that local adaptation rescues numerous species that would otherwise become extinct."}
div_1000 <- array(NA,dim=c(19,11,3,3),dimnames=list(c("d_zero","d_minus9","d_minus8","d_minus7","d_minus6","d_minus5","d_minus4","d_minus3","d_minus2","d_01","d_02","d_03","d_04","d_05","d_06","d_07","d_08","d_09","d_10"),c("h_0_","h_01_","h_02_","h_03_","h_04_","h_05_","h_06_","h_07_","h_08_","h_09_","h_10_"),c("alpha","gamma","beta"),c("mean","CV","SDE")))

setwd("data/")
for(i in 1:length(d_lev)){
  for(j in 1:2){
    ## Read in population size values
    result <- paste(h_lev[j],d_lev[i],"_res.mat",sep="")
    r <- R.matlab::readMat(toString(result))$N[,,1000]    # alpha
    div_1000[i,j,1,1] <- mean(diversity(r[rowSums(r) != 0,],index="invsimpson")) # mean
    div_1000[i,j,1,2] <- CV(diversity(r[rowSums(r) != 0,],index="invsimpson")) # CV
    div_1000[i,j,1,3] <- standard_error(diversity(r[rowSums(r) != 0,],index="invsimpson")) # SE

    # gamma
    div_1000[i,j,2,1] <- if (is.null(dim(r[rowSums(r) != 0,])[1])) {NA} else if (dim(r[rowSums(r) != 0,])[1] < 2) {NA} else {diversity(colSums(r[rowSums(r) != 0,]),index="invsimpson")}

    # beta
    div_1000[i,j,3,1] <- if (is.null(dim(r[rowSums(r) != 0,])[1])) {NA} else if (dim(r[rowSums(r) != 0,])[1] < 2) {NA} else {mean(diversity(colSums(r[rowSums(r) != 0,]),index="invsimpson") / diversity(r[rowSums(r) != 0,],index="invsimpson"))}

    div_1000[i,j,3,2] <- if (is.null(dim(r[rowSums(r) != 0,])[1])) {NA} else if (dim(r[rowSums(r) != 0,])[1] < 2) {NA} else {CV(diversity(colSums(r[rowSums(r) != 0,]),index="invsimpson") / diversity(r[rowSums(r) != 0,],index="invsimpson"))}

    div_1000[i,j,3,3] <- if (is.null(dim(r[rowSums(r) != 0,])[1])) {NA} else if (dim(r[rowSums(r) != 0,])[1] < 2) {NA} else {standard_error(diversity(colSums(r[rowSums(r) != 0,]),index="invsimpson") / diversity(r[rowSums(r) != 0,],index="invsimpson"))}
  }
}

## Plot of mean and CV for gamma diversity
##t1000 only
plot(1:length(d_lev),div_1000[,1,2,1],type="n",ylim=c(0,15), xaxt="n", yaxt="n",xlab="dispersal (d)",ylab="diversity",main="")
axis(1, at=1:19, labels=c(0,expression(10^-9),expression(10^-8),expression(10^-7),expression(10^-6),expression(10^-5),expression(10^-4),expression(10^-3),expression(10^-2),0.1,0.2,0.3,0.4,0.5,0.6,0.7,0.8,0.9,1), las=1)
axis(2, at=c(0:15), las=1)
points(1:length(d_lev),div_1000[,1,2,1],pch=19,col="black",cex=2, type="b") # gamma

points(1:length(d_lev),div_1000[,1,3,1],pch=21,cex=2, type="b") # beta
arrows(1:length(d_lev), div_1000[,1,3,1], 1:length(d_lev), div_1000[,1,3,1]+div_1000[,1,3,3], length=0.05, angle=90, code=2, lwd=1, col="black")
arrows(1:length(d_lev), div_1000[,1,3,1], 1:length(d_lev), div_1000[,1,3,1]-div_1000[,1,3,3], length=0.05, angle=90, code=2, lwd=1, col="black")

points(1:length(d_lev),div_1000[,1,1,1],pch=19,col="darkgray",cex=2, type="b") # alpha
arrows(1:length(d_lev), div_1000[,1,1,1], 1:length(d_lev), div_1000[,1,1,1]+div_1000[,1,1,3], length=0.05, angle=90, code=2, lwd=1, col="black")
arrows(1:length(d_lev), div_1000[,1,1,1], 1:length(d_lev), div_1000[,1,1,1]-div_1000[,1,1,3], length=0.05, angle=90, code=2, lwd=1, col="black")

# Add h2 = 0.1
# gamma
points(1:length(d_lev),div_1000[,2,2,1],pch=19,col="black",cex=2)
lines(1:length(d_lev),div_1000[,2,2,1],lty=5)
# beta
points(1:length(d_lev),div_1000[,2,3,1],pch=21,cex=2)
lines(1:length(d_lev),div_1000[,2,3,1],lty=5)

arrows(1:length(d_lev), div_1000[,2,3,1], 1:length(d_lev), div_1000[,2,3,1]+div_1000[,2,3,3], length=0.05, angle=90, code=2, lwd=1, col="black")
arrows(1:length(d_lev), div_1000[,2,3,1], 1:length(d_lev), div_1000[,2,3,1]-div_1000[,2,3,3], length=0.05, angle=90, code=2, lwd=1, col="black")
# alpha
points(1:length(d_lev),div_1000[,2,1,1],pch=19,col="darkgray",cex=2)
lines(1:length(d_lev),div_1000[,2,1,1],lty=5)
arrows(1:length(d_lev), div_1000[,2,1,1], 1:length(d_lev), div_1000[,2,1,1]+div_1000[,2,1,3], length=0.05, angle=90, code=2, lwd=1, col="black")
arrows(1:length(d_lev), div_1000[,2,1,1], 1:length(d_lev), div_1000[,2,1,1]-div_1000[,2,1,3], length=0.05, angle=90, code=2, lwd=1, col="black")

legend("topright",legend = c(expression(gamma),expression(beta),expression(alpha)), pt.bg=c("black","white","gray"),pch=21,pt.cex=1.5,bty="n",y.intersp=1)
```

\newpage

```{r fig3, echo=FALSE, fig.show="hold", out.width="25%", fig.cap = "(A) Plot with among-site similarity plotted across dispersal levels for $h^2$=0 and $h^2$=0.1, for dispersal levels XXX. (B) Plot with average among-site similarity (error bars at ± 1 standard error are given but very small) across $h2$ levels."}
setwd("data/")
## Network similarity plot
cd <- array(NA,dim=c(19,11,2),dimnames=list(c("d_zero","d_minus9","d_minus8","d_minus7","d_minus6","d_minus5","d_minus4","d_minus3","d_minus2","d_01","d_02","d_03","d_04","d_05","d_06","d_07","d_08","d_09","d_10"),c("h_0_","h_01_","h_02_","h_03_","h_04_","h_05_","h_06_","h_07_","h_08_","h_09_","h_10_"),c("mean","SE")))

## Plot of among-site community similarity
# Plot labels
l <- rep(NA,8)
l[1] <- expression("d = 0")
l[2] <- expression("d = 10^-5")
l[3] <- expression("d = 10^-4")
l[4] <- expression("d = 10^-3")
l[5] <- expression("d = 0.1")
l[6] <- expression("d = 0.5")
l[7] <- expression("d = 0.8")
l[8] <- expression("d = 1")
count <- 1

par(mfrow=(c(8,2)))
par(mar=c(2,2,1.5,1))
for(j in 1:2){
  for(i in c(1,6,7,8,10,14,17,19)){
    count <- count+1
    ## Read in population size values
    result <- paste(h_lev[j],d_lev[i],"_res.mat",sep="")
    r <- R.matlab::readMat(toString(result))$N[,,1000]
    xy <- R.matlab::readMat(toString(result))$xy

    ##Hellinger-distance transform community data
    comm_hell <- decostand(r, "hellinger")
    comm_hell <- vegdist(comm_hell, "euclidean")
    comm_hell <- as.matrix(comm_hell)
    cd[i,j,1] <- mean(comm_hell)
    cd[i,j,2] <- standard_error(comm_hell)

    ##W1 = community
    ##Construct similarity graphs
    # Convert all values where one site has 0 species present to distance 0
    s1000 <- rowSums(r)
    s1000[s1000 > 0] <- 1
    for(k in 1:50){
      for(b in 1:50){
        if(s1000[k] == 0 | s1000[b] == 0){comm_hell[k,b] <- 1}
      }
    }
    diag(comm_hell) <- 1

    ## affinity graph
    K <- 49
    alpha <- 0.5
    W1 <- affinityMatrix(comm_hell, K, alpha)

    comm_dist <- data.frame(t(combn(1:50,2))[,1])
    comm_dist$to <- t(combn(1:50,2))[,2]
    comm_dist$dist <- as.numeric(W1[lower.tri(W1, diag=F)])
    colnames(comm_dist) <- c('from', 'to', 'dist')
    # White colors for edges between sites with no species present
    comm_dist$from_s <- NA
    comm_dist$to_s <- NA
    comm_dist$color <- "none"
    for(k in 1:50){
      comm_dist$from_s[comm_dist$from == k] <- s1000[k]
      comm_dist$to_s[comm_dist$to == k] <- s1000[k]
    }
    for(k in 1:nrow(comm_dist)){
      ifelse(comm_dist$from_s[k] + comm_dist$to_s[k] < 2,comm_dist$color[k] <- "white",comm_dist$color[k] <- "red")
    }

    qgraph(comm_dist[,1:3], esize=5, vsize = 2, gray=F, edge.color = comm_dist$color, directed=F, labels = 1:50, layout = xy, maximum = 10.98745, minimum = .035, label.cex=3, title = l[count])
  }
}
```

\newpage

<a name="appendix">Appendix 1</a>
================================================================================

```{r figS1, echo = FALSE, eval = TRUE, fig.width = 5, fig.height = 5, fig.cap = "Figure S1. Spatial maps showing position of all 50 sites in the simulation, their value for the single environmental variable E, and their value for initial Simpson's diversity at the beginning of the simulation.", warning=FALSE,message=FALSE}

## Figure S1. Plot of location of sites in landscape, associated E values, and associated diversity values
setwd("data/")
result <- paste(h_lev[2],d_lev[8],"_res.mat",sep="")
dat  <- R.matlab::readMat(toString(result))

layout(matrix(c(1,3,2,4),ncol=2), width = c(2,1),height = c(1,1))
pal <- colorRampPalette(c("white", "blue"))
plot(dat$xy,col="grey90", fill=TRUE, warnings=F, xlab="X coord",ylab="Y coord")
options(warn=0)
points(dat$xy[,1], dat$xy[,2],pch = 21,bg = pal(10)[as.numeric(cut(dat$E,breaks = 10))],col="black")

legend_image <- as.raster(matrix(rev(pal(10)), ncol=1))
plot(c(0,2),c(0,1),type = 'n', axes = F,xlab = '', ylab = '', main = 'E value')
text(x=1.5, y = seq(0,1,l=5), labels = seq(0,1,l=5))
rasterImage(legend_image, 0, 0, 1, 1) ## set new for overplot w/ next plot

pal <- colorRampPalette(c("white", "orange"))
a  <- vegan::diversity(dat$N[,,1],index="invsimpson")
a[is.finite(a)==FALSE] <- 0
plot(dat$xy,col="grey90", fill=TRUE, warnings=F, xlab="X coord",ylab="Y coord")
options(warn=0)
points(dat$xy[,1], dat$xy[,2],pch = 21,bg = pal(10)[as.numeric(cut(a,breaks = 10))],col="black")

legend_image <- as.raster(matrix(rev(pal(10)), ncol=1))
plot(c(0,2),c(0,1),type = 'n', axes = F,xlab = '', ylab = '', main = 'inverse Simpson diversity')
text(x=1.5, y = seq(0,1,l=6), labels = seq(0,15,l=6))
rasterImage(legend_image, 0, 0, 1, 1) ## set new for overplot w/ next plot
```

\newpage

```{r figS2, echo=FALSE, fig.show="hold", fig.cap = "Plots with (A) gamma, (B) alpha, and (C) beta diversity across levels of dispersal (x-axis), for values of heritability ranging from 0-1."}
setwd("data/")
## Inverse Simpson's diversity across additional heritability levels
for(i in 1:length(d_lev)){
  for(j in 3:length(h_lev)){
    ## Read in population size values
    result <- paste(h_lev[j],d_lev[i],"_res.mat",sep="")
    r <- R.matlab::readMat(toString(result))$N[,,1000]    # alpha
    div_1000[i,j,1,1] <- mean(diversity(r[rowSums(r) != 0,],index="invsimpson")) # mean
    div_1000[i,j,1,2] <- CV(diversity(r[rowSums(r) != 0,],index="invsimpson")) # CV
    div_1000[i,j,1,3] <- standard_error(diversity(r[rowSums(r) != 0,],index="invsimpson")) # SE

    # gamma
    div_1000[i,j,2,1] <- if (is.null(dim(r[rowSums(r) != 0,])[1])) {NA} else if (dim(r[rowSums(r) != 0,])[1] < 2) {NA} else {diversity(colSums(r[rowSums(r) != 0,]),index="invsimpson")}

    # beta
    div_1000[i,j,3,1] <- if (is.null(dim(r[rowSums(r) != 0,])[1])) {NA} else if (dim(r[rowSums(r) != 0,])[1] < 2) {NA} else {mean(diversity(colSums(r[rowSums(r) != 0,]),index="invsimpson") / diversity(r[rowSums(r) != 0,],index="invsimpson"))}

    div_1000[i,j,3,2] <- if (is.null(dim(r[rowSums(r) != 0,])[1])) {NA} else if (dim(r[rowSums(r) != 0,])[1] < 2) {NA} else {CV(diversity(colSums(r[rowSums(r) != 0,]),index="invsimpson") / diversity(r[rowSums(r) != 0,],index="invsimpson"))}

    div_1000[i,j,3,3] <- if (is.null(dim(r[rowSums(r) != 0,])[1])) {NA} else if (dim(r[rowSums(r) != 0,])[1] < 2) {NA} else {standard_error(diversity(colSums(r[rowSums(r) != 0,]),index="invsimpson") / diversity(r[rowSums(r) != 0,],index="invsimpson"))}
  }
}

## Plot results across heritability levels
## Plot of mean and SE for gamma diversity
##t1000 only
par(mfcol=c(3,1),mai = c(0.3, 0.3, 0.1, 0.1))
# gamma
plot(1:length(d_lev),div_1000[,1,2,1],type="n",ylim=c(0,15), xaxt="n", yaxt="n",xlab="dispersal (d)",ylab="diversity",main="")
axis(1, at=1:19, labels=c(0,expression(10^-9),expression(10^-8),expression(10^-7),expression(10^-6),expression(10^-5),expression(10^-4),expression(10^-3),expression(10^-2),0.1,0.2,0.3,0.4,0.5,0.6,0.7,0.8,0.9,1), las=1)
axis(2, at=c(0:15), las=1)
# h2=0
points(1:length(d_lev),div_1000[,1,2,1],pch=19,col="black",cex=.7, type="b") # gamma
#arrows(1:length(d_lev), div_1000[,1,2,1], 1:length(d_lev), div_1000[,1,2,1]+div_1000[,1,2,3], length=0.05, angle=90, code=2, lwd=1, col="black")
#arrows(1:length(d_lev), div_1000[,1,2,1], 1:length(d_lev), div_1000[,1,2,1]-div_1000[,1,2,3], length=0.05, angle=90, code=2, lwd=1, col="black")
# loop across h2
for(i in 2:length(h_lev)){
  points(1:length(d_lev),div_1000[,i,2,1],pch=19,col="black",cex=.7, type="b") # gamma
  #arrows(1:length(d_lev),div_1000[,i,2,1],1:length(d_lev),div_1000[,i,2,1]+div_1000[,i,2,3],length=0.05, angle=90, code=2, lwd=1,col="black")
  #arrows(1:length(d_lev),div_1000[,i,2,1],1:length(d_lev),div_1000[,i,2,1]-div_1000[,i,2,3],length=0.05, angle=90, code=2, lwd=1, col="black")
}

# alpha
plot(1:length(d_lev),div_1000[,1,2,1],type="n",ylim=c(0,15), xaxt="n", yaxt="n",xlab="dispersal (d)",ylab="diversity",main="")
axis(1, at=1:19, labels=c(0,expression(10^-9),expression(10^-8),expression(10^-7),expression(10^-6),expression(10^-5),expression(10^-4),expression(10^-3),expression(10^-2),0.1,0.2,0.3,0.4,0.5,0.6,0.7,0.8,0.9,1), las=1)
axis(2, at=c(0:15), las=1)
# h2=0
points(1:length(d_lev),div_1000[,1,1,1],pch=19,col="black",cex=.7, type="b") # alpha
#arrows(1:length(d_lev), div_1000[,1,1,1], 1:length(d_lev), div_1000[,1,1,1]+div_1000[,1,1,3], length=0.05, angle=90, code=2, lwd=1, col="black")
#arrows(1:length(d_lev), div_1000[,1,1,1], 1:length(d_lev), div_1000[,1,1,1]-div_1000[,1,1,3], length=0.05, angle=90, code=2, lwd=1, col="black")
# loop across h2
for(i in 2:length(h_lev)){
  points(1:length(d_lev),div_1000[,i,1,1],pch=19,col="black",cex=.7, type="b") # alpha
  #arrows(1:length(d_lev),div_1000[,i,1,1],1:length(d_lev),div_1000[,i,1,1]+div_1000[,i,1,3],length=0.05, angle=90, code=2, lwd=1,col="black")
  #arrows(1:length(d_lev),div_1000[,i,1,1],1:length(d_lev),div_1000[,i,1,1]-div_1000[,i,1,3],length=0.05, angle=90, code=2, lwd=1, col="black")
}

# beta
plot(1:length(d_lev),div_1000[,1,2,1],type="n",ylim=c(0,15), xaxt="n", yaxt="n",xlab="dispersal (d)",ylab="diversity",main="")
axis(1, at=1:19, labels=c(0,expression(10^-9),expression(10^-8),expression(10^-7),expression(10^-6),expression(10^-5),expression(10^-4),expression(10^-3),expression(10^-2),0.1,0.2,0.3,0.4,0.5,0.6,0.7,0.8,0.9,1), las=1)
axis(2, at=c(0:15), las=1)
# h2=0
points(1:length(d_lev),div_1000[,1,3,1],pch=19,col="black",cex=.7,type="b")
#arrows(1:length(d_lev), div_1000[,1,3,1], 1:length(d_lev), div_1000[,1,3,1]+div_1000[,1,3,3], length=0.05, angle=90, code=2, lwd=1, col="black")
#arrows(1:length(d_lev), div_1000[,1,3,1], 1:length(d_lev), div_1000[,1,3,1]-div_1000[,1,3,3], length=0.05, angle=90, code=2, lwd=1, col="black")
# loop across h2
for(i in 2:length(h_lev)){
  points(1:length(d_lev),div_1000[,i,3,1],pch=19,col="black",cex=.7,type="b")
  #arrows(1:length(d_lev),div_1000[,i,3,1],1:length(d_lev),div_1000[,i,3,1]+div_1000[,i,3,3],length=0.05, angle=90, code=2, lwd=1,col="black")
  #arrows(1:length(d_lev),div_1000[,i,3,1],1:length(d_lev),div_1000[,i,3,1]-div_1000[,i,3,3],length=0.05, angle=90, code=2, lwd=1, col="black")
}
```

### 2023-06-14 manuscript/manuscript_v04.Rmd
Moved mscript, added res file, began res walkthrough

--- 2023-06-07 manuscript_v04.Rmd
@@
 output:
-  pdf_document: default
   bookdown::pdf_document2: default

### 2023-06-14 manuscript/results_jp.Rmd
Moved mscript, added res file, began res walkthrough

---
title: "Walk-through of HMSC results"
author: "Jelena H. Pantel"
date: "`r file.mtime(knitr::current_input())`"
output:
  bookdown::pdf_document2: default
  html_document: default
  word_document:
    reference_docx: docx_template.docx
---

```{r setup,echo=FALSE,warning=FALSE,message=FALSE}
library(bookdown)
library(knitr)
library(here)
library(rmarkdown)
library(Hmsc)
library(beanplot)
library(bayesplot)
library(posterior)
```

The goal of our analysis is to estimate the relative effects of intrinsic population dynamics, environmental drivers, spatial structure, and trait evolution for driving community composition in the metacommunity. We used HMSC to build a linear Bayesian hierarchical model. The structure of the resulting Hmsc object is as follows:

```{r tidy=TRUE, tidy.opts=list(width.cutoff=60)}
datapath <- here::here("output")
result <- "h_01_d_minus3"  # Enter desired simulation conditions here
res <- load(paste(datapath,"/",result,"_hmsc.RData",sep=""))
```

For each of the 15 species, we have the following regression coefficient estimates:

* 1. intercept
* 2-16. the impacts of $N_{j,t-1}$ for $N_{i,t}$, for *j* = 1-15)
    + e.g. for species 1: 2 is the impact of $N_{1,t-1}$ for $N_{1,t}$
    + 3. the impact of $N_{2,t-1}$ for $N_{1,t}$
    + 4. the impact of $N_{3,t-1}$ for $N_{1,t}$
    + 5. the impact of $N_{4,t-1}$ for $N_{1,t}$
    + 6. the impact of $N_{5,t-1}$ for $N_{1,t}$
    + 7. the impact of $N_{6,t-1}$ for $N_{1,t}$
    + 8. the impact of $N_{7,t-1}$ for $N_{1,t}$
    + 9. the impact of $N_{8,t-1}$ for $N_{1,t}$
    + 10. the impact of $N_{9,t-1}$ for $N_{1,t}$
    + 11. the impact of $N_{10,t-1}$ for $N_{1,t}$
    + 12. the impact of $N_{11,t-1}$ for $N_{1,t}$
    + 13. the impact of $N_{12,t-1}$ for $N_{1,t}$
    + 14. the impact of $N_{13,t-1}$ for $N_{1,t}$
    + 15. the impact of $N_{14,t-1}$ for $N_{1,t}$
    + 16. the impact of $N_{15,t-1}$ for $N_{1,t}$
* 17. The impact of environment (*E*) for $N_{i,t}$
* 18-32. the impacts of $x_{j,t-1}$ for $N_{i,t}$, for *j* = 1-15)
    + e.g. for species 1: 18 is the impact of $x_{1,t-1}$ for $N_{1,t}$
    + 19. the impact of $x_{2,t-1}$ for $N_{1,t}$
    + 20. the impact of $x_{3,t-1}$ for $N_{1,t}$
    + 21. the impact of $x_{4,t-1}$ for $N_{1,t}$
    + 22. the impact of $x_{5,t-1}$ for $N_{1,t}$
    + 23. the impact of $x_{6,t-1}$ for $N_{1,t}$
    + 24. the impact of $x_{7,t-1}$ for $N_{1,t}$
    + 25. the impact of $x_{8,t-1}$ for $N_{1,t}$
    + 26. the impact of $x_{9,t-1}$ for $N_{1,t}$
    + 27. the impact of $x_{10,t-1}$ for $N_{1,t}$
    + 28. the impact of $x_{11,t-1}$ for $N_{1,t}$
    + 29. the impact of $x_{12,t-1}$ for $N_{1,t}$
    + 30. the impact of $x_{13,t-1}$ for $N_{1,t}$
    + 31. the impact of $x_{14,t-1}$ for $N_{1,t}$
    + 32. the impact of $x_{15,t-1}$ for $N_{1,t}$
* 33-47. the impacts of $x_{j,t-1} \cdot  N_{j,t-1}$  for $N_{i,t}$, for *j* = 1-15) $x_{1,t-1} \cdot  N_{1,t-1}$ for $N_{1,t}$
    + e.g. for species 1: 33 is the impact of $x_{1,t-1} \cdot  N_{1,t-1}$ for $N_{1,t}$
    + 34. the impact of $x_{2,t-1} \cdot  N_{2,t-1}$ for $N_{1,t}$
    + 35. the impact of $x_{3,t-1} \cdot  N_{3,t-1}$ for $N_{1,t}$
    + 36. the impact of $x_{4,t-1} \cdot  N_{4,t-1}$ for $N_{1,t}$
    + 37. the impact of $x_{5,t-1} \cdot  N_{5,t-1}$ for $N_{1,t}$
    + 38. the impact of $x_{6,t-1} \cdot  N_{6,t-1}$ for $N_{1,t}$
    + 39. the impact of $x_{7,t-1} \cdot  N_{7,t-1}$ for $N_{1,t}$
    + 40. the impact of $x_{8,t-1} \cdot  N_{8,t-1}$ for $N_{1,t}$
    + 41. the impact of $x_{9,t-1} \cdot  N_{9,t-1}$ for $N_{1,t}$
    + 42. the impact of $x_{10,t-1} \cdot  N_{10,t-1}$ for $N_{1,t}$
    + 43. the impact of $x_{11,t-1} \cdot  N_{11,t-1}$ for $N_{1,t}$
    + 44. the impact of $x_{12,t-1} \cdot  N_{12,t-1}$ for $N_{1,t}$
    + 45. the impact of $x_{13,t-1} \cdot  N_{13,t-1}$ for $N_{1,t}$
    + 46. the impact of $x_{14,t-1} \cdot  N_{14,t-1}$ for $N_{1,t}$
    + 47. the impact of $x_{15,t-1} \cdot  N_{15,t-1}$ for $N_{1,t}$

This repeats for all *j* = 15 species, i.e. coefficients 48-94 are for species 2, coefficients 95-141 are for species 3, etc. In order to evaluate results, we combine all iterations across all chains as a dataframe, and the predictor variables are indices of the dataframe columns (according to the numbering above).

```{r tidy=TRUE, tidy.opts=list(width.cutoff=60), figures-3, fig.show="hold", out.width="50%"}
m.post = Hmsc::convertToCodaObject(m.1)
m.df <- as.data.frame(rbind(m.post$Beta[[1]],m.post$Beta[[2]]))
## intercept
B.I <- m.df[,seq(1,705,by=47)]
beanplot::beanplot(B.I,log="",names=c("Sp1","Sp2","Sp3","Sp4","Sp5","Sp6","Sp7","Sp8","Sp9","Sp10","Sp11","Sp12","Sp13","Sp14","Sp15"),ylab=expression('N'["i,t"]),main="Intercept")

## Fixed effects: environment
B.E <- m.df[,seq(17,705,by=47)]
beanplot::beanplot(B.E,names=c("Sp1","Sp2","Sp3","Sp4","Sp5","Sp6","Sp7","Sp8","Sp9","Sp10","Sp11","Sp12","Sp13","Sp14","Sp15"),ylab=expression('N'["i,t"]),main=expression("Environment: " ~  beta ~ "*" ~ "E"[i]))
```

```{r tidy=TRUE, tidy.opts=list(width.cutoff=60), figures-3, fig.show="hold", out.width="50%"}
m.post = Hmsc::convertToCodaObject(m.1)
m.draw <- posterior::as_draws_array(m.post$Beta)
# Intercept
bayesplot::mcmc_trace(m.draw,pars=vars(seq(1,705,by=47)),size=0.1)
#mcmc_dens(m.draw,pars=vars(seq(17,705,by=47)))
bayesplot::mcmc_dens_overlay(m.draw,pars=vars(seq(1,705,by=47))) + yaxis_text(TRUE) + ylab("density")
# Environment
bayesplot::mcmc_trace(m.draw,pars=vars(seq(17,705,by=47)),size=0.1)
bayesplot::mcmc_dens_overlay(m.draw,pars=vars(seq(17,705,by=47))) + yaxis_text(TRUE) + ylab("density")
```

### 2023-12-06 manuscript_v04.Rmd

@@
 ---
-title: "Detection of eco-evolutionary dynamics in metacommunities using Joint Species Distribution Models"
-author: "Jelena H. Pantel, Ruben Hermann"
-date: Ecological Modelling, Faculty of Biology, University of Duisburg-Essen, Universitätsstraße 5, 45141 Essen, Germany
+title: "Detection of eco-evolutionary dynamics in metacommunities using Joint Species
+  Distribution Models"
+author: "Jelena H. Pantel, Ruben J. Hermann"
+date: "Ecological Modelling, Faculty of Biology, University of Duisburg-Essen, Universitätsstraße
+  5, 45141 Essen, Germany"
 output:
-  pdf_document: default
-  bookdown::pdf_document2: default
-  html_document: default
+  pdf_document:
+    toc: no
+  bookdown::html_document2: default
+  bookdown::pdf_document2:
+    toc: no
   word_document:
@@
 ```{r setup,echo=FALSE,warning=FALSE,message=FALSE}
-library(bookdown)
 library(R.matlab)
@@
 library(rsvg)
+
+library(gridExtra)
+library(tidyverse)
+library(patchwork)
+```
+```{r, include=FALSE}
+source("../code/R/rename.R")
+source("../code/R/post_fig.R")
 ```
@@
 At the same time, populations of individual species are evolving in ways that can be impacted by the same factors, as sites present different selection pressures and site connectivity can impact gene flow and genetic drift.
-Identifying the relative impacts of environmental, spatial, biotic, and other drivers of community composition across spatial and temporal scales has been greatly facilitated by joint species distribution models, but these models have not yet considered the role that evolution can play in driving changes in composition.
-I used Heirarchical Models of Species Communities (HMSC) to analyze data from a simulation model of an evolving metacommunity, in order to establish whether these models can sufficiently quantify the contribution of phenotypic evolution for metacommunity composition.
-The models successfully partitioned variance contributed by environmental, spatial, and phenotypic drivers, and also estimated site- and year-specific covariance.
-I also included estimates for eco-evolutionary interactions, and the model successfully estimated their role as well.
-The study of eco-evolutionary dynamics may require data that reflects numerous complex, interacting processes and its necessary to have flexible, generalized statistical models to analyze this data.
+Identifying the relative impacts of environmental, spatial, biotic, and other drivers of community composition across spatial and temporal scales has been greatly facilitated by joint species distribution models, but these models have not yet considered the role that microevolution can play in driving changes in composition.
+We used Heirarchical Models of Species Communities (HMSC) to analyze data from a simulation model of an evolving metacommunity, to establish whether these models can sufficiently quantify the contribution of phenotypic evolution for metacommunity composition.
+The models successfully partitioned variance contributed by environmental, spatial, and evolving phenotypic drivers, and also estimated site- and year-specific covariance.
+We also included estimates for eco-evolutionary interactions, and the model successfully estimated their role as well.
+The study of eco-evolutionary dynamics may require data that reflects numerous complex, interacting processes and it is necessary to have flexible, generalized statistical models to analyze this data.
 HMSC models present one promising path for analysis of eco-evolutionary dynamics in multi-species communities.
@@
 2014 [@Pollock2014]; Pichler & Hartig 2021 [@Pichler2021]) or multi-species occupancy models (REF; Devarajan et al.
-2020).
+2020 [@Devarajan2020]).
 These models can incorporate critical components that structure variance in biological data such as observer, measurement, and process error, error propagation, the existence of present but undetected species or life stages (REF) and other forms of imperfect detection (REF).
@@
 spOccupancy, Doser et al.
-2022), and there are intriguing extensions of these occupancy models with consideration of the error and data types particularly associated with eDNA (e.g.
-EDNAOCCUPANCY, Dorazio & Erickson 2018) and citizen science data (Altwegg & Nichols 2019).
+2022 [@Doser2022]), and there are intriguing extensions of these occupancy models with consideration of the error and data types particularly associated with eDNA (e.g.
+EDNAOCCUPANCY, Dorazio & Erickson 2018 [@Dorazio2018]) and citizen science data (Altwegg & Nichols 2019 [@Altwegg2019]).
 
 In addition to ecological processes, biodiversity at different levels of ecological organization – populations, communities, and ecosystems – are partially determined by the genotypic and phenotypic properties of organisms as well.
-These ecological processes can impact selection on genes and phenotypes, potentially resulting in dynamics feedback loops between ecological and evolutionary processes (Lion 2018; Barbour et al.
-2022).
+These ecological processes can impact selection on genes and phenotypes, potentially resulting in dynamics feedback loops between ecological and evolutionary processes (Lion 2018 [@Lion2018]; Barbour et al.
+2022 [@Barbour2022]).
 Although an increasing number of theoretical (REF) and empirical studies have evaluated the impacts of evolutionary dynamics for community assebmly and metacommunity dynamics (REF), analytical methods to consider this additional complexity lag behind the scope of data collected in eco-evolutionary studies.
 A series of methods papers (Hairston et al.
-2005; Ellner et al.
-2011; Govaert et al.
-2016) has made progress in utilizing an ANOVA framework to partition variance (in phenotypes or in other ecological response variables) across contributing ecological and evolutionary components.
+2005 [@Hairston2005]; Ellner et al.
+2011 [@Ellner2011]; Govaert et al.
+2016 [@Govaert2016]) has made progress in utilizing an ANOVA framework to partition variance (in phenotypes or in other ecological response variables) across contributing ecological and evolutionary components.
 These approaches have been applied to categorical ecological and evolutionary treatments with two levels.
 For example, the study of terHorst et al.
-(2014) evaluated the impacts of a categorical evolutionary treatment, where host plants were experimentally adapted to a dry or wet soil environment, and a categorical environmental treatment, where experiments were conducted in a dry or wet soil environment, for bacterial and fungal community richness and diversity.
-Hiltunen and Becks (2014) compared the relative importance of the (ecological) change in prey density and the (evolutionary) change in prey defense traits for the change in predator density from one time point to the next.
+(2014) [@terHorst2014] evaluated the impacts of a categorical evolutionary treatment, where host plants were experimentally adapted to a dry or wet soil environment, and a categorical environmental treatment, where experiments were conducted in a dry or wet soil environment, for bacterial and fungal community richness and diversity.
+Hiltunen and Becks (2014) [@Hiltunen2014] compared the relative importance of the (ecological) change in prey density and the (evolutionary) change in prey defense traits for the change in predator density from one time point to the next.
 However, these methods are ultimately limited by the same requirements that can limit application of ANOVA - data should follow a normal distribution, the processes that structure the target response variable must be linear and additive, and complicated structuring mechanisms must be reduced into one or a few target categories labeled as ‘ecology’ or ‘evolution’.
 Flexible models, which can vary depending on the mechanisms and structure of the diverse kinds of data that most studies of eco-evolutionary dynamics will collect, are needed for rigorous inference about eco-evolutionary processes.
-These models should consider the mechanistic basis of data structure, with realistic models of variance and uncertainty, and ideally can be used to make predictions about future states of the system.
+These models should consider the causal basis of data structure, with realistic models of variance and uncertainty, and ideally can be used to make predictions about future states of the system.
 
@@
 Lasky et al.
-(2020) developed an integrated reaction norm model, linking genetic, phenotypic, and demographic processes, and Benito Garzón et al.
-(2019) presented a species distribution model with local adaptation and phenotypic plasticity ($\Delta$ SDMs).
+(2020) [@Lasky2020] developed an integrated reaction norm model, linking genetic, phenotypic, and demographic processes, and Benito Garzón et al.
+(2019) [@Garzon2019] presented a species distribution model with local adaptation and phenotypic plasticity ($\Delta$ SDMs).
 The model of Lasky et al.
@@
 One particular metacommunity statistical model, the Hierarchical Modelling of Species Communities (HMSC) framework (Ovaskainen et al.
-2017; Tikhonov et al.
-2020), was built explicitly to consider the multitude of processes that can structure species occupancy and abundance data across time and space and is thus well suited to consider analysis of metacommunity eco-evolutionary data (Leibold et al.
-2022).
+2017 [@Ovaskainen2017HMSC]; Tikhonov et al.
+2020 [@Tikhonov2020]), was built explicitly to consider the multitude of processes that can structure species occupancy and abundance data across time and space (Leibold et al.
+2022 [@Leibold2022]) and is thus well suited to consider analysis of metacommunity eco-evolutionary data.
 This model of environmental filtering via variation and covariation in how species respond to their environment considers the impacts of traits and phylogenetic relationships, and uses latent variables to account for the numerous unobserved environmental and spatial features that are difficult to exhaustively measure in community data.
 The associated R package has immense flexibility and accompanying data analysis examples and code, and has been applied to study environmental responses of diverse assemblages of organisms (REF).
-In this study, I use a simulation of an evolving metacommunity model to generate realistic time series of population sizes and mean trait values for multiple species in a spatially and environmentally complex landscape.
-I then use HMSC to analyze the resulting data, to determine whether HMSC can successfully estimate the impacts of trait evolution for community composition, and to generate relative contributions of fixed environmental, spatial, and trait evolution drivers as well as random effects of spatial and temporal covariance among species.
-Finally, in order to determine whether HMSC is useful for hypothesis testing about the drivers of eco-evolutionary dynamics in communities in a landscape, I also conducted a few numerical experiments with the evolving metacommunity simulation - I varied the degree of environmental fluctuations in the landscape, and I varied the strength of species interactions, to determine whether these potential drivers of increased selection pressure alter the importance of evolution for metacommunity composition.
-For these experiments, I hypothesized that XX and YY.
-I found that XXXX.
+In this study, we use a simulation of an evolving metacommunity model to generate realistic time series of population sizes and mean trait values for multiple species in a spatially and environmentally complex landscape.
+We then use HMSC to analyze the resulting data, to determine whether HMSC can successfully estimate the impacts of trait evolution for community composition, and to generate relative contributions of fixed environmental, spatial, and trait evolution drivers as well as random effects of spatial and temporal covariance among species.
+Finally, in order to determine whether HMSC is useful for hypothesis testing about the drivers of eco-evolutionary dynamics in communities in a landscape, we also conducted numerical experiments with the evolving metacommunity simulation - we varied the degree of environmental fluctuations in the landscape and the strength of environmental selection, to determine whether these potential drivers altered the importance of evolution for metacommunity composition.
+We hypothesized these simulated conditions would have differing impacts on the relative importance of trait evolution for community composition.
+Increasing strength of selection should increase the speed of local adaptation and reduce genetic diversity, and therefore we expected a smaller detected impact of evolution for metacommunity composition in our system.
+We also expected that increasing the environmental variability would increase the impact of trait evolution, as the environment continuously presents a changing selection pressure that drives evolutionary change.
 
@@
 Population growth for species in the metacommunity simulation follows a Leslie-Gower model (a discrete-time version of a Lotka-Volterra model; Beverton & Holt 1957; Leslie & Gower 1958).
-I consider the impact of trait evolution for growth using a discrete time quantitative genetic model of evolutionary rescue (Gomulkiewicz & Holt 1995).
+We consider the impact of trait evolution for growth using a discrete time quantitative genetic model of evolutionary rescue (Gomulkiewicz & Holt 1995).
 The model for population size is as follows:
@@
 This model was used in Pantel & Becks (2023) to evaluate the consequences of adaptive evolution for coexistence in a three-species system.
-To expand this model to a metacommunity, I consider the evolution of multiple species (*S* = 15) in a landscape of patches (*k* = 50).
+To expand this model to a metacommunity, we consider the evolution of multiple species (*S* = 15) in a landscape of patches (*k* = 50).
 The patches have values of an environmental property *E* (drawn from a uniform distribution ranging from 0 to 1) that determines the local optimum phenotype *E*, i.e.
@@
 
-To simulate phenotypic evolution as well as species growth and competition in the landscape, I used parameter values of \(W_{max}\) = 2, *P* = 1, *w* = 10, $N_c$ = 100, and *p* = 0.001.
+To simulate phenotypic evolution as well as species growth and competition in the landscape, we used parameter values of \(W_{max}\) = 2, *P* = 1, *w* = 10, $N_c$ = 100, and *p* = 0.001.
 The simulated metacommunity was initialized in a way that mimics community assembly in small vernal pools that periodically flood (and thus are reset seasonally).
-In the simulated pool system, local communities are drawn at random from the regional species pool, and thus their initial degree of maladaptation in the pool is initially random as well.
+In the simulated pool system, local communities are drawn at random from the regional species pool, and thus their initial degree of maladaptation in the pool is random as well.
 The initial species richness of each site $s_{0k}$ was drawn from a random Poisson distribution, $s_{0k} \sim Pois(0.75)$, the initial population size for all species $N_{0i} \sim Pois(10)$, and the initial degree of maladaptation \(\beta_{i0} \sim Gamma(0.75,1)\).
-From this \(\beta_{i0}\), the initial distance to the local patch's optimum phenotype was calculated as $d_{i0}=\sqrt{B_{i0}(w+P)}$ and the initial trait value for each species $x_{i0}$ was calculated by subtracting $d_{i0}$ from the lowest site environmental value $E_k$.
+From this \(\beta_{i0}\), the initial distance to the local patch's optimum phenotype was calculated as $d_{i0}=\sqrt{B_{i0}(w+P)}$ and the initial trait value for each species $x_{i0}$ was calculated by subtracting $d_{i0}$ from the local environmental value $E_k$ (Figure \@ref(fig:figS1)).
 
 The species interaction matrix was fixed throughout the simulation, with \(\alpha_{ii} = 0.00125\) and \(\alpha_{ij} \sim U(0.0015,1)\).
-After draws for all random parameter values, the same resulting initial community was used for all simulations, to allow direct comparison across dispersal and heritability values (Figure \@ref(fig:figS1)).
+After draws for all random parameter values, the same resulting initial community was used for all simulations, to allow direct comparison across dispersal and heritability values.
 Simulations were run across a range of dispersal values (identical for all species; *d* = 0, $10^{-9},10^{-8},10^{-7},10^{-6},10^{-5},10^{-4},10^{-3},10^{-2}$, 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 1) and heritability values (identical for all species: $h^2$ = 0, 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 1).
@@
 
-To determine the relative influence of environmental and spatial properties, population dynamics, and evolutionary dynamics for community structure, I applied hierarchical modelling of species communities (HMSC, Ovaskainen et al.
+To determine the relative influence of environmental and spatial properties, population dynamics, and evolutionary dynamics for community structure, we applied hierarchical modelling of species communities (HMSC, Ovaskainen et al.
 2017; Tikhonov et al.
@@
 2020).
-The fixed effects were (1) $E_k$, the value of the site’s environmental parameter, (2) the value of each species' population size the time step before $N_{ik,t-1}$, (3) the absolute value of the change in each species' trait value from the previous time step to the next $|\Delta x_{ik,t-1 \to t}|$, (4) and an interaction term for population size by change in trait value $N_{ik,t-1} \times |\Delta x_{ik,t-1 \to t}|$, which led to a total of 46 predictors.
-The random effects were modeled at the site-level ($\epsilon_{ik} \sim N(0, \Omega^S)$, where $\Omega^S$ is the site species-to-species covariance matrix) and at the day level ($\epsilon_{it} \sim N(0, \Omega^T)$, where $\Omega^T$ is the time species-to-species covariance matrix) using a latent variable approach.
+The fixed effects were (1) $E_k$, the site’s environmental value, (2) the value of each species' population size the time step before $N_{ik,t-1}$, (3) the absolute change in each species' trait value from the previous time step to the next $|\Delta x_{ik,t-1 \to t}|$, (4) and an interaction term for population size by change in trait value $N_{ik,t-1} \times |\Delta x_{ik,t-1 \to t}|$, which led to a total of 46 predictors.
+The random effects were modeled at the site-level ($\epsilon_{ik} \sim N(0, \Omega^S)$, where $\Omega^S$ is the site species-to-species covariance matrix) and at the time step level ($\epsilon_{it} \sim N(0, \Omega^T)$, where $\Omega^T$ is the time species-to-species covariance matrix) using a latent variable approach.
 In this approach, species-to-species covariance matrices are computed as $\Omega = \lambda \intercal \lambda$, where $\lambda$ are the loadings of each species with latent variables, and latent variables are modeled as $\epsilon_{ik} = \sum_B \eta_{kB} \lambda_{Bi}$ and $\epsilon_{it} = \sum_B \eta_{tB} \lambda_{Bi}$ for *B* latent variables considered (Ovaskainen et al.
@@
 2020 [@Tikhonov2020]).
-Analysis and was conducted in R (version 4.1.1) using packages ‘Hmsc’, 'R.matlab', and XX (Bengtsson 2022; Tikhonov et al.
+Simulation results were implemented using 'R.matlab' (Bengtsson 2022), analysis and was conducted in R (version 4.1.1) using packages ‘Hmsc’, and XX (Tikhonov et al.
 2022; XX).
@@
 
+**Simulation model**
 The simulation model followed population size and trait value dynamics for species in the landscape across 1000 time steps, and results depended on the dispersal level and heritability values (Figure \@ref(fig:fig1)).
@@
 
+```{r, echo = FALSE, eval = TRUE, warning=FALSE, message=FALSE, results='hide'}
+setwd("../output")
+load("h_01_d_minus3_hmsc_v01.RData")
+s <- ncol(m.1$Y)
+npred <- ncol(m.1$XData)+1
+# Return the identity of the species that did not go extinct
+spec_id <- colnames(X)[grep("xy",colnames(X))] |> gsub(pattern="xy",replacement="") |> as.numeric()
+# Convert HMSC to MCMC object
+m.post = Hmsc::convertToCodaObject(m.1)
+m.df <- as.data.frame(rbind(m.post$Beta[[1]],m.post$Beta[[2]]))
+# Rename beta coefficients by group
+name <- rename(m.df,s,npred,spec_id)
+m.df <- name$m.df
+```
+
+**Statistical model**
+The goal of fitting the observed data to the HMSC model was to determine whether HMSC could successfully estimate the overall effects of four main drivers of population size for each of the species in the simulated metacommunity: environment, population size at the previous time step, evolution, thus change in species' trait $x$, and the interaction between population size and change in trait $x$.
+The full simulated time series was used as data to fit to the HMSC model, and therefore (1) the predictive power of the model was high ($R^2$ > 0.98 for all species, in all four model scenarios) and (2) using the population size at the previous time step was sufficient to explain almost of all the variation in the HMSC model (Appendix XX VarPart results).
+Instead of comparing the environmental, density, and evolutionary predictors on their importance for overall population size and explanatory power, we instead look at the values of the estimated effect size for scaled (to mean zero and unit variance) predictors for a single set of representative simulation runs, when dispersal $d = 10^{-3}$ and heritability $h^2 = 0.1$ for all species.
+
+There was no observed strong, consistent effect of changing the strength of selection (weak selection: $P = 1, w = 10$, strong selection: $P = 0.1, w = 0.2$) or the degree of environmental variability (no environmental variability: $v[E] = 0$, environmental variability: $v[E] = 0.01$) for the overall Intercept, the effects of environment, the intraspecific predictors (Figure \@ref(fig:fig4)), or the interspecific impacts of species densities on densities of other species (Figure \@ref(fig:fig5)).
+However, these differing simulation scenarios did impact the interspecific trait evolution predictors: scenarios with weak selection had a much stronger impact of $|\Delta x_{ik,t-1 \to t}|$ for abundances of other species, and scenarios without environmental variability had a much stronger impact of the interaction term for population size by change in trait value $N_{ik,t-1} \times |\Delta x_{ik,t-1 \to t}|$ (Figure \@ref(fig:fig5)).
+
+It is interesting to look in more depth at the impacts that trait evolution - quantified as the absolute value of the change in the population's trait value $x$ from one time step to the next ($|\Delta x_{ik,t-1 \to t}|$), have for the population sizes of other species ($y_i$).
+In Scenario 1 (weak selection, no environmental variability), we observe that $|\Delta x_{15}|$ has a very strong impact for $y_{13}$, with a posterior mean of `r mean(m.df$Bx13.15)` (and 95% HPDI, or 95% highest probability density interval = `r quantile(m.df$Bx13.15,probs=c(0.025,0.975))`).
+Figure 6 \@ref(fig:fig6) shows that high population densities of species 13 correlate with high trait changes of species 15 in all 5 patches, where both species co-exist.
+These indicate that species 15 had its greatest trait changes at the start of the simulation during which species 13 reached high population densities.
+As species 15 declines due to the increase of species 2, species 13 recovers again its population size (Fig.
+6b,h,j).
+Similarly we also observe a strong impact of $|\Delta x_{2}|$ for $y_{15}$, with a posterior mean of `r mean(m.df$Bx15.2)` (and 95% HPDI, or 95% highest probability density interval = `r quantile(m.df$Bx15.2,probs=c(0.025,0.975))`).
+It is important to note that interspecific impacts of $|\Delta x_{i}|$ are always indirect, as there is no direct interaction between $x_i$ and $y_j$.
+The only possible form of influence of is an increase in $y_i$ through evolution, leading to changes in intraspecific competition.
+We studied these posteriors in more detail, to provide an example on how to interpret these parameter estimations by the HMSC model (see Appendix XX).
+
+Another interesting result is seen when interpreting the interaction term included in the model.
+The model considers the change in trait from one time step to the next ($|\Delta x_{ik,t-1 \to t}|$), the population size at time $t-1$, $N_{ik,t-1}$, and a term for the interaction between these $N_{ik,t-1} \times |\Delta x_{ik,t-1 \to t}|$.
+The portion of the model that considers these predictors is as follows:
+
+$$ y_i = \beta_1 \cdot |\Delta x_{ik,t-1 \to t}| + \beta_2 \cdot N_{ik,t-1} + \beta_3 \cdot N_{ik,t-1} \cdot |\Delta x_{ik,t-1 \to t}| $$
+
+We can better understand these interaction effects by looking at a particular example, in this case for Scenario 1 (weak selection, no environmental variability).
+We look at Species 1, which has the following posterior estimates for the main and interaction terms:
+
+```{r, echo = FALSE, eval = TRUE}
+# Load data
+load("../output/h_01_d_minus3_hmsc_v01.RData")
+# Settings for number of species and their identity
+s <- ncol(m.1$Y)
+npred <- ncol(m.1$XData)+1
+# Return the identity of the species that did not go extinct
+spec_id <- colnames(X)[grep("xy",colnames(X))] |> gsub(pattern="xy",replacement="") |> as.numeric()
+# Convert HMSC to MCMC object
+m.post = Hmsc::convertToCodaObject(m.1)
+m.df <- as.data.frame(rbind(m.post$Beta[[1]],m.post$Beta[[2]]))
+# Rename beta coefficients by group
+name <- rename(m.df,s,npred,spec_id)
+m.df <- name$m.df
+## Extract parameters for Species 11
+i <- 1
+m.df.1 <- m.df[,((i-1)*npred+1):(npred*i)]
+# All beta coefficients where HPDI doesn't include 0 ####
+a <- apply(m.df.1,2,function(x) quantile(x,probs=c(0.025,0.975)))
+b <- apply(a,2,function(x) prod(x[1],x[2]) < 0)
+B.not0 <- m.df.1[,b==FALSE]
+
+tab <- data.frame(B0.1 = c(mean(m.df.1$B0.1),quantile(m.df.1$B0.1,probs=c(0.025,0.975))),
+              Bn.1= c(mean(m.df.1$Bn1.1),quantile(m.df.1$Bn1.1,probs=c(0.025,0.975))),
+              Bx1.1 = c(mean(m.df.1$Bx1.1),quantile(m.df.1$Bx1.1,probs=c(0.025,0.975))),
+              Bnx1.1 = c(mean(m.df.1$Bnx1.1),quantile(m.df.1$Bnx1.1,probs=c(0.025,0.975))))
+rownames(tab) <- c("mean", "2.5% HPDI","96.5% HPDI")
+
+knitr::kable(tab,digits=4)
+```
+
+This means that the absolute value of trait evolution $|\Delta x_{1,t-1 \to t}|$ had a positive impact on population size $N_{1,t}$, and the population size at the previous time step $N_{1,t-1}$ also had a small positive effect, but the effect of trait evolution depended on the population size (Figure \@ref(fig:fig7)).
+These interaction terms are interesting to interpret from a perspective of better understanding eco-evolutionary interactions (where the impacts of evolution for population size depend on the value of population size).
+Figure (Figure \@ref(fig:fig5)) indicates stronger interaction effects in the weak selection scenarios (1 and 2).
+We can look in more depth at the interaction effects and evaluate whether they are additive, synergistic, or antagonistic.
+We calculated posterior distributions for $\beta_3 - (\beta_1 + \beta_2)$ and plot these in Figure \@ref(fig:fig8).
+Values with negative HPDI are antagonistic, with HPDI overlapping 0 are additive, and with positive HPDI are synergistic.
+We observed substantially more antagonisic interactions of population size and trait evolution in the scenarios with weak selection (Scenarios 1 and 3).
+
 Discussion
@@
 
+STOPPED HERE.
 I have analysed data collected by Herman Bumpus [@Bumpus1898] on the relationship between sparrow (*Passer domesticus*) total length and surival following an unusually severe storm.
@@
 --->
+When interpreting estimated posteriors of the HMSC model, for the different parameters, it is important to consider that these are based on correlations.
+Such a correlation may be the result of a direct impact from one parameter, the effect of an unmeasured parameter or the result of different parameter interacting.
+Therefore, when interpreting their effects on the communitiy processes they need to be investigated further.
+As an example for this we examined the higher posterior of $|\Delta x_{15}|$ on $y_{13}$ and of $|\Delta x_{2}|$ on $y_{15}$ (for further detail see Appendix XX).
+In this example we were able to show that the effect of $|\Delta x_{2}|$ on $y_{15}$ was a direct consequence of the strong competition coefficient between both species and as species 2 was the stronger competitor (through higher population size) it led to the decline of $y_{15}$.
+At the same time the effect of $|\Delta x_{15}|$ on $y_{13}$ was the result of a correlation without causation, as species 13 has no strong competition with species 15 (or 2).
+Therefore, before concluding effects of the estimated posteriors these parameters need to be further analyzed.
+Furthermore, results from species with low densities need to be interpreted with care as the low population size might inflate the resulting posterior.
 
@@
 h_lev <- c("h_0_","h_01_","h_02_","h_03_","h_04_","h_05_","h_06_","h_07_","h_08_","h_09_","h_10_")
-setwd("data/")
+setwd("../data")
 result <- paste(h_lev[1],d_lev[1],"_res.mat",sep="")
@@
 
-setwd("data/")
+setwd("../data")
 for(i in 1:length(d_lev)){
@@
 
-```{r fig3, echo=FALSE, fig.show="hold", out.width="25%", fig.cap = "(A) Plot with among-site similarity plotted across dispersal levels for $h^2$=0 and $h^2$=0.1, for dispersal levels XXX.
+```{r fig3, echo=FALSE, fig.show="hold", out.width="40%", fig.cap = "(A) Plot with among-site similarity plotted across dispersal levels for $h^2$=0 and $h^2$=0.1, for dispersal levels XXX.
 (B) Plot with average among-site similarity (error bars at ± 1 standard error are given but very small) across $h2$ levels."}
-setwd("data/")
+setwd("../data")
 ## Network similarity plot
@@
 
+```{r fig4, echo=FALSE, fig.show="hold", out.width="100%", fig.cap = "Plot with posterior effect size estimates for regression coefficients for (a,f,k,p) Intercept, (b,g,l,q) environment, (c,h,m,r) Intraspecific population size, (d,i,n,s) Intraspecific delta X, and (e,j,o,t) Intraspecific population size X delta X.
+Row 1 is scenario 1 (weak selection, no environmental variability), row 2 is scenario 2 (strong selection, no environmental variability), row 3 is scenario 2 (weak selection, environmental variability), and row 4 is scenario 4 (strong selection, no environmental variability)."}
+# Load data
+load("../output/h_01_d_minus3_hmsc_v01.RData")
+# Settings for number of species and their identity
+s <- ncol(m.1$Y)
+npred <- ncol(m.1$XData)+1
+# Return the identity of the species that did not go extinct
+spec_id <- colnames(X)[grep("xy",colnames(X))] |> gsub(pattern="xy",replacement="") |> as.numeric()
+# Convert HMSC to MCMC object
+m.post = Hmsc::convertToCodaObject(m.1)
+m.df <- as.data.frame(rbind(m.post$Beta[[1]],m.post$Beta[[2]]))
+# Rename beta coefficients by group
+name <- rename(m.df,s,npred,spec_id)
+m.df <- name$m.df
+# Call plotting function and return list for this scenario
+s1 <- post_fig(m.df,s)
+## Repeat for the other 3 scenarios
+# Scenario 2
+load("../output/h_01_d_minus3_hmsc_v02.RData")
+s <- ncol(m.1$Y)
+npred <- ncol(m.1$XData)+1
+spec_id <- colnames(X)[grep("xy",colnames(X))] |> gsub(pattern="xy",replacement="") |> as.numeric()
+m.post = Hmsc::convertToCodaObject(m.1)
+m.df <- as.data.frame(rbind(m.post$Beta[[1]],m.post$Beta[[2]]))
+name <- rename(m.df,s,npred,spec_id)
+m.df <- name$m.df
+s2 <- post_fig(m.df,s)
+# Scenario 3
+load("../output/h_01_d_minus3_hmsc_v03.RData")
+s <- ncol(m.1$Y)
+npred <- ncol(m.1$XData)+1
+spec_id <- colnames(X)[grep("xy",colnames(X))] |> gsub(pattern="xy",replacement="") |> as.numeric()
+m.post = Hmsc::convertToCodaObject(m.1)
+m.df <- as.data.frame(rbind(m.post$Beta[[1]],m.post$Beta[[2]]))
+name <- rename(m.df,s,npred,spec_id)
+m.df <- name$m.df
+s3 <- post_fig(m.df,s)
+# Scenario 4
+load("../output/h_01_d_minus3_hmsc_v04.RData")
+s <- ncol(m.1$Y)
+npred <- ncol(m.1$XData)+1
+spec_id <- colnames(X)[grep("xy",colnames(X))] |> gsub(pattern="xy",replacement="") |> as.numeric()
+m.post = Hmsc::convertToCodaObject(m.1)
+m.df <- as.data.frame(rbind(m.post$Beta[[1]],m.post$Beta[[2]]))
+name <- rename(m.df,s,npred,spec_id)
+m.df <- name$m.df
+s4 <- post_fig(m.df,s)
+## Arrange plots
+#gridExtra::grid.arrange(s1[[1]],s2[[1]],s3[[1]],s4[[1]],s1[[2]],s2[[2]],s3[[2]],s4[[2]],s1[[3]],s2[[3]],s3[[3]],s4[[3]],s1[[5]],s2[[5]],s3[[5]],s4[[5]],s1[[7]],s2[[7]],s3[[7]],s4[[7]],ncol=5,nrow=4,layout_matrix=rbind(c(1,5,9,13,17),c(2,6,10,14,18),c(3,7,11,15,19),c(4,8,12,16,20)))
+cowplot::plot_grid(s1[[1]],s2[[1]],s3[[1]],s4[[1]],s1[[2]],s2[[2]],s3[[2]],s4[[2]],s1[[3]],s2[[3]],s3[[3]],s4[[3]],s1[[5]],s2[[5]],s3[[5]],s4[[5]],s1[[7]],s2[[7]],s3[[7]],s4[[7]], labels=c("a","b","c","d","e","f","g","h","i","j","k","l","m","n","o","p","q","r","s","t"), ncol=5,nrow=4,byrow=FALSE,label_size=6)
+```
+
+\newpage
+
+```{r fig5, echo=FALSE, fig.show="hold", out.width="100%", fig.cap = "Plot with posterior effect size estimates for regression coefficients for (a,d,g,j) interpecific population size, (b,e,h,k) interspecific delta X, and (c,f,i,l) interspecific population size X delta X.
+Row 1 is scenario 1, row 2 is scenario 2, row 3 is scenario 3, and row 4 is scenario 4."}
+## Arrange plots
+#gridExtra::grid.arrange(s1[[4]],s2[[4]],s3[[4]],s4[[4]],s1[[6]],s2[[6]],s3[[6]],s4[[6]],s1[[8]],s2[[8]],s3[[8]],s4[[8]],ncol=3,nrow=4,layout_matrix=rbind(c(1,5,9),c(2,6,10),c(3,7,11),c(4,8,12)))
+cowplot::plot_grid(s1[[4]],s2[[4]],s3[[4]],s4[[4]],s1[[6]],s2[[6]],s3[[6]],s4[[6]],s1[[8]],s2[[8]],s3[[8]],s4[[8]],ncol=3,nrow=4,byrow=FALSE,labels=c("a","b","c","d","e","f","g","h","i","j","k","l"),label_size=6)
+```
+
+\newpage
+
+```{r fig6, echo=FALSE, fig.show="hold", warning=FALSE,out.width="90%", fig.cap = "Plot showing interpretation of impact of DeltaX-15 for Y-13."}
+# Load data
+res  <- R.matlab::readMat(paste("../data/",result,"_res.mat",sep="")) # Scenario 1
+# Identify sites where Species 15 and 13 co-occur
+pa <- res$N
+pa[pa > 0] <- 1
+pa <- pa[,c(13,15),]
+a <- apply(pa,1,function(x) colSums(x) ==2)
+# To get sites where both species co-occur at the same time
+b <- which(colSums(a) > 0)
+###Creating a 3d arrays to easier manipulate the data
+#For the X data
+X3d <-- array(numeric(),c(50,34,999)) 
+for (i in 0:998){
+  X3d[,,(i+1)] = as.matrix(X[((i*50+1):((i+1)*50)),])
+}
+colnames(X3d) <- colnames(X)
+#For the Y data
+Y3d <-- array(numeric(),c(50,11,999)) 
+for (i in 0:998){
+  Y3d[,,(i+1)] = as.matrix(Y[((i*50+1):((i+1)*50)),])
+}
+# Plot
+patches <- list()
+patches_l<-list()
+p_X<-list()
+p_Y<-list()
+for (i in 1:length(b)) {
+patches[[i]] <- data.frame(x15=X3d[b[i],23,],y02=Y3d[b[i],2,],y13=Y3d[b[i],9,],y15=Y3d[b[i],11,],time=2:1000)  
+patches_l[[i]] <- pivot_longer(patches[[i]],cols=2:4,values_to="Y",names_to = "sp")
+p_X[[i]] <-ggplot(patches_l[[i]],aes(x=x15,y=Y,color=sp,linetype=sp))+
+  geom_point(size=2,alpha=0.5)+
+  ylim(c(0,700))+
+  scale_color_manual(values = c("coral3","darkorchid4","rosybrown1"))+
+  theme_bw()+
+  xlab(NULL)+ylab(NULL)+
+  theme()+
+  guides(color = FALSE,linetype=F)
+
+p_Y[[i]]<-ggplot(patches_l[[i]],aes(x=time,y=Y,color=sp,linetype=sp))+
+  geom_line(linewidth=2)+
+  ylim(c(0,700))+
+  scale_color_manual(values = c("coral3","darkorchid4","rosybrown1"))+
+  theme_bw()+
+  ylab(NULL)+xlab(NULL)+
+  theme(axis.text.y = element_blank(),axis.text.x = element_blank())+
+  guides(color = FALSE,linetype=F)
+}
+
+p_X[[1]]+p_X[[2]]+p_X[[3]]+ylab("Specis density")+
+  p_X[[4]]+p_X[[5]]+xlab("Delta x of Species 15")+
+  p_Y[[1]]+p_Y[[2]]+p_Y[[3]]+p_Y[[4]]+p_Y[[5]]+xlab("Time")+
+  theme(axis.text.x=element_text())+
+  scale_x_continuous(breaks = c(0,250,500,750,1000))+
+  plot_layout(nrow=5,byrow=F)+
+  plot_annotation(tag_levels = list(c("a","c","e","g","i","b","d","f","h","j")))
+```
+
+\newpage
+
+```{r fig7, echo=FALSE, fig.show="hold", out.width="60%", fig.cap = "Interpretation of interaction effect for Species 1, Scenario 1.
+Model-predicted value for $N_{1t}$ across a range of $|deltaX|$ values, using the mean (black line), 2.5% (orange lines), and 97.5% (green lines) values of $N_{1,t-1}$.
+Observed values from the simulation are shown as grey points."}
+# 1 generate posterior predictions for y2 ~ y_t-1_1
+m.df.1 <- m.df.1[,c(1,2,14,25)]
+# range of observed Y_t-1_2.2, after scaling
+Yt1_1.1 <- scale(m.1$X[,"y1"])
+Yt1_1.1_range <- sample(Yt1_1.1,100,replace=FALSE)
+Yt1_1.1_range2 <- seq(from=min(Yt1_1.1),to=max(Yt1_1.1),length.out=100)
+Yx_1.1 <- scale(m.1$X[,"x1"])
+Yx_1.1_range <- sample(Yx_1.1,100,replace=FALSE)
+Yx_1.1_range2 <- seq(from=min(Yx_1.1),to=max(Yx_1.1),length.out=100)
+Yt1X_1.1 <- scale(m.1$X[,"xy1"])
+# calculate predicted values for Y1 across range of Yx_1.1
+yhat_no_int <- apply(m.df.1,1,function(x) (x[1] + Yx_1.1_range2*x[3]))
+# Plot mean and 95% HPDI lines for this regression
+center <- mean(m.df.1[,1]) + Yx_1.1_range2*mean(m.df.1[,3])
+# Plot mean and 95% HPDI lines for this regression
+# Calculate predicted values for Y1 across range of Yt1_1.1
+vals <- quantile(Yt1_1.1_range2,probs=c(.025,0.5,.975))
+cred_025 <- apply(m.df.1,1,function(x) (x[1] + Yx_1.1_range2*x[3] + (vals[1])*x[2]) + (Yx_1.1_range2*vals[1]*x[4]))
+cred_975 <- apply(m.df.1,1,function(x) (x[1] + Yx_1.1_range2*x[3] + (vals[3])*x[2]) + (Yx_1.1_range2*vals[3]*x[4]))
+center <- mean(m.df.1[,1]) + Yx_1.1_range2*mean(m.df.1[,3]) + vals[2]*mean(m.df.1[,2]) + (Yx_1.1_range2*vals[2]*mean(m.df.1[,4]))
+# Plot
+plot(Yx_1.1_range2,center,type="l",col="black",xlim=c(min(Yt1_1.1),max(Yt1_1.1)),ylim=c(0,12),xlab="deltaX_1.1",ylab="predicted Yt_1")
+samp <- sample(1:2000,50)
+for(i in 1:50){
+  lines(Yx_1.1_range2,cred_025[,samp[i]],col="orange",lwd=0.2)
+}
+samp <- sample(1:2000,50)
+for(i in 1:50){
+  lines(Yx_1.1_range2,cred_975[,samp[i]],col="green",lwd=0.2)
+}
+## Add overall mean and observed values
+lines(Yx_1.1_range2,center,col="black")
+points(Yt1_1.1,scale(m.1$Y[,"y1"]),cex=.3,pch=19,col="lightgrey")
+```
+
+\newpage
+
+```{r fig8, echo=FALSE, fig.show="hold", out.width="60%", fig.cap = "Antagonistic (HPDI < 0), additive (HPDI includes 0), or synergistic (HPDI > 0) eco-evolutionary interaction effects for (a) scenario 1 (weak selection, no environmental variability), (b) scenario 2 (strong selection, no environmental variability), (c) scenario 3 (weak selection, environmental variability), and (d) scenario 4 (strong selection, environmental variability)."}
+# Load data
+load("../output/h_01_d_minus3_hmsc_v01.RData")
+# Settings for number of species and their identity
+s <- ncol(m.1$Y)
+npred <- ncol(m.1$XData)+1
+# Return the identity of the species that did not go extinct
+spec_id <- colnames(X)[grep("xy",colnames(X))] |> gsub(pattern="xy",replacement="") |> as.numeric()
+# Convert HMSC to MCMC object
+m.post = Hmsc::convertToCodaObject(m.1)
+m.df <- as.data.frame(rbind(m.post$Beta[[1]],m.post$Beta[[2]]))
+# Rename beta coefficients by group
+name <- rename(m.df,s,npred,spec_id)
+m.df <- name$m.df
+### New code from this point forwards ###
+int <- cbind(colnames(m.df)[name$Nt1],colnames(m.df)[name$NtdX],colnames(m.df)[name$Nt1dX])
+post_add <- array(NA,dim=c(nrow(m.df),nrow(int)))
+for(i in 1:nrow(int)){
+  add <- as.data.frame(m.df[int[i,]])
+  post_add[,i] <- add[,3] - (add[,1] + add[,2])
+}
+post_add <- data.frame(post_add)
+colnames(post_add) <- colnames(m.df)[name$Nt1dX]
+add <- post_add[,order(colMeans(post_add))]
+plist <- list()
+plist[[1]] <- bayesplot::mcmc_intervals(add,point_size=2) +
+    ggplot2::coord_cartesian(xlim=c(-4,0.8)) + ggplot2::theme(text = ggplot2::element_text(size=6))
+## Repeat for the other 3 scenarios
+# Scenario 2
+load("../output/h_01_d_minus3_hmsc_v02.RData")
+s <- ncol(m.1$Y)
+npred <- ncol(m.1$XData)+1
+spec_id <- colnames(X)[grep("xy",colnames(X))] |> gsub(pattern="xy",replacement="") |> as.numeric()
+m.post = Hmsc::convertToCodaObject(m.1)
+m.df <- as.data.frame(rbind(m.post$Beta[[1]],m.post$Beta[[2]]))
+name <- rename(m.df,s,npred,spec_id)
+m.df <- name$m.df
+### New code from this point forwards ###
+int <- cbind(colnames(m.df)[name$Nt1],colnames(m.df)[name$NtdX],colnames(m.df)[name$Nt1dX])
+post_add <- array(NA,dim=c(nrow(m.df),nrow(int)))
+for(i in 1:nrow(int)){
+  add <- as.data.frame(m.df[int[i,]])
+  post_add[,i] <- add[,3] - (add[,1] + add[,2])
+}
+post_add <- data.frame(post_add)
+colnames(post_add) <- colnames(m.df)[name$Nt1dX]
+add <- post_add[,order(colMeans(post_add))]
+plist[[2]] <- bayesplot::mcmc_intervals(add,point_size=2) +
+    ggplot2::coord_cartesian(xlim=c(-4,0.8)) + ggplot2::theme(text = ggplot2::element_text(size=6))
+# Scenario 3
+load("../output/h_01_d_minus3_hmsc_v03.RData")
+s <- ncol(m.1$Y)
+npred <- ncol(m.1$XData)+1
+spec_id <- colnames(X)[grep("xy",colnames(X))] |> gsub(pattern="xy",replacement="") |> as.numeric()
+m.post = Hmsc::convertToCodaObject(m.1)
+m.df <- as.data.frame(rbind(m.post$Beta[[1]],m.post$Beta[[2]]))
+name <- rename(m.df,s,npred,spec_id)
+m.df <- name$m.df
+### New code from this point forwards ###
+int <- cbind(colnames(m.df)[name$Nt1],colnames(m.df)[name$NtdX],colnames(m.df)[name$Nt1dX])
+post_add <- array(NA,dim=c(nrow(m.df),nrow(int)))
+for(i in 1:nrow(int)){
+  add <- as.data.frame(m.df[int[i,]])
+  post_add[,i] <- add[,3] - (add[,1] + add[,2])
+}
+post_add <- data.frame(post_add)
+colnames(post_add) <- colnames(m.df)[name$Nt1dX]
+add <- post_add[,order(colMeans(post_add))]
+plist[[3]] <- bayesplot::mcmc_intervals(add,point_size=2) +
+    ggplot2::coord_cartesian(xlim=c(-4,0.8)) + ggplot2::theme(text = ggplot2::element_text(size=6))
+# Scenario 4
+load("../output/h_01_d_minus3_hmsc_v04.RData")
+s <- ncol(m.1$Y)
+npred <- ncol(m.1$XData)+1
+spec_id <- colnames(X)[grep("xy",colnames(X))] |> gsub(pattern="xy",replacement="") |> as.numeric()
+m.post = Hmsc::convertToCodaObject(m.1)
+m.df <- as.data.frame(rbind(m.post$Beta[[1]],m.post$Beta[[2]]))
+name <- rename(m.df,s,npred,spec_id)
+m.df <- name$m.df
+### New code from this point forwards ###
+int <- cbind(colnames(m.df)[name$Nt1],colnames(m.df)[name$NtdX],colnames(m.df)[name$Nt1dX])
+post_add <- array(NA,dim=c(nrow(m.df),nrow(int)))
+for(i in 1:nrow(int)){
+  add <- as.data.frame(m.df[int[i,]])
+  post_add[,i] <- add[,3] - (add[,1] + add[,2])
+}
+post_add <- data.frame(post_add)
+colnames(post_add) <- colnames(m.df)[name$Nt1dX]
+add <- post_add[,order(colMeans(post_add))]
+plist[[4]] <- bayesplot::mcmc_intervals(add,point_size=2) +
+    ggplot2::coord_cartesian(xlim=c(-4,0.8)) + ggplot2::theme(text = ggplot2::element_text(size=6))
+
+gridExtra::grid.arrange(plist[[1]],plist[[2]],plist[[3]],plist[[4]],ncol=1,nrow=4,layout_matrix=rbind(1,2,3,4))
+```
+
+\newpage
+
 <a name="appendix">Appendix 1</a>
@@
 ## Figure S1. Plot of location of sites in landscape, associated E values, and associated diversity values
-setwd("data/")
+setwd("../data")
 result <- paste(h_lev[2],d_lev[8],"_res.mat",sep="")
@@
 
-```{r figS2, echo=FALSE, fig.show="hold", fig.cap = "Plots with (A) gamma, (B) alpha, and (C) beta diversity across levels of dispersal (x-axis), for values of heritability ranging from 0-1."}
-setwd("data/")
+```{r figS2, echo=FALSE, fig.show="hold", fig.cap = "Plots of diversity (y-axis) across dispersal levels (x-axis) for simulations at different levels of heritability ($h^2$=0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 1).
+(a) Regional ($\\gamma$), (b) local ($\\alpha$), and (c) among-site ($\\beta$) diversity values are shown."}
+# Figure S2. Plot of diversity (y-axis) across dispersal levels (x-axis) for simulations at different levels of heritability (h2).
+setwd("../data")
 ## Inverse Simpson's diversity across additional heritability levels
@@
 ```
+
+\newpage
+
+<a name="appendix">Appendix 2</a>
+================================================================================
+
+The goal of this appendix is to better interpret the interaction effects.

### 2023-12-06 manuscript_v04.Rmd
attempt to upload offline v05 for change comparison

@@
   Distribution Models"
-author: "Jelena H. Pantel, Ruben J. Hermann"
+author: "Jelena H. Pantel, Ruben Hermann"
 date: "Ecological Modelling, Faculty of Biology, University of Duisburg-Essen, Universitätsstraße
@@
 output:
-  pdf_document:
-    toc: no
-  bookdown::html_document2: default
   bookdown::pdf_document2:
     toc: no
+    fig_caption: yes
+    latex_engine: xelatex
+  bookdown::html_document2: default
   word_document:
@@
 - \pagestyle{fancy}
-- \fancyhead[L]{Pantel, JH}
+- \fancyhead[L]{Pantel and Hermann}
 - \fancyhead[R]{Detection of ecoevo dynamics}
@@
 source("../code/R/post_fig.R")
+source("../code/R/get_deltaX.R")
+source("../code/R/get_div.R")
 ```
@@
 Biodiversity at the metacommunity scale is typically influenced by a number of environmental, spatial, biotic, and stochastic factors.
-At the same time, populations of individual species are evolving in ways that can be impacted by the same factors, as sites present different selection pressures and site connectivity can impact gene flow and genetic drift.
+At the same time, populations of individual species can evolve and this process can be impacted by the same factors, as sites present different selection pressures and site connectivity can impact gene flow and genetic drift.
 Identifying the relative impacts of environmental, spatial, biotic, and other drivers of community composition across spatial and temporal scales has been greatly facilitated by joint species distribution models, but these models have not yet considered the role that microevolution can play in driving changes in composition.
@@
 The initial species richness of each site $s_{0k}$ was drawn from a random Poisson distribution, $s_{0k} \sim Pois(0.75)$, the initial population size for all species $N_{0i} \sim Pois(10)$, and the initial degree of maladaptation \(\beta_{i0} \sim Gamma(0.75,1)\).
-From this \(\beta_{i0}\), the initial distance to the local patch's optimum phenotype was calculated as $d_{i0}=\sqrt{B_{i0}(w+P)}$ and the initial trait value for each species $x_{i0}$ was calculated by subtracting $d_{i0}$ from the local environmental value $E_k$ (Figure \@ref(fig:figS1)).
+From this \(\beta_{i0}\), the initial distance to the local patch's optimum phenotype was calculated as $d_{i0}=\sqrt{B_{i0}(w+P)}$ and the initial trait value for each species $x_{i0}$ was calculated by subtracting $d_{i0}$ from the lowest site environmental value $E_k$ (Figure \@ref(fig:figS1)).
 
@@
 The simulations were run in MATLAB (version R2022a) for 1000 time steps and produced records of $N_{ikt}$ and $x_{ikt}$.
+To test the impacts of strength of selection and environmental variability, we repeated simulations at a single, representative level of heritability and dispersal ($d = 10^{-3}$, $h^2 = 0.1$) in four different scenarios: Scenario 1 (weak selection, fixed environment: $P = 1, w = 10, v[E] = 0$), Scenario 2 (strong selection, fixed environment: $P = 0.1, w = 0.2, v[E] = 0$), Scenario 3 (weak selection, variable environment: $P = 1, w = 10, v[E] = 0.01$), and Scenario 4 (strong selection, variable environment: P = 0.1, w = 0.2, v[E] = 0.01).
 
@@
 2020).
-The fixed effects were (1) $E_k$, the site’s environmental value, (2) the value of each species' population size the time step before $N_{ik,t-1}$, (3) the absolute change in each species' trait value from the previous time step to the next $|\Delta x_{ik,t-1 \to t}|$, (4) and an interaction term for population size by change in trait value $N_{ik,t-1} \times |\Delta x_{ik,t-1 \to t}|$, which led to a total of 46 predictors.
+The fixed effects were (1) $E_k$, the value of the site’s environmental parameter, (2) the value of each species' population size the time step before $N_{ik,t-1}$, (3) the absolute value of the change in each species' trait value from the previous time step to the next $|\Delta x_{ik,t-1 \to t}|$, (4) and an interaction term for population size by change in trait value $N_{ik,t-1} \times |\Delta x_{ik,t-1 \to t}|$, which led to a total of 46 predictors.
 The random effects were modeled at the site-level ($\epsilon_{ik} \sim N(0, \Omega^S)$, where $\Omega^S$ is the site species-to-species covariance matrix) and at the time step level ($\epsilon_{it} \sim N(0, \Omega^T)$, where $\Omega^T$ is the time species-to-species covariance matrix) using a latent variable approach.
@@
 2020 [@Tikhonov2020]).
-Simulation results were implemented using 'R.matlab' (Bengtsson 2022), analysis and was conducted in R (version 4.1.1) using packages ‘Hmsc’, and XX (Tikhonov et al.
+Analysis and was conducted in R (version 4.1.1) using packages ‘Hmsc’, 'R.matlab', and XX (Bengtsson 2022; Tikhonov et al.
 2022; XX).
@@
 **Statistical model**
-The goal of fitting the observed data to the HMSC model was to determine whether HMSC could successfully estimate the overall effects of four main drivers of population size for each of the species in the simulated metacommunity: environment, population size at the previous time step, evolution, thus change in species' trait $x$, and the interaction between population size and change in trait $x$.
-The full simulated time series was used as data to fit to the HMSC model, and therefore (1) the predictive power of the model was high ($R^2$ > 0.98 for all species, in all four model scenarios) and (2) using the population size at the previous time step was sufficient to explain almost of all the variation in the HMSC model (Appendix XX VarPart results).
-Instead of comparing the environmental, density, and evolutionary predictors on their importance for overall population size and explanatory power, we instead look at the values of the estimated effect size for scaled (to mean zero and unit variance) predictors for a single set of representative simulation runs, when dispersal $d = 10^{-3}$ and heritability $h^2 = 0.1$ for all species.
-
-There was no observed strong, consistent effect of changing the strength of selection (weak selection: $P = 1, w = 10$, strong selection: $P = 0.1, w = 0.2$) or the degree of environmental variability (no environmental variability: $v[E] = 0$, environmental variability: $v[E] = 0.01$) for the overall Intercept, the effects of environment, the intraspecific predictors (Figure \@ref(fig:fig4)), or the interspecific impacts of species densities on densities of other species (Figure \@ref(fig:fig5)).
-However, these differing simulation scenarios did impact the interspecific trait evolution predictors: scenarios with weak selection had a much stronger impact of $|\Delta x_{ik,t-1 \to t}|$ for abundances of other species, and scenarios without environmental variability had a much stronger impact of the interaction term for population size by change in trait value $N_{ik,t-1} \times |\Delta x_{ik,t-1 \to t}|$ (Figure \@ref(fig:fig5)).
+The goal of fitting the observed data to the HMSC model was to determine whether HMSC could successfully estimate the overall effects of four main drivers of population size for each of the species in the simulated metacommunity: environment, population size at the previous time step, evolution in species' trait $x$, and the interaction between population size and change in trait $x$.
+The full simulated time series was used as data to fit to the HMSC model, and therefore the (1) the predictive power of the model was high ($R^2$ > 0.98 for all species, in all four model scenarios) and (2) using the population size at the previous time step was sufficient to explain almost of all the variation in the HMSC model (Appendix XX VarPart results).
+Instead of comparing the environmental, biotic, and evolutionary predictors on their importance for overall population size and explanatory power, we instead look at the values of the estimated effect size for scaled (to mean zero and unit variance) predictors for a single set of representative simulation runs, when dispersal $d = 10^{-3}$ and heritability $h^2 = 0.1$ for all species.
+
+There was no observed strong, consistent effect of changing the strength of selection (weak selection: $P = 1, w = 10$, strong selection: $P = 0.1, w = 0.2$) or the degree of environmental variability (no environmental variability: $v[E] = 0$, environmental variability: $v[E] = 0.01$) for the overall Intercept, the effects of environment, the intraspecific predictors (Figure \@ref(fig:fig4)), or the interspecific impacts of species abundances on abundances of other species (Figure \@ref(fig:fig5)).
+However, these differing simulation scenarios did impact the interspecific trait evolution predictors: scenarios with weak selection (1,3) had a much stronger impact of $|\Delta x_{ik,t-1 \to t}|$ for abundances of other species, and scenarios without environmental variability (1,2) had a much stronger impact of the interaction term for population size by change in trait value $N_{ik,t-1} \times |\Delta x_{ik,t-1 \to t}|$ (Figure \@ref(fig:fig5)).
+These results can be explained by evaluating the patterns of change in species composition ($\beta$ diversity) and change in trait values ($|\Delta x|$) over the course of the simulation time series.
+Scenarios with strong selection (2 and 4) experienced more rapid loss of locally maladapted species during the initial 'community assembly' phase of the simulation (see in the strong initial drop in $\beta$ diversity in Figure \@ref(fig:fig7)).
+This loss of maladapted species meant there was less opportunity for local adaptation occurring in the same time frame as the most substantial shifts in species abundances (Figure \@ref(fig:fig7)), which would produce the stronger effect size coefficients $\beta_{|\Delta x|}$ observed in the weak selection scenarios in Figure \@ref(fig:fig5).
+Though dispersal and environmental variability (in scenarios 3 and 4) did lead to more local adaptation and changes in species abundances and local composition later in the time series (Figure \@ref(fig:fig7)), these were not correlated with each other and thus did not produce strong impacts of trait evolution for species abundances.
 
 It is interesting to look in more depth at the impacts that trait evolution - quantified as the absolute value of the change in the population's trait value $x$ from one time step to the next ($|\Delta x_{ik,t-1 \to t}|$), have for the population sizes of other species ($y_i$).
-In Scenario 1 (weak selection, no environmental variability), we observe that $|\Delta x_{15}|$ has a very strong impact for $y_{13}$, with a posterior mean of `r mean(m.df$Bx13.15)` (and 95% HPDI, or 95% highest probability density interval = `r quantile(m.df$Bx13.15,probs=c(0.025,0.975))`).
-Figure 6 \@ref(fig:fig6) shows that high population densities of species 13 correlate with high trait changes of species 15 in all 5 patches, where both species co-exist.
-These indicate that species 15 had its greatest trait changes at the start of the simulation during which species 13 reached high population densities.
-As species 15 declines due to the increase of species 2, species 13 recovers again its population size (Fig.
-6b,h,j).
-Similarly we also observe a strong impact of $|\Delta x_{2}|$ for $y_{15}$, with a posterior mean of `r mean(m.df$Bx15.2)` (and 95% HPDI, or 95% highest probability density interval = `r quantile(m.df$Bx15.2,probs=c(0.025,0.975))`).
-It is important to note that interspecific impacts of $|\Delta x_{i}|$ are always indirect, as there is no direct interaction between $x_i$ and $y_j$.
-The only possible form of influence of is an increase in $y_i$ through evolution, leading to changes in intraspecific competition.
-We studied these posteriors in more detail, to provide an example on how to interpret these parameter estimations by the HMSC model (see Appendix XX).
+In Scenario 1 (weak selection, no environmental variability), we observe that $|\Delta x_{15}|$ Species 15 has a very strong impact for $y_{13}$, with a posterior mean of `r mean(m.df$Bx13.15)` (and 95% HPDI, or 95% highest probability density interval = `r quantile(m.df$Bx13.15,probs=c(0.025,0.975))`).
+See Figure \@ref(fig:fig8).
+XXXXX.
 
@@
 
-This means that the absolute value of trait evolution $|\Delta x_{1,t-1 \to t}|$ had a positive impact on population size $N_{1,t}$, and the population size at the previous time step $N_{1,t-1}$ also had a small positive effect, but the effect of trait evolution depended on the population size (Figure \@ref(fig:fig7)).
+This means that the absolute value of trait evolution $|\Delta x_{1,t-1 \to t}|$ had a positive impact on population size $N_{1,t}$, and the population size at the previous time step $N_{1,t-1}$ also had a small positive effect, but the effect of trait evolution depended on the population size (Figure \@ref(fig:fig9)).
 These interaction terms are interesting to interpret from a perspective of better understanding eco-evolutionary interactions (where the impacts of evolution for population size depend on the value of population size).
@@
 We can look in more depth at the interaction effects and evaluate whether they are additive, synergistic, or antagonistic.
-We calculated posterior distributions for $\beta_3 - (\beta_1 + \beta_2)$ and plot these in Figure \@ref(fig:fig8).
+We calculated posterior distributions for $\beta_3 - (\beta_1 + \beta_2)$ and plot these in Figure \@ref(fig:fig10).
 Values with negative HPDI are antagonistic, with HPDI overlapping 0 are additive, and with positive HPDI are synergistic.
 We observed substantially more antagonisic interactions of population size and trait evolution in the scenarios with weak selection (Scenarios 1 and 3).
+
+The importance of evolution for community composition can be observed by considering model predictions for the simulated dynamics when there is no evolution.
+In this instance, we replace the $|\Delta x|$ values and the interaction effects $N_{ik,t-1} \cdot |\Delta x_{ik,t-1 \to t}|$ in the predictor matrix to 0 and consider the predicted values compared to observed.
+The deviations seen in Figure \@ref(fig:fig10) indicate that failing to consider evolution in the system (and interactions with population size) does impact the predicted population size.
+This is less pronounced in this simulated example, likely due to the perfect time series included in the HMSC model and thus the high explanatory power of the $N_{ik,t-1}$ predictors.
+[note: try average |DeltaX|, try to separate most explanatory N_t-1 values (intra or inter?).
+This needs more work.]
 
@@
 --->
-When interpreting estimated posteriors of the HMSC model, for the different parameters, it is important to consider that these are based on correlations.
-Such a correlation may be the result of a direct impact from one parameter, the effect of an unmeasured parameter or the result of different parameter interacting.
-Therefore, when interpreting their effects on the communitiy processes they need to be investigated further.
-As an example for this we examined the higher posterior of $|\Delta x_{15}|$ on $y_{13}$ and of $|\Delta x_{2}|$ on $y_{15}$ (for further detail see Appendix XX).
-In this example we were able to show that the effect of $|\Delta x_{2}|$ on $y_{15}$ was a direct consequence of the strong competition coefficient between both species and as species 2 was the stronger competitor (through higher population size) it led to the decline of $y_{15}$.
-At the same time the effect of $|\Delta x_{15}|$ on $y_{13}$ was the result of a correlation without causation, as species 13 has no strong competition with species 15 (or 2).
-Therefore, before concluding effects of the estimated posteriors these parameters need to be further analyzed.
-Furthermore, results from species with low densities need to be interpreted with care as the low population size might inflate the resulting posterior.
 
@@
 ```{r fig6, echo=FALSE, fig.show="hold", warning=FALSE,out.width="90%", fig.cap = "Plot showing interpretation of impact of DeltaX-15 for Y-13."}
-# Load data
+# # Load data
 res  <- R.matlab::readMat(paste("../data/",result,"_res.mat",sep="")) # Scenario 1
@@
 for (i in 1:length(b)) {
-patches[[i]] <- data.frame(x15=X3d[b[i],23,],y02=Y3d[b[i],2,],y13=Y3d[b[i],9,],y15=Y3d[b[i],11,],time=2:1000)  
+ patches[[i]] <- data.frame(x15=X3d[b[i],23,],y04=Y3d[b[i],3,],y13=Y3d[b[i],9,],y15=Y3d[b[i],11,],time=2:1000)
  patches_l[[i]] <- pivot_longer(patches[[i]],cols=2:4,values_to="Y",names_to = "sp")
@@
    ylim(c(0,700))+
-  scale_color_manual(values = c("coral3","darkorchid4","rosybrown1"))+
+   scale_color_manual(values = c("gray47","darkorchid4","rosybrown1"))+
    theme_bw()+
@@
 
-```{r fig7, echo=FALSE, fig.show="hold", out.width="60%", fig.cap = "Interpretation of interaction effect for Species 1, Scenario 1.
+```{r fig7, echo=FALSE, warning=FALSE, message=FALSE, fig.show="hold", out.width="100%", fig.cap = "Time series of diversity, N, deltaX."}
+d_lev <- c("d_zero","d_minus9","d_minus8","d_minus7","d_minus6","d_minus5","d_minus4","d_minus3","d_minus2","d_01","d_02","d_03","d_04","d_05","d_06","d_07","d_08","d_09","d_10")
+h_lev <- c("h_0_","h_01_","h_02_","h_03_","h_04_","h_05_","h_06_","h_07_","h_08_","h_09_","h_10_")
+setwd("../data")
+z <- 2
+f <- 8
+result <- paste(h_lev[z],d_lev[f],sep="")
+# Scenario 1
+res  <- R.matlab::readMat(paste(result,"_res.mat",sep=""))
+r <- res$N
+r1 <- get_deltaX(res) # setup for dX plot
+delta_x_time_s1 <- r1$delta_x_time
+tot_delta_x_s1 <- r1$tot_delta_x
+r2 <- get_div(r)
+div_s1 <- r2$div # setup for diversity plot
+totN1 <- r2$totN # setup for N plot
+# Scenario 2
+res  <- R.matlab::readMat(paste(result,"_res_v02.mat",sep=""))
+r <- res$N
+r1 <- get_deltaX(res)
+delta_x_time_s2 <- r1$delta_x_time
+tot_delta_x_s2 <- r1$tot_delta_x
+r2 <- get_div(r)
+div_s2 <- r2$div # setup for diversity plot
+totN2 <- r2$totN # setup for N plot
+# Scenario 3
+res  <- R.matlab::readMat(paste(result,"_res_v03.mat",sep=""))
+r <- res$N
+r1 <- get_deltaX(res)
+delta_x_time_s3 <- r1$delta_x_time
+tot_delta_x_s3 <- r1$tot_delta_x
+r2 <- get_div(r)
+div_s3 <- r2$div # setup for diversity plot
+totN3 <- r2$totN # setup for N plot
+# Scenario 4
+res  <- R.matlab::readMat(paste(result,"_res_v04.mat",sep=""))
+r <- res$N
+r1 <- get_deltaX(res)
+delta_x_time_s4 <- r1$delta_x_time
+tot_delta_x_s4 <- r1$tot_delta_x
+r2 <- get_div(r)
+div_s4 <- r2$div # setup for diversity plot
+totN4 <- r2$totN # setup for N plot
+
+## Diversity as time series
+# Scenario 1
+div_gg <- reshape2::melt(div_s1[,,1])
+colnames(div_gg) <- c("time","div","mean")
+div_gg$se <- reshape2::melt(div_s1[,,2])[,3]
+div_list <- list()
+div_list[[1]] <- ggplot2::ggplot(div_gg,ggplot2::aes(x=time,y=mean,col=div)) + ggplot2::geom_line(size=0.05) + ggplot2::geom_point(size=0.2) + ggplot2::geom_errorbar(ggplot2::aes(ymin=mean-se, ymax=mean+se), width=0.02, position=ggplot2::position_dodge(.05), size=0.05) + ggplot2::coord_cartesian(xlim=c(0,200))  + theme(legend.position = "none")
+# Scenario 2
+div_gg <- reshape2::melt(div_s2[,,1])
+colnames(div_gg) <- c("time","div","mean")
+div_gg$se <- reshape2::melt(div_s1[,,2])[,3]
+div_list[[2]] <- ggplot2::ggplot(div_gg,ggplot2::aes(x=time,y=mean,col=div)) + ggplot2::geom_line(size=0.05) + ggplot2::geom_point(size=0.2) + ggplot2::geom_errorbar(ggplot2::aes(ymin=mean-se, ymax=mean+se), width=0.02, position=ggplot2::position_dodge(.05), size=0.05) + ggplot2::coord_cartesian(xlim=c(0,200)) + theme(legend.position = "none")
+# Scenario 3
+div_gg <- reshape2::melt(div_s3[,,1])
+colnames(div_gg) <- c("time","div","mean")
+div_gg$se <- reshape2::melt(div_s1[,,2])[,3]
+div_list[[3]] <- ggplot2::ggplot(div_gg,ggplot2::aes(x=time,y=mean,col=div)) + ggplot2::geom_line(size=0.05) + ggplot2::geom_point(size=0.2) + ggplot2::geom_errorbar(ggplot2::aes(ymin=mean-se, ymax=mean+se), width=0.02, position=ggplot2::position_dodge(.05), size=0.05) + ggplot2::coord_cartesian(xlim=c(0,200)) + theme(legend.position = "none")
+# Scenario 4
+div_gg <- reshape2::melt(div_s4[,,1])
+colnames(div_gg) <- c("time","div","mean")
+div_gg$se <- reshape2::melt(div_s1[,,2])[,3]
+div_list[[4]] <- ggplot2::ggplot(div_gg,ggplot2::aes(x=time,y=mean,col=div)) + ggplot2::geom_line(size=0.05) + ggplot2::geom_point(size=0.2) + ggplot2::geom_errorbar(ggplot2::aes(ymin=mean-se, ymax=mean+se), width=0.02, position=ggplot2::position_dodge(.05), size=0.05) + ggplot2::coord_cartesian(xlim=c(0,200)) + theme(legend.position = "none")
+
+## Time series plot of total N per scenario (by species)
+mycols2 <- c("maroon4","coral3","peachpuff3","gray47","orange3","darkorange","gray33","lightgoldenrod2","gray96","cadetblue4","gray40","lightseagreen","darkorchid4","midnightblue","rosybrown1" )
+# Scenario 1
+totN_s <- data.frame(t(totN1))
+colnames(totN_s) <- gsub("X","s",colnames(totN_s))
+totN_s <- reshape2::melt(totN_s)
+totN_s$time <- rep(1:1000,11)
+N_list <- list()
+N_list[[1]] <- ggplot2::ggplot(totN_s,ggplot2::aes(x=time,y=value,col=variable)) + ggplot2::geom_line() + ggplot2::coord_cartesian(xlim=c(0,200),ylim=c(0,15000)) + scale_color_manual(values=mycols2[-e]) + theme(legend.position = "none")
+# Scenario 2
+totN_s <- data.frame(t(totN2))
+colnames(totN_s) <- gsub("X","s",colnames(totN_s))
+totN_s <- reshape2::melt(totN_s)
+totN_s$time <- rep(1:1000,11)
+N_list[[2]] <- ggplot2::ggplot(totN_s,ggplot2::aes(x=time,y=value,col=variable)) + ggplot2::geom_line() + ggplot2::coord_cartesian(xlim=c(0,200),ylim=c(0,15000)) + scale_color_manual(values=mycols2[-e]) + theme(legend.position = "none")
+# Scenario 3
+totN_s <- data.frame(t(totN3))
+colnames(totN_s) <- gsub("X","s",colnames(totN_s))
+totN_s <- reshape2::melt(totN_s)
+totN_s$time <- rep(1:1000,11)
+N_list[[3]] <- ggplot2::ggplot(totN_s,ggplot2::aes(x=time,y=value,col=variable)) + ggplot2::geom_line() + ggplot2::coord_cartesian(xlim=c(0,200),ylim=c(0,15000)) + scale_color_manual(values=mycols2[-e]) + theme(legend.position = "none")
+# Scenario 4
+totN_s <- data.frame(t(totN4))
+colnames(totN_s) <- gsub("X","s",colnames(totN_s))
+totN_s <- reshape2::melt(totN_s)
+totN_s$time <- rep(1:1000,11)
+N_list[[4]] <- ggplot2::ggplot(totN_s,ggplot2::aes(x=time,y=value,col=variable)) + ggplot2::geom_line() + ggplot2::coord_cartesian(xlim=c(0,200),ylim=c(0,15000)) + scale_color_manual(values=mycols2[-e]) + theme(legend.position = "none")
+
+## Time series plot of total |deltaX| per scenario
+tot_delta_x_s1 <- data.frame(tot_delta_x_s1)
+tot_delta_x_s1$time <- 2:1000
+plist <- list()
+plist[[1]] <- ggplot2::ggplot(tot_delta_x_s1,ggplot2::aes(x=time,y=tot_delta_x_s1)) + ggplot2::geom_line(size=0.2) + ggplot2::coord_cartesian(ylim=c(0,2))
+
+tot_delta_x_s2 <- data.frame(tot_delta_x_s2)
+tot_delta_x_s2$time <- 2:1000
+plist[[2]] <- ggplot2::ggplot(tot_delta_x_s2,ggplot2::aes(x=time,y=tot_delta_x_s2)) + ggplot2::geom_line(size=0.2) + ggplot2::coord_cartesian(ylim=c(0,2))
+
+tot_delta_x_s3 <- data.frame(tot_delta_x_s3)
+tot_delta_x_s3$time <- 2:1000
+plist[[3]] <- ggplot2::ggplot(tot_delta_x_s3,ggplot2::aes(x=time,y=tot_delta_x_s3)) + ggplot2::geom_line(size=0.2) + ggplot2::coord_cartesian(ylim=c(0,2))
+
+tot_delta_x_s4 <- data.frame(tot_delta_x_s4)
+tot_delta_x_s4$time <- 2:1000
+plist[[4]] <- ggplot2::ggplot(tot_delta_x_s4,ggplot2::aes(x=time,y=tot_delta_x_s4)) + ggplot2::geom_line(size=0.2) + ggplot2::coord_cartesian(ylim=c(0,2))
+
+# Need to add legend
+cowplot::plot_grid(div_list[[1]],div_list[[2]],div_list[[3]],div_list[[4]],N_list[[1]],N_list[[2]],N_list[[3]],N_list[[4]],plist[[1]],plist[[2]],plist[[3]],plist[[4]],byrow=TRUE,rel_widths = c(1,1,1,1,1,1,1,1,2,2,2,2),labels=c("a","b","c","d","e","f","g","h","i","j","k","l"),label_size=10)
+```
+
+\newpage
+
+```{r fig8, echo=FALSE, fig.show="hold", out.width="60%", fig.cap = "Interpretation of interaction effect for Species 1, Scenario 1.
 Model-predicted value for $N_{1t}$ across a range of $|deltaX|$ values, using the mean (black line), 2.5% (orange lines), and 97.5% (green lines) values of $N_{1,t-1}$.
@@
 
-```{r fig8, echo=FALSE, fig.show="hold", out.width="60%", fig.cap = "Antagonistic (HPDI < 0), additive (HPDI includes 0), or synergistic (HPDI > 0) eco-evolutionary interaction effects for (a) scenario 1 (weak selection, no environmental variability), (b) scenario 2 (strong selection, no environmental variability), (c) scenario 3 (weak selection, environmental variability), and (d) scenario 4 (strong selection, environmental variability)."}
+```{r fig9, echo=FALSE, fig.show="hold", out.width="60%", fig.cap = "Antagonistic (HPDI < 0), additive (HPDI includes 0), or synergistic (HPDI > 0) eco-evolutionary interaction effects for (a) scenario 1 (weak selection, no environmental variability), (b) scenario 2 (strong selection, no environmental variability), (c) scenario 3 (weak selection, environmental variability), and (d) scenario 4 (strong selection, environmental variability)."}
 # Load data
@@
 
+```{r fig10, echo=FALSE, fig.show="hold", out.width="100%", fig.cap = "Predicted plots with |deltaX| = 0"}
+# Load results
+load("../output/h_01_d_minus3_hmsc_v01.RData")
+mX <- X
+mX[,24:34] <- 0
+pd.1 <- predict(m.1,XData=mX)
+# What do the predicted vs. observed values look like?
+## Calculate HPDI for each data point
+s <- ncol(m.1$Y)
+pdv.1 <- lapply(pd.1,as.vector)
+pdm.1 <- do.call(rbind, pdv.1)
+pd_hdpi_25.1 <- apply(pdm.1,2,quantile,probs=0.025)
+pd_hdpi_975.1 <- apply(pdm.1,2,quantile,probs=0.975)
+pdm_hdpi_25.1 <- matrix(pd_hdpi_25.1,nrow=dim(pd.1[[1]])[1],ncol=dim(pd.1[[1]])[2],byrow=FALSE)
+pdm_hdpi_975.1 <- matrix(pd_hdpi_975.1,nrow=dim(pd.1[[1]])[1],ncol=dim(pd.1[[1]])[2],byrow=FALSE)
+## Plot time series with 95% HDPI
+## Identify which species are absent (N=0) in each site-by-year instance
+plotY <- Y
+plot_25 <- pdm_hdpi_25.1
+plot_975 <- pdm_hdpi_975.1
+a <- apply(plotY,2,min)
+for(i in 1:s){
+  b <- which(plotY[,i]==a[i])
+  plotY[b,i] <- NA
+  plot_25[b,i] <- NA
+  plot_975[b,i] <- NA
+}
+## Plot
+par(mfrow=(c(10,5)))
+par(mar=c(1,1,1,1))
+for(j in 1:50){
+  plot(plotY[Random$site==j,1],type="n",ylim=c(-1,800),ylab=NA,xlab=NA,xaxs="i",yaxs="i")
+  lines(plot_25[Random$site==j,1],col="red")
+  lines(plot_975[Random$site==j,1],col="red")
+  lines(plotY[Random$site==j,1],col="black")
+  for(s in 2:s){
+    lines(plot_25[Random$site==j,s],col="red")
+    lines(plot_975[Random$site==j,s],col="red")
+    lines(plotY[Random$site==j,s],col="black")
+  }
+}
+
+# We zoom into a few sites to understand what the plot is telling us:
+factor <- matrix(NA,nrow=nrow(Random),ncol=ncol(Random),dimnames=list(NULL,colnames(Random)))
+factor[,1] <- as.numeric(as.character(Random$site))
+factor[,2] <- as.numeric(as.character(Random$year))
+factor[,3] <- as.numeric(as.character(Random$sample))
+factor <- as.data.frame(factor)
+
+par(mfrow=c(6,1))
+k <- c(5,12,15,20,49)
+#k <- c(17,21,30,35,40)
+j <- k[1]
+plot(plotY[factor$site==j & factor$year < 200,1],type="n",ylim=c(-1,800),ylab=NA,xlab=NA,xaxs="i",yaxs="i")
+lines(plot_25[factor$site==j & factor$year < 200,1],col="red")
+lines(plot_975[factor$site==j & factor$year < 200,1],col="red")
+lines(plotY[factor$site==j & factor$year < 200,1],col="black")
+for(s in 2:s){
+lines(plot_25[factor$site==j & factor$year < 200,s],col="red")
+lines(plot_975[factor$site==j & factor$year < 200,s],col="red")
+lines(plotY[factor$site==j & factor$year < 200,s],col="black")
+}
+
+j <- k[2]
+plot(plotY[factor$site==j & factor$year < 200,1],type="n",ylim=c(-1,800),ylab=NA,xlab=NA,xaxs="i",yaxs="i")
+lines(plot_25[factor$site==j & factor$year < 200,1],col="red")
+lines(plot_975[factor$site==j & factor$year < 200,1],col="red")
+lines(plotY[factor$site==j & factor$year < 200,1],col="black")
+for(s in 2:s){
+lines(plot_25[factor$site==j & factor$year < 200,s],col="red")
+lines(plot_975[factor$site==j & factor$year < 200,s],col="red")
+lines(plotY[factor$site==j & factor$year < 200,s],col="black")
+}
+
+j <- k[3]
+plot(plotY[factor$site==j & factor$year < 200,1],type="n",ylim=c(-1,800),ylab=NA,xlab=NA,xaxs="i",yaxs="i")
+lines(plot_25[factor$site==j & factor$year < 200,1],col="red")
+lines(plot_975[factor$site==j & factor$year < 200,1],col="red")
+lines(plotY[factor$site==j & factor$year < 420,1],col="black")
+for(s in 2:s){
+lines(plot_25[factor$site==j & factor$year < 200,s],col="red")
+lines(plot_975[factor$site==j & factor$year < 200,s],col="red")
+lines(plotY[factor$site==j & factor$year < 200,s],col="black")
+}
+
+j <- k[4]
+plot(plotY[factor$site==j & factor$year < 200,1],type="n",ylim=c(-1,800),ylab=NA,xlab=NA,xaxs="i",yaxs="i")
+lines(plot_25[factor$site==j & factor$year < 200,1],col="red")
+lines(plot_975[factor$site==j & factor$year < 200,1],col="red")
+lines(plotY[factor$site==j & factor$year < 200,1],col="black")
+for(s in 2:s){
+lines(plot_25[factor$site==j & factor$year < 200,s],col="red")
+lines(plot_975[factor$site==j & factor$year < 200,s],col="red")
+lines(plotY[factor$site==j & factor$year < 200,s],col="black")
+}
+
+j <- k[5]
+plot(plotY[factor$site==j & factor$year < 200,1],type="n",ylim=c(-1,800),ylab=NA,xlab=NA,xaxs="i",yaxs="i")
+lines(plot_25[factor$site==j & factor$year < 200,1],col="red")
+lines(plot_975[factor$site==j & factor$year < 200,1],col="red")
+lines(plotY[factor$site==j & factor$year < 200,1],col="black")
+for(s in 2:s){
+lines(plot_25[factor$site==j & factor$year < 200,s],col="red")
+lines(plot_975[factor$site==j & factor$year < 200,s],col="red")
+lines(plotY[factor$site==j & factor$year < 200,s],col="black")
+}
+
+j <- 48
+plot(plotY[factor$site==j & factor$year < 800,1],type="n",ylim=c(-1,800),ylab=NA,xlab=NA,xaxs="i",yaxs="i")
+lines(plot_25[factor$site==j & factor$year < 800,1],col="red")
+lines(plot_975[factor$site==j & factor$year < 800,1],col="red")
+lines(plotY[factor$site==j & factor$year < 800,1],col="black")
+for(s in 2:s){
+lines(plot_25[factor$site==j & factor$year < 800,s],col="red")
+lines(plot_975[factor$site==j & factor$year < 800,s],col="red")
+lines(plotY[factor$site==j & factor$year < 800,s],col="black")
+}
+```
+
+\newpage
+
 <a name="appendix">Appendix 1</a>

### 2024-04-20 Appendix_S1_v01.Rmd

---
title: "Appendix S1. Linear regression models and non-linear population dynamics"
author:
  - Jelena H. Pantel^[Laboratoire Chrono-environnement,UMR 6249 CNRS-UFC, 16 Route de Gray, 25030 Besançon cedex, France, jelena.pantel@univ-fcomte.fr]
  - Ruben J. Hermann^[University of Duisburg-Essen, Universitätsstraße 5, 45141 Essen, Germany, ruben.hermann@uni-due.de]
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::pdf_document2:
    toc: no
    fig_caption: yes
    latex_engine: xelatex
  bookdown::html_document2: default
  word_document:
    reference_docx: docx_template.docx
header-includes:
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \fancyhead[L]{Author Pantel and Hermann}
- \fancyhead[R]{Detection of ecoevo dynamics}
- \usepackage{lineno}
- \linenumbers
linestretch: 2
link-citations: yes
linkcolor: blue
csl: the-american-naturalist.csl
bibliography: refs.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
```{r setup,echo=FALSE,message=FALSE}
library(Hmsc)
```
```{r,echo=FALSE,message=FALSE,warning=FALSE}
# Wrap long lines
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy=TRUE)
```

# 1 species, logistic growth
Population growth over time in a single species is first modelled using a Beverton-Holt (discrete-time, logistic) model (@Beverton1957), using an intra-specific competition coefficient for density-dependent growth.

$$ N_{i,t+1} = \frac{r_i N_{i,t}}{1 + \alpha_{ii} N_{i,t}}$$

$$ N_{i,t+1} =\frac{\hat{W}e^ \frac{-[(\frac{w+(1-h^2)P} {P+w})(E-x_{i,t})]^2}{2(P+w)}N_{i,t}} {1 + \alpha_{ii}N_{i,t} + \sum_{j \neq i}^{S} \alpha_{ij}N_{j,t}} $$

```{r}

```

## Generating the data frames
For the diagnosis first need to convert the chains into a coda object and extract the effective sample size (ESS), which denotes how many MCMC iteration were independent (e.g. having no autocorrelation) from previous iterations, and  potential scale reduction factors (PSRF), which describe how well the MCMC chains converged around a value and if the two chains differ from each other. Therefore a high value for ESS (close to the number of actual iterations multiplied by the number of chains) is preferable and a PSRF value of 1 represents both chains converging on the same mean.
Note we do not check the predictive and explanatory power of the model, as we have a perfect data set by the time series data of the simulations, which leads to $R^2$ > .98.
```{r eval=FALSE}
load("../../output/h_01_d_minus3_hmsc_v01.RData")
m.post <- convertToCodaObject(m.1)
ess.beta<- effectiveSize(m.post$Beta)
psrf.beta <- gelman.diag(m.post$Beta, multivariate=FALSE)$psrf
```

# References

```{r}
1 + 1
knitr::knit_exit()
```

### 2024-04-20 Appendix_S1_v01.Rmd
Added BH sim + arima

@@
     latex_engine: xelatex
+    extra_dependencies: ["float"]
   bookdown::html_document2: default
@@
 - \linenumbers
-linestretch: 2
 link-citations: yes
@@
 library(Hmsc)
+library(ggplot2)
+library(tidyr)
 ```
@@
 # Wrap long lines
-knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy=TRUE)
+knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy=TRUE,out.width="50%")
 ```
 
-# 1 species, logistic growth
-Population growth over time in a single species is first modelled using a Beverton-Holt (discrete-time, logistic) model (@Beverton1957), using an intra-specific competition coefficient for density-dependent growth.
+# One species, logistic growth
+Population growth over time in a single species is first modelled using a Beverton-Holt (discrete-time, logistic) model (@Beverton1957), using an intra-specific competition coefficient for density-dependent growth (@Hart2013).
 
 $$ N_{i,t+1} = \frac{r_i N_{i,t}}{1 + \alpha_{ii} N_{i,t}}$$
+Note that in this model, the system is at equilibrium when $N_{i,t+1} = N_{i,t}$, and therefore:
+
+$$ N^* = N^*\frac{r_i}{1 + \alpha_{ii} N^*} $$
+$$ 1 = \frac{r_i}{1 + \alpha_{ii} N^*} $$
+$$ N^* = \frac{r_i - 1}{\alpha_{ii}} $$
+## Population dynamics simulation
+In the metacommunity simulation in the main text, a species resides in a site with an initial population size $N_{i,0} \sim Pois(10)$, a growth rate $r_i$ that depends on the local environmental value $E_k$ and the species trait $x_i$, and a fixed intra-specific competition coefficient of $\alpha_{ii}$ = 0.00125.
+We simulate population growth here:
+
+```{r}
+set.seed(42)
+# Simulate initial species population growth
+N1.0 <- rpois(1,10)
+r1.0 <- 1.67
+alpha.11 <- 0.00125
+# model function
+disc_log <- function(r, N0, alpha) {
+    Nt1 <- (r*N0) / (1+alpha*N0)
+    return(Nt1)
+}
+# Simulation of model for t time steps
+t <- 30
+N <- rep(NA, t)
+N[1] <- N1.0
+for (i in 2:t) {
+  N[i] <- disc_log(r=r1.0, N0=N[i-1], alpha=alpha.11)
+}
+```
+```{r figs1, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs1}Population size \\textit{N} over time \\textit{t} for a discrete-time logistic growth model, with parameters $r_i$ = 1.67, $N_{1,0}$ = 14, and $\\alpha_{11}$ = 0.00125."}
+# Plot simulation: ggplot
+dat <- as.data.frame(N)
+dat$time <- 1:t
+ggplot2::ggplot(dat, aes(time, N)) + geom_point() + geom_hline(yintercept = ((r1.0 - 1)/alpha.11),
+    linetype = "dashed", color = "gray")
+```
+
+## Linear statistical model
+
+We fit the population time series data to a first-order auto-regressive model to predict $N_{t+1}$ as a function of $N_t$:
+
+$$ N_{t+1} = \beta_0 + \beta_1 N_t + \epsilon_t $$
+```{r}
+m.1 <- arima(log(N),order=c(1,0,0))
+#plotting the series along with the fitted values
+m.1.fit <- log(N) - residuals(m.1)
+dat$ar1.fit <- m.1.fit
+```
+```{r figs2, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs2}Population size over time (black line) with fitted values from a first-order autoregressive model (red dashed line)."}
+colors <- c("data" = "black", "model fit" = "blue")
+ggplot(data=dat,aes(x=time)) + geom_point(aes(y=log(N),color="data")) + geom_point(aes(y=ar1.fit, color = "model fit"), alpha=0.5) +
+  geom_line(aes(y=log(N), color="data"), linewidth=.4) +
+  geom_line(aes(y=ar1.fit, color = "model fit"), alpha=0.5, linewidth=.4) +
+  theme_bw() +
+  xlab("time") +
+  ylab("N") +
+  labs(color="Legend") +
+  scale_color_manual(values = colors)
+```
+
+We can also examine density dependence by plotting $\Delta N = N_{t+1} - N_t$ vs. $N_t$:
+
+```{r}
+dN <- rep(NA, (t-1))
+for (i in 1:(t-1)) {
+  dN[i] <- log(N[i+1])-log(N[i])
+}
+```
+```{r figs3, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs2}Population size over time (black line) with fitted values from a first-order autoregressive model (red dashed line)."}
+d.dN <- as.data.frame(dN)
+d.dN$Nt <- log(dat$N[1:(t-1)])
+ggplot2::ggplot(d.dN, aes(Nt, dN)) + geom_point()
+```
+
+```{r}
+knitr::knit_exit()
+```
+
+In the metacommunity simulation in the main text, a species $i$ resides in a site $k$ with an environmental optimum trait value $E_k \sim unif(0,1)$, an initial population size $N_{i,0} \sim Pois(10)$, and an initial degree of maladaptation $\beta_{0,i} \sim Gamma(0.75,1)$.
+From this \(\beta_{0,i}\), the initial distance to the local patch's optimum phenotype was calculated as $d_{0,i}=\sqrt{B_{0,i}(w+P)}$ and the initial trait value for each species $x_{0,i}$ is calculated by subtracting $d_{0,i}$ from the local environmental optimum value $E_k$.
+Here I simulate a species that is initially suited for the environment ($d_{0,i}$)
+
+Internal question 1 - how does the $\alpha$ version compared to the $K$ version?
+
+```{r}
+set.seed(42)
+# Simulate initial species population growth
+N1.0 <- rpois(1,10)
+# Behind-the-scenes initial r1.0
+E1 <- runif(1,0,1)
+beta1.0 <- rgamma(1,0.75,1)
+w <- 10
+P <- 1
+Wmax <- 2
+What <- Wmax*sqrt(w / (P+w))
+h2 <- 0
+d1.0 <- sqrt(beta1.0*(w+P))
+x1.0 <- E1 - d1.0
+# Emergent initial r1.0
+r1.0 <- What*exp( -(((w+(1-h2)*P)/(P+w))*(E1-x1.0))^2/ (2*(P+w)))
+alpha.11 <- 0.00125
+
+# model function
+disc_log <- function(r, N0, alpha) {
+    Nt1 <- (r*N0) / (1+alpha*N0)
+    return(Nt1)
+}
+
+# Simulation of model for t time steps
+N <- rep(NA, t)
+N[1] <- N1.0
+t <- 30
+for (i in 2:t) {
+  N[i] <- disc_log(r=r1.0, N0=N[i-1], alpha=alpha.11)
+    N0 <- N[i]
+}
+# Plot simulation: ggplot
+dat <- as.data.frame(N)
+dat$time <- as.numeric(rownames(dat))
+ggplot2::ggplot(dat, aes(time, N)) + geom_point() + geom_hline(yintercept = (1/alpha.11),
+    linetype = "dashed", color = "gray")
+# Additional plot: how does growth rate change with time?
+range <- seq(0,800,by=0.01)
+rec <- rep(NA, length(range))
+for (i in 1:length(range)) {
+    rec[i] <- r1.0 / (1+alpha.11*range[i])
+}
+plot(y=rec,x=range,ylim=c(-1,2))
+```
+
+```{r}
+# Parameter values to use for simulation
+r <- 1.21
+K <- 1400
+N0 <- 4
+t <- 30
+# model function
+disc_log <- function(r, N0, K) {
+    Nt1 <- N0 + r * N0 * (1 - N0/K)
+    return(Nt1)
+}
+# Simulation of model for t time steps
+N <- rep(NA, t)
+for (i in 1:t) {
+    N[i] <- disc_log(r, N0, K)
+    N0 <- N[i]
+}
+# Plot simulation: ggplot
+dat <- as.data.frame(N)
+dat$time <- as.numeric(rownames(dat))
+ggplot2::ggplot(dat, aes(time, N)) + geom_point() + geom_hline(yintercept = K,
+    linetype = "dashed", color = "gray")
+```
 
@@
 # References
-
-```{r}
-1 + 1
-knitr::knit_exit()
-```

### 2024-04-20 Appendix_S1_v01.Rmd
Finished 1 species discrete logistic growth + linear model

@@
 library(tidyr)
+library(coda)
 ```
@@
 $$ N_{i,t+1} = \frac{r_i N_{i,t}}{1 + \alpha_{ii} N_{i,t}}$$
-Note that in this model, the system is at equilibrium when $N_{i,t+1} = N_{i,t}$, and therefore:
+Note that in this model, the system is at equilibrium when $N_{i,t+1} = N_{i,t}$, and therefore: $$ N^* = N^*\frac{r_i}{1 + \alpha_{ii} N^*} $$
 
-$$ N^* = N^*\frac{r_i}{1 + \alpha_{ii} N^*} $$
 $$ 1 = \frac{r_i}{1 + \alpha_{ii} N^*} $$
+
 $$ N^* = \frac{r_i - 1}{\alpha_{ii}} $$
+
 ## Population dynamics simulation
@@
 
-We fit the population time series data to a first-order auto-regressive model to predict $N_{t+1}$ as a function of $N_t$:
+We fit the population time series data to a first-order auto-regressive model to predict $N_{t+1}$ as a function of $N_t$, and compare that to a linear regression:
 
 $$ N_{t+1} = \beta_0 + \beta_1 N_t + \epsilon_t $$
-```{r}
-m.1 <- arima(log(N),order=c(1,0,0))
+```{r,warning=FALSE}
+# Fit the model
+m.1.ar <- arima(x = log(N), order = c(1,0,0), include.mean = T, method="CSS")
+m.1.lm <- lm(log(dat$N[2:t]) ~ log(dat$N[1:(t-1)]))
 #plotting the series along with the fitted values
-m.1.fit <- log(N) - residuals(m.1)
-dat$ar1.fit <- m.1.fit
+m.1.ar.fit <- log(N) - residuals(m.1.ar)
+m.1.lm.fit <- log(dat$N[2:t]) - m.1.lm$resid
+dat$ar1.fit <- m.1.ar.fit
+dat$lm.fit <- NA
+dat$lm.fit[2:t] <- m.1.lm.fit
 ```
-```{r figs2, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs2}Population size over time (black line) with fitted values from a first-order autoregressive model (red dashed line)."}
-colors <- c("data" = "black", "model fit" = "blue")
-ggplot(data=dat,aes(x=time)) + geom_point(aes(y=log(N),color="data")) + geom_point(aes(y=ar1.fit, color = "model fit"), alpha=0.5) +
+```{r figs2, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs2}Population size over time (black line) with fitted values from a first-order autoregressive model (red dashed line).",warning=FALSE}
+colors <- c("data" = "black", "AR1 fit" = "gray", "lm fit"="orange")
+ggplot(data=dat,aes(x=time)) +
+  geom_point(aes(y=log(N),color="data")) +
+  geom_point(aes(y=ar1.fit, color = "AR1 fit"), alpha=0.5) +
+  geom_point(aes(y=lm.fit, color = "lm fit"), alpha=0.5) +
   geom_line(aes(y=log(N), color="data"), linewidth=.4) +
-  geom_line(aes(y=ar1.fit, color = "model fit"), alpha=0.5, linewidth=.4) +
+  geom_line(aes(y=ar1.fit, color = "AR1 fit"), alpha=0.5, linewidth=.4) +
+  geom_line(aes(y=ar1.fit, color = "lm fit"), alpha=0.5, linewidth=.4) +
   theme_bw() +
@@
 
-We can also examine density dependence by plotting $\Delta N = N_{t+1} - N_t$ vs. $N_t$:
+The linear model is a good fit, and $N_{t+1}$ and $N_t$ are well-represented by a linear function:
 
-```{r}
+```{r figs3, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs3}Population size (logarithm) at one time step $N_{t+1}$ as a function of log-population size in the previous time step $N_t$.
+"}
 dN <- rep(NA, (t-1))
@@
 }
-```
-```{r figs3, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs2}Population size over time (black line) with fitted values from a first-order autoregressive model (red dashed line)."}
 d.dN <- as.data.frame(dN)
 d.dN$Nt <- log(dat$N[1:(t-1)])
+d.dN$Nt_plus_1 <- log(dat$N[2:t])
+ggplot2::ggplot(d.dN, aes(x=Nt,y=Nt_plus_1)) + geom_point()
+```
+
+We also examine density dependence by plotting $\Delta N = N_{t+1} - N_t$ vs. $N_t$:
+
+```{r figs4, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs4} Change in population size from one time step to the next $N_{t+1}$ as a function of $N_{t+1}$"}
+d.dN <- as.data.frame(dN)
+d.dN$Nt <- log(dat$N[1:(t-1)])
+d.dN$Nt_plus_1 <- log(dat$N[2:t])
 ggplot2::ggplot(d.dN, aes(Nt, dN)) + geom_point()
 ```
+## Bayesian linear statistical model: HMSC
+We can estimate the same model parameters using HMSC:
+
+```{r}
+# prepare data in HMSC format
+Y <- as.matrix(log(dat$N[2:t]))
+XData <- data.frame(x=log(dat$N[1:(t-1)]))
+m.1.hmsc <- Hmsc(Y=Y,XData=XData,XFormula=~x)
+# Bayesian model parameters
+nChains <- 2
+thin <- 5
+samples <- 1000
+transient <- 500*thin
+verbose <- 500*thin
+# sample MCMC
+m.1.sample <- sampleMcmc(m.1.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
+```
+
+```{r}
+m.post.hmsc <- convertToCodaObject(m.1.sample)
+summary(m.post.hmsc$Beta)
+plot(m.post.hmsc$Beta)
+```
+
+These estimates match well with those from the AR1 and linear model:
+
+```{r}
+# AR1 coefficients (recall that the intercept is the term below multipled by 1 - phi1)
+m.1.ar$coef
+m.1.ar$coef[2]*(1-m.1.ar$coef[1])
+# linear model
+summary(m.1.lm)$coefficients[1:2,1:2]
+# Bayesian estimates
+summary(m.post.hmsc$Beta)$statistics[1:2,1:2]
+```
+
+```{r figs5, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs5} Observed (grey) and model-fit (blue) values for population size at time t (x-axis) and t+1 (y-axis).",results='hide',fig.keep='all'}
+Gradient <- constructGradient(m.1.sample,focalVariable="x",ngrid=29)
+predY <- predict(m.1.sample,Gradient=Gradient,expected=TRUE)
+#preds <-computePredictedValues(m.1.sample)
+plotGradient(m.1.sample,Gradient,pred=predY,showData=T,measure="Y",main="",xlab="N_t",ylab="predicted N_t+1")
+```
+
+## Conclusions
+In this example, a first-order auto-regressive model works well, bypassing the need to estimate logistic growth parameters $r_i$ and $\alpha_{ii}$.
+The density-dependence dynamics ($\Delta N \sim f(N_t)$) show an overall declining trend over time.
+The Bayesian estimation implemented in HMSC gives good parameter estimates.
 

### 2024-05-02 Appendix_S1_v01.Rmd
trying to update

@@
 
+# One species, logistic growth, environmental covariate
+We now consider using a linear model to analyze population growth when the species growth rate is impacted by a single environmental covariate.
+
+## Growth depends on environment
+
+First we add environment-dependent growth rate. The growth rate $r_i$ becomes:
+
+$$ r_i = \hat{W}e^{-(E-x_{i,t})^2}$$
+
+Here, $\hat{W}$ is the maximal population growth rate (set to )
+
+equation becomes:
+
+$$ N_{i,t+1} = \frac{r_i N_{i,t}}{1 + \alpha_{ii} N_{i,t}}$$
+
+We make this not linked to then linked to growth and in each instance run a linear model to show no effect is detected or is detected as it should be.
+E can vary randomly, as long as it links to growth.
+
+```{r}
+# Simulate initial species population growth
+N1.0 <- rpois(1,10)
+r1.0 <- 1.67
+alpha.11 <- 0.00125
+# model function
+disc_log <- function(r, N0, alpha) {
+    Nt1 <- (r*N0) / (1+alpha*N0)
+    return(Nt1)
+}
+# Simulation of model for t time steps
+t <- 30
+N <- rep(NA, t)
+N[1] <- N1.0
+for (i in 2:t) {
+  N[i] <- disc_log(r=r1.0, N0=N[i-1], alpha=alpha.11)
+}
+```
+
 In the metacommunity simulation in the main text, a species $i$ resides in a site $k$ with an environmental optimum trait value $E_k \sim unif(0,1)$, an initial population size $N_{i,0} \sim Pois(10)$, and an initial degree of maladaptation $\beta_{0,i} \sim Gamma(0.75,1)$.

### 2024-05-02 Appendix_S1_v01.Rmd
Adding environment-dependent growth rate

@@
 
-Here, $\hat{W}$ is the maximal population growth rate (set to )
+Here, $\hat{W}$ is the maximal population growth rate (set to `r r1.0` as above), $E$ is the local environmental trait optimum value, and $x_{i,t}$ is species *i* trait value at time *t*.
+We simulate the environment $E$ value fluctuating randomly over time, and use a linear model to fit $E$ as a covariate:
+
+```{r}
+# Simulate initial species population growth with environment fluctuations
+N1.0 <- rpois(1,10)
+r1.0 <- 1.67
+alpha.11 <- 0.00125
+E.0 <- 0.8
+x1.0 <- 0.8
+# model function
+disc_log_E <- function(r, N0, alpha, E, x) {
+    Nt1 <- ((r*exp(-(E-x)^2))*N0) / (1+alpha*N0)
+    return(Nt1)
+}
+# Simulation of model for t time steps
+t <- 30
+N <- rep(NA, t)
+N[1] <- N1.0
+E <- rep(NA, t)
+E[1] <- E.0
+for (i in 2:t) {
+  N[i] <- disc_log_E(r=r1.0, N0=N[i-1], alpha=alpha.11, E=E[i-1], x=x1.0)
+  E[i] <- E[i-1] + rnorm(1,0,.1)
+}
+```
+```{r figs1, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs1}Population size \\textit{N} over time \\textit{t} for a discrete-time logistic growth model, with parameters $r_i$ = 1.67, $N_{1,0}$ = 14, and $\\alpha_{11}$ = 0.00125."}
+# Plot simulation: ggplot
+dat <- as.data.frame(cbind(N,E))
+dat$time <- 1:t
+ggplot2::ggplot(dat, aes(time, N)) + geom_point() + geom_hline(yintercept = ((r1.0 - 1)/alpha.11),
+    linetype = "dashed", color = "gray")
+```
+
+```{r}
+r1.0 <- 1.67
+E <- 0.8
+x1.0 <- 0.7
+x_range <- seq(0,2,by=0.01)
+r_range <- r1.0*exp(-(E-x_range)^2)
+plot(x_range,r_range)
+```
+
+# Simulate initial species population growth
+N1.0 <- rpois(1,10)
+r1.0 <- 1.67
+alpha.11 <- 0.00125
 

### 2024-05-02 Appendix_S1_v01.Rmd
Env-dependent growth

--- 2024-05-02 Appendix_S1_v01.Rmd
@@
 library(coda)
+library(patchwork)
 ```
@@
   geom_line(aes(y=ar1.fit, color = "AR1 fit"), alpha=0.5, linewidth=.4) +
-  geom_line(aes(y=ar1.fit, color = "lm fit"), alpha=0.5, linewidth=.4) +
+  geom_line(aes(y=lm.fit, color = "lm fit"), alpha=0.5, linewidth=.4) +
   theme_bw() +
@@
 
-```{r figs5, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs5} Observed (grey) and model-fit (blue) values for population size at time t (x-axis) and t+1 (y-axis).",results='hide',fig.keep='all'}
+```{r figs5, fig.pos = "H", fig.show="hold", fig.cap = "\\label{fig:figs5} Observed (grey) and model-fit (blue) values for population size at time t (x-axis) and t+1 (y-axis).",results='hide',fig.keep='all'}
 Gradient <- constructGradient(m.1.sample,focalVariable="x",ngrid=29)
@@
 
-```{r}
-knitr::knit_exit()
-```
-
 # One species, logistic growth, environmental covariate
@@
 
-Here, $\hat{W}$ is the maximal population growth rate (set to )
+Here, $\hat{W}$ is the maximal population growth rate (set to `r r1.0` as above), $E$ is the local environmental trait optimum value, and $x_{i,t}$ is species *i* trait value at time *t*.
+We see that if $E = x_{i,t}$ then the growth rate is at the value $r$ = `r r1.0`.
+Here, we begin with $E = x_{i,t} = 0.8$, then simulate the environment $E$ value fluctuating randomly over time, and finally use a linear model to fit $E$ as a covariate.
+
+```{r}
+# Simulate initial species population growth with environment fluctuations
+N1.0 <- 10
+r1.0 <- 1.67
+alpha.11 <- 0.00125
+E.0 <- 0.8
+x1.0 <- 0.8
+# model function
+disc_log_E <- function(r, N0, alpha, E, x) {
+    Nt1 <- ((r*exp(-(E-x)^2))*N0) / (1+alpha*N0)
+    return(Nt1)
+}
+# Simulation of model for t time steps
+t <- 40
+N <- rep(NA, t)
+N[1] <- N1.0
+E <- rep(NA, t)
+E[1] <- E.0
+for (i in 2:t) {
+  N[i] <- disc_log_E(r=r1.0, N0=N[i-1], alpha=alpha.11, E=E[i-1], x=x1.0)
+  E[i] <- E[i-1] + rnorm(1,0,.1)
+}
+```
+```{r figs6, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs1}Population size \\textit{N} over time \\textit{t} for a discrete-time logistic growth model, with parameters $r_i$ = 1.67, $N_{1,0}$ = 14, and $\\alpha_{11}$ = 0.00125."}
+# Plot simulation: ggplot
+dat <- as.data.frame(cbind(N,E))
+dat$time <- 1:t
+p1 <- ggplot2::ggplot(dat, aes(time, N)) + geom_point() + geom_hline(yintercept = ((r1.0 - 1)/alpha.11),linetype = "dashed", color = "gray")
+p2 <- ggplot2::ggplot(dat, aes(time, E)) + geom_point()
+p1 + p2
+```
+
+## Linear statistical model with environmental covariate
+
+We now include environment $E$ as a covariate in the linear model:
+
+$$ N_{t} = \beta_0 + \beta_1 N_{t-1} + \beta_2 E_{t-1}+ \epsilon_t $$
+
+```{r,warning=FALSE}
+# Fit the model
+m.2.ar <- arima(x = log(N), order = c(1,0,0), include.mean = T, method="CSS", xreg=E)
+m.2.lm <- lm(log(dat$N[2:t]) ~ log(dat$N[1:(t-1)]) + E[1:(t-1)])
+#plotting the series along with the fitted values
+m.2.ar.fit <- log(N) - residuals(m.2.ar)
+m.2.lm.fit <- log(dat$N[2:t]) - m.2.lm$resid
+dat$ar2.fit <- m.2.ar.fit
+dat$lm2.fit <- NA
+dat$lm2.fit[2:t] <- m.2.lm.fit
+```
+```{r figs7, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs2}Population size over time (black line) with fitted values from a first-order autoregressive model (red dashed line).",warning=FALSE}
+colors <- c("data" = "black", "AR1 fit" = "gray", "lm fit"="orange")
+ggplot(data=dat,aes(x=time)) +
+  geom_point(aes(y=log(N),color="data")) +
+  geom_point(aes(y=ar2.fit, color = "AR1 fit"), alpha=0.5) +
+  geom_point(aes(y=lm2.fit, color = "lm fit"), alpha=0.5) +
+  geom_line(aes(y=log(N), color="data"), linewidth=.4) +
+  geom_line(aes(y=ar2.fit, color = "AR1 fit"), alpha=0.5, linewidth=.4) +
+  geom_line(aes(y=lm2.fit, color = "lm fit"), alpha=0.5, linewidth=.4) +
+  theme_bw() +
+  xlab("time") +
+  ylab("N") +
+  labs(color="Legend") +
+  scale_color_manual(values = colors)
+```
+The linear model is a good fit when including the environmental covariate. $N_{t+1}$ and $N_t$ are no longer well-represented by a linear function without taking $E$ into account.
+
+```{r figs8, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs3}Population size (logarithm) at one time step $N_{t+1}$ as a function of log-population size in the previous time step $N_t$.
+"}
+dN <- rep(NA, (t-1))
+for (i in 1:(t-1)) {
+  dN[i] <- log(N[i+1])-log(N[i])
+}
+d.dN <- as.data.frame(dN)
+d.dN$Nt <- log(dat$N[1:(t-1)])
+d.dN$Nt_plus_1 <- log(dat$N[2:t])
+p1 <- ggplot2::ggplot(d.dN, aes(x=Nt,y=Nt_plus_1)) + geom_point()
+
+d.dN <- as.data.frame(dN)
+d.dN$Nt <- log(dat$N[1:(t-1)])
+d.dN$Nt_plus_1 <- log(dat$N[2:t])
+d.dN$E <- dat$E[2:t]
+p2 <- ggplot2::ggplot(d.dN, aes(x=E,y=Nt_plus_1)) + geom_point()
+p3 <- ggplot2::ggplot(d.dN, aes(Nt, dN)) + geom_point()
+p4 <- ggplot2::ggplot(d.dN, aes(E, dN)) + geom_point()
+
+p1+p2+p3+p4
+```
+
+## Bayesian linear statistical model: HMSC
+We can estimate the same model parameters using HMSC:
+
+```{r}
+# prepare data in HMSC format
+Y <- as.matrix(log(dat$N[2:t]))
+XData <- data.frame(x=cbind(log(dat$N[1:(t-1)])),E[1:(t-1)])
+colnames(XData)[2] <- "E"
+m.2.hmsc <- Hmsc(Y=Y,XData=XData,XFormula=~x+E)
+# Bayesian model parameters
+nChains <- 2
+thin <- 5
+samples <- 1000
+transient <- 500*thin
+verbose <- 500*thin
+# sample MCMC
+m.2.sample <- sampleMcmc(m.2.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
+```
+
+```{r}
+m2.post.hmsc <- convertToCodaObject(m.2.sample)
+summary(m2.post.hmsc$Beta)
+plot(m2.post.hmsc$Beta)
+```
+
+These estimates match well with those from the AR1 and linear model:
+
+```{r}
+# AR1 coefficients (recall that the intercept is the term below multipled by 1 - phi1)
+m.2.ar$coef
+m.2.ar$coef[2]*(1-m.2.ar$coef[1])
+# linear model
+summary(m.2.lm)$coefficients[1:3,1:2]
+# Bayesian estimates
+summary(m2.post.hmsc$Beta)$statistics[1:3,1:2]
+```
+
+We recall that the interpretation of the coefficients in an arimaX (arima with covariates) model is difficult.
+They do not give the impact on $N_t$ per unit increase in X as in a regression.
+So we do not interpret the causation implied by the coefficient in the arimaX model.
+In the regression model, we can see that $E$ has a positive impact on $N_t$.
+
+```{r figs9, fig.pos = "H", fig.show="hold", fig.cap = "\\label{fig:figs5} Observed (grey) and model-fit (blue) values for population size at time t (x-axis) and t+1 (y-axis).",results='hide',fig.keep='all'}
+Gradient <- constructGradient(m.2.sample,focalVariable="E",non.focalVariables=list(x=list(2)),ngrid=39)
+predY <- predict(m.2.sample,Gradient=Gradient,expected=TRUE)
+#preds <-computePredictedValues(m.1.sample)
+plotGradient(m.2.sample,Gradient,pred=predY,showData=T,measure="Y",main="",xlab="E_t",ylab="predicted N_t+1")
+```
+
+## Conclusions
+In this example, the linear regression again works well to describe the impact of $E_t$ for $N_t$.
+We can see the response of course depends on the population size value, as the gradient plot is generally linear
+
+a first-order auto-regressive model works well, bypassing the need to estimate logistic growth parameters $r_i$ and $\alpha_{ii}$.
+The density-dependence dynamics ($\Delta N \sim f(N_t)$) show an overall declining trend over time.
+The Bayesian estimation implemented in HMSC gives good parameter estimates.
+
+```{r}
+knitr::knit_exit()
+```
+
+```{r}
+blah <- data.frame(cbind(log(dat$N[2:t]),log(dat$N[1:(t-1)]),E[1:(t-1)]))
+colnames(blah) <- c("Nt1","Nt","E")
+mod <- lm(Nt1 ~ Nt + E,data=blah)
+jtools::effect_plot(mod,pred=E,plot.points = TRUE)
+```
+
+# Simulate initial species population growth
+N1.0 <- rpois(1,10)
+r1.0 <- 1.67
+alpha.11 <- 0.00125
 

### 2024-05-03 Appendix_S1_v01.Rmd
Env lm: quadratic term added

@@
 library(patchwork)
+library(jtools)
+library(bayesplot)
 ```
@@
 
+```{r}
+lm_dat <- data.frame(cbind(log(dat$N[2:t]),log(dat$N[1:(t-1)])))
+colnames(lm_dat) <- c("Nt1","Nt")
+ggplot(lm_dat,aes(Nt,Nt1)) +
+  stat_summary(fun.data= mean_cl_normal) + 
+  geom_smooth(method='lm')
+```
+
 ## Conclusions
@@
 ```
-```{r figs6, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs1}Population size \\textit{N} over time \\textit{t} for a discrete-time logistic growth model, with parameters $r_i$ = 1.67, $N_{1,0}$ = 14, and $\\alpha_{11}$ = 0.00125."}
+```{r figs6, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs1}Population size \\textit{N} over time \\textit{t} for a discrete-time logistic growth model, with parameters $r_i$ = 1.67, $N_{1,0}$ = 14, and $\\alpha_{11}$ = 0.00125.
+Relationship between E and Nt is also shown."}
 # Plot simulation: ggplot
@@
 m.2.ar <- arima(x = log(N), order = c(1,0,0), include.mean = T, method="CSS", xreg=E)
-m.2.lm <- lm(log(dat$N[2:t]) ~ log(dat$N[1:(t-1)]) + E[1:(t-1)])
+m.2.lm <- lm(log(dat$N[2:t]) ~ log(dat$N[1:(t-1)]) + log(E[1:(t-1)]))
 #plotting the series along with the fitted values
@@
 ```
-The linear model is a good fit when including the environmental covariate. $N_{t+1}$ and $N_t$ are no longer well-represented by a linear function without taking $E$ into account.
 
@@
 d.dN$Nt_plus_1 <- log(dat$N[2:t])
-d.dN$E <- dat$E[2:t]
+d.dN$E <- log(dat$E[2:t])
 p2 <- ggplot2::ggplot(d.dN, aes(x=E,y=Nt_plus_1)) + geom_point()
@@
 p1+p2+p3+p4
+```
+
+The linear model is a good fit when including the environmental covariate.
+$N_{t+1}$ and $N_t$ can still be captured by a linear relationship.
+However we see that the relationship between $N_{t+1}$ and $E_t$ is non-linear.
+This tells us that the lm is good for predictions, but not for inference (for capturing well the relationsip between the predictor and response variable).
+The use of linear relationships in JSDMs is discussed in [@Ingram2020], and in many applications (e.g.
+[@Erickson2023]) quadratic terms are used, which create bell-shaped response curves that may better match species with optimal niches (as opposed to linear, monotonically increasing relationships between population size and environmental predictors).
+We thus include a quadratic term for $E_t$ to provide a better fit to the data.
+
+```{r}
+df <- data.frame(cbind(log(dat$N[2:t]),log(dat$N[1:(t-1)]),E[1:(t-1)],E[1:(t-1)]^2))
+colnames(df) <- c("Nt1","Nt","E","Esq")
+m.2.lm <- lm(Nt1 ~ Nt + E + Esq,data=df)
+```
+
+```{r figs9, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs3}Population size (logarithm) at one time step $N_{t+1}$ as a function of log-population size in the previous time step $N_t$.
+"}
+jtools::effect_plot(m.2.lm,pred=E,plot.points = TRUE)
+ggplot(data = blah, aes(x = E, y = Nt1)) +
+  geom_point() +
+  geom_smooth(method = "lm", formula = y ~ x + I(x^2))
 ```
@@
 Y <- as.matrix(log(dat$N[2:t]))
-XData <- data.frame(x=cbind(log(dat$N[1:(t-1)])),E[1:(t-1)])
-colnames(XData)[2] <- "E"
-m.2.hmsc <- Hmsc(Y=Y,XData=XData,XFormula=~x+E)
+XData <- df
+m.2.hmsc <- Hmsc(Y=Y,XData=XData,XFormula=~Nt+E+Esq)
 # Bayesian model parameters
@@
 summary(m2.post.hmsc$Beta)
-plot(m2.post.hmsc$Beta)
+bayesplot::mcmc_trace(m2.post.hmsc$Beta)
+bayesplot::mcmc_areas(m2.post.hmsc$Beta,area_method = c("equal height"))
 ```
@@
 # linear model
-summary(m.2.lm)$coefficients[1:3,1:2]
+summary(m.2.lm)$coefficients[1:4,1:2]
 # Bayesian estimates
-summary(m2.post.hmsc$Beta)$statistics[1:3,1:2]
+summary(m2.post.hmsc$Beta)$statistics[1:4,1:2]
 ```
@@
 ```{r figs9, fig.pos = "H", fig.show="hold", fig.cap = "\\label{fig:figs5} Observed (grey) and model-fit (blue) values for population size at time t (x-axis) and t+1 (y-axis).",results='hide',fig.keep='all'}
-Gradient <- constructGradient(m.2.sample,focalVariable="E",non.focalVariables=list(x=list(2)),ngrid=39)
-predY <- predict(m.2.sample,Gradient=Gradient,expected=TRUE)
-#preds <-computePredictedValues(m.1.sample)
+Gradient <- constructGradient(m.2.sample,focalVariable="E",non.focalVariables=list(Nt=list(2),Esq=list(2)),ngrid=39)
+predY <- predict(m.2.sample,XData=Gradient$XDataNew,expected=TRUE)
 plotGradient(m.2.sample,Gradient,pred=predY,showData=T,measure="Y",main="",xlab="E_t",ylab="predicted N_t+1")
+# Can't figure out curved gradient
+pr <- predict(m.2.hmsc,m.2.sample,)
 ```
@@
 knitr::knit_exit()
-```
-
-```{r}
-blah <- data.frame(cbind(log(dat$N[2:t]),log(dat$N[1:(t-1)]),E[1:(t-1)]))
-colnames(blah) <- c("Nt1","Nt","E")
-mod <- lm(Nt1 ~ Nt + E,data=blah)
-jtools::effect_plot(mod,pred=E,plot.points = TRUE)
 ```

### 2024-05-04 Appendix_S1_v01.Rmd
Finished 1 species logistic ~ Environment

@@
 jtools::effect_plot(m.2.lm,pred=E,plot.points = TRUE)
-ggplot(data = blah, aes(x = E, y = Nt1)) +
+ggplot(data = df, aes(x = E, y = Nt1)) +
   geom_point() +
@@
 
-```{r figs9, fig.pos = "H", fig.show="hold", fig.cap = "\\label{fig:figs5} Observed (grey) and model-fit (blue) values for population size at time t (x-axis) and t+1 (y-axis).",results='hide',fig.keep='all'}
+```{r figs10, fig.pos = "H", fig.show="hold", fig.cap = "\\label{fig:figs5} Observed (grey) and model-fit (blue) values for population size at time t (x-axis) and t+1 (y-axis).",results='hide',fig.keep='all'}
 Gradient <- constructGradient(m.2.sample,focalVariable="E",non.focalVariables=list(Nt=list(2),Esq=list(2)),ngrid=39)
@@
 plotGradient(m.2.sample,Gradient,pred=predY,showData=T,measure="Y",main="",xlab="E_t",ylab="predicted N_t+1")
-# Can't figure out curved gradient
-pr <- predict(m.2.hmsc,m.2.sample,)
+# Can't figure out curved gradient using E and E^2
+#pr <- predict(m.2.hmsc,m.2.sample,)
 ```
@@
 ## Conclusions
-In this example, the linear regression again works well to describe the impact of $E_t$ for $N_t$.
-We can see the response of course depends on the population size value, as the gradient plot is generally linear
-
-a first-order auto-regressive model works well, bypassing the need to estimate logistic growth parameters $r_i$ and $\alpha_{ii}$.
-The density-dependence dynamics ($\Delta N \sim f(N_t)$) show an overall declining trend over time.
-The Bayesian estimation implemented in HMSC gives good parameter estimates.
+In this example, the linear regression again works well to describe the impact of $E_t$ for $N_t$ when using the quadratic formulation.
+The arimaX model works well for fitting and subsequent prediction, but less well for inference about the impacts of $E$.
+From the quadratic regression terms for $E$, we correctly see that the population size is maximal at the species trait value and decreases away from that value.
+We will continue to use log-transformed abundance and now introduce quadratic terms for the environmental parameter.
 

### 2024-05-04 Appendix_S1_v01.Rmd
Begin 2-species competition + lm

@@
 library(bayesplot)
+library(reshape2)
 ```
@@
 
-```{r}
-knitr::knit_exit()
-```
+# Two species, logistic growth, competition
+XXXXX
+## Interspecific competition
+The growth equation for each species now becomes:
 
-# Simulate initial species population growth
-N1.0 <- rpois(1,10)
-r1.0 <- 1.67
-alpha.11 <- 0.00125
-
-equation becomes:
-
-$$ N_{i,t+1} = \frac{r_i N_{i,t}}{1 + \alpha_{ii} N_{i,t}}$$
-
-We make this not linked to then linked to growth and in each instance run a linear model to show no effect is detected or is detected as it should be.
-E can vary randomly, as long as it links to growth.
+$$ N_{i,t+1} = \frac{r_i N_{i,t}}{1 + \alpha_{ii} N_{i,t} + \alpha_{ij} N_{j,t}}$$
+We use a distinct growth rate for the 2nd species and introduce the interspecific interaction coefficient $\alpha_{ij}$.
 
 ```{r}
-# Simulate initial species population growth
-N1.0 <- rpois(1,10)
+# Simulate initial species population growth with environment fluctuations
+N1.0 <- 10
+N2.0 <- 10
 r1.0 <- 1.67
+r2.0 <- 1.7
 alpha.11 <- 0.00125
+alpha.22 <- 0.00125
+alpha.12 <- 0.008
+alpha.21 <- 0.008725
 # model function
-disc_log <- function(r, N0, alpha) {
-    Nt1 <- (r*N0) / (1+alpha*N0)
-    return(Nt1)
+disc_LV_comp <- function(r1,r2,N1.0,N2.0,alpha.11,alpha.22,alpha.12,alpha.21) {
+    Nt1 <- (r1*N1.0) / (1+alpha.11*N1.0+alpha.12*N2.0)
+    Nt2 <- (r2*N2.0) / (1+alpha.22*N2.0+alpha.21*N1.0)
+    return(c(Nt1,Nt2))
 }
 # Simulation of model for t time steps
-t <- 30
-N <- rep(NA, t)
-N[1] <- N1.0
+t <- 40
+N <- array(NA,dim=c(t,2))
+N <- as.data.frame(N)
+colnames(N) <- c("N1","N2")
+N$N1[1] <- N1.0
+N$N2[1] <- N2.0
 for (i in 2:t) {
-  N[i] <- disc_log(r=r1.0, N0=N[i-1], alpha=alpha.11)
+  res <- disc_LV_comp(r1=r1.0,r2=r2.0,N1.0=N[i-1,1],N2.0=N[i-1,2],alpha.11=alpha.11,alpha.22=alpha.22,alpha.12=alpha.12,alpha.21=alpha.21)
+  N$N1[i] <- res[1]
+  N$N2[i] <- res[2]
 }
 ```
-
-In the metacommunity simulation in the main text, a species $i$ resides in a site $k$ with an environmental optimum trait value $E_k \sim unif(0,1)$, an initial population size $N_{i,0} \sim Pois(10)$, and an initial degree of maladaptation $\beta_{0,i} \sim Gamma(0.75,1)$.
-From this \(\beta_{0,i}\), the initial distance to the local patch's optimum phenotype was calculated as $d_{0,i}=\sqrt{B_{0,i}(w+P)}$ and the initial trait value for each species $x_{0,i}$ is calculated by subtracting $d_{0,i}$ from the local environmental optimum value $E_k$.
-Here I simulate a species that is initially suited for the environment ($d_{0,i}$)
-
-Internal question 1 - how does the $\alpha$ version compared to the $K$ version?
-
-```{r}
-set.seed(42)
-# Simulate initial species population growth
-N1.0 <- rpois(1,10)
-# Behind-the-scenes initial r1.0
-E1 <- runif(1,0,1)
-beta1.0 <- rgamma(1,0.75,1)
-w <- 10
-P <- 1
-Wmax <- 2
-What <- Wmax*sqrt(w / (P+w))
-h2 <- 0
-d1.0 <- sqrt(beta1.0*(w+P))
-x1.0 <- E1 - d1.0
-# Emergent initial r1.0
-r1.0 <- What*exp( -(((w+(1-h2)*P)/(P+w))*(E1-x1.0))^2/ (2*(P+w)))
-alpha.11 <- 0.00125
-
-# model function
-disc_log <- function(r, N0, alpha) {
-    Nt1 <- (r*N0) / (1+alpha*N0)
-    return(Nt1)
-}
-
-# Simulation of model for t time steps
-N <- rep(NA, t)
-N[1] <- N1.0
-t <- 30
-for (i in 2:t) {
-  N[i] <- disc_log(r=r1.0, N0=N[i-1], alpha=alpha.11)
-    N0 <- N[i]
-}
+```{r figs11, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs1}Population size \\textit{N} over time \\textit{t} for a discrete-time logistic growth model, with parameters $r_i$ = 1.67, $N_{1,0}$ = 14, and $\\alpha_{11}$ = 0.00125.
+Relationship between E and Nt is also shown."}
 # Plot simulation: ggplot
-dat <- as.data.frame(N)
-dat$time <- as.numeric(rownames(dat))
-ggplot2::ggplot(dat, aes(time, N)) + geom_point() + geom_hline(yintercept = (1/alpha.11),
-    linetype = "dashed", color = "gray")
-# Additional plot: how does growth rate change with time?
-range <- seq(0,800,by=0.01)
-rec <- rep(NA, length(range))
-for (i in 1:length(range)) {
-    rec[i] <- r1.0 / (1+alpha.11*range[i])
-}
-plot(y=rec,x=range,ylim=c(-1,2))
+N$time <- 1:t
+dat <- melt(N, id.vars="time")
+ggplot2::ggplot(dat, aes(time, value, col=variable)) + geom_point() + geom_hline(yintercept = ((r1.0 - 1)/alpha.11),linetype = "dashed", color = "gray")
 ```
-
-```{r}
-# Parameter values to use for simulation
-r <- 1.21
-K <- 1400
-N0 <- 4
-t <- 30
-# model function
-disc_log <- function(r, N0, K) {
-    Nt1 <- N0 + r * N0 * (1 - N0/K)
-    return(Nt1)
-}
-# Simulation of model for t time steps
-N <- rep(NA, t)
-for (i in 1:t) {
-    N[i] <- disc_log(r, N0, K)
-    N0 <- N[i]
-}
-# Plot simulation: ggplot
-dat <- as.data.frame(N)
-dat$time <- as.numeric(rownames(dat))
-ggplot2::ggplot(dat, aes(time, N)) + geom_point() + geom_hline(yintercept = K,
-    linetype = "dashed", color = "gray")
-```
-
-$$ N_{i,t+1} =\frac{\hat{W}e^ \frac{-[(\frac{w+(1-h^2)P} {P+w})(E-x_{i,t})]^2}{2(P+w)}N_{i,t}} {1 + \alpha_{ii}N_{i,t} + \sum_{j \neq i}^{S} \alpha_{ij}N_{j,t}} $$
-
-```{r}
-
-```
-
-## Generating the data frames
-For the diagnosis first need to convert the chains into a coda object and extract the effective sample size (ESS), which denotes how many MCMC iteration were independent (e.g.
-having no autocorrelation) from previous iterations, and  potential scale reduction factors (PSRF), which describe how well the MCMC chains converged around a value and if the two chains differ from each other.
-Therefore a high value for ESS (close to the number of actual iterations multiplied by the number of chains) is preferable and a PSRF value of 1 represents both chains converging on the same mean.
-Note we do not check the predictive and explanatory power of the model, as we have a perfect data set by the time series data of the simulations, which leads to $R^2$ > .98.
-```{r eval=FALSE}
-load("../../output/h_01_d_minus3_hmsc_v01.RData")
-m.post <- convertToCodaObject(m.1)
-ess.beta<- effectiveSize(m.post$Beta)
-psrf.beta <- gelman.diag(m.post$Beta, multivariate=FALSE)$psrf
-```
+## Linear statistical model with environmental covariate
+XX
+## Bayesian linear statistical model: HMSC
+XX
+## Conclusions
+XX
 

### 2024-05-04 Appendix_S1_v01.Rmd
began LV competition, observed bifurcation

@@
 ## Linear statistical model with environmental covariate
-XX
+
+We fit each species' population time series to an arimaX and lm with population size of the others species as a covariate.
+
+$$ N_{1,t} = \beta_0 + \beta_1 N_{1,t-1} + \beta_2 N_{2,t-1}+ \epsilon_t $$
+
+```{r figs12, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs1}Population size \\textit{N} over time \\textit{t} for a discrete-time logistic growth model, with parameters $r_i$ = 1.67, $N_{1,0}$ = 14, and $\\alpha_{11}$ = 0.00125.
+Relationship between E and Nt is also shown."}
+# Plot simulation: ggplot
+dat <- as.data.frame(cbind(N$N1,N$N2))
+colnames(dat) <- c("N1","N2")
+dat$time <- 1:t
+p1 <- ggplot2::ggplot(dat, aes(time, N1)) + geom_point() + geom_hline(yintercept = ((r1.0 - 1)/alpha.11),linetype = "dashed", color = "gray")
+p2 <- ggplot2::ggplot(dat, aes(time, N2)) + geom_point()
+p1 + p2
+```
+
+```{r,warning=FALSE}
+## Fit the model for Species 1
+m.3.ar.n1 <- arima(x = log(N$N1), order = c(1,0,0), include.mean = T, method="CSS", xreg=N$N2)
+m.3.lm.n1 <- lm(log(dat$N1[2:t]) ~ log(dat$N1[1:(t-1)]) + log(dat$N2[1:(t-1)]))
+#plotting the series along with the fitted values
+m.3.ar.fit.n1 <- log(N$N1) - residuals(m.3.ar.n1)
+m.3.lm.fit.n1 <- log(dat$N1[2:t]) - m.3.lm.n1$resid
+dat$ar3.fit.n1 <- m.3.ar.fit.n1
+dat$lm3.fit.n1 <- NA
+dat$lm3.fit.n1[2:t] <- m.3.lm.fit.n1
+## Species 2
+m.3.ar.n2 <- arima(x = log(N$N2), order = c(1,0,0), include.mean = T, method="CSS", xreg=N$N1)
+m.3.lm.n2 <- lm(log(dat$N2[2:t]) ~ log(dat$N2[1:(t-1)]) + log(dat$N1[1:(t-1)]))
+#plotting the series along with the fitted values
+m.3.ar.fit.n2 <- log(N$N2) - residuals(m.3.ar.n2)
+m.3.lm.fit.n2 <- log(dat$N2[2:t]) - m.3.lm.n2$resid
+dat$ar3.fit.n2 <- m.3.ar.fit.n2
+dat$lm3.fit.n2 <- NA
+dat$lm3.fit.n2[2:t] <- m.3.lm.fit.n2
+```
+
+```{r figs12, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs2}Population size over time (black line) with fitted values from a first-order autoregressive model (red dashed line).",warning=FALSE}
+colors <- c("data" = "black", "AR1 fit" = "gray", "lm fit"="orange")
+p1 <- ggplot(data=dat,aes(x=time)) +
+  geom_point(aes(y=log(N1),color="data")) +
+  geom_point(aes(y=ar3.fit.n1, color = "AR1 fit"), alpha=0.5) +
+  geom_point(aes(y=lm3.fit.n1, color = "lm fit"), alpha=0.5) +
+  geom_line(aes(y=log(N1), color="data"), linewidth=.4) +
+  geom_line(aes(y=ar3.fit.n1, color = "AR1 fit"), alpha=0.5, linewidth=.4) +
+  geom_line(aes(y=lm3.fit.n1, color = "lm fit"), alpha=0.5, linewidth=.4) +
+  theme_bw() +
+  xlab("time") +
+  ylab("N1") +
+  labs(color="Legend") +
+  scale_color_manual(values = colors)
+p2 <- ggplot(data=dat,aes(x=time)) +
+  geom_point(aes(y=log(N2),color="data")) +
+  geom_point(aes(y=ar3.fit.n2, color = "AR1 fit"), alpha=0.5) +
+  geom_point(aes(y=lm3.fit.n2, color = "lm fit"), alpha=0.5) +
+  geom_line(aes(y=log(N2), color="data"), linewidth=.4) +
+  geom_line(aes(y=ar3.fit.n2, color = "AR1 fit"), alpha=0.5, linewidth=.4) +
+  geom_line(aes(y=lm3.fit.n2, color = "lm fit"), alpha=0.5, linewidth=.4) +
+  theme_bw() +
+  xlab("time") +
+  ylab("N2") +
+  labs(color="Legend") +
+  scale_color_manual(values = colors)
+p1+p2
+```
+
+```{r figs13, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs3}Population size (logarithm) at one time step $N_{t+1}$ as a function of log-population size in the previous time step $N_t$.
+"}
+dN1 <- rep(NA, (t-1))
+for (i in 1:(t-1)) {
+  dN1[i] <- log(N$N1[i+1])-log(N$N1[i])
+}
+d.dN1 <- as.data.frame(dN1)
+d.dN1$Nt <- log(dat$N1[1:(t-1)])
+d.dN1$Nt_plus_1 <- log(dat$N1[2:t])
+p1 <- ggplot2::ggplot(d.dN1, aes(x=Nt,y=Nt_plus_1)) + geom_point()
+
+d.dN1 <- as.data.frame(dN1)
+d.dN1$Nt <- log(dat$N1[1:(t-1)])
+d.dN1$Nt_plus_1 <- log(dat$N1[2:t])
+d.dN1$N2 <- log(dat$N2[2:t])
+p2 <- ggplot2::ggplot(d.dN1, aes(x=N2,y=Nt_plus_1)) + geom_point()
+p3 <- ggplot2::ggplot(d.dN1, aes(Nt, dN1)) + geom_point()
+p4 <- ggplot2::ggplot(d.dN1, aes(N2, dN1)) + geom_point()
+
+p1+p2+p3+p4
+```
+
+XXXXX
+
+```{r}
+df <- data.frame(cbind(log(dat$N1[2:t]),log(dat$N1[1:(t-1)]),log(dat$N2[1:(t-1)])))
+colnames(df) <- c("Nt1","N1","N2")
+m.3.lm <- lm(Nt1 ~ N1 + N2,data=df)
+```
+
+```{r figs9, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs3}Population size (logarithm) at one time step $N_{t+1}$ as a function of log-population size in the previous time step $N_t$.
+"}
+jtools::effect_plot(m.3.lm,pred=N2,plot.points = TRUE)
+ggplot(data = df, aes(x = N2, y = Nt1)) +
+  geom_point() +
+  geom_smooth(method = "lm", formula = y ~ x)
+```
+UH OH !!!!
+
+## Bayesian linear statistical model: HMSC
+We can estimate the same model parameters using HMSC:
+
+```{r}
+# prepare data in HMSC format
+Y <- as.matrix(log(dat$N[2:t]))
+XData <- df
+m.2.hmsc <- Hmsc(Y=Y,XData=XData,XFormula=~Nt+E+Esq)
+# Bayesian model parameters
+nChains <- 2
+thin <- 5
+samples <- 1000
+transient <- 500*thin
+verbose <- 500*thin
+# sample MCMC
+m.2.sample <- sampleMcmc(m.2.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
+```
+
+```{r}
+m2.post.hmsc <- convertToCodaObject(m.2.sample)
+summary(m2.post.hmsc$Beta)
+bayesplot::mcmc_trace(m2.post.hmsc$Beta)
+bayesplot::mcmc_areas(m2.post.hmsc$Beta,area_method = c("equal height"))
+```
+
+These estimates match well with those from the AR1 and linear model:
+
+```{r}
+# AR1 coefficients (recall that the intercept is the term below multipled by 1 - phi1)
+m.2.ar$coef
+m.2.ar$coef[2]*(1-m.2.ar$coef[1])
+# linear model
+summary(m.2.lm)$coefficients[1:4,1:2]
+# Bayesian estimates
+summary(m2.post.hmsc$Beta)$statistics[1:4,1:2]
+```
+
+We recall that the interpretation of the coefficients in an arimaX (arima with covariates) model is difficult.
+They do not give the impact on $N_t$ per unit increase in X as in a regression.
+So we do not interpret the causation implied by the coefficient in the arimaX model.
+In the regression model, we can see that $E$ has a positive impact on $N_t$.
+
+```{r figs10, fig.pos = "H", fig.show="hold", fig.cap = "\\label{fig:figs5} Observed (grey) and model-fit (blue) values for population size at time t (x-axis) and t+1 (y-axis).",results='hide',fig.keep='all'}
+Gradient <- constructGradient(m.2.sample,focalVariable="E",non.focalVariables=list(Nt=list(2),Esq=list(2)),ngrid=39)
+predY <- predict(m.2.sample,XData=Gradient$XDataNew,expected=TRUE)
+plotGradient(m.2.sample,Gradient,pred=predY,showData=T,measure="Y",main="",xlab="E_t",ylab="predicted N_t+1")
+# Can't figure out curved gradient using E and E^2
+#pr <- predict(m.2.hmsc,m.2.sample,)
+```
+
 ## Bayesian linear statistical model: HMSC

### 2024-05-11 Appendix_S1_v01.Rmd
Used HMSC for species interactions + E change

@@
 # One species, logistic growth
-Population growth over time in a single species is first modelled using a Beverton-Holt (discrete-time, logistic) model (@Beverton1957), using an intra-specific competition coefficient for density-dependent growth (@Hart2013).
+Population growth over time in a single species is first modelled using a Beverton-Holt (discrete-time, logistic) model (@Beverton1957), using an intra-specific competition coefficient for density-dependent growth (@Hart2013) - thus $\alpha$ = $1/K$.
 
@@
 
-We fit the population time series data to a first-order auto-regressive model to predict $N_{t+1}$ as a function of $N_t$, and compare that to a linear regression:
+We fit the population time series data to a first-order auto-regressive model to predict $N_{t+1}$ as a function of $N_t$, and compare that to a linear regression.
+We use ln-N after @Ives1995, as also discussed in @Certain2018 and @Olivenca2021.
 
@@
 
-```{r figs12, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs2}Population size over time (black line) with fitted values from a first-order autoregressive model (red dashed line).",warning=FALSE}
+```{r figs13, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs2}Population size over time (black line) with fitted values from a first-order autoregressive model (red dashed line).",warning=FALSE}
 colors <- c("data" = "black", "AR1 fit" = "gray", "lm fit"="orange")
@@
 
-```{r figs13, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs3}Population size (logarithm) at one time step $N_{t+1}$ as a function of log-population size in the previous time step $N_t$.
+```{r figs14, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs3}Population size (logarithm) at one time step $N_{t+1}$ as a function of log-population size in the previous time step $N_t$.
 "}
@@
 colnames(df) <- c("Nt1","N1","N2")
-m.3.lm <- lm(Nt1 ~ N1 + N2,data=df)
+m.3.lm <- lm(Nt1 ~ I(1/N1) + I(1/N2),data=df)
+
+cor(df$Nt1, predict(m.3.lm))
+summary(m.3.lm)
+xx <- seq(0,100, length=1000)
+prediction<-data.frame(N2=xx,N1=100)
+plot(df$N2, df$Nt1, xlab = "N2", ylab="N.t+1" , pch=16)
+lines(prediction$N2, predict(m.3.lm, prediction), lty=2,col="red",lwd=3)
+
 ```
 
-```{r figs9, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs3}Population size (logarithm) at one time step $N_{t+1}$ as a function of log-population size in the previous time step $N_t$.
+```{r figs15, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs3}Population size (logarithm) at one time step $N_{t+1}$ as a function of log-population size in the previous time step $N_t$.
 "}
@@
 ```
-UH OH !!!!
+We can already see that the linear model does not provide a useful fit to the data points (neither does taking the inverse function help).
+The regression parameters can predict well the population size of the other species, but they are not informative for inference.
+This has been observed previously (@Certain2018), and non-linear least squares models (among others) are more often used to estimate parameters for dynamics resulting from these kinds of species interaction models (@Kloppers2013; @Muhlbauer2020; @Olivenca2021).
+The inability of the linear models to reproduce the interaction coefficients is more clearly demonstrated in @Certain2018.
+However, they do indicate that the slope of the linear model can reflect the direction of effect for the interacting species.
+
+Instead of further exploring techniques to fit to time-series derived from non-linear competition processes, we look more closely at how HMSC manages to fit data emerging from competition dynamics and make inference about species interactions.
+Instead of estimating full interaction coefficients, or assuming sparse interactions, they instead assume that a reduced number of linear combinations of species abundances that are most relevant to determining future growth rates for species in the community can bypass the 'curse of dimensionality' problem.
+This is well presented in @Ovaskainen2017.
+We apply that approach here:
 
 ## Bayesian linear statistical model: HMSC
-We can estimate the same model parameters using HMSC:
+We can fit the data to a linear model using HMSC, with latent variables to capture the species associations. These are implemented by including a random effect at the time level.
 
@@
 # prepare data in HMSC format
-Y <- as.matrix(log(dat$N[2:t]))
-XData <- df
-m.2.hmsc <- Hmsc(Y=Y,XData=XData,XFormula=~Nt+E+Esq)
+Y <- as.matrix(cbind(log(dat$N1[2:t]),log(dat$N2[2:t])))
+XData <- df[,2:3]
+studyDesign = data.frame(sample = as.factor(1:(t-1)))
+rL = HmscRandomLevel(units = studyDesign$sample)
+m.3.hmsc = Hmsc(Y = Y, XData = XData, studyDesign = studyDesign, ranLevels = list(sample = rL))
 # Bayesian model parameters
@@
 # sample MCMC
-m.2.sample <- sampleMcmc(m.2.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
+m.3.sample <- sampleMcmc(m.3.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
 ```
@@
 ```{r}
-m2.post.hmsc <- convertToCodaObject(m.2.sample)
-summary(m2.post.hmsc$Beta)
-bayesplot::mcmc_trace(m2.post.hmsc$Beta)
-bayesplot::mcmc_areas(m2.post.hmsc$Beta,area_method = c("equal height"))
+m3.post.hmsc <- convertToCodaObject(m.3.sample)
+summary(m3.post.hmsc$Beta)
+bayesplot::mcmc_trace(m3.post.hmsc$Beta)
+bayesplot::mcmc_areas(m3.post.hmsc$Beta,area_method = c("equal height"))
+
+OmegaCor = computeAssociations(m.3.sample)
+supportLevel = 0.95
+toPlot = ((OmegaCor[[1]]$support>supportLevel)
++ (OmegaCor[[1]]$support<(1-supportLevel))>0)*OmegaCor[[1]]$mean
+corrplot::corrplot(toPlot, method = "color",col = colorRampPalette(c("blue","white","red"))(200),title = paste("random effect level:", m.3.sample$rLNames[1]), mar=c(0,0,1,0))
 ```
@@
 # AR1 coefficients (recall that the intercept is the term below multipled by 1 - phi1)
-m.2.ar$coef
-m.2.ar$coef[2]*(1-m.2.ar$coef[1])
+m.3.ar.n1$coef
+m.3.ar.n1$coef[2]*(1-m.3.ar.n1$coef[1])
 # linear model
-summary(m.2.lm)$coefficients[1:4,1:2]
+summary(m.3.lm.n1)$coefficients[1:3,1:2]
 # Bayesian estimates
-summary(m2.post.hmsc$Beta)$statistics[1:4,1:2]
+summary(m3.post.hmsc$Beta)$statistics[1:3,1:2]
 ```
 
-We recall that the interpretation of the coefficients in an arimaX (arima with covariates) model is difficult.
-They do not give the impact on $N_t$ per unit increase in X as in a regression.
-So we do not interpret the causation implied by the coefficient in the arimaX model.
-In the regression model, we can see that $E$ has a positive impact on $N_t$.
-
-```{r figs10, fig.pos = "H", fig.show="hold", fig.cap = "\\label{fig:figs5} Observed (grey) and model-fit (blue) values for population size at time t (x-axis) and t+1 (y-axis).",results='hide',fig.keep='all'}
-Gradient <- constructGradient(m.2.sample,focalVariable="E",non.focalVariables=list(Nt=list(2),Esq=list(2)),ngrid=39)
-predY <- predict(m.2.sample,XData=Gradient$XDataNew,expected=TRUE)
-plotGradient(m.2.sample,Gradient,pred=predY,showData=T,measure="Y",main="",xlab="E_t",ylab="predicted N_t+1")
-# Can't figure out curved gradient using E and E^2
-#pr <- predict(m.2.hmsc,m.2.sample,)
+```{r}
+# Simulate initial species population growth with environment fluctuations
+N1.0 <- 10
+N2.0 <- 10
+r1.0 <- 1.7
+r2.0 <- 1.7
+alpha.11 <- 0.01
+alpha.22 <- 0.01
+alpha.12 <- 0.005
+alpha.21 <- 0.01
+# model function
+disc_LV_comp <- function(r1,r2,N1.0,N2.0,alpha.11,alpha.22,alpha.12,alpha.21) {
+    Nt1 <- (r1*N1.0) / (1+alpha.11*N1.0+alpha.12*N2.0)
+    Nt2 <- (r2*N2.0) / (1+alpha.22*N2.0+alpha.21*N1.0)
+    return(c(Nt1,Nt2))
+}
+# Simulation of model for t time steps
+t <- 40
+N <- array(NA,dim=c(t,2))
+N <- as.data.frame(N)
+colnames(N) <- c("N1","N2")
+N$N1[1] <- N1.0
+N$N2[1] <- N2.0
+for (i in 2:t) {
+  res <- disc_LV_comp(r1=r1.0,r2=r2.0,N1.0=N[i-1,1],N2.0=N[i-1,2],alpha.11=alpha.11,alpha.22=alpha.22,alpha.12=alpha.12,alpha.21=alpha.21)
+  N$N1[i] <- res[1]
+  N$N2[i] <- res[2]
+}
+```
+```{r}
+# Plot simulation: ggplot
+N$time <- 1:t
+dat <- melt(N, id.vars="time")
+ggplot2::ggplot(dat, aes(time, value, col=variable)) + geom_point() + geom_hline(yintercept = ((r1.0 - 1)/alpha.11),linetype = "dashed", color = "gray")
 ```
 
-## Bayesian linear statistical model: HMSC
-XX
-## Conclusions
-XX
+```{r}
+# Plot simulation: ggplot
+dat <- as.data.frame(cbind(N$N1,N$N2))
+colnames(dat) <- c("N1","N2")
+dat$time <- 1:t
+df <- data.frame(cbind(log(dat$N1[2:t]),log(dat$N2[2:t])))
+colnames(df) <- c("Nt1","Nt2")
+# prepare data in HMSC format
+Y <- as.matrix(cbind(df$Nt1,df$Nt2))
+XData <- data.frame(cbind(log(dat$N1[1:(t-1)]),log(dat$N2[1:(t-1)])))
+studyDesign = data.frame(sample = as.factor(1:(t-1)))
+rL = HmscRandomLevel(units = studyDesign$sample)
+m.4.hmsc = Hmsc(Y = Y, XData = XData, XFormula = ~1, studyDesign = studyDesign, ranLevels = list(sample = rL))
+# Bayesian model parameters
+nChains <- 2
+thin <- 5
+samples <- 1000
+transient <- 500*thin
+verbose <- 500*thin
+# sample MCMC
+m.4.sample <- sampleMcmc(m.4.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
+```
+
+```{r}
+m4.post.hmsc <- convertToCodaObject(m.4.sample)
+summary(m4.post.hmsc$Beta)
+bayesplot::mcmc_trace(m4.post.hmsc$Beta)
+bayesplot::mcmc_areas(m4.post.hmsc$Beta,area_method = c("equal height"))
+
+OmegaCor = computeAssociations(m.4.sample)
+supportLevel = 0.95
+toPlot = ((OmegaCor[[1]]$support>supportLevel)
++ (OmegaCor[[1]]$support<(1-supportLevel))>0)*OmegaCor[[1]]$mean
+corrplot::corrplot(toPlot, method = "color",col = colorRampPalette(c("blue","white","red"))(200),title = paste("random effect level:", m.4.sample$rLNames[1]), mar=c(0,0,1,0))
+```
+
+The HMSC model can estimate the species associations well.
+
+```{r}
+# explanatory power
+preds = computePredictedValues(m.4.sample)
+#evaluateModelFit(hM = m.4.sample, predY = preds)
+# predictive power / cross-validation
+#partition = createPartition(m.4.sample, nfolds = 2)
+#preds = computePredictedValues(m.4.sample, partition = partition, nParallel = nChains)
+#evaluateModelFit(hM = m.4.sample, predY = preds)
+## xx
+#preds = computePredictedValues(m.4.sample, partition=partition, partition.sp=c(1,2), mcmcStep=10, nParallel = nChains)
+#evaluateModelFit(hM=m.4.sample, predY=preds)
+# xx
+#etaPost=getPostEstimate(m.4.sample, "Eta")
+#lambdaPost=getPostEstimate(m.4.sample, "Lambda")
+#biPlot(m.4.sample, etaPost = etaPost, lambdaPost = lambdaPost, factors = c(1,2),"X1")
+```
+```{r}
+df$n1_25 <- NA
+df$n1_975 <- NA
+df$n2_25 <- NA
+df$n2_975 <- NA
+for(i in 1:39){
+  red <- preds[i,,]
+  a <- apply(red,1,quantile,probs=c(.025,.975))
+  df$n1_25[i] <- a[1,1]
+  df$n1_975[i] <- a[2,1]
+  df$n2_25[i] <- a[1,2]
+  df$n2_975[i] <- a[2,2]
+}
+plot(df$Nt1,pch=19,col="black",ylim=c(0,5))
+points(df$Nt2,pch=19,col="orange")
+lines(df$n1_25,col="blue")
+lines(df$n1_975,col="blue")
+lines(df$n2_25,col="blue")
+lines(df$n2_975,col="blue")
+```
+
+# 4 species growth, competition, with environmental change
+
+The goal of the process was to use HMSC's latent variable approach to study the results of non-linear processes in ecology using a linear model.
+We can infer the direction and magnitude of the species interaction.
+We now consider an environmental covariate that changes over time and impacts each species growth.
+We will use HMSC from this point forward.
+
+```{r}
+# Simulate initial species population growth with environment fluctuations
+N1.0 <- 10
+N2.0 <- 10
+r1.0 <- 1.7
+r2.0 <- 1.7
+alpha.11 <- 0.01
+alpha.22 <- 0.01
+alpha.12 <- 0.005
+alpha.21 <- 0.01
+E.0 <- 0.8
+x1.0 <- 0.6
+x2.0 <- 0.8
+
+# model function
+disc_LV_E <- function(r1,r2,N1.0,N2.0,alpha.11,alpha.22,alpha.12,alpha.21,E,x1,x2) {
+    Nt1 <- ((r1*exp(-(E-x1)^2))*N1.0) / (1+alpha.11*N1.0+alpha.12*N2.0)
+    Nt2 <- ((r2*exp(-(E-x2)^2))*N2.0) / (1+alpha.22*N2.0+alpha.21*N1.0)
+    return(c(Nt1,Nt2))
+}
+# Simulation of model for t time steps
+t <- 40
+N <- array(NA,dim=c(t,2))
+N <- as.data.frame(N)
+colnames(N) <- c("N1","N2")
+N$N1[1] <- N1.0
+N$N2[1] <- N2.0
+E <- rep(NA, t)
+E[1] <- E.0
+for (i in 2:t) {
+  res <- disc_LV_E(r1=r1.0,r2=r2.0,N1.0=N[i-1,1],N2.0=N[i-1,2],alpha.11=alpha.11,alpha.22=alpha.22,alpha.12=alpha.12,alpha.21=alpha.21,E=E[i-1],x1=x1.0,x2=x2.0)
+  N$N1[i] <- res[1]
+  N$N2[i] <- res[2]
+  E[i] <- E[i-1] + rnorm(1,0,.1)
+}
+```
+```{r}
+# Plot simulation: ggplot
+N$time <- 1:t
+dat <- melt(N, id.vars="time")
+ggplot2::ggplot(dat, aes(time, value, col=variable)) + geom_point() + geom_hline(yintercept = ((r1.0 - 1)/alpha.11),linetype = "dashed", color = "gray")
+```
+```{r}
+# Plot simulation: ggplot
+dat <- as.data.frame(cbind(log(N$N1),log(N$N2)))
+colnames(dat) <- c("N1","N2")
+dat$time <- 1:t
+df <- data.frame(cbind(dat$N1[2:t],dat$N2[2:t]))
+colnames(df) <- c("Nt1","Nt2")
+# prepare data in HMSC format
+Y <- as.matrix(cbind(df$Nt1,df$Nt2))
+XData <- data.frame(cbind(dat$N1[1:(t-1)],dat$N2[1:(t-1)]),E[1:(t-1)],E[1:(t-1)]^2)
+colnames(XData) <- c("n1","n2","E","Esq")
+studyDesign = data.frame(sample = as.factor(1:(t-1)))
+rL = HmscRandomLevel(units = studyDesign$sample)
+m.5.hmsc = Hmsc(Y = Y, XData = XData, XFormula = ~E+Esq, studyDesign = studyDesign, ranLevels = list(sample = rL))
+# Bayesian model parameters
+nChains <- 2
+thin <- 5
+samples <- 1000
+transient <- 500*thin
+verbose <- 500*thin
+# sample MCMC
+m.5.sample <- sampleMcmc(m.5.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
+```
+
+```{r}
+m5.post.hmsc <- convertToCodaObject(m.5.sample)
+summary(m5.post.hmsc$Beta)
+bayesplot::mcmc_trace(m5.post.hmsc$Beta)
+bayesplot::mcmc_areas(m5.post.hmsc$Beta,area_method = c("equal height"))
+```
+
+```{r}
+# Bayesian estimates
+summary(m5.post.hmsc$Beta)$statistics[1:4,1:2]
+```
+
+```{r}
+OmegaCor = computeAssociations(m.5.sample)
+supportLevel = 0.7
+toPlot = ((OmegaCor[[1]]$support>supportLevel)
++ (OmegaCor[[1]]$support<(1-supportLevel))>0)*OmegaCor[[1]]$mean
+corrplot::corrplot(toPlot, method = "color",col = colorRampPalette(c("blue","white","red"))(200),title = paste("random effect level:", m.5.sample$rLNames[1]), mar=c(0,0,1,0))
+```
+
+```{r}
+Gradient <- constructGradient(m.5.sample,focalVariable="E",non.focalVariables=list(Esq=list(2)),ngrid=39)
+predY <- predict(m.5.sample,XData=Gradient$XDataNew,expected=TRUE)
+a <- plotGradient(m.5.sample,Gradient,pred=predY,showData=T,measure="Y",index=1,main="",xlab="E_t",ylab="predicted N1_t+1")
+b <- plotGradient(m.5.sample,Gradient,pred=predY,showData=T,measure="Y",index=2,main="",xlab="E_t",ylab="predicted N2_t+1")
+
+postBeta = getPostEstimate(m.5.sample, parName = "Beta")
+plotBeta(m.5.sample, post = postBeta, param = "Mean", supportLevel = 0.7)
+```
+
+XX Interpret environmental terms
 

### 2024-05-17 manuscript/manuscript_v11.Rmd
Added new draft mscript

---
title: "Detection of eco-evolutionary dynamics in metacommunities using Joint Species
  Distribution Models"
author: "Jelena H. Pantel, Ruben J. Hermann"
date: "Ecological Modelling, Faculty of Biology, University of Duisburg-Essen, Universitätsstraße
  5, 45141 Essen, Germany"
output:
  bookdown::pdf_document2:
    toc: no
    fig_caption: yes
    latex_engine: xelatex
    number_sections: true
  bookdown::html_document2: default
  word_document:
    reference_docx: docx_template.docx
header-includes:
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \fancyhead[L]{Author XXX}
- \fancyhead[R]{Detection of ecoevo dynamics}
- \usepackage{lineno}
- \linenumbers
linestretch: 2
link-citations: yes
linkcolor: blue
csl: the-american-naturalist.csl
bibliography: refs.bib
---

```{r setup,echo=FALSE,warning=FALSE,message=FALSE}
library(R.matlab)
library(Hmsc)
```

\newpage

Title
===============================================================================

Detection of eco-evolutionary dynamics in metacommunities using Joint Species
  Distribution Models

Abstract
===============================================================================

XXXXX

Introduction
================================================================================

XXXXX

Methods
================================================================================

## Simulation model

We simulated growth and competition dynamics for a multi-species assemblage in a patchy landscape, with site variation in one environmental property (Figure \@ref(fig:fig1)).

```{r fig1, echo = FALSE, eval = TRUE, fig.width = 5, fig.height = 5, fig.cap = "Spatial maps showing position of all 50 sites in the simulation, their value for the single environmental variable E, and their value for initial Simpson's diversity at the beginning of the simulation.", warning=FALSE,message=FALSE}

## Figure 1. Plot of location of sites in landscape, associated E values, and associated diversity values
result <- paste("../data/h_01_d_minus3_res.mat",sep="")
dat  <- R.matlab::readMat(toString(result))

layout(matrix(c(1,3,2,4),ncol=2), width = c(2,1),height = c(1,1))
pal <- colorRampPalette(c("white", "blue"))
plot(dat$xy,col="grey90", fill=TRUE, warnings=F, xlab="X coord",ylab="Y coord")
options(warn=0)
points(dat$xy[,1], dat$xy[,2],pch = 21,bg = pal(10)[as.numeric(cut(dat$E,breaks = 10))],col="black")

legend_image <- as.raster(matrix(rev(pal(10)), ncol=1))
plot(c(0,2),c(0,1),type = 'n', axes = F,xlab = '', ylab = '', main = 'E value')
text(x=1.5, y = seq(0,1,l=5), labels = seq(0,1,l=5))
rasterImage(legend_image, 0, 0, 1, 1) ## set new for overplot w/ next plot

pal <- colorRampPalette(c("white", "orange"))
a  <- vegan::diversity(dat$N[,,1],index="invsimpson")
a[is.finite(a)==FALSE] <- 0
plot(dat$xy,col="grey90", fill=TRUE, warnings=F, xlab="X coord",ylab="Y coord")
options(warn=0)
points(dat$xy[,1], dat$xy[,2],pch = 21,bg = pal(10)[as.numeric(cut(a,breaks = 10))],col="black")

legend_image <- as.raster(matrix(rev(pal(10)), ncol=1))
plot(c(0,2),c(0,1),type = 'n', axes = F,xlab = '', ylab = '', main = 'inverse Simpson diversity')
text(x=1.5, y = seq(0,1,l=6), labels = seq(0,15,l=6))
rasterImage(legend_image, 0, 0, 1, 1, cex=0.75) ## set new for overplot w/ next plot
```

### Environmental variation

Population growth for species in the metacommunity simulation follows a Leslie-Gower model (a discrete-time version of a Lotka-Volterra model [@Beverton1957] [@Leslie1958]). We consider the impact of trait evolution for growth using a discrete time quantitative genetic model of evolutionary rescue [@Gomulkiewicz1995]. The model for population size is as follows:

$$ N_{i,t+1} =\frac{\hat{W}e^ \frac{-[(\frac{w+(1-h^2)P} {P+w})(E-x_{i,t})]^2}{2(P+w)}N_{i,t}} {1 + \alpha_{ii}N_{i,t} + \sum_{j \neq i}^{S} \alpha_{ij}N_{j,t}} $$
where $N_i,t$ is the population size of species *i* at time *t*, \( \hat{W} \) is calculated as \( \hat{W}=W_{max}\sqrt(\frac{w}{P+w}) \), \(W_{max}\) is the species' maximum per-capita growth rate, *w* is the width of the Gaussian fitness function (which determines the strength of selection, as increasing values indicate a weaker reduction in fitness with distance from optimum trait value), *P* is the width of the distribution of the phenotype *x*, $h^2$ is the heritability of the trait *x*, *E* is the local environmental optimum trait value, $x_{i,t}$ is the trait value of species *i* at time *t*, $\alpha_{ii}$ is the intraspecific competition coefficient (the per capita impact of species *i* on itself) and $\alpha_{ij}$ is the interspecific competition coefficient. Populations have a critical density $N_c$, below which the population is subject to extinction due to demographic stochasticity at a probability of *p* [@Gomulkiewicz1995].

This model was used in Pantel & Becks [@Pantel2023] to evaluate the consequences of adaptive evolution for coexistence in a three-species system. To expand this model to a metacommunity, we consider the evolution of multiple species (*S* = 15) in a landscape of patches (*k* = 50). The patches have values of an environmental property *E* (drawn from a uniform distribution ranging from 0 to 1) that determines the local optimum phenotype *E* = $x_{opt}$, i.e. where species experience their absolute fitness $W_{max}$, and patches also have spatial locations *X* and *Y* (both drawn from U(0, 1)). The patches thus have a connectivity matrix **D** (here given by their Euclidean distance), as well as a connectivity matrix **C** that is a Gaussian function of **D** and species dispersal rate $d_i$.

To simulate trait evolution together with species growth and competition in the landscape, we used parameter values of \(W_{max}\) = 2, *P* = 1, *w* = 10, $N_c$ = 100, and *p* = 0.001. The simulated metacommunity was initialized in a way that mimics community assembly in small vernal pools that periodically flood (and thus are reset seasonally). In the simulated pool system, local communities are drawn at random from the regional species pool, and thus their initial degree of maladaptation in the pool is random as well. The initial species richness of each site $s_{0k}$ was drawn from a random Poisson distribution, $s_{0k} \sim Pois(0.75)$, the initial population size for all species $N_{0i} \sim Pois(10)$, and the initial degree of maladaptation \(\beta_{i0} \sim Gamma(0.75,1)\). From this \(\beta_{i0}\), the initial distance to the local patch's optimum phenotype was calculated as $d_{i0}=\sqrt{B_{i0}(w+P)}$ and the initial trait value for each species $x_{i0}$ was calculated by subtracting $d_{i0}$ from the local environmental optimum value $E_k$ (Appendix S1, Figure S1.1).

The species interaction matrix was fixed throughout the simulation, with \(\alpha_{ii} = 0.00125\) and \(\alpha_{ij} \sim U(0.0015,1)\). After draws for all random parameter values, the same resulting initial community was used for all simulations, to allow direct comparison across dispersal and heritability values. Simulations were run across a range of dispersal values (identical for all species; *d* = 0, $10^{-9},10^{-8},10^{-7},10^{-6},10^{-5},10^{-4},10^{-3},10^{-2}$, 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 1) and heritability values (identical for all species: $h^2$ = 0, 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 1). The simulations were run in MATLAB (version R2023a) for 1000 time steps and produced records of $N_{ikt}$ and $x_{ikt}$. To test the impacts of strength of selection and environmental variability, we repeated simulations at a single, representative level of heritability and dispersal ($d = 10^{-3}$, $h^2 = 0.1$) in four different scenarios: Scenario 1 (weak selection, fixed environment: $P = 1, w = 10, v[E] = 0$), Scenario 2 (strong selection, fixed environment: $P = 0.1, w = 0.2, v[E] = 0$), Scenario 3 (weak selection, variable environment: $P = 1, w = 10, v[E] = 0.01$), and Scenario 4 (strong selection, variable environment: P = 0.1, w = 0.2, v[E] = 0.01). Environmental variability was implemented by adding random noise to $E_k$ every 10 time steps ($E_{k,t+1} \sim N(E_{k,t},v[E])$).

*Statistical model*

To determine the relative influence of environmental and spatial properties, intra- and interspecific population dynamics, and evolutionary dynamics for community structure, we applied hierarchical modelling of species communities (HMSC, [@Ovaskainen2017HMSC] [@Tikhonov2020]) to model the abundance at each site and year for each species $N_{ikt}$. The abundance values were modeled as $N_{ikt} \sim N(L_{ikt},\sigma_2)$, where the linear predictors **L** are the sum of fixed and random effects **L^F^** + **L^R^** [@Tikhonov2020]. The fixed effects were (1) $E_k$, the value of the site’s environmental parameter (predictor 1), (2) the value of each species' population size the time step before $N_{ik,t-1}$ (predictors 2-16; see [@Sandal2022]), (3) the absolute value of the change in each species' trait value from the previous time step to the next $|\Delta x_{ik,t-1 \to t}|$ (predictors 17-31), (4) and an interaction term for population size by change in trait value $N_{ik,t-1} \times |\Delta x_{ik,t-1 \to t}|$ (predictors 32-46), which led to a total of 46 predictors. The random effects were modeled at the site-level ($\epsilon_{ik} \sim N(0, \Omega^S)$, where $\Omega^S$ is the site species-to-species covariance matrix) and at the time step level ($\epsilon_{it} \sim N(0, \Omega^T)$, where $\Omega^T$ is the temporal species-to-species covariance matrix) using a latent variable approach. In this approach, species-to-species covariance matrices are computed as $\Omega = \lambda \intercal \lambda$, where $\lambda$ are the loadings of each species with latent variables, and latent variables are modeled as $\epsilon_{ik} = \sum_B \eta_{kB} \lambda_{Bi}$ (site) and $\epsilon_{it} = \sum_B \eta_{tB} \lambda_{Bi}$ (time) for *B* latent variables considered [@Ovaskainen2016]. The spatial latent variables represent a spatial model, where species respond to some latent spatial predictor associated with the *xy* coordinates of the sites [@Tikhonov2020]. Analysis and was conducted in R (version 4.2.3) using packages ‘Hmsc’ and 'R.matlab' ([@Bengtsson2022]; [@Tikhonov2022]).

Results
===============================================================================

XXXX

Discussion
================================================================================

XXXXX

References
===============================================================================

<div id="refs"></div>

<a name="appendix">Figures & Tables</a>
================================================================================

### 2024-06-27 manuscript/Appendix_S1_v01.Rmd
More simulation conditions, simple cases PredictGradient working

--- 2024-05-11 Appendix_S1_v01.Rmd
@@
 Gradient <- constructGradient(m.2.sample,focalVariable="E",non.focalVariables=list(Nt=list(2),Esq=list(2)),ngrid=39)
+# Esq is manually constructed as gradient-produced E^2
+Gradient$XDataNew$Esq <- Gradient$XDataNew$E^2
 predY <- predict(m.2.sample,XData=Gradient$XDataNew,expected=TRUE)
 plotGradient(m.2.sample,Gradient,pred=predY,showData=T,measure="Y",main="",xlab="E_t",ylab="predicted N_t+1")
-# Can't figure out curved gradient using E and E^2
-#pr <- predict(m.2.hmsc,m.2.sample,)
 ```

### 2024-07-23 manuscript/Appendix_S1_v01.Rmd
Finished competition + env change

@@
 # Two species, logistic growth, competition
-XXXXX
+Here we investigate how a second species impacts the inference we can make from linear models.
+
 ## Interspecific competition
@@
 ```{r}
-# Simulate initial species population growth with environment fluctuations
+# Simulate initial species population growth with a second species present
 N1.0 <- 10
@@
 ```
-```{r figs11, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs1}Population size \\textit{N} over time \\textit{t} for a discrete-time logistic growth model, with parameters $r_i$ = 1.67, $N_{1,0}$ = 14, and $\\alpha_{11}$ = 0.00125.
-Relationship between E and Nt is also shown."}
+```{r figs11, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs1}Population size \\textit{N} over time \\textit{t} for a discrete-time logistic growth model, with parameters $r_1$ = 1.67, $r_2$ = 1.7, $N_{i,0}$ = 10, $\\alpha_{ii}$ = 0.00125, $\\alpha_{12}$ = 0.008, and $\\alpha_{21}$ = 0.008725."}
 # Plot simulation: ggplot
@@
 ```
-## Linear statistical model with environmental covariate
+## Linear statistical model with covariate for both species
 
@@
 
-```{r figs12, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs1}Population size \\textit{N} over time \\textit{t} for a discrete-time logistic growth model, with parameters $r_i$ = 1.67, $N_{1,0}$ = 14, and $\\alpha_{11}$ = 0.00125.
-Relationship between E and Nt is also shown."}
+```{r figs12, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs1}Population size \\textit{N} over time \\textit{t} for a discrete-time logistic growth model, with competition."}
 # Plot simulation: ggplot
@@
 
-```{r figs13, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs2}Population size over time (black line) with fitted values from a first-order autoregressive model (red dashed line).",warning=FALSE}
+```{r figs13, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs2}Population size over time (black line) with fitted values from a first-order autoregressive model (gray line) and from a linear model (orange line).",warning=FALSE}
 colors <- c("data" = "black", "AR1 fit" = "gray", "lm fit"="orange")
@@
 
-```{r figs14, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs3}Population size (logarithm) at one time step $N_{t+1}$ as a function of log-population size in the previous time step $N_t$.
-"}
+```{r figs14, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs3}Population size (logarithm) at one time step $N_{t+1}$ as a function of log-population size of Species 1 and Species 2 in the previous time step $N_t$.
+Also shown is the change in population size of Species 1 as a function of $N_t$ for both species."}
 dN1 <- rep(NA, (t-1))
@@
 
-XXXXX
+We can see that the population size of Species 2 does not have a linear relationship with that of Species 1.
 
@@
 
-```{r figs15, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs3}Population size (logarithm) at one time step $N_{t+1}$ as a function of log-population size in the previous time step $N_t$.
-"}
+```{r figs15, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs3}Population size (logarithm) at one time step $N_{t+1}$ as a function of log-population size of the second species in the previous time step $N_t$.
+Fitted (line) and observed (points) values are shown."}
 jtools::effect_plot(m.3.lm,pred=N2,plot.points = TRUE)
@@
 bayesplot::mcmc_areas(m3.post.hmsc$Beta,area_method = c("equal height"))
+```
 
+Now we look at the estimates for the species associations:
+
+```{r}
 OmegaCor = computeAssociations(m.3.sample)
@@
 
-These estimates match well with those from the AR1 and linear model:
+The model fixed effect estimates match well with those from the AR1 and linear model.
+However, the model up to this point includes $N_{t1}$ and $N_{t2}$ as fixed predictors.
+I change that here, as it better matches what can be done in large multi-species models, and the residual associations will better match the effects species have on one another.
 
@@
 
+I try again with a next example where the only difference between species is the $\alpha_{ij}$.
+
 ```{r}
-# Simulate initial species population growth with environment fluctuations
+# Simulate initial species population growth with a second species present
 N1.0 <- 10
@@
 
-The HMSC model can estimate the species associations well.
+The HMSC model correctly shows the negative association between the species (shown here as correlations - JSDM CHapter 7 tells us the intr-specific correlation is always 1).
 
@@
 
-# 4 species growth, competition, with environmental change
+The residual associations should not be interpreted as interaction coefficients, but rather used for prediction.
+We can see here they are effective for this (shown above is the observed and model-predicted 95% CI).
+
+# Species growth, competition, with environmental change
 
@@
 We now consider an environmental covariate that changes over time and impacts each species growth.
-We will use HMSC from this point forward.
+We will use only HMSC from this point forward, as the species interactions represent a departure from what a linear or AR model can manage.
+
+In this example, the local environmental optimum trait value $E = 0.8$, and Species 1 $x_1 = 0.6$, while $x_1 = 0.8$.
 
@@
 ```
+We can see that the Environment has an overall negative impact for both species as $E$ drifts randomly towards more positive values (where Species 2 is closer because it has a higher trait value).
+We also see that the residual species association effects are strongly positive.
+We evaluate this by greatly increasing the effects if interspecific competition (while decreasing the strength of environmental variation).
 
-XX Interpret environmental terms
+```{r}
+# Simulate initial species population growth with environment fluctuations
+N1.0 <- 10
+N2.0 <- 10
+r1.0 <- 1.7
+r2.0 <- 1.7
+alpha.11 <- 0.01
+alpha.22 <- 0.01
+alpha.12 <- 0.015
+alpha.21 <- 0.02
+E.0 <- 0.8
+x1.0 <- 0.8
+x2.0 <- 0.8
+
+# model function
+disc_LV_E <- function(r1,r2,N1.0,N2.0,alpha.11,alpha.22,alpha.12,alpha.21,E,x1,x2) {
+    Nt1 <- ((r1*exp(-(E-x1)^2))*N1.0) / (1+alpha.11*N1.0+alpha.12*N2.0)
+    Nt2 <- ((r2*exp(-(E-x2)^2))*N2.0) / (1+alpha.22*N2.0+alpha.21*N1.0)
+    return(c(Nt1,Nt2))
+}
+# Simulation of model for t time steps
+t <- 40
+N <- array(NA,dim=c(t,2))
+N <- as.data.frame(N)
+colnames(N) <- c("N1","N2")
+N$N1[1] <- N1.0
+N$N2[1] <- N2.0
+E <- rep(NA, t)
+E[1] <- E.0
+for (i in 2:t) {
+  res <- disc_LV_E(r1=r1.0,r2=r2.0,N1.0=N[i-1,1],N2.0=N[i-1,2],alpha.11=alpha.11,alpha.22=alpha.22,alpha.12=alpha.12,alpha.21=alpha.21,E=E[i-1],x1=x1.0,x2=x2.0)
+  N$N1[i] <- res[1]
+  N$N2[i] <- res[2]
+  E[i] <- E[i-1] + rnorm(1,0,0.001)
+}
+```
+```{r}
+# Plot simulation: ggplot
+N$time <- 1:t
+dat <- melt(N, id.vars="time")
+ggplot2::ggplot(dat, aes(time, value, col=variable)) + geom_point() + geom_hline(yintercept = ((r1.0 - 1)/alpha.11),linetype = "dashed", color = "gray")
+```
+```{r echo=FALSE}
+# Plot simulation: ggplot
+dat <- as.data.frame(cbind(log(N$N1),log(N$N2)))
+colnames(dat) <- c("N1","N2")
+dat$time <- 1:t
+df <- data.frame(cbind(dat$N1[2:t],dat$N2[2:t]))
+colnames(df) <- c("Nt1","Nt2")
+# prepare data in HMSC format
+Y <- as.matrix(cbind(df$Nt1,df$Nt2))
+XData <- data.frame(cbind(dat$N1[1:(t-1)],dat$N2[1:(t-1)]),E[1:(t-1)],E[1:(t-1)]^2)
+colnames(XData) <- c("n1","n2","E","Esq")
+studyDesign = data.frame(sample = as.factor(1:(t-1)))
+rL = HmscRandomLevel(units = studyDesign$sample)
+m.6.hmsc = Hmsc(Y = Y, XData = XData, XFormula = ~E+Esq, studyDesign = studyDesign, ranLevels = list(sample = rL))
+# Bayesian model parameters
+nChains <- 2
+thin <- 5
+samples <- 2000
+transient <- 100*thin
+verbose <- 500*thin
+# sample MCMC
+m.6.sample <- sampleMcmc(m.6.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
+```
+```{r}
+OmegaCor = computeAssociations(m.6.sample)
+supportLevel = 0.7
+toPlot = ((OmegaCor[[1]]$support>supportLevel)
++ (OmegaCor[[1]]$support<(1-supportLevel))>0)*OmegaCor[[1]]$mean
+corrplot::corrplot(toPlot, method = "color",col = colorRampPalette(c("blue","white","red"))(200),title = paste("random effect level:", m.6.sample$rLNames[1]), mar=c(0,0,1,0))
+OmegaCor
+```
+```{r}
+Gradient <- constructGradient(m.6.sample,focalVariable="E",non.focalVariables=list(Esq=list(2)),ngrid=39)
+predY <- predict(m.6.sample,XData=Gradient$XDataNew,expected=TRUE)
+a <- plotGradient(m.6.sample,Gradient,pred=predY,showData=T,measure="Y",index=1,main="",xlab="E_t",ylab="predicted N1_t+1")
+b <- plotGradient(m.6.sample,Gradient,pred=predY,showData=T,measure="Y",index=2,main="",xlab="E_t",ylab="predicted N2_t+1")
+
+postBeta = getPostEstimate(m.6.sample, parName = "Beta")
+plotBeta(m.6.sample, post = postBeta, param = "Mean", supportLevel = 0.7)
+```
+With the reduced impact of the environment, we now see the strong negative associations between the two competing species. We also correctly see no impact of environment for species abundances.
+
+# Species growth, competition, environmental change, and trait evolution
 

### 2024-07-26 manuscript/Appendix_S1_v01.Rmd
Began Species Evol, added new vignette with evol hypotheses

@@
 We also see that the residual species association effects are strongly positive.
+
+We use variation partition to estimate the relative importance of environment and species associations for species abundances.
+
+```{r warning=FALSE}
+VP <- computeVariancePartitioning(m.5.sample,group=c(1,1,1),groupnames="Env")
+plotVariancePartitioning(m.5.sample,VP,args.legend=list(cex=0.75,bg="transparent"))
+```
+
 We evaluate this by greatly increasing the effects if interspecific competition (while decreasing the strength of environmental variation).
@@
 
+```{r warning=FALSE}
+VP <- computeVariancePartitioning(m.6.sample,group=c(1,1,1),groupnames="Env")
+plotVariancePartitioning(m.6.sample,VP,args.legend=list(cex=0.75,bg="transparent"))
+```
+
 # Species growth, competition, environmental change, and trait evolution
 
+We now consider that trait evolution can occur, with an impact on the species growth rate / fitness in the context of the environment.
+
+## Trait evolution
+
+Previously the population growth rate was fixed as \( r_i = \hat{W}e^{-(E-x_{i,t})^2}\). However, we now introduce the quantitative genetic model of evolutionary rescue [@Gomulkiewicz1995].
+
+The growth equation for each species now becomes:
+
+$$ N_{i,t+1} =\frac{\hat{W}e^ \frac{-[(\frac{w+(1-h^2)P} {P+w})(E-x_{i,t})]^2}{2(P+w)}N_{i,t}} {1 + \alpha_{ii}N_{i,t} + \alpha_{ij} N_{j,t}} $$
+
+where \( \hat{W} \) is calculated as \( \hat{W}=W_{max}\sqrt(\frac{w}{P+w}) \), \(W_{max}\) is the species' maximum per-capita growth rate, *w* is the width of the Gaussian fitness function (which determines the strength of selection, as increasing values indicate a weaker reduction in fitness with distance from optimum trait value), *P* is the width of the distribution of the phenotype *x*, and $h^2$ is the heritability of the trait *x*.
+For the simulation we use \(W_{max}\) = 2, *P* = 1, and *w* = 10.
+
+The change in the average trait value each time step is given by:
+
+$$ d_{i,t+1} = k d_{i,t} $$
+where \( k = \frac{w + (1-h^2) P}{w + P} \) and \( d_{i,t} = E_t - x_{i,t} \).
+
+## Population dynamics simulation
+
+We use the same examples as above, with weaker then stronger strength of interspecific interactions.
+
+```{r}
+# Case 1: Weaker interactions, stronger environment
+# Simulate initial species population growth with environment fluctuations
+N1.0 <- 10
+N2.0 <- 10
+alpha.11 <- 0.01
+alpha.22 <- 0.01
+alpha.12 <- 0.005
+alpha.21 <- 0.01
+E.0 <- 0.8
+x1.0 <- 0.6
+x2.0 <- 0.8
+P <- 1
+w <- 10
+Wmax <- 2
+h2 <- 0.1
+k <- (w+(1-h2)*P)/(P+w)
+
+# model function
+disc_LV_evol <- function(N1.0,N2.0,alpha.11,alpha.22,alpha.12,alpha.21,E,x1.0,x2.0,P,w,Wmax,h2) {
+  What <- Wmax*sqrt(w/(P+w))
+  r1 <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E-x1.0))^2)/(2*(P+w)))
+  Nt1 <- (r1*N1.0) / (1+alpha.11*N1.0+alpha.12*N2.0)
+  r2 <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E-x2.0))^2)/(2*(P+w)))
+  Nt2 <- (r2*N2.0) / (1+alpha.22*N2.0+alpha.21*N1.0)
+  return(c(Nt1,Nt2,r1,r2))
+}
+# Simulation of model for t time steps
+t <- 40
+N <- array(NA,dim=c(t,2))
+N <- as.data.frame(N)
+colnames(N) <- c("N1","N2")
+x <- array(NA,dim=c(t,2))
+x <- as.data.frame(x)
+colnames(x) <- c("x1","x2")
+r <- array(NA,dim=c(t,2))
+r <- as.data.frame(r)
+colnames(r) <- c("r1","r2")
+N$N1[1] <- N1.0
+N$N2[1] <- N2.0
+x$x1[1] <- x1.0
+x$x2[1] <- x2.0
+r$r1[1] <- exp((-(((w+(1-h2)*P)/(P+w))*(E.0-x1.0))^2)/(2*(P+w)))
+r$r2[1] <- exp((-(((w+(1-h2)*P)/(P+w))*(E.0-x2.0))^2)/(2*(P+w)))
+E <- rep(NA, t)
+E[1] <- E.0
+for (i in 2:t) {
+  res <- disc_LV_evol(N1.0=N[i-1,1],N2.0=N[i-1,2],alpha.11=alpha.11,alpha.22=alpha.22,alpha.12=alpha.12,alpha.21=alpha.21,E=E[i-1],x1=x[i-1,1],x2=x[i-1,2],P=P,w=w,Wmax=Wmax,h2=h2)
+  N$N1[i] <- res[1]
+  N$N2[i] <- res[2]
+  r$r1[i] <- res[3]
+  r$r2[i] <- res[4]
+  # trait change
+  d1 <- E[i-1] - x[i-1,1]
+  d2 <- E[i-1] - x[i-1,2]
+  d1.1 <- k*d1
+  d2.1 <- k*d2
+  x$x1[i] <- E[i-1] - d1.1
+  x$x2[i] <- E[i-1] - d2.1
+  # environmental change
+  E[i] <- E[i-1] + rnorm(1,0,.01)
+}
+```
+```{r}
+# Plot simulation: ggplot
+N$time <- 1:t
+dat <- melt(N, id.vars="time")
+ggplot2::ggplot(dat, aes(time, value, col=variable)) + geom_point() + geom_hline(yintercept = ((r1.0 - 1)/alpha.11),linetype = "dashed", color = "gray")
+```
+
+The overall research goal is to determine the effect of trait evolution for species abundances.
+We use the absolute value of the change in trait value from one time point to the next (the total amount of trait change, regardless of direction) as a fixed effect in the HMSC.
+
+```{r}
+# Plot simulation: ggplot
+dat <- as.data.frame(cbind(log(N$N1),log(N$N2),x$x1,x$x2))
+colnames(dat) <- c("N1","N2","x1","x2")
+dat$time <- 1:t
+df <- data.frame(cbind(dat$N1[2:t],dat$N2[2:t],dat$x1[2:t],dat$x2[2:t]))
+colnames(df) <- c("Nt1","Nt2","xt1","xt2")
+# prepare data in HMSC format
+Y <- as.matrix(cbind(df$Nt1,df$Nt2))
+XData <- data.frame(cbind(dat$N1[1:(t-1)],dat$N2[1:(t-1)]),E[1:(t-1)],E[1:(t-1)]^2,abs(dat$x1[2:t] - dat$x1[1:(t-1)]),abs(dat$x2[2:t] - dat$x2[1:(t-1)]))
+colnames(XData) <- c("n1","n2","E","Esq","dx1","dx2")
+studyDesign = data.frame(sample = as.factor(1:(t-1)))
+rL = HmscRandomLevel(units = studyDesign$sample)
+m.7.hmsc = Hmsc(Y = Y, XData = XData, XFormula = ~E+Esq+dx1+dx2, studyDesign = studyDesign, ranLevels = list(sample = rL))
+# Bayesian model parameters
+nChains <- 2
+thin <- 5
+samples <- 2000
+transient <- 100*thin
+verbose <- 500*thin
+# sample MCMC
+m.7.sample <- sampleMcmc(m.7.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
+```
+```{r}
+m7.post.hmsc <- convertToCodaObject(m.7.sample)
+summary(m7.post.hmsc$Beta)
+bayesplot::mcmc_trace(m7.post.hmsc$Beta)
+bayesplot::mcmc_areas(m7.post.hmsc$Beta,area_method = c("equal height"))
+```
+
+```{r}
+# Bayesian estimates
+summary(m7.post.hmsc$Beta)$statistics[1:10,1:2]
+```
+
+```{r}
+OmegaCor = computeAssociations(m.7.sample)
+supportLevel = 0.7
+toPlot = ((OmegaCor[[1]]$support>supportLevel)
++ (OmegaCor[[1]]$support<(1-supportLevel))>0)*OmegaCor[[1]]$mean
+corrplot::corrplot(toPlot, method = "color",col = colorRampPalette(c("blue","white","red"))(200),title = paste("random effect level:", m.7.sample$rLNames[1]), mar=c(0,0,1,0))
+OmegaCor
+```
+
+```{r}
+Gradient <- constructGradient(m.7.sample,focalVariable="E",non.focalVariables=list(Esq=list(2)),ngrid=39)
+predY <- predict(m.7.sample,XData=Gradient$XDataNew,expected=TRUE)
+a <- plotGradient(m.7.sample,Gradient,pred=predY,showData=T,measure="Y",index=1,main="",xlab="E_t",ylab="predicted N1_t+1")
+b <- plotGradient(m.7.sample,Gradient,pred=predY,showData=T,measure="Y",index=2,main="",xlab="E_t",ylab="predicted N2_t+1")
+
+postBeta = getPostEstimate(m.7.sample, parName = "Beta")
+plotBeta(m.7.sample, post = postBeta, param = "Mean", supportLevel = 0.7)
+```
+```{r warning=FALSE}
+VP <- computeVariancePartitioning(m.7.sample,group=c(1,1,1,2,3),groupnames=c("Env","Sp1","Sp2"))
+plotVariancePartitioning(m.7.sample,VP,cols=c("white","skyblue","darkgrey","orange"),args.legend=list(cex=0.75,bg="transparent"))
+```
+We can also look at a gradient plot for the effect of evolution in Species 1 for abundance in Species 1 and Species 2.
+
+```{r}
+Gradient <- constructGradient(m.7.sample,focalVariable="dx1",non.focalVariables=list(E=list(2),Esq=list(2),dx2=list(2)),ngrid=39)
+Gradient$XDataNew$Esq <- Gradient$XDataNew$E^2
+predY <- predict(m.7.sample,XData=Gradient$XDataNew,expected=TRUE)
+a <- plotGradient(m.7.sample,Gradient,pred=predY,showData=T,measure="Y",index=1,main="",xlab="dx1_t",ylab="predicted N1_t+1")
+b <- plotGradient(m.7.sample,Gradient,pred=predY,showData=T,measure="Y",index=2,main="",xlab="dx1_t",ylab="predicted N2_t+1")
+
+postBeta = getPostEstimate(m.7.sample, parName = "Beta")
+plotBeta(m.7.sample, post = postBeta, param = "Mean", supportLevel = 0.7)
+```
+
+We repeat this plot for the effect of evolution in Species 2 for abundance in Species 1 and Species 2.
+
+```{r}
+Gradient <- constructGradient(m.7.sample,focalVariable="dx2",non.focalVariables=list(E=list(2),Esq=list(2),dx1=list(2)),ngrid=39)
+Gradient$XDataNew$Esq <- Gradient$XDataNew$E^2
+predY <- predict(m.7.sample,XData=Gradient$XDataNew,expected=TRUE)
+a <- plotGradient(m.7.sample,Gradient,pred=predY,showData=T,measure="Y",index=1,main="",xlab="dx2_t",ylab="predicted N1_t+1")
+b <- plotGradient(m.7.sample,Gradient,pred=predY,showData=T,measure="Y",index=2,main="",xlab="dx2_t",ylab="predicted N2_t+1")
+```
+
+THESE SHOULD NOT BE THE SAME
+
+We repeat the analysis for
+
 # References

### 2024-07-26 manuscript/Appendix_S2_v01.Rmd
Began Species Evol, added new vignette with evol hypotheses

--- 2024-06-27 manuscript/Appendix_S1_v01.Rmd
@@
 ---
-title: "Appendix S1. Linear regression models and non-linear population dynamics"
+title: "Appendix S2. Evolutionary drivers of species and community dynamics"
 author:
@@
 
-# One species, logistic growth
-Population growth over time in a single species is first modelled using a Beverton-Holt (discrete-time, logistic) model (@Beverton1957), using an intra-specific competition coefficient for density-dependent growth (@Hart2013) - thus $\alpha$ = $1/K$.
+# Background
 
-$$ N_{i,t+1} = \frac{r_i N_{i,t}}{1 + \alpha_{ii} N_{i,t}}$$
-Note that in this model, the system is at equilibrium when $N_{i,t+1} = N_{i,t}$, and therefore: $$ N^* = N^*\frac{r_i}{1 + \alpha_{ii} N^*} $$
+In Appendix S1.5, we evaluated the ability of the statistical model to estimate the impact of trait evolution for species abundances.
+Trait evolution's impact for species abundances can be modeled as a predictor in different ways.
+Here we derive the numerical form of the trait $x_i$ that we hypothesize has the most impact on species abundances.
 
-$$ 1 = \frac{r_i}{1 + \alpha_{ii} N^*} $$
+# Model for population growth, competition, environmental change, and evolution
 
-$$ N^* = \frac{r_i - 1}{\alpha_{ii}} $$
+The growth equation for each species is:
 
-## Population dynamics simulation
-In the metacommunity simulation in the main text, a species resides in a site with an initial population size $N_{i,0} \sim Pois(10)$, a growth rate $r_i$ that depends on the local environmental value $E_k$ and the species trait $x_i$, and a fixed intra-specific competition coefficient of $\alpha_{ii}$ = 0.00125.
-We simulate population growth here:
+$$ N_{i,t+1} =\frac{\hat{W}e^ \frac{-[(\frac{w+(1-h^2)P} {P+w})(E-x_{i,t})]^2}{2(P+w)}N_{i,t}} {1 + \alpha_{ii}N_{i,t} + \alpha_{ij} N_{j,t}} $$
+
+We simulate population growth under particular parameter values, to generate population dynamics where species interactions, environment, and trait evolution all drive population dynamics.
 
 ```{r}
-set.seed(42)
-# Simulate initial species population growth
-N1.0 <- rpois(1,10)
-r1.0 <- 1.67
-alpha.11 <- 0.00125
-# model function
-disc_log <- function(r, N0, alpha) {
-    Nt1 <- (r*N0) / (1+alpha*N0)
-    return(Nt1)
-}
-# Simulation of model for t time steps
-t <- 30
-N <- rep(NA, t)
-N[1] <- N1.0
-for (i in 2:t) {
-  N[i] <- disc_log(r=r1.0, N0=N[i-1], alpha=alpha.11)
-}
-```
-```{r figs1, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs1}Population size \\textit{N} over time \\textit{t} for a discrete-time logistic growth model, with parameters $r_i$ = 1.67, $N_{1,0}$ = 14, and $\\alpha_{11}$ = 0.00125."}
-# Plot simulation: ggplot
-dat <- as.data.frame(N)
-dat$time <- 1:t
-ggplot2::ggplot(dat, aes(time, N)) + geom_point() + geom_hline(yintercept = ((r1.0 - 1)/alpha.11),
-    linetype = "dashed", color = "gray")
-```
-
-## Linear statistical model
-
-We fit the population time series data to a first-order auto-regressive model to predict $N_{t+1}$ as a function of $N_t$, and compare that to a linear regression.
-We use ln-N after @Ives1995, as also discussed in @Certain2018 and @Olivenca2021.
-
-$$ N_{t+1} = \beta_0 + \beta_1 N_t + \epsilon_t $$
-```{r,warning=FALSE}
-# Fit the model
-m.1.ar <- arima(x = log(N), order = c(1,0,0), include.mean = T, method="CSS")
-m.1.lm <- lm(log(dat$N[2:t]) ~ log(dat$N[1:(t-1)]))
-#plotting the series along with the fitted values
-m.1.ar.fit <- log(N) - residuals(m.1.ar)
-m.1.lm.fit <- log(dat$N[2:t]) - m.1.lm$resid
-dat$ar1.fit <- m.1.ar.fit
-dat$lm.fit <- NA
-dat$lm.fit[2:t] <- m.1.lm.fit
-```
-```{r figs2, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs2}Population size over time (black line) with fitted values from a first-order autoregressive model (red dashed line).",warning=FALSE}
-colors <- c("data" = "black", "AR1 fit" = "gray", "lm fit"="orange")
-ggplot(data=dat,aes(x=time)) +
-  geom_point(aes(y=log(N),color="data")) +
-  geom_point(aes(y=ar1.fit, color = "AR1 fit"), alpha=0.5) +
-  geom_point(aes(y=lm.fit, color = "lm fit"), alpha=0.5) +
-  geom_line(aes(y=log(N), color="data"), linewidth=.4) +
-  geom_line(aes(y=ar1.fit, color = "AR1 fit"), alpha=0.5, linewidth=.4) +
-  geom_line(aes(y=lm.fit, color = "lm fit"), alpha=0.5, linewidth=.4) +
-  theme_bw() +
-  xlab("time") +
-  ylab("N") +
-  labs(color="Legend") +
-  scale_color_manual(values = colors)
-```
-
-The linear model is a good fit, and $N_{t+1}$ and $N_t$ are well-represented by a linear function:
-
-```{r figs3, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs3}Population size (logarithm) at one time step $N_{t+1}$ as a function of log-population size in the previous time step $N_t$.
-"}
-dN <- rep(NA, (t-1))
-for (i in 1:(t-1)) {
-  dN[i] <- log(N[i+1])-log(N[i])
-}
-d.dN <- as.data.frame(dN)
-d.dN$Nt <- log(dat$N[1:(t-1)])
-d.dN$Nt_plus_1 <- log(dat$N[2:t])
-ggplot2::ggplot(d.dN, aes(x=Nt,y=Nt_plus_1)) + geom_point()
-```
-
-We also examine density dependence by plotting $\Delta N = N_{t+1} - N_t$ vs. $N_t$:
-
-```{r figs4, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs4} Change in population size from one time step to the next $N_{t+1}$ as a function of $N_{t+1}$"}
-d.dN <- as.data.frame(dN)
-d.dN$Nt <- log(dat$N[1:(t-1)])
-d.dN$Nt_plus_1 <- log(dat$N[2:t])
-ggplot2::ggplot(d.dN, aes(Nt, dN)) + geom_point()
-```
-## Bayesian linear statistical model: HMSC
-We can estimate the same model parameters using HMSC:
-
-```{r}
-# prepare data in HMSC format
-Y <- as.matrix(log(dat$N[2:t]))
-XData <- data.frame(x=log(dat$N[1:(t-1)]))
-m.1.hmsc <- Hmsc(Y=Y,XData=XData,XFormula=~x)
-# Bayesian model parameters
-nChains <- 2
-thin <- 5
-samples <- 1000
-transient <- 500*thin
-verbose <- 500*thin
-# sample MCMC
-m.1.sample <- sampleMcmc(m.1.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
-```
-
-```{r}
-m.post.hmsc <- convertToCodaObject(m.1.sample)
-summary(m.post.hmsc$Beta)
-plot(m.post.hmsc$Beta)
-```
-
-These estimates match well with those from the AR1 and linear model:
-
-```{r}
-# AR1 coefficients (recall that the intercept is the term below multipled by 1 - phi1)
-m.1.ar$coef
-m.1.ar$coef[2]*(1-m.1.ar$coef[1])
-# linear model
-summary(m.1.lm)$coefficients[1:2,1:2]
-# Bayesian estimates
-summary(m.post.hmsc$Beta)$statistics[1:2,1:2]
-```
-
-```{r figs5, fig.pos = "H", fig.show="hold", fig.cap = "\\label{fig:figs5} Observed (grey) and model-fit (blue) values for population size at time t (x-axis) and t+1 (y-axis).",results='hide',fig.keep='all'}
-Gradient <- constructGradient(m.1.sample,focalVariable="x",ngrid=29)
-predY <- predict(m.1.sample,Gradient=Gradient,expected=TRUE)
-#preds <-computePredictedValues(m.1.sample)
-plotGradient(m.1.sample,Gradient,pred=predY,showData=T,measure="Y",main="",xlab="N_t",ylab="predicted N_t+1")
-```
-
-```{r}
-lm_dat <- data.frame(cbind(log(dat$N[2:t]),log(dat$N[1:(t-1)])))
-colnames(lm_dat) <- c("Nt1","Nt")
-ggplot(lm_dat,aes(Nt,Nt1)) +
-  stat_summary(fun.data= mean_cl_normal) + 
-  geom_smooth(method='lm')
-```
-
-## Conclusions
-In this example, a first-order auto-regressive model works well, bypassing the need to estimate logistic growth parameters $r_i$ and $\alpha_{ii}$.
-The density-dependence dynamics ($\Delta N \sim f(N_t)$) show an overall declining trend over time.
-The Bayesian estimation implemented in HMSC gives good parameter estimates.
-
-# One species, logistic growth, environmental covariate
-We now consider using a linear model to analyze population growth when the species growth rate is impacted by a single environmental covariate.
-
-## Growth depends on environment
-
-First we add environment-dependent growth rate. The growth rate $r_i$ becomes:
-
-$$ r_i = \hat{W}e^{-(E-x_{i,t})^2}$$
-
-Here, $\hat{W}$ is the maximal population growth rate (set to `r r1.0` as above), $E$ is the local environmental trait optimum value, and $x_{i,t}$ is species *i* trait value at time *t*.
-We see that if $E = x_{i,t}$ then the growth rate is at the value $r$ = `r r1.0`.
-Here, we begin with $E = x_{i,t} = 0.8$, then simulate the environment $E$ value fluctuating randomly over time, and finally use a linear model to fit $E$ as a covariate.
-
-```{r}
-# Simulate initial species population growth with environment fluctuations
-N1.0 <- 10
-r1.0 <- 1.67
-alpha.11 <- 0.00125
-E.0 <- 0.8
-x1.0 <- 0.8
-# model function
-disc_log_E <- function(r, N0, alpha, E, x) {
-    Nt1 <- ((r*exp(-(E-x)^2))*N0) / (1+alpha*N0)
-    return(Nt1)
-}
-# Simulation of model for t time steps
-t <- 40
-N <- rep(NA, t)
-N[1] <- N1.0
-E <- rep(NA, t)
-E[1] <- E.0
-for (i in 2:t) {
-  N[i] <- disc_log_E(r=r1.0, N0=N[i-1], alpha=alpha.11, E=E[i-1], x=x1.0)
-  E[i] <- E[i-1] + rnorm(1,0,.1)
-}
-```
-```{r figs6, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs1}Population size \\textit{N} over time \\textit{t} for a discrete-time logistic growth model, with parameters $r_i$ = 1.67, $N_{1,0}$ = 14, and $\\alpha_{11}$ = 0.00125.
-Relationship between E and Nt is also shown."}
-# Plot simulation: ggplot
-dat <- as.data.frame(cbind(N,E))
-dat$time <- 1:t
-p1 <- ggplot2::ggplot(dat, aes(time, N)) + geom_point() + geom_hline(yintercept = ((r1.0 - 1)/alpha.11),linetype = "dashed", color = "gray")
-p2 <- ggplot2::ggplot(dat, aes(time, E)) + geom_point()
-p1 + p2
-```
-
-## Linear statistical model with environmental covariate
-
-We now include environment $E$ as a covariate in the linear model:
-
-$$ N_{t} = \beta_0 + \beta_1 N_{t-1} + \beta_2 E_{t-1}+ \epsilon_t $$
-
-```{r,warning=FALSE}
-# Fit the model
-m.2.ar <- arima(x = log(N), order = c(1,0,0), include.mean = T, method="CSS", xreg=E)
-m.2.lm <- lm(log(dat$N[2:t]) ~ log(dat$N[1:(t-1)]) + log(E[1:(t-1)]))
-#plotting the series along with the fitted values
-m.2.ar.fit <- log(N) - residuals(m.2.ar)
-m.2.lm.fit <- log(dat$N[2:t]) - m.2.lm$resid
-dat$ar2.fit <- m.2.ar.fit
-dat$lm2.fit <- NA
-dat$lm2.fit[2:t] <- m.2.lm.fit
-```
-```{r figs7, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs2}Population size over time (black line) with fitted values from a first-order autoregressive model (red dashed line).",warning=FALSE}
-colors <- c("data" = "black", "AR1 fit" = "gray", "lm fit"="orange")
-ggplot(data=dat,aes(x=time)) +
-  geom_point(aes(y=log(N),color="data")) +
-  geom_point(aes(y=ar2.fit, color = "AR1 fit"), alpha=0.5) +
-  geom_point(aes(y=lm2.fit, color = "lm fit"), alpha=0.5) +
-  geom_line(aes(y=log(N), color="data"), linewidth=.4) +
-  geom_line(aes(y=ar2.fit, color = "AR1 fit"), alpha=0.5, linewidth=.4) +
-  geom_line(aes(y=lm2.fit, color = "lm fit"), alpha=0.5, linewidth=.4) +
-  theme_bw() +
-  xlab("time") +
-  ylab("N") +
-  labs(color="Legend") +
-  scale_color_manual(values = colors)
-```
-
-```{r figs8, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs3}Population size (logarithm) at one time step $N_{t+1}$ as a function of log-population size in the previous time step $N_t$.
-"}
-dN <- rep(NA, (t-1))
-for (i in 1:(t-1)) {
-  dN[i] <- log(N[i+1])-log(N[i])
-}
-d.dN <- as.data.frame(dN)
-d.dN$Nt <- log(dat$N[1:(t-1)])
-d.dN$Nt_plus_1 <- log(dat$N[2:t])
-p1 <- ggplot2::ggplot(d.dN, aes(x=Nt,y=Nt_plus_1)) + geom_point()
-
-d.dN <- as.data.frame(dN)
-d.dN$Nt <- log(dat$N[1:(t-1)])
-d.dN$Nt_plus_1 <- log(dat$N[2:t])
-d.dN$E <- log(dat$E[2:t])
-p2 <- ggplot2::ggplot(d.dN, aes(x=E,y=Nt_plus_1)) + geom_point()
-p3 <- ggplot2::ggplot(d.dN, aes(Nt, dN)) + geom_point()
-p4 <- ggplot2::ggplot(d.dN, aes(E, dN)) + geom_point()
-
-p1+p2+p3+p4
-```
-
-The linear model is a good fit when including the environmental covariate.
-$N_{t+1}$ and $N_t$ can still be captured by a linear relationship.
-However we see that the relationship between $N_{t+1}$ and $E_t$ is non-linear.
-This tells us that the lm is good for predictions, but not for inference (for capturing well the relationsip between the predictor and response variable).
-The use of linear relationships in JSDMs is discussed in [@Ingram2020], and in many applications (e.g.
-[@Erickson2023]) quadratic terms are used, which create bell-shaped response curves that may better match species with optimal niches (as opposed to linear, monotonically increasing relationships between population size and environmental predictors).
-We thus include a quadratic term for $E_t$ to provide a better fit to the data.
-
-```{r}
-df <- data.frame(cbind(log(dat$N[2:t]),log(dat$N[1:(t-1)]),E[1:(t-1)],E[1:(t-1)]^2))
-colnames(df) <- c("Nt1","Nt","E","Esq")
-m.2.lm <- lm(Nt1 ~ Nt + E + Esq,data=df)
-```
-
-```{r figs9, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs3}Population size (logarithm) at one time step $N_{t+1}$ as a function of log-population size in the previous time step $N_t$.
-"}
-jtools::effect_plot(m.2.lm,pred=E,plot.points = TRUE)
-ggplot(data = df, aes(x = E, y = Nt1)) +
-  geom_point() +
-  geom_smooth(method = "lm", formula = y ~ x + I(x^2))
-```
-
-## Bayesian linear statistical model: HMSC
-We can estimate the same model parameters using HMSC:
-
-```{r}
-# prepare data in HMSC format
-Y <- as.matrix(log(dat$N[2:t]))
-XData <- df
-m.2.hmsc <- Hmsc(Y=Y,XData=XData,XFormula=~Nt+E+Esq)
-# Bayesian model parameters
-nChains <- 2
-thin <- 5
-samples <- 1000
-transient <- 500*thin
-verbose <- 500*thin
-# sample MCMC
-m.2.sample <- sampleMcmc(m.2.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
-```
-
-```{r}
-m2.post.hmsc <- convertToCodaObject(m.2.sample)
-summary(m2.post.hmsc$Beta)
-bayesplot::mcmc_trace(m2.post.hmsc$Beta)
-bayesplot::mcmc_areas(m2.post.hmsc$Beta,area_method = c("equal height"))
-```
-
-These estimates match well with those from the AR1 and linear model:
-
-```{r}
-# AR1 coefficients (recall that the intercept is the term below multipled by 1 - phi1)
-m.2.ar$coef
-m.2.ar$coef[2]*(1-m.2.ar$coef[1])
-# linear model
-summary(m.2.lm)$coefficients[1:4,1:2]
-# Bayesian estimates
-summary(m2.post.hmsc$Beta)$statistics[1:4,1:2]
-```
-
-We recall that the interpretation of the coefficients in an arimaX (arima with covariates) model is difficult.
-They do not give the impact on $N_t$ per unit increase in X as in a regression.
-So we do not interpret the causation implied by the coefficient in the arimaX model.
-In the regression model, we can see that $E$ has a positive impact on $N_t$.
-
-```{r figs10, fig.pos = "H", fig.show="hold", fig.cap = "\\label{fig:figs5} Observed (grey) and model-fit (blue) values for population size at time t (x-axis) and t+1 (y-axis).",results='hide',fig.keep='all'}
-Gradient <- constructGradient(m.2.sample,focalVariable="E",non.focalVariables=list(Nt=list(2),Esq=list(2)),ngrid=39)
-# Esq is manually constructed as gradient-produced E^2
-Gradient$XDataNew$Esq <- Gradient$XDataNew$E^2
-predY <- predict(m.2.sample,XData=Gradient$XDataNew,expected=TRUE)
-plotGradient(m.2.sample,Gradient,pred=predY,showData=T,measure="Y",main="",xlab="E_t",ylab="predicted N_t+1")
-```
-
-## Conclusions
-In this example, the linear regression again works well to describe the impact of $E_t$ for $N_t$ when using the quadratic formulation.
-The arimaX model works well for fitting and subsequent prediction, but less well for inference about the impacts of $E$.
-From the quadratic regression terms for $E$, we correctly see that the population size is maximal at the species trait value and decreases away from that value.
-We will continue to use log-transformed abundance and now introduce quadratic terms for the environmental parameter.
-
-# Two species, logistic growth, competition
-XXXXX
-## Interspecific competition
-The growth equation for each species now becomes:
-
-$$ N_{i,t+1} = \frac{r_i N_{i,t}}{1 + \alpha_{ii} N_{i,t} + \alpha_{ij} N_{j,t}}$$
-We use a distinct growth rate for the 2nd species and introduce the interspecific interaction coefficient $\alpha_{ij}$.
-
-```{r}
+# Case 1: Weaker interactions, stronger environment
 # Simulate initial species population growth with environment fluctuations
@@
 N2.0 <- 10
-r1.0 <- 1.67
-r2.0 <- 1.7
-alpha.11 <- 0.00125
-alpha.22 <- 0.00125
-alpha.12 <- 0.008
-alpha.21 <- 0.008725
-# model function
-disc_LV_comp <- function(r1,r2,N1.0,N2.0,alpha.11,alpha.22,alpha.12,alpha.21) {
-    Nt1 <- (r1*N1.0) / (1+alpha.11*N1.0+alpha.12*N2.0)
-    Nt2 <- (r2*N2.0) / (1+alpha.22*N2.0+alpha.21*N1.0)
-    return(c(Nt1,Nt2))
-}
-# Simulation of model for t time steps
-t <- 40
-N <- array(NA,dim=c(t,2))
-N <- as.data.frame(N)
-colnames(N) <- c("N1","N2")
-N$N1[1] <- N1.0
-N$N2[1] <- N2.0
-for (i in 2:t) {
-  res <- disc_LV_comp(r1=r1.0,r2=r2.0,N1.0=N[i-1,1],N2.0=N[i-1,2],alpha.11=alpha.11,alpha.22=alpha.22,alpha.12=alpha.12,alpha.21=alpha.21)
-  N$N1[i] <- res[1]
-  N$N2[i] <- res[2]
-}
-```
-```{r figs11, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs1}Population size \\textit{N} over time \\textit{t} for a discrete-time logistic growth model, with parameters $r_i$ = 1.67, $N_{1,0}$ = 14, and $\\alpha_{11}$ = 0.00125.
-Relationship between E and Nt is also shown."}
-# Plot simulation: ggplot
-N$time <- 1:t
-dat <- melt(N, id.vars="time")
-ggplot2::ggplot(dat, aes(time, value, col=variable)) + geom_point() + geom_hline(yintercept = ((r1.0 - 1)/alpha.11),linetype = "dashed", color = "gray")
-```
-## Linear statistical model with environmental covariate
-
-We fit each species' population time series to an arimaX and lm with population size of the others species as a covariate.
-
-$$ N_{1,t} = \beta_0 + \beta_1 N_{1,t-1} + \beta_2 N_{2,t-1}+ \epsilon_t $$
-
-```{r figs12, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs1}Population size \\textit{N} over time \\textit{t} for a discrete-time logistic growth model, with parameters $r_i$ = 1.67, $N_{1,0}$ = 14, and $\\alpha_{11}$ = 0.00125.
-Relationship between E and Nt is also shown."}
-# Plot simulation: ggplot
-dat <- as.data.frame(cbind(N$N1,N$N2))
-colnames(dat) <- c("N1","N2")
-dat$time <- 1:t
-p1 <- ggplot2::ggplot(dat, aes(time, N1)) + geom_point() + geom_hline(yintercept = ((r1.0 - 1)/alpha.11),linetype = "dashed", color = "gray")
-p2 <- ggplot2::ggplot(dat, aes(time, N2)) + geom_point()
-p1 + p2
-```
-
-```{r,warning=FALSE}
-## Fit the model for Species 1
-m.3.ar.n1 <- arima(x = log(N$N1), order = c(1,0,0), include.mean = T, method="CSS", xreg=N$N2)
-m.3.lm.n1 <- lm(log(dat$N1[2:t]) ~ log(dat$N1[1:(t-1)]) + log(dat$N2[1:(t-1)]))
-#plotting the series along with the fitted values
-m.3.ar.fit.n1 <- log(N$N1) - residuals(m.3.ar.n1)
-m.3.lm.fit.n1 <- log(dat$N1[2:t]) - m.3.lm.n1$resid
-dat$ar3.fit.n1 <- m.3.ar.fit.n1
-dat$lm3.fit.n1 <- NA
-dat$lm3.fit.n1[2:t] <- m.3.lm.fit.n1
-## Species 2
-m.3.ar.n2 <- arima(x = log(N$N2), order = c(1,0,0), include.mean = T, method="CSS", xreg=N$N1)
-m.3.lm.n2 <- lm(log(dat$N2[2:t]) ~ log(dat$N2[1:(t-1)]) + log(dat$N1[1:(t-1)]))
-#plotting the series along with the fitted values
-m.3.ar.fit.n2 <- log(N$N2) - residuals(m.3.ar.n2)
-m.3.lm.fit.n2 <- log(dat$N2[2:t]) - m.3.lm.n2$resid
-dat$ar3.fit.n2 <- m.3.ar.fit.n2
-dat$lm3.fit.n2 <- NA
-dat$lm3.fit.n2[2:t] <- m.3.lm.fit.n2
-```
-
-```{r figs13, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs2}Population size over time (black line) with fitted values from a first-order autoregressive model (red dashed line).",warning=FALSE}
-colors <- c("data" = "black", "AR1 fit" = "gray", "lm fit"="orange")
-p1 <- ggplot(data=dat,aes(x=time)) +
-  geom_point(aes(y=log(N1),color="data")) +
-  geom_point(aes(y=ar3.fit.n1, color = "AR1 fit"), alpha=0.5) +
-  geom_point(aes(y=lm3.fit.n1, color = "lm fit"), alpha=0.5) +
-  geom_line(aes(y=log(N1), color="data"), linewidth=.4) +
-  geom_line(aes(y=ar3.fit.n1, color = "AR1 fit"), alpha=0.5, linewidth=.4) +
-  geom_line(aes(y=lm3.fit.n1, color = "lm fit"), alpha=0.5, linewidth=.4) +
-  theme_bw() +
-  xlab("time") +
-  ylab("N1") +
-  labs(color="Legend") +
-  scale_color_manual(values = colors)
-p2 <- ggplot(data=dat,aes(x=time)) +
-  geom_point(aes(y=log(N2),color="data")) +
-  geom_point(aes(y=ar3.fit.n2, color = "AR1 fit"), alpha=0.5) +
-  geom_point(aes(y=lm3.fit.n2, color = "lm fit"), alpha=0.5) +
-  geom_line(aes(y=log(N2), color="data"), linewidth=.4) +
-  geom_line(aes(y=ar3.fit.n2, color = "AR1 fit"), alpha=0.5, linewidth=.4) +
-  geom_line(aes(y=lm3.fit.n2, color = "lm fit"), alpha=0.5, linewidth=.4) +
-  theme_bw() +
-  xlab("time") +
-  ylab("N2") +
-  labs(color="Legend") +
-  scale_color_manual(values = colors)
-p1+p2
-```
-
-```{r figs14, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs3}Population size (logarithm) at one time step $N_{t+1}$ as a function of log-population size in the previous time step $N_t$.
-"}
-dN1 <- rep(NA, (t-1))
-for (i in 1:(t-1)) {
-  dN1[i] <- log(N$N1[i+1])-log(N$N1[i])
-}
-d.dN1 <- as.data.frame(dN1)
-d.dN1$Nt <- log(dat$N1[1:(t-1)])
-d.dN1$Nt_plus_1 <- log(dat$N1[2:t])
-p1 <- ggplot2::ggplot(d.dN1, aes(x=Nt,y=Nt_plus_1)) + geom_point()
-
-d.dN1 <- as.data.frame(dN1)
-d.dN1$Nt <- log(dat$N1[1:(t-1)])
-d.dN1$Nt_plus_1 <- log(dat$N1[2:t])
-d.dN1$N2 <- log(dat$N2[2:t])
-p2 <- ggplot2::ggplot(d.dN1, aes(x=N2,y=Nt_plus_1)) + geom_point()
-p3 <- ggplot2::ggplot(d.dN1, aes(Nt, dN1)) + geom_point()
-p4 <- ggplot2::ggplot(d.dN1, aes(N2, dN1)) + geom_point()
-
-p1+p2+p3+p4
-```
-
-XXXXX
-
-```{r}
-df <- data.frame(cbind(log(dat$N1[2:t]),log(dat$N1[1:(t-1)]),log(dat$N2[1:(t-1)])))
-colnames(df) <- c("Nt1","N1","N2")
-m.3.lm <- lm(Nt1 ~ I(1/N1) + I(1/N2),data=df)
-
-cor(df$Nt1, predict(m.3.lm))
-summary(m.3.lm)
-xx <- seq(0,100, length=1000)
-prediction<-data.frame(N2=xx,N1=100)
-plot(df$N2, df$Nt1, xlab = "N2", ylab="N.t+1" , pch=16)
-lines(prediction$N2, predict(m.3.lm, prediction), lty=2,col="red",lwd=3)
-
-```
-
-```{r figs15, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs3}Population size (logarithm) at one time step $N_{t+1}$ as a function of log-population size in the previous time step $N_t$.
-"}
-jtools::effect_plot(m.3.lm,pred=N2,plot.points = TRUE)
-ggplot(data = df, aes(x = N2, y = Nt1)) +
-  geom_point() +
-  geom_smooth(method = "lm", formula = y ~ x)
-```
-We can already see that the linear model does not provide a useful fit to the data points (neither does taking the inverse function help).
-The regression parameters can predict well the population size of the other species, but they are not informative for inference.
-This has been observed previously (@Certain2018), and non-linear least squares models (among others) are more often used to estimate parameters for dynamics resulting from these kinds of species interaction models (@Kloppers2013; @Muhlbauer2020; @Olivenca2021).
-The inability of the linear models to reproduce the interaction coefficients is more clearly demonstrated in @Certain2018.
-However, they do indicate that the slope of the linear model can reflect the direction of effect for the interacting species.
-
-Instead of further exploring techniques to fit to time-series derived from non-linear competition processes, we look more closely at how HMSC manages to fit data emerging from competition dynamics and make inference about species interactions.
-Instead of estimating full interaction coefficients, or assuming sparse interactions, they instead assume that a reduced number of linear combinations of species abundances that are most relevant to determining future growth rates for species in the community can bypass the 'curse of dimensionality' problem.
-This is well presented in @Ovaskainen2017.
-We apply that approach here:
-
-## Bayesian linear statistical model: HMSC
-We can fit the data to a linear model using HMSC, with latent variables to capture the species associations. These are implemented by including a random effect at the time level.
-
-```{r}
-# prepare data in HMSC format
-Y <- as.matrix(cbind(log(dat$N1[2:t]),log(dat$N2[2:t])))
-XData <- df[,2:3]
-studyDesign = data.frame(sample = as.factor(1:(t-1)))
-rL = HmscRandomLevel(units = studyDesign$sample)
-m.3.hmsc = Hmsc(Y = Y, XData = XData, studyDesign = studyDesign, ranLevels = list(sample = rL))
-# Bayesian model parameters
-nChains <- 2
-thin <- 5
-samples <- 1000
-transient <- 500*thin
-verbose <- 500*thin
-# sample MCMC
-m.3.sample <- sampleMcmc(m.3.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
-```
-
-```{r}
-m3.post.hmsc <- convertToCodaObject(m.3.sample)
-summary(m3.post.hmsc$Beta)
-bayesplot::mcmc_trace(m3.post.hmsc$Beta)
-bayesplot::mcmc_areas(m3.post.hmsc$Beta,area_method = c("equal height"))
-
-OmegaCor = computeAssociations(m.3.sample)
-supportLevel = 0.95
-toPlot = ((OmegaCor[[1]]$support>supportLevel)
-+ (OmegaCor[[1]]$support<(1-supportLevel))>0)*OmegaCor[[1]]$mean
-corrplot::corrplot(toPlot, method = "color",col = colorRampPalette(c("blue","white","red"))(200),title = paste("random effect level:", m.3.sample$rLNames[1]), mar=c(0,0,1,0))
-```
-
-These estimates match well with those from the AR1 and linear model:
-
-```{r}
-# AR1 coefficients (recall that the intercept is the term below multipled by 1 - phi1)
-m.3.ar.n1$coef
-m.3.ar.n1$coef[2]*(1-m.3.ar.n1$coef[1])
-# linear model
-summary(m.3.lm.n1)$coefficients[1:3,1:2]
-# Bayesian estimates
-summary(m3.post.hmsc$Beta)$statistics[1:3,1:2]
-```
-
-```{r}
-# Simulate initial species population growth with environment fluctuations
-N1.0 <- 10
-N2.0 <- 10
-r1.0 <- 1.7
-r2.0 <- 1.7
-alpha.11 <- 0.01
-alpha.22 <- 0.01
-alpha.12 <- 0.005
-alpha.21 <- 0.01
-# model function
-disc_LV_comp <- function(r1,r2,N1.0,N2.0,alpha.11,alpha.22,alpha.12,alpha.21) {
-    Nt1 <- (r1*N1.0) / (1+alpha.11*N1.0+alpha.12*N2.0)
-    Nt2 <- (r2*N2.0) / (1+alpha.22*N2.0+alpha.21*N1.0)
-    return(c(Nt1,Nt2))
-}
-# Simulation of model for t time steps
-t <- 40
-N <- array(NA,dim=c(t,2))
-N <- as.data.frame(N)
-colnames(N) <- c("N1","N2")
-N$N1[1] <- N1.0
-N$N2[1] <- N2.0
-for (i in 2:t) {
-  res <- disc_LV_comp(r1=r1.0,r2=r2.0,N1.0=N[i-1,1],N2.0=N[i-1,2],alpha.11=alpha.11,alpha.22=alpha.22,alpha.12=alpha.12,alpha.21=alpha.21)
-  N$N1[i] <- res[1]
-  N$N2[i] <- res[2]
-}
-```
-```{r}
-# Plot simulation: ggplot
-N$time <- 1:t
-dat <- melt(N, id.vars="time")
-ggplot2::ggplot(dat, aes(time, value, col=variable)) + geom_point() + geom_hline(yintercept = ((r1.0 - 1)/alpha.11),linetype = "dashed", color = "gray")
-```
-
-```{r}
-# Plot simulation: ggplot
-dat <- as.data.frame(cbind(N$N1,N$N2))
-colnames(dat) <- c("N1","N2")
-dat$time <- 1:t
-df <- data.frame(cbind(log(dat$N1[2:t]),log(dat$N2[2:t])))
-colnames(df) <- c("Nt1","Nt2")
-# prepare data in HMSC format
-Y <- as.matrix(cbind(df$Nt1,df$Nt2))
-XData <- data.frame(cbind(log(dat$N1[1:(t-1)]),log(dat$N2[1:(t-1)])))
-studyDesign = data.frame(sample = as.factor(1:(t-1)))
-rL = HmscRandomLevel(units = studyDesign$sample)
-m.4.hmsc = Hmsc(Y = Y, XData = XData, XFormula = ~1, studyDesign = studyDesign, ranLevels = list(sample = rL))
-# Bayesian model parameters
-nChains <- 2
-thin <- 5
-samples <- 1000
-transient <- 500*thin
-verbose <- 500*thin
-# sample MCMC
-m.4.sample <- sampleMcmc(m.4.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
-```
-
-```{r}
-m4.post.hmsc <- convertToCodaObject(m.4.sample)
-summary(m4.post.hmsc$Beta)
-bayesplot::mcmc_trace(m4.post.hmsc$Beta)
-bayesplot::mcmc_areas(m4.post.hmsc$Beta,area_method = c("equal height"))
-
-OmegaCor = computeAssociations(m.4.sample)
-supportLevel = 0.95
-toPlot = ((OmegaCor[[1]]$support>supportLevel)
-+ (OmegaCor[[1]]$support<(1-supportLevel))>0)*OmegaCor[[1]]$mean
-corrplot::corrplot(toPlot, method = "color",col = colorRampPalette(c("blue","white","red"))(200),title = paste("random effect level:", m.4.sample$rLNames[1]), mar=c(0,0,1,0))
-```
-
-The HMSC model can estimate the species associations well.
-
-```{r}
-# explanatory power
-preds = computePredictedValues(m.4.sample)
-#evaluateModelFit(hM = m.4.sample, predY = preds)
-# predictive power / cross-validation
-#partition = createPartition(m.4.sample, nfolds = 2)
-#preds = computePredictedValues(m.4.sample, partition = partition, nParallel = nChains)
-#evaluateModelFit(hM = m.4.sample, predY = preds)
-## xx
-#preds = computePredictedValues(m.4.sample, partition=partition, partition.sp=c(1,2), mcmcStep=10, nParallel = nChains)
-#evaluateModelFit(hM=m.4.sample, predY=preds)
-# xx
-#etaPost=getPostEstimate(m.4.sample, "Eta")
-#lambdaPost=getPostEstimate(m.4.sample, "Lambda")
-#biPlot(m.4.sample, etaPost = etaPost, lambdaPost = lambdaPost, factors = c(1,2),"X1")
-```
-```{r}
-df$n1_25 <- NA
-df$n1_975 <- NA
-df$n2_25 <- NA
-df$n2_975 <- NA
-for(i in 1:39){
-  red <- preds[i,,]
-  a <- apply(red,1,quantile,probs=c(.025,.975))
-  df$n1_25[i] <- a[1,1]
-  df$n1_975[i] <- a[2,1]
-  df$n2_25[i] <- a[1,2]
-  df$n2_975[i] <- a[2,2]
-}
-plot(df$Nt1,pch=19,col="black",ylim=c(0,5))
-points(df$Nt2,pch=19,col="orange")
-lines(df$n1_25,col="blue")
-lines(df$n1_975,col="blue")
-lines(df$n2_25,col="blue")
-lines(df$n2_975,col="blue")
-```
-
-# 4 species growth, competition, with environmental change
-
-The goal of the process was to use HMSC's latent variable approach to study the results of non-linear processes in ecology using a linear model.
-We can infer the direction and magnitude of the species interaction.
-We now consider an environmental covariate that changes over time and impacts each species growth.
-We will use HMSC from this point forward.
-
-```{r}
-# Simulate initial species population growth with environment fluctuations
-N1.0 <- 10
-N2.0 <- 10
-r1.0 <- 1.7
-r2.0 <- 1.7
 alpha.11 <- 0.01
@@
 x2.0 <- 0.8
+P <- 1
+w <- 10
+Wmax <- 2
+h2 <- 0.1
+k <- (w+(1-h2)*P)/(P+w)
 
 # model function
-disc_LV_E <- function(r1,r2,N1.0,N2.0,alpha.11,alpha.22,alpha.12,alpha.21,E,x1,x2) {
-    Nt1 <- ((r1*exp(-(E-x1)^2))*N1.0) / (1+alpha.11*N1.0+alpha.12*N2.0)
-    Nt2 <- ((r2*exp(-(E-x2)^2))*N2.0) / (1+alpha.22*N2.0+alpha.21*N1.0)
-    return(c(Nt1,Nt2))
+disc_LV_evol <- function(N1.0,N2.0,alpha.11,alpha.22,alpha.12,alpha.21,E,x1.0,x2.0,P,w,Wmax,h2) {
+  What <- Wmax*sqrt(w/(P+w))
+  r1 <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E-x1.0))^2)/(2*(P+w)))
+  Nt1 <- (r1*N1.0) / (1+alpha.11*N1.0+alpha.12*N2.0)
+  r2 <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E-x2.0))^2)/(2*(P+w)))
+  Nt2 <- (r2*N2.0) / (1+alpha.22*N2.0+alpha.21*N1.0)
+  return(c(Nt1,Nt2,r1,r2))
 }
@@
 colnames(N) <- c("N1","N2")
+x <- array(NA,dim=c(t,2))
+x <- as.data.frame(x)
+colnames(x) <- c("x1","x2")
+r <- array(NA,dim=c(t,2))
+r <- as.data.frame(r)
+colnames(r) <- c("r1","r2")
 N$N1[1] <- N1.0
 N$N2[1] <- N2.0
+x$x1[1] <- x1.0
+x$x2[1] <- x2.0
+r$r1[1] <- exp((-(((w+(1-h2)*P)/(P+w))*(E.0-x1.0))^2)/(2*(P+w)))
+r$r2[1] <- exp((-(((w+(1-h2)*P)/(P+w))*(E.0-x2.0))^2)/(2*(P+w)))
 E <- rep(NA, t)
@@
 for (i in 2:t) {
-  res <- disc_LV_E(r1=r1.0,r2=r2.0,N1.0=N[i-1,1],N2.0=N[i-1,2],alpha.11=alpha.11,alpha.22=alpha.22,alpha.12=alpha.12,alpha.21=alpha.21,E=E[i-1],x1=x1.0,x2=x2.0)
+  res <- disc_LV_evol(N1.0=N[i-1,1],N2.0=N[i-1,2],alpha.11=alpha.11,alpha.22=alpha.22,alpha.12=alpha.12,alpha.21=alpha.21,E=E[i-1],x1=x[i-1,1],x2=x[i-1,2],P=P,w=w,Wmax=Wmax,h2=h2)
   N$N1[i] <- res[1]
   N$N2[i] <- res[2]
-  E[i] <- E[i-1] + rnorm(1,0,.1)
+  r$r1[i] <- res[3]
+  r$r2[i] <- res[4]
+  # trait change
+  d1 <- E[i-1] - x[i-1,1]
+  d2 <- E[i-1] - x[i-1,2]
+  d1.1 <- k*d1
+  d2.1 <- k*d2
+  x$x1[i] <- E[i-1] - d1.1
+  x$x2[i] <- E[i-1] - d2.1
+  # environmental change
+  E[i] <- E[i-1] + rnorm(1,0,.01)
 }
@@
 dat <- melt(N, id.vars="time")
-ggplot2::ggplot(dat, aes(time, value, col=variable)) + geom_point() + geom_hline(yintercept = ((r1.0 - 1)/alpha.11),linetype = "dashed", color = "gray")
+ggplot2::ggplot(dat, aes(time, value, col=variable)) + geom_point() + geom_hline(yintercept = (1/alpha.11),linetype = "dashed", color = "gray")
 ```
+
+# Hypotheses for evolution as a driver of species abundances
+
+The growth equation is not linear, yet we fit abundance data to a heirarchical model with linear predictors because in most natural systems, a single theoretical growth model can't adequately capture the complexity in a given system.
+Linear models don't capture mechanistic relationships, but instead correlative.
+They are useful for inferring direction and magnitude of effect sizes.
+In Appendix S1, we explore how non-linear dynamics are captured when fit to a linear model.
+Here, we explore different forms to encode evolution in the trait $x$ that determines the organism's fitness in this system.
+
+## H1. Raw trait value drives species abundances
+
 ```{r}
 # Plot simulation: ggplot
-dat <- as.data.frame(cbind(log(N$N1),log(N$N2)))
-colnames(dat) <- c("N1","N2")
+dat <- as.data.frame(cbind(log(N$N1),log(N$N2),x$x1,x$x2))
+colnames(dat) <- c("N1","N2","x1","x2")
 dat$time <- 1:t
-df <- data.frame(cbind(dat$N1[2:t],dat$N2[2:t]))
-colnames(df) <- c("Nt1","Nt2")
-# prepare data in HMSC format
-Y <- as.matrix(cbind(df$Nt1,df$Nt2))
-XData <- data.frame(cbind(dat$N1[1:(t-1)],dat$N2[1:(t-1)]),E[1:(t-1)],E[1:(t-1)]^2)
-colnames(XData) <- c("n1","n2","E","Esq")
-studyDesign = data.frame(sample = as.factor(1:(t-1)))
-rL = HmscRandomLevel(units = studyDesign$sample)
-m.5.hmsc = Hmsc(Y = Y, XData = XData, XFormula = ~E+Esq, studyDesign = studyDesign, ranLevels = list(sample = rL))
-# Bayesian model parameters
-nChains <- 2
-thin <- 5
-samples <- 1000
-transient <- 500*thin
-verbose <- 500*thin
-# sample MCMC
-m.5.sample <- sampleMcmc(m.5.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
+df <- data.frame(cbind(dat$N1[2:t],dat$N2[2:t],dat$x1[2:t],dat$x2[2:t]))
+colnames(df) <- c("Nt1","Nt2","xt1","xt2")
+# Plot
+p1 <- ggplot2::ggplot(df, aes(xt1, Nt1)) + geom_point()
+p2 <- ggplot2::ggplot(df, aes(xt2, Nt1)) + geom_point()
+p3 <- ggplot2::ggplot(df, aes(xt1, Nt2)) + geom_point()
+p4 <- ggplot2::ggplot(df, aes(xt2, Nt2)) + geom_point()
+p1 + p2 + p3 + p4
+```
+
+We do not hypothesize that the raw trait value is the best predictor of species abundances, because the impacts of this evolved trait depend on the context of the environment.
+
+## H2. Absolute value of trait distance from optimum drives species abundances
+
+```{r}
+# Plot simulation: ggplot
+df$E <- E[2:t]
+df$abs_Ex1 <- abs(df$E - df$xt1)
+df$abs_Ex2 <- abs(df$E - df$xt2)
+# Plot
+p1 <- ggplot2::ggplot(df, aes(abs_Ex1, Nt1)) + geom_point()
+p2 <- ggplot2::ggplot(df, aes(abs_Ex2, Nt1)) + geom_point()
+p3 <- ggplot2::ggplot(df, aes(abs_Ex1, Nt2)) + geom_point()
+p4 <- ggplot2::ggplot(df, aes(abs_Ex2, Nt2)) + geom_point()
+p1 + p2 + p3 + p4
+```
+
+We hypothesize that this represents the true impact on species abundances, via the direct impact this measure has on fitness and consequent population growth.
+However, we also acknowledge this is likely difficult to measure in most empirical systems.
+
+## H3. Absolute value of trait change from one time to the next drives species abundances
+
+We now onsider the magnitude of trait change from one time step to the next as a driver for species abundances.
+
+```{r}
+df$dx1 <- abs(dat$x1[2:t] - dat$x1[1:(t-1)])
+df$dx2 <- abs(dat$x2[2:t] - dat$x2[1:(t-1)])
+# Plot
+p1 <- ggplot2::ggplot(df, aes(dx1, Nt1)) + geom_point()
+p2 <- ggplot2::ggplot(df, aes(dx2, Nt1)) + geom_point()
+p3 <- ggplot2::ggplot(df, aes(dx1, Nt2)) + geom_point()
+p4 <- ggplot2::ggplot(df, aes(dx2, Nt2)) + geom_point()
+p1 + p2 + p3 + p4
+```
+
+The measures correlate highly with one another, as the amount of trait change (the amount of evolution) is determined by the distance to the optimum trait value (given that P and w remain constant in the simulation).
+
+```{r}
+cor(df$dx1, df$abs_Ex1)
+cor(df$dx2, df$abs_Ex2)
+# Plot
+p1 <- ggplot2::ggplot(df, aes(abs_Ex1,dx1)) + geom_point()
+p2 <- ggplot2::ggplot(df, aes(abs_Ex2,dx2)) + geom_point()
+p1 + p2
+```
+
+# Statistical model for evolution as a driver of species abundances
+
+```{r}
+knitr::knit_exit()
 ```
@@
 ```{r}
-m5.post.hmsc <- convertToCodaObject(m.5.sample)
-summary(m5.post.hmsc$Beta)
-bayesplot::mcmc_trace(m5.post.hmsc$Beta)
-bayesplot::mcmc_areas(m5.post.hmsc$Beta,area_method = c("equal height"))
+m7.post.hmsc <- convertToCodaObject(m.7.sample)
+summary(m7.post.hmsc$Beta)
+bayesplot::mcmc_trace(m7.post.hmsc$Beta)
+bayesplot::mcmc_areas(m7.post.hmsc$Beta,area_method = c("equal height"))
 ```
@@
 # Bayesian estimates
-summary(m5.post.hmsc$Beta)$statistics[1:4,1:2]
+summary(m7.post.hmsc$Beta)$statistics[1:10,1:2]
 ```
@@
 ```{r}
-OmegaCor = computeAssociations(m.5.sample)
+OmegaCor = computeAssociations(m.7.sample)
 supportLevel = 0.7
@@
 + (OmegaCor[[1]]$support<(1-supportLevel))>0)*OmegaCor[[1]]$mean
-corrplot::corrplot(toPlot, method = "color",col = colorRampPalette(c("blue","white","red"))(200),title = paste("random effect level:", m.5.sample$rLNames[1]), mar=c(0,0,1,0))
+corrplot::corrplot(toPlot, method = "color",col = colorRampPalette(c("blue","white","red"))(200),title = paste("random effect level:", m.7.sample$rLNames[1]), mar=c(0,0,1,0))
+OmegaCor
 ```
@@
 ```{r}
-Gradient <- constructGradient(m.5.sample,focalVariable="E",non.focalVariables=list(Esq=list(2)),ngrid=39)
-predY <- predict(m.5.sample,XData=Gradient$XDataNew,expected=TRUE)
-a <- plotGradient(m.5.sample,Gradient,pred=predY,showData=T,measure="Y",index=1,main="",xlab="E_t",ylab="predicted N1_t+1")
-b <- plotGradient(m.5.sample,Gradient,pred=predY,showData=T,measure="Y",index=2,main="",xlab="E_t",ylab="predicted N2_t+1")
+Gradient <- constructGradient(m.7.sample,focalVariable="E",non.focalVariables=list(Esq=list(2)),ngrid=39)
+predY <- predict(m.7.sample,XData=Gradient$XDataNew,expected=TRUE)
+a <- plotGradient(m.7.sample,Gradient,pred=predY,showData=T,measure="Y",index=1,main="",xlab="E_t",ylab="predicted N1_t+1")
+b <- plotGradient(m.7.sample,Gradient,pred=predY,showData=T,measure="Y",index=2,main="",xlab="E_t",ylab="predicted N2_t+1")
 
-postBeta = getPostEstimate(m.5.sample, parName = "Beta")
-plotBeta(m.5.sample, post = postBeta, param = "Mean", supportLevel = 0.7)
+postBeta = getPostEstimate(m.7.sample, parName = "Beta")
+plotBeta(m.7.sample, post = postBeta, param = "Mean", supportLevel = 0.7)
+```
+```{r warning=FALSE}
+VP <- computeVariancePartitioning(m.7.sample,group=c(1,1,1,2,3),groupnames=c("Env","Sp1","Sp2"))
+plotVariancePartitioning(m.7.sample,VP,cols=c("white","skyblue","darkgrey","orange"),args.legend=list(cex=0.75,bg="transparent"))
+```
+We can also look at a gradient plot for the effect of evolution in Species 1 for abundance in Species 1 and Species 2.
+
+```{r}
+Gradient <- constructGradient(m.7.sample,focalVariable="dx1",non.focalVariables=list(E=list(2),Esq=list(2),dx2=list(2)),ngrid=39)
+Gradient$XDataNew$Esq <- Gradient$XDataNew$E^2
+predY <- predict(m.7.sample,XData=Gradient$XDataNew,expected=TRUE)
+a <- plotGradient(m.7.sample,Gradient,pred=predY,showData=T,measure="Y",index=1,main="",xlab="dx1_t",ylab="predicted N1_t+1")
+b <- plotGradient(m.7.sample,Gradient,pred=predY,showData=T,measure="Y",index=2,main="",xlab="dx1_t",ylab="predicted N2_t+1")
+
+postBeta = getPostEstimate(m.7.sample, parName = "Beta")
+plotBeta(m.7.sample, post = postBeta, param = "Mean", supportLevel = 0.7)
 ```
 
-XX Interpret environmental terms
+We repeat this plot for the effect of evolution in Species 2 for abundance in Species 1 and Species 2.
+
+```{r}
+Gradient <- constructGradient(m.7.sample,focalVariable="dx2",non.focalVariables=list(E=list(2),Esq=list(2),dx1=list(2)),ngrid=39)
+Gradient$XDataNew$Esq <- Gradient$XDataNew$E^2
+predY <- predict(m.7.sample,XData=Gradient$XDataNew,expected=TRUE)
+a <- plotGradient(m.7.sample,Gradient,pred=predY,showData=T,measure="Y",index=1,main="",xlab="dx2_t",ylab="predicted N1_t+1")
+b <- plotGradient(m.7.sample,Gradient,pred=predY,showData=T,measure="Y",index=2,main="",xlab="dx2_t",ylab="predicted N2_t+1")
+```
+
+THESE SHOULD NOT BE THE SAME
+
+We repeat the analysis for
 

### 2024-07-26 manuscript/Appendix_S2_v01.Rmd
Evolution alt hypotheses tested

@@
 
+We hypothesize that \( \lvert E_t - x_{i,t} \rvert \) will be the best predictor of species abundances, and that \( \lvert x_{t+1} - x_{i,t} \rvert \) will have similar explanatory power.
+
+```{r}
+# prepare data in HMSC format
+dat <- as.data.frame(cbind(log(N$N1),log(N$N2),x$x1,x$x2))
+colnames(dat) <- c("N1","N2","x1","x2")
+dat$time <- 1:t
+df <- data.frame(cbind(dat$N1[2:t],dat$N2[2:t],dat$x1[2:t],dat$x2[2:t]))
+colnames(df) <- c("Nt1","Nt2","xt1","xt2")
+df$dx1 <- abs(dat$x1[2:t] - dat$x1[1:(t-1)])
+df$dx2 <- abs(dat$x2[2:t] - dat$x2[1:(t-1)])
+# Y matrix
+Y <- as.matrix(cbind(df$Nt1,df$Nt2))
+# X matrix
+XData <- data.frame(cbind(dat$N1[1:(t-1)],dat$N2[1:(t-1)]),E[1:(t-1)],E[1:(t-1)]^2,dat$x1[1:(t-1)],dat$x2[1:(t-1)],abs(dat$x1[2:t] - dat$x1[1:(t-1)]),abs(dat$x2[2:t] - dat$x2[1:(t-1)]),abs(E[1:(t-1)] - dat$x1[1:(t-1)]),abs(E[1:(t-1)] - dat$x2[1:(t-1)]))
+colnames(XData) <- c("n1","n2","E","Esq","x1","x2","dx1","dx2","dEx1","dEx2")
+# Prepare HMSC model
+studyDesign = data.frame(sample = as.factor(1:(t-1)))
+rL = HmscRandomLevel(units = studyDesign$sample)
+# Design and fit 4 alternative HMSC models
+models = list()
+for(i in 1:4){
+  XFormula = switch(i, ~1, ~E+Esq+x1+x2, ~E+Esq+dEx1+dEx2, ~E+Esq+dx1+dx2)
+  m = Hmsc(Y = Y, XData = XData, XFormula = XFormula, studyDesign = studyDesign, ranLevels = list(sample = rL))
+  models[[i]] = m
+}
+# Bayesian model parameters
+nChains <- 2
+thin <- 5
+samples <- 2000
+transient <- 100*thin
+verbose <- 500*thin
+for(i in 1:4){
+  models[[i]] = sampleMcmc(models[[i]],thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
+}
+```
+
+We check the explanatory and predictive power of each model using cross-validation.
+
+```{r}
+# After Ovaskainen & Abrego 2020, Chapter 7
+partition = createPartition(m,nfolds=2,column="sample")
+partition.sp = c(1,2)
+result = matrix(NA,nrow=3,ncol=4)
+for(i in 1:4){
+  m = models[[i]]
+  # Explanatory power
+  preds = computePredictedValues(m)
+  MF = evaluateModelFit(hM=m,predY=preds)
+  result[1,i] = mean(MF$R2)
+  # Predictive power based on cross-validation
+  preds = computePredictedValues(m,partition=partition)
+  MF = evaluateModelFit(hM=m,predY=preds)
+  result[2,i] <- mean(MF$R2)
+  # Predictive power based on conditional cross-validation
+  preds = computePredictedValues(m,partition=partition,partition.sp=partition.sp,mcmcStep=100)
+  MF = evaluateModelFit(hM=m,predY=preds)
+  result[3,i] = mean(MF$R2)
+}
+```
+
+```{r}
+result
+```
+
+We can see that the \( \lvert E_t - x_{i,t} \rvert \) and \( \lvert x_{t+1} - x_{i,t} \rvert \) do have similar explanatory power, however, the species traits were in fact the stronger predictor of species abundances.
+I finally ask whether a model with raw trait values and change in trait values has imprived explanatory power (if the magnitude of evolutionary change has any additional explanatory role).
+
+```{r}
+# Design and fit 4 alternative HMSC models
+XFormula = ~E+Esq+x1+x2+dx1+dx2
+models[[5]] = Hmsc(Y = Y, XData = XData, XFormula = XFormula, studyDesign = studyDesign, ranLevels = list(sample = rL))
+models[[5]] =sampleMcmc(models[[5]],thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
+# Explanatory power
+result2 <- matrix(NA,nrow=3,ncol=5)
+result2[,1:4] <- result
+for(i in 5){
+  m = models[[i]]
+  # Explanatory power
+  preds = computePredictedValues(m)
+  MF = evaluateModelFit(hM=m,predY=preds)
+  result2[1,i] = mean(MF$R2)
+  # Predictive power based on cross-validation
+  preds = computePredictedValues(m,partition=partition)
+  MF = evaluateModelFit(hM=m,predY=preds)
+  result2[2,i] <- mean(MF$R2)
+  # Predictive power based on conditional cross-validation
+  preds = computePredictedValues(m,partition=partition,partition.sp=partition.sp,mcmcStep=100)
+  MF = evaluateModelFit(hM=m,predY=preds)
+  result2[3,i] = mean(MF$R2)
+}
+```
 ```{r}

### 2024-09-07 manuscript/Appendix_S1_v01.Rmd
Full draft complete

@@
 
-We use the same examples as above, with weaker then stronger strength of interspecific interactions.
+We use one of the same examples as above, with weak strength of interspecific interactions.
 
@@
 dat <- melt(N, id.vars="time")
-ggplot2::ggplot(dat, aes(time, value, col=variable)) + geom_point() + geom_hline(yintercept = ((r1.0 - 1)/alpha.11),linetype = "dashed", color = "gray")
+ggplot2::ggplot(dat, aes(time, value, col=variable)) + geom_point()
 ```
@@
 The overall research goal is to determine the effect of trait evolution for species abundances.
+We hypothesize that \( \lvert x_{t+1} - x_{i,t} \rvert \) is an important driver of species abundances, alongside the environment and species interactions.
+We plan to fit a statistical model to estimate the impact of $E$, $E^2$, and \( \lvert x_{t+1} - x_{i,t} \rvert \) as fixed effects, and time points as a random effect (to estimate species-to-species associations).
+However, we first plot the effects of interest as scatter plots with species abundances.
+
+```{r}
+dat2 <- as.data.frame(cbind(log(N$N1),log(N$N2),x$x1,x$x2))
+colnames(dat2) <- c("N1","N2","x1","x2")
+dat2$time <- 1:t
+df <- data.frame(cbind(dat2$N1[2:t],dat2$N2[2:t],dat2$x1[2:t],dat2$x2[2:t]))
+colnames(df) <- c("Nt1","Nt2","xt1","xt2")
+df$dx1 <- abs(dat2$x1[2:t] - dat2$x1[1:(t-1)])
+df$dx2 <- abs(dat2$x2[2:t] - dat2$x2[1:(t-1)])
+# Plot
+p1 <- ggplot2::ggplot(df, aes(dx1, Nt1)) + geom_point()
+p2 <- ggplot2::ggplot(df, aes(dx2, Nt1)) + geom_point()
+p3 <- ggplot2::ggplot(df, aes(dx1, Nt2)) + geom_point()
+p4 <- ggplot2::ggplot(df, aes(dx2, Nt2)) + geom_point()
+p1 + p2 + p3 + p4
+```
+We see that the effects are not going to be well described by regression coefficients, as there are distinct impacts of change in trait values for population abundances at low vs.
+high change in trait value.
+We consider instead the impacts of trait change for species growth rates.
+
+```{r}
+dat2$r1 <- r$r1
+dat2$r2 <- r$r2
+df$r1 <- dat2$r1[2:t]
+df$r2 <- dat2$r2[2:t]
+
+df$dr1 <- abs(dat2$r1[2:t] - dat2$r1[1:(t-1)])
+df$dr2 <- abs(dat2$r2[2:t] - dat2$r2[1:(t-1)])
+# Plot
+p1 <- ggplot2::ggplot(df, aes(dx1, r1)) + geom_point()
+p2 <- ggplot2::ggplot(df, aes(dx2, r1)) + geom_point()
+p3 <- ggplot2::ggplot(df, aes(dx1, r2)) + geom_point()
+p4 <- ggplot2::ggplot(df, aes(dx2, r2)) + geom_point()
+p1 + p2 + p3 + p4
+```
+
+We now use the species growth rates as the response variable in the model, with $E$, $E^2$, and \( \lvert x_{t+1} - x_{i,t} \rvert \) as fixed effects, and time points as a random effect (to estimate species-to-species associations).
 We use the absolute value of the change in trait value from one time point to the next (the total amount of trait change, regardless of direction) as a fixed effect in the HMSC.
@@
 # Plot simulation: ggplot
-dat <- as.data.frame(cbind(log(N$N1),log(N$N2),x$x1,x$x2))
-colnames(dat) <- c("N1","N2","x1","x2")
+dat <- as.data.frame(cbind(r$r1,r$r2,x$x1,x$x2))
+colnames(dat) <- c("r1","r2","x1","x2")
 dat$time <- 1:t
-df <- data.frame(cbind(dat$N1[2:t],dat$N2[2:t],dat$x1[2:t],dat$x2[2:t]))
-colnames(df) <- c("Nt1","Nt2","xt1","xt2")
+df <- data.frame(cbind(dat$r1[2:t],dat$r2[2:t],dat$x1[2:t],dat$x2[2:t]))
+colnames(df) <- c("rt1","rt2","xt1","xt2")
 # prepare data in HMSC format
-Y <- as.matrix(cbind(df$Nt1,df$Nt2))
-XData <- data.frame(cbind(dat$N1[1:(t-1)],dat$N2[1:(t-1)]),E[1:(t-1)],E[1:(t-1)]^2,abs(dat$x1[2:t] - dat$x1[1:(t-1)]),abs(dat$x2[2:t] - dat$x2[1:(t-1)]))
-colnames(XData) <- c("n1","n2","E","Esq","dx1","dx2")
+Y <- as.matrix(cbind(df$rt1,df$rt2))
+XData <- data.frame(cbind(E[1:(t-1)],E[1:(t-1)]^2,abs(dat$x1[2:t] - dat$x1[1:(t-1)]),abs(dat$x2[2:t] - dat$x2[1:(t-1)])))
+colnames(XData) <- c("E","Esq","dx1","dx2")
 studyDesign = data.frame(sample = as.factor(1:(t-1)))
@@
 samples <- 2000
-transient <- 100*thin
+transient <- 1000*thin
 verbose <- 500*thin

### 2024-09-07 manuscript/Appendix_S2_v01.Rmd
Full draft complete

@@
 ```{r}
+colnames(result) <- c("null","xt","dEx","dx")
+rownames(result) <- c("R2_expl","R2_pred","R2_pred_cross")
 result
@@
 
-We can see that the \( \lvert E_t - x_{i,t} \rvert \) and \( \lvert x_{t+1} - x_{i,t} \rvert \) do have similar explanatory power, however, the species traits were in fact the stronger predictor of species abundances.
-I finally ask whether a model with raw trait values and change in trait values has imprived explanatory power (if the magnitude of evolutionary change has any additional explanatory role).
+An important conclusion from this analysis is that the \( \lvert E_t - x_{i,t} \rvert \) and \( \lvert x_{t+1} - x_{i,t} \rvert \) do have similar explanatory power, and that - even though the species traits were the stronger predictor of species abundances - the model with total change in trait value can be used to assess the question "How does evolution impact species abundances?"
 
 ```{r}
-# Design and fit 4 alternative HMSC models
-XFormula = ~E+Esq+x1+x2+dx1+dx2
-models[[5]] = Hmsc(Y = Y, XData = XData, XFormula = XFormula, studyDesign = studyDesign, ranLevels = list(sample = rL))
-models[[5]] =sampleMcmc(models[[5]],thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
-# Explanatory power
-result2 <- matrix(NA,nrow=3,ncol=5)
-result2[,1:4] <- result
-for(i in 5){
-  m = models[[i]]
-  # Explanatory power
-  preds = computePredictedValues(m)
-  MF = evaluateModelFit(hM=m,predY=preds)
-  result2[1,i] = mean(MF$R2)
-  # Predictive power based on cross-validation
-  preds = computePredictedValues(m,partition=partition)
-  MF = evaluateModelFit(hM=m,predY=preds)
-  result2[2,i] <- mean(MF$R2)
-  # Predictive power based on conditional cross-validation
-  preds = computePredictedValues(m,partition=partition,partition.sp=partition.sp,mcmcStep=100)
-  MF = evaluateModelFit(hM=m,predY=preds)
-  result2[3,i] = mean(MF$R2)
-}
-```
-```{r}
-knitr::knit_exit()
-```
-
-```{r}
-m7.post.hmsc <- convertToCodaObject(m.7.sample)
-summary(m7.post.hmsc$Beta)
-bayesplot::mcmc_trace(m7.post.hmsc$Beta)
-bayesplot::mcmc_areas(m7.post.hmsc$Beta,area_method = c("equal height"))
+m.post.hmsc <- convertToCodaObject(models[[4]])
+bayesplot::mcmc_trace(m.post.hmsc$Beta)
+bayesplot::mcmc_areas(m.post.hmsc$Beta,area_method = c("equal height"))
 ```
@@
 # Bayesian estimates
-summary(m7.post.hmsc$Beta)$statistics[1:10,1:2]
+summary(m.post.hmsc$Beta)$statistics[1:10,1:2]
 ```
@@
 ```{r}
-OmegaCor = computeAssociations(m.7.sample)
+OmegaCor = computeAssociations(models[[4]])
 supportLevel = 0.7
@@
 + (OmegaCor[[1]]$support<(1-supportLevel))>0)*OmegaCor[[1]]$mean
-corrplot::corrplot(toPlot, method = "color",col = colorRampPalette(c("blue","white","red"))(200),title = paste("random effect level:", m.7.sample$rLNames[1]), mar=c(0,0,1,0))
+corrplot::corrplot(toPlot, method = "color",col = colorRampPalette(c("blue","white","red"))(200),title = paste("random effect level:", models[[4]]$rLNames[1]), mar=c(0,0,1,0))
 OmegaCor
-```
-
-```{r}
-Gradient <- constructGradient(m.7.sample,focalVariable="E",non.focalVariables=list(Esq=list(2)),ngrid=39)
-predY <- predict(m.7.sample,XData=Gradient$XDataNew,expected=TRUE)
-a <- plotGradient(m.7.sample,Gradient,pred=predY,showData=T,measure="Y",index=1,main="",xlab="E_t",ylab="predicted N1_t+1")
-b <- plotGradient(m.7.sample,Gradient,pred=predY,showData=T,measure="Y",index=2,main="",xlab="E_t",ylab="predicted N2_t+1")
-
-postBeta = getPostEstimate(m.7.sample, parName = "Beta")
-plotBeta(m.7.sample, post = postBeta, param = "Mean", supportLevel = 0.7)
-```
-```{r warning=FALSE}
-VP <- computeVariancePartitioning(m.7.sample,group=c(1,1,1,2,3),groupnames=c("Env","Sp1","Sp2"))
-plotVariancePartitioning(m.7.sample,VP,cols=c("white","skyblue","darkgrey","orange"),args.legend=list(cex=0.75,bg="transparent"))
 ```
@@
 ```{r}
-Gradient <- constructGradient(m.7.sample,focalVariable="dx1",non.focalVariables=list(E=list(2),Esq=list(2),dx2=list(2)),ngrid=39)
+Gradient <- constructGradient(models[[4]],focalVariable="dx1",non.focalVariables=list(E=list(2),Esq=list(2),dx2=list(2)),ngrid=39)
 Gradient$XDataNew$Esq <- Gradient$XDataNew$E^2
-predY <- predict(m.7.sample,XData=Gradient$XDataNew,expected=TRUE)
-a <- plotGradient(m.7.sample,Gradient,pred=predY,showData=T,measure="Y",index=1,main="",xlab="dx1_t",ylab="predicted N1_t+1")
-b <- plotGradient(m.7.sample,Gradient,pred=predY,showData=T,measure="Y",index=2,main="",xlab="dx1_t",ylab="predicted N2_t+1")
+predY <- predict(models[[4]],XData=Gradient$XDataNew,expected=TRUE)
+a <- plotGradient(models[[4]],Gradient,pred=predY,showData=T,measure="Y",index=1,main="",xlab="dx1_t",ylab="predicted N1_t+1")
+b <- plotGradient(models[[4]],Gradient,pred=predY,showData=T,measure="Y",index=2,main="",xlab="dx1_t",ylab="predicted N2_t+1")
 
-postBeta = getPostEstimate(m.7.sample, parName = "Beta")
-plotBeta(m.7.sample, post = postBeta, param = "Mean", supportLevel = 0.7)
+postBeta = getPostEstimate(models[[4]], parName = "Beta")
+plotBeta(models[[4]], post = postBeta, param = "Mean", supportLevel = 0.7)
 ```
-
 We repeat this plot for the effect of evolution in Species 2 for abundance in Species 1 and Species 2.
@@
 ```{r}
-Gradient <- constructGradient(m.7.sample,focalVariable="dx2",non.focalVariables=list(E=list(2),Esq=list(2),dx1=list(2)),ngrid=39)
+Gradient <- constructGradient(models[[4]],focalVariable="dx2",non.focalVariables=list(E=list(2),Esq=list(2),dx1=list(2)),ngrid=39)
 Gradient$XDataNew$Esq <- Gradient$XDataNew$E^2
-predY <- predict(m.7.sample,XData=Gradient$XDataNew,expected=TRUE)
-a <- plotGradient(m.7.sample,Gradient,pred=predY,showData=T,measure="Y",index=1,main="",xlab="dx2_t",ylab="predicted N1_t+1")
-b <- plotGradient(m.7.sample,Gradient,pred=predY,showData=T,measure="Y",index=2,main="",xlab="dx2_t",ylab="predicted N2_t+1")
+predY <- predict(models[[4]],XData=Gradient$XDataNew,expected=TRUE)
+a <- plotGradient(models[[4]],Gradient,pred=predY,showData=T,measure="Y",index=1,main="",xlab="dx2_t",ylab="predicted N1_t+1")
+b <- plotGradient(models[[4]],Gradient,pred=predY,showData=T,measure="Y",index=2,main="",xlab="dx2_t",ylab="predicted N2_t+1")
 ```
+To evaluate the relative importance of evolution for species abundances, we use variation partition:
 
-THESE SHOULD NOT BE THE SAME
-
-We repeat the analysis for
-
-# References
+```{r warning=FALSE}
+VP <- computeVariancePartitioning(models[[4]],group=c(1,1,1,2,3),groupnames=c("Env","Sp1","Sp2"))
+plotVariancePartitioning(models[[4]],VP,cols=c("white","skyblue","darkgrey","orange"),args.legend=list(cex=0.75,bg="transparent"))
+```
+We see that trait evolution in Species 1 is an important driver of abundances in both Species 1 and Species 2.

### 2024-09-08 manuscript/Appendix_S1_v01.Rmd
Completed App2, App1 evolution, knit pdf

--- 2024-07-23 manuscript/Appendix_S1_v01.Rmd
@@
 
+```{r warning=FALSE}
+VP <- computeVariancePartitioning(m.6.sample,group=c(1,1,1),groupnames="Env")
+plotVariancePartitioning(m.6.sample,VP,args.legend=list(cex=0.75,bg="transparent"))
+```
+
 # Species growth, competition, environmental change, and trait evolution
 
+We now consider that trait evolution can occur, with an impact on the species growth rate / fitness in the context of the environment.
+
+## Trait evolution
+
+Previously the population growth rate was fixed as \( r_i = \hat{W}e^{-(E-x_{i,t})^2}\). However, we now introduce the quantitative genetic model of evolutionary rescue [@Gomulkiewicz1995].
+
+The growth equation for each species now becomes:
+
+$$ N_{i,t+1} =\frac{\hat{W}e^ \frac{-[(\frac{w+(1-h^2)P} {P+w})(E-x_{i,t})]^2}{2(P+w)}N_{i,t}} {1 + \alpha_{ii}N_{i,t} + \alpha_{ij} N_{j,t}} $$
+
+where \( \hat{W} \) is calculated as \( \hat{W}=W_{max}\sqrt(\frac{w}{P+w}) \), \(W_{max}\) is the species' maximum per-capita growth rate, *w* is the width of the Gaussian fitness function (which determines the strength of selection, as increasing values indicate a weaker reduction in fitness with distance from optimum trait value), *P* is the width of the distribution of the phenotype *x*, and $h^2$ is the heritability of the trait *x*.
+For the simulation we use \(W_{max}\) = 2, *P* = 1, and *w* = 2.
+
+The change in the average trait value each time step is given by:
+
+$$ d_{i,t+1} = k d_{i,t} $$
+where \( k = \frac{w + (1-h^2) P}{w + P} \) and \( d_{i,t} = E_t - x_{i,t} \).
+
+## Population dynamics simulation
+
+We use one of the same examples as above, with weak strength of interspecific interactions.
+We use initial trait values that favor the weaker competitor, and we simulate random increasing shifts in the local environment.
+
+```{r}
+# Case 1: Weaker interactions, stronger environment
+# Simulate initial species population growth with environment fluctuations
+N1.0 <- 10
+N2.0 <- 10
+alpha.11 <- 0.01
+alpha.22 <- 0.01
+alpha.12 <- 0.005
+alpha.21 <- 0.01
+E.0 <- 0.8
+x1.0 <- 0.1
+x2.0 <- 0.8
+P <- 1
+w <- 2
+Wmax <- 2
+h2 <- 1
+k <- (w + (1 - h2) * P)/(P + w)
+
+# model function
+disc_LV_evol <- function(N1.0,N2.0,alpha.11,alpha.22,alpha.12,alpha.21,E,x1.0,x2.0,P,w,Wmax,h2) {
+  What <- Wmax*sqrt(w/(P+w))
+  r1 <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E-x1.0))^2)/(2*(P+w)))
+  Nt1 <- (r1*N1.0) / (1+alpha.11*N1.0+alpha.12*N2.0)
+  r2 <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E-x2.0))^2)/(2*(P+w)))
+  Nt2 <- (r2*N2.0) / (1+alpha.22*N2.0+alpha.21*N1.0)
+  return(c(Nt1,Nt2,r1,r2))
+}
+# Simulation of model for t time steps
+t <- 40
+N <- array(NA,dim=c(t,2))
+N <- as.data.frame(N)
+colnames(N) <- c("N1","N2")
+x <- array(NA,dim=c(t,2))
+x <- as.data.frame(x)
+colnames(x) <- c("x1","x2")
+r <- array(NA,dim=c(t,2))
+r <- as.data.frame(r)
+colnames(r) <- c("r1","r2")
+N$N1[1] <- N1.0
+N$N2[1] <- N2.0
+x$x1[1] <- x1.0
+x$x2[1] <- x2.0
+What <- Wmax*sqrt(w/(P+w))
+r$r1[1] <- What*(exp((-(((w+(1-h2)*P)/(P+w))*(E.0-x1.0))^2)/(2*(P+w))))
+r$r2[1] <- What*(exp((-(((w+(1-h2)*P)/(P+w))*(E.0-x2.0))^2)/(2*(P+w))))
+E <- rep(NA, t)
+E[1] <- E.0
+for (i in 2:t) {
+  res <- disc_LV_evol(N1.0=N[i-1,1],N2.0=N[i-1,2],alpha.11=alpha.11,alpha.22=alpha.22,alpha.12=alpha.12,alpha.21=alpha.21,E=E[i-1],x1=x[i-1,1],x2=x[i-1,2],P=P,w=w,Wmax=Wmax,h2=h2)
+  N$N1[i] <- res[1]
+  N$N2[i] <- res[2]
+  r$r1[i] <- res[3]
+  r$r2[i] <- res[4]
+  # trait change
+  d1 <- E[i-1] - x[i-1,1]
+  d2 <- E[i-1] - x[i-1,2]
+  d1.1 <- k*d1
+  d2.1 <- k*d2
+  x$x1[i] <- E[i-1] - d1.1
+  x$x2[i] <- E[i-1] - d2.1
+  # environmental change
+  E[i] <- E[i-1] + abs(rnorm(1, 0, 0.05))
+}
+```
+```{r}
+# Plot simulation: ggplot
+N$time <- 1:t
+dat <- melt(N, id.vars="time")
+ggplot2::ggplot(dat, aes(time, value, col=variable)) + geom_point()
+```
+
+The overall research goal is to determine the effect of trait evolution for species abundances.
+We hypothesize that \( \lvert x_{t+1} - x_{i,t} \rvert \) is an important driver of species abundances, alongside the environment and species interactions.
+We plan to fit a statistical model to estimate the impact of $E$, $E^2$, and \( \lvert x_{t+1} - x_{i,t} \rvert \) as fixed effects, and time points as a random effect (to estimate species-to-species associations).
+We first plot the effects of interest as scatter plots with species abundances.
+
+```{r}
+dat2 <- as.data.frame(cbind(log(N$N1),log(N$N2),x$x1,x$x2))
+colnames(dat2) <- c("N1","N2","x1","x2")
+dat2$time <- 1:t
+df <- data.frame(cbind(dat2$N1[2:t],dat2$N2[2:t],dat2$x1[2:t],dat2$x2[2:t]))
+colnames(df) <- c("Nt1","Nt2","xt1","xt2")
+df$dx1 <- abs(dat2$x1[2:t] - dat2$x1[1:(t-1)])
+df$dx2 <- abs(dat2$x2[2:t] - dat2$x2[1:(t-1)])
+# Plot
+p1 <- ggplot2::ggplot(df, aes(dx1, Nt1)) + geom_point()
+p2 <- ggplot2::ggplot(df, aes(dx2, Nt1)) + geom_point()
+p3 <- ggplot2::ggplot(df, aes(dx1, Nt2)) + geom_point()
+p4 <- ggplot2::ggplot(df, aes(dx2, Nt2)) + geom_point()
+p1 + p2 + p3 + p4
+```
+We expect that change in traits in Species 1 will have an effect on Species abundances, and we do not expect a strong impact of change in traits in Species 2.
+
+We now use the species abundances as the response variable in the model, with $E$, $E^2$, and \( \lvert x_{t+1} - x_{i,t} \rvert \) as fixed effects, and time points as a random effect (to estimate species-to-species associations).
+We use the absolute value of the change in trait value from one time point to the next (the total amount of trait change, regardless of direction) as a fixed effect in the HMSC.
+
+```{r}
+# Plot simulation: ggplot
+dat <- as.data.frame(cbind(log(N$N1), log(N$N2), x$x1, x$x2))
+colnames(dat) <- c("N1", "N2", "x1", "x2")
+dat$time <- 1:t
+df <- data.frame(cbind(dat$N1[2:t], dat$N2[2:t], dat$x1[2:t], dat$x2[2:t]))
+colnames(df) <- c("Nt1", "Nt2", "xt1", "xt2")
+df$dx1 <- abs(dat$x1[2:t] - dat$x1[1:(t - 1)])
+df$dx2 <- abs(dat$x2[2:t] - dat$x2[1:(t - 1)])
+# prepare data in HMSC format
+Y <- as.matrix(cbind(df$Nt1, df$Nt2))
+XData <- data.frame(cbind(E[1:(t-1)],E[1:(t-1)]^2,abs(dat$x1[2:t] - dat$x1[1:(t-1)]),abs(dat$x2[2:t] - dat$x2[1:(t-1)])))
+colnames(XData) <- c("E","Esq","dx1","dx2")
+studyDesign = data.frame(sample = as.factor(1:(t-1)))
+rL = HmscRandomLevel(units = studyDesign$sample)
+m.7.hmsc = Hmsc(Y = Y, XData = XData, XFormula = ~E+Esq+dx1+dx2, studyDesign = studyDesign, ranLevels = list(sample = rL))
+# Bayesian model parameters
+nChains <- 2
+thin <- 5
+samples <- 2000
+transient <- 1000*thin
+verbose <- 500*thin
+# sample MCMC
+m.7.sample <- sampleMcmc(m.7.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
+```
+```{r}
+m7.post.hmsc <- convertToCodaObject(m.7.sample)
+summary(m7.post.hmsc$Beta)
+bayesplot::mcmc_trace(m7.post.hmsc$Beta)
+bayesplot::mcmc_areas(m7.post.hmsc$Beta,area_method = c("equal height"))
+```
+
+```{r}
+# Bayesian estimates
+cbind(summary(m7.post.hmsc$Beta)$statistics[1:10, 1], summary(m7.post.hmsc$Beta)$quantiles[, c(1, 5)])
+```
+
+```{r}
+OmegaCor = computeAssociations(m.7.sample)
+supportLevel = 0.95
+toPlot = ((OmegaCor[[1]]$support>supportLevel)
++ (OmegaCor[[1]]$support<(1-supportLevel))>0)*OmegaCor[[1]]$mean
+corrplot::corrplot(toPlot, method = "color",col = colorRampPalette(c("blue","white","red"))(200),title = paste("random effect level:", m.7.sample$rLNames[1]), mar=c(0,0,1,0))
+OmegaCor
+```
+
+```{r}
+Gradient <- constructGradient(m.7.sample,focalVariable="E",non.focalVariables=list(Esq=list(2),dx1=list(1),dx2=list(1)),ngrid=39)
+# Esq is manually constructed as gradient-produced E^2
+Gradient$XDataNew$Esq <- Gradient$XDataNew$E^2
+predY <- predict(m.7.sample,XData=Gradient$XDataNew,expected=TRUE)
+a <- plotGradient(m.7.sample,Gradient,pred=predY,showData=T,measure="Y",index=1,main="",xlab="E_t",ylab="predicted N1_t+1")
+b <- plotGradient(m.7.sample,Gradient,pred=predY,showData=T,measure="Y",index=2,main="",xlab="E_t",ylab="predicted N2_t+1")
+
+postBeta = getPostEstimate(m.7.sample, parName = "Beta")
+plotBeta(m.7.sample, post = postBeta, param = "Mean", supportLevel = 0.95)
+```
+```{r warning=FALSE}
+VP <- computeVariancePartitioning(m.7.sample,group=c(1,1,1,2,3),groupnames=c("Env","Sp1","Sp2"))
+plotVariancePartitioning(m.7.sample,VP,cols=c("white","skyblue","darkgrey","orange"),args.legend=list(cex=0.75,bg="transparent"))
+```
+We can also look at a gradient plot for the effect of evolution in Species 1 for abundance in Species 1 and Species 2.
+
+```{r}
+Gradient <- constructGradient(m.7.sample,focalVariable="dx1",non.focalVariables=list(E=list(1),Esq=list(2),dx2=list(1)),ngrid=39)
+Gradient$XDataNew$Esq <- Gradient$XDataNew$E^2
+predY <- predict(m.7.sample,XData=Gradient$XDataNew,expected=TRUE)
+a <- plotGradient(m.7.sample,Gradient,pred=predY,showData=T,measure="Y",index=1,main="",xlab="dx1_t",ylab="predicted N1_t+1")
+b <- plotGradient(m.7.sample,Gradient,pred=predY,showData=T,measure="Y",index=2,main="",xlab="dx1_t",ylab="predicted N2_t+1")
+
+postBeta = getPostEstimate(m.7.sample, parName = "Beta")
+plotBeta(m.7.sample, post = postBeta, param = "Mean", supportLevel = 0.95)
+```
+
+We repeat this plot for the effect of evolution in Species 2 for abundance in Species 1 and Species 2.
+
+```{r}
+Gradient <- constructGradient(m.7.sample,focalVariable="dx2",non.focalVariables=list(E=list(1),Esq=list(2),dx1=list(1)),ngrid=39)
+Gradient$XDataNew$Esq <- Gradient$XDataNew$E^2
+predY <- predict(m.7.sample,XData=Gradient$XDataNew,expected=TRUE)
+a <- plotGradient(m.7.sample,Gradient,pred=predY,showData=T,measure="Y",index=1,main="",xlab="dx2_t",ylab="predicted N1_t+1")
+b <- plotGradient(m.7.sample,Gradient,pred=predY,showData=T,measure="Y",index=2,main="",xlab="dx2_t",ylab="predicted N2_t+1")
+```
+
+The analysis indicates that we can successfully use the HMSC analysis to partition drivers of species abundances that include evolving traits.
+We see that change in Species 1 trait is a strong driver of abundances in Species1 and 2, and that trait change in Species 2 has a weaker impact.
+This aligns well with the observed simulation results.
+
+Though some of the main effect coefficients aren't well reflected in the data, we recall the model has strong predictive power when the elements are combined, with high predictive power to the observed data.
+
+```{r echo=FALSE}
+pd.1 <- predict(m.7.sample,XData=XData)
+# What do the predicted vs. observed values look like?
+## Calculate HPDI for each data point
+s <- ncol(m.7.sample$Y)
+pdv.1 <- lapply(pd.1,as.vector)
+pdm.1 <- do.call(rbind, pdv.1)
+pd_hdpi_25.1 <- apply(pdm.1,2,quantile,probs=0.025)
+pd_hdpi_975.1 <- apply(pdm.1,2,quantile,probs=0.975)
+pdm_hdpi_25.1 <- matrix(pd_hdpi_25.1,nrow=dim(pd.1[[1]])[1],ncol=dim(pd.1[[1]])[2],byrow=FALSE)
+pdm_hdpi_975.1 <- matrix(pd_hdpi_975.1,nrow=dim(pd.1[[1]])[1],ncol=dim(pd.1[[1]])[2],byrow=FALSE)
+plotY <- Y
+plot_25 <- pdm_hdpi_25.1
+plot_975 <- pdm_hdpi_975.1
+
+p_25 <- matrix(0,nrow=39,ncol=s)
+p_975 <- matrix(0,nrow=39,ncol=s)
+p_Y <- matrix(0,nrow=39,ncol=s)
+for (i in 1:s) {
+  p_25[,i] <- plot_25[,i]
+  p_975[,i] <- plot_975[,i]
+  p_Y[,i] <- plotY[,i]
+}
+d_25 <-as.data.frame(p_25)
+d_975 <-as.data.frame(p_975)
+d_Y <-as.data.frame(p_Y)
+
+d_25$time <- 2:40
+d_975$time <- 2:40
+d_Y$time <- 2:40
+
+d_25l <- pivot_longer(d_25,cols=1:2,names_to = "species",values_to = "Y")
+d_975l <- pivot_longer(d_975,cols=1:2,names_to = "species",values_to = "Y")
+d_Yl <- pivot_longer(d_Y,cols=1:2,names_to = "species",values_to = "Y")
+
+ggplot(d_Yl,aes(x=time,y=Y,group=species))+
+  geom_line(data=d_25l,aes(x=time,y=Y,group=species),color="orange",linewidth=0.35)+
+  geom_line(data=d_975l,aes(x=time,y=Y,group=species),color="orange",linewidth=0.35)+
+  geom_line(color="black",linewidth=0.35)+
+  theme_classic()+
+  ylab("Population density")
+```
+
+However, we can see the coefficients are less informative for inference.
+It is specific to this set of simulation conditions that the degree of evolution occurs at the beginning of the simulation, when the species is more mal-adapted.
+To better consider how trait evolution drives changes in population size, we use this change as the response variable Y in the predictive model.
+
+```{r}
+# Plot simulation: ggplot
+dat <- as.data.frame(cbind(log(N$N1), log(N$N2), x$x1, x$x2))
+colnames(dat) <- c("N1", "N2", "x1", "x2")
+dat$time <- 1:t
+df <- data.frame(cbind(dat$N1[2:t], dat$N2[2:t], dat$x1[2:t], dat$x2[2:t]))
+colnames(df) <- c("Nt1", "Nt2", "xt1", "xt2")
+df$dN1 <- abs(dat$N1[2:t] - dat$N1[1:(t - 1)])
+df$dN2 <- abs(dat$N2[2:t] - dat$N2[1:(t - 1)])
+df$dx1 <- abs(dat$x1[2:t] - dat$x1[1:(t - 1)])
+df$dx2 <- abs(dat$x2[2:t] - dat$x2[1:(t - 1)])
+# prepare data in HMSC format
+Y <- as.matrix(cbind(df$dN1, df$dN2))
+XData <- data.frame(cbind(E[1:(t-1)],E[1:(t-1)]^2,abs(dat$x1[2:t] - dat$x1[1:(t-1)]),abs(dat$x2[2:t] - dat$x2[1:(t-1)])))
+colnames(XData) <- c("E","Esq","dx1","dx2")
+studyDesign = data.frame(sample = as.factor(1:(t-1)))
+rL = HmscRandomLevel(units = studyDesign$sample)
+m.8.hmsc = Hmsc(Y = Y, XData = XData, XFormula = ~E+Esq+dx1+dx2, studyDesign = studyDesign, ranLevels = list(sample = rL))
+# Bayesian model parameters
+nChains <- 2
+thin <- 5
+samples <- 2000
+transient <- 1000*thin
+verbose <- 500*thin
+# sample MCMC
+m.8.sample <- sampleMcmc(m.8.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
+```
+```{r}
+m8.post.hmsc <- convertToCodaObject(m.8.sample)
+summary(m8.post.hmsc$Beta)
+bayesplot::mcmc_trace(m8.post.hmsc$Beta)
+bayesplot::mcmc_areas(m8.post.hmsc$Beta,area_method = c("equal height"))
+```
+
+```{r}
+Gradient <- constructGradient(m.8.sample,focalVariable="dx1",non.focalVariables=list(E=list(1),Esq=list(2),dx2=list(1)),ngrid=39)
+Gradient$XDataNew$Esq <- Gradient$XDataNew$E^2
+predY <- predict(m.8.sample,XData=Gradient$XDataNew,expected=TRUE)
+a <- plotGradient(m.8.sample,Gradient,pred=predY,showData=T,measure="Y",index=1,main="",xlab="dx1_t",ylab="predicted dN1_t+1")
+b <- plotGradient(m.8.sample,Gradient,pred=predY,showData=T,measure="Y",index=2,main="",xlab="dx1_t",ylab="predicted dN2_t+1")
+
+postBeta = getPostEstimate(m.8.sample, parName = "Beta")
+plotBeta(m.8.sample, post = postBeta, param = "Mean", supportLevel = 0.95)
+```
+
+We repeat this plot for the effect of evolution in Species 2 for abundance in Species 1 and Species 2.
+
+```{r}
+Gradient <- constructGradient(m.8.sample,focalVariable="dx2",non.focalVariables=list(E=list(1),Esq=list(2),dx1=list(1)),ngrid=39)
+Gradient$XDataNew$Esq <- Gradient$XDataNew$E^2
+predY <- predict(m.8.sample,XData=Gradient$XDataNew,expected=TRUE)
+a <- plotGradient(m.8.sample,Gradient,pred=predY,showData=T,measure="Y",index=1,main="",xlab="dx2_t",ylab="predicted dN1_t+1")
+b <- plotGradient(m.8.sample,Gradient,pred=predY,showData=T,measure="Y",index=2,main="",xlab="dx2_t",ylab="predicted dN2_t+1")
+```
+
+```{r warning=FALSE}
+VP <- computeVariancePartitioning(m.8.sample,group=c(1,1,1,2,3),groupnames=c("Env","Sp1","Sp2"))
+plotVariancePartitioning(m.8.sample,VP,cols=c("white","skyblue","darkgrey","orange"),args.legend=list(cex=0.75,bg="transparent"))
+```
+Here we can see that trait evolution in Species 1 had a strong positive relationship with the *change* in population size for both species. Trait evolution in Species 2 had much less of an effect.
+
 # References

### 2024-09-08 manuscript/Appendix_S2_v01.Rmd
Completed App2, App1 evolution, knit pdf

@@
 ```{r}
+set.seed(42)
 # Case 1: Weaker interactions, stronger environment
@@
 E.0 <- 0.8
-x1.0 <- 0.6
+x1.0 <- 0.1
 x2.0 <- 0.8
 P <- 1
-w <- 10
+w <- 2
 Wmax <- 2
-h2 <- 0.1
-k <- (w+(1-h2)*P)/(P+w)
+h2 <- 1
+k <- (w + (1 - h2) * P)/(P + w)
 
@@
 x$x2[1] <- x2.0
-r$r1[1] <- exp((-(((w+(1-h2)*P)/(P+w))*(E.0-x1.0))^2)/(2*(P+w)))
-r$r2[1] <- exp((-(((w+(1-h2)*P)/(P+w))*(E.0-x2.0))^2)/(2*(P+w)))
+What <- Wmax*sqrt(w/(P+w))
+r$r1[1] <- What*(exp((-(((w+(1-h2)*P)/(P+w))*(E.0-x1.0))^2)/(2*(P+w))))
+r$r2[1] <- What*(exp((-(((w+(1-h2)*P)/(P+w))*(E.0-x2.0))^2)/(2*(P+w))))
 E <- rep(NA, t)
@@
   # environmental change
-  E[i] <- E[i-1] + rnorm(1,0,.01)
+  E[i] <- E[i-1] + abs(rnorm(1,0,0.05))
 }
@@
 
+We can check the impact evolution had for this system by re-running dynamics with $h^2 = 0$:
+
+```{r echo = FALSE}
+h2 <- 0
+k <- (w+(1-h2)*P)/(P+w)
+# Simulation of model for t time steps
+Na <- array(NA,dim=c(t,2))
+Na <- as.data.frame(Na)
+colnames(Na) <- c("N1","N2")
+xa <- array(NA,dim=c(t,2))
+xa <- as.data.frame(xa)
+colnames(xa) <- c("x1","x2")
+ra <- array(NA,dim=c(t,2))
+ra <- as.data.frame(ra)
+colnames(ra) <- c("r1","r2")
+Na$N1[1] <- N1.0
+Na$N2[1] <- N2.0
+xa$x1[1] <- x1.0
+xa$x2[1] <- x2.0
+What <- Wmax*sqrt(w/(P+w))
+ra$r1[1] <- What*(exp((-(((w+(1-h2)*P)/(P+w))*(E.0-x1.0))^2)/(2*(P+w))))
+ra$r2[1] <- What*(exp((-(((w+(1-h2)*P)/(P+w))*(E.0-x2.0))^2)/(2*(P+w))))
+for (i in 2:t) {
+  res <- disc_LV_evol(N1.0=Na[i-1,1],N2.0=Na[i-1,2],alpha.11=alpha.11,alpha.22=alpha.22,alpha.12=alpha.12,alpha.21=alpha.21,E=E[i-1],x1=xa[i-1,1],x2=xa[i-1,2],P=P,w=w,Wmax=Wmax,h2=h2)
+  Na$N1[i] <- res[1]
+  Na$N2[i] <- res[2]
+  ra$r1[i] <- res[3]
+  ra$r2[i] <- res[4]
+  # trait change
+  d1 <- E[i-1] - xa[i-1,1]
+  d2 <- E[i-1] - xa[i-1,2]
+  d1.1 <- k*d1
+  d2.1 <- k*d2
+  xa$x1[i] <- E[i-1] - d1.1
+  xa$x2[i] <- E[i-1] - d2.1
+}
+# Plot simulation: ggplot
+Na$time <- 1:t
+data <- melt(Na, id.vars="time")
+ggplot2::ggplot(data, aes(time, value, col=variable)) + geom_point() + geom_hline(yintercept = (1/alpha.11),linetype = "dashed", color = "gray")
+```
+
 # Hypotheses for evolution as a driver of species abundances
 
-The growth equation is not linear, yet we fit abundance data to a heirarchical model with linear predictors because in most natural systems, a single theoretical growth model can't adequately capture the complexity in a given system.
+The growth equation is not linear, yet we fit abundance data to a hierarchical model with linear predictors because in most natural systems, a single theoretical growth model can't adequately capture the complexity in a given system.
 Linear models don't capture mechanistic relationships, but instead correlative.
@@
 # Bayesian estimates
-summary(m.post.hmsc$Beta)$statistics[1:10,1:2]
+cbind(summary(m.post.hmsc$Beta)$statistics[1:10,1],summary(m.post.hmsc$Beta)$quantiles[,c(1,5)])
 ```

### 2024-09-10 manuscript/Appendix_S1_v01.Rmd
Started spatial sim

@@
 library(reshape2)
+library(MASS)
 ```
@@
 
+# Species growth, competition, environmental change, and trait evolution in a spatially structured environment
+
+We now consider that trait evolution can occur, with an impact on the species growth rate / fitness in the context of the environment.
+
+A final consideration to add to the model is the existence of multiple sites that may have correlation structure in the variance of species abundances.
+We consider this in our simulation by adding multiple (10) sites, with no vs.
+high spatial covariance.
+We then consider spatial random effects in HMSC, where covariance is considered as a function of distance between sites.
+
+## Population dynamics simulation
+
+We use the same example as above, but now with 10 sites.
+To illustrate how HMSC can estimate the impacts of distance-dependent spatial covariance, we duplicate the population time series obtained in the example above, but we add in site variance and among-site covariance.
+
+```{r}
+# Case 1: Weaker interactions, stronger environment
+# Simulate initial species population growth with environment fluctuations
+N1.0 <- 10
+N2.0 <- 10
+alpha.11 <- 0.01
+alpha.22 <- 0.01
+alpha.12 <- 0.005
+alpha.21 <- 0.01
+E.0 <- 0.8
+x1.0 <- 0.1
+x2.0 <- 0.8
+P <- 1
+w <- 2
+Wmax <- 2
+h2 <- 1
+k <- (w + (1 - h2) * P)/(P + w)
+
+# model function
+disc_LV_evol <- function(N1.0,N2.0,alpha.11,alpha.22,alpha.12,alpha.21,E,x1.0,x2.0,P,w,Wmax,h2) {
+  What <- Wmax*sqrt(w/(P+w))
+  r1 <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E-x1.0))^2)/(2*(P+w)))
+  Nt1 <- (r1*N1.0) / (1+alpha.11*N1.0+alpha.12*N2.0)
+  r2 <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E-x2.0))^2)/(2*(P+w)))
+  Nt2 <- (r2*N2.0) / (1+alpha.22*N2.0+alpha.21*N1.0)
+  return(c(Nt1,Nt2,r1,r2))
+}
+# Simulation of model for t time steps, i sites
+j <- 10
+t <- 40
+N <- array(NA,dim=c(j*t,2))
+N <- as.data.frame(N)
+colnames(N) <- c("N1","N2")
+x <- array(NA,dim=c(j*t,2))
+x <- as.data.frame(x)
+colnames(x) <- c("x1","x2")
+r <- array(NA,dim=c(j*t,2))
+r <- as.data.frame(r)
+colnames(r) <- c("r1","r2")
+N$N1[1:j] <- N1.0
+N$N2[1:j] <- N2.0
+x$x1[1:j] <- x1.0
+x$x2[1:j] <- x2.0
+What <- Wmax*sqrt(w/(P+w))
+r$r1[1:j] <- What*(exp((-(((w+(1-h2)*P)/(P+w))*(E.0-x1.0))^2)/(2*(P+w))))
+r$r2[1:j] <- What*(exp((-(((w+(1-h2)*P)/(P+w))*(E.0-x2.0))^2)/(2*(P+w))))
+E <- rep(NA, t)
+E[1] <- E.0
+# Keep track of site and time
+study <- array(NA,dim=c(j*t,3))
+study <- as.data.frame(study)
+colnames(study) <- c("obs","site","time")
+study$obs <- 1:dim(study)[1]
+study$site <- rep(1:j,times=t)
+study$time <- rep(1:t,each=j)
+
+for (i in 2:t) {
+  for(z in 1:j){
+    res <- disc_LV_evol(N1.0=N[study$site==z & study$time==i-1,1],N2.0=N[study$site==z & study$time==i-1,2],alpha.11=alpha.11,alpha.22=alpha.22,alpha.12=alpha.12,alpha.21=alpha.21,E=E[i-1],x1=x[study$site==z & study$time==i-1,1],x2=x[study$site==z & study$time==i-1,2],P=P,w=w,Wmax=Wmax,h2=h2)
+    N$N1[study$site==z & study$time==i] <- res[1]
+    N$N2[study$site==z & study$time==i] <- res[2]
+    r$r1[study$site==z & study$time==i] <- res[3]
+    r$r2[study$site==z & study$time==i] <- res[4]
+    # trait change
+    d1 <- E[i-1] - x[study$site==z & study$time==i-1,1]
+    d2 <- E[i-1] - x[study$site==z & study$time==i-1,2]
+    d1.1 <- k*d1
+    d2.1 <- k*d2
+    x$x1[study$site==z & study$time==i] <- E[i-1] - d1.1
+    x$x2[study$site==z & study$time==i] <- E[i-1] - d2.1
+  }
+  # site-level random effect
+  sigma <- 0
+  sigma.spatial <- 2
+  alpha.spatial <- 0.5
+  # random site locations
+  xycoords = matrix(runif(2*j), ncol = 2)
+  Sigma = sigma.spatial^2*exp(-as.matrix(dist(xycoords))/alpha.spatial)
+  # draw from covariance matrix
+  a = mvrnorm(mu=rep(0,j), Sigma = Sigma)
+  N$N1[study$time==i] <- N$N1[study$time==i] + a
+  a = mvrnorm(mu=rep(0,j), Sigma = Sigma)
+  N$N2[study$time==i] <- N$N2[study$time==i] + a
+  # environmental change
+  E[i] <- E[i-1] + abs(rnorm(1, 0, 0.05))
+}
+```
+```{r}
+# Plot simulation: ggplot
+N$site <- study$site
+N$time <- study$time
+dat <- melt(N, id.vars=c("site","time"))
+ggplot2::ggplot(dat, aes(time, value, col=variable)) + geom_point() + facet_wrap(~site,nrow=2)
+```
+
+The overall research goal is to determine the effect of trait evolution for species abundances.
+We hypothesize that \( \lvert x_{t+1} - x_{i,t} \rvert \) is an important driver of species abundances, alongside the environment and species interactions.
+We plan to fit a statistical model to estimate the impact of $E$, $E^2$, and \( \lvert x_{t+1} - x_{i,t} \rvert \) as fixed effects, and time points as a random effect (to estimate species-to-species associations).
+We first plot the effects of interest as scatter plots with species abundances.
+
+```{r}
+dat2 <- as.data.frame(cbind(log(N$N1),log(N$N2),x$x1,x$x2))
+colnames(dat2) <- c("N1","N2","x1","x2")
+dat2$time <- 1:t
+df <- data.frame(cbind(dat2$N1[2:t],dat2$N2[2:t],dat2$x1[2:t],dat2$x2[2:t]))
+colnames(df) <- c("Nt1","Nt2","xt1","xt2")
+df$dx1 <- abs(dat2$x1[2:t] - dat2$x1[1:(t-1)])
+df$dx2 <- abs(dat2$x2[2:t] - dat2$x2[1:(t-1)])
+# Plot
+p1 <- ggplot2::ggplot(df, aes(dx1, Nt1)) + geom_point()
+p2 <- ggplot2::ggplot(df, aes(dx2, Nt1)) + geom_point()
+p3 <- ggplot2::ggplot(df, aes(dx1, Nt2)) + geom_point()
+p4 <- ggplot2::ggplot(df, aes(dx2, Nt2)) + geom_point()
+p1 + p2 + p3 + p4
+```
+We expect that change in traits in Species 1 will have an effect on Species abundances, and we do not expect a strong impact of change in traits in Species 2.
+
+We now use the species abundances as the response variable in the model, with $E$, $E^2$, and \( \lvert x_{t+1} - x_{i,t} \rvert \) as fixed effects, and time points as a random effect (to estimate species-to-species associations).
+We use the absolute value of the change in trait value from one time point to the next (the total amount of trait change, regardless of direction) as a fixed effect in the HMSC.
+
+```{r}
+# Plot simulation: ggplot
+dat <- as.data.frame(cbind(log(N$N1), log(N$N2), x$x1, x$x2))
+colnames(dat) <- c("N1", "N2", "x1", "x2")
+dat$time <- 1:t
+df <- data.frame(cbind(dat$N1[2:t], dat$N2[2:t], dat$x1[2:t], dat$x2[2:t]))
+colnames(df) <- c("Nt1", "Nt2", "xt1", "xt2")
+df$dx1 <- abs(dat$x1[2:t] - dat$x1[1:(t - 1)])
+df$dx2 <- abs(dat$x2[2:t] - dat$x2[1:(t - 1)])
+# prepare data in HMSC format
+Y <- as.matrix(cbind(df$Nt1, df$Nt2))
+XData <- data.frame(cbind(E[1:(t-1)],E[1:(t-1)]^2,abs(dat$x1[2:t] - dat$x1[1:(t-1)]),abs(dat$x2[2:t] - dat$x2[1:(t-1)])))
+colnames(XData) <- c("E","Esq","dx1","dx2")
+studyDesign = data.frame(sample = as.factor(1:(t-1)))
+rL = HmscRandomLevel(units = studyDesign$sample)
+m.7.hmsc = Hmsc(Y = Y, XData = XData, XFormula = ~E+Esq+dx1+dx2, studyDesign = studyDesign, ranLevels = list(sample = rL))
+# Bayesian model parameters
+nChains <- 2
+thin <- 5
+samples <- 2000
+transient <- 1000*thin
+verbose <- 500*thin
+# sample MCMC
+m.7.sample <- sampleMcmc(m.7.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
+```
+
+Previously the population growth rate was fixed as \( r_i = \hat{W}e^{-(E-x_{i,t})^2}\). However, we now introduce the quantitative genetic model of evolutionary rescue [@Gomulkiewicz1995].
+
 # References

### 2024-09-11 manuscript/Appendix_S1_v01.Rmd
spatial not working

--- 2024-09-08 manuscript/Appendix_S1_v01.Rmd
@@
 
+# Species growth, competition, environmental change, and trait evolution in a spatially structured environment
+
+We now consider that trait evolution can occur, with an impact on the species growth rate / fitness in the context of the environment.
+
+A final consideration to add to the model is the existence of multiple sites that may have correlation structure in the variance of species abundances.
+We consider this in our simulation by adding multiple (10) sites, with no vs.
+high spatial covariance.
+We then consider spatial random effects in HMSC, where covariance is considered as a function of distance between sites.
+
+## Population dynamics simulation
+
+We use the same example as above, but now with 10 sites.
+To illustrate how HMSC can estimate the impacts of distance-dependent spatial covariance, we duplicate the population time series obtained in the example above, but we add in site variance and among-site covariance.
+
+```{r}
+# Case 1: Weaker interactions, stronger environment
+# Simulate initial species population growth with environment fluctuations
+N1.0 <- 10
+N2.0 <- 10
+alpha.11 <- 0.01
+alpha.22 <- 0.01
+alpha.12 <- 0.005
+alpha.21 <- 0.01
+E.0 <- 0.8
+x1.0 <- 0.1
+x2.0 <- 0.8
+P <- 1
+w <- 2
+Wmax <- 2
+h2 <- 1
+k <- (w + (1 - h2) * P)/(P + w)
+
+# model function
+disc_LV_evol <- function(N1.0,N2.0,alpha.11,alpha.22,alpha.12,alpha.21,E,x1.0,x2.0,P,w,Wmax,h2) {
+  What <- Wmax*sqrt(w/(P+w))
+  r1 <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E-x1.0))^2)/(2*(P+w)))
+  Nt1 <- (r1*N1.0) / (1+alpha.11*N1.0+alpha.12*N2.0)
+  r2 <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E-x2.0))^2)/(2*(P+w)))
+  Nt2 <- (r2*N2.0) / (1+alpha.22*N2.0+alpha.21*N1.0)
+  return(c(Nt1,Nt2,r1,r2))
+}
+# Simulation of model for t time steps, i sites
+j <- 10
+# random site locations
+xycoords = matrix(runif(2*j), ncol = 2)
+rownames(xycoords) <- 1:j
+t <- 40
+N <- array(NA,dim=c(j*t,2))
+N <- as.data.frame(N)
+colnames(N) <- c("N1","N2")
+x <- array(NA,dim=c(j*t,2))
+x <- as.data.frame(x)
+colnames(x) <- c("x1","x2")
+r <- array(NA,dim=c(j*t,2))
+r <- as.data.frame(r)
+colnames(r) <- c("r1","r2")
+N$N1[1:j] <- N1.0
+N$N2[1:j] <- N2.0
+x$x1[1:j] <- x1.0
+x$x2[1:j] <- x2.0
+What <- Wmax*sqrt(w/(P+w))
+r$r1[1:j] <- What*(exp((-(((w+(1-h2)*P)/(P+w))*(E.0-x1.0))^2)/(2*(P+w))))
+r$r2[1:j] <- What*(exp((-(((w+(1-h2)*P)/(P+w))*(E.0-x2.0))^2)/(2*(P+w))))
+E <- rep(NA, t)
+E[1] <- E.0
+# Keep track of site and time
+study <- array(NA,dim=c(j*t,3))
+study <- as.data.frame(study)
+colnames(study) <- c("obs","site","time")
+study$obs <- 1:dim(study)[1]
+study$site <- rep(1:j,times=t)
+study$time <- rep(1:t,each=j)
+
+for (i in 2:t) {
+  for(z in 1:j){
+    res <- disc_LV_evol(N1.0=N[study$site==z & study$time==i-1,1],N2.0=N[study$site==z & study$time==i-1,2],alpha.11=alpha.11,alpha.22=alpha.22,alpha.12=alpha.12,alpha.21=alpha.21,E=E[i-1],x1=x[study$site==z & study$time==i-1,1],x2=x[study$site==z & study$time==i-1,2],P=P,w=w,Wmax=Wmax,h2=h2)
+    N$N1[study$site==z & study$time==i] <- res[1]
+    N$N2[study$site==z & study$time==i] <- res[2]
+    r$r1[study$site==z & study$time==i] <- res[3]
+    r$r2[study$site==z & study$time==i] <- res[4]
+    # trait change
+    d1 <- E[i-1] - x[study$site==z & study$time==i-1,1]
+    d2 <- E[i-1] - x[study$site==z & study$time==i-1,2]
+    d1.1 <- k*d1
+    d2.1 <- k*d2
+    x$x1[study$site==z & study$time==i] <- E[i-1] - d1.1
+    x$x2[study$site==z & study$time==i] <- E[i-1] - d2.1
+  }
+  # site-level random effect
+  sigma <- 0
+  sigma.spatial <- 2
+  alpha.spatial <- 0.5
+  Sigma = sigma.spatial^2*exp(-as.matrix(dist(xycoords))/alpha.spatial)
+  # draw from covariance matrix
+  a = mvrnorm(mu=rep(0,j), Sigma = Sigma)
+  N$N1[study$time==i] <- N$N1[study$time==i] + a
+  a = mvrnorm(mu=rep(0,j), Sigma = Sigma)
+  N$N2[study$time==i] <- N$N2[study$time==i] + a
+  # environmental change
+  E[i] <- E[i-1] + abs(rnorm(1, 0, 0.05))
+}
+```
+```{r}
+# Plot simulation: ggplot
+N$site <- study$site
+N$time <- study$time
+dat <- melt(N, id.vars=c("site","time"))
+ggplot2::ggplot(dat, aes(time, value, col=variable)) + geom_point() + facet_wrap(~site,nrow=2)
+```
+We now fit a spatial random effect, and add that to the HMSC model.
+
+```{r}
+# Plot simulation: ggplot
+#N[N <0] <- 0
+dat <- as.data.frame(cbind(log(N$N1), log(N$N2), x$x1, x$x2))
+colnames(dat) <- c("N1", "N2", "x1", "x2")
+dat$time <- 1:t
+df <- data.frame(cbind(dat$N1[dat$time !=1], dat$N2[dat$time !=1], dat$x1[dat$time !=1], dat$x2[dat$time !=1]))
+colnames(df) <- c("Nt1", "Nt2", "xt1", "xt2")
+df$dx1 <- abs(dat$x1[dat$time != 1] - dat$x1[dat$time != t])
+df$dx2 <- abs(dat$x2[dat$time != 1] - dat$x2[dat$time != t])
+# prepare data in HMSC format
+Y <- as.matrix(cbind(df$Nt1, df$Nt2))
+XData <- data.frame(cbind(rep(E[1:(t-1)],times=j),rep(E[1:(t-1)],times=j)^2,abs(dat$x1[dat$time != 1] - dat$x1[dat$time != t]),abs(dat$x2[dat$time != 1] - dat$x2[dat$time != t])))
+colnames(XData) <- c("E","Esq","dx1","dx2")
+# Update study design to include spatial random effects
+studyDesign = data.frame(sample = as.factor(1:(j*(t-1))),time=rep(1:(t-1),each=j),site=rep(1:j,times=(t-1)))
+studyDesign$time <- as.factor(studyDesign$time)
+studyDesign$site <- as.factor(studyDesign$site)
+rL.time = HmscRandomLevel(units = studyDesign$time)
+rL.spatial = HmscRandomLevel(sData = xycoords)
+# Fit HMSC model
+m.9.hmsc = Hmsc(Y = Y, XData = XData, XFormula = ~E+Esq+dx1+dx2, studyDesign = studyDesign, ranLevels = list("time" = rL.time,"site"=rL.spatial))
+# Bayesian model parameters
+nChains <- 2
+thin <- 5
+samples <- 2000
+transient <- 1000*thin
+verbose <- 500*thin
+# sample MCMC
+m.9.sample <- sampleMcmc(m.9.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
+```
+
+```{r}
+m9.post.hmsc <- convertToCodaObject(m.9.sample)
+summary(m9.post.hmsc$Beta)
+bayesplot::mcmc_trace(m9.post.hmsc$Beta)
+bayesplot::mcmc_areas(m9.post.hmsc$Beta,area_method = c("equal height"))
+```
+```{r warning=FALSE}
+VP <- computeVariancePartitioning(m.9.sample,group=c(1,1,1,2,3),groupnames=c("Env","Sp1","Sp2"))
+plotVariancePartitioning(m.9.sample,VP,cols=c("white","skyblue","darkgrey","orange","black"),args.legend=list(cex=0.75,bg="transparent"))
+```
+
+Previously the population growth rate was fixed as \( r_i = \hat{W}e^{-(E-x_{i,t})^2}\). However, we now introduce the quantitative genetic model of evolutionary rescue [@Gomulkiewicz1995].
+
 # References

### 2024-09-11 manuscript/Appendix_S1_v01.Rmd
Updated random effects, spatial working

@@
 ## Bayesian linear statistical model: HMSC
-We can fit the data to a linear model using HMSC, with latent variables to capture the species associations. These are implemented by including a random effect at the time level.
+We can fit the data to a linear model using HMSC, with latent variables to capture the species associations.
+These are implemented by including a random effect at the time level.
+For this analysis, we do not expect the impact of Species 1 and 2 that results from competition to vary over time - competition will always have negative impacts for species abundances.
+Our goal is to estimate the overall impact of species covariance across all of the time points.
+We do not estimate these impacts as a function of the lag in time points.
+In other words, we assume that each time point has random variance and that species may covary in their response at each time point.
+We do not consider that species covariance is a function of distance between sampled time points (see @Ovaskainen2020, equation 5.9).
+We instead assume that site loadings $\eta$ are independent among the sampled time points.
 
@@
 studyDesign = data.frame(sample = as.factor(1:(t-1)))
-rL = HmscRandomLevel(units = studyDesign$sample)
-m.3.hmsc = Hmsc(Y = Y, XData = XData, studyDesign = studyDesign, ranLevels = list(sample = rL))
+studyDesign$sample <- as.factor(studyDesign$sample)
+rL.sample = HmscRandomLevel(units = studyDesign$sample)
+m.3.hmsc = Hmsc(Y = Y, XData = XData, studyDesign = studyDesign, ranLevels = list(sample = rL.sample))
 # Bayesian model parameters
@@
 The model fixed effect estimates match well with those from the AR1 and linear model.
-However, the model up to this point includes $N_{t1}$ and $N_{t2}$ as fixed predictors.
-I change that here, as it better matches what can be done in large multi-species models, and the residual associations will better match the effects species have on one another.
 
@@
 studyDesign = data.frame(sample = as.factor(1:(t-1)))
-rL = HmscRandomLevel(units = studyDesign$sample)
-m.4.hmsc = Hmsc(Y = Y, XData = XData, XFormula = ~1, studyDesign = studyDesign, ranLevels = list(sample = rL))
+studyDesign$sample <- as.factor(studyDesign$sample)
+rL.sample = HmscRandomLevel(units = studyDesign$sample)
+m.4.hmsc = Hmsc(Y = Y, XData = XData, XFormula = ~1, studyDesign = studyDesign, ranLevels = list(sample = rL.sample))
 # Bayesian model parameters
@@
 
-The HMSC model correctly shows the negative association between the species (shown here as correlations - JSDM CHapter 7 tells us the intr-specific correlation is always 1).
+The HMSC model correctly shows the negative association between the species (shown here as correlations - JSDM CHapter 7 tells us the intra-specific correlation is always 1).
 
@@
 
-We evaluate this by greatly increasing the effects if interspecific competition (while decreasing the strength of environmental variation).
+We evaluate this by greatly increasing the effects of interspecific competition (while decreasing the strength of environmental variation).
 
@@
 colnames(XData) <- c("E","Esq","dx1","dx2")
-studyDesign = data.frame(sample = as.factor(1:(t-1)))
-rL = HmscRandomLevel(units = studyDesign$sample)
-m.7.hmsc = Hmsc(Y = Y, XData = XData, XFormula = ~E+Esq+dx1+dx2, studyDesign = studyDesign, ranLevels = list(sample = rL))
+# studyDesign = data.frame(sample = as.factor(1:(t-1)))
+# rL = HmscRandomLevel(units = studyDesign$sample)
+time1 <- 1:t
+time1 <- as.data.frame(time1)
+time1 <- subset(time1,time1>1)
+rownames(XData) <- row.names(time1)
+studyDesign = data.frame(sample = as.factor(1:(t-1)),time = row.names(XData))
+studyDesign$time <- as.factor(studyDesign$time)
+studyDesign$sample <- as.factor(studyDesign$sample)
+rL.time = HmscRandomLevel(sData=time1)
+rL.sample = HmscRandomLevel(units = studyDesign$sample)
+m.7.hmsc = Hmsc(Y = Y, XData = XData, XFormula = ~E+Esq+dx1+dx2, studyDesign = studyDesign, ranLevels = list(sample = rL.sample,time=rL.time))
 # Bayesian model parameters
@@
 VP <- computeVariancePartitioning(m.7.sample,group=c(1,1,1,2,3),groupnames=c("Env","Sp1","Sp2"))
-plotVariancePartitioning(m.7.sample,VP,cols=c("white","skyblue","darkgrey","orange"),args.legend=list(cex=0.75,bg="transparent"))
+plotVariancePartitioning(m.7.sample,VP,cols=c("white","skyblue","darkgrey","orange","black"),args.legend=list(cex=0.75,bg="transparent"))
 ```
@@
 colnames(XData) <- c("E","Esq","dx1","dx2")
-studyDesign = data.frame(sample = as.factor(1:(t-1)))
-rL = HmscRandomLevel(units = studyDesign$sample)
-m.8.hmsc = Hmsc(Y = Y, XData = XData, XFormula = ~E+Esq+dx1+dx2, studyDesign = studyDesign, ranLevels = list(sample = rL))
+# studyDesign = data.frame(sample = as.factor(1:(t-1)))
+# rL = HmscRandomLevel(units = studyDesign$sample)
+
+time1 <- 1:t
+time1 <- as.data.frame(time1)
+time1 <- subset(time1,time1>1)
+rownames(XData) <- row.names(time1)
+studyDesign = data.frame(sample = as.factor(1:(t-1)),time = row.names(XData))
+studyDesign$time <- as.factor(studyDesign$time)
+studyDesign$sample <- as.factor(studyDesign$sample)
+rL.time = HmscRandomLevel(sData=time1)
+rL.sample = HmscRandomLevel(units = studyDesign$sample)
+
+m.8.hmsc = Hmsc(Y = Y, XData = XData, XFormula = ~E+Esq+dx1+dx2, studyDesign = studyDesign, ranLevels = list(sample = rL.sample,time=rL.time))
 # Bayesian model parameters
@@
 VP <- computeVariancePartitioning(m.8.sample,group=c(1,1,1,2,3),groupnames=c("Env","Sp1","Sp2"))
-plotVariancePartitioning(m.8.sample,VP,cols=c("white","skyblue","darkgrey","orange"),args.legend=list(cex=0.75,bg="transparent"))
+plotVariancePartitioning(m.8.sample,VP,cols=c("white","skyblue","darkgrey","orange","black"),args.legend=list(cex=0.75,bg="transparent"))
 ```
 Here we can see that trait evolution in Species 1 had a strong positive relationship with the *change* in population size for both species. Trait evolution in Species 2 had much less of an effect.
+
+```{r}
+knitr::knit_exit()
+```
 

### 2024-09-12 manuscript/Appendix_S1_v01.Rmd
Finished spatial, knit

--- 2024-09-11 manuscript/Appendix_S1_v01.Rmd
@@
 studyDesign = data.frame(sample = as.factor(1:(t-1)))
-rL = HmscRandomLevel(units = studyDesign$sample)
-m.3.hmsc = Hmsc(Y = Y, XData = XData, studyDesign = studyDesign, ranLevels = list(sample = rL))
+studyDesign$sample <- as.factor(studyDesign$sample)
+rL.sample = HmscRandomLevel(units = studyDesign$sample)
+m.3.hmsc = Hmsc(Y = Y, XData = XData, XFormula = ~1, studyDesign = studyDesign, ranLevels = list(sample = rL.sample))
 # Bayesian model parameters
@@
 # Bayesian estimates
-summary(m3.post.hmsc$Beta)$statistics[1:3,1:2]
+summary(m3.post.hmsc$Beta)$statistics[1:2,1:2]
 ```
@@
 studyDesign = data.frame(sample = as.factor(1:(t-1)))
-rL = HmscRandomLevel(units = studyDesign$sample)
-m.4.hmsc = Hmsc(Y = Y, XData = XData, XFormula = ~1, studyDesign = studyDesign, ranLevels = list(sample = rL))
+studyDesign$sample <- as.factor(studyDesign$sample)
+rL.sample = HmscRandomLevel(units = studyDesign$sample)
+m.4.hmsc = Hmsc(Y = Y, XData = XData, XFormula = ~1,studyDesign = studyDesign, ranLevels = list(sample = rL.sample))
 # Bayesian model parameters
@@
 OmegaCor = computeAssociations(m.5.sample)
-supportLevel = 0.7
+supportLevel = 0.95
 toPlot = ((OmegaCor[[1]]$support>supportLevel)
@@
 ```
-
-We evaluate this by greatly increasing the effects if interspecific competition (while decreasing the strength of environmental variation).
+We evaluate this by greatly increasing the effects of interspecific competition (while decreasing the strength of environmental variation).
 
@@
 ```
-```{r echo=FALSE}
+```{r}
 # Plot simulation: ggplot
@@
 OmegaCor = computeAssociations(m.6.sample)
-supportLevel = 0.7
+supportLevel = 0.95
 toPlot = ((OmegaCor[[1]]$support>supportLevel)
@@
 ```
-With the reduced impact of the environment, we now see the strong negative associations between the two competing species. We also correctly see no impact of environment for species abundances.
+With the reduced impact of the environment, we now see the strong negative associations between the two competing species.
+We also correctly see a relatively weak impact of environment for species abundances.
 
@@
 rL = HmscRandomLevel(units = studyDesign$sample)
-m.7.hmsc = Hmsc(Y = Y, XData = XData, XFormula = ~E+Esq+dx1+dx2, studyDesign = studyDesign, ranLevels = list(sample = rL))
+m.7.hmsc = Hmsc(Y = Y, XData = XData, XFormula = ~E+Esq+dx1+dx2, studyDesign = studyDesign, ranLevels = list(sample = rL.sample))
 # Bayesian model parameters
@@
 rL = HmscRandomLevel(units = studyDesign$sample)
-m.8.hmsc = Hmsc(Y = Y, XData = XData, XFormula = ~E+Esq+dx1+dx2, studyDesign = studyDesign, ranLevels = list(sample = rL))
+m.8.hmsc = Hmsc(Y = Y, XData = XData, XFormula = ~E+Esq+dx1+dx2, studyDesign = studyDesign, ranLevels = list(sample = rL.sample))
 # Bayesian model parameters
@@
   }
+  # environmental change
+  E[i] <- E[i-1] + abs(rnorm(1, 0, 0.05))
+}
 # site-level random effect
@@
 a = mvrnorm(mu=rep(0,j), Sigma = Sigma)
+for(i in 1:t){
   N$N1[study$time==i] <- N$N1[study$time==i] + a
+}
 a = mvrnorm(mu=rep(0,j), Sigma = Sigma)
+for(i in 1:t){
   N$N2[study$time==i] <- N$N2[study$time==i] + a
-  # environmental change
-  E[i] <- E[i-1] + abs(rnorm(1, 0, 0.05))
 }
@@
 # Plot simulation: ggplot
-#N[N <0] <- 0
+#N[N <0] <- 0.01
 dat <- as.data.frame(cbind(log(N$N1), log(N$N2), x$x1, x$x2))
@@
 # Update study design to include spatial random effects
-studyDesign = data.frame(sample = as.factor(1:(j*(t-1))),time=rep(1:(t-1),each=j),site=rep(1:j,times=(t-1)))
-studyDesign$time <- as.factor(studyDesign$time)
+samp1 <- 1:(j*t)
+samp1 <- as.data.frame(samp1)
+samp1 <- subset(samp1,samp1>10)
+rownames(XData) <- row.names(samp1)
+studyDesign = data.frame(sample=1:(j*(t-1)),site=study$site[study$time > 1])
+studyDesign$sample <- as.factor(studyDesign$sample)
 studyDesign$site <- as.factor(studyDesign$site)
-rL.time = HmscRandomLevel(units = studyDesign$time)
 rL.spatial = HmscRandomLevel(sData = xycoords)  
+rL.sample = HmscRandomLevel(units = studyDesign$sample)
 # Fit HMSC model
-m.9.hmsc = Hmsc(Y = Y, XData = XData, XFormula = ~E+Esq+dx1+dx2, studyDesign = studyDesign, ranLevels = list("time" = rL.time,"site"=rL.spatial))
+m.9.hmsc = Hmsc(Y = Y, XData = XData, XFormula = ~E+Esq+dx1+dx2, studyDesign = studyDesign, ranLevels = list("sample"=rL.sample,"site"=rL.spatial))
 # Bayesian model parameters
@@
 VP <- computeVariancePartitioning(m.9.sample,group=c(1,1,1,2,3),groupnames=c("Env","Sp1","Sp2"))
-plotVariancePartitioning(m.9.sample,VP,cols=c("white","skyblue","darkgrey","orange","black"),args.legend=list(cex=0.75,bg="transparent"))
-```
-
-Previously the population growth rate was fixed as \( r_i = \hat{W}e^{-(E-x_{i,t})^2}\). However, we now introduce the quantitative genetic model of evolutionary rescue [@Gomulkiewicz1995].
+plotVariancePartitioning(m.9.sample,VP,cols=c("white","skyblue","darkgrey","orange","pink"),args.legend=list(cex=0.75,bg="transparent"))
+```
+
+```{r}
+OmegaCor = computeAssociations(m.9.sample)
+supportLevel = 0.95
+toPlot = ((OmegaCor[[1]]$support>supportLevel)
++ (OmegaCor[[1]]$support<(1-supportLevel))>0)*OmegaCor[[1]]$mean
+corrplot::corrplot(toPlot, method = "color",col = colorRampPalette(c("blue","white","red"))(200),title = paste("random effect level:", m.9.sample$rLNames[1]), mar=c(0,0,1,0))
+OmegaCor
+toPlot = ((OmegaCor[[2]]$support>supportLevel)
++ (OmegaCor[[2]]$support<(1-supportLevel))>0)*OmegaCor[[2]]$mean
+corrplot::corrplot(toPlot, method = "color",col = colorRampPalette(c("blue","white","red"))(200),title = paste("random effect level:", m.9.sample$rLNames[2]), mar=c(0,0,1,0))
+```
+We can see that the model considers spatial random effects by removing the spatial covariance, and only including spatial variance.
+
+```{r}
+# Case 1: Weaker interactions, stronger environment
+# Simulate initial species population growth with environment fluctuations
+N1.0 <- 10
+N2.0 <- 10
+alpha.11 <- 0.01
+alpha.22 <- 0.01
+alpha.12 <- 0.005
+alpha.21 <- 0.01
+E.0 <- 0.8
+x1.0 <- 0.1
+x2.0 <- 0.8
+P <- 1
+w <- 2
+Wmax <- 2
+h2 <- 1
+k <- (w + (1 - h2) * P)/(P + w)
+
+# model function
+disc_LV_evol <- function(N1.0,N2.0,alpha.11,alpha.22,alpha.12,alpha.21,E,x1.0,x2.0,P,w,Wmax,h2) {
+  What <- Wmax*sqrt(w/(P+w))
+  r1 <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E-x1.0))^2)/(2*(P+w)))
+  Nt1 <- (r1*N1.0) / (1+alpha.11*N1.0+alpha.12*N2.0)
+  r2 <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E-x2.0))^2)/(2*(P+w)))
+  Nt2 <- (r2*N2.0) / (1+alpha.22*N2.0+alpha.21*N1.0)
+  return(c(Nt1,Nt2,r1,r2))
+}
+# Simulation of model for t time steps, i sites
+j <- 10
+# random site locations
+xycoords = matrix(runif(2*j), ncol = 2)
+rownames(xycoords) <- 1:j
+t <- 40
+N <- array(NA,dim=c(j*t,2))
+N <- as.data.frame(N)
+colnames(N) <- c("N1","N2")
+x <- array(NA,dim=c(j*t,2))
+x <- as.data.frame(x)
+colnames(x) <- c("x1","x2")
+r <- array(NA,dim=c(j*t,2))
+r <- as.data.frame(r)
+colnames(r) <- c("r1","r2")
+N$N1[1:j] <- N1.0
+N$N2[1:j] <- N2.0
+x$x1[1:j] <- x1.0
+x$x2[1:j] <- x2.0
+What <- Wmax*sqrt(w/(P+w))
+r$r1[1:j] <- What*(exp((-(((w+(1-h2)*P)/(P+w))*(E.0-x1.0))^2)/(2*(P+w))))
+r$r2[1:j] <- What*(exp((-(((w+(1-h2)*P)/(P+w))*(E.0-x2.0))^2)/(2*(P+w))))
+E <- rep(NA, t)
+E[1] <- E.0
+# Keep track of site and time
+study <- array(NA,dim=c(j*t,3))
+study <- as.data.frame(study)
+colnames(study) <- c("obs","site","time")
+study$obs <- 1:dim(study)[1]
+study$site <- rep(1:j,times=t)
+study$time <- rep(1:t,each=j)
+
+for (i in 2:t) {
+  for(z in 1:j){
+    res <- disc_LV_evol(N1.0=N[study$site==z & study$time==i-1,1],N2.0=N[study$site==z & study$time==i-1,2],alpha.11=alpha.11,alpha.22=alpha.22,alpha.12=alpha.12,alpha.21=alpha.21,E=E[i-1],x1=x[study$site==z & study$time==i-1,1],x2=x[study$site==z & study$time==i-1,2],P=P,w=w,Wmax=Wmax,h2=h2)
+    N$N1[study$site==z & study$time==i] <- res[1]
+    N$N2[study$site==z & study$time==i] <- res[2]
+    r$r1[study$site==z & study$time==i] <- res[3]
+    r$r2[study$site==z & study$time==i] <- res[4]
+    # trait change
+    d1 <- E[i-1] - x[study$site==z & study$time==i-1,1]
+    d2 <- E[i-1] - x[study$site==z & study$time==i-1,2]
+    d1.1 <- k*d1
+    d2.1 <- k*d2
+    x$x1[study$site==z & study$time==i] <- E[i-1] - d1.1
+    x$x2[study$site==z & study$time==i] <- E[i-1] - d2.1
+  }
+  # environmental change
+  E[i] <- E[i-1] + abs(rnorm(1, 0, 0.05))
+}
+# site-level random effect
+sigma <- 0
+sigma.spatial <- 10
+alpha.spatial <- 0.5
+Sigma = sigma.spatial^2*exp(-as.matrix(dist(xycoords))/alpha.spatial)
+# draw from covariance matrix
+a = mvrnorm(mu=rep(0,j), Sigma = Sigma)
+for(i in 1:t){
+  N$N1[study$time==i] <- N$N1[study$time==i] + a
+}
+a = mvrnorm(mu=rep(0,j), Sigma = Sigma)
+for(i in 1:t){
+  N$N2[study$time==i] <- N$N2[study$time==i] + a
+}
+```
+```{r}
+# Plot simulation: ggplot
+N$site <- study$site
+N$time <- study$time
+dat <- melt(N, id.vars=c("site","time"))
+ggplot2::ggplot(dat, aes(time, value, col=variable)) + geom_point() + facet_wrap(~site,nrow=2)
+```
+```{r}
+# Plot simulation: ggplot
+N[N <0] <- 0.01
+dat <- as.data.frame(cbind(log(N$N1), log(N$N2), x$x1, x$x2))
+colnames(dat) <- c("N1", "N2", "x1", "x2")
+dat$time <- 1:t
+df <- data.frame(cbind(dat$N1[dat$time !=1], dat$N2[dat$time !=1], dat$x1[dat$time !=1], dat$x2[dat$time !=1]))
+colnames(df) <- c("Nt1", "Nt2", "xt1", "xt2")
+df$dx1 <- abs(dat$x1[dat$time != 1] - dat$x1[dat$time != t])
+df$dx2 <- abs(dat$x2[dat$time != 1] - dat$x2[dat$time != t])
+# prepare data in HMSC format
+Y <- as.matrix(cbind(df$Nt1, df$Nt2))
+XData <- data.frame(cbind(rep(E[1:(t-1)],times=j),rep(E[1:(t-1)],times=j)^2,abs(dat$x1[dat$time != 1] - dat$x1[dat$time != t]),abs(dat$x2[dat$time != 1] - dat$x2[dat$time != t])))
+colnames(XData) <- c("E","Esq","dx1","dx2")
+# Update study design to include spatial random effects
+samp1 <- 1:(j*t)
+samp1 <- as.data.frame(samp1)
+samp1 <- subset(samp1,samp1>10)
+rownames(XData) <- row.names(samp1)
+studyDesign = data.frame(sample=1:(j*(t-1)),site=study$site[study$time > 1])
+studyDesign$sample <- as.factor(studyDesign$sample)
+studyDesign$site <- as.factor(studyDesign$site)
+rL.spatial = HmscRandomLevel(sData = xycoords)  
+rL.sample = HmscRandomLevel(units = studyDesign$sample)
+# Fit HMSC model
+m.10.hmsc = Hmsc(Y = Y, XData = XData, XFormula = ~E+Esq+dx1+dx2, studyDesign = studyDesign, ranLevels = list("sample"=rL.sample,"site"=rL.spatial))
+# Bayesian model parameters
+nChains <- 2
+thin <- 5
+samples <- 2000
+transient <- 1000*thin
+verbose <- 500*thin
+# sample MCMC
+m.10.sample <- sampleMcmc(m.10.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
+```
+```{r}
+m10.post.hmsc <- convertToCodaObject(m.10.sample)
+summary(m10.post.hmsc$Beta)
+bayesplot::mcmc_trace(m10.post.hmsc$Beta)
+bayesplot::mcmc_areas(m10.post.hmsc$Beta,area_method = c("equal height"))
+```
+```{r warning=FALSE}
+VP <- computeVariancePartitioning(m.10.sample,group=c(1,1,1,2,3),groupnames=c("Env","Sp1","Sp2"))
+plotVariancePartitioning(m.10.sample,VP,cols=c("white","skyblue","darkgrey","orange","pink"),args.legend=list(cex=0.75,bg="transparent"))
+```
+
+```{r}
+OmegaCor = computeAssociations(m.9.sample)
+supportLevel = 0.95
+toPlot = ((OmegaCor[[1]]$support>supportLevel)
++ (OmegaCor[[1]]$support<(1-supportLevel))>0)*OmegaCor[[1]]$mean
+corrplot::corrplot(toPlot, method = "color",col = colorRampPalette(c("blue","white","red"))(200),title = paste("random effect level:", m.9.sample$rLNames[1]), mar=c(0,0,1,0))
+OmegaCor
+toPlot = ((OmegaCor[[2]]$support>supportLevel)
++ (OmegaCor[[2]]$support<(1-supportLevel))>0)*OmegaCor[[2]]$mean
+corrplot::corrplot(toPlot, method = "color",col = colorRampPalette(c("blue","white","red"))(200),title = paste("random effect level:", m.9.sample$rLNames[2]), mar=c(0,0,1,0))
+```
 

### 2024-09-13 manuscript/Appendix_S1_v01.Rmd
Finished AppS1 all scenarios, knit

@@
 
+The HMSC model correctly shows the negative association between the species (shown here as correlations - JSDM CHapter 7 tells us the intra-specific correlation is always 1).
+
+```{r}
+preds = computePredictedValues(m.3.sample)
+df <- data.frame(cbind(log(dat$N1[2:t]),log(dat$N2[2:t]),log(dat$N1[1:(t-1)]),log(dat$N2[1:(t-1)])))
+colnames(df) <- c("Nt1","Nt2","N1","N2")
+df$n1_25 <- NA
+df$n1_975 <- NA
+df$n2_25 <- NA
+df$n2_975 <- NA
+for(i in 1:39){
+  red <- preds[i,,]
+  a <- apply(red,1,quantile,probs=c(.025,.975))
+  df$n1_25[i] <- a[1,1]
+  df$n1_975[i] <- a[2,1]
+  df$n2_25[i] <- a[1,2]
+  df$n2_975[i] <- a[2,2]
+}
+plot(df$Nt1,pch=19,col="black",ylim=c(0,5))
+points(df$Nt2,pch=19,col="orange")
+lines(df$n1_25,col="black")
+lines(df$n1_975,col="black")
+lines(df$n2_25,col="orange")
+lines(df$n2_975,col="orange")
+```
+
 I try again with a next example where the only difference between species is the $\alpha_{ij}$.
@@
 plotVariancePartitioning(m.6.sample,VP,args.legend=list(cex=0.75,bg="transparent"))
+```
+```{r}
+preds = computePredictedValues(m.6.sample)
+df$n1_25 <- NA
+df$n1_975 <- NA
+df$n2_25 <- NA
+df$n2_975 <- NA
+for(i in 1:39){
+  red <- preds[i,,]
+  a <- apply(red,1,quantile,probs=c(.025,.975))
+  df$n1_25[i] <- a[1,1]
+  df$n1_975[i] <- a[2,1]
+  df$n2_25[i] <- a[1,2]
+  df$n2_975[i] <- a[2,2]
+}
+plot(df$Nt1,pch=19,col="black",ylim=c(0,5))
+points(df$Nt2,pch=19,col="orange")
+lines(df$n1_25,col="black")
+lines(df$n1_975,col="black")
+lines(df$n2_25,col="orange")
+lines(df$n2_975,col="orange")
 ```

### 2024-10-01 manuscript/Appendix_S3_v01.Rmd
added sim, arranged folders, began metacom scenario

--- 2024-09-08 manuscript/Appendix_S2_v01.Rmd
@@
 ---
-title: "Appendix S2. Evolutionary drivers of species and community dynamics"
+title: "Appendix S3. Hypotheses and tests for increased importance of evolutionary drivers for species and community dynamics"
 author:
@@
 
-In Appendix S1.5, we evaluated the ability of the statistical model to estimate the impact of trait evolution for species abundances.
-Trait evolution's impact for species abundances can be modeled as a predictor in different ways.
-Here we derive the numerical form of the trait $x_i$ that we hypothesize has the most impact on species abundances.
+First I establish a metacommunity simulation scenario where environment, spatial random effects, and year-based Random Effects (species interactions) all play a role in driving species abundances.
+I then add in trait evolution, and look at its relative importance.
 

### 2024-10-01 code/R/metacom_evol_sim_same_init.R

@@ -1,0 +1,294 @@
+## Evolving metacommunity simulation
+## J.H. Pantel & Ruben Hermann
+## June 26 2024
+## Scenario 1. Weak selection, fixed environment, all dispersal and heritability levels
+## # P = 1, w = 10, v[E] = 0
+  
+#set.seed(1)
+
+###################################
+## Step 0. Small functions    ##
+###################################
+# I needed to write some small function, which lets me run the dispersal lines smoother
+# Here I check if the drawn runif number is smaller then the likelihood of dispersal
+# if so then one individual with be migrating
+# x = rlist (list of randomly generated numbers (runif) for each individual in a patch of species i)
+# i = species number (is part of the loop for dispersal)
+check <- function(x){
+    x <=d[i]
+}
+
+# This function is the randome sample to chose amone sites where the species can immigrate,
+# weighted by connectivitx.
+# It takes simply a vector of all patches in, so:
+# x = site/1:50 and i = species number (is part of the loop for dispersal)
+disp_patch <- function(x){
+    sample(x=site[-x],size=mig_count[x],replace=T,prob=connec[i,-x,x])
+}
+#Function to update the trait value, where the environmental value is substracted
+#by the distance of the trait (which is changed by evo)
+#x = The number of the species
+#xt_up = the updated trait value, which is saved into xt
+up_trait <- function(x){
+  xt_up<-E-dt[,x,t]
+}
+###################################
+## Step 1. Initialize species    ##
+###################################
+patch = 30   # Number of patches in the landscape
+spec = 10    # Number of species in the community
+time = 1000  # Number of time steps to run the simulation
+N = array(numeric(),c(patch,spec,time))   # Record of abundances of all species over time patch x spec x time
+P = array(1,c(1,spec))    # Phenotypic variances for all species
+w = array(10,c(1,spec))   # Width of fitness curve
+Wmax = 2    # Maximum population growth rate for all species
+Wt = array(numeric(),c(patch,spec,time))   # Record of fitness vlaues over time, patch x spec x time
[…]

### 2024-10-21 code/R/fig1.R
4 panel plot, aligned in 1 row (not clean)

@@ -9,6 +9,7 @@ library(ggplot2)
 library(patchwork)
 library(bayesplot)
 library(reshape2)
+library(cowplot)
 
 ## set random seed
 set.seed(42)
@@ -20,8 +21,8 @@ N0 <- c(10,10)
 r <- c(1.7,1.7)
 alpha.11 <- 0.01
 alpha.22 <- 0.01
-alpha.12 <- 0.005
-alpha.21 <- 0.01
+alpha.12 <- 0.015
+alpha.21 <- 0.02
 alpha <- matrix(c(alpha.11,alpha.21,alpha.12,alpha.22),nrow=2,byrow=FALSE) # careful to make sure you have correct interaction matrix
 E.0 <- 0.8
 x <- c(0.6,0.8)
@@ -60,8 +61,6 @@ m.1.sample <- sampleMcmc(m.1.hmsc,thin=thin,sample=samples,transient=transient,n
 pred <- predict(m.1.sample)
 hold1 <- list()
 hold2 <- list()
-#y <- array(NA,length(r)*(t-1))
-#yrep <- array(NA,dim=c(length(pred),length(r)*(t-1)))
 for(i in 1:length(r)){
   hold1[[i]] <- as.numeric(Y[,i])
   data <- lapply(pred, function(x) x[, colnames(df)[i], drop = FALSE])
@@ -70,12 +69,33 @@ for(i in 1:length(r)){
 y <- unlist(hold1) # vector of each species' observed N across time points
 yrep <- as.matrix(data.frame(hold2))
 ### Mod1 plot ###
+par(mgp=c(2.2,0.45,0), tcl=-0.4, mar=c(3.3,3.6,1.1,1.1))
 color_scheme_set("brightblue")
 #ppc_dens_overlay(y, yrep[1:50, ])
-ppc_intervals(y,yrep,x=rep(dat$time[1:(t-1)],times=2),prob = 0.50) +
-  labs(x = "time",y = "N",)
-
-
-
-
+a <- ppc_intervals(y,yrep,x=rep(dat$time[1:(t-1)],times=2),prob = 0.95) +
+  labs(x = "time",y = "ln(N)",) +
+  theme(plot.margin = unit(c(0, 0, 0, 0), "cm"))
+### Gradient plot ###
+Gradient <- constructGradient(m.1.sample,focalVariable="E",non.focalVariables=list(Esq=list(2)),ngrid=39)
+Gradient$XDataNew$Esq <- Gradient$XDataNew$E^2
[…]

### 2024-10-21 code/R/fig1.R
Started part 2

@@ -12,17 +12,17 @@ library(reshape2)
 library(cowplot)
 
 ## set random seed
-set.seed(42)
+set.seed(8)
 
-# 1a. Sim: Two species, logistic growth + competition, environmental covariate ---------
+# 1. Sim: Two species, logistic growth + competition, environmental covariate ---------
 ### Sim1. Two species, Leslie-Gower competition, environmental covariate
 # Initial conditions
 N0 <- c(10,10)
 r <- c(1.7,1.7)
 alpha.11 <- 0.01
 alpha.22 <- 0.01
-alpha.12 <- 0.015
-alpha.21 <- 0.02
+alpha.12 <- 0.005
+alpha.21 <- 0.01
 alpha <- matrix(c(alpha.11,alpha.21,alpha.12,alpha.22),nrow=2,byrow=FALSE) # careful to make sure you have correct interaction matrix
 E.0 <- 0.8
 x <- c(0.6,0.8)
@@ -36,8 +36,13 @@ E <- rep(NA, t)
 E[1] <- E.0
 for (i in 2:t) {
   N[i,] <- disc_LV_E(r=r,N0=N[i-1,],alpha=alpha,E=E[i-1],x=x)
-  E[i] <- E[i-1] + rnorm(1,0,.1)
+  E[i] <- E[i-1] + rnorm(1,0,0.1)
 }
+
+N$time <- 1:t
+dat <- melt(N, id.vars="time")
+ggplot2::ggplot(dat, aes(time, value, col=variable)) + geom_point() + geom_hline(yintercept = ((r[1] - 1)/alpha.11),linetype = "dashed", color = "gray")
+
 ### Mod1: HMSC model fit 1 ###
 # prepare data in HMSC format
 dat <- as.data.frame(log(N))
@@ -69,33 +74,56 @@ for(i in 1:length(r)){
 y <- unlist(hold1) # vector of each species' observed N across time points
 yrep <- as.matrix(data.frame(hold2))
 ### Mod1 plot ###
-par(mgp=c(2.2,0.45,0), tcl=-0.4, mar=c(3.3,3.6,1.1,1.1))
+par(mgp=c(2.2,0.45,0), tcl=-0.4, mar=c(2,2,1,1))
 color_scheme_set("brightblue")
 #ppc_dens_overlay(y, yrep[1:50, ])
 a <- ppc_intervals(y,yrep,x=rep(dat$time[1:(t-1)],times=2),prob = 0.95) +
   labs(x = "time",y = "ln(N)",) +
[…]

### 2024-10-21 code/R/fig1.R
Finished Sim2, began Sim3 spatial

@@ -116,14 +116,119 @@ alpha.12 <- 0.005
 alpha.21 <- 0.01
 alpha <- matrix(c(alpha.11,alpha.21,alpha.12,alpha.22),nrow=2,byrow=FALSE) # careful to make sure you have correct interaction matrix
 E.0 <- 0.8
-x <- c(0.1,0.8)
+x.0 <- c(0.1,0.8)
 P <- 1
 w <- 2
 Wmax <- 2
 h2 <- 1
 k <- (w + (1 - h2) * P)/(P + w)
+# Simulation of model for t time steps
+# Simulation of model for t time steps
+t <- 40
+N <- array(NA,dim=c(t,length(r)))
+N <- as.data.frame(N)
+colnames(N) <- paste0("N",1:length(r))
+N[1,] <- N0
+E <- rep(NA, t)
+E[1] <- E.0
+x <- array(NA,dim=c(t,length(r)))
+x <- as.data.frame(x)
+colnames(x) <- paste0("x",1:length(r))
+x[1,] <- x.0
+r <- array(NA,dim=c(t,length(r)))
+r <- as.data.frame(r)
+colnames(r) <- paste0("r",1:length(r))
+What <- Wmax*sqrt(w/(P+w))
+r[1,] <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E[1]-x[1,]))^2)/(2*(P+w)))
+for (i in 2:t) {
+  res <- disc_LV_evol(N0=N[i-1,],alpha=alpha,E=E[i-1],x=x[i-1,],P=P,w=w,Wmax=Wmax,h2=h2)
+  N[i,] <- res$Nt1
+  r[i,] <- res$r
+  # trait change
+  d <- E[i-1] - x[i-1,]
+  d1 <- k*d
+  x[i,] <- E[i-1] - d1
+  # environmental change
+  E[i] <- E[i-1] + abs(rnorm(1, 0, 0.05))
+}
+
+# Plot simulation: ggplot
+N$time <- 1:t
+dat <- melt(N, id.vars="time")
+ggplot2::ggplot(dat, aes(time, value, col=variable)) + geom_point()
+
+### Mod2: HMSC model fit 2 ###
+# prepare data in HMSC format
+dat <- as.data.frame(cbind(log(N[,1:2]),x))
+dat$time <- 1:t
+df <- data.frame(dat[(2:t),-5])
[…]

### 2024-10-21 code/R/fig1.R
Finished Sim3, crude plots

@@ -10,6 +10,7 @@ library(patchwork)
 library(bayesplot)
 library(reshape2)
 library(cowplot)
+library(MASS)
 
 ## set random seed
 set.seed(8)
@@ -179,9 +180,6 @@ transient <- 1000*thin
 verbose <- 500*thin
 # sample MCMC
 m.2.sample <- sampleMcmc(m.2.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
-
-
-
 # prepare for plot
 pred2 <- predict(m.2.sample)
 hold1 <- list()
@@ -193,7 +191,7 @@ for(i in 1:length(r)){
 }
 y <- unlist(hold1) # vector of each species' observed N across time points
 yrep <- as.matrix(data.frame(hold2))
-### Mod1 plot ###
+### Mod2 plot ###
 par(mgp=c(2.2,0.45,0), tcl=-0.4, mar=c(2,2,1,1))
 color_scheme_set("brightblue")
 #ppc_dens_overlay(y, yrep[1:50, ])
@@ -229,6 +227,153 @@ plot_grid(a,plot_grid(b,c,nrow=2),d,e,nrow=1,rel_widths=c(2,1,1,1))
 # 3. Sim: Two species, logistic growth + competition, environmental covariate in spatially structured environment ---------
 ### Sim3. Two species, Leslie-Gower competition, environmental covariate, evolution, multiple sites
 # Initial conditions
+# Simulation of model for t time steps, i sites
+j <- 10
+# random site locations
+xycoords = matrix(runif(2*j), ncol = 2)
+rownames(xycoords) <- 1:j
+t <- 40
+N <- array(NA,dim=c(j*t,length(r0)))
+N <- as.data.frame(N)
+colnames(N) <- paste0("N",1:length(r0))
+x <- array(NA,dim=c(j*t,length(r0)))
+x <- as.data.frame(x)
+colnames(x) <- paste0("x",1:length(r0))
+r <- array(NA,dim=c(j*t,length(r0)))
+r <- as.data.frame(r)
+colnames(r) <- paste0("r",1:length(r0))
+N[1:j,] <- rep(N0,each=j)
+E <- rep(NA, t)
+E[1] <- E.0
+x[1:j,] <- rep(x.0,each=j)
+What <- Wmax*sqrt(w/(P+w))
[…]

### 2024-10-22 code/R/1_metacom_sim_v01.R
Addition of fluctuating E

@@ -152,12 +152,17 @@ h2 = array(h2_vals[2],c(1,spec)) #which h2_value do you want to use? - just usin
 k = ((w+((1-h2)*P))/(w+P))
 d = array(d_vals[8],c(1,spec))  #which d_value do you want to use? - just using 3 to test
 
+load("./data/h_01_d_minus3_res.RData")
+
 ###################################################
 ## Step 3. Evolving metacommunity simulations    ##
 ###################################################
 
-for (t in 2:time) {
+for (t in 2:200) {
     t
+    #E fluctuation
+    if (t%%10==0) {E = E + abs(rnorm(50,0,0.05))}
+  
     pop = N[,,t] # Write this generation's population size into a temporary record 
     pop[N[,,t-1]==0] = 0 # Carry over all extinct species
 
@@ -289,4 +294,4 @@ z <- 2
 f <- 8
 print(paste(h2_names[z],d_names[f],sep=""))
 result <- paste(h_lev[z],d_lev[f],sep="")
-save.image(paste("././data/",result,"_res.RData",sep=""))
+save.image(paste("././data/",result,"_res_short.RData",sep=""))

### 2024-10-22 code/R/2_results_sim.R

@@ -11,7 +11,7 @@ library(patchwork)
 
 ##
 ####  Step 1. Read in the data and get it into shape --------------
-load("././data/h_01_d_minus3_res.RData") # Scenario 1
+load("././data/h_01_d_minus3_res_short.RData") # Scenario 1
 ##
 ################### Extinction checks from JP ###################
 # Address extinct species

### 2024-10-22 code/R/1_metacom_sim_v01.R
Additions for variation in environmental changes

@@ -45,7 +45,9 @@ Wmax = 2    # Maximum population growth rate for all species
 Wt = array(numeric(),c(patch,spec,time))   # Record of fitness vlaues over time, patch x spec x time
 dt = array(numeric(),c(patch,spec,time))   # Record of distance from optimum value over time, patch x spec x time
 xt = array(9999,c(patch,spec,time))     # Record of phenotype value over time, patch x spec x time
+Et = array(numeric(),c(patch,time))   # Record of environmental value over time, patch x spec x time
 
+var = 0.1 #Adding variation to the environmental value for no variation set to 0 (var is the sd of the normal distribution with a mean of 0)
 ## Heritability
 # Scenario 1: No evolution
 #h2 = array(0,c(1,spec))
@@ -87,6 +89,7 @@ for (i in 1:spec) {
 
 ## Environmental properties
 E = array(runif(patch),c(patch,1))    # Each patch has an environmental optimum value E. d, distance between species phenotype and patch optimum, will equal some phenotype value of the species x minus this E quantity. So d = x - E
+Et[,1] = E
 
 ## Initial occupancy
 # Random initial occupancy
@@ -152,7 +155,7 @@ h2 = array(h2_vals[2],c(1,spec)) #which h2_value do you want to use? - just usin
 k = ((w+((1-h2)*P))/(w+P))
 d = array(d_vals[8],c(1,spec))  #which d_value do you want to use? - just using 3 to test
 
-load("./data/h_01_d_minus3_res.RData")
+#load("./data/h_01_d_minus3_res.RData")
 
 ###################################################
 ## Step 3. Evolving metacommunity simulations    ##
@@ -160,9 +163,10 @@ load("./data/h_01_d_minus3_res.RData")
 
 for (t in 2:200) {
     t
-    #E fluctuation
-    if (t%%10==0) {E = E + abs(rnorm(50,0,0.05))}
-  
+    #E fluctuation, here with directional change (with the abs() function)
+    if (t%%10==0) {E = E + abs(rnorm(50,0,var))}
+    Et[,t] = E
+    
     pop = N[,,t] # Write this generation's population size into a temporary record 
     pop[N[,,t-1]==0] = 0 # Carry over all extinct species
 
@@ -288,10 +292,17 @@ for (t in 2:200) {
[…]

### 2024-10-22 code/R/2_results_sim.R

@@ -11,7 +11,7 @@ library(patchwork)
 
 ##
 ####  Step 1. Read in the data and get it into shape --------------
-load("././data/h_01_d_minus3_res_short.RData") # Scenario 1
+load("././data/h_01_d_minus3_res_short_Edirect_sd0.1.RData") # Scenario 1
 ##
 ################### Extinction checks from JP ###################
 # Address extinct species
@@ -32,7 +32,7 @@ for(i in 3:dim(xt)[3]){
 ## fitness of species: -abs(abs(E)-x) --> so that negative values represent low fitness
 fit_x <-  array(NA,c(dim(xt)[1],dim(xt)[2],dim(xt)[3]),dimnames=list(NULL,c(paste("y",1:dim(xt)[2],"",sep="")),NULL))
 for(i in 1:dim(xt)[3]){
-  fit_x[,,i] <- -abs(sweep(abs(xt[,,i]),1,E,"-"))
+  fit_x[,,i] <- -abs(sweep(abs(xt[,,i]),1,Et[,i],"-"))
 }
 
 ## average fitness of patch: -abs(abs(E)-x) --> so that negative values represent low fitness
@@ -128,7 +128,7 @@ mtext("Population dynamics in each plot over time", side = 3, line = 0, outer = 
 par(mfrow=(c(10,5)))
 par(mai=c(0.1,0.1,0.1,0.1),oma=c(2,2,2,1))
 for(j in 1:nrow(xt)){
-  plot(xt[j,1,],col=mycols2[1],ylim=c(min(xt,na.rm=T),1),axes=FALSE,type="n",ylab=NA,xlab=NA,xaxs="i",yaxs="i")
+  plot(xt[j,1,],col=mycols2[1],ylim=c(min(xt,na.rm=T),max(xt,na.rm=T)),axes=FALSE,type="n",ylab=NA,xlab=NA,xaxs="i",yaxs="i")
   if (j %in% c(1,6,11,16,21,26,31,36,41,46))
     axis(2,col="grey40",col.axis="grey20",labels=T)
   if (j %in% c(46,47,48,49,50))
@@ -190,6 +190,20 @@ for(j in 1:nrow(sp_rich)){
 }
 mtext("Species richness in each patch over time", side = 3, line = 0, outer = TRUE)
 
+## Plot of environmental value for given patch over time
[…]

### 2024-10-22 code/R/6_emp_hmsc.R
Addition of the hmsc script for the empirical data, without any diagnostic and result script so far

@@ -1,0 +1,138 @@
+# Title. emp_hmsc_v01.R
+# Author. R.J. Hermann
+# Date 24-08-2023
+# Description. This script contains analyses for enmpircial data of the HMSC_ecoevo project and associated manuscript Hermann & Pantel 202x.
+
+#### Libraries ---------------------------------------------------------------
+library(tidyverse)
+library(Hmsc)
+library(tictoc)
+
+#### Step 1. Loading the empirical data from the publication Jewell & Bell 2023 --------------
+dat_ab <- read.csv("empirical data/Phase1_abundance.csv")
+dat_ph <- read.csv("empirical data/Phase1_phenotypes.csv")
+
+#Deleting abundance data of Week 14 and Week 0, because they did not measure the FrondArea on that Weeks
+dat_ab <- subset(dat_ab,Week<14)
+dat_ab <- subset(dat_ab,Week>0)
+
+#Changing the H(for high), M (for moderate) and L (for low) for the actual values of the experiment
+#C.Light is the amount of layers of 50% shade cloth (described as percentage in the publication)
+dat_ph$C.Light[dat_ph$C.Light=="L"]<- 3
+dat_ph$C.Light[dat_ph$C.Light=="M"]<- 12
+dat_ph$C.Light[dat_ph$C.Light=="H"]<- 50
+
+dat_ab$C.Light[dat_ab$C.Light=="L"]<- 3
+dat_ab$C.Light[dat_ab$C.Light=="M"]<- 12
+dat_ab$C.Light[dat_ab$C.Light=="H"]<- 50
+
+#C.Nutrients is the concentrations of added nutrients, we use her the µg/L concentrations for dissolved phosphor.
+dat_ph$C.Nutrients[dat_ph$C.Nutrients=="L"]<- 10
+dat_ph$C.Nutrients[dat_ph$C.Nutrients=="M"]<- 40
+dat_ph$C.Nutrients[dat_ph$C.Nutrients=="H"]<- 160
+
+dat_ab$C.Nutrients[dat_ab$C.Nutrients=="L"]<- 10
+dat_ab$C.Nutrients[dat_ab$C.Nutrients=="M"]<- 40
+dat_ab$C.Nutrients[dat_ab$C.Nutrients=="H"]<- 160
+
+#Setting them to numeric, as they were characters before
+dat_ab$C.Light <- as.numeric(dat_ab$C.Light)
[…]

### 2024-10-22 code/R/fig1.R
Finished Mod3, not clean

@@ -78,7 +78,7 @@ yrep <- as.matrix(data.frame(hold2))
 par(mgp=c(2.2,0.45,0), tcl=-0.4, mar=c(2,2,1,1))
 color_scheme_set("brightblue")
 #ppc_dens_overlay(y, yrep[1:50, ])
-a <- ppc_intervals(y,yrep,x=rep(dat$time[1:(t-1)],times=2),prob = 0.95) +
+p1 <- ppc_intervals(y,yrep,x=rep(dat$time[1:(t-1)],times=2),prob = 0.95) +
   labs(x = "time",y = "ln(N)",) +
   theme(plot.margin = unit(c(0, 0, 0, 0), "cm")) +
   theme(legend.position = "none")
@@ -87,25 +87,25 @@ Gradient <- constructGradient(m.1.sample,focalVariable="E",non.focalVariables=li
 Gradient$XDataNew$Esq <- Gradient$XDataNew$E^2
 predY <- predict(m.1.sample,XData=Gradient$XDataNew,expected=TRUE)
 plotGradient(m.1.sample,Gradient,pred=predY,showData=T,measure="Y",index=1,main="",xlab="E_t",ylab="predicted N1_t+1",showPosteriorSupport=FALSE,cex.axis=0.75)
-b <- recordPlot()
+p2 <- recordPlot()
 plotGradient(m.1.sample,Gradient,pred=predY,showData=T,measure="Y",index=2,main="",xlab="E_t",ylab="predicted N2_t+1",showPosteriorSupport=FALSE,cex.axis=0.75)
-c <- recordPlot()
+p3 <- recordPlot()
 ### Beta posterior plot ###
 m.post = Hmsc::convertToCodaObject(m.1.sample)
 m.beta <- as.data.frame(rbind(m.post$Beta[[1]],m.post$Beta[[2]]))
 colnames(m.beta) <- c("Int1","E1","Esq1","Int2","E2","Esq2")
-d <- bayesplot::mcmc_areas(m.beta) + scale_x_continuous(limits=c(-20,30))
+p4 <- bayesplot::mcmc_areas(m.beta) + scale_x_continuous(limits=c(-20,30))
 #postBeta = getPostEstimate(m.1.sample, parName = "Beta")
 #plotBeta(m.1.sample, post = postBeta, param = "Mean", supportLevel = 0.7)
 #d <- recordPlot()
 ### variance partitioning ###
[…]

### 2024-10-22 code/R/fig1.R
Improved Mod3 plots

@@ -333,14 +333,16 @@ plotGradient(m.3.sample,Gradient,pred=predY,showData=T,measure="Y",index=1,main=
 p11 <- recordPlot()
 plotGradient(m.3.sample,Gradient,pred=predY,showData=T,measure="Y",index=2,main="",xlab="E_t",ylab="predicted N2_t+1",showPosteriorSupport=FALSE,cex.axis=0.75)
 p12 <- recordPlot()
-
+### Beta posterior plot ###
 m3.post.hmsc <- convertToCodaObject(m.3.sample)
-summary(m3.post.hmsc$Beta)
+m.beta <- as.data.frame(rbind(m3.post.hmsc$Beta[[1]],m3.post.hmsc$Beta[[2]]))
+colnames(m.beta) <- c("Int1","E1","Esq1","dx1_1","dx1_2","Int2","E2","Esq2","dx2_1","dx2_2")
 p13 <- bayesplot::mcmc_trace(m3.post.hmsc$Beta)
-p14 <- bayesplot::mcmc_areas(m3.post.hmsc$Beta,area_method = c("equal height"))
+p14 <- bayesplot::mcmc_areas(m.beta) + scale_x_continuous(limits=c(-20,30))
 
 VP <- computeVariancePartitioning(m.3.sample,group=c(1,1,1,2,3),groupnames=c("Env","Sp1","Sp2"))
-p15 <- plotVariancePartitioning(m.3.sample,VP,cols=c("white","skyblue","darkgrey","orange","pink"),args.legend=list(cex=0.75,bg="transparent"))
+plotVariancePartitioning(m.3.sample,VP,cols=c("white","skyblue","darkgrey","orange","pink"),args.legend=list(cex=0.75,bg="transparent"))
+p15 <- recordPlot()
 
 # prepare for plot
 pred3 <- predict(m.3.sample)
@@ -372,6 +374,18 @@ gg_dat$site <- rep(study$site[study$time !=1],times=2)
 gg_dat$time <- rep(study$time[study$time !=1],times=2)
 gg_dat$ci_025 <- apply(yrep,2,quantile,probs=.025)
 gg_dat$ci_975 <- apply(yrep,2,quantile,probs=.975)
[…]

### 2024-10-23 code/R/fig1.R
Finished full-panel figure, cleaned for draft

@@ -40,9 +40,12 @@ for (i in 2:t) {
   E[i] <- E[i-1] + rnorm(1,0,0.1)
 }
 
-N$time <- 1:t
-dat <- melt(N, id.vars="time")
-ggplot2::ggplot(dat, aes(time, value, col=variable)) + geom_point() + geom_hline(yintercept = ((r[1] - 1)/alpha.11),linetype = "dashed", color = "gray")
+### Plot of time series ###
+gdat <- N
+gdat$time <- 1:t
+dat <- melt(gdat, id.vars="time")
+pa1 <- ggplot2::ggplot(dat, aes(time, value, col=variable)) + geom_point() + geom_hline(yintercept = ((r[1] - 1)/alpha.11),linetype = "dashed", color = "gray")
+### ###
 
 ### Mod1: HMSC model fit 1 ###
 # prepare data in HMSC format
@@ -75,7 +78,7 @@ for(i in 1:length(r)){
 y <- unlist(hold1) # vector of each species' observed N across time points
 yrep <- as.matrix(data.frame(hold2))
 ### Mod1 plot ###
-par(mgp=c(2.2,0.45,0), tcl=-0.4, mar=c(2,2,1,1))
+par(mgp=c(2,0.45,0), tcl=-0.4, mar=c(1.3,1.2,0,0))
 color_scheme_set("brightblue")
 #ppc_dens_overlay(y, yrep[1:50, ])
 p1 <- ppc_intervals(y,yrep,x=rep(dat$time[1:(t-1)],times=2),prob = 0.95) +
@@ -86,26 +89,25 @@ p1 <- ppc_intervals(y,yrep,x=rep(dat$time[1:(t-1)],times=2),prob = 0.95) +
 Gradient <- constructGradient(m.1.sample,focalVariable="E",non.focalVariables=list(Esq=list(2)),ngrid=39)
 Gradient$XDataNew$Esq <- Gradient$XDataNew$E^2
 predY <- predict(m.1.sample,XData=Gradient$XDataNew,expected=TRUE)
-plotGradient(m.1.sample,Gradient,pred=predY,showData=T,measure="Y",index=1,main="",xlab="E_t",ylab="predicted N1_t+1",showPosteriorSupport=FALSE,cex.axis=0.75)
[…]

### 2024-10-23 manuscript/manuscript_v11.Rmd
Updated manuscript to blank, checked code/R

@@
   Distribution Models"
-author: "Jelena H. Pantel, Ruben J. Hermann"
-date: "Ecological Modelling, Faculty of Biology, University of Duisburg-Essen, Universitätsstraße
-  5, 45141 Essen, Germany"
+author:
+  - Jelena H. Pantel^[Laboratoire Chrono-environnement,UMR 6249 CNRS-UFC, 16 Route de Gray, 25030 Besançon cedex, France, jelena.pantel@univ-fcomte.fr]
+  - Ruben J. Hermann^[University of Duisburg-Essen, Universitätsstraße 5, 45141 Essen, Germany, ruben.hermann@uni-due.de]
+date: "`r format(Sys.time(), '%d %B, %Y')`"
 output:
@@
     latex_engine: xelatex
+    extra_dependencies: ["float"]
     number_sections: true
@@
 - \pagestyle{fancy}
-- \fancyhead[L]{Author XXX}
+- \fancyhead[L]{Author Pantel and Hermann}
 - \fancyhead[R]{Detection of ecoevo dynamics}
@@
 ```{r setup,echo=FALSE,warning=FALSE,message=FALSE}
-library(R.matlab)
 library(Hmsc)
@@
 
-Detection of eco-evolutionary dynamics in metacommunities using Joint Species
-  Distribution Models
+Detection of eco-evolutionary dynamics in metacommunities using Joint Species Distribution Models
 
@@
 
-We simulated growth and competition dynamics for a multi-species assemblage in a patchy landscape, with site variation in one environmental property (Figure \@ref(fig:fig1)).
-
-```{r fig1, echo = FALSE, eval = TRUE, fig.width = 5, fig.height = 5, fig.cap = "Spatial maps showing position of all 50 sites in the simulation, their value for the single environmental variable E, and their value for initial Simpson's diversity at the beginning of the simulation.", warning=FALSE,message=FALSE}
-
-## Figure 1. Plot of location of sites in landscape, associated E values, and associated diversity values
-result <- paste("../data/h_01_d_minus3_res.mat",sep="")
-dat  <- R.matlab::readMat(toString(result))
-
-layout(matrix(c(1,3,2,4),ncol=2), width = c(2,1),height = c(1,1))
-pal <- colorRampPalette(c("white", "blue"))
-plot(dat$xy,col="grey90", fill=TRUE, warnings=F, xlab="X coord",ylab="Y coord")
-options(warn=0)
-points(dat$xy[,1], dat$xy[,2],pch = 21,bg = pal(10)[as.numeric(cut(dat$E,breaks = 10))],col="black")
-
-legend_image <- as.raster(matrix(rev(pal(10)), ncol=1))
-plot(c(0,2),c(0,1),type = 'n', axes = F,xlab = '', ylab = '', main = 'E value')
-text(x=1.5, y = seq(0,1,l=5), labels = seq(0,1,l=5))
-rasterImage(legend_image, 0, 0, 1, 1) ## set new for overplot w/ next plot
-
-pal <- colorRampPalette(c("white", "orange"))
-a  <- vegan::diversity(dat$N[,,1],index="invsimpson")
-a[is.finite(a)==FALSE] <- 0
-plot(dat$xy,col="grey90", fill=TRUE, warnings=F, xlab="X coord",ylab="Y coord")
-options(warn=0)
-points(dat$xy[,1], dat$xy[,2],pch = 21,bg = pal(10)[as.numeric(cut(a,breaks = 10))],col="black")
-
-legend_image <- as.raster(matrix(rev(pal(10)), ncol=1))
-plot(c(0,2),c(0,1),type = 'n', axes = F,xlab = '', ylab = '', main = 'inverse Simpson diversity')
-text(x=1.5, y = seq(0,1,l=6), labels = seq(0,15,l=6))
-rasterImage(legend_image, 0, 0, 1, 1, cex=0.75) ## set new for overplot w/ next plot
-```
+We simulated growth and competition dynamics for a multi-species assemblage in a patchy landscape, with site variation in one environmental property.
 
@@
 Populations have a critical density $N_c$, below which the population is subject to extinction due to demographic stochasticity at a probability of *p* [@Gomulkiewicz1995].
-
-This model was used in Pantel & Becks [@Pantel2023] to evaluate the consequences of adaptive evolution for coexistence in a three-species system.
-To expand this model to a metacommunity, we consider the evolution of multiple species (*S* = 15) in a landscape of patches (*k* = 50).
-The patches have values of an environmental property *E* (drawn from a uniform distribution ranging from 0 to 1) that determines the local optimum phenotype *E* = $x_{opt}$, i.e.
-where species experience their absolute fitness $W_{max}$, and patches also have spatial locations *X* and *Y* (both drawn from U(0, 1)).
-The patches thus have a connectivity matrix **D** (here given by their Euclidean distance), as well as a connectivity matrix **C** that is a Gaussian function of **D** and species dispersal rate $d_i$.
-
-To simulate trait evolution together with species growth and competition in the landscape, we used parameter values of \(W_{max}\) = 2, *P* = 1, *w* = 10, $N_c$ = 100, and *p* = 0.001.
-The simulated metacommunity was initialized in a way that mimics community assembly in small vernal pools that periodically flood (and thus are reset seasonally).
-In the simulated pool system, local communities are drawn at random from the regional species pool, and thus their initial degree of maladaptation in the pool is random as well.
-The initial species richness of each site $s_{0k}$ was drawn from a random Poisson distribution, $s_{0k} \sim Pois(0.75)$, the initial population size for all species $N_{0i} \sim Pois(10)$, and the initial degree of maladaptation \(\beta_{i0} \sim Gamma(0.75,1)\).
-From this \(\beta_{i0}\), the initial distance to the local patch's optimum phenotype was calculated as $d_{i0}=\sqrt{B_{i0}(w+P)}$ and the initial trait value for each species $x_{i0}$ was calculated by subtracting $d_{i0}$ from the local environmental optimum value $E_k$ (Appendix S1, Figure S1.1).
-
-The species interaction matrix was fixed throughout the simulation, with \(\alpha_{ii} = 0.00125\) and \(\alpha_{ij} \sim U(0.0015,1)\).
-After draws for all random parameter values, the same resulting initial community was used for all simulations, to allow direct comparison across dispersal and heritability values.
-Simulations were run across a range of dispersal values (identical for all species; *d* = 0, $10^{-9},10^{-8},10^{-7},10^{-6},10^{-5},10^{-4},10^{-3},10^{-2}$, 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 1) and heritability values (identical for all species: $h^2$ = 0, 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 1).
-The simulations were run in MATLAB (version R2023a) for 1000 time steps and produced records of $N_{ikt}$ and $x_{ikt}$.
-To test the impacts of strength of selection and environmental variability, we repeated simulations at a single, representative level of heritability and dispersal ($d = 10^{-3}$, $h^2 = 0.1$) in four different scenarios: Scenario 1 (weak selection, fixed environment: $P = 1, w = 10, v[E] = 0$), Scenario 2 (strong selection, fixed environment: $P = 0.1, w = 0.2, v[E] = 0$), Scenario 3 (weak selection, variable environment: $P = 1, w = 10, v[E] = 0.01$), and Scenario 4 (strong selection, variable environment: P = 0.1, w = 0.2, v[E] = 0.01).
-Environmental variability was implemented by adding random noise to $E_k$ every 10 time steps ($E_{k,t+1} \sim N(E_{k,t},v[E])$).
-
-*Statistical model*
-
-To determine the relative influence of environmental and spatial properties, intra- and interspecific population dynamics, and evolutionary dynamics for community structure, we applied hierarchical modelling of species communities (HMSC, [@Ovaskainen2017HMSC] [@Tikhonov2020]) to model the abundance at each site and year for each species $N_{ikt}$.
-The abundance values were modeled as $N_{ikt} \sim N(L_{ikt},\sigma_2)$, where the linear predictors **L** are the sum of fixed and random effects **L^F^** + **L^R^** [@Tikhonov2020].
-The fixed effects were (1) $E_k$, the value of the site’s environmental parameter (predictor 1), (2) the value of each species' population size the time step before $N_{ik,t-1}$ (predictors 2-16; see [@Sandal2022]), (3) the absolute value of the change in each species' trait value from the previous time step to the next $|\Delta x_{ik,t-1 \to t}|$ (predictors 17-31), (4) and an interaction term for population size by change in trait value $N_{ik,t-1} \times |\Delta x_{ik,t-1 \to t}|$ (predictors 32-46), which led to a total of 46 predictors.
-The random effects were modeled at the site-level ($\epsilon_{ik} \sim N(0, \Omega^S)$, where $\Omega^S$ is the site species-to-species covariance matrix) and at the time step level ($\epsilon_{it} \sim N(0, \Omega^T)$, where $\Omega^T$ is the temporal species-to-species covariance matrix) using a latent variable approach.
-In this approach, species-to-species covariance matrices are computed as $\Omega = \lambda \intercal \lambda$, where $\lambda$ are the loadings of each species with latent variables, and latent variables are modeled as $\epsilon_{ik} = \sum_B \eta_{kB} \lambda_{Bi}$ (site) and $\epsilon_{it} = \sum_B \eta_{tB} \lambda_{Bi}$ (time) for *B* latent variables considered [@Ovaskainen2016].
-The spatial latent variables represent a spatial model, where species respond to some latent spatial predictor associated with the *xy* coordinates of the sites [@Tikhonov2020].
-Analysis and was conducted in R (version 4.2.3) using packages ‘Hmsc’ and 'R.matlab' ([@Bengtsson2022]; [@Tikhonov2022]).
 

### 2024-10-27 manuscript/manuscript_v11.Rmd

@@
 
-## Simulation model
+## Linear statistical models for one- and two-species models of growth, competition, and trait evolution
+
+We evaluated whether linear regression models can appropriately capture the dynamics that drive species abundances in more complex communities.
+We fit data generated from simulation models with increasing complexity to Bayesian linear regression models.
+These simulation models were (i) a single-species model of discrete-time logistic population growth, (ii) expanded to include a randomly fluctuating environmental property $E$ that impacts the species' population growth rate, (iii) a two-species model of logistic growth and competition that was then (iv) expanded to consider a randomly fluctuating environmental property, (v) and further expanded to consider evolution in a trait $x$ that is selected by the environmental property, and finally (vi) the model of two-species growth, competition, environmental change, and evolution in a spatially structured (10-patch) habitat.
+
+For linear regression models to fit to the simulated population size data, we began with models using population size and environment as fixed effects ($N_t \sim N_{t+1} + E_t + E_t^2$), and progressed to a mixed model using environment and trait evolution as fixed effects ($N_t \sim E_t + E_t^2 + |\Delta x|$) and time and spatial distance as random effects.
+For consistency, all linear regression models were implemented using HMSC (Hierarchical Modelling of Species Communities [@Ovaskainen2017HMSC] [@Tikhonov2020]).
+HMSC uses a latent variable approach as an alternative to directly modelling species interaction coefficients, estimating a reduced number of linear combinations of species abundances that best predict future population size values.
+For linear models fit to data generated in simulations *iii*, *iv*, *v*, and *vi* above, the time random effect was implemented to capture species interactions (as part of the temporal species association matrix estimated by the model), and the spatial random effect was implemented in model *vi* to capture spatial covariance in abundances as a function of distance between sites.
+Full methodological details are given in Appendix S1.
+
+## Evolution in metacommunities and HMSC
 
@@
 
-XXXX
+## Linear statistical models for one- and two-species models of growth, competition, and trait evolution
+
+Linear regression models can estimate the impacts of intra- and interspecific competition, environmental variation, and trait evolution for population size in simple one- and two-species models of population growth.
+In simple cases (one-species models, linear and quadratic relationships between environment and population size), the regression coefficients successfully reflect the direction and magnitude of the relationships included in the non-linear simulation model (the impact of $N_t$ and $E_t$ for $N_{t+1}$).
+However, as the processes included in the simulation models become slightly more complex (competition with a second species, spatial structure), regression coefficients were useful for prediction but not for recreating the underlying processes driving changes in population size (Appendix S1; REF?).
+
+To consider species interactions and environmental fluctuations (model *iv*; Figure 1a \@ref(fig:fig1)), we used the latent variable modelling approach implemented in HMSC.
+The linear model successfully estimated coefficients for fixed environmental effects that captured the impacts of environment for population size in both species (Figure 1b,c) and partitioned the relative importance of the environment and species interactions for species abundances (Figure 1d).
+The model coefficients could also predict the population size values over time (Figure 1a), though the quality of this prediction varied depending on the level of environmental noise (Appendix S1).
+
+```{r, out.width="90%", include=TRUE, fig.align="center", fig.cap=c("your caption"), echo=FALSE}
+knitr::include_graphics("../output/fig1.pdf")
+```
+
+The HMSC model was successfully modelled changes in species abundances when trait evolution was introduced (model *v*; Figure 1e \@ref(fig:fig1)).
+Fixed effect coefficients for the environmental driver and the total trait change from one time to the next (|\Delta x|) matched well with observed data (Figure 1f,g).
+The model estimated the relative importance of the environmental driver, trait evolution in both species, and species interactions (Figure 1h), and predicted the observed species abundances well (Figure 1e).
+Spatial structure was also reproduced in the coefficients of the linear HMSC model by using spatial random effects to capture covariance in species abundances as a function of spatial distance between sites in the 10-patch simulation model (model *vi*; Figure 1i,j,k \@ref(fig:fig1)).
+The variation partition successfully reflects separation of environmental drivers from trait evolution, species interactions, and spatial random effects (Figure 1l), and we show in Appendix S1 that the relative importance of trait evolution increases when we use a simulation model with less spatial and environmental variation and weak species interactions.
+
+## Evolution in metacommunities and HMSC
+
+We simulated growth and competition dynamics for a multi-species assemblage in a patchy landscape, with site variation in one environmental property.
+
+### Environmental variation
+
+Population growth for spec
 

### 2024-10-28 code/R/1_metacom_sim_v01.R
Changes to dt added

@@ -18,20 +18,29 @@ check <- function(x){
     x <=d[i]
 }
 
-# This function is the randome sample to chose amone sites where the species can immigrate,
+# This function is the random sample to chose among sites where the species can immigrate,
 # weighted by connectivitx.
 # It takes simply a vector of all patches in, so:
 # x = site/1:50 and i = species number (is part of the loop for dispersal)
 disp_patch <- function(x){
     sample(x=site[-x],size=mig_count[x],replace=T,prob=connec[i,-x,x])
 }
-#Function to update the trait value, where the environmental value is substracted
+#Function to update the trait value, where the environmental value is subtracted
 #by the distance of the trait (which is changed by evo)
 #x = The number of the species
 #xt_up = the updated trait value, which is saved into xt
 up_trait <- function(x){
   xt_up<-E-dt[,x,t]
 }
+
+#Function to update the trait distance value, to the optimal value
+#by the new E value
+#x = The number of the species
+#dt_up = the updated trait distance value, which is saved into dt
+up_dist <- function(x){
+  dt_up <- E-xt[,x,t]
+}
+
 ###################################
 ## Step 1. Initialize species    ##
 ###################################
@@ -185,13 +194,17 @@ for (t in 2:200) {
     last=N[,,t-1]
     
     ## Carry over all populations that are above the critical population density
-    pop[last>=100] = last[last>=100] #this had to be moved up, as R does not like the NAs, when accesing position information
+    pop[last>=100] = last[last>=100] #this had to be moved up, as R does not like the NAs, when accessing position information
     pop[pop>extprob & pop < 100] = last[N[,,t-1]>0 & N[,,t-1]<100 & pop>extprob] # Surviving populations
 
-  
-
-    ## Growth of all remaining populations
-    dt[,,t] = sweep(dt[,,t-1],2,k,"*") # Update distance to optimum phenotype
+    
+## Growth of all remaining populations
+    if (t%%10==0) {
+      dt[,,t] = sapply(1:spec,up_dist) # updating trait distance to the new environmental optimum
+      dt[,,t] = sweep(dt[,,t],2,k,"*")}  # update distance with evolution to the new optimum
+    if (t%%10!=0) {dt[,,t] = sweep(dt[,,t-1],2,k,"*")}  # update distance with evolution to the new optimum
+    
+    # Update distance to optimum phenotype
     # Revise the value for the population that went stochastically extinct this generation
     if (nrow(extinct)>0) {
         for (X in 1:nrow(extinct)) {

### 2024-10-28 code/R/1_metacom_sim_v01.R
Small correction

@@ -38,7 +38,7 @@ up_trait <- function(x){
 #x = The number of the species
 #dt_up = the updated trait distance value, which is saved into dt
 up_dist <- function(x){
-  dt_up <- E-xt[,x,t]
+  dt_up <- E-xt[,x,t-1]
 }
 
 ###################################

### 2024-10-28 code/R/7_metacom_sim_v02.R
Code for all d and h2 levels

@@ -1,0 +1,309 @@
+## Evolving metacommunity simulation
+## J.H. Pantel & Ruben Hermann
+## October 27 2024
+## Scenario 1. across d and h2 levels
+## # P = 1, w = 2, v[E] = 0
+  
+set.seed(1)
+
+###################################
+## Step 0. Small functions    ##
+###################################
+# I needed to write some small function, which lets me run the dispersal lines smoother
+# Here I check if the drawn runif number is smaller then the likelihood of dispersal
+# if so then one individual with be migrating
+# x = rlist (list of randomly generated numbers (runif) for each individual in a patch of species i)
+# i = species number (is part of the loop for dispersal)
+check <- function(x){
+    x <=d[i]
+}
+
+# This function is the randome sample to chose amone sites where the species can immigrate,
+# weighted by connectivitx.
+# It takes simply a vector of all patches in, so:
+# x = site/1:50 and i = species number (is part of the loop for dispersal)
+disp_patch <- function(x){
+    sample(x=site[-x],size=mig_count[x],replace=T,prob=connec[i,-x,x])
+}
+#Function to update the trait value, where the environmental value is substracted
+#by the distance of the trait (which is changed by evo)
+#x = The number of the species
+#xt_up = the updated trait value, which is saved into xt
+up_trait <- function(x){
+  xt_up<-E-dt[,x,t]
+}
+
+###################################
+## Step 1. Initialize species    ##
+###################################
+patch = 50   # Number of patches in the landscape
+spec = 15    # Number of species in the community
+time = 1000  # Number of time steps to run the simulation
+N = array(numeric(),c(patch,spec,time))   # Record of abundances of all species over time patch x spec x time
+P = array(1,c(1,spec))    # Phenotypic variances for all species
+w = array(2,c(1,spec))   # Width of fitness curve
+Wmax = 2    # Maximum population growth rate for all species
+Wt = array(numeric(),c(patch,spec,time))   # Record of fitness vlaues over time, patch x spec x time
+dt = array(numeric(),c(patch,spec,time))   # Record of distance from optimum value over time, patch x spec x time
[…]

### 2024-10-29 manuscript/Appendix_S4_v01.Rmd
Updated MC sims (across d, h2). Started walk-through on sim+hmsc tests across h2 levels.

---
title: "Appendix S4. The relative importance of evolution in populations and communities"
author:
  - Jelena H. Pantel^[Laboratoire Chrono-environnement,UMR 6249 CNRS-UFC, 16 Route de Gray, 25030 Besançon cedex, France, jelena.pantel@univ-fcomte.fr]
  - Ruben J. Hermann^[University of Duisburg-Essen, Universitätsstraße 5, 45141 Essen, Germany, ruben.hermann@uni-due.de]
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::pdf_document2:
    toc: no
    fig_caption: yes
    latex_engine: xelatex
    extra_dependencies: ["float"]
  bookdown::html_document2: default
  word_document:
    reference_docx: docx_template.docx
header-includes:
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \fancyhead[L]{Author Pantel and Hermann}
- \fancyhead[R]{Detection of ecoevo dynamics}
- \usepackage{lineno}
- \linenumbers
link-citations: yes
linkcolor: blue
csl: the-american-naturalist.csl
bibliography: refs.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
```{r setup,echo=FALSE,message=FALSE}
library(ecoevor)
library(ggplot2)
library(reshape2)
library(Hmsc)
library(bayesplot)
```
```{r,echo=FALSE,message=FALSE,warning=FALSE}
# Wrap long lines
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy=TRUE,out.width="50%")
```

# Logistic growth and trait evolution

We can create simulations of evolutionary rescue in a population, where adaptive evolution rescues a population from extinction, at a speed that depends on the heritability level in the system. Population dynamics are modelled as:

$$ N_{i,t+1} =\frac{\hat{W}e^ \frac{-[(\frac{w+(1-h^2)P} {P+w})(E-x_{i,t})]^2}{2(P+w)}N_{i,t}} {1 + \alpha_{ii}N_{i,t} + \alpha_{ij} N_{j,t}} $$

where \( \hat{W} \) is calculated as \( \hat{W}=W_{max}\sqrt(\frac{w}{P+w}) \), \(W_{max}\) is the species' maximum per-capita growth rate, *w* is the width of the Gaussian fitness function (which determines the strength of selection, as increasing values indicate a weaker reduction in fitness with distance from optimum trait value), *P* is the width of the distribution of the phenotype *x*, and $h^2$ is the heritability of the trait *x*. For the simulation we use \(W_{max}\) = 2, *P* = 1, and *w* = 2.

The change in the average trait value each time step is given by:

$$ d_{i,t+1} = k d_{i,t} $$
where \( k = \frac{w + (1-h^2) P}{w + P} \) and \( d_{i,t} = E_t - x_{i,t} \).

## Population dynamics simulation

```{r}
### 1 species, h2=0 ###
# Initial conditions
N0 <- 10
alpha.11 <- 0.01
alpha <- matrix(alpha.11,byrow=FALSE)
E.0 <- 0.8
P <- 1
w <- 0.5
# Draw initial trait value using degree of initial maladaptation
B0 <- 1
d0 = sqrt(B0*(w+P))
x.0 = E.0-d0
Wmax <- 2
h2 <- 0
k <- (w + (1 - h2) * P)/(P + w)
# Simulation of model for t time steps
# Simulation of model for t time steps
t <- 60
N <- array(NA,dim=c(t,length(N0)))
N <- as.data.frame(N)
colnames(N) <- paste0("N",1:length(N0))
N[1,] <- N0
E <- rep(NA, t)
E[1] <- E.0
x <- array(NA,dim=c(t,length(N0)))
x <- as.data.frame(x)
colnames(x) <- paste0("x",1:length(N0))
x[1,] <- x.0
r <- array(NA,dim=c(t,length(N0)))
r <- as.data.frame(r)
colnames(r) <- paste0("r",1:length(N0))
What <- Wmax*sqrt(w/(P+w))
r[1,] <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E[1]-x[1,]))^2)/(2*(P+w)))
for (i in 2:t) {
  res <- disc_LV_evol(N0=N[i-1,],alpha=alpha,E=E[i-1],x=x[i-1,],P=P,w=w,Wmax=Wmax,h2=h2)
  N[i,] <- res$Nt1
  r[i,] <- res$r
  # trait change
  d <- E[i-1] - x[i-1,]
  d1 <- k*d
  x[i,] <- E[i-1] - d1
  # environmental change
  E[i] <- E[i-1] + abs(rnorm(1, 0, 0))
}
# Arrange to save across scenarios
gdat <- N
gdat$time <- 1:t
gdat$h2 <- h2
gdat$x1 <- x$x1
gdat$E <- E
```
```{r}
### 1 species, h2=0.01 ###
# Initial conditions
N0 <- 10
alpha.11 <- 0.01
alpha <- matrix(alpha.11,byrow=FALSE)
E.0 <- 0.8
P <- 1
w <- 0.5
# Draw initial trait value using degree of initial maladaptation
B0 <- 1
d0 = sqrt(B0*(w+P))
x.0 = E.0-d0
Wmax <- 2
h2 <- 0.01
k <- (w + (1 - h2) * P)/(P + w)
# Simulation of model for t time steps
# Simulation of model for t time steps
t <- 60
N <- array(NA,dim=c(t,length(N0)))
N <- as.data.frame(N)
colnames(N) <- paste0("N",1:length(N0))
N[1,] <- N0
E <- rep(NA, t)
E[1] <- E.0
x <- array(NA,dim=c(t,length(N0)))
x <- as.data.frame(x)
colnames(x) <- paste0("x",1:length(N0))
x[1,] <- x.0
r <- array(NA,dim=c(t,length(N0)))
r <- as.data.frame(r)
colnames(r) <- paste0("r",1:length(N0))
What <- Wmax*sqrt(w/(P+w))
r[1,] <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E[1]-x[1,]))^2)/(2*(P+w)))
for (i in 2:t) {
  res <- disc_LV_evol(N0=N[i-1,],alpha=alpha,E=E[i-1],x=x[i-1,],P=P,w=w,Wmax=Wmax,h2=h2)
  N[i,] <- res$Nt1
  r[i,] <- res$r
  # trait change
  d <- E[i-1] - x[i-1,]
  d1 <- k*d
  x[i,] <- E[i-1] - d1
  # environmental change
  E[i] <- E[i-1] + abs(rnorm(1, 0, 0))
}
# Arrange to save across scenarios
gdat <- rbind(gdat,data.frame(N,time=1:t,h2=rep(h2,t),x1=x$x1,E=E))
```
```{r}
### 1 species, h2=0.02 ###
# Initial conditions
N0 <- 10
alpha.11 <- 0.01
alpha <- matrix(alpha.11,byrow=FALSE)
E.0 <- 0.8
P <- 1
w <- 0.5
# Draw initial trait value using degree of initial maladaptation
B0 <- 1
d0 = sqrt(B0*(w+P))
x.0 = E.0-d0
Wmax <- 2
h2 <- 0.02
k <- (w + (1 - h2) * P)/(P + w)
# Simulation of model for t time steps
# Simulation of model for t time steps
t <- 60
N <- array(NA,dim=c(t,length(N0)))
N <- as.data.frame(N)
colnames(N) <- paste0("N",1:length(N0))
N[1,] <- N0
E <- rep(NA, t)
E[1] <- E.0
x <- array(NA,dim=c(t,length(N0)))
x <- as.data.frame(x)
colnames(x) <- paste0("x",1:length(N0))
x[1,] <- x.0
r <- array(NA,dim=c(t,length(N0)))
r <- as.data.frame(r)
colnames(r) <- paste0("r",1:length(N0))
What <- Wmax*sqrt(w/(P+w))
r[1,] <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E[1]-x[1,]))^2)/(2*(P+w)))
for (i in 2:t) {
  res <- disc_LV_evol(N0=N[i-1,],alpha=alpha,E=E[i-1],x=x[i-1,],P=P,w=w,Wmax=Wmax,h2=h2)
  N[i,] <- res$Nt1
  r[i,] <- res$r
  # trait change
  d <- E[i-1] - x[i-1,]
  d1 <- k*d
  x[i,] <- E[i-1] - d1
  # environmental change
  E[i] <- E[i-1] + abs(rnorm(1, 0, 0))
}
# Arrange to save across scenarios
gdat <- rbind(gdat,data.frame(N,time=1:t,h2=rep(h2,t),x1=x$x1,E=E))
```
```{r}
### 1 species, h2=0.03 ###
# Initial conditions
N0 <- 10
alpha.11 <- 0.01
alpha <- matrix(alpha.11,byrow=FALSE)
E.0 <- 0.8
P <- 1
w <- 0.5
# Draw initial trait value using degree of initial maladaptation
B0 <- 1
d0 = sqrt(B0*(w+P))
x.0 = E.0-d0
Wmax <- 2
h2 <- 0.03
k <- (w + (1 - h2) * P)/(P + w)
# Simulation of model for t time steps
# Simulation of model for t time steps
t <- 60
N <- array(NA,dim=c(t,length(N0)))
N <- as.data.frame(N)
colnames(N) <- paste0("N",1:length(N0))
N[1,] <- N0
E <- rep(NA, t)
E[1] <- E.0
x <- array(NA,dim=c(t,length(N0)))
x <- as.data.frame(x)
colnames(x) <- paste0("x",1:length(N0))
x[1,] <- x.0
r <- array(NA,dim=c(t,length(N0)))
r <- as.data.frame(r)
colnames(r) <- paste0("r",1:length(N0))
What <- Wmax*sqrt(w/(P+w))
r[1,] <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E[1]-x[1,]))^2)/(2*(P+w)))
for (i in 2:t) {
  res <- disc_LV_evol(N0=N[i-1,],alpha=alpha,E=E[i-1],x=x[i-1,],P=P,w=w,Wmax=Wmax,h2=h2)
  N[i,] <- res$Nt1
  r[i,] <- res$r
  # trait change
  d <- E[i-1] - x[i-1,]
  d1 <- k*d
  x[i,] <- E[i-1] - d1
  # environmental change
  E[i] <- E[i-1] + abs(rnorm(1, 0, 0))
}
# Arrange to save across scenarios
gdat <- rbind(gdat,data.frame(N,time=1:t,h2=rep(h2,t),x1=x$x1,E=E))
```
```{r}
### 1 species, h2=0.04 ###
# Initial conditions
N0 <- 10
alpha.11 <- 0.01
alpha <- matrix(alpha.11,byrow=FALSE)
E.0 <- 0.8
P <- 1
w <- 0.5
# Draw initial trait value using degree of initial maladaptation
B0 <- 1
d0 = sqrt(B0*(w+P))
x.0 = E.0-d0
Wmax <- 2
h2 <- 0.04
k <- (w + (1 - h2) * P)/(P + w)
# Simulation of model for t time steps
# Simulation of model for t time steps
t <- 60
N <- array(NA,dim=c(t,length(N0)))
N <- as.data.frame(N)
colnames(N) <- paste0("N",1:length(N0))
N[1,] <- N0
E <- rep(NA, t)
E[1] <- E.0
x <- array(NA,dim=c(t,length(N0)))
x <- as.data.frame(x)
colnames(x) <- paste0("x",1:length(N0))
x[1,] <- x.0
r <- array(NA,dim=c(t,length(N0)))
r <- as.data.frame(r)
colnames(r) <- paste0("r",1:length(N0))
What <- Wmax*sqrt(w/(P+w))
r[1,] <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E[1]-x[1,]))^2)/(2*(P+w)))
for (i in 2:t) {
  res <- disc_LV_evol(N0=N[i-1,],alpha=alpha,E=E[i-1],x=x[i-1,],P=P,w=w,Wmax=Wmax,h2=h2)
  N[i,] <- res$Nt1
  r[i,] <- res$r
  # trait change
  d <- E[i-1] - x[i-1,]
  d1 <- k*d
  x[i,] <- E[i-1] - d1
  # environmental change
  E[i] <- E[i-1] + abs(rnorm(1, 0, 0))
}
# Arrange to save across scenarios
gdat <- rbind(gdat,data.frame(N,time=1:t,h2=rep(h2,t),x1=x$x1,E=E))
```
```{r}
### 1 species, h2=0.05 ###
# Initial conditions
N0 <- 10
alpha.11 <- 0.01
alpha <- matrix(alpha.11,byrow=FALSE)
E.0 <- 0.8
P <- 1
w <- 0.5
# Draw initial trait value using degree of initial maladaptation
B0 <- 1
d0 = sqrt(B0*(w+P))
x.0 = E.0-d0
Wmax <- 2
h2 <- 0.05
k <- (w + (1 - h2) * P)/(P + w)
# Simulation of model for t time steps
# Simulation of model for t time steps
t <- 60
N <- array(NA,dim=c(t,length(N0)))
N <- as.data.frame(N)
colnames(N) <- paste0("N",1:length(N0))
N[1,] <- N0
E <- rep(NA, t)
E[1] <- E.0
x <- array(NA,dim=c(t,length(N0)))
x <- as.data.frame(x)
colnames(x) <- paste0("x",1:length(N0))
x[1,] <- x.0
r <- array(NA,dim=c(t,length(N0)))
r <- as.data.frame(r)
colnames(r) <- paste0("r",1:length(N0))
What <- Wmax*sqrt(w/(P+w))
r[1,] <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E[1]-x[1,]))^2)/(2*(P+w)))
for (i in 2:t) {
  res <- disc_LV_evol(N0=N[i-1,],alpha=alpha,E=E[i-1],x=x[i-1,],P=P,w=w,Wmax=Wmax,h2=h2)
  N[i,] <- res$Nt1
  r[i,] <- res$r
  # trait change
  d <- E[i-1] - x[i-1,]
  d1 <- k*d
  x[i,] <- E[i-1] - d1
  # environmental change
  E[i] <- E[i-1] + abs(rnorm(1, 0, 0))
}
# Arrange to save across scenarios
gdat <- rbind(gdat,data.frame(N,time=1:t,h2=rep(h2,t),x1=x$x1,E=E))
```
```{r}
### 1 species, h2=0.075 ###
# Initial conditions
N0 <- 10
alpha.11 <- 0.01
alpha <- matrix(alpha.11,byrow=FALSE)
E.0 <- 0.8
P <- 1
w <- 0.5
# Draw initial trait value using degree of initial maladaptation
B0 <- 1
d0 = sqrt(B0*(w+P))
x.0 = E.0-d0
Wmax <- 2
h2 <- 0.075
k <- (w + (1 - h2) * P)/(P + w)
# Simulation of model for t time steps
# Simulation of model for t time steps
t <- 60
N <- array(NA,dim=c(t,length(N0)))
N <- as.data.frame(N)
colnames(N) <- paste0("N",1:length(N0))
N[1,] <- N0
E <- rep(NA, t)
E[1] <- E.0
x <- array(NA,dim=c(t,length(N0)))
x <- as.data.frame(x)
colnames(x) <- paste0("x",1:length(N0))
x[1,] <- x.0
r <- array(NA,dim=c(t,length(N0)))
r <- as.data.frame(r)
colnames(r) <- paste0("r",1:length(N0))
What <- Wmax*sqrt(w/(P+w))
r[1,] <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E[1]-x[1,]))^2)/(2*(P+w)))
for (i in 2:t) {
  res <- disc_LV_evol(N0=N[i-1,],alpha=alpha,E=E[i-1],x=x[i-1,],P=P,w=w,Wmax=Wmax,h2=h2)
  N[i,] <- res$Nt1
  r[i,] <- res$r
  # trait change
  d <- E[i-1] - x[i-1,]
  d1 <- k*d
  x[i,] <- E[i-1] - d1
  # environmental change
  E[i] <- E[i-1] + abs(rnorm(1, 0, 0))
}
# Arrange to save across scenarios
gdat <- rbind(gdat,data.frame(N,time=1:t,h2=rep(h2,t),x1=x$x1,E=E))
```
```{r}
### 1 species, h2=0.1 ###
# Initial conditions
N0 <- 10
alpha.11 <- 0.01
alpha <- matrix(alpha.11,byrow=FALSE)
E.0 <- 0.8
P <- 1
w <- 0.5
# Draw initial trait value using degree of initial maladaptation
B0 <- 1
d0 = sqrt(B0*(w+P))
x.0 = E.0-d0
Wmax <- 2
h2 <- 0.1
k <- (w + (1 - h2) * P)/(P + w)
# Simulation of model for t time steps
# Simulation of model for t time steps
t <- 60
N <- array(NA,dim=c(t,length(N0)))
N <- as.data.frame(N)
colnames(N) <- paste0("N",1:length(N0))
N[1,] <- N0
E <- rep(NA, t)
E[1] <- E.0
x <- array(NA,dim=c(t,length(N0)))
x <- as.data.frame(x)
colnames(x) <- paste0("x",1:length(N0))
x[1,] <- x.0
r <- array(NA,dim=c(t,length(N0)))
r <- as.data.frame(r)
colnames(r) <- paste0("r",1:length(N0))
What <- Wmax*sqrt(w/(P+w))
r[1,] <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E[1]-x[1,]))^2)/(2*(P+w)))
for (i in 2:t) {
  res <- disc_LV_evol(N0=N[i-1,],alpha=alpha,E=E[i-1],x=x[i-1,],P=P,w=w,Wmax=Wmax,h2=h2)
  N[i,] <- res$Nt1
  r[i,] <- res$r
  # trait change
  d <- E[i-1] - x[i-1,]
  d1 <- k*d
  x[i,] <- E[i-1] - d1
  # environmental change
  E[i] <- E[i-1] + abs(rnorm(1, 0, 0))
}
# Arrange to save across scenarios
gdat <- rbind(gdat,data.frame(N,time=1:t,h2=rep(h2,t),x1=x$x1,E=E))
```
```{r}
### 1 species, h2=0.4 ###
# Initial conditions
N0 <- 10
alpha.11 <- 0.01
alpha <- matrix(alpha.11,byrow=FALSE)
E.0 <- 0.8
P <- 1
w <- 0.5
# Draw initial trait value using degree of initial maladaptation
B0 <- 1
d0 = sqrt(B0*(w+P))
x.0 = E.0-d0
Wmax <- 2
h2 <- 0.4
k <- (w + (1 - h2) * P)/(P + w)
# Simulation of model for t time steps
# Simulation of model for t time steps
t <- 60
N <- array(NA,dim=c(t,length(N0)))
N <- as.data.frame(N)
colnames(N) <- paste0("N",1:length(N0))
N[1,] <- N0
E <- rep(NA, t)
E[1] <- E.0
x <- array(NA,dim=c(t,length(N0)))
x <- as.data.frame(x)
colnames(x) <- paste0("x",1:length(N0))
x[1,] <- x.0
r <- array(NA,dim=c(t,length(N0)))
r <- as.data.frame(r)
colnames(r) <- paste0("r",1:length(N0))
What <- Wmax*sqrt(w/(P+w))
r[1,] <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E[1]-x[1,]))^2)/(2*(P+w)))
for (i in 2:t) {
  res <- disc_LV_evol(N0=N[i-1,],alpha=alpha,E=E[i-1],x=x[i-1,],P=P,w=w,Wmax=Wmax,h2=h2)
  N[i,] <- res$Nt1
  r[i,] <- res$r
  # trait change
  d <- E[i-1] - x[i-1,]
  d1 <- k*d
  x[i,] <- E[i-1] - d1
  # environmental change
  E[i] <- E[i-1] + abs(rnorm(1, 0, 0))
}
# Arrange to save across scenarios
gdat <- rbind(gdat,data.frame(N,time=1:t,h2=rep(h2,t),x1=x$x1,E=E))
```
```{r}
### 1 species, h2=0.9 ###
# Initial conditions
N0 <- 10
alpha.11 <- 0.01
alpha <- matrix(alpha.11,byrow=FALSE)
E.0 <- 0.8
P <- 1
w <- 0.5
# Draw initial trait value using degree of initial maladaptation
B0 <- 1
d0 = sqrt(B0*(w+P))
x.0 = E.0-d0
Wmax <- 2
h2 <- 0.9
k <- (w + (1 - h2) * P)/(P + w)
# Simulation of model for t time steps
# Simulation of model for t time steps
t <- 60
N <- array(NA,dim=c(t,length(N0)))
N <- as.data.frame(N)
colnames(N) <- paste0("N",1:length(N0))
N[1,] <- N0
E <- rep(NA, t)
E[1] <- E.0
x <- array(NA,dim=c(t,length(N0)))
x <- as.data.frame(x)
colnames(x) <- paste0("x",1:length(N0))
x[1,] <- x.0
r <- array(NA,dim=c(t,length(N0)))
r <- as.data.frame(r)
colnames(r) <- paste0("r",1:length(N0))
What <- Wmax*sqrt(w/(P+w))
r[1,] <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E[1]-x[1,]))^2)/(2*(P+w)))
for (i in 2:t) {
  res <- disc_LV_evol(N0=N[i-1,],alpha=alpha,E=E[i-1],x=x[i-1,],P=P,w=w,Wmax=Wmax,h2=h2)
  N[i,] <- res$Nt1
  r[i,] <- res$r
  # trait change
  d <- E[i-1] - x[i-1,]
  d1 <- k*d
  x[i,] <- E[i-1] - d1
  # environmental change
  E[i] <- E[i-1] + abs(rnorm(1, 0, 0))
}
# Arrange to save across scenarios
gdat <- rbind(gdat,data.frame(N,time=1:t,h2=rep(h2,t),x1=x$x1,E=E))
```
```{r}
### 1 species, h2=1 ###
# Initial conditions
N0 <- 10
alpha.11 <- 0.01
alpha <- matrix(alpha.11,byrow=FALSE)
E.0 <- 0.8
P <- 1
w <- 0.5
# Draw initial trait value using degree of initial maladaptation
B0 <- 1
d0 = sqrt(B0*(w+P))
x.0 = E.0-d0
Wmax <- 2
h2 <- 1
k <- (w + (1 - h2) * P)/(P + w)
# Simulation of model for t time steps
# Simulation of model for t time steps
t <- 60
N <- array(NA,dim=c(t,length(N0)))
N <- as.data.frame(N)
colnames(N) <- paste0("N",1:length(N0))
N[1,] <- N0
E <- rep(NA, t)
E[1] <- E.0
x <- array(NA,dim=c(t,length(N0)))
x <- as.data.frame(x)
colnames(x) <- paste0("x",1:length(N0))
x[1,] <- x.0
r <- array(NA,dim=c(t,length(N0)))
r <- as.data.frame(r)
colnames(r) <- paste0("r",1:length(N0))
What <- Wmax*sqrt(w/(P+w))
r[1,] <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E[1]-x[1,]))^2)/(2*(P+w)))
for (i in 2:t) {
  res <- disc_LV_evol(N0=N[i-1,],alpha=alpha,E=E[i-1],x=x[i-1,],P=P,w=w,Wmax=Wmax,h2=h2)
  N[i,] <- res$Nt1
  r[i,] <- res$r
  # trait change
  d <- E[i-1] - x[i-1,]
  d1 <- k*d
  x[i,] <- E[i-1] - d1
  # environmental change
  E[i] <- E[i-1] + abs(rnorm(1, 0, 0))
}
# Arrange to save across scenarios
gdat <- rbind(gdat,data.frame(N,time=1:t,h2=rep(h2,t),x1=x$x1,E=E))
```

```{r}
### Plot of time series ###
gdat$h2 <- as.factor(gdat$h2)
ggplot2::ggplot(gdat, aes(time, N1, col=h2)) + geom_point()
### ###
```

## HMSC model fit

The goal of fitting the data to a statistical model is to estimate the relative important of trait evolution for the population dynamics.

```{r}
### Model 1, h2=00 ###
dat <- cbind(log(gdat$N1[gdat$h2==0]),gdat$x1[gdat$h2==0])
dat <- as.data.frame(dat)
colnames(dat) <- c("N1","x1")
dat$time <- 1:t
dat <- as.data.frame(dat)
df <- data.frame(dat[(2:t),-3])
colnames(df) <- c("Nt1", "xt1")
df$dx1 <- abs(dat$x1[2:t] - dat$x1[1:(t - 1)])
# prepare data in HMSC format
Y <- as.matrix(df$Nt1)
XData <- data.frame(cbind(E[1:(t-1)],E[1:(t-1)]^2,abs(dat$x1[2:t] - dat$x1[1:(t-1)])))
colnames(XData) <- c("E","Esq","dx1")

studyDesign = data.frame(sample = as.factor(1:(t-1)))
rL = HmscRandomLevel(units = studyDesign$sample)

m.1.hmsc = Hmsc(Y = Y, XData = XData, XFormula = ~E+Esq+dx1, studyDesign = studyDesign, ranLevels = list(sample = rL))
# Bayesian model parameters
nChains <- 2
thin <- 5
samples <- 2000
transient <- 1000*thin
verbose <- 500*thin
# sample MCMC
m.1.sample <- sampleMcmc(m.1.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
m1.post.hmsc <- convertToCodaObject(m.1.sample)
```

```{r warning=FALSE}
VP1 <- computeVariancePartitioning(m.1.sample,group=c(1,1,1,2),groupnames=c("Env","x1"))
plotVariancePartitioning(m.1.sample,VP1,cols=c("white","skyblue","darkgrey"),args.legend=list(cex=0.75,bg="transparent"))
```
```{r}
# prepare for plot
pred2 <- predict(m.1.sample)
hold1 <- list()
hold2 <- list()
for(i in 1:length(N0)){
  hold1[[i]] <- as.numeric(Y[,i])
  data <- lapply(pred2, function(x) x[, colnames(x)[i], drop = FALSE])
  hold2[[i]] <- matrix(unlist(data), nrow=length(pred2), byrow=TRUE)
}
y <- unlist(hold1) # vector of each species' observed N across time points
yrep <- as.matrix(data.frame(hold2))
### Mod2 plot ###
par(mgp=c(2,0.45,0), tcl=-0.4, mar=c(1.3,1.2,0,0))
color_scheme_set("brightblue")
#ppc_dens_overlay(y, yrep[1:50, ])
ppc_intervals(y,yrep,x=dat$time[1:(t-1)],prob = 0.95) +
  labs(x = "time",y = "ln(N)",) +
  theme(plot.margin = unit(c(0, 0, 0, 0), "cm")) +
  theme(legend.position = "none")
```

```{r}
### Model 2, h2=0.01 ###
dat <- cbind(log(gdat$N1[gdat$h2==0.01]),gdat$x1[gdat$h2==0.01])
dat <- as.data.frame(dat)
colnames(dat) <- c("N1","x1")
dat$time <- 1:t
dat <- as.data.frame(dat)
df <- data.frame(dat[(2:t),-3])
colnames(df) <- c("Nt1", "xt1")
df$dx1 <- abs(dat$x1[2:t] - dat$x1[1:(t - 1)])
# prepare data in HMSC format
Y <- as.matrix(df$Nt1)
XData <- data.frame(cbind(E[1:(t-1)],E[1:(t-1)]^2,abs(dat$x1[2:t] - dat$x1[1:(t-1)])))
colnames(XData) <- c("E","Esq","dx1")

studyDesign = data.frame(sample = as.factor(1:(t-1)))
rL = HmscRandomLevel(units = studyDesign$sample)

m.2.hmsc = Hmsc(Y = Y, XData = XData, XFormula = ~E+Esq+dx1, studyDesign = studyDesign, ranLevels = list(sample = rL))
# Bayesian model parameters
nChains <- 2
thin <- 5
samples <- 2000
transient <- 1000*thin
verbose <- 500*thin
# sample MCMC
m.2.sample <- sampleMcmc(m.2.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
m2.post.hmsc <- convertToCodaObject(m.2.sample)
```

```{r warning=FALSE}
VP2 <- computeVariancePartitioning(m.2.sample,group=c(1,1,1,2),groupnames=c("Env","x1"))
plotVariancePartitioning(m.2.sample,VP2,cols=c("white","skyblue","darkgrey"),args.legend=list(cex=0.75,bg="transparent"))
```
```{r}
# prepare for plot
pred2 <- predict(m.2.sample)
hold1 <- list()
hold2 <- list()
for(i in 1:length(N0)){
  hold1[[i]] <- as.numeric(Y[,i])
  data <- lapply(pred2, function(x) x[, colnames(x)[i], drop = FALSE])
  hold2[[i]] <- matrix(unlist(data), nrow=length(pred2), byrow=TRUE)
}
y <- unlist(hold1) # vector of each species' observed N across time points
yrep <- as.matrix(data.frame(hold2))
### Mod2 plot ###
par(mgp=c(2,0.45,0), tcl=-0.4, mar=c(1.3,1.2,0,0))
color_scheme_set("brightblue")
#ppc_dens_overlay(y, yrep[1:50, ])
ppc_intervals(y,yrep,x=dat$time[1:(t-1)],prob = 0.95) +
  labs(x = "time",y = "ln(N)",) +
  theme(plot.margin = unit(c(0, 0, 0, 0), "cm")) +
  theme(legend.position = "none")
```

```{r}
### Model 3, h2=0.02 ###
dat <- cbind(log(gdat$N1[gdat$h2==0.02]),gdat$x1[gdat$h2==0.02])
dat <- as.data.frame(dat)
colnames(dat) <- c("N1","x1")
dat$time <- 1:t
dat <- as.data.frame(dat)
df <- data.frame(dat[(2:t),-3])
colnames(df) <- c("Nt1", "xt1")
df$dx1 <- abs(dat$x1[2:t] - dat$x1[1:(t - 1)])
# prepare data in HMSC format
Y <- as.matrix(df$Nt1)
XData <- data.frame(cbind(E[1:(t-1)],E[1:(t-1)]^2,abs(dat$x1[2:t] - dat$x1[1:(t-1)])))
colnames(XData) <- c("E","Esq","dx1")

studyDesign = data.frame(sample = as.factor(1:(t-1)))
rL = HmscRandomLevel(units = studyDesign$sample)

m.3.hmsc = Hmsc(Y = Y, XData = XData, XFormula = ~E+Esq+dx1, studyDesign = studyDesign, ranLevels = list(sample = rL))
# Bayesian model parameters
nChains <- 2
thin <- 5
samples <- 2000
transient <- 1000*thin
verbose <- 500*thin
# sample MCMC
m.3.sample <- sampleMcmc(m.3.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
m3.post.hmsc <- convertToCodaObject(m.3.sample)
```

```{r warning=FALSE}
VP3 <- computeVariancePartitioning(m.3.sample,group=c(1,1,1,2),groupnames=c("Env","x1"))
plotVariancePartitioning(m.3.sample,VP3,cols=c("white","skyblue","darkgrey"),args.legend=list(cex=0.75,bg="transparent"))
```
```{r}
# prepare for plot
pred2 <- predict(m.3.sample)
hold1 <- list()
hold2 <- list()
for(i in 1:length(N0)){
  hold1[[i]] <- as.numeric(Y[,i])
  data <- lapply(pred2, function(x) x[, colnames(x)[i], drop = FALSE])
  hold2[[i]] <- matrix(unlist(data), nrow=length(pred2), byrow=TRUE)
}
y <- unlist(hold1) # vector of each species' observed N across time points
yrep <- as.matrix(data.frame(hold2))
### Mod2 plot ###
par(mgp=c(2,0.45,0), tcl=-0.4, mar=c(1.3,1.2,0,0))
color_scheme_set("brightblue")
#ppc_dens_overlay(y, yrep[1:50, ])
ppc_intervals(y,yrep,x=dat$time[1:(t-1)],prob = 0.95) +
  labs(x = "time",y = "ln(N)",) +
  theme(plot.margin = unit(c(0, 0, 0, 0), "cm")) +
  theme(legend.position = "none")
```

```{r}
### Model 4, h2=0.03 ###
dat <- cbind(log(gdat$N1[gdat$h2==0.03]),gdat$x1[gdat$h2==0.03])
dat <- as.data.frame(dat)
colnames(dat) <- c("N1","x1")
dat$time <- 1:t
dat <- as.data.frame(dat)
df <- data.frame(dat[(2:t),-3])
colnames(df) <- c("Nt1", "xt1")
df$dx1 <- abs(dat$x1[2:t] - dat$x1[1:(t - 1)])
# prepare data in HMSC format
Y <- as.matrix(df$Nt1)
XData <- data.frame(cbind(E[1:(t-1)],E[1:(t-1)]^2,abs(dat$x1[2:t] - dat$x1[1:(t-1)])))
colnames(XData) <- c("E","Esq","dx1")

studyDesign = data.frame(sample = as.factor(1:(t-1)))
rL = HmscRandomLevel(units = studyDesign$sample)

m.4.hmsc = Hmsc(Y = Y, XData = XData, XFormula = ~E+Esq+dx1, studyDesign = studyDesign, ranLevels = list(sample = rL))
# Bayesian model parameters
nChains <- 2
thin <- 5
samples <- 2000
transient <- 1000*thin
verbose <- 500*thin
# sample MCMC
m.4.sample <- sampleMcmc(m.4.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
m4.post.hmsc <- convertToCodaObject(m.4.sample)
```

```{r warning=FALSE}
VP4 <- computeVariancePartitioning(m.4.sample,group=c(1,1,1,2),groupnames=c("Env","x1"))
plotVariancePartitioning(m.4.sample,VP4,cols=c("white","skyblue","darkgrey"),args.legend=list(cex=0.75,bg="transparent"))
```
```{r}
# prepare for plot
pred2 <- predict(m.4.sample)
hold1 <- list()
hold2 <- list()
for(i in 1:length(N0)){
  hold1[[i]] <- as.numeric(Y[,i])
  data <- lapply(pred2, function(x) x[, colnames(x)[i], drop = FALSE])
  hold2[[i]] <- matrix(unlist(data), nrow=length(pred2), byrow=TRUE)
}
y <- unlist(hold1) # vector of each species' observed N across time points
yrep <- as.matrix(data.frame(hold2))
### Mod2 plot ###
par(mgp=c(2,0.45,0), tcl=-0.4, mar=c(1.3,1.2,0,0))
color_scheme_set("brightblue")
#ppc_dens_overlay(y, yrep[1:50, ])
ppc_intervals(y,yrep,x=dat$time[1:(t-1)],prob = 0.95) +
  labs(x = "time",y = "ln(N)",) +
  theme(plot.margin = unit(c(0, 0, 0, 0), "cm")) +
  theme(legend.position = "none")
```
```{r}
### Model 5, h2=0.04 ###
dat <- cbind(log(gdat$N1[gdat$h2==0.04]),gdat$x1[gdat$h2==0.04])
dat <- as.data.frame(dat)
colnames(dat) <- c("N1","x1")
dat$time <- 1:t
dat <- as.data.frame(dat)
df <- data.frame(dat[(2:t),-3])
colnames(df) <- c("Nt1", "xt1")
df$dx1 <- abs(dat$x1[2:t] - dat$x1[1:(t - 1)])
# prepare data in HMSC format
Y <- as.matrix(df$Nt1)
XData <- data.frame(cbind(E[1:(t-1)],E[1:(t-1)]^2,abs(dat$x1[2:t] - dat$x1[1:(t-1)])))
colnames(XData) <- c("E","Esq","dx1")

studyDesign = data.frame(sample = as.factor(1:(t-1)))
rL = HmscRandomLevel(units = studyDesign$sample)

m.5.hmsc = Hmsc(Y = Y, XData = XData, XFormula = ~E+Esq+dx1, studyDesign = studyDesign, ranLevels = list(sample = rL))
# Bayesian model parameters
nChains <- 2
thin <- 5
samples <- 2000
transient <- 1000*thin
verbose <- 500*thin
# sample MCMC
m.5.sample <- sampleMcmc(m.5.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
m5.post.hmsc <- convertToCodaObject(m.5.sample)
```

```{r warning=FALSE}
VP5 <- computeVariancePartitioning(m.5.sample,group=c(1,1,1,2),groupnames=c("Env","x1"))
plotVariancePartitioning(m.5.sample,VP5,cols=c("white","skyblue","darkgrey"),args.legend=list(cex=0.75,bg="transparent"))
```
```{r}
# prepare for plot
pred2 <- predict(m.5.sample)
hold1 <- list()
hold2 <- list()
for(i in 1:length(N0)){
  hold1[[i]] <- as.numeric(Y[,i])
  data <- lapply(pred2, function(x) x[, colnames(x)[i], drop = FALSE])
  hold2[[i]] <- matrix(unlist(data), nrow=length(pred2), byrow=TRUE)
}
y <- unlist(hold1) # vector of each species' observed N across time points
yrep <- as.matrix(data.frame(hold2))
### Mod2 plot ###
par(mgp=c(2,0.45,0), tcl=-0.4, mar=c(1.3,1.2,0,0))
color_scheme_set("brightblue")
#ppc_dens_overlay(y, yrep[1:50, ])
ppc_intervals(y,yrep,x=dat$time[1:(t-1)],prob = 0.95) +
  labs(x = "time",y = "ln(N)",) +
  theme(plot.margin = unit(c(0, 0, 0, 0), "cm")) +
  theme(legend.position = "none")
```
```{r}
### Model 6, h2=0.05 ###
dat <- cbind(log(gdat$N1[gdat$h2==0.05]),gdat$x1[gdat$h2==0.05])
dat <- as.data.frame(dat)
colnames(dat) <- c("N1","x1")
dat$time <- 1:t
dat <- as.data.frame(dat)
df <- data.frame(dat[(2:t),-3])
colnames(df) <- c("Nt1", "xt1")
df$dx1 <- abs(dat$x1[2:t] - dat$x1[1:(t - 1)])
# prepare data in HMSC format
Y <- as.matrix(df$Nt1)
XData <- data.frame(cbind(E[1:(t-1)],E[1:(t-1)]^2,abs(dat$x1[2:t] - dat$x1[1:(t-1)])))
colnames(XData) <- c("E","Esq","dx1")

studyDesign = data.frame(sample = as.factor(1:(t-1)))
rL = HmscRandomLevel(units = studyDesign$sample)

m.6.hmsc = Hmsc(Y = Y, XData = XData, XFormula = ~E+Esq+dx1, studyDesign = studyDesign, ranLevels = list(sample = rL))
# Bayesian model parameters
nChains <- 2
thin <- 5
samples <- 2000
transient <- 1000*thin
verbose <- 500*thin
# sample MCMC
m.6.sample <- sampleMcmc(m.6.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
m6.post.hmsc <- convertToCodaObject(m.6.sample)
```

```{r warning=FALSE}
VP6 <- computeVariancePartitioning(m.6.sample,group=c(1,1,1,2),groupnames=c("Env","x1"))
plotVariancePartitioning(m.6.sample,VP6,cols=c("white","skyblue","darkgrey"),args.legend=list(cex=0.75,bg="transparent"))
```
```{r}
# prepare for plot
pred2 <- predict(m.6.sample)
hold1 <- list()
hold2 <- list()
for(i in 1:length(N0)){
  hold1[[i]] <- as.numeric(Y[,i])
  data <- lapply(pred2, function(x) x[, colnames(x)[i], drop = FALSE])
  hold2[[i]] <- matrix(unlist(data), nrow=length(pred2), byrow=TRUE)
}
y <- unlist(hold1) # vector of each species' observed N across time points
yrep <- as.matrix(data.frame(hold2))
### Mod2 plot ###
par(mgp=c(2,0.45,0), tcl=-0.4, mar=c(1.3,1.2,0,0))
color_scheme_set("brightblue")
#ppc_dens_overlay(y, yrep[1:50, ])
ppc_intervals(y,yrep,x=dat$time[1:(t-1)],prob = 0.95) +
  labs(x = "time",y = "ln(N)",) +
  theme(plot.margin = unit(c(0, 0, 0, 0), "cm")) +
  theme(legend.position = "none")
```

```{r}
### Model 7, h2=0.075 ###
dat <- cbind(log(gdat$N1[gdat$h2==0.075]),gdat$x1[gdat$h2==0.075])
dat <- as.data.frame(dat)
colnames(dat) <- c("N1","x1")
dat$time <- 1:t
dat <- as.data.frame(dat)
df <- data.frame(dat[(2:t),-3])
colnames(df) <- c("Nt1", "xt1")
df$dx1 <- abs(dat$x1[2:t] - dat$x1[1:(t - 1)])
# prepare data in HMSC format
Y <- as.matrix(df$Nt1)
XData <- data.frame(cbind(E[1:(t-1)],E[1:(t-1)]^2,abs(dat$x1[2:t] - dat$x1[1:(t-1)])))
colnames(XData) <- c("E","Esq","dx1")

studyDesign = data.frame(sample = as.factor(1:(t-1)))
rL = HmscRandomLevel(units = studyDesign$sample)

m.7.hmsc = Hmsc(Y = Y, XData = XData, XFormula = ~E+Esq+dx1, studyDesign = studyDesign, ranLevels = list(sample = rL))
# Bayesian model parameters
nChains <- 2
thin <- 5
samples <- 2000
transient <- 1000*thin
verbose <- 500*thin
# sample MCMC
m.7.sample <- sampleMcmc(m.7.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
m7.post.hmsc <- convertToCodaObject(m.7.sample)
```

```{r warning=FALSE}
VP7 <- computeVariancePartitioning(m.7.sample,group=c(1,1,1,2),groupnames=c("Env","x1"))
plotVariancePartitioning(m.7.sample,VP7,cols=c("white","skyblue","darkgrey"),args.legend=list(cex=0.75,bg="transparent"))
```
```{r}
# prepare for plot
pred2 <- predict(m.7.sample)
hold1 <- list()
hold2 <- list()
for(i in 1:length(N0)){
  hold1[[i]] <- as.numeric(Y[,i])
  data <- lapply(pred2, function(x) x[, colnames(x)[i], drop = FALSE])
  hold2[[i]] <- matrix(unlist(data), nrow=length(pred2), byrow=TRUE)
}
y <- unlist(hold1) # vector of each species' observed N across time points
yrep <- as.matrix(data.frame(hold2))
### Mod2 plot ###
par(mgp=c(2,0.45,0), tcl=-0.4, mar=c(1.3,1.2,0,0))
color_scheme_set("brightblue")
#ppc_dens_overlay(y, yrep[1:50, ])
ppc_intervals(y,yrep,x=dat$time[1:(t-1)],prob = 0.95) +
  labs(x = "time",y = "ln(N)",) +
  theme(plot.margin = unit(c(0, 0, 0, 0), "cm")) +
  theme(legend.position = "none")
```
```{r}
### Model 8, h2=0.1 ###
dat <- cbind(log(gdat$N1[gdat$h2==0.1]),gdat$x1[gdat$h2==0.1])
dat <- as.data.frame(dat)
colnames(dat) <- c("N1","x1")
dat$time <- 1:t
dat <- as.data.frame(dat)
df <- data.frame(dat[(2:t),-3])
colnames(df) <- c("Nt1", "xt1")
df$dx1 <- abs(dat$x1[2:t] - dat$x1[1:(t - 1)])
# prepare data in HMSC format
Y <- as.matrix(df$Nt1)
XData <- data.frame(cbind(E[1:(t-1)],E[1:(t-1)]^2,abs(dat$x1[2:t] - dat$x1[1:(t-1)])))
colnames(XData) <- c("E","Esq","dx1")

studyDesign = data.frame(sample = as.factor(1:(t-1)))
rL = HmscRandomLevel(units = studyDesign$sample)

m.8.hmsc = Hmsc(Y = Y, XData = XData, XFormula = ~E+Esq+dx1, studyDesign = studyDesign, ranLevels = list(sample = rL))
# Bayesian model parameters
nChains <- 2
thin <- 5
samples <- 2000
transient <- 1000*thin
verbose <- 500*thin
# sample MCMC
m.8.sample <- sampleMcmc(m.8.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
m8.post.hmsc <- convertToCodaObject(m.8.sample)
```

```{r warning=FALSE}
VP8 <- computeVariancePartitioning(m.8.sample,group=c(1,1,1,2),groupnames=c("Env","x1"))
plotVariancePartitioning(m.8.sample,VP8,cols=c("white","skyblue","darkgrey"),args.legend=list(cex=0.75,bg="transparent"))
```
```{r}
# prepare for plot
pred2 <- predict(m.8.sample)
hold1 <- list()
hold2 <- list()
for(i in 1:length(N0)){
  hold1[[i]] <- as.numeric(Y[,i])
  data <- lapply(pred2, function(x) x[, colnames(x)[i], drop = FALSE])
  hold2[[i]] <- matrix(unlist(data), nrow=length(pred2), byrow=TRUE)
}
y <- unlist(hold1) # vector of each species' observed N across time points
yrep <- as.matrix(data.frame(hold2))
### Mod2 plot ###
par(mgp=c(2,0.45,0), tcl=-0.4, mar=c(1.3,1.2,0,0))
color_scheme_set("brightblue")
#ppc_dens_overlay(y, yrep[1:50, ])
ppc_intervals(y,yrep,x=dat$time[1:(t-1)],prob = 0.95) +
  labs(x = "time",y = "ln(N)",) +
  theme(plot.margin = unit(c(0, 0, 0, 0), "cm")) +
  theme(legend.position = "none")
```
```{r}
### Model 9, h2=0.4 ###
dat <- cbind(log(gdat$N1[gdat$h2==0.4]),gdat$x1[gdat$h2==0.4])
dat <- as.data.frame(dat)
colnames(dat) <- c("N1","x1")
dat$time <- 1:t
dat <- as.data.frame(dat)
df <- data.frame(dat[(2:t),-3])
colnames(df) <- c("Nt1", "xt1")
df$dx1 <- abs(dat$x1[2:t] - dat$x1[1:(t - 1)])
# prepare data in HMSC format
Y <- as.matrix(df$Nt1)
XData <- data.frame(cbind(E[1:(t-1)],E[1:(t-1)]^2,abs(dat$x1[2:t] - dat$x1[1:(t-1)])))
colnames(XData) <- c("E","Esq","dx1")

studyDesign = data.frame(sample = as.factor(1:(t-1)))
rL = HmscRandomLevel(units = studyDesign$sample)

m.9.hmsc = Hmsc(Y = Y, XData = XData, XFormula = ~E+Esq+dx1, studyDesign = studyDesign, ranLevels = list(sample = rL))
# Bayesian model parameters
nChains <- 2
thin <- 5
samples <- 2000
transient <- 1000*thin
verbose <- 500*thin
# sample MCMC
m.9.sample <- sampleMcmc(m.9.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
m9.post.hmsc <- convertToCodaObject(m.9.sample)
```

```{r warning=FALSE}
VP9 <- computeVariancePartitioning(m.9.sample,group=c(1,1,1,2),groupnames=c("Env","x1"))
plotVariancePartitioning(m.9.sample,VP9,cols=c("white","skyblue","darkgrey"),args.legend=list(cex=0.75,bg="transparent"))
```
```{r}
# prepare for plot
pred2 <- predict(m.9.sample)
hold1 <- list()
hold2 <- list()
for(i in 1:length(N0)){
  hold1[[i]] <- as.numeric(Y[,i])
  data <- lapply(pred2, function(x) x[, colnames(x)[i], drop = FALSE])
  hold2[[i]] <- matrix(unlist(data), nrow=length(pred2), byrow=TRUE)
}
y <- unlist(hold1) # vector of each species' observed N across time points
yrep <- as.matrix(data.frame(hold2))
### Mod2 plot ###
par(mgp=c(2,0.45,0), tcl=-0.4, mar=c(1.3,1.2,0,0))
color_scheme_set("brightblue")
#ppc_dens_overlay(y, yrep[1:50, ])
ppc_intervals(y,yrep,x=dat$time[1:(t-1)],prob = 0.95) +
  labs(x = "time",y = "ln(N)",) +
  theme(plot.margin = unit(c(0, 0, 0, 0), "cm")) +
  theme(legend.position = "none")
```
```{r}
### Model 10, h2=0.9 ###
dat <- cbind(log(gdat$N1[gdat$h2==0.9]),gdat$x1[gdat$h2==0.9])
dat <- as.data.frame(dat)
colnames(dat) <- c("N1","x1")
dat$time <- 1:t
dat <- as.data.frame(dat)
df <- data.frame(dat[(2:t),-3])
colnames(df) <- c("Nt1", "xt1")
df$dx1 <- abs(dat$x1[2:t] - dat$x1[1:(t - 1)])
# prepare data in HMSC format
Y <- as.matrix(df$Nt1)
XData <- data.frame(cbind(E[1:(t-1)],E[1:(t-1)]^2,abs(dat$x1[2:t] - dat$x1[1:(t-1)])))
colnames(XData) <- c("E","Esq","dx1")

studyDesign = data.frame(sample = as.factor(1:(t-1)))
rL = HmscRandomLevel(units = studyDesign$sample)

m.10.hmsc = Hmsc(Y = Y, XData = XData, XFormula = ~E+Esq+dx1, studyDesign = studyDesign, ranLevels = list(sample = rL))
# Bayesian model parameters
nChains <- 2
thin <- 5
samples <- 2000
transient <- 1000*thin
verbose <- 500*thin
# sample MCMC
m.10.sample <- sampleMcmc(m.10.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
m10.post.hmsc <- convertToCodaObject(m.10.sample)
```

```{r warning=FALSE}
VP10 <- computeVariancePartitioning(m.10.sample,group=c(1,1,1,2),groupnames=c("Env","x1"))
plotVariancePartitioning(m.10.sample,VP10,cols=c("white","skyblue","darkgrey"),args.legend=list(cex=0.75,bg="transparent"))
```
```{r}
# prepare for plot
pred2 <- predict(m.10.sample)
hold1 <- list()
hold2 <- list()
for(i in 1:length(N0)){
  hold1[[i]] <- as.numeric(Y[,i])
  data <- lapply(pred2, function(x) x[, colnames(x)[i], drop = FALSE])
  hold2[[i]] <- matrix(unlist(data), nrow=length(pred2), byrow=TRUE)
}
y <- unlist(hold1) # vector of each species' observed N across time points
yrep <- as.matrix(data.frame(hold2))
### Mod2 plot ###
par(mgp=c(2,0.45,0), tcl=-0.4, mar=c(1.3,1.2,0,0))
color_scheme_set("brightblue")
#ppc_dens_overlay(y, yrep[1:50, ])
ppc_intervals(y,yrep,x=dat$time[1:(t-1)],prob = 0.95) +
  labs(x = "time",y = "ln(N)",) +
  theme(plot.margin = unit(c(0, 0, 0, 0), "cm")) +
  theme(legend.position = "none")
```
```{r}
### Model 11, h2=1 ###
dat <- cbind(log(gdat$N1[gdat$h2==1]),gdat$x1[gdat$h2==1])
dat <- as.data.frame(dat)
colnames(dat) <- c("N1","x1")
dat$time <- 1:t
dat <- as.data.frame(dat)
df <- data.frame(dat[(2:t),-3])
colnames(df) <- c("Nt1", "xt1")
df$dx1 <- abs(dat$x1[2:t] - dat$x1[1:(t - 1)])
# prepare data in HMSC format
Y <- as.matrix(df$Nt1)
XData <- data.frame(cbind(E[1:(t-1)],E[1:(t-1)]^2,abs(dat$x1[2:t] - dat$x1[1:(t-1)])))
colnames(XData) <- c("E","Esq","dx1")

studyDesign = data.frame(sample = as.factor(1:(t-1)))
rL = HmscRandomLevel(units = studyDesign$sample)

m.11.hmsc = Hmsc(Y = Y, XData = XData, XFormula = ~E+Esq+dx1, studyDesign = studyDesign, ranLevels = list(sample = rL))
# Bayesian model parameters
nChains <- 2
thin <- 5
samples <- 2000
transient <- 1000*thin
verbose <- 500*thin
# sample MCMC
m.11.sample <- sampleMcmc(m.11.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
m11.post.hmsc <- convertToCodaObject(m.11.sample)
```

```{r warning=FALSE}
VP11 <- computeVariancePartitioning(m.11.sample,group=c(1,1,1,2),groupnames=c("Env","x1"))
plotVariancePartitioning(m.11.sample,VP11,cols=c("white","skyblue","darkgrey"),args.legend=list(cex=0.75,bg="transparent"))
```
```{r}
# prepare for plot
pred2 <- predict(m.11.sample)
hold1 <- list()
hold2 <- list()
for(i in 1:length(N0)){
  hold1[[i]] <- as.numeric(Y[,i])
  data <- lapply(pred2, function(x) x[, colnames(x)[i], drop = FALSE])
  hold2[[i]] <- matrix(unlist(data), nrow=length(pred2), byrow=TRUE)
}
y <- unlist(hold1) # vector of each species' observed N across time points
yrep <- as.matrix(data.frame(hold2))
### Mod2 plot ###
par(mgp=c(2,0.45,0), tcl=-0.4, mar=c(1.3,1.2,0,0))
color_scheme_set("brightblue")
#ppc_dens_overlay(y, yrep[1:50, ])
ppc_intervals(y,yrep,x=dat$time[1:(t-1)],prob = 0.95) +
  labs(x = "time",y = "ln(N)",) +
  theme(plot.margin = unit(c(0, 0, 0, 0), "cm")) +
  theme(legend.position = "none")
```

```{r}
mc_dat_1 <- mcmc_intervals_data(m1.post.hmsc$Beta, pars="B[dx1 (C4), sp1 (S1)]")
mc_dat_2 <- mcmc_intervals_data(m2.post.hmsc$Beta, pars="B[dx1 (C4), sp1 (S1)]")
mc_dat_3 <- mcmc_intervals_data(m3.post.hmsc$Beta, pars="B[dx1 (C4), sp1 (S1)]")
mc_dat_4 <- mcmc_intervals_data(m4.post.hmsc$Beta, pars="B[dx1 (C4), sp1 (S1)]")
mc_dat_5 <- mcmc_intervals_data(m5.post.hmsc$Beta, pars="B[dx1 (C4), sp1 (S1)]")
mc_dat_6 <- mcmc_intervals_data(m6.post.hmsc$Beta, pars="B[dx1 (C4), sp1 (S1)]")
mc_dat_7 <- mcmc_intervals_data(m7.post.hmsc$Beta, pars="B[dx1 (C4), sp1 (S1)]")
mc_dat_8 <- mcmc_intervals_data(m8.post.hmsc$Beta, pars="B[dx1 (C4), sp1 (S1)]")
mc_dat_9 <- mcmc_intervals_data(m9.post.hmsc$Beta, pars="B[dx1 (C4), sp1 (S1)]")
mc_dat_10 <- mcmc_intervals_data(m10.post.hmsc$Beta, pars="B[dx1 (C4), sp1 (S1)]")
mc_dat_11 <- mcmc_intervals_data(m11.post.hmsc$Beta, pars="B[dx1 (C4), sp1 (S1)]")

combined <- rbind(mc_dat_1,mc_dat_2,mc_dat_3,mc_dat_4,mc_dat_5,mc_dat_6,mc_dat_7,mc_dat_8,mc_dat_9,mc_dat_10,mc_dat_11)
combined$model <- rep(c("0","0.01","0.02","0.03","0.04","0.05","0.075","0.1","0.4","0.9","1"),times=c(dim(mc_dat_1)[1],dim(mc_dat_2)[1],dim(mc_dat_3)[1],dim(mc_dat_4)[1],dim(mc_dat_5)[1],dim(mc_dat_6)[1],dim(mc_dat_7)[1],dim(mc_dat_8)[1],dim(mc_dat_9)[1],dim(mc_dat_10)[1],dim(mc_dat_11)[1]))

# make the plot using ggplot 
theme_set(bayesplot::theme_default())
pos <- position_nudge(y = c(0,0.1,0.2,0.3,0.4,0.5,0.6,0.7,0.8,0.9,1))
ggplot(combined, aes(x = m, y = parameter, color = model)) + 
  geom_linerange(aes(xmin = l, xmax = h), position = pos, size=2)+
  geom_linerange(aes(xmin = ll, xmax = hh), position = pos)+
  geom_point(aes(color=model),position = pos)

ggplot(combined, aes(x = m, y = parameter, color = model)) + 
  geom_linerange(aes(xmin = l, xmax = h), position = pos, size=2)+
  geom_linerange(aes(xmin = ll, xmax = hh), position = pos)+
  geom_point(aes(color=model),position = pos)+
  coord_flip()+
  geom_vline(xintercept=0,linetype="dashed")
```

# References

### 2024-10-29 code/R/3_main_v02.R

@@ -1,0 +1,161 @@
+# Title. main_v02.R
+# Author. J.H. Pantel
+# Date 28-10-2024
+# Description. This script contains analyses for the HMSC_ecoevo project and associated manuscript Hermann & Pantel 202x.
+
+#### Libraries ---------------------------------------------------------------
+library(Hmsc)
+library(tictoc)
+
+#### Step 1. Run MATLAB simulations --------------
+
+### Scenario 1. ###
+# Weak selection, fixed environment, all dispersal and heritability levels
+# P = 1, w = 10, v[E] = 0
+# /code/R/metacom_evol_sim_same_init_v01.m
+
+### Scenario 2. ###
+# Strong selection, fixed environment, d = 10^-3, h2 = 0.1
+# P = 0.1, w = 0.2, v[E] = 0
+# HMSC_ecoevo/code/simulation/metacom_evol_sim_same_init_v02.m
+
+### Scenario 3. ###
+# Weak selection, variable environment, d = 10^-3, h2 = 0.1
+# P = 1, w = 10, v[E] = 0.01
+# HMSC_ecoevo/code/simulation/metacom_evol_sim_same_init_v03.m
+
+### Scenario 4. ###
+# Strong selection, variable environment, d = 10^-3, h2 = 0.1
+# P = 0.1, w = 0.2, v[E] = 0.01
+# HMSC_ecoevo/code/simulation/metacom_evol_sim_same_init_v04.m
+
+#### Step 2. Prepare for HMSC analyses --------------
+d_lev <- c("d_zero","d_minus9","d_minus8","d_minus7","d_minus6","d_minus5","d_minus4","d_minus3","d_minus2","d_01","d_02","d_03","d_04","d_05","d_06","d_07","d_08","d_09","d_10")
+h_lev <- c("h_0_","h_01_","h_02_","h_03_","h_04_","h_05_","h_06_","h_07_","h_08_","h_09_","h_10_")
+#setwd("./data")
+
+# HMSC for single results condition
+z <- 2
+f <- 4
+
+print(paste(h_lev[z],d_lev[f],sep=""))
+result <- paste(h_lev[z],d_lev[f],sep="")
+load(paste("/Users/jhpantel/Nextcloud/Pantel/data/",result,"_res_v02.RData",sep="")) # Scenario 1
+# res  <- R.matlab::readMat(paste(result,"_res_v02.mat",sep="")) # Scenario 2
[…]

### 2024-10-29 code/R/7_metacom_sim_v02.R

@@ -4,6 +4,8 @@
 ## Scenario 1. across d and h2 levels
 ## # P = 1, w = 2, v[E] = 0
   
+library(vegan)
+library(ggplot2)
 set.seed(1)
 
 ###################################
@@ -15,22 +17,30 @@ set.seed(1)
 # x = rlist (list of randomly generated numbers (runif) for each individual in a patch of species i)
 # i = species number (is part of the loop for dispersal)
 check <- function(x){
-    x <=d[i]
+  x <=d[i]
 }
 
-# This function is the randome sample to chose amone sites where the species can immigrate,
+# This function is the random sample to chose among sites where the species can immigrate,
 # weighted by connectivitx.
 # It takes simply a vector of all patches in, so:
 # x = site/1:50 and i = species number (is part of the loop for dispersal)
 disp_patch <- function(x){
-    sample(x=site[-x],size=mig_count[x],replace=T,prob=connec[i,-x,x])
+  sample(x=site[-x],size=mig_count[x],replace=T,prob=connec[i,-x,x])
 }
-#Function to update the trait value, where the environmental value is substracted
+#Function to update the trait value, where the environmental value is subtracted
 #by the distance of the trait (which is changed by evo)
 #x = The number of the species
 #xt_up = the updated trait value, which is saved into xt
 up_trait <- function(x){
   xt_up<-E-dt[,x,t]
+}
+
+#Function to update the trait distance value, to the optimal value
+#by the new E value
+#x = The number of the species
+#dt_up = the updated trait distance value, which is saved into dt
+up_dist <- function(x){
+  dt_up <- E-xt[,x,t-1]
 }
 
 ###################################
@@ -46,7 +56,9 @@ Wmax = 2    # Maximum population growth rate for all species
 Wt = array(numeric(),c(patch,spec,time))   # Record of fitness vlaues over time, patch x spec x time
 dt = array(numeric(),c(patch,spec,time))   # Record of distance from optimum value over time, patch x spec x time
[…]

### 2024-10-29 code/R/1_metacom_sim_v01.R
New branch for updating the script!

@@ -170,10 +170,10 @@ d = array(d_vals[8],c(1,spec))  #which d_value do you want to use? - just using 
 ## Step 3. Evolving metacommunity simulations    ##
 ###################################################
 
-for (t in 2:200) {
+for (t in 2:time) {
     t
     #E fluctuation, here with directional change (with the abs() function)
-    if (t%%10==0) {E = E + abs(rnorm(50,0,var))}
+    if (t%%10==0) {E = E + rnorm(50,0,var)}
     Et[,t] = E
     
     pop = N[,,t] # Write this generation's population size into a temporary record 
@@ -306,11 +306,11 @@ for (t in 2:200) {
 
 #### Step 4. Save results
 #A small extra step, if the simulation ran for only 200 time steps
-dt <- dt[,,1:200]
-N <- N[,,1:200]
-Wt <- Wt[,,1:200]
-xt <- xt[,,1:200]
-Et <- Et[,1:200]
+#dt <- dt[,,1:200]
+#N <- N[,,1:200]
+#Wt <- Wt[,,1:200]
+#xt <- xt[,,1:200]
+#Et <- Et[,1:200]
 
 d_lev <- c("d_zero","d_minus9","d_minus8","d_minus7","d_minus6","d_minus5","d_minus4","d_minus3","d_minus2","d_01","d_02","d_03","d_04","d_05","d_06","d_07","d_08","d_09","d_10")
 h_lev <- c("h_0_","h_01_","h_02_","h_03_","h_04_","h_05_","h_06_","h_07_","h_08_","h_09_","h_10_")
@@ -318,4 +318,4 @@ z <- 2
 f <- 8
 print(paste(h2_names[z],d_names[f],sep=""))
 result <- paste(h_lev[z],d_lev[f],sep="")
-save.image(paste("././data/",result,"_res_short_Edirect_sd0.1.RData",sep=""))
+save.image(paste("././data/",result,"_res_short_E_var0.1.RData",sep=""))

### 2024-10-29 code/R/1_metacom_sim_v01.R
Updated 3_main script

@@ -47,6 +47,7 @@ up_dist <- function(x){
 patch = 50   # Number of patches in the landscape
 spec = 15    # Number of species in the community
 time = 1000  # Number of time steps to run the simulation
+cut = 1000   # Number when to stop the simulation early
 N = array(numeric(),c(patch,spec,time))   # Record of abundances of all species over time patch x spec x time
 P = array(1,c(1,spec))    # Phenotypic variances for all species
 w = array(2,c(1,spec))   # Width of fitness curve
@@ -173,7 +174,7 @@ d = array(d_vals[8],c(1,spec))  #which d_value do you want to use? - just using 
 for (t in 2:time) {
     t
     #E fluctuation, here with directional change (with the abs() function)
-    if (t%%10==0) {E = E + rnorm(50,0,var)}
+    if (t%%10==0) {E = E + rnorm(patch,0,var)}
     Et[,t] = E
     
     pop = N[,,t] # Write this generation's population size into a temporary record 
@@ -297,6 +298,7 @@ for (t in 2:time) {
             Wt[A[z,1],A[z,2],t] = NA
         }
     }
+    if (t >= cut) {break}
 }
 
 ## use randsample to randomly sample among the available sites (the other 49) with weights according to connectivity
@@ -306,11 +308,11 @@ for (t in 2:time) {
 
 #### Step 4. Save results
 #A small extra step, if the simulation ran for only 200 time steps
-#dt <- dt[,,1:200]
-#N <- N[,,1:200]
-#Wt <- Wt[,,1:200]
-#xt <- xt[,,1:200]
-#Et <- Et[,1:200]
+dt <- dt[,,1:cut]
+N <- N[,,1:cut]
+Wt <- Wt[,,1:cut]
+xt <- xt[,,1:cut]
+Et <- Et[,1:cut]
 
 d_lev <- c("d_zero","d_minus9","d_minus8","d_minus7","d_minus6","d_minus5","d_minus4","d_minus3","d_minus2","d_01","d_02","d_03","d_04","d_05","d_06","d_07","d_08","d_09","d_10")
 h_lev <- c("h_0_","h_01_","h_02_","h_03_","h_04_","h_05_","h_06_","h_07_","h_08_","h_09_","h_10_")

### 2024-10-29 code/R/3_main_v01.R

@@ -53,10 +53,17 @@ d <- b[c] # generation of last occurrence for species that go extinct
 
 # Arrange all data for HMSC
 ## site-by-species abundance data as a time series
-Y <- array(NA,dim=c(50000,15),dimnames=list(NULL,c("y1","y2","y3","y4","y5","y6","y7","y8","y9","y10","y11","y12","y13","y14","y15")))
-count <- seq(1,50000,by=50)
-for(i in 1:1000){
-  Y[(count[i]:(count[i]+49)),] <- N[,,i]
+spec_name = NULL
+trait_name = NULL
+for (i in 1:spec){
+  spec_name[i] = paste("y",i,"",sep="")
+  trait_name[i] = paste("x",i,"",sep="")
+}
+
+Y <- array(NA,dim=c((patch*cut),spec),dimnames=list(NULL,spec_name))
+count <- seq(1,(patch*cut),by=patch)
+for(i in 1:cut){
+  Y[(count[i]:(count[i]+(patch-1))),] <- N[,,i]
 }
 
 ## site-by-species trait data as a time series
@@ -66,25 +73,25 @@ trait_init[trait_init == 9999] <- NaN
 trait_x[,,1] <- trait_init
 ## site-by-species CHANGE in trait value as a time series
 ## absolute value of change in trait value
-delta_x <- array(NA,c(50,15,999),dimnames=list(NULL,c("x1","x2","x3","x4","x5","x6","x7","x8","x9","x10","x11","x12","x13","x14","x15"),NULL))
-for(i in 2:time){
+delta_x <- array(NA,c(patch,spec,(cut-1)),dimnames=list(NULL,trait_name,NULL))
+for(i in 2:cut){
   delta_x[,,i-1] <- abs(trait_x[,,i] - trait_x[,,i-1])
 }
-delta_x_time <- array(NA,dim=c(49950,15),dimnames=list(NULL,c("x1","x2","x3","x4","x5","x6","x7","x8","x9","x10","x11","x12","x13","x14","x15")))
-count <- seq(1,49950,by=50)
-for(i in 1:999){
-  delta_x_time[(count[i]:(count[i]+49)),] <- delta_x[,,i]
+delta_x_time <- array(NA,dim=c((patch*cut-patch),spec),dimnames=list(NULL,trait_name))
+count <- seq(1,(patch*cut-patch),by=patch)
+for(i in 1:(cut-1)){
[…]

### 2024-10-30 code/R/3_main_v01.R
Changed code according to review on E --> Et

@@ -110,12 +110,16 @@ rL.time <- Hmsc::HmscRandomLevel(units = unique(Random$year))
 rL.time = Hmsc::setPriors(rL.time,nfMin=1,nfMax=5)
 
 ## Fixed effects: environment
-env_hold <- array(NA,dim=c(dim(Y)[1],dim(E)[2]))
-for(i in 1:dim(E)[2]){ # for each environmental variable
-  env_hold[,i] <- rep(E[,i],(cut-1))
-  colnames(env_hold)[i] <- paste("env",i,sep="")
+env_names = NULL
+for (i in 1:dim(E)[2]){
+  env_names[i] = paste("env",i,"",sep="")
 }
-
+env_hold <- array(NA,dim=c(dim(Y)[1],1),dimnames = list(NULL,env_names))
+count <- seq(1,(patch*cut-patch),by=patch)
+for(i in 1:(cut-1)){ # for each environmental variable
+  env_hold[(count[i]:(count[i]+(patch-1))),] <- Et[,i+1]
+  
+}
 ## Interaction effects: site abundance the year before X change in species trait value
 X_delta_x <- X * delta_x_time
 colnames(X_delta_x) <- paste("x",colnames(X_delta_x),sep="")

### 2024-10-30 code/R/3_main_v01.R
Change 1 to dim(E)[2]

@@ -114,7 +114,7 @@ env_names = NULL
 for (i in 1:dim(E)[2]){
   env_names[i] = paste("env",i,"",sep="")
 }
-env_hold <- array(NA,dim=c(dim(Y)[1],1),dimnames = list(NULL,env_names))
+env_hold <- array(NA,dim=c(dim(Y)[1],dim(E)[2]),dimnames = list(NULL,env_names))
 count <- seq(1,(patch*cut-patch),by=patch)
 for(i in 1:(cut-1)){ # for each environmental variable
   env_hold[(count[i]:(count[i]+(patch-1))),] <- Et[,i+1]

### 2024-11-01 code/R/7_metacom_sim_v02.R
Re-run metacom sim, updated to main

@@ -49,6 +49,7 @@ up_dist <- function(x){
 patch = 50   # Number of patches in the landscape
 spec = 15    # Number of species in the community
 time = 1000  # Number of time steps to run the simulation
+cut = 200   # Number when to stop the simulation early
 N = array(numeric(),c(patch,spec,time))   # Record of abundances of all species over time patch x spec x time
 P = array(1,c(1,spec))    # Phenotypic variances for all species
 w = array(2,c(1,spec))   # Width of fitness curve
@@ -58,7 +59,7 @@ dt = array(numeric(),c(patch,spec,time))   # Record of distance from optimum val
 xt = array(9999,c(patch,spec,time))     # Record of phenotype value over time, patch x spec x time
 Et = array(numeric(),c(patch,time))   # Record of environmental value over time, patch x spec x time
 
-var = 0.1 #Adding variation to the environmental value for no variation set to 0 (var is the sd of the normal distribution with a mean of 0)
+var = 0 #Adding variation to the environmental value for no variation set to 0 (var is the sd of the normal distribution with a mean of 0)
 ## Heritability
 # Scenario 1: No evolution
 #h2 = array(0,c(1,spec))
@@ -180,7 +181,7 @@ for(v.d in 1:length(d_vals)){
     ## Step 3. Evolving metacommunity simulations    ##
     ###################################################
     
-    for (t in 2:200) {
+    for (t in 2:time) {
       t
       #E fluctuation, here with directional change (with the abs() function)
       if (t%%10==0) {E = E + abs(rnorm(50,0,var))}
@@ -301,20 +302,21 @@ for(v.d in 1:length(d_vals)){
       ## Check to see if population trait information remains for extinct populations
       A = which(pop==0 & N[,,t-1]>0,arr.ind=T)
       if(nrow(A)>0){
-        for (z in 1:nrow(A)) { # For each species / site that went extinct for other reasons
-          xt[A[z,1],A[z,2],t] = NA
-          dt[A[z,1],A[z,2],t] = NA
-          Wt[A[z,1],A[z,2],t] = NA
+        for (za in 1:nrow(A)) { # For each species / site that went extinct for other reasons
+          xt[A[za,1],A[za,2],t] = NA
[…]

### 2024-11-01 code/R/3_main_v02.R
Another z to zb edit, re-ran all sims

@@ -36,7 +36,7 @@ h_lev <- c("h_0_","h_01_","h_02_","h_03_","h_04_","h_05_","h_06_","h_07_","h_08_
 
 # HMSC for single results condition
 z <- 2
-f <- 4
+f <- 8
 
 print(paste(h_lev[z],d_lev[f],sep=""))
 result <- paste(h_lev[z],d_lev[f],sep="")

### 2024-11-01 code/R/7_metacom_sim_v02.R

@@ -195,9 +195,9 @@ for(v.d in 1:length(d_vals)){
       pop[N[,,t-1]>0 & N[,,t-1] < 100]  = runif(length(pop[N[,,t-1]>0 & N[,,t-1] < 100]))
       extinct<-which(pop>0&pop<extprob,arr.ind=T) # Record position of species that went stochastically extinct
       if (nrow(extinct)>0) {
-        for (z in 1:nrow(extinct)){   #For each species that is stochastically extinct
-          xt[extinct[z,1],extinct[z,2],t] =0
-          dt[extinct[z,1],extinct[z,2],t] =0
+        for (zb in 1:nrow(extinct)){   #For each species that is stochastically extinct
+          xt[extinct[zb,1],extinct[zb,2],t] =0
+          dt[extinct[zb,1],extinct[zb,2],t] =0
         }
       }
       pop[pop>0&pop<extprob]=0  # Stochastically extinct populations set to zero for this generation    

### 2024-11-01 code/R/1_metacom_sim_v01.R
Edited another z to zb

@@ -185,9 +185,9 @@ for (t in 2:time) {
     pop[N[,,t-1]>0 & N[,,t-1] < 100]  = runif(length(pop[N[,,t-1]>0 & N[,,t-1] < 100]))
     extinct<-which(pop>0&pop<extprob,arr.ind=T) # Record position of species that went stochastically extinct
     if (nrow(extinct)>0) {
-        for (z in 1:nrow(extinct)){   #For each species that is stochastically extinct
-            xt[extinct[z,1],extinct[z,2],t] =0
-            dt[extinct[z,1],extinct[z,2],t] =0
+        for (zb in 1:nrow(extinct)){   #For each species that is stochastically extinct
+            xt[extinct[zb,1],extinct[zb,2],t] =0
+            dt[extinct[zb,1],extinct[zb,2],t] =0
         }
     }
     pop[pop>0&pop<extprob]=0  # Stochastically extinct populations set to zero for this generation    

### 2024-11-01 code/R/5_results_hmsc.R
testing all 5 scripts. Need to change script 5 in main.

@@ -14,7 +14,7 @@ load("./data/h_01_d_minus3_hmsc_v01.RData")
 
 #### Step 1. Evaluate posteriors --------------
 m.post = Hmsc::convertToCodaObject(m.1)
-m.df <- as.data.frame(rbind(m.post$Beta[[1]],m.post$Beta[[2]]))
+m.df <- do.call(rbind, m.post$Beta)
 
 #### Create Variance Paritioning plot
 s <- ncol(m.1$Y)
@@ -50,14 +50,14 @@ colnames(B.int)[1] <- "Int1"
 bayesplot::mcmc_areas(B.int,rhat=xc) + legend_none() +  theme(text = element_text(size = 20))
 
 # N ~ Environment ##########
-B.E <- m.df[,2]
+B.E <- m.df[,seq(2,length(colnames(m.df)),npred)]
 # All values, colored by overlap of HPDI with 0
 xc<-NULL
 for(i in 1:ncol(B.E)) xc[i]<-as.numeric(sum(colnames(B.E)[i]==not0))+1
 theme_set(theme_grey())
 #Renaming it to E and the according species number
 for(i in 1:(s-1)){
-  B.E <- cbind(B.E,m.df[,2+(i*npred)])
+  #B.E <- cbind(B.E,m.df[,2+(i*npred)])
   colnames(B.E)[i+1] <- paste0("E", (i+1), sep = "")
 }
 colnames(B.E)[1] <- "E1"

### 2024-11-01 code/R/5_results_hmsc.R
Generalized for number of chains, fixed error (I received at least) for B.E (beta for env).

@@ -1,6 +1,6 @@
 # Title. results.R
 # Author. J.H. Pantel
-# Date 28-08-2023
+# Date 01-11-2024
 # Description. This script contains analyses for the HMSC_ecoevo project and associated manuscript Hermann & Pantel 202x.
 
 #### Libraries ---------------------------------------------------------------
@@ -14,7 +14,7 @@ load("./data/h_01_d_minus3_hmsc_v01.RData")
 
 #### Step 1. Evaluate posteriors --------------
 m.post = Hmsc::convertToCodaObject(m.1)
-m.df <- as.data.frame(rbind(m.post$Beta[[1]],m.post$Beta[[2]]))
+m.df <- do.call(rbind, m.post$Beta)
 
 #### Create Variance Paritioning plot
 s <- ncol(m.1$Y)
@@ -50,14 +50,14 @@ colnames(B.int)[1] <- "Int1"
 bayesplot::mcmc_areas(B.int,rhat=xc) + legend_none() +  theme(text = element_text(size = 20))
 
 # N ~ Environment ##########
-B.E <- m.df[,2]
+B.E <- m.df[,seq(2,length(colnames(m.df)),npred)]
 # All values, colored by overlap of HPDI with 0
 xc<-NULL
 for(i in 1:ncol(B.E)) xc[i]<-as.numeric(sum(colnames(B.E)[i]==not0))+1
 theme_set(theme_grey())
 #Renaming it to E and the according species number
 for(i in 1:(s-1)){
-  B.E <- cbind(B.E,m.df[,2+(i*npred)])
+  #B.E <- cbind(B.E,m.df[,2+(i*npred)])
   colnames(B.E)[i+1] <- paste0("E", (i+1), sep = "")
 }
 colnames(B.E)[1] <- "E1"

### 2024-11-01 code/R/3_main_v01.R
All 5 files work with each other from beginning to end. (check y-axis labels for beta-E plot). No more changes to these in main unless we agree. This is starting point.

@@ -40,7 +40,7 @@ f <- 8
 
 print(paste(h_lev[z],d_lev[f],sep=""))
 result <- paste(h_lev[z],d_lev[f],sep="")
-load(paste("././data/",result,"_res.RData",sep="")) # Scenario 1
+load(paste("././data/",result,"_res_short_E_var0.1.RData",sep="")) # Scenario 1
 # res  <- R.matlab::readMat(paste(result,"_res_v02.mat",sep="")) # Scenario 2
 # res  <- R.matlab::readMat(paste(result,"_res_v03.mat",sep="")) # Scenario 3
 # res  <- R.matlab::readMat(paste(result,"_res_v04.mat",sep="")) # Scenario 4
@@ -146,7 +146,7 @@ if (length(c) > 0) { # if there are any extinct species
 #XFormula = ~env1+x1+x2+x3+x4+x5+x6+x7+x8+x9+x10+x11+x12+x13+x14+x15
 sp <- s[-e]
 sp
-XFormula = ~env1+x1+x2+x3+x6+x7+x8+x9+x10+x12+x13+x14+x15
+XFormula = ~env1+x1+x3+x4+x6+x7+x8+x9+x10+x12+x13+x14+x15
 m <- Hmsc::Hmsc(Y=Y,XData=X,XFormula = XFormula,studyDesign=Random,ranLevels=list("site"=rL.spatial,"year"=rL.time),distr="normal",XScale=TRUE,YScale=TRUE)
 tic()
 
@@ -158,7 +158,7 @@ verbose <- 50*thin
 
 m.1 <- Hmsc::sampleMcmc(m,samples=samples,transient=transient,nChains=nChains,nParallel=nChains,verbose=verbose,thin=thin,updater = list(GammaEta = FALSE))
 a <- toc()
-save.image(file=file.path(paste("./data/",h_lev[z],d_lev[f],"_hmsc_v01_short.RData",sep=""))) # Scenario 1
+save.image(file=file.path(paste("./data/",h_lev[z],d_lev[f],"_hmsc_v01_short_E_var0.1.RData",sep=""))) # Scenario 1
 # save.image(file=file.path(paste(h_lev[z],d_lev[f],"_hmsc_v02.RData",sep=""))) # Scenario 2
 # save.image(file=file.path(paste(h_lev[z],d_lev[f],"_hmsc_v03.RData",sep=""))) # Scenario 3
[…]

### 2024-11-01 code/R/4_diagnostics.R

@@ -13,7 +13,7 @@ library(bayesplot)
 
 #### Checking the MCMC chains --------------
 #### Step 0 Load the data --------------
-load("./data/h_01_d_minus3_hmsc_v01.RData")
+load("./data/h_01_d_minus3_hmsc_v01_short_E_var0.1.RData")
 
 #### Step 1 Create the necessary data.frame --------------
 
@@ -28,9 +28,9 @@ npred <- ncol(m.1$XData)+1
 sp<-colnames(m.1$Y)
 sp_n<-as.numeric(sub("y","",colnames(m.1$Y)))
 VP <- computeVariancePartitioning(m.1, group = c(1,2,rep(3,s)), groupnames = c("Intercept","Env","deltaX"))
-pdf("./output/vp2.pdf",width=8,height=4)
+#pdf("./output/vp2.pdf",width=8,height=4)
 plotVariancePartitioning(m.1, VP = VP, args.legend=list(cex=0.75,bg="transparent"))
-dev.off()
+#dev.off()
 
 #### Step 2 Creatin simpple histograms --------------
 #Creating a histogram of the effective sample size (ess) of all Beta parameters

### 2024-11-01 code/R/3_main_v02.R
HMSC script for my sims

@@ -1,6 +1,6 @@
-# Title. main_v02.R
+# Title. main_v01.R
 # Author. J.H. Pantel
-# Date 28-10-2024
+# Date 24-08-2023
 # Description. This script contains analyses for the HMSC_ecoevo project and associated manuscript Hermann & Pantel 202x.
 
 #### Libraries ---------------------------------------------------------------
@@ -35,8 +35,8 @@ h_lev <- c("h_0_","h_01_","h_02_","h_03_","h_04_","h_05_","h_06_","h_07_","h_08_
 #setwd("./data")
 
 # HMSC for single results condition
-z <- 2
-f <- 8
+z <- 4
+f <- 1
 
 print(paste(h_lev[z],d_lev[f],sep=""))
 result <- paste(h_lev[z],d_lev[f],sep="")
@@ -53,10 +53,17 @@ d <- b[c] # generation of last occurrence for species that go extinct
 
 # Arrange all data for HMSC
 ## site-by-species abundance data as a time series
-Y <- array(NA,dim=c(50000,15),dimnames=list(NULL,c("y1","y2","y3","y4","y5","y6","y7","y8","y9","y10","y11","y12","y13","y14","y15")))
-count <- seq(1,50000,by=50)
-for(i in 1:1000){
-  Y[(count[i]:(count[i]+49)),] <- N[,,i]
+spec_name = NULL
+trait_name = NULL
+for (i in 1:spec){
+  spec_name[i] = paste("y",i,"",sep="")
+  trait_name[i] = paste("x",i,"",sep="")
+}
+
+Y <- array(NA,dim=c((patch*cut),spec),dimnames=list(NULL,spec_name))
+count <- seq(1,(patch*cut),by=patch)
+for(i in 1:cut){
+  Y[(count[i]:(count[i]+(patch-1))),] <- N[,,i]
 }
 
 ## site-by-species trait data as a time series
@@ -66,25 +73,25 @@ trait_init[trait_init == 9999] <- NaN
 trait_x[,,1] <- trait_init
 ## site-by-species CHANGE in trait value as a time series
 ## absolute value of change in trait value
-delta_x <- array(NA,c(50,15,999),dimnames=list(NULL,c("x1","x2","x3","x4","x5","x6","x7","x8","x9","x10","x11","x12","x13","x14","x15"),NULL))
[…]

### 2024-11-03 manuscript/Appendix_S4_v01.Rmd
began mc sim analysis

@@
 ```
+```{r}
+m2.post.hmsc <- convertToCodaObject(m.2.sample)
+summary(m2.post.hmsc$Beta)
+bayesplot::mcmc_trace(m2.post.hmsc$Beta)
+bayesplot::mcmc_areas(m2.post.hmsc$Beta,area_method = c("equal height"))
+```
+
+```{r}
+Gradient <- constructGradient(m.2.sample,focalVariable="dx1",non.focalVariables=list(E=list(2),Esq=list(2)),ngrid=59)
+# Esq is manually constructed as gradient-produced E^2
+Gradient$XDataNew$Esq <- Gradient$XDataNew$E^2
+predY <- predict(m.2.sample,XData=Gradient$XDataNew,expected=TRUE)
+plotGradient(m.2.sample,Gradient,pred=predY,showData=T,measure="Y",main="",xlab="dx1_t",ylab="predicted N_t+1")
+```
 

### 2024-11-06 code/R/7_metacom_sim_v02.R
Preliminary results for h2 and d metacom sim

@@ -439,3 +439,14 @@ colnames(mean_div.gg)[1:4] <- c("d_lev","h_lev","level","mean")
 mean_div.gg$level <- as.factor(mean_div.gg$level)
 
 p <- ggplot2::ggplot(mean_div.gg,ggplot2::aes(x=d_lev,y=mean,col=level,linetype=h_lev)) + ggplot2::geom_point(aes(fill=level),size=2,shape=21,col="black") + ggplot2::geom_line(aes(group=interaction(level,h_lev)),size=0.25,col="black") + ggplot2::geom_errorbar(ggplot2::aes(ymin=mean-se, ymax=mean+se), width=0.2, size=0.25, col="black", linetype=1) + ggplot2::coord_cartesian(ylim=c(0,15)) + ggplot2::theme_classic() + ggplot2::ylab("diversity") + ggplot2::xlab("dispersal") + scale_x_discrete(labels=c(0,expression(10^-9),expression(10^-8),expression(10^-7),expression(10^-6),expression(10^-5),expression(10^-4),expression(10^-3),expression(10^-2),0.1,0.2,0.3,0.4,0.5,0.6,0.7,0.8,0.9,1)) + scale_fill_manual(values=c("black","white", "darkgrey"))
+
+
+load("~/Nextcloud/Pantel/data/h_01_d_zero_hmsc_v02.RData")
+#### Step 1. Evaluate posteriors --------------
+m.post = Hmsc::convertToCodaObject(m.1)
+m.df <- do.call(rbind, m.post$Beta)
+#### Create Variance Paritioning plot
+s <- ncol(m.1$Y)
+VP <- computeVariancePartitioning(m.1, group = c(1,1,rep(2,s)), groupnames = c("Env","deltaX"))
+#pdf("./output/vp2.pdf",width=8,height=4)
+plotVariancePartitioning(m.1, VP = VP, args.legend=list(cex=0.75,bg="transparent"),cols=c("white","skyblue","darkgrey","orange","pink"))

### 2024-11-06 code/R/fig1.R

@@ -102,7 +102,7 @@ p4 <- bayesplot::mcmc_areas(m.beta) + scale_x_continuous(limits=c(-20,30)) +
 ### variance partitioning ###
 par(mgp=c(2,0.45,0), tcl=-0.4, mar=c(1.3,1.2,0.5,0.5))
 VP <- computeVariancePartitioning(m.1.sample, group = c(1, 1, 1), groupnames = "Env")
-plotVariancePartitioning(m.1.sample, VP, args.legend = list(cex = 0.6, bg = "transparent"),cex.axis=0.6,main="")
+plotVariancePartitioning(m.1.sample, VP, args.legend = list(cex = 0.6, bg = "transparent"),cex.axis=0.6,main="",cols=c("white","orange"))
 p5 <- recordPlot()
 ### ### ### ### ### ### ###
 ### all plots together ###

### 2024-11-06 manuscript/manuscript_v11.Rmd
Added Intro to mscript, knit to pdf

@@
 
-XXXXX
+Analysis of biodiversity at large spatial scales can be complex, because numerous environmental properties, spatial structure and connectivity, interactions between species, as well as stochastic processes such as drift and priority effects operate and determine observed patterns.
+Statistical models in ecology have made tremendous progress in reflecting these numerous ecological processes at multiple scales.
+There are now a diversity of joint species distribution models [JSDMs; @Pollock2014] [@Pichler2021] or multi-species occupancy models [@Devarajan2020] [@Poggiato2021], which can incorporate these diverse components that structure variance in biological data.
+Many of these models increasingly have implementations in R packages (e.g.
+spOccupancy, [@Doser2022]), and have been used to measure the relative importance of different ecological drivers in fungi, birds, mammals, phytoplankton, fish, and other taxa, from local to continental scales (e.g.
+[@Abrego2020] [@Antao2022] [@Keppeler2022]).
+
+There are some remaining drivers of community composition that aren't currently reflected in these models.
+The dynamics of populations and communities, and resulting biodiversity patterns, are also influenced by the genotypic and phenotypic properties of organisms as well.
+Although traits can be considered in some JSDMs, they typically require fixed values for species' traits ([@Ovaskainen2017HMSC] [@Tikhonov2020]; but see @Abrego2024).
+However, ecological dynamic processes can impact selection on genes and phenotypes, potentially resulting in interactions or feedback loops between ecological and evolutionary processes [@Lion2018] [@Barbour2022].
+Although an increasing number of theoretical and empirical studies have evaluated the impact of evolutionary dynamics for community assembly and metacommunity dynamics [@Loeuille2008] [@Osmond2013] [@Pantel2015] [@Toju2017], analytical methods to consider this additional complexity lag behind the scope of data collected in eco-evolutionary studies.
+Some methods exist that ultimately apply an ANOVA framework to partition variance in traits or growth rates across contributions of ecological and evolutionary components ([@Hairston2005] [@Ellner2011] [@Govaert2016]).
+These methods tend to be cited more frequently for their potential than for their actual application to empirical data (but see [@terHorst2014] [@Hiltunen2014]), because this approach is ultimately limited by the same requirements that can limit application of ANOVA - data should follow a normal distribution, the processes that structure the target response variable must be linear and additive, and complicated structuring mechanisms must be reduced into one or a few target categories labeled as ‘ecology’ or ‘evolution’.
+Flexible models, which can vary depending on the mechanisms and structure of the diverse kinds of data that most studies of eco-evolutionary dynamics will collect, are needed for rigorous inference about eco-evolutionary processes [@Pantel2023].
+
+Two existing statistical models are potentially well suited for analysis of eco-evolutionary data at the microevolutionary and metacommunity scale but also have some considerations before their application to complex community data.
+Lasky et al.
+(2020) [@Lasky2020] developed an integrated reaction norm model, linking genetic, phenotypic, and demographic processes, and Benito Garzón et al.
+(2019) [@Garzon2019] presented a species distribution model with local adaptation and phenotypic plasticity ($\Delta$ SDMs).
+The model of Lasky et al.
+is an excellent candidate for spatially complex population dynamics, and awaits an extension to the multispecies level.
+The $\Delta$ SDM models described by Benito Garzón et al.
+truly disentangle the role that plastic and evolutionary trait divergence can play in species distributions, but they also require common garden experimental data across sites to estimate these effects.
+Observational data collected in metacommunity surveys may only have observational data across sites or time points, as large-scale experimental measures of reaction norms required by the $\Delta$ SDM model may be difficult to obtain.
+
+One particular metacommunity statistical model, the Hierarchical Modelling of Species Communities (HMSC) framework [@Ovaskainen2017HMSC] [@Tikhonov2020], explicitly considers the multitude of processes that can structure community composition across time and space [@Leibold2022] and is thus well suited to consider analysis of metacommunity eco-evolutionary data.
+This model of environmental filtering via variation and covariation in how species respond to their environment considers the impacts of (fixed) trait values and phylogenetic relationships, and uses latent variables to account for the numerous unobserved environmental and spatial features that are difficult to exhaustively measure in community surveys.
+The associated R package has immense flexibility, and has been applied to study environmental and abiotic drivers of diverse assemblages of organisms (e.g.
+[@Sandal2022] [@Weigel2023]).
+HMSC is potentially useful for studying eco-evolutionary drivers of community structure, but the model has not previously been applied to include heritable changes in trait values over time.
+
+In this study, we use simulations of population dynamics, environmental drivers, species interactions, and trait evolution in populations, communities, and at the metacommunity scale to generate realistic time series of population sizes and mean trait values.
+We then use HMSC to analyze the resulting data, to determine whether HMSC can successfully estimate the impacts of trait evolution for community composition, and to generate relative contributions of fixed environmental, spatial, and trait evolution drivers as well as random effects of spatial and temporal covariance among species.
+In order to determine whether HMSC is useful for hypothesis testing about the drivers of eco-evolutionary dynamics in communities in a structured landscape, we XXX SIMULATIONS XXX to determine whether these potential drivers altered the importance of evolution for metacommunity composition.
+We hypothesized that XXX.
+We also XXX EMPIRICAL XXX.
 

### 2024-11-09 code/R/fig4.R
Draft code for simulations figure

@@ -1,0 +1,151 @@
+## Code for Figure 4
+## J.H. Pantel
+## 09-11-2024
+
+## libraries
+library(ggplot2)
+library(ggalluvial)
+library(Hmsc)
+library(vegan)
+
+## Figure 4a. Alpha diversity across d and h2 levels
+## functions
+CV <- function(x){
+  a <- mean(x,na.rm=T)	#Get the average of all values
+  b <- (x - a)^2					#Squared deviations from mean
+  c <- sqrt((sum(b,na.rm=T))/(length(x[is.na(x) == F])-1)) #sum squared deviations, divide by sample N-1, take square root
+  d <- c/a  #coefficient of variation
+  return(CV=d)
+}
+standard_error <- function(x){
+  st_err <- sd(x) / sqrt(length(x))	
+  
+  return(st_err)
+}
+## Read in and save relevant data
+div_200 <- array(NA,dim=c(19,11,3,3),dimnames=list(c("d_zero","d_minus9","d_minus8","d_minus7","d_minus6","d_minus5","d_minus4","d_minus3","d_minus2","d_01","d_02","d_03","d_04","d_05","d_06","d_07","d_08","d_09","d_10"),c("h_0_","h_01_","h_02_","h_03_","h_04_","h_05_","h_06_","h_07_","h_08_","h_09_","h_10_"),c("alpha","gamma","beta"),c("mean","CV","SDE")))
+d_lev <- c("d_zero","d_minus9","d_minus8","d_minus7","d_minus6","d_minus5","d_minus4","d_minus3","d_minus2","d_01","d_02","d_03","d_04","d_05","d_06","d_07","d_08","d_09","d_10")
+h_lev <- c("h_0_","h_01_","h_02_","h_03_","h_04_","h_05_","h_06_","h_07_","h_08_","h_09_","h_10_")
+## Inverse Simpson's diversity across additional heritability levels
+for(i in 1:length(d_lev)){
+  for(j in 1:length(h_lev)){
+    ## Read in population size values
+    result <- paste("/Users/jhpantel/Nextcloud/Pantel/data/",h_lev[j],d_lev[i],"_res_v02.RData",sep="")
+    tmp.env <- new.env()
+    load(toString(result),envir=tmp.env)
+    r <- get("N",pos=tmp.env) # alpha
+    rm(tmp.env)
+    r <- r[,,200]
[…]

### 2024-11-09 code/R/1_metacom_sim_v02.R
Cleaning Fig4 code, repeat sims

@@ -1,6 +1,6 @@
 ## Evolving metacommunity simulation
 ## J.H. Pantel & Ruben Hermann
-## October 27 2024
+## November 09 2024
 ## Scenario 1. across d and h2 levels
 ## # P = 1, w = 2, v[E] = 0
   
@@ -321,132 +321,8 @@ for(v.d in 1:length(d_vals)){
     h_lev <- c("h_0_","h_01_","h_02_","h_03_","h_04_","h_05_","h_06_","h_07_","h_08_","h_09_","h_10_")
     print(paste(h2_names[v.h2],d_names[v.d],sep=""))
     result <- paste(h_lev[v.h2],d_lev[v.d],sep="")
-    save.image(paste("/Users/jhpantel/Nextcloud/Pantel/data/",result,"_res_v02.RData",sep=""))
+    save.image(paste("./data/mc_v02/",result,"_res_v02.RData",sep=""))
     #### Remove results
     rm(list=setdiff(ls(), c("d_vals","h2_vals","d_names","h2_names","v.d","v.h2")))
   }
 }
-
-###################################
-## Step 5. Evaluate data.        ##
-###################################
-CV <- function(x){
-  a <- mean(x,na.rm=T)	#Get the average of all values
-  b <- (x - a)^2					#Squared deviations from mean
-  c <- sqrt((sum(b,na.rm=T))/(length(x[is.na(x) == F])-1)) #sum squared deviations, divide by sample N-1, take square root
-  d <- c/a  #coefficient of variation
-  return(CV=d)
-}
-
-standard_error <- function(x){
-  st_err <- sd(x) / sqrt(length(x))	
-  
-  return(st_err)
-}
-####### Figure 1. Plot of diversity (y-axis) across dispersal levels (x-axis) for simulations at different levels of heritability (h2). #######
[…]

### 2024-11-09 code/R/1_metacom_sim_v02.R
Added h2=0.05

@@ -159,9 +159,9 @@ alpha[lower.tri(alpha,diag=F)]=0 + (.0015-0)*runif(length(alpha[lower.tri(alpha,
 
 ## Set up dispersal and heritability to vary across a loop
 d_vals = c(0, 10^-9, 10^-8, 10^-7, 10^-6, 10^-5, 10^-4, 10^-3, 10^-2, 10^-1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 1)
-h2_vals = c(0, 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 1)
+h2_vals = c(0, 0.05, 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 1)
 d_names = c("0", "minus9", "minus8", "minus7", "minus6", "minus5", "minus4", "minus3", "minus2", "01", "02", "03", "04", "05", "06", "07", "08", "09", "10")
-h2_names = c("0", "01", "02", "03", "04", "05", "06", "07", "08", "09", "10")
+h2_names = c("0", "005", "01", "02", "03", "04", "05", "06", "07", "08", "09", "10")
 
 ## Save initial community conditions
 save.image("./data/init_v02.RData")
@@ -171,7 +171,7 @@ rm(list=setdiff(ls(), c("d_vals","h2_vals","d_names","h2_names","v.d","v.h2")))
 
 ## Vary h2 and d values across simulation runs
 for(v.d in 1:length(d_vals)){
-  for(v.h2 in 1:length(h2_vals)){
+  for(v.h2 in 2:3){
     load("./data/init_v02.RData")
     d = array(d_vals[v.d],c(1,spec))
     h2 = array(h2_vals[v.h2],c(1,spec))
@@ -318,7 +318,7 @@ for(v.d in 1:length(d_vals)){
     xt <- xt[,,1:cut]
     Et <- Et[,1:cut]
     d_lev <- c("d_zero","d_minus9","d_minus8","d_minus7","d_minus6","d_minus5","d_minus4","d_minus3","d_minus2","d_01","d_02","d_03","d_04","d_05","d_06","d_07","d_08","d_09","d_10")
-    h_lev <- c("h_0_","h_01_","h_02_","h_03_","h_04_","h_05_","h_06_","h_07_","h_08_","h_09_","h_10_")
[…]

### 2024-11-09 code/R/1_metacom_sim_v02.R
Updated h2=0.005

@@ -171,7 +171,7 @@ rm(list=setdiff(ls(), c("d_vals","h2_vals","d_names","h2_names","v.d","v.h2")))
 
 ## Vary h2 and d values across simulation runs
 for(v.d in 1:length(d_vals)){
-  for(v.h2 in 2:3){
+  for(v.h2 in 2){
     load("./data/init_v02.RData")
     d = array(d_vals[v.d],c(1,spec))
     h2 = array(h2_vals[v.h2],c(1,spec))

### 2024-11-09 code/R/fig4.R
Started mc conept figure

@@ -7,6 +7,7 @@ library(ggplot2)
 library(ggalluvial)
 library(Hmsc)
 library(vegan)
+library(cowplot)
 
 ## Figure 4a. Alpha diversity across d and h2 levels
 ## functions
@@ -30,7 +31,7 @@ h_lev <- c("h_0_","h_01_","h_02_","h_03_","h_04_","h_05_","h_06_","h_07_","h_08_
 for(i in 1:length(d_lev)){
   for(j in 1:length(h_lev)){
     ## Read in population size values
-    result <- paste("/Users/jhpantel/Nextcloud/Pantel/data/",h_lev[j],d_lev[i],"_res_v02.RData",sep="")
+    result <- paste("./data/mc_v02/",h_lev[j],d_lev[i],"_res_v02.RData",sep="")
     tmp.env <- new.env()
     load(toString(result),envir=tmp.env)
     r <- get("N",pos=tmp.env) # alpha
@@ -143,6 +144,41 @@ p3 <- ggplot(sub3,aes(y = Percentage, x = d, fill = as.character(level))) +
   theme(panel.grid.major = element_blank(), 
         axis.text.y = element_blank(), 
         axis.ticks.y = element_blank(),legend.position=c(0.8,0.5),legend.text = element_text(size=7),legend.title = element_blank())
+
+## Figure 4a. Conceptual MC figure
+load(paste("./data/mc_v02/",h_lev[2],d_lev[8],"_res_v02.RData",sep=""))
+mc <- data.frame(x=rep(NA,50*3),y=rep(NA,50*3),t=rep(c(0,100,200),each=50),gam=rep(NA,50*3),xt=rep(NA,50*3))
+# t0
+mc[1:50,c(1,2)] <- xy
+a  <- diversity(N[,,1],index="invsimpson")
+a[is.finite(a)==FALSE] <- 0
+mc[1:50,4] <- a # alpha diversity
+b <- rowSums(xt[,,1]*N[,,1],na.rm=TRUE) / rowSums(N[,,1])
+b[is.finite(b)==FALSE] <- 0
+mc[1:50,5] <- b # community weighted mean trait value
+# t100
+mc[51:100,c(1,2)] <- xy
+a  <- diversity(N[,,100],index="invsimpson")
+a[is.finite(a)==FALSE] <- 0
+mc[51:100,4] <- a # alpha diversity
[…]

### 2024-11-09 code/R/fig4.R
Size by gamma, need to update by ts relative gamma

@@ -174,11 +174,11 @@ b[is.finite(b)==FALSE] <- 0
 mc[101:150,5] <- b # community weighted mean trait value
 
 sub1 <- mc[mc$t==0,]
-a1 <- ggplot(sub1,aes(x,y)) + geom_point(aes(color=xt)) + scale_color_gradient(low = "white", high = "skyblue") + theme(legend.position = "none")
+a1 <- ggplot(sub1,aes(x,y)) + geom_point(aes(color=xt,size=gam)) + scale_color_gradient(low = "white", high = "skyblue") + theme(legend.position = "none")
 sub2 <- mc[mc$t==100,]
-a2 <- ggplot(sub2,aes(x,y)) + geom_point(aes(color=xt)) + scale_color_gradient(low = "white", high = "skyblue") + theme(legend.position = "none")
+a2 <- ggplot(sub2,aes(x,y)) + geom_point(aes(color=xt,size=gam)) + scale_color_gradient(low = "white", high = "skyblue") + theme(legend.position = "none")
 sub3 <- mc[mc$t==200,]
-a3 <- ggplot(sub3,aes(x,y)) + geom_point(aes(color=xt)) + scale_color_gradient(low = "white", high = "skyblue") + theme(legend.position = "none")
+a3 <- ggplot(sub3,aes(x,y)) + geom_point(aes(color=xt,size=gam)) + scale_color_gradient(low = "white", high = "skyblue") + theme(legend.position = "none")
 
 d <- plot_grid(p0,plot_grid(p1,p2,p3,nrow=3),ncol=2,rel_widths=c(2,1.5))
 

### 2024-11-10 code/R/1_metacom_sim_v02.R
No h2=0.05, edit Fig4

@@ -159,9 +159,9 @@ alpha[lower.tri(alpha,diag=F)]=0 + (.0015-0)*runif(length(alpha[lower.tri(alpha,
 
 ## Set up dispersal and heritability to vary across a loop
 d_vals = c(0, 10^-9, 10^-8, 10^-7, 10^-6, 10^-5, 10^-4, 10^-3, 10^-2, 10^-1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 1)
-h2_vals = c(0, 0.05, 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 1)
+h2_vals = c(0, 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 1)
 d_names = c("0", "minus9", "minus8", "minus7", "minus6", "minus5", "minus4", "minus3", "minus2", "01", "02", "03", "04", "05", "06", "07", "08", "09", "10")
-h2_names = c("0", "005", "01", "02", "03", "04", "05", "06", "07", "08", "09", "10")
+h2_names = c("0", "01", "02", "03", "04", "05", "06", "07", "08", "09", "10")
 
 ## Save initial community conditions
 save.image("./data/init_v02.RData")
@@ -318,7 +318,7 @@ for(v.d in 1:length(d_vals)){
     xt <- xt[,,1:cut]
     Et <- Et[,1:cut]
     d_lev <- c("d_zero","d_minus9","d_minus8","d_minus7","d_minus6","d_minus5","d_minus4","d_minus3","d_minus2","d_01","d_02","d_03","d_04","d_05","d_06","d_07","d_08","d_09","d_10")
-    h_lev <- c("h_0_","h_005_","h_01_","h_02_","h_03_","h_04_","h_05_","h_06_","h_07_","h_08_","h_09_","h_10_")
+    h_lev <- c("h_0_","h_01_","h_02_","h_03_","h_04_","h_05_","h_06_","h_07_","h_08_","h_09_","h_10_")
     print(paste(h2_names[v.h2],d_names[v.d],sep=""))
     result <- paste(h_lev[v.h2],d_lev[v.d],sep="")
     save.image(paste("./data/mc_v02/",result,"_res_v02.RData",sep=""))

### 2024-11-10 code/R/fig4.R

@@ -9,7 +9,7 @@ library(Hmsc)
 library(vegan)
 library(cowplot)
 
-## Figure 4a. Alpha diversity across d and h2 levels
+#### Figure 4a. Alpha diversity across d and h2 levels ####
 ## functions
 CV <- function(x){
   a <- mean(x,na.rm=T)	#Get the average of all values
@@ -71,9 +71,9 @@ for(i in 1:length(h2_vals)){
 sub <- mean_div.gg[mean_div.gg$level=="gamma",]
 sub2 <- sub[(sub$h2==0 | sub$h2==0.1 | sub$h2==0.3 | sub$h2==0.9),]
   
-p0 <- ggplot(sub2,aes(x=d_lev,y=mean)) + geom_point(aes(colour=h2)) + scale_colour_gradient(low="lightgrey",high="forestgreen")+ geom_line(aes(group=interaction(level,h_lev)),size=0.25,col="black",linewidth=0.1) + coord_cartesian(ylim=c(0,15)) + theme_classic() + ylab(expression(gamma*" diversity")) + xlab("dispersal") + scale_x_discrete(labels=c(0,expression(10^-9),expression(10^-8),expression(10^-7),expression(10^-6),expression(10^-5),expression(10^-4),expression(10^-3),expression(10^-2),0.1,0.2,0.3,0.4,0.5,0.6,0.7,0.8,0.9,1))
+p0 <- ggplot(sub2,aes(x=d_lev,y=mean)) + geom_point(aes(colour=h2)) + scale_colour_gradient(low="lightgrey",high="forestgreen")+ geom_line(aes(group=interaction(level,h_lev)),col="black",linewidth=0.1) + coord_cartesian(ylim=c(0,15)) + theme_classic() + ylab(expression(gamma*" diversity")) + xlab("dispersal") + scale_x_discrete(labels=c(0,expression(10^-9),expression(10^-8),expression(10^-7),expression(10^-6),expression(10^-5),expression(10^-4),expression(10^-3),expression(10^-2),0.1,0.2,0.3,0.4,0.5,0.6,0.7,0.8,0.9,1))
 
-## Figure 4b. Stacked bar chart of mean VarPart results across d and h2 levels tested
+#### Figure 4b. Stacked bar chart of mean VarPart results across d and h2 levels tested ####
 ## Enter VarPart data
[…]

### 2024-11-10 code/R/fig4.R
HMSC results in data folder, updated Fig4

@@ -87,7 +87,7 @@ counter <- 1
 for(v.d in c(1,4,8)){ # dlev
   for(v.h2 in c(2,4,10)){ # hlev
     ## Read in population size values
-    result <- paste("/Users/jhpantel/Nextcloud/Pantel/data/",h_lev[v.h2],d_lev[v.d],"_hmsc_v02.RData",sep="")
+    result <- paste("./data/mc_v02/",h_lev[v.h2],d_lev[v.d],"_hmsc_v02.RData",sep="")
     load(result)
     s <- ncol(m.1$Y)
     VP <- computeVariancePartitioning(m.1, group = c(1,1,rep(2,s)), groupnames = c("Env","deltaX"))

### 2024-11-10 code/R/fig4.R
Fig4 with draft 'concept' figure

@@ -145,15 +145,9 @@ p3 <- ggplot(sub3,aes(y = Percentage, x = d, fill = as.character(level))) +
         axis.text.y = element_blank(), 
         axis.ticks.y = element_blank(),legend.position=c(0.8,0.5),legend.text = element_text(size=7),legend.title = element_blank())
 
-d <- plot_grid(p0,plot_grid(p1,p2,p3,nrow=3),ncol=2,rel_widths=c(2,1.5))
-
-pdf("./output/fig4.pdf",paper="a4r")
-d
-dev.off()
-
 #### Figure 4a. Conceptual MC figure ####
 load(paste("./data/mc_v02/",h_lev[2],d_lev[8],"_res_v02.RData",sep=""))
-mc <- data.frame(x=rep(NA,50*3),y=rep(NA,50*3),t=rep(c(0,100,200),each=50),gam=rep(NA,50*3),xt=rep(NA,50*3))
+mc <- data.frame(x=rep(NA,50*3),y=rep(NA,50*3),t=rep(c(0,10,200),each=50),div_a=rep(NA,50*3),xt=rep(NA,50*3))
 # t0
 mc[1:50,c(1,2)] <- xy
 a  <- diversity(N[,,1],index="invsimpson")
@@ -162,12 +156,12 @@ mc[1:50,4] <- a # alpha diversity
 b <- rowSums(xt[,,1]*N[,,1],na.rm=TRUE) / rowSums(N[,,1])
 b[is.finite(b)==FALSE] <- 0
 mc[1:50,5] <- b # community weighted mean trait value
-# t100
+# t10
 mc[51:100,c(1,2)] <- xy
-a  <- diversity(N[,,100],index="invsimpson")
+a  <- diversity(N[,,10],index="invsimpson")
 a[is.finite(a)==FALSE] <- 0
 mc[51:100,4] <- a # alpha diversity
-b <- rowSums(xt[,,100]*N[,,100],na.rm=TRUE) / rowSums(N[,,100])
+b <- rowSums(xt[,,10]*N[,,10],na.rm=TRUE) / rowSums(N[,,10])
 b[is.finite(b)==FALSE] <- 0
 mc[51:100,5] <- b # community weighted mean trait value
 # t200
@@ -179,9 +173,10 @@ b <- rowSums(xt[,,200]*N[,,200],na.rm=TRUE) / rowSums(N[,,200])
 b[is.finite(b)==FALSE] <- 0
 mc[101:150,5] <- b # community weighted mean trait value
 
[…]

### 2024-11-10 manuscript/Appendix_S4_v01.Rmd

@@
 ```
-
+```{r}
+m11.post.hmsc <- convertToCodaObject(m.11.sample)
+summary(m11.post.hmsc$Beta)
+bayesplot::mcmc_trace(m11.post.hmsc$Beta)
+bayesplot::mcmc_areas(m11.post.hmsc$Beta,area_method = c("equal height"))
+```
 ```{r warning=FALSE}
@@
 
+We can see that the variation in Y is a combination of the relative importance of density-dependence and of trait evolution.
+We can see that relative importance changes across $h^2$ levels.
+What drives changes in this relative importance?
+
+```{r}
+d_dat <- gdat[gdat$time != 1,c(1,2,3,5)]
+d_dat <- as.data.frame(d_dat)
+d_dat$dxt <- NA
+d_dat$ddt <- NA
+d_dat$dN <- NA
+d_dat$dd <- NA
+for(i in 1:11){
+  sub <- gdat[gdat$h2==levels(gdat$h2)[i],]
+  d_dat$dxt[d_dat$h2==levels(d_dat$h2)[i]] <- abs(sub$x1[2:60]-sub$x1[1:59]) # absolute value of change in trait value
+  d_dat$ddt[d_dat$h2==levels(d_dat$h2)[i]] <- alpha[1,1]*sub$N1[2:60] - alpha[1,1]*sub$N1[1:59]
+  d_dat$dN[d_dat$h2==levels(d_dat$h2)[i]] <- sub$N1[2:60] - sub$N1[1:59]
+  d_dat$dd[d_dat$h2==levels(d_dat$h2)[i]] <- alpha[1,1]*d_dat$N1[d_dat$h2==levels(d_dat$h2)[i]]
+}
+```
+```{r}
+### Plot of time xx ###
+d_dat <- d_dat[d_dat$h2 != 0,]
+d_dat$ln_N <- log(d_dat$N1)
+ggplot2::ggplot(d_dat, aes(dxt, dN, col=h2)) + geom_line() + geom_point()
+ggplot2::ggplot(d_dat, aes(ddt, dN, col=h2)) + geom_line() + geom_point()
+### ###
+```
+
 # References

### 2024-11-10 manuscript/manuscript_v11.Rmd
draft Intro,Methods,Results for Fig4.

@@
 
-Title
-===============================================================================
+# Title
 
@@
 
-Abstract
-===============================================================================
+# Abstract
 
@@
 
-Introduction
-================================================================================
+# Introduction
 
@@
 Some methods exist that ultimately apply an ANOVA framework to partition variance in traits or growth rates across contributions of ecological and evolutionary components ([@Hairston2005] [@Ellner2011] [@Govaert2016]).
-These methods tend to be cited more frequently for their potential than for their actual application to empirical data (but see [@terHorst2014] [@Hiltunen2014]), because this approach is ultimately limited by the same requirements that can limit application of ANOVA - data should follow a normal distribution, the processes that structure the target response variable must be linear and additive, and complicated structuring mechanisms must be reduced into one or a few target categories labeled as ‘ecology’ or ‘evolution’.
+These methods tend to be cited more frequently for their potential than for their actual application to empirical data (but see [@terHorst2014] [@Hiltunen2014]), because this approach is ultimately limited by the same requirements that can limit application of ANOVA - data should follow a normal distribution, the processes that structure the target response variable must be linear and additive, and complicated structuring mechanisms must be reduced into one or a few target categories labeled as 'ecology' or 'evolution'.
 Flexible models, which can vary depending on the mechanisms and structure of the diverse kinds of data that most studies of eco-evolutionary dynamics will collect, are needed for rigorous inference about eco-evolutionary processes [@Pantel2023].
@@
 
-In this study, we use simulations of population dynamics, environmental drivers, species interactions, and trait evolution in populations, communities, and at the metacommunity scale to generate realistic time series of population sizes and mean trait values.
+In this study, we use simulations of population dynamics, environmental drivers, species interactions, and trait evolution in populations, communities, at the local and metacommunity scale, to generate simulated time series of population sizes and mean trait values.
 We then use HMSC to analyze the resulting data, to determine whether HMSC can successfully estimate the impacts of trait evolution for community composition, and to generate relative contributions of fixed environmental, spatial, and trait evolution drivers as well as random effects of spatial and temporal covariance among species.
-In order to determine whether HMSC is useful for hypothesis testing about the drivers of eco-evolutionary dynamics in communities in a structured landscape, we XXX SIMULATIONS XXX to determine whether these potential drivers altered the importance of evolution for metacommunity composition.
-We hypothesized that XXX.
-We also XXX EMPIRICAL XXX.
+We asked three main questions: (i) do linear statistical models (like HMSC) accurately capture the impact of non-linear population, community, and evolutionary dynamic processes?, (ii) can HMSC be used to evaluate hypothesized drivers of the relative importance of evolution for community composition at the metacommunity scale?, and (iii) can we better understand the role of evolution for community dynamics in experimental systems using HMSC?
 
-Methods
-================================================================================
+# Methods
 
@@
 
-We evaluated whether linear regression models can appropriately capture the dynamics that drive species abundances in more complex communities.
+We evaluated whether linear regression models can appropriately capture the dynamics that drive species abundances in relatively simple populations and communities.
 We fit data generated from simulation models with increasing complexity to Bayesian linear regression models.
@@
 For linear regression models to fit to the simulated population size data, we began with models using population size and environment as fixed effects ($N_t \sim N_{t+1} + E_t + E_t^2$), and progressed to a mixed model using environment and trait evolution as fixed effects ($N_t \sim E_t + E_t^2 + |\Delta x|$) and time and spatial distance as random effects.
-For consistency, all linear regression models were implemented using HMSC (Hierarchical Modelling of Species Communities [@Ovaskainen2017HMSC] [@Tikhonov2020]).
+For consistency, all linear regression models were implemented as heirarchical Bayesian models using HMSC (Hierarchical Modelling of Species Communities [@Ovaskainen2017HMSC] [@Tikhonov2020]).
 HMSC uses a latent variable approach as an alternative to directly modelling species interaction coefficients, estimating a reduced number of linear combinations of species abundances that best predict future population size values.
@@
 Full methodological details are given in Appendix S1.
+All simulations were run in R (@R2024) and HMSC models were implemented using the `Hmsc` package (@Hmsc2022).
 
@@
 
-We simulated growth and competition dynamics for a multi-species assemblage in a patchy landscape, with site variation in one environmental property.
+###Simulation model
 
-### Environmental variation
-
+We also simulated growth and competition dynamics for a multi-species assemblage in a patchy landscape, with site variation in one environmental property, and with evolution in a trait that is selected on by the environment.
 Population growth for species in the metacommunity simulation follows a Leslie-Gower model (a discrete-time version of a Lotka-Volterra model [@Beverton1957] [@Leslie1958]).
@@
 
-$$ N_{i,t+1} =\frac{\hat{W}e^ \frac{-[(\frac{w+(1-h^2)P} {P+w})(E-x_{i,t})]^2}{2(P+w)}N_{i,t}} {1 + \alpha_{ii}N_{i,t} + \sum_{j \neq i}^{S} \alpha_{ij}N_{j,t}} $$
-where $N_i,t$ is the population size of species *i* at time *t*, \( \hat{W} \) is calculated as \( \hat{W}=W_{max}\sqrt(\frac{w}{P+w}) \), \(W_{max}\) is the species' maximum per-capita growth rate, *w* is the width of the Gaussian fitness function (which determines the strength of selection, as increasing values indicate a weaker reduction in fitness with distance from optimum trait value), *P* is the width of the distribution of the phenotype *x*, $h^2$ is the heritability of the trait *x*, *E* is the local environmental optimum trait value, $x_{i,t}$ is the trait value of species *i* at time *t*, $\alpha_{ii}$ is the intraspecific competition coefficient (the per capita impact of species *i* on itself) and $\alpha_{ij}$ is the interspecific competition coefficient.
+$$ N_{i,t+1} =\frac{\hat{W}e^ \frac{-[(\frac{w+(1-h^2)P} {P+w})(E-x_{i,t})]^2}{2(P+w)}N_{i,t}} {1 + \alpha_{ii}N_{i,t} + \sum_{j \neq i}^{S} \alpha_{ij}N_{j,t}} $$ where $N_i,t$ is the population size of species *i* at time *t*, $\hat{W}$ is calculated as $\hat{W}=W_{max}\sqrt(\frac{w}{P+w})$, $W_{max}$ is the species' maximum per-capita growth rate, *w* is the width of the Gaussian fitness function (which determines the strength of selection, as increasing values indicate a weaker reduction in fitness with distance from optimum trait value), *P* is the width of the distribution of the phenotype *x*, $h^2$ is the heritability of the trait *x*, *E* is the local environmental optimum trait value, $x_{i,t}$ is the trait value of species *i* at time *t*, $\alpha_{ii}$ is the intraspecific competition coefficient (the per capita impact of species *i* on itself) and $\alpha_{ij}$ is the interspecific competition coefficient.
 Populations have a critical density $N_c$, below which the population is subject to extinction due to demographic stochasticity at a probability of *p* [@Gomulkiewicz1995].
 
-Results
-===============================================================================
+This model was used in Pantel & Becks [@Pantel2023] to evaluate the consequences of adaptive evolution for coexistence in a three-species system.
+To expand this model to a metacommunity, we consider the evolution of multiple species (*S* = 15) in a landscape of patches (*k* = 50).
+The patches have values of an environmental property $E \sim U(0,1)$ that determines the local optimum phenotype *E* = $x_{opt}$, i.e.
+where species experience their absolute fitness $W_{max}$, and patches also have spatial locations *X* and *Y* (both drawn from $U(0, 1)$).
+The patches thus have a connectivity matrix **D** (here given by their Euclidean distance), as well as a connectivity matrix **C** that is a Gaussian function of **D** and species dispersal rate $d_i$.
+
+We used the evolving metacommunity model to test whether landscape connectivity (the overall dispersal rate $d$) and the speed of evolution (evaluated as $h^2$) influenced the relative importance of trait evolution ($|\Delta x|$) for metacommunity composition (evaluated as $V_{|\Delta x|}$, the proportion of variation in species abundances explained by $|\Delta x|$ as estimated by the HMSC model fit to the simulated data across $d$ and $h^2$ levels).
+We hypothesized that the overall importance of trait evolution ($V_{|\Delta x|}$) would decrease with increasing dispersal rate - at low dispersal, locally mal-adapted species are less likely to emigrate to more optimal patches and will experience strong local selection.
+Population sizes in the mal-adapted populations will also likely be lower, leading to reduced relative importance of density-dependent intra- and interspecific competition.
+We also hypothesized that heritability ($h^2$) would mediate the relative importance of dispersal rate - at higher $h^2$, species will evolve quickly and we expected the importance of trait evolution $V_{|\Delta x|}$ to depend on the amount of time that mal-adapted populations persist.
+Therefore simulations run using higher $h^2$ would have decreased importance of evolution $V_{|\Delta x|}$ for community composition.
+For the same reason, we hypothesized generally that, for a given dispersal rate $d$, the importance of evolution $V_{|\Delta x|}$ would decrease with increasing $h^2$.
+
+All simulations used the same parameter values ($W_{max}$ = 2, *P* = 1, *w* = 2, $N_c$ = 100, *p* = 0.001), species interaction matrix ($\alpha_{ii} = 0.00125$; $\alpha_{ij} \sim U(0.0015,1)$), and initial conditions (initial richness $s_{0k} \sim Pois(0.75)$; initial population size $N_{0i} \sim Pois(10)$; initial degree of maladaptation $\beta_{i0} \sim Gamma(0.75,1)$, initial distance to the local patch's optimum phenotype $d_{i0}=\sqrt{B_{i0}(w+P)}$, initial trait value $x_{i0} = E_k - d_{i0}$; Appendix S4).
+Simulations were run across a range of dispersal values (identical for all species; *d* = 0, $10^{-9},10^{-8},10^{-7},10^{-6},10^{-5},10^{-4},10^{-3},10^{-2}$, 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 1) and heritability values (identical for all species: $h^2$ = 0, 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 1) for 200 time steps, producing records of $N_{ikt}$ and $x_{ikt}$.
+
+###Statistical model
+
+To determine the relative influence of environmental and spatial properties, intra- and interspecific population dynamics, and evolutionary dynamics for community structure, we applied hierarchical modelling of species communities (HMSC, [@Ovaskainen2017HMSC] [@Tikhonov2020]) to model the log abundance at each site and year for each species $ln(N_{ikt})$.
+The abundance values were modeled as $N_{ikt} \sim N(L_{ikt},\sigma_2)$, where the linear predictors **L** are the sum of fixed and random effects **L^F^** + **L^R^** [@Tikhonov2020].
+The fixed effects were $E_k$, $E_k^2$, and the absolute value of the change in each species' trait value from the previous time step to the next $|\Delta x_{ik,t-1 \to t}|$.
+Random effects included time (the day of the sample) and space.
+The spatial latent variables represent a spatial model, where species respond to some latent spatial predictor associated with the *xy* coordinates of the sites [@Tikhonov2020].
+
+# Results
 
@@
 The HMSC model was successfully modelled changes in species abundances when trait evolution was introduced (model *v*; Figure 1e \@ref(fig:fig1)).
-Fixed effect coefficients for the environmental driver and the total trait change from one time to the next (|\Delta x|) matched well with observed data (Figure 1f,g).
+Fixed effect coefficients for the environmental driver and the total trait change from one time to the next (\|\Delta x\|) matched well with observed data (Figure 1f,g).
 The model estimated the relative importance of the environmental driver, trait evolution in both species, and species interactions (Figure 1h), and predicted the observed species abundances well (Figure 1e).
@@
 
-We simulated growth and competition dynamics for a multi-species assemblage in a patchy landscape, with site variation in one environmental property.
+The simulation model followed population size and trait value dynamics for species in the landscape across 200 time steps (Figure 4a \@ref(fig:fig4)).
+Resulting biodiversity dynamics depended on the dispersal level and heritability values (Figure 4b \@ref(fig:fig4)).
+For dispersal, results mirrored some expectations from existing metacommunity models, i.e.
+that regional $\gamma$ diversity decreases with increasing dispersal level ([@Mouquet2003]; Figure 4b \@ref(fig:fig4)).
+The inclusion of trait heritability had a very strong effect on community diversity, reflecting that local adaptation can rescue species and contribute to richness at the local and metacommunity scale.
+Species diversity at the regional $\gamma$ level increased with increasing $h^2$ levels (Figure 4b \@ref(fig:fig4)).
 
-### Environmental variation
+Analysis using the HMSC model for community time series across $d$ and $h^2$ levels indicated that the importance of evolution for community composition ($V_{|\Delta x|}$) decreased with increasing dispersal and with increasing heritability (Figure 4c,d,e \@ref(fig:fig4)).
 
-Population growth for spec
+```{r, out.width="90%", include=TRUE, fig.align="center", fig.cap=c("your caption"), echo=FALSE}
+knitr::include_graphics("../output/fig4.pdf")
+```
 
-Discussion
-================================================================================
+# Discussion
 
@@
 
-References
-===============================================================================
+# References
 
-<div id="refs"></div>
+::: {#refs}
+:::
 
-<a name="appendix">Figures & Tables</a>
-================================================================================
+# <a name="appendix">Figures & Tables</a>

### 2024-11-10 manuscript/manuscript_v11.Rmd
Addition of empirical data

@@
 
-###Statistical model
+### Statistical model
 
@@
 The spatial latent variables represent a spatial model, where species respond to some latent spatial predictor associated with the *xy* coordinates of the sites [@Tikhonov2020].
+
+### Applying HMSC on empirical data
+
+To test the applicability of HMSC on ecological and evolutionary effects in empirical data, we applied HMSC on the experimental data of Jewel & Bell (2023) [@Jewell2023].
+The authors conducted a community selection experiments in which they grew four floating and short-lived aquatic plants (the angiosperms *Lemna minor* here designated as Lm, *Spirodela polyrhiza* designated as Sp and *Wolffia columbiana* designated as Wc, and the liverwort *Ricciocarpos natans* designated as Rn) under different environmental conditions for twelve weeks.
+The environmental conditions were defined by three differed amount of shadings of the water pool, 3%, 12% and 50%, and different nutrient concentrations, with low (dissolved nitrogen (DN) = 200 μg L−1, dissolved phosphorous (DP) = 10 μg L−1)), medium (DN = 800 μg L−1, DP = 40 μg L−1), and high concentration (DN = 3200 μg L−1, DP = 160 μg L−1) of nutrients.
+They combined each shading with each nutrient level, resulting in nine different environmental conditions.
+Throughout the experiment they measured biweekly, from week one to week eleven, the frond size of ten individuals from each species, as body size trait, in each environmental condition alongside their corresponding species density.
+We calculated the average trait change as $\Delta x$ = $\^{x}_{ijt} - \^{x}_{ijt-1}$, where $\^{x}_{ij}$ described the mean trait of species $i$ in environment $j$ and week $t$.
+As no $\Delta \^{x}$ value could be calculated for week one of the experiments, it was omitted from the analysis.
+We fitted two HMSC models to the data: An 'evolution model' in which $\Delta \^{x}$ of each species and environmental conditions were applied as explanatory variable, and a 'no-evolution model', in which only the environmental conditions were included.
+Species density was log-transformed species before applying it to the model, therefore we could use normal distribution in the model.
+Additionally, we applied replicates (*n* = 2) and time as random effects.
+For each model we ran four Markov chain Monte-Carlos' for 1,875,000 iterations from which the first 625,000 were cut off as the transient phase and with a thinning factor of 5,000 250 samples were taken from each chain.
+To test the impact of evolution by the model, we extracted the posterior distributions of the estimated beta effects.
+These were multiplied with the $\Delta \^{x}$ of the corresponding species and summed up together with the intercept.
 
@@
 
+### Empirical data
+
+The empirical data [@Jewell2023] showed the community dynamics of four plant species under different envrionmental conditions, additionally the front size of each species was measured to determine its trait change over time.
+We calculated the mean variance partitioning (VP) over all species, to evaluate the impact of evolutionary (change of the front size trait) and ecological (light and nutrient treatments) effects on changes on species density.
+Almost one fifth (18.6%) of the variance in species density could be explained by the variance in $\Delta \^{x}$ (Fig.
+Xa).
+This is due to the reduction of the VP of the ecological effects by a total of 14.4% (from 37% to 26.6% for the amount of shading and from 31% to 27.1% for the nutrient concentrations) and the reduction of 4.2% of the random effects (from 18.7% to 15% for time and from 13.3% to 12.8% for replicates) (Fig, X).
+While $\Delta \^{x}$ of the species Lm, Sp and Wc showed a similarly low impact, on average 3.6%, $\Delta \^{x}$ of species Rn showed a high impact on variance of species density, 7.8%, which can be explained by its high $\Delta \^{x}$ changes over time (Fig.
+SX).
+The high density intervals (HDI) of 95% estimated beta effects showed a negative correlation of Light shading with species density, while nutrient concentration correlated positively with density (Fig.
+Xc).
+Changes in $\Delta \^{x}$ impacted densities differently.
+At one hand changes of species Rn frond size correlated negatively with is density, on the other hand changes in the frond size of species Wc showed a positive correlation with the other species Sp and Lm (Fig.
+Xc).
+The impact of evolution (of all species summed up), impacted species density over time, yet did not show a clear direction over the course of the experiment (Fig.
+Xd).
+
+```{r, out.width="90%", include=TRUE, fig.align="center", fig.cap=c("The estimated effects of species trait changes and environmental treatments.
+The variance partitioning of all fixed effects of the HMSC model (a) including trait changes with envrionmental treatments and (b) only including the environmental treamtents.
+The (b) high density interval of 95% showed a clear pattern for the different environmental treatments, yet differing results for trait changes and their impact on species density.
+This can be seen in (d) the shifting predictions of density over time due to the total trait changes of all species."), echo=FALSE}
+knitr::include_graphics("../output/fig5.pdf")
+```
+
 # Discussion
 
-XXXXX
+### Empirical data
+
+By implementing the trait changes of the four different species (Lm, Rn, Sp and Wc) in the HMSC model we were able to inferr the relative impact of the trait changes on the community dynamics.
+This application shows is advantages over previous methods,which estimate the impact of trait changes (thus evolution), as it is not depended on the data structure and simplicity [@Hairston2005] [@Ellner2011], can be implemented on multiple species[@Lasky2020]] and does not require additional common garden experiments [@Garzon2019].
+Therefore this allows for a straight forward implentation of HMSC on a variety of data structures as lone as a clear response variable (e.g.
+species density) and ecological (e.g.
+environmental conditions) as well as evolutionary (e.g.
+heritable trait change) are measured.
+An addtional advantage of HMSC, and JSDMs in general, is the ability to use the empirical data to predict a species response to changes in the environmental variables, even beyond the measured values [@Tikhonov2020] [@Ovaskainen2015].
+With the implementation this approach can be combined with the fixed effects to answer furter questions, such as: How does trait change impact the species density under changing environmental conditions?
+And how does the impact of trait changes depend on specific environmental condtions?
+
+At the same time a correct interpretation of such predictions as well es the estimated posterior beta distribution require a thorough understand of the experimental system.
+For example we often observed a negative correlation between trait change and species density in our simulations and also in species Rn of the empirical data, yet they derived from different mechanisms.
+In our simulations trait changes were greater when a species trait was further away from the optimal trait value, this greater distance resulted also in lower fitness and thus lower population size, while a species with higher ftiness and thus higher population size had smaller trait changes.
+Therefore, HMSC coirrelates a high trait changes with a low popualtion size and thus estimates a negative beta effect.
+Such effects can be expected in empirical data of i.e.
+evolutionary rescue experiments, where a species has to adapt to a deleterious environment before going extinct [@Gomulkiewicz1995][@Bell2009].
+In the empirical data this negative correlation was the result of the speceis Rn investing into a larger frond size (Fig.
+SX), this likely lead to a lower investment into reproduction, while still mantaining a big surface on the water surface to capture the reduced light.
 
@@
 # <a name="appendix">Figures & Tables</a>
+ 7 7 7 7 7 7 7 7 7 7 7 7 7 7 7 7 7 7 7 7 7 7 7 7 7 7 7 7 7 7 7 7 7 7 7

### 2024-11-11 code/R/fig4.R
Removed spatial RanEf from HMSC model for d=0

@@ -84,7 +84,27 @@ h2_vals = c(0, 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 1)
 ## Get VarPart results across HMSC output
 begin <- seq(1,36,4)
 counter <- 1
-for(v.d in c(1,4,8)){ # dlev
+for(v.d in c(1)){ # dlev
+  for(v.h2 in c(2,4,10)){ # hlev
+    ## Read in population size values
+    result <- paste("/Users/jhpantel/Nextcloud/Pantel/data/mc_v02/",h_lev[v.h2],d_lev[v.d],"_hmsc_v02.RData",sep="")
+    load(result)
+    s <- ncol(m.1$Y)
+    VP <- computeVariancePartitioning(m.1, group = c(1,1,rep(2,s)), groupnames = c("Env","deltaX"))
+    ng = dim(VP$vals)[1]
+    leg = VP$groupnames
+    for (r in 1:m.1$nr) {
+      leg = c(leg, paste("Random: ", m.1$rLNames[r], sep = ""))
+    }
+    means = round(100 * rowMeans(VP$vals), 1)
+    means = c(means[1],means[2],0,means[3])
+    data_bar[begin[counter]:(begin[counter]+3),3] <- means
+    data_bar[begin[counter]:(begin[counter]+3),1] <- d_vals[v.d]
+    data_bar[begin[counter]:(begin[counter]+3),2] <- h2_vals[v.h2]
+    counter <- counter + 1
+  }
+}
+for(v.d in c(4,8)){ # dlev
   for(v.h2 in c(2,4,10)){ # hlev
     ## Read in population size values
     result <- paste("./data/mc_v02/",h_lev[v.h2],d_lev[v.d],"_hmsc_v02.RData",sep="")

### 2024-11-11 manuscript/Appendix_S5_v01.Rmd

---
  title: "Appendix S5. The interpretation of the estimated trait change beta"
author:
  - Jelena H. Pantel^[Laboratoire Chrono-environnement,UMR 6249 CNRS-UFC, 16 Route de Gray, 25030 Besançon cedex, France, jelena.pantel@univ-fcomte.fr]
- Ruben J. Hermann^[University of Duisburg-Essen, Universitätsstraße 5, 45141 Essen, Germany, ruben.hermann@uni-due.de]
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::pdf_document2:
  toc: no
fig_caption: yes
latex_engine: xelatex
extra_dependencies: ["float"]
bookdown::html_document2: default
word_document:
  reference_docx: docx_template.docx
header-includes:
  - \usepackage{fancyhdr}
- \pagestyle{fancy}
- \fancyhead[L]{Author Pantel and Hermann}
- \fancyhead[R]{Detection of ecoevo dynamics}
- \usepackage{lineno}
- \linenumbers
link-citations: yes
linkcolor: blue
csl: the-american-naturalist.csl
bibliography: refs.bib
---

  ```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
```{r setup,echo=FALSE,message=FALSE}
library(ecoevor)
library(ggplot2)
library(reshape2)
library(Hmsc)
library(bayesplot)
```
```{r,echo=FALSE,message=FALSE,warning=FALSE}
# Wrap long lines
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy=TRUE,out.width="50%")
```

# Empirical data

To correctly interpret correlation of trait change with population density of the species (as seen in Fig. X) we first need to check the raw data.

```{r, out.width="90%", include=TRUE, fig.align="center", fig.cap=c("Figure S05 1"), echo=FALSE}
knitr::include_graphics("../output/fig_S05_1.pdf")
```

From the empirical data we can detect that species Rn is drastically increasing in frond size over the course of the experiment, with a peak in trait change during week 5 and 7. Simultaneously it is only slowly increasing in population density, with slightly lower density during week 5. This leads to the highest value in $\Delta \hat{x}$ during a lower value in the population density of species Rn. Similarly species Wc had great trait changes, during time points when species Lm and Sp showed an increase in population density. HMSC draws a linear correlation between both and in (\@ref(fig:fig2)) it can be seen how HMSC infers these correlations (the logarithmic scale of $\Delta \hat{x}$ was used for clearance).

```{r, out.width="90%", include=TRUE, fig.align="center", fig.cap=c("Figure S05 2"), echo=FALSE}
knitr::include_graphics("../output/fig_S05_2.pdf")
```

What should be noted is that, species Rn was the only to increase in frond size over this experiment, which is likely the result of its decreased in population size. At the same time species Wc decreased in frond size, thus occupying less space on the water surface thus freeing space for the other species to occupy. Though this interpretation needs to be taken with caution as more information is missing for a more careful interpretation.

### 2024-11-11 manuscript/manuscript_v11.Rmd

--- 2024-11-10 manuscript/manuscript_v11.Rmd
@@
 The spatial latent variables represent a spatial model, where species respond to some latent spatial predictor associated with the *xy* coordinates of the sites [@Tikhonov2020].
+
+### Applying HMSC on empirical data
+
+To test the applicability of HMSC on ecological and evolutionary effects in empirical data, we applied HMSC on the experimental data of Jewel & Bell (2023) [@Jewell2023].
+The authors conducted a community selection experiments in which they grew four floating and short-lived aquatic plants (the angiosperms *Lemna minor* here designated as Lm, *Spirodela polyrhiza* designated as Sp and *Wolffia columbiana* designated as Wc, and the liverwort *Ricciocarpos natans* designated as Rn) under different environmental conditions for twelve weeks.
+The environmental conditions were defined by three differed amount of shadings of the water pool, 3%, 12% and 50%, and different nutrient concentrations, with low (dissolved nitrogen (DN) = 200 μg L−1, dissolved phosphorous (DP) = 10 μg L−1)), medium (DN = 800 μg L−1, DP = 40 μg L−1), and high concentration (DN = 3200 μg L−1, DP = 160 μg L−1) of nutrients.
+They combined each shading with each nutrient level, resulting in nine different environmental conditions.
+Throughout the experiment they measured biweekly, from week one to week eleven, the frond size of ten individuals from each species, as body size trait, in each environmental condition alongside their corresponding species density.
+We calculated the average trait change as $\Delta x$ = $\^{x}_{ijt} - \^{x}_{ijt-1}$, where $\^{x}_{ij}$ described the mean trait of species $i$ in environment $j$ and week $t$.
+As no $\Delta \hat{x}$ value could be calculated for week one of the experiments, it was omitted from the analysis.
+We fitted two HMSC models to the data: An 'evolution model' in which $\Delta \hat{x}$ of each species and environmental conditions were applied as explanatory variable, and a 'no-evolution model', in which only the environmental conditions were included.
+Species density was log-transformed species before applying it to the model, therefore we could use normal distribution in the model.
+Additionally, we applied replicates (*n* = 2) and time as random effects.
+For each model we ran four Markov chain Monte-Carlos' for 1,875,000 iterations from which the first 625,000 were cut off as the transient phase and with a thinning factor of 5,000 250 samples were taken from each chain.
+To test the impact of evolution by the model, we extracted the posterior distributions of the estimated beta effects.
+These were multiplied with the $\Delta \hat{x}$ of the corresponding species and summed up together with the intercept.
 
@@
 
+### Empirical data
+
+The empirical data [@Jewell2023] showed the community dynamics of four plant species under different environmental conditions, additionally the front size of each species was measured to determine its trait change over time.
+We calculated the mean variance partitioning (VP) over all species, to evaluate the impact of evolutionary (change of the front size trait) and ecological (light and nutrient treatments) effects on changes on species density.
+Almost one fifth (18.6%) of the variance in species density could be explained by the variance in $\Delta \hat{x}$ (\@ref(fig:fig5)a).
+This is due to the reduction of the VP of the ecological effects by a total of 14.4% (from 37% to 26.6% for the amount of shading and from 31% to 27.1% for the nutrient concentrations) and the reduction of 4.2% of the random effects (from 18.7% to 15% for time and from 13.3% to 12.8% for replicates) (\@ref(fig:fig5)).
+While $\Delta \hat{x}$ of the species Lm, Sp and Wc showed a similarly low impact, on average 3.6%, $\Delta \hat{x}$ of species Rn showed a high impact on variance of species density, 7.8%, which can be explained by its high $\Delta \hat{x}$ changes over time (see Appendix S05).
+The high density intervals (HDI) of 95% estimated beta effects showed a negative correlation of Light shading with species density, while nutrient concentration correlated positively with density (\@ref(fig:fig5)c).
+Changes in $\Delta \hat{x}$ impacted densities differently.
+At one hand changes of species Rn frond size correlated negatively with is density, on the other hand changes in the frond size of species Wc showed a positive correlation with the other species Sp and Lm (\@ref(fig:fig5)c).
+The impact of evolution (of all species summed up), impacted species density over time, yet did not show a clear direction over the course of the experiment (\@ref(fig:fig5)d).
+
+```{r, out.width="90%", include=TRUE, fig.align="center", fig.cap=c("The estimated effects of species trait changes and environmental treatments.
+The variance partitioning of all fixed effects of the HMSC model (a) including trait changes with envrionmental treatments and (b) only including the environmental treamtents.
+The (b) high density interval of 95% showed a clear pattern for the different environmental treatments, yet differing results for trait changes and their impact on species density.
+This can be seen in (d) the shifting predictions of density over time due to the total trait changes of all species."), echo=FALSE}
+knitr::include_graphics("../output/fig5.pdf")
+```
+
 # Discussion
 
-XXXXX
+### Empirical data
+
+By implementing the trait changes of the four different species (Lm, Rn, Sp and Wc) in the HMSC model we were able to infer the relative impact of the trait changes on the community dynamics.
+This application shows is advantages over previous methods,which estimate the impact of trait changes (thus evolution), as it is not depended on the data structure and simplicity [@Hairston2005] [@Ellner2011], can be implemented on multiple species[@Lasky2020] and does not require additional common garden experiments [@Garzon2019].
+Therefore this allows for a straight forward implementation of HMSC on a variety of data structures as lone as a clear response variable (e.g.
+species density) and ecological (e.g.
+environmental conditions) as well as evolutionary (e.g.
+heritable trait change) are measured.
+An additional advantage of HMSC, and JSDMs in general, is the ability to use the empirical data to predict a species response to changes in the environmental variables, even beyond the measured values [@Tikhonov2020] [@Ovaskainen2015].
+With the implementation this approach can be combined with the fixed effects to answer further questions, such as: How does trait change impact the species density under changing environmental conditions?
+And how does the impact of trait changes depend on specific environmental conditions?
+
+At the same time a correct interpretation of such predictions as well es the estimated posterior beta distribution require a thorough understand of the experimental system.
+For example we often observed a negative correlation between trait change and species density in our simulations and also in species Rn of the empirical data, yet they derived from different mechanisms.
+In our simulations trait changes were greater when a species trait was further away from the optimal trait value, this greater distance resulted also in lower fitness and thus lower population size, while a species with higher fitness and thus higher population size had smaller trait changes.
+Therefore, HMSC correlates a high trait changes with a low population size and thus estimates a negative beta effect.
+Such effects can be expected in empirical data of i.e.
+evolutionary rescue experiments, where a species has to adapt to a deleterious environment before going extinct [@Gomulkiewicz1995][@Bell2009].
+In the empirical data this negative correlation was the result of the species Rn investing into a larger frond size (Fig.
+SX), this likely lead to a lower investment into reproduction, while still maintaining a similar total frond area on the water surface to capture the reduced light.
 

### 2024-11-11 code/R/fig4.R
Updated fig4 no spatial RanEf for d=0

@@ -87,7 +87,7 @@ counter <- 1
 for(v.d in c(1)){ # dlev
   for(v.h2 in c(2,4,10)){ # hlev
     ## Read in population size values
-    result <- paste("/Users/jhpantel/Nextcloud/Pantel/data/mc_v02/",h_lev[v.h2],d_lev[v.d],"_hmsc_v02.RData",sep="")
+    result <- paste("/Users/jhpantel/Nextcloud/Pantel/data/",h_lev[v.h2],d_lev[v.d],"_hmsc_v02_nospace.RData",sep="")
     load(result)
     s <- ncol(m.1$Y)
     VP <- computeVariancePartitioning(m.1, group = c(1,1,rep(2,s)), groupnames = c("Env","deltaX"))

### 2024-11-11 code/R/fig4.R
Updated code and fig4 to exclude spatial RanEf when d=0

@@ -87,7 +87,7 @@ counter <- 1
 for(v.d in c(1)){ # dlev
   for(v.h2 in c(2,4,10)){ # hlev
     ## Read in population size values
-    result <- paste("/Users/jhpantel/Nextcloud/Pantel/data/",h_lev[v.h2],d_lev[v.d],"_hmsc_v02_nospace.RData",sep="")
+    result <- paste("./data/mc_v02/",h_lev[v.h2],d_lev[v.d],"_hmsc_v02_nospace.RData",sep="")
     load(result)
     s <- ncol(m.1$Y)
     VP <- computeVariancePartitioning(m.1, group = c(1,1,rep(2,s)), groupnames = c("Env","deltaX"))

### 2024-11-11 manuscript/manuscript_v11.Rmd
Draft mscript11

@@
 author:
-  - Jelena H. Pantel^[Laboratoire Chrono-environnement,UMR 6249 CNRS-UFC, 16 Route de Gray, 25030 Besançon cedex, France, jelena.pantel@univ-fcomte.fr]
-  - Ruben J. Hermann^[University of Duisburg-Essen, Universitätsstraße 5, 45141 Essen, Germany, ruben.hermann@uni-due.de]
+- "Jelena H. Pantel^[Laboratoire Chrono-environnement,UMR 6249 CNRS-UFC, 16 Route
+  de Gray, 25030 Besançon cedex, France, jelena.pantel@univ-fcomte.fr]"
+- "Ruben J. Hermann^[University of Duisburg-Essen, Universitätsstraße 5, 45141 Essen,
+  Germany, ruben.hermann@uni-due.de]"
 date: "`r format(Sys.time(), '%d %B, %Y')`"
 output:
+  word_document:
+    reference_docx: docx_template.docx
+  bookdown::html_document2: default
   bookdown::pdf_document2:
@@
     latex_engine: xelatex
-    extra_dependencies: ["float"]
-    number_sections: true
-  bookdown::html_document2: default
-  word_document:
-    reference_docx: docx_template.docx
+    extra_dependencies: float
+    number_sections: yes
+  pdf_document:
+    toc: no
 header-includes:
@@
 linkcolor: blue
-csl: the-american-naturalist.csl
+csl: "the-american-naturalist.csl"
 bibliography: refs.bib
@@
 library(Hmsc)
+library(knitr)
+library(kableExtra)
+library(formatR)
+library(rmarkdown)
 ```
@@
 The dynamics of populations and communities, and resulting biodiversity patterns, are also influenced by the genotypic and phenotypic properties of organisms as well.
-Although traits can be considered in some JSDMs, they typically require fixed values for species' traits ([@Ovaskainen2017HMSC] [@Tikhonov2020]; but see @Abrego2024).
-However, ecological dynamic processes can impact selection on genes and phenotypes, potentially resulting in interactions or feedback loops between ecological and evolutionary processes [@Lion2018] [@Barbour2022].
+Although traits can be considered in some JSDMs, they typically require fixed values for species' traits (e.g.
+[@Ovaskainen2017HMSC] [@Tikhonov2020]; but see @Abrego2024).
+However, ecological dynamics can impact selection on genes and phenotypes, potentially resulting in interactions or feedback loops between ecological and evolutionary processes [@Lion2018] [@Barbour2022].
 Although an increasing number of theoretical and empirical studies have evaluated the impact of evolutionary dynamics for community assembly and metacommunity dynamics [@Loeuille2008] [@Osmond2013] [@Pantel2015] [@Toju2017], analytical methods to consider this additional complexity lag behind the scope of data collected in eco-evolutionary studies.
-Some methods exist that ultimately apply an ANOVA framework to partition variance in traits or growth rates across contributions of ecological and evolutionary components ([@Hairston2005] [@Ellner2011] [@Govaert2016]).
-These methods tend to be cited more frequently for their potential than for their actual application to empirical data (but see [@terHorst2014] [@Hiltunen2014]), because this approach is ultimately limited by the same requirements that can limit application of ANOVA - data should follow a normal distribution, the processes that structure the target response variable must be linear and additive, and complicated structuring mechanisms must be reduced into one or a few target categories labeled as 'ecology' or 'evolution'.
+Some methods exist that ultimately apply ANOVA to partition variance in traits or growth rates across contributions of ecological and evolutionary components ([@Hairston2005] [@Ellner2011] [@Govaert2016]; additional methods for continuous data are also described in @Ellner2011).
+These methods are more frequently cited for their potential rather than for their actual application to empirical data (but see [@terHorst2014] [@Hiltunen2014]), because this approach is ultimately limited by the same requirements that can limit application of ANOVA - data should follow a normal distribution, the processes that structure the target response variable must be linear and additive, and complicated structuring mechanisms must be reduced into one or a few target categories labeled as 'ecology' or 'evolution'.
 Flexible models, which can vary depending on the mechanisms and structure of the diverse kinds of data that most studies of eco-evolutionary dynamics will collect, are needed for rigorous inference about eco-evolutionary processes [@Pantel2023].
@@
 truly disentangle the role that plastic and evolutionary trait divergence can play in species distributions, but they also require common garden experimental data across sites to estimate these effects.
-Observational data collected in metacommunity surveys may only have observational data across sites or time points, as large-scale experimental measures of reaction norms required by the $\Delta$ SDM model may be difficult to obtain.
+Observational data collected in metacommunity surveys may only have survey data across sites or time points, as large-scale experimental measures of reaction norms required by the $\Delta$ SDM model may be difficult to obtain.
 
-One particular metacommunity statistical model, the Hierarchical Modelling of Species Communities (HMSC) framework [@Ovaskainen2017HMSC] [@Tikhonov2020], explicitly considers the multitude of processes that can structure community composition across time and space [@Leibold2022] and is thus well suited to consider analysis of metacommunity eco-evolutionary data.
+One particular metacommunity statistical model, the Hierarchical Modelling of Species Communities (HMSC) framework [@Ovaskainen2017HMSC] [@Tikhonov2020], explicitly considers the multitude of processes that can structure community composition across time and space [@Leibold2022] and is potentially well suited to consider analysis of metacommunity eco-evolutionary processes.
 This model of environmental filtering via variation and covariation in how species respond to their environment considers the impacts of (fixed) trait values and phylogenetic relationships, and uses latent variables to account for the numerous unobserved environmental and spatial features that are difficult to exhaustively measure in community surveys.
-The associated R package has immense flexibility, and has been applied to study environmental and abiotic drivers of diverse assemblages of organisms (e.g.
+It has been demonstrated that this statistical model reflects the diverse array of community assembly and metacommunity processes that structure species abundances (i.e.
+species sorting, environmental filtering, priority effects; [@Little2018] [@Ovaskainen2020] [@Leibold2022]).
+The associated R package has been applied to study environmental and abiotic drivers of diverse assemblages of organisms (e.g.
 [@Sandal2022] [@Weigel2023]).
@@
 
-In this study, we use simulations of population dynamics, environmental drivers, species interactions, and trait evolution in populations, communities, at the local and metacommunity scale, to generate simulated time series of population sizes and mean trait values.
-We then use HMSC to analyze the resulting data, to determine whether HMSC can successfully estimate the impacts of trait evolution for community composition, and to generate relative contributions of fixed environmental, spatial, and trait evolution drivers as well as random effects of spatial and temporal covariance among species.
-We asked three main questions: (i) do linear statistical models (like HMSC) accurately capture the impact of non-linear population, community, and evolutionary dynamic processes?, (ii) can HMSC be used to evaluate hypothesized drivers of the relative importance of evolution for community composition at the metacommunity scale?, and (iii) can we better understand the role of evolution for community dynamics in experimental systems using HMSC?
+In this study, we use simulations of population dynamics, environmental drivers, species interactions, and trait evolution in populations, communities, at the local and metacommunity scale, to study eco-evolutionary dynamics.
+After generating simulated time series of population sizes and mean trait values, we then use HMSC to analyze the resulting data, to determine whether HMSC can successfully estimate the impacts of trait evolution for community composition, and to generate relative contributions of fixed environmental, spatial, and trait evolution drivers as well as random effects of spatial and temporal covariance among species.
+We asked three main questions: (i) can linear statistical models (like HMSC) accurately capture the impact of non-linear population, community, and evolutionary dynamic processes?, (ii) can HMSC be used to evaluate hypothesized drivers of the relative importance of evolution for community composition at the metacommunity scale?, and (iii) can we better understand the role of evolution for community dynamics in experimental systems using HMSC?
 
@@
 We evaluated whether linear regression models can appropriately capture the dynamics that drive species abundances in relatively simple populations and communities.
-We fit data generated from simulation models with increasing complexity to Bayesian linear regression models.
-These simulation models were (i) a single-species model of discrete-time logistic population growth, (ii) expanded to include a randomly fluctuating environmental property $E$ that impacts the species' population growth rate, (iii) a two-species model of logistic growth and competition that was then (iv) expanded to consider a randomly fluctuating environmental property, (v) and further expanded to consider evolution in a trait $x$ that is selected by the environmental property, and finally (vi) the model of two-species growth, competition, environmental change, and evolution in a spatially structured (10-patch) habitat.
+We fit data generated from simulation models with increasing complexity (\@ref(tab:tab1)) to Bayesian linear regression models.
+These simulation models were (A) a single-species model of discrete-time logistic population growth, (B) expanded to include a randomly fluctuating environmental property $E$ that impacts the species' population growth rate, (C) a two-species model of logistic growth and competition that was then (D) expanded to consider a randomly fluctuating environmental property, (E) and further expanded to consider evolution in a trait $x$ that is selected by the environmental property, and finally (F) the model of two-species growth, competition, environmental change, and evolution in a spatially structured (10-patch) habitat.
 
-For linear regression models to fit to the simulated population size data, we began with models using population size and environment as fixed effects ($N_t \sim N_{t+1} + E_t + E_t^2$), and progressed to a mixed model using environment and trait evolution as fixed effects ($N_t \sim E_t + E_t^2 + |\Delta x|$) and time and spatial distance as random effects.
+```{r tab1, echo=FALSE}
+df <- data.frame(model=c("1 species, logistic","1 species, logistic, env","2 species, logistic, competition","2 species, logistic, competition, env","2 species, logistic, competition, env, trait evolution","2 species, logistic, competition, env, trait evolution, spatial structure"), row.names=LETTERS[1:6])
+
+df$formula <- c("$N_{i,t+1} = \\frac{r_i N_{i,t}}{1+\\alpha_{ii} N_{i,t}}$",
+                "A, $r_i = \\hat{W}e^{-(E_t - x_{i,t})^2}$",
+                "$N_{i,t+1} = \\frac{r_i N_{i,t}}{1+\\alpha_{ii} N_{i,t} + \\alpha_{ij} N_{j,t}}$",
+                "C, $r_i = \\hat{W}e^{-(E_t - x_{i,t})^2}$",
+                "C, $r_i = \\hat{W}e^\\frac{-[(\\frac{w+(1-h^2)P}{P+w})(E_t - x_{i,t})]^2}{2 (P+w)}$",
+                "E, $N_{i,k,t+1} = N_{i,k,t} + \\sigma_k + \\sigma_{kl}$")
+kable(df, "latex", align="l", booktabs=TRUE, escape = F, caption = 'Models for population growth')  %>%
+footnote(general=c("$i,j=$","species,","$t=$","time,","$k,l=$","site"),
+         number=c("See Appendix S1 for parameter values used in simulation"),
+         footnote_as_chunk=TRUE, 
+         escape=FALSE)
+```
+
+To fit linear regression models to the simulated population size data, we began with models using population size and environment (including the second-order term, to better reflect species optimum values) as fixed effects ($N_t \sim N_{t+1} + E_t + E_t^2$; for simulation models *A* and *B*), and progressed to a mixed model using environment and trait evolution as fixed effects ($N_t \sim E_t + E_t^2 + |\Delta x|$; for simulation models *C*, *D*, and *E*) and time (models *C*, *D*, and *E*) and spatial distance (model *F*) as random effects.
 For consistency, all linear regression models were implemented as heirarchical Bayesian models using HMSC (Hierarchical Modelling of Species Communities [@Ovaskainen2017HMSC] [@Tikhonov2020]).
 HMSC uses a latent variable approach as an alternative to directly modelling species interaction coefficients, estimating a reduced number of linear combinations of species abundances that best predict future population size values.
-For linear models fit to data generated in simulations *iii*, *iv*, *v*, and *vi* above, the time random effect was implemented to capture species interactions (as part of the temporal species association matrix estimated by the model), and the spatial random effect was implemented in model *vi* to capture spatial covariance in abundances as a function of distance between sites.
+For linear models fit to data generated in simulations *C*, *D*, *E*, and *F* above, the time random effect was implemented to capture species interactions (as part of the temporal species association matrix estimated by the model), and the spatial random effect was implemented in model *F* to capture spatial covariance in abundances as a function of distance between sites.
 Full methodological details are given in Appendix S1.
@@
 
-###Simulation model
+### Simulation model
 
 We also simulated growth and competition dynamics for a multi-species assemblage in a patchy landscape, with site variation in one environmental property, and with evolution in a trait that is selected on by the environment.
-Population growth for species in the metacommunity simulation follows a Leslie-Gower model (a discrete-time version of a Lotka-Volterra model [@Beverton1957] [@Leslie1958]).
-We consider the impact of trait evolution for growth using a discrete time quantitative genetic model of evolutionary rescue [@Gomulkiewicz1995].
-The model for population size is as follows:
+Population growth for species in the metacommunity simulation followed a Leslie-Gower model (a discrete-time version of a Lotka-Volterra model [@Beverton1957] [@Leslie1958]).
+We considered the impact of trait evolution for growth using a discrete time quantitative genetic model of evolutionary rescue [@Gomulkiewicz1995].
+The model for population size was:
 
-$$ N_{i,t+1} =\frac{\hat{W}e^ \frac{-[(\frac{w+(1-h^2)P} {P+w})(E-x_{i,t})]^2}{2(P+w)}N_{i,t}} {1 + \alpha_{ii}N_{i,t} + \sum_{j \neq i}^{S} \alpha_{ij}N_{j,t}} $$ where $N_i,t$ is the population size of species *i* at time *t*, $\hat{W}$ is calculated as $\hat{W}=W_{max}\sqrt(\frac{w}{P+w})$, $W_{max}$ is the species' maximum per-capita growth rate, *w* is the width of the Gaussian fitness function (which determines the strength of selection, as increasing values indicate a weaker reduction in fitness with distance from optimum trait value), *P* is the width of the distribution of the phenotype *x*, $h^2$ is the heritability of the trait *x*, *E* is the local environmental optimum trait value, $x_{i,t}$ is the trait value of species *i* at time *t*, $\alpha_{ii}$ is the intraspecific competition coefficient (the per capita impact of species *i* on itself) and $\alpha_{ij}$ is the interspecific competition coefficient.
+$$ N_{i,t+1} =\frac{\hat{W}e^ \frac{-[(\frac{w+(1-h^2)P} {P+w})(E-x_{i,t})]^2}{2(P+w)}N_{i,t}} {1 + \alpha_{ii}N_{i,t} + \sum_{j \neq i}^{S} \alpha_{ij}N_{j,t}} $$ where $N_i,t$ is the population size of species *i* at time *t*, $\hat{W}$ is calculated as $\hat{W}=W_{max}\sqrt(\frac{w}{P+w})$, $W_{max}$ is the species' maximum per-capita growth rate, *w* is the width of the Gaussian fitness function (which determines the strength of selection), *P* is the width of the distribution of the phenotype *x*, $h^2$ is the heritability of the trait *x*, *E* is the local environmental optimum trait value, $x_{i,t}$ is the trait value of species *i* at time *t*, $\alpha_{ii}$ is the intraspecific competition coefficient (the per capita impact of species *i* on itself) and $\alpha_{ij}$ is the interspecific competition coefficient.
 Populations have a critical density $N_c$, below which the population is subject to extinction due to demographic stochasticity at a probability of *p* [@Gomulkiewicz1995].
@@
 
-We used the evolving metacommunity model to test whether landscape connectivity (the overall dispersal rate $d$) and the speed of evolution (evaluated as $h^2$) influenced the relative importance of trait evolution ($|\Delta x|$) for metacommunity composition (evaluated as $V_{|\Delta x|}$, the proportion of variation in species abundances explained by $|\Delta x|$ as estimated by the HMSC model fit to the simulated data across $d$ and $h^2$ levels).
-We hypothesized that the overall importance of trait evolution ($V_{|\Delta x|}$) would decrease with increasing dispersal rate - at low dispersal, locally mal-adapted species are less likely to emigrate to more optimal patches and will experience strong local selection.
-Population sizes in the mal-adapted populations will also likely be lower, leading to reduced relative importance of density-dependent intra- and interspecific competition.
+We used the evolving metacommunity model to test whether landscape connectivity (the overall dispersal rate $d$) and the speed of evolution (evaluated by varying $h^2$) influenced the relative importance of trait evolution ($|\Delta x|$) for metacommunity composition $N_{i,t,k}$ (evaluated as $V_{|\Delta x|}$, the proportion of variation in species abundances explained by $|\Delta x|$ as estimated by the HMSC model fit to the simulated data across $d$ and $h^2$ levels).
+We hypothesized that the overall importance of trait evolution ($V_{|\Delta x|}$) would decrease with increasing dispersal rate - at low dispersal rates, locally mal-adapted species are less likely to emigrate to more optimal patches and will experience strong local selection, leading to a large amount of trait evolution.
+Population sizes for these initially mal-adapted populations will also likely be lower, leading to reduced relative importance of density-dependent intra- and interspecific competition.
 We also hypothesized that heritability ($h^2$) would mediate the relative importance of dispersal rate - at higher $h^2$, species will evolve quickly and we expected the importance of trait evolution $V_{|\Delta x|}$ to depend on the amount of time that mal-adapted populations persist.
-Therefore simulations run using higher $h^2$ would have decreased importance of evolution $V_{|\Delta x|}$ for community composition.
+Therefore simulations with higher $h^2$ would have decreased importance of evolution $V_{|\Delta x|}$ for community composition.
 For the same reason, we hypothesized generally that, for a given dispersal rate $d$, the importance of evolution $V_{|\Delta x|}$ would decrease with increasing $h^2$.
@@
 To determine the relative influence of environmental and spatial properties, intra- and interspecific population dynamics, and evolutionary dynamics for community structure, we applied hierarchical modelling of species communities (HMSC, [@Ovaskainen2017HMSC] [@Tikhonov2020]) to model the log abundance at each site and year for each species $ln(N_{ikt})$.
-The abundance values were modeled as $N_{ikt} \sim N(L_{ikt},\sigma_2)$, where the linear predictors **L** are the sum of fixed and random effects **L^F^** + **L^R^** [@Tikhonov2020].
+The log-abundance values were modeled as $ln(N_{ikt}) \sim N(L_{ikt},\sigma_2)$, where the linear predictors **L** are the sum of fixed and random effects **L^F^** + **L^R^** [@Tikhonov2020].
 The fixed effects were $E_k$, $E_k^2$, and the absolute value of the change in each species' trait value from the previous time step to the next $|\Delta x_{ik,t-1 \to t}|$.
@@
 The spatial latent variables represent a spatial model, where species respond to some latent spatial predictor associated with the *xy* coordinates of the sites [@Tikhonov2020].
+When there was no dispersal in the system ($d=0$), spatial random effects were not included in the model.
 
@@
 The authors conducted a community selection experiments in which they grew four floating and short-lived aquatic plants (the angiosperms *Lemna minor* here designated as Lm, *Spirodela polyrhiza* designated as Sp and *Wolffia columbiana* designated as Wc, and the liverwort *Ricciocarpos natans* designated as Rn) under different environmental conditions for twelve weeks.
-The environmental conditions were defined by three differed amount of shadings of the water pool, 3%, 12% and 50%, and different nutrient concentrations, with low (dissolved nitrogen (DN) = 200 μg L−1, dissolved phosphorous (DP) = 10 μg L−1)), medium (DN = 800 μg L−1, DP = 40 μg L−1), and high concentration (DN = 3200 μg L−1, DP = 160 μg L−1) of nutrients.
+The environmental conditions were defined by three differed amount of shadings of the water pool, 3%, 12% and 50%, and different nutrient concentrations, with low (dissolved nitrogen (DN) = 200 $\mu g L^{-1}$, dissolved phosphorous (DP) = 10 $\mu g L^{-1}$, medium (DN = 800 $\mu g L^{-1}$, DP = 40 $\mu g L^{-1}$), and high concentration (DN = 3200 $\mu g L^{-1}$, DP = 160 $\mu g L^{-1}$) of nutrients.
 They combined each shading with each nutrient level, resulting in nine different environmental conditions.
 Throughout the experiment they measured biweekly, from week one to week eleven, the frond size of ten individuals from each species, as body size trait, in each environmental condition alongside their corresponding species density.
-We calculated the average trait change as $\Delta x$ = $\^{x}_{ijt} - \^{x}_{ijt-1}$, where $\^{x}_{ij}$ described the mean trait of species $i$ in environment $j$ and week $t$.
+We calculated the average trait change as $\Delta x = x_{ijt} - x_{ijt-1}$, where $x_{ij}$ described the mean trait of species $i$ in environment $j$ and week $t$.
 As no $\Delta \hat{x}$ value could be calculated for week one of the experiments, it was omitted from the analysis.
@@
 Linear regression models can estimate the impacts of intra- and interspecific competition, environmental variation, and trait evolution for population size in simple one- and two-species models of population growth.
-In simple cases (one-species models, linear and quadratic relationships between environment and population size), the regression coefficients successfully reflect the direction and magnitude of the relationships included in the non-linear simulation model (the impact of $N_t$ and $E_t$ for $N_{t+1}$).
-However, as the processes included in the simulation models become slightly more complex (competition with a second species, spatial structure), regression coefficients were useful for prediction but not for recreating the underlying processes driving changes in population size (Appendix S1; REF?).
+In simple cases (one-species models, linear and quadratic relationships between environment and population size; models *A* and *B*), the regression coefficients can successfully reflect the direction and magnitude of the relationships included in the non-linear simulation model (the impact of $N_t$ and $E_t$ for $N_{t+1}$).
+However, as the processes included in the simulation models become slightly more complex (competition with a second species, spatial structure; models *C*, *D*, *E*, *F*), regression coefficients were useful for prediction but not for recreating the underlying processes driving changes in population size (Appendix S1).
 
-To consider species interactions and environmental fluctuations (model *iv*; Figure 1a \@ref(fig:fig1)), we used the latent variable modelling approach implemented in HMSC.
-The linear model successfully estimated coefficients for fixed environmental effects that captured the impacts of environment for population size in both species (Figure 1b,c) and partitioned the relative importance of the environment and species interactions for species abundances (Figure 1d).
+To consider species interactions and environmental fluctuations (model *D*; Figure 1a \@ref(fig:fig1)), we used the latent variable modelling approach implemented in HMSC.
+The linear model successfully estimated coefficients for fixed environmental effects that captured the impacts of environment for population size in both species (Figure 1b) and partitioned the relative importance of the environment and species interactions for species abundances (Figure 1c).
 The model coefficients could also predict the population size values over time (Figure 1a), though the quality of this prediction varied depending on the level of environmental noise (Appendix S1).
 
-```{r, out.width="90%", include=TRUE, fig.align="center", fig.cap=c("your caption"), echo=FALSE}
+```{r fig1, out.width="90%", include=TRUE, fig.align="center", fig.cap=c("Simulated and model-estimated values for population dynamics in two-species systems.
+Panels a, d, and g show log-population size (y-axis) over time (x-axis) for species $i$ and $j$ when they grow and compete for resources in sites with random environmental fluctuations over time.
+Species either have fixed (panel a) or evolving trait values (panels d, g) and are present in either 1 (panels a, d; simulation models D and E) or multiple sites (panel g; model F).
+Graphs in panels b, e, and h show predicted log-population size ($ln(N)$) for species 1 and 2 across gradients of the environment ($E + E^2$) and the change in trait values of species 1 ($|\\Delta x_1|$).
+Graphs in panels c, f, and i show the results of a variation partition after fitting data to the HMSC statistical model.
+The variation partition plot shows the proportion of variation explained by environmental, trait evolution ($|\\Delta x_1|$ and $|\\Delta x_2|$), time random, and spatial random predictors for $ln(N)$ for species 1 and 2.
+Error bars and shading represent the 95% highest probability density interval for posterior predictions."), echo=FALSE}
 knitr::include_graphics("../output/fig1.pdf")
@@
 
-The HMSC model was successfully modelled changes in species abundances when trait evolution was introduced (model *v*; Figure 1e \@ref(fig:fig1)).
-Fixed effect coefficients for the environmental driver and the total trait change from one time to the next (\|\Delta x\|) matched well with observed data (Figure 1f,g).
-The model estimated the relative importance of the environmental driver, trait evolution in both species, and species interactions (Figure 1h), and predicted the observed species abundances well (Figure 1e).
-Spatial structure was also reproduced in the coefficients of the linear HMSC model by using spatial random effects to capture covariance in species abundances as a function of spatial distance between sites in the 10-patch simulation model (model *vi*; Figure 1i,j,k \@ref(fig:fig1)).
-The variation partition successfully reflects separation of environmental drivers from trait evolution, species interactions, and spatial random effects (Figure 1l), and we show in Appendix S1 that the relative importance of trait evolution increases when we use a simulation model with less spatial and environmental variation and weak species interactions.
+The HMSC model successfully modelled changes in species abundances when trait evolution was introduced (model *E*; Figure 1d \@ref(fig:fig1)).
+Trait evolution played a strong role for population size, especially the effect of $|\Delta x_1|$ for $ln(N_1)$ and $ln(N_2)$ (gradient plots in Figure 1e).
+The model estimated the relative importance of the environmental driver, trait evolution in both species, and species interactions (Figure 1f), and predicted the observed species abundances well (Figure 1d).
+Spatial structure was also reproduced in the coefficients of the linear HMSC model by using spatial random effects to capture covariance in species abundances as a function of spatial distance between sites in the 10-patch simulation model (model *F*; Figure 1g,h,i \@ref(fig:fig1)).
+The variation partition successfully reflected separation of environmental drivers from trait evolution, species interactions, and spatial random effects (Figure 1i), and we show in Appendix S1 that the relative importance of trait evolution increases when we use a simulation model with less spatial and environmental variation and weak species interactions.
 
@@
 
-The simulation model followed population size and trait value dynamics for species in the landscape across 200 time steps (Figure 4a \@ref(fig:fig4)).
-Resulting biodiversity dynamics depended on the dispersal level and heritability values (Figure 4b \@ref(fig:fig4)).
+The simulation model followed population size and trait value dynamics for species in the landscape across 200 time steps (Figure 2a \@ref(fig:fig2)).
+Resulting biodiversity dynamics depended on the dispersal level and heritability values (Figure 2b \@ref(fig:fig2)).
 For dispersal, results mirrored some expectations from existing metacommunity models, i.e.
-that regional $\gamma$ diversity decreases with increasing dispersal level ([@Mouquet2003]; Figure 4b \@ref(fig:fig4)).
-The inclusion of trait heritability had a very strong effect on community diversity, reflecting that local adaptation can rescue species and contribute to richness at the local and metacommunity scale.
-Species diversity at the regional $\gamma$ level increased with increasing $h^2$ levels (Figure 4b \@ref(fig:fig4)).
+that regional $\gamma$ diversity decreases with increasing dispersal level ([@Mouquet2003]; Figure 2b \@ref(fig:fig2)).
+The inclusion of trait heritability had a strong effect on community diversity, reflecting that local adaptation can rescue species and contribute to richness at the local and metacommunity scale.
+Regional species diversity ($\gamma$) increased with increasing $h^2$ levels (Figure 2b \@ref(fig:fig2)).
 
-Analysis using the HMSC model for community time series across $d$ and $h^2$ levels indicated that the importance of evolution for community composition ($V_{|\Delta x|}$) decreased with increasing dispersal and with increasing heritability (Figure 4c,d,e \@ref(fig:fig4)).
+Analysis using the HMSC model for community time series across $d$ and $h^2$ levels indicated that the importance of evolution for community composition ($V_{|\Delta x|}$) decreased with increasing dispersal and with increasing heritability (Figure 2c,d,e \@ref(fig:fig2)).
+The importance of spatial random effects increased with dispersal rate, and overall the importance of evolution decreased with increasing $h^2$.
 
-```{r, out.width="90%", include=TRUE, fig.align="center", fig.cap=c("your caption"), echo=FALSE}
-knitr::include_graphics("../output/fig4.pdf")
+```{r fig2, out.width="90%", include=TRUE, fig.align="center", fig.cap=c("Importance of trait evolution for change in community composition in metacommunities.
+(a) We developed a simulation model for population growth, competition, and trait evolution in 15 species that inhabit a patchy landscape.
+The landscape has 50 sites (colored points in the XY space), with distinct local environments that select for diverse patterns of species diversity (size of points) and average trait values (color of points; results are shown at time $t=0, 10, 200$ for a simulation with $d=10^-3$ and $h^2 = 0.1$).
+(b) Regional metacommunity $\\Gamma$ diversity varied across levels of dispersal and heritability ($h^2$).
+(c) Variation partition plots after fitting metacommunity time series of species abundances $ln(N_{ikt})$ to HMSC models with environment, trait evolution, and random effects (space and time) as predictors.
+The top panel is for $h^2 = 0.1$, the middle panel is for $h^2 = 0.3$, and the bottom panel is for $h^2 = 0.9$."), echo=FALSE}
+knitr::include_graphics("../output/fig2.pdf")
 ```

### 2024-11-11 manuscript/manuscript_v12.Rmd
Draft 12 with full Discussion

--- 2024-11-11 manuscript/manuscript_v11.Rmd
@@
 
-```{r, out.width="90%", include=TRUE, fig.align="center", fig.cap=c("The estimated effects of species trait changes and environmental treatments.
+```{r fig3, out.width="90%", include=TRUE, fig.align="center", fig.cap=c("The estimated effects of species trait changes and environmental treatments.
 The variance partitioning of all fixed effects of the HMSC model (a) including trait changes with envrionmental treatments and (b) only including the environmental treamtents.
@@
 
-### Empirical data
+*Main findings*
+The analysis of community composition often requires statistical models that can consider the combined influences of environmental drivers and spatial connectivity.
+A more difficult, yet critical, element to incorporate is species interactions.
+Although the causal role of species interactions is difficult to interpret from statistical models (but see [@Dubart2022] [@Luo2022]) in systems with few species), joint species distribution models (JSDMs) have implemented spatial and temporal random effects that can effectively capture these species interactions using a latent variable approach (but note the resulting species covariance matrices will also reflect other processes besides species interactions; [@Poggiato2021]).
+Finally, one element that has not been fully implemented into joint species distribution models is dynamic trait evolution (see [@Garzon2019] for a model that does consider reaction norms from in-situ reciprocal transplant experiments).
+We used an existing JSDM and developed simulation models of evolving populations in increasing complex scenarios to determine whether the signal of trait evolution could be estimated from data and compared to other drivers of community structure.
+We also applied this statistical model (HMSC, hierarchical models of species distributions; [@Ovaskainen2017HMSC] [@Tikhonov2020]) to an existing empirical dataset (from the experimental study of community eco-evolutionary dynamics in aquatic plants of [@Jewell2023]) to show how HMSC can be used to model the changes in species composition over time as a function of environmental drivers, trait change, and temporal random effects.
+Overall, there is no barrier to including dynamic trait change as a fixed effect in a JSDM model, and the resulting effect size coefficient estimates can be used to (i) estimate the relative importance of trait evolution for community composition and (ii) predict the impacts of future trait change in the system.
 
+*Discussion regarding simple case studies*
+At the same time, fixed effects for the impacts of trait evolution on population size should be interpreted with a great deal of caution.
+For example, our simulations reveal instances where trait evolution may occur at the same time as large changes in population size, and this co-occurrence isn't always causal (e.g.
+when selection is weak but trait heritability is high, and species are introduced at low initial population size).
+In this instance, the regression coefficient from the model may be negative and strong.
+By using simulations of trait evolution and population growth in simple systems (1- and 2-species systems, at single and simplified multi-patch scales; \@ref(fig:fig1); Appendix S1), we can better interpret effect size estimates.
+However, these precautions are not unique to including trait evolution as a predictor of community composition.
+For example, previous research has shown that the effect size coefficients from linear and AR models using $N_{i,t-1}$ to predict $N_{i,t}$ or $N_{j,t}$ are not equivalent to the intra- and interspecific interaction coefficients that determine species density dependence and should not automatically be used to interpret the direction and magnitude of species interactions ([@Kloppers2013] [@Certain2018] [@Muhlbauer2020] [@Olivenca2021]).
+Our findings for using HMSC to estimate the impact of trait evolution ($|\Delta x_i|$) for species abundances are similar - (i) the relative importance of trait evolution estimated by a variation partition accurately reflects the relative strength of simulated species interactions vs.
+magnitude of trait change and (ii) the coefficients greatly increase the predictive ability of the statistical model for simulated data (Appendix S1).
+HMSC is therefore a useful tool for understanding the role that evolution can play in community dynamics.
+
+*Discussion regarding metacom simulations*
+This was also observed when using HMSC to evaluate the importance of evolution for metacommunity structure, using simulated data from a model of population growth, competition, dispersal, and trait evolution in a spatially structured habitat (Figure 2a \@ref(fig:fig1); Appendix S4).
+First of all, the simulations themselves revealed an intriguing role of adaptive evolution for rescuing individual species from extinction (i.e.
+[@Gomulkiewicz1995]) and as a consequence greatly increasing biodiversity at the local and the landscape scale (i.e.
+community rescue, [@Low2015] [@Bell2019]; Figure 2b \@ref(fig:fig2)).
+The recovery of species via local adaptation had consequent effects for coexistence, as species that were initially mal-adapted could recover in population size and exert strong competitive effects on other species.
+Because HMSC can accurately estimate the relative importance of trait evolution for community composition (Appendix S4), we were thus able to test hypotheses about the drivers of increased or decreased impacts of trait evolution for community structure.
+Our simulation models fit to HMSC confirmed our hypotheses that the importance of evolution would decrease with increasing landscape connectivity (modeled as species dispersal rate $d$) and the overall speed of adaptive evolution (modeled as heritability $h^2$).
+Previous theoretical models have been shown that these two properties can have a strong influence on resulting community composition ([@Urban2008] [@Loeuille2008] [@Urban2009] [@DeMeester2016]), but our demonstration that the HMSC statistical model can accurately estimate $R^2$ values for the relative importance of trait evolution can allow hypotheses from theoretical models to be tested using empirical data and compared across different study systems.
+
+*Discussion regarding empirical data*
 By implementing the trait changes of the four different species (Lm, Rn, Sp and Wc) in the HMSC model we were able to infer the relative impact of the trait changes on the community dynamics.
@@
 
+*Takehome points and conclusions*
+There has been a strong general interest in estimating effect sizes of evolution for ecological processes [@Hairston2005]; [@Ellner2011]; [@Pantel2015]; [@Barbour2022]), but in many empirical studies these estimates focus on one or a few species where the impacts of their evolution is strong enough to play a clear and detectable role in the corresponding ecological response variable of interest (e.g.
+population growth rate, community diversity, likelihood of food web collapse).
+There is no general knowledge for overall distributions of effect sizes for the impact of evolution for species composition in communities, nor are there studies that evaluate drivers of shifts in these distributions of effect sizes in complex communities (see [@Hermann2024] for estimates in a predator-prey system).
+We present here distributions of effect sizes of evolution for species abundances (i.e.
+Figure 1c,f,i \@ref(fig:fig1); Figure 2c \@ref(fig:fig2); Figure 3c \@ref(fig:fig3)) in simulated and experimental aquatic plant communities, and expect that these effect size coefficients can be useful for better understanding the relative importance of evolution as a driver of community composition.
+
+An increasing number of studies collect and analyze high-resolution time series of populations and trait values for single [@Rudman2022] and multiple [@Thomas2018] [@Gibert2023] [@Han2023] [@Shen2023] species.
+With increasing access to ecological and evolutionary data comes an increase in the number of interacting processes that should be considered to identify causal mechanisms that structure this data [@Pantel2023].
+HMSC, and joint species distribution models generally, can be expanded to further include dynamic trait change that occurs during the course of community ecological studies.
+Two critically important additions that will increase the application to field or experimental observation data will be (1) bidirectional causality and eco-evolutionary feedbacks and (2) the relative contribution of plasticity and genotypic evolution for trait change (i.e.
+[@Gibert2022]).
+It will also be necessary to improve knowledge of the distributions of eco-evolutionary effects and how these are influenced by causal processes that change the speed of evolution or the strength of species interactions.
+Although these interacting eco-evolutionary dynamic processes represent an increased degree of complexity for data analysis and interpretation, these mechanistic processes can be translated into simpler statistical models to estimate ecological and evolutionary drivers of community composition.
+
 # References

### 2024-11-13 manuscript/manuscript_v13.Rmd
Added mscript draft 13 to address RH comments

--- 2024-11-11 manuscript/manuscript_v12.Rmd
@@
 Analysis of biodiversity at large spatial scales can be complex, because numerous environmental properties, spatial structure and connectivity, interactions between species, as well as stochastic processes such as drift and priority effects operate and determine observed patterns.
-Statistical models in ecology have made tremendous progress in reflecting these numerous ecological processes at multiple scales.
-There are now a diversity of joint species distribution models [JSDMs; @Pollock2014] [@Pichler2021] or multi-species occupancy models [@Devarajan2020] [@Poggiato2021], which can incorporate these diverse components that structure variance in biological data.
-Many of these models increasingly have implementations in R packages (e.g.
-spOccupancy, [@Doser2022]), and have been used to measure the relative importance of different ecological drivers in fungi, birds, mammals, phytoplankton, fish, and other taxa, from local to continental scales (e.g.
+Statistical models in ecology have made substantial progress in reflecting these numerous ecological processes at multiple scales.
+There are now a diversity of joint species distribution models [JSDMs; @Pollock2014] [@Pichler2021] and multi-species occupancy models [@Devarajan2020] [@Poggiato2021], which can incorporate these diverse components that structure variance in biological data.
+Many of these models have implementations in R packages (e.g.
+spOccupancy, [@Doser2022]), and are used to measure the relative importance of different ecological drivers in e.g.
+communities of fungi, birds, mammals, phytoplankton, fish, and other taxa, from local to continental scales (e.g.
 [@Abrego2020] [@Antao2022] [@Keppeler2022]).
 
-There are some remaining drivers of community composition that aren't currently reflected in these models.
-The dynamics of populations and communities, and resulting biodiversity patterns, are also influenced by the genotypic and phenotypic properties of organisms as well.
+Despite the variety of components implemented in these models, there are some additional drivers of community composition which are not currently reflected.
+The dynamics of populations and communities, and resulting biodiversity patterns, are also influenced by the genotypic and phenotypic properties of organisms.
 Although traits can be considered in some JSDMs, they typically require fixed values for species' traits (e.g.
@@
 Although an increasing number of theoretical and empirical studies have evaluated the impact of evolutionary dynamics for community assembly and metacommunity dynamics [@Loeuille2008] [@Osmond2013] [@Pantel2015] [@Toju2017], analytical methods to consider this additional complexity lag behind the scope of data collected in eco-evolutionary studies.
-Some methods exist that ultimately apply ANOVA to partition variance in traits or growth rates across contributions of ecological and evolutionary components ([@Hairston2005] [@Ellner2011] [@Govaert2016]; additional methods for continuous data are also described in @Ellner2011).
+Some methods exist that ultimately apply ANOVA to partition variance in traits or growth rates across contributions of ecological and evolutionary components ([@Hairston2005] [@Govaert2016]; methods for discrete and continuous analyses are described in @Ellner2011).
 These methods are more frequently cited for their potential rather than for their actual application to empirical data (but see [@terHorst2014] [@Hiltunen2014]), because this approach is ultimately limited by the same requirements that can limit application of ANOVA - data should follow a normal distribution, the processes that structure the target response variable must be linear and additive, and complicated structuring mechanisms must be reduced into one or a few target categories labeled as 'ecology' or 'evolution'.
-Flexible models, which can vary depending on the mechanisms and structure of the diverse kinds of data that most studies of eco-evolutionary dynamics will collect, are needed for rigorous inference about eco-evolutionary processes [@Pantel2023].
+Flexible models, which can better reflect the mechanisms and structure of the diverse kinds of data that most studies of eco-evolutionary dynamics will collect, are needed for rigorous inference about eco-evolutionary processes [@Pantel2023].
 
@@
 
-One particular metacommunity statistical model, the Hierarchical Modelling of Species Communities (HMSC) framework [@Ovaskainen2017HMSC] [@Tikhonov2020], explicitly considers the multitude of processes that can structure community composition across time and space [@Leibold2022] and is potentially well suited to consider analysis of metacommunity eco-evolutionary processes.
-This model of environmental filtering via variation and covariation in how species respond to their environment considers the impacts of (fixed) trait values and phylogenetic relationships, and uses latent variables to account for the numerous unobserved environmental and spatial features that are difficult to exhaustively measure in community surveys.
-It has been demonstrated that this statistical model reflects the diverse array of community assembly and metacommunity processes that structure species abundances (i.e.
+One particular metacommunity statistical model, the Hierarchical Modelling of Species Communities (HMSC) framework [@Ovaskainen2017HMSC] [@Tikhonov2020], explicitly considers a multitude of ecological processes that can structure community composition across time and space [@Leibold2022] and is potentially well suited to consider analysis of metacommunity eco-evolutionary processes.
+This model infers the role of environmental filtering via variation and covariation in how species respond to their environment, considers the impacts of (fixed) trait values and phylogenetic relationships, and uses latent variables to account for the numerous unobserved environmental and spatial features that are difficult to exhaustively measure in community surveys.
+Previous studies have demonstrated that this statistical model can reflect the diverse array of community assembly and metacommunity processes that structure species abundances (i.e.
 species sorting, environmental filtering, priority effects; [@Little2018] [@Ovaskainen2020] [@Leibold2022]).
@@
 In this study, we use simulations of population dynamics, environmental drivers, species interactions, and trait evolution in populations, communities, at the local and metacommunity scale, to study eco-evolutionary dynamics.
-After generating simulated time series of population sizes and mean trait values, we then use HMSC to analyze the resulting data, to determine whether HMSC can successfully estimate the impacts of trait evolution for community composition, and to generate relative contributions of fixed environmental, spatial, and trait evolution drivers as well as random effects of spatial and temporal covariance among species.
+After generating simulated time series of population sizes and mean trait values, we then use HMSC to analyze the resulting data, to determine whether HMSC can successfully estimate the impacts of trait evolution for community composition, and to generate relative contributions of fixed environmental and trait evolution drivers as well as random effects of spatial and temporal covariance among species.
 We asked three main questions: (i) can linear statistical models (like HMSC) accurately capture the impact of non-linear population, community, and evolutionary dynamic processes?, (ii) can HMSC be used to evaluate hypothesized drivers of the relative importance of evolution for community composition at the metacommunity scale?, and (iii) can we better understand the role of evolution for community dynamics in experimental systems using HMSC?
@@
 
-To fit linear regression models to the simulated population size data, we began with models using population size and environment (including the second-order term, to better reflect species optimum values) as fixed effects ($N_t \sim N_{t+1} + E_t + E_t^2$; for simulation models *A* and *B*), and progressed to a mixed model using environment and trait evolution as fixed effects ($N_t \sim E_t + E_t^2 + |\Delta x|$; for simulation models *C*, *D*, and *E*) and time (models *C*, *D*, and *E*) and spatial distance (model *F*) as random effects.
-For consistency, all linear regression models were implemented as heirarchical Bayesian models using HMSC (Hierarchical Modelling of Species Communities [@Ovaskainen2017HMSC] [@Tikhonov2020]).
+To fit linear regression models to the simulated population size data, we began with models using population size (model *A*) and environment (including the second-order term, to better reflect species optimum values; model *B*= as fixed effects ($N_t \sim N_{t+1} + E_t + E_t^2$), and progressed to a mixed model using environment and trait evolution as fixed effects ($N_t \sim E_t + E_t^2 + |\Delta x|$; for simulation model *E*) and time (models *C*, *D*, *E*, and *F*) and spatial distance (model *F*) as random effects.
+For consistency, all linear regression models were implemented as hierarchical Bayesian models using HMSC (Hierarchical Modelling of Species Communities [@Ovaskainen2017HMSC] [@Tikhonov2020]).
 HMSC uses a latent variable approach as an alternative to directly modelling species interaction coefficients, estimating a reduced number of linear combinations of species abundances that best predict future population size values.
@@
 The patches have values of an environmental property $E \sim U(0,1)$ that determines the local optimum phenotype *E* = $x_{opt}$, i.e.
-where species experience their absolute fitness $W_{max}$, and patches also have spatial locations *X* and *Y* (both drawn from $U(0, 1)$).
+where species experience their maximum absolute fitness $W_{max}$, and patches also have spatial locations *X* and *Y* (both drawn from $U(0, 1)$).
 The patches thus have a connectivity matrix **D** (here given by their Euclidean distance), as well as a connectivity matrix **C** that is a Gaussian function of **D** and species dispersal rate $d_i$.
 
-We used the evolving metacommunity model to test whether landscape connectivity (the overall dispersal rate $d$) and the speed of evolution (evaluated by varying $h^2$) influenced the relative importance of trait evolution ($|\Delta x|$) for metacommunity composition $N_{i,t,k}$ (evaluated as $V_{|\Delta x|}$, the proportion of variation in species abundances explained by $|\Delta x|$ as estimated by the HMSC model fit to the simulated data across $d$ and $h^2$ levels).
+We used the evolving metacommunity model to test whether landscape connectivity (the overall dispersal rate $d$) and the speed of evolution (varying $h^2$) influenced the relative importance of trait evolution ($|\Delta x|$) for metacommunity composition $N_{i,t,k}$ (evaluated as $V_{|\Delta x|}$, the proportion of variation in species abundances explained by $|\Delta x|$ as estimated by the HMSC model fit to the simulated data across $d$ and $h^2$ levels).
 We hypothesized that the overall importance of trait evolution ($V_{|\Delta x|}$) would decrease with increasing dispersal rate - at low dispersal rates, locally mal-adapted species are less likely to emigrate to more optimal patches and will experience strong local selection, leading to a large amount of trait evolution.
-Population sizes for these initially mal-adapted populations will also likely be lower, leading to reduced relative importance of density-dependent intra- and interspecific competition.
+Population sizes for initial, mal-adapted populations will also likely be lower, leading to reduced relative importance of density-dependent intra- and interspecific competition.
 We also hypothesized that heritability ($h^2$) would mediate the relative importance of dispersal rate - at higher $h^2$, species will evolve quickly and we expected the importance of trait evolution $V_{|\Delta x|}$ to depend on the amount of time that mal-adapted populations persist.
-Therefore simulations with higher $h^2$ would have decreased importance of evolution $V_{|\Delta x|}$ for community composition.
+Therefore, simulations with higher $h^2$ would have decreased importance of evolution $V_{|\Delta x|}$ for community composition.
 For the same reason, we hypothesized generally that, for a given dispersal rate $d$, the importance of evolution $V_{|\Delta x|}$ would decrease with increasing $h^2$.

### 2024-11-13 manuscript/manuscript_v13.Rmd
Edits mscript v13

@@
 
-$$ N_{i,t+1} =\frac{\hat{W}e^ \frac{-[(\frac{w+(1-h^2)P} {P+w})(E-x_{i,t})]^2}{2(P+w)}N_{i,t}} {1 + \alpha_{ii}N_{i,t} + \sum_{j \neq i}^{S} \alpha_{ij}N_{j,t}} $$ where $N_i,t$ is the population size of species *i* at time *t*, $\hat{W}$ is calculated as $\hat{W}=W_{max}\sqrt(\frac{w}{P+w})$, $W_{max}$ is the species' maximum per-capita growth rate, *w* is the width of the Gaussian fitness function (which determines the strength of selection), *P* is the width of the distribution of the phenotype *x*, $h^2$ is the heritability of the trait *x*, *E* is the local environmental optimum trait value, $x_{i,t}$ is the trait value of species *i* at time *t*, $\alpha_{ii}$ is the intraspecific competition coefficient (the per capita impact of species *i* on itself) and $\alpha_{ij}$ is the interspecific competition coefficient.
+$$ N_{i,t+1} =\frac{\hat{W}e^ \frac{-[(\frac{w+(1-h^2)P} {P+w})(E-x_{i,t})]^2}{2(P+w)}N_{i,t}} {1 + \alpha_{ii}N_{i,t} + \sum_{j \neq i}^{S} \alpha_{ij}N_{j,t}} $$ where $N_i,t$ is the population size of species *i* at time *t*, $\hat{W}$ is calculated as $\hat{W}=W_{max}\sqrt(\frac{w}{P+w})$, $W_{max}$ is the species' maximum per-capita growth rate, *w* is the width of the Gaussian fitness function (which determines the strength of selection), *P* is the width of the distribution of the phenotype *x*, $h^2$ is the heritability of the trait *x*, *E* is the local environmental optimum trait value ($x_{opt}$), $x_{i,t}$ is the trait value of species *i* at time *t*, $\alpha_{ii}$ is the intraspecific competition coefficient (the per capita impact of species *i* on itself) and $\alpha_{ij}$ is the interspecific competition coefficient.
 Populations have a critical density $N_c$, below which the population is subject to extinction due to demographic stochasticity at a probability of *p* [@Gomulkiewicz1995].
@@
 
-To determine the relative influence of environmental and spatial properties, intra- and interspecific population dynamics, and evolutionary dynamics for community structure, we applied hierarchical modelling of species communities (HMSC, [@Ovaskainen2017HMSC] [@Tikhonov2020]) to model the log abundance at each site and year for each species $ln(N_{ikt})$.
+To determine the relative influence of environmental and spatial properties, intra- and interspecific population dynamics, and evolutionary dynamics for community structure, we applied hierarchical modelling of species communities (HMSC, [@Ovaskainen2017HMSC] [@Tikhonov2020]) to model the log-abundance at each site and year for each species.
 The log-abundance values were modeled as $ln(N_{ikt}) \sim N(L_{ikt},\sigma_2)$, where the linear predictors **L** are the sum of fixed and random effects **L^F^** + **L^R^** [@Tikhonov2020].
@@
 The authors conducted a community selection experiments in which they grew four floating and short-lived aquatic plants (the angiosperms *Lemna minor* here designated as Lm, *Spirodela polyrhiza* designated as Sp and *Wolffia columbiana* designated as Wc, and the liverwort *Ricciocarpos natans* designated as Rn) under different environmental conditions for twelve weeks.
-The environmental conditions were defined by three differed amount of shadings of the water pool, 3%, 12% and 50%, and different nutrient concentrations, with low (dissolved nitrogen (DN) = 200 $\mu g L^{-1}$, dissolved phosphorous (DP) = 10 $\mu g L^{-1}$, medium (DN = 800 $\mu g L^{-1}$, DP = 40 $\mu g L^{-1}$), and high concentration (DN = 3200 $\mu g L^{-1}$, DP = 160 $\mu g L^{-1}$) of nutrients.
-They combined each shading with each nutrient level, resulting in nine different environmental conditions.
-Throughout the experiment they measured biweekly, from week one to week eleven, the frond size of ten individuals from each species, as body size trait, in each environmental condition alongside their corresponding species density.
-We calculated the average trait change as $\Delta x = x_{ijt} - x_{ijt-1}$, where $x_{ij}$ described the mean trait of species $i$ in environment $j$ and week $t$.
-As no $\Delta \hat{x}$ value could be calculated for week one of the experiments, it was omitted from the analysis.
-We fitted two HMSC models to the data: An 'evolution model' in which $\Delta \hat{x}$ of each species and environmental conditions were applied as explanatory variable, and a 'no-evolution model', in which only the environmental conditions were included.
+The environmental conditions were defined by three different shade cover levels for the experimental pools, 3%, 12% and 50%, and different nutrient concentrations, with low (dissolved nitrogen (DN) = 200 $\mu g L^{-1}$, dissolved phosphorous (DP) = 10 $\mu g L^{-1}$, medium (DN = 800 $\mu g L^{-1}$, DP = 40 $\mu g L^{-1}$), and high (DN = 3200 $\mu g L^{-1}$, DP = 160 $\mu g L^{-1}$) concentrations of nutrients.
+They combined each shade and nutrient level, resulting in nine different environmental conditions.
+For the eleven weeks of the experiment, the frond size of ten individuals from each species, as well as population density were measured.
+To use the data with our HMSC analysis, we calculated the average trait change as $\Delta x = x_{ijt} - x_{ijt-1}$, where $x_{ij}$ described the mean trait of species $i$ in environment $j$ and week $t$.
+
+To determine the relative importance of evolution for community composition, we fitted two HMSC models to the data: An 'evolution model' in which $\Delta \hat{x}$ of each species and environmental conditions were used as explanatory variables, and a 'no-evolution model', in which only the environmental conditions were included.
 Species density was log-transformed species before applying it to the model, therefore we could use normal distribution in the model.
-Additionally, we applied replicates (*n* = 2) and time as random effects.
-For each model we ran four Markov chain Monte-Carlos' for 1,875,000 iterations from which the first 625,000 were cut off as the transient phase and with a thinning factor of 5,000 250 samples were taken from each chain.
-To test the impact of evolution by the model, we extracted the posterior distributions of the estimated beta effects.
-These were multiplied with the $\Delta \hat{x}$ of the corresponding species and summed up together with the intercept.
+We also included replicates (*n* = 2) and time as random effects.
+For each model we ran four MCMC chains to generate 1000 samples of the posterior distribution for all model parameters.
+To evaluate the model-estimated role of evolution for predicted population size, we used the posterior distributions of the estimated effect size of $|\Delta x|$ multiplied by the observed $\Delta \hat{x}$ of each species,added to the model-estimated intercept.
 

### 2024-11-22 manuscript/Appendix_S1_v02.Rmd
Cleaning up Appendix S1

--- 2024-09-13 manuscript/Appendix_S1_v01.Rmd
@@
 csl: the-american-naturalist.csl
-bibliography: refs.bib
+bibliography: refs_S1.bib
 ---
@@
 library(MASS)
+library(ecoevor)
+library(Hmisc)
 ```
@@
 ## Population dynamics simulation
-In the metacommunity simulation in the main text, a species resides in a site with an initial population size $N_{i,0} \sim Pois(10)$, a growth rate $r_i$ that depends on the local environmental value $E_k$ and the species trait $x_i$, and a fixed intra-specific competition coefficient of $\alpha_{ii}$ = 0.00125.
-We simulate population growth here:
+We simulate population growth using $r_i = 1.67$ and $\alpha_{ii}$ = 0.00125, with an initial population size $N_{1,0} = 10$.
+This is implemented using the `disc_log` function from the `ecoevor` package [https://github.com/jhpantel/ecoevoR](https://github.com/jhpantel/ecoevoR).
 
@@
 # Simulate initial species population growth
-N1.0 <- rpois(1,10)
+N1.0 <- 10
 r1.0 <- 1.67
 alpha.11 <- 0.00125
-# model function
-disc_log <- function(r, N0, alpha) {
-    Nt1 <- (r*N0) / (1+alpha*N0)
-    return(Nt1)
-}
 # Simulation of model for t time steps
@@
 ggplot2::ggplot(dat, aes(time, N)) + geom_point() + geom_hline(yintercept = ((r1.0 - 1)/alpha.11),
-    linetype = "dashed", color = "gray")
+    linetype = "dashed", color = "gray") +
+  theme_classic()
 ```
@@
 ```
-```{r figs2, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs2}Population size over time (black line) with fitted values from a first-order autoregressive model (red dashed line).",warning=FALSE}
+```{r figs2, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs2}Population size over time (black line) with fitted values from a first-order autoregressive (grey) and linear regression (orange) model.",warning=FALSE}
 colors <- c("data" = "black", "AR1 fit" = "gray", "lm fit"="orange")
@@
 d.dN$Nt_plus_1 <- log(dat$N[2:t])
-ggplot2::ggplot(d.dN, aes(x=Nt,y=Nt_plus_1)) + geom_point()
+ggplot2::ggplot(d.dN, aes(x=Nt,y=Nt_plus_1)) + geom_point() + theme_classic()
 ```
@@
 
-```{r figs4, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs4} Change in population size from one time step to the next $N_{t+1}$ as a function of $N_{t+1}$"}
+```{r figs4, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs4} Change in population size from one time step to the next $\\Delta N$ as a function of $N_{t+1}$"}
 d.dN <- as.data.frame(dN)
@@
 d.dN$Nt_plus_1 <- log(dat$N[2:t])
-ggplot2::ggplot(d.dN, aes(Nt, dN)) + geom_point()
+ggplot2::ggplot(d.dN, aes(Nt, dN)) + geom_point() + theme_classic()
 ```
@@
 
-```{r}
+```{r results='hide', message=FALSE, warning=FALSE}
 # prepare data in HMSC format
@@
 
-```{r figs5, fig.pos = "H", fig.show="hold", fig.cap = "\\label{fig:figs5} Observed (grey) and model-fit (blue) values for population size at time t (x-axis) and t+1 (y-axis).",results='hide',fig.keep='all'}
+```{r figs5, fig.pos = "H", fig.show="hold", fig.cap = "\\label{fig:figs5} Observed (grey) and model-fit (blue) values for population size at time t (x-axis) and t+1 (y-axis) using HMSC.",results='hide',fig.keep='all'}
 Gradient <- constructGradient(m.1.sample,focalVariable="x",ngrid=29)
 predY <- predict(m.1.sample,Gradient=Gradient,expected=TRUE)
-#preds <-computePredictedValues(m.1.sample)
 plotGradient(m.1.sample,Gradient,pred=predY,showData=T,measure="Y",main="",xlab="N_t",ylab="predicted N_t+1")
@@
 
-```{r}
+```{r figs6, fig.pos = "H", fig.show="hold", fig.cap = "\\label{fig:figs6} Observed (grey) and model-fit (blue) values for population size at time t (x-axis) and t+1 (y-axis) using lm.",results='hide',fig.keep='all',warning=FALSE,message=FALSE}
 lm_dat <- data.frame(cbind(log(dat$N[2:t]),log(dat$N[1:(t-1)])))
@@
   stat_summary(fun.data= mean_cl_normal) + 
-  geom_smooth(method='lm')
+  geom_smooth(method='lm') + theme_classic()
 ```
@@
 Here, we begin with $E = x_{i,t} = 0.8$, then simulate the environment $E$ value fluctuating randomly over time, and finally use a linear model to fit $E$ as a covariate.
+We implement the simulation using the `disc_log_E` function from the `ecoevor` package.
 
@@
 x1.0 <- 0.8
-# model function
-disc_log_E <- function(r, N0, alpha, E, x) {
-    Nt1 <- ((r*exp(-(E-x)^2))*N0) / (1+alpha*N0)
-    return(Nt1)
-}
 # Simulation of model for t time steps
@@
 ```
-```{r figs6, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs1}Population size \\textit{N} over time \\textit{t} for a discrete-time logistic growth model, with parameters $r_i$ = 1.67, $N_{1,0}$ = 14, and $\\alpha_{11}$ = 0.00125.
-Relationship between E and Nt is also shown."}
+```{r figs7, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs7}Population size \\textit{N} over time \\textit{t} for a discrete-time logistic growth model, with parameters $r_i$ = 1.67, $N_{1,0}$ = 14, and $\\alpha_{11}$ = 0.00125.
+The value of E over time is also shown."}
 # Plot simulation: ggplot
@@
 dat$time <- 1:t
-p1 <- ggplot2::ggplot(dat, aes(time, N)) + geom_point() + geom_hline(yintercept = ((r1.0 - 1)/alpha.11),linetype = "dashed", color = "gray")
-p2 <- ggplot2::ggplot(dat, aes(time, E)) + geom_point()
+p1 <- ggplot2::ggplot(dat, aes(time, N)) + geom_point() + geom_hline(yintercept = ((r1.0 - 1)/alpha.11),linetype = "dashed", color = "gray") + theme_classic()
+p2 <- ggplot2::ggplot(dat, aes(time, E)) + geom_point() + theme_classic()
 p1 + p2
@@
 ```
-```{r figs7, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs2}Population size over time (black line) with fitted values from a first-order autoregressive model (red dashed line).",warning=FALSE}
+
+```{r figs8, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs8} Population size over time (black line) with fitted values from a first-order autoregressive model (gray) and a linear model (orange).",warning=FALSE}
+
 colors <- c("data" = "black", "AR1 fit" = "gray", "lm fit"="orange")
@@
 
-```{r figs8, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs3}Population size (logarithm) at one time step $N_{t+1}$ as a function of log-population size in the previous time step $N_t$.
-"}
+The linear model is a good fit when including the environmental covariate.
+$N_{t+1}$ and $N_t$ can still be captured by a linear relationship.
+However we see that the relationship between $N_{t+1}$ and $E_t$ is non-linear.
+
+```{r figs9, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs9}Population size (logarithm) at one time step $N_{t+1}$ as a function of log-population size in the previous time step $N_t$.
+Also shown are change in population size $\\Delta N$ with population size and E."}
 dN <- rep(NA, (t-1))
@@
 d.dN$Nt_plus_1 <- log(dat$N[2:t])
-p1 <- ggplot2::ggplot(d.dN, aes(x=Nt,y=Nt_plus_1)) + geom_point()
+p1 <- ggplot2::ggplot(d.dN, aes(x=Nt,y=Nt_plus_1)) + geom_point() + theme_classic()
 
@@
 d.dN$E <- log(dat$E[2:t])
-p2 <- ggplot2::ggplot(d.dN, aes(x=E,y=Nt_plus_1)) + geom_point()
-p3 <- ggplot2::ggplot(d.dN, aes(Nt, dN)) + geom_point()
-p4 <- ggplot2::ggplot(d.dN, aes(E, dN)) + geom_point()
+p2 <- ggplot2::ggplot(d.dN, aes(x=E,y=Nt_plus_1)) + geom_point() + theme_classic()
+p3 <- ggplot2::ggplot(d.dN, aes(Nt, dN)) + geom_point() + theme_classic()
+p4 <- ggplot2::ggplot(d.dN, aes(E, dN)) + geom_point() + theme_classic()
 
@@
 
-The linear model is a good fit when including the environmental covariate.
-$N_{t+1}$ and $N_t$ can still be captured by a linear relationship.
-However we see that the relationship between $N_{t+1}$ and $E_t$ is non-linear.
 This tells us that the lm is good for predictions, but not for inference (for capturing well the relationsip between the predictor and response variable).
@@
 
-```{r figs9, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs3}Population size (logarithm) at one time step $N_{t+1}$ as a function of log-population size in the previous time step $N_t$.
-"}
+```{r figs10, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs10} Population size (logarithm) at one time step $N_{t+1}$ as a function of environment E.
+Linear and quadratic fits are shown."}
 jtools::effect_plot(m.2.lm,pred=E,plot.points = TRUE)
@@
   geom_point() +
-  geom_smooth(method = "lm", formula = y ~ x + I(x^2))
+  geom_smooth(method = "lm", formula = y ~ x + I(x^2)) + theme_classic()
 ```
@@
 
-```{r}
+```{r results='hide', warning=FALSE, message=FALSE}
 # prepare data in HMSC format
@@
 
+```{r}
+knitr::knit_exit()
+```
+
 We recall that the interpretation of the coefficients in an arimaX (arima with covariates) model is difficult.
@@
 
-```{r figs10, fig.pos = "H", fig.show="hold", fig.cap = "\\label{fig:figs5} Observed (grey) and model-fit (blue) values for population size at time t (x-axis) and t+1 (y-axis).",results='hide',fig.keep='all'}
+```{r figs11, fig.pos = "H", fig.show="hold", fig.cap = "\\label{fig:figs11} Observed (grey) and model-fit (blue) values for population size at time t (x-axis) and t+1 (y-axis).",results='hide',fig.keep='all'}
 Gradient <- constructGradient(m.2.sample,focalVariable="E",non.focalVariables=list(Nt=list(2),Esq=list(2)),ngrid=39)
@@
 ```
-```{r figs11, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs1}Population size \\textit{N} over time \\textit{t} for a discrete-time logistic growth model, with parameters $r_1$ = 1.67, $r_2$ = 1.7, $N_{i,0}$ = 10, $\\alpha_{ii}$ = 0.00125, $\\alpha_{12}$ = 0.008, and $\\alpha_{21}$ = 0.008725."}
+```{r figs12, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs12}Population size \\textit{N} over time \\textit{t} for a discrete-time logistic growth model, with parameters $r_1$ = 1.67, $r_2$ = 1.7, $N_{i,0}$ = 10, $\\alpha_{ii}$ = 0.00125, $\\alpha_{12}$ = 0.008, and $\\alpha_{21}$ = 0.008725."}
 # Plot simulation: ggplot
@@
 
-```{r figs12, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs1}Population size \\textit{N} over time \\textit{t} for a discrete-time logistic growth model, with competition."}
+```{r figs13, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs13}Population size \\textit{N} over time \\textit{t} for a discrete-time logistic growth model, with competition."}
 # Plot simulation: ggplot
@@
 
-```{r figs13, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs2}Population size over time (black line) with fitted values from a first-order autoregressive model (gray line) and from a linear model (orange line).",warning=FALSE}
+```{r figs14, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs14}Population size over time (black line) with fitted values from a first-order autoregressive model (gray line) and from a linear model (orange line).",warning=FALSE}
 colors <- c("data" = "black", "AR1 fit" = "gray", "lm fit"="orange")
@@
 
-```{r figs14, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs3}Population size (logarithm) at one time step $N_{t+1}$ as a function of log-population size of Species 1 and Species 2 in the previous time step $N_t$.
+```{r figs15, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs15}Population size (logarithm) at one time step $N_{t+1}$ as a function of log-population size of Species 1 and Species 2 in the previous time step $N_t$.
 Also shown is the change in population size of Species 1 as a function of $N_t$ for both species."}
@@
 
-```{r figs15, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs3}Population size (logarithm) at one time step $N_{t+1}$ as a function of log-population size of the second species in the previous time step $N_t$.
+```{r figs16, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs16}Population size (logarithm) at one time step $N_{t+1}$ as a function of log-population size of the second species in the previous time step $N_t$.
 Fitted (line) and observed (points) values are shown."}

### 2024-11-25 manuscript/Appendix_S1_v02.Rmd
Worked on Appendix S1

@@
 
-```{r figs9, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs9}Population size (logarithm) at one time step $N_{t+1}$ as a function of log-population size in the previous time step $N_t$.
+```{r figs9, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs9}Population size (logarithm) at one time step $N_{t+1}$ as a function of log-population size in the previous time step $N_t$ and of environmental value $E_t$.
 Also shown are change in population size $\\Delta N$ with population size and E."}
@@
 
-```{r}
-knitr::knit_exit()
-```
-
 We recall that the interpretation of the coefficients in an arimaX (arima with covariates) model is difficult.
@@
 
-```{r figs11, fig.pos = "H", fig.show="hold", fig.cap = "\\label{fig:figs11} Observed (grey) and model-fit (blue) values for population size at time t (x-axis) and t+1 (y-axis).",results='hide',fig.keep='all'}
+```{r figs11, fig.pos = "H", fig.show="hold", fig.cap = "\\label{fig:figs11} Observed (grey) and model-fit (blue) values for population size at time t+1 (y-axis) as a function of environment $E_t$ (x-axis).",results='hide',fig.keep='all'}
 Gradient <- constructGradient(m.2.sample,focalVariable="E",non.focalVariables=list(Nt=list(2),Esq=list(2)),ngrid=39)
@@
 dat <- melt(N, id.vars="time")
-ggplot2::ggplot(dat, aes(time, value, col=variable)) + geom_point() + geom_hline(yintercept = ((r1.0 - 1)/alpha.11),linetype = "dashed", color = "gray")
+ggplot2::ggplot(dat, aes(time, value, col=variable)) + geom_point() + geom_hline(yintercept = ((r1.0 - 1)/alpha.11),linetype = "dashed", color = "gray") + theme_classic()
 ```
@@
 
-```{r figs13, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs13}Population size \\textit{N} over time \\textit{t} for a discrete-time logistic growth model, with competition."}
+```{r figs13, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs13}Population size \\textit{N} over time \\textit{t} for a discrete-time logistic growth model, with competition between 2 species ($N_1$ and $N_2$)."}
 # Plot simulation: ggplot
@@
 dat$time <- 1:t
-p1 <- ggplot2::ggplot(dat, aes(time, N1)) + geom_point() + geom_hline(yintercept = ((r1.0 - 1)/alpha.11),linetype = "dashed", color = "gray")
-p2 <- ggplot2::ggplot(dat, aes(time, N2)) + geom_point()
+p1 <- ggplot2::ggplot(dat, aes(time, N1)) + geom_point() + geom_hline(yintercept = ((r1.0 - 1)/alpha.11),linetype = "dashed", color = "gray") + theme_classic()
+p2 <- ggplot2::ggplot(dat, aes(time, N2)) + geom_point() + theme_classic()
 p1 + p2
@@
   xlab("time") +
-  ylab("N1") +
+  ylab("log(N1)") +
   labs(color="Legend") +
@@
   xlab("time") +
-  ylab("N2") +
+  ylab("log(N2)") +
   labs(color="Legend") +
@@
 
-```{r figs15, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs15}Population size (logarithm) at one time step $N_{t+1}$ as a function of log-population size of Species 1 and Species 2 in the previous time step $N_t$.
+```{r figs15, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs15}Population size (logarithm) in species 1 at one time step $N_{1,t+1}$ as a function of log-population size of Species 1 and Species 2 in the previous time step $N_t$.
 Also shown is the change in population size of Species 1 as a function of $N_t$ for both species."}
@@
 d.dN1$Nt_plus_1 <- log(dat$N1[2:t])
-p1 <- ggplot2::ggplot(d.dN1, aes(x=Nt,y=Nt_plus_1)) + geom_point()
+p1 <- ggplot2::ggplot(d.dN1, aes(x=Nt,y=Nt_plus_1)) + geom_point() + xlab("log(N1)") + ylab("log(N1_t+1)") + theme_classic()
 
@@
 d.dN1$N2 <- log(dat$N2[2:t])
-p2 <- ggplot2::ggplot(d.dN1, aes(x=N2,y=Nt_plus_1)) + geom_point()
-p3 <- ggplot2::ggplot(d.dN1, aes(Nt, dN1)) + geom_point()
-p4 <- ggplot2::ggplot(d.dN1, aes(N2, dN1)) + geom_point()
+p2 <- ggplot2::ggplot(d.dN1, aes(x=N2,y=Nt_plus_1)) + geom_point() + theme_classic() + xlab("log(N2)") + ylab("log(N1_t+1)")
+p3 <- ggplot2::ggplot(d.dN1, aes(Nt, dN1)) + geom_point() + xlab("log(N1)") + theme_classic()
+p4 <- ggplot2::ggplot(d.dN1, aes(N2, dN1)) + geom_point() + theme_classic() + xlab("log(N2)")
 
@@
 
-```{r}
+```{r figs16, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs16}Population size (logarithm) in species 1 at one time step $N_{1,t+1}$ as a function of log-population size of Species 2 in the previous time step $N_t$."}
 df <- data.frame(cbind(log(dat$N1[2:t]),log(dat$N1[1:(t-1)]),log(dat$N2[1:(t-1)])))
@@
 lines(prediction$N2, predict(m.3.lm, prediction), lty=2,col="red",lwd=3)
-
-```
-
-```{r figs16, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs16}Population size (logarithm) at one time step $N_{t+1}$ as a function of log-population size of the second species in the previous time step $N_t$.
-Fitted (line) and observed (points) values are shown."}
+```
+
+```{r figs17, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs17} Linear model-predicted population size (logarithm) at one time step $N_{t+1}$ as a function of log-population size of the second species in the previous time step $N_t$.
+Fitted (line) and observed (points) values are shown.", warning=FALSE}
 jtools::effect_plot(m.3.lm,pred=N2,plot.points = TRUE)
-ggplot(data = df, aes(x = N2, y = Nt1)) +
-  geom_point() +
-  geom_smooth(method = "lm", formula = y ~ x)
-```
+# ggplot(data = df, aes(x = N2, y = Nt1)) +
+#   geom_point() +
+#   geom_smooth(method = "lm", formula = y ~ x) + theme_classic()
+```
+
 We can already see that the linear model does not provide a useful fit to the data points (neither does taking the inverse function help).
@@
 This is well presented in @Ovaskainen2017.
-We apply that approach here:
+We apply that approach here.
 
@@
 
-```{r}
+```{r figs18, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs18} Observed (points) and HMSC model-predicted population size (logarithm) for Species 1 (black) and Species 2 (orange), using environment $E_t$ as a fixed effect predictor and specifying temporal random effects to capture the species interactions."}
 preds = computePredictedValues(m.3.sample)
@@
 
-I try again with a next example where the only difference between species is the $\alpha_{ij}$.
+```{r}
+knitr::knit_exit()
+```
+
+To show that the HMSC model correctly allocates variance to the contributing components of the model (here environment, intra-specific and interspecific density-dependence, implemented as temporal random effects), we simulate a new example where the only difference between species is the $\alpha_{ij}$.
 
@@
 dat <- melt(N, id.vars="time")
-ggplot2::ggplot(dat, aes(time, value, col=variable)) + geom_point() + geom_hline(yintercept = ((r1.0 - 1)/alpha.11),linetype = "dashed", color = "gray")
+ggplot2::ggplot(dat, aes(time, value, col=variable)) + geom_point() + geom_hline(yintercept = ((r1.0 - 1)/alpha.11),linetype = "dashed", color = "gray") = theme_classic()
 ```

### 2024-11-26 manuscript/Appendix_S1_v02.Rmd
Updated S1, knit.

@@
 
-# One species, logistic growth
+# Model A: One species, logistic growth
 Population growth over time in a single species is first modelled using a Beverton-Holt (discrete-time, logistic) model (@Beverton1957), using an intra-specific competition coefficient for density-dependent growth (@Hart2013) - thus $\alpha$ = $1/K$.
@@
 ```
-```{r figs1, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs1}Population size \\textit{N} over time \\textit{t} for a discrete-time logistic growth model, with parameters $r_i$ = 1.67, $N_{1,0}$ = 14, and $\\alpha_{11}$ = 0.00125."}
+```{r figs1, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs1}Population size \\textit{N} over time \\textit{t} for a discrete-time logistic growth model, with parameters $r_i$ = 1.67, $N_{1,0}$ = 10, and $\\alpha_{11}$ = 0.00125."}
 # Plot simulation: ggplot
@@
 
-# One species, logistic growth, environmental covariate
+# Model B: One species, logistic growth, environmental covariate
 We now consider using a linear model to analyze population growth when the species growth rate is impacted by a single environmental covariate.
@@
 
-# Two species, logistic growth, competition
+# Model C: Two species, logistic growth, competition
 Here we investigate how a second species impacts the inference we can make from linear models.
@@
 We use a distinct growth rate for the 2nd species and introduce the interspecific interaction coefficient $\alpha_{ij}$.
-
-```{r}
-# Simulate initial species population growth with a second species present
-N1.0 <- 10
-N2.0 <- 10
-r1.0 <- 1.67
-r2.0 <- 1.7
+The simulation is generated using the `disc_LV_comp` function in the `ecoevor` package.
+
+```{r}
+# Initial conditions
+N0 <- c(10,10)
+r <- c(1.67,1.7)
 alpha.11 <- 0.00125
@@
 alpha.21 <- 0.008725
-# model function
-disc_LV_comp <- function(r1,r2,N1.0,N2.0,alpha.11,alpha.22,alpha.12,alpha.21) {
-    Nt1 <- (r1*N1.0) / (1+alpha.11*N1.0+alpha.12*N2.0)
-    Nt2 <- (r2*N2.0) / (1+alpha.22*N2.0+alpha.21*N1.0)
-    return(c(Nt1,Nt2))
-}
+alpha <- matrix(c(alpha.11,alpha.21,alpha.12,alpha.22),nrow=2,byrow=FALSE)
+# make sure the interaction matrix is correct
 # Simulation of model for t time steps
@@
 colnames(N) <- c("N1","N2")
-N$N1[1] <- N1.0
-N$N2[1] <- N2.0
+N[1,] <- N0
 for (i in 2:t) {
-  res <- disc_LV_comp(r1=r1.0,r2=r2.0,N1.0=N[i-1,1],N2.0=N[i-1,2],alpha.11=alpha.11,alpha.22=alpha.22,alpha.12=alpha.12,alpha.21=alpha.21)
-  N$N1[i] <- res[1]
-  N$N2[i] <- res[2]
+  N[i,] <- disc_LV_comp(r=r,N0=N[i-1,],alpha=alpha)
 }
@@
 dat <- melt(N, id.vars="time")
-ggplot2::ggplot(dat, aes(time, value, col=variable)) + geom_point() + geom_hline(yintercept = ((r1.0 - 1)/alpha.11),linetype = "dashed", color = "gray") + theme_classic()
+ggplot2::ggplot(dat, aes(time, value, col=variable)) + geom_point() + geom_hline(yintercept = ((r1.0 - 1)/alpha.11),linetype = "dashed", color = "gray") + theme_classic() + ylab("N")
 ```
@@
 
-```{r figs16, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs16}Population size (logarithm) in species 1 at one time step $N_{1,t+1}$ as a function of log-population size of Species 2 in the previous time step $N_t$."}
+```{r}
 df <- data.frame(cbind(log(dat$N1[2:t]),log(dat$N1[1:(t-1)]),log(dat$N2[1:(t-1)])))
@@
 m.3.lm <- lm(Nt1 ~ I(1/N1) + I(1/N2),data=df)
-
 cor(df$Nt1, predict(m.3.lm))
 summary(m.3.lm)
+```
+
+```{r figs16, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs16}Population size (logarithm) in species 1 at one time step $N_{1,t+1}$ as a function of log-population size of Species 2 in the previous time step $N_t$."}
 xx <- seq(0,100, length=1000)
@@
 ```{r figs17, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs17} Linear model-predicted population size (logarithm) at one time step $N_{t+1}$ as a function of log-population size of the second species in the previous time step $N_t$.
-Fitted (line) and observed (points) values are shown.", warning=FALSE}
+Fitted (line) and observed (points) values are shown.", warning=FALSE, message=FALSE}
 jtools::effect_plot(m.3.lm,pred=N2,plot.points = TRUE)
@@
 # prepare data in HMSC format
-Y <- as.matrix(cbind(log(dat$N1[2:t]),log(dat$N2[2:t])))
-XData <- df[,2:3]
+dat <- as.data.frame(log(N))
+dat$time <- 1:t
+df <- data.frame(dat[(2:t),-3])
+Y <- as.matrix(df)
+XData <- data.frame(dat[1:(t-1),-3])
+colnames(XData) <- c(paste0("n",1:length(r)))
 studyDesign = data.frame(sample = as.factor(1:(t-1)))
@@
 
-The model fixed effect estimates match well with those from the AR1 and linear model.
+The direction of HMSC model fixed effect estimates matches well with those from the AR1 and linear model.
 
@@
 ```
-
 The HMSC model correctly shows the negative association between the species (shown here as correlations - JSDM CHapter 7 tells us the intra-specific correlation is always 1).
 
-```{r figs18, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs18} Observed (points) and HMSC model-predicted population size (logarithm) for Species 1 (black) and Species 2 (orange), using environment $E_t$ as a fixed effect predictor and specifying temporal random effects to capture the species interactions."}
+```{r figs18, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs18} Observed (points) and HMSC model-predicted population size (logarithm) for Species 1 (black) and Species 2 (orange), using environment $E_t$ as a fixed effect predictor and specifying temporal random effects to capture the species interactions.
+Lines represent the 95% highest probability density interval (HPDI) for predictions at each time point."}
 preds = computePredictedValues(m.3.sample)
-df <- data.frame(cbind(log(dat$N1[2:t]),log(dat$N2[2:t]),log(dat$N1[1:(t-1)]),log(dat$N2[1:(t-1)])))
+df <- data.frame(cbind(dat$N1[2:t],dat$N2[2:t],dat$N1[1:(t-1)],dat$N2[1:(t-1)]))
 colnames(df) <- c("Nt1","Nt2","N1","N2")
@@
 
-```{r}
-knitr::knit_exit()
-```
-
 To show that the HMSC model correctly allocates variance to the contributing components of the model (here environment, intra-specific and interspecific density-dependence, implemented as temporal random effects), we simulate a new example where the only difference between species is the $\alpha_{ij}$.
@@
 ```{r}
-# Simulate initial species population growth with a second species present
-N1.0 <- 10
-N2.0 <- 10
-r1.0 <- 1.7
-r2.0 <- 1.7
+# Initial conditions
+N0 <- c(10,10)
+r <- c(1.7,1.7)
 alpha.11 <- 0.01
@@
 alpha.21 <- 0.01
-# model function
-disc_LV_comp <- function(r1,r2,N1.0,N2.0,alpha.11,alpha.22,alpha.12,alpha.21) {
-    Nt1 <- (r1*N1.0) / (1+alpha.11*N1.0+alpha.12*N2.0)
-    Nt2 <- (r2*N2.0) / (1+alpha.22*N2.0+alpha.21*N1.0)
-    return(c(Nt1,Nt2))
-}
+alpha <- matrix(c(alpha.11,alpha.21,alpha.12,alpha.22),nrow=2,byrow=FALSE) # careful to make sure you have correct interaction matrix
 # Simulation of model for t time steps
@@
 colnames(N) <- c("N1","N2")
-N$N1[1] <- N1.0
-N$N2[1] <- N2.0
+N[1,] <- N0
 for (i in 2:t) {
-  res <- disc_LV_comp(r1=r1.0,r2=r2.0,N1.0=N[i-1,1],N2.0=N[i-1,2],alpha.11=alpha.11,alpha.22=alpha.22,alpha.12=alpha.12,alpha.21=alpha.21)
-  N$N1[i] <- res[1]
-  N$N2[i] <- res[2]
-}
-```
-```{r}
+  N[i,] <- disc_LV_comp(r=r,N0=N[i-1,],alpha=alpha)
+}
+```
+```{r figs19, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs19} Population size \\textit{N} over time \\textit{t} for a discrete-time logistic growth model with competition, with parameters $r_1 = r_2$ = 1.7, $N_{i,0}$ = 10, $\\alpha_{ii}$ = 0.01, $\\alpha_{12}$ = 0.005, and $\\alpha_{21}$ = 0.01."}
 # Plot simulation: ggplot
@@
 dat <- melt(N, id.vars="time")
-ggplot2::ggplot(dat, aes(time, value, col=variable)) + geom_point() + geom_hline(yintercept = ((r1.0 - 1)/alpha.11),linetype = "dashed", color = "gray") = theme_classic()
+ggplot2::ggplot(dat, aes(time, value, col=variable)) + geom_point() + geom_hline(yintercept = ((r1.0 - 1)/alpha.11),linetype = "dashed", color = "gray") + theme_classic()
 ```
@@
 preds = computePredictedValues(m.4.sample)
-#evaluateModelFit(hM = m.4.sample, predY = preds)
-# predictive power / cross-validation
-#partition = createPartition(m.4.sample, nfolds = 2)
-#preds = computePredictedValues(m.4.sample, partition = partition, nParallel = nChains)
-#evaluateModelFit(hM = m.4.sample, predY = preds)
-## xx
-#preds = computePredictedValues(m.4.sample, partition=partition, partition.sp=c(1,2), mcmcStep=10, nParallel = nChains)
-#evaluateModelFit(hM=m.4.sample, predY=preds)
-# xx
-#etaPost=getPostEstimate(m.4.sample, "Eta")
-#lambdaPost=getPostEstimate(m.4.sample, "Lambda")
-#biPlot(m.4.sample, etaPost = etaPost, lambdaPost = lambdaPost, factors = c(1,2),"X1")
-```
-```{r}
+```
+```{r figs20, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs20} Observed (points) and HMSC model-predicted population size (logarithm) for Species 1 (black) and Species 2 (orange), using environment $E_t$ as a fixed effect predictor and specifying temporal random effects to capture the species interactions.
+Lines represent the 95% HPDI for predictions at each time point."}
 df$n1_25 <- NA
@@
 points(df$Nt2,pch=19,col="orange")
-lines(df$n1_25,col="blue")
-lines(df$n1_975,col="blue")
-lines(df$n2_25,col="blue")
-lines(df$n2_975,col="blue")
+lines(df$n1_25,col="black")
+lines(df$n1_975,col="black")
+lines(df$n2_25,col="orange")
+lines(df$n2_975,col="orange")
 ```
@@
 The residual associations should not be interpreted as interaction coefficients, but rather used for prediction.
-We can see here they are effective for this (shown above is the observed and model-predicted 95% CI).
-
-# Species growth, competition, with environmental change
+We can see here they are effective for this (shown above is the observed and model-predicted 95% highest probability density interval - HPDI).
+
+# Model D: Species growth, competition, with environmental change
 
@@
 
-In this example, the local environmental optimum trait value $E = 0.8$, and Species 1 $x_1 = 0.6$, while $x_1 = 0.8$.
-
-```{r}
-# Simulate initial species population growth with environment fluctuations
-N1.0 <- 10
-N2.0 <- 10
-r1.0 <- 1.7
-r2.0 <- 1.7
+In this example, the local environmental optimum trait value $E = 0.8$, and Species 1 $x_1 = 0.6$, while $x_2 = 0.8$.
+The simulation is generated using the `disc_LV_E` function in the `ecoevor` package.
+
+```{r}
+# Initial conditions
+N0 <- c(10,10)
+r <- c(1.7,1.7)
 alpha.11 <- 0.01
@@
 alpha.21 <- 0.01
+alpha <- matrix(c(alpha.11,alpha.21,alpha.12,alpha.22),nrow=2,byrow=FALSE)
 E.0 <- 0.8
-x1.0 <- 0.6
-x2.0 <- 0.8
-
-# model function
-disc_LV_E <- function(r1,r2,N1.0,N2.0,alpha.11,alpha.22,alpha.12,alpha.21,E,x1,x2) {
-    Nt1 <- ((r1*exp(-(E-x1)^2))*N1.0) / (1+alpha.11*N1.0+alpha.12*N2.0)
-    Nt2 <- ((r2*exp(-(E-x2)^2))*N2.0) / (1+alpha.22*N2.0+alpha.21*N1.0)
-    return(c(Nt1,Nt2))
-}
+x <- c(0.6,0.8)
 # Simulation of model for t time steps
 t <- 40
-N <- array(NA,dim=c(t,2))
+N <- array(NA,dim=c(t,length(N0)))
 N <- as.data.frame(N)
-colnames(N) <- c("N1","N2")
-N$N1[1] <- N1.0
-N$N2[1] <- N2.0
+colnames(N) <- paste0("N",1:length(N0))
+N[1,] <- N0
 E <- rep(NA, t)
@@
 for (i in 2:t) {
-  res <- disc_LV_E(r1=r1.0,r2=r2.0,N1.0=N[i-1,1],N2.0=N[i-1,2],alpha.11=alpha.11,alpha.22=alpha.22,alpha.12=alpha.12,alpha.21=alpha.21,E=E[i-1],x1=x1.0,x2=x2.0)
-  N$N1[i] <- res[1]
-  N$N2[i] <- res[2]
-  E[i] <- E[i-1] + rnorm(1,0,.1)
-}
-```
-```{r}
+  N[i,] <- disc_LV_E(r=r,N0=N[i-1,],alpha=alpha,E=E[i-1],x=x)
+  E[i] <- E[i-1] + rnorm(1,0,0.1)
+}
+```
+```{r figs21, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs21} Population size \\textit{N} over time \\textit{t} for a discrete-time logistic growth model with competition and a changing environment, with parameters $r_1 = r_2$ = 1.7, $N_{i,0}$ = 10, $\\alpha_{ii}$ = 0.01, $\\alpha_{12}$ = 0.005, and $\\alpha_{21}$ = 0.01."}
 # Plot simulation: ggplot
-N$time <- 1:t
-dat <- melt(N, id.vars="time")
-ggplot2::ggplot(dat, aes(time, value, col=variable)) + geom_point() + geom_hline(yintercept = ((r1.0 - 1)/alpha.11),linetype = "dashed", color = "gray")
-```
-```{r}
-# Plot simulation: ggplot
-dat <- as.data.frame(cbind(log(N$N1),log(N$N2)))
-colnames(dat) <- c("N1","N2")
+gdat <- N
+gdat$time <- 1:t
+dat <- melt(gdat, id.vars="time")
+ggplot2::ggplot(dat, aes(time, value, col=variable)) + geom_point() + geom_hline(yintercept = ((r[1] - 1)/alpha.11),linetype = "dashed", color = "gray") + theme_classic() + ylab("N")
+```
+```{r}
+dat <- as.data.frame(log(N))
 dat$time <- 1:t
-df <- data.frame(cbind(dat$N1[2:t],dat$N2[2:t]))
-colnames(df) <- c("Nt1","Nt2")
+df <- data.frame(dat[(2:t),-3])
+Y <- as.matrix(df)
 # prepare data in HMSC format
-Y <- as.matrix(cbind(df$Nt1,df$Nt2))
-XData <- data.frame(cbind(dat$N1[1:(t-1)],dat$N2[1:(t-1)]),E[1:(t-1)],E[1:(t-1)]^2)
-colnames(XData) <- c("n1","n2","E","Esq")
+XData <- data.frame(dat[1:(t-1),-3],E[1:(t-1)],E[1:(t-1)]^2)
+colnames(XData) <- c(paste0("n",1:length(r)),"E","Esq")
 studyDesign = data.frame(sample = as.factor(1:(t-1)))
@@
 ```
-We can see that the Environment has an overall negative impact for both species as $E$ drifts randomly towards more positive values (where Species 2 is closer because it has a higher trait value).
-We also see that the residual species association effects are strongly positive.
+We can see that the Environment has an overall negative impact for Species 1 as $E$ drifts randomly towards more positive values (and thus further from Species 1 trait value).
+We also see that the residual species association for species 1 and 2 are negative.
 
@@
 ```
-We evaluate this by greatly increasing the effects of interspecific competition (while decreasing the strength of environmental variation).
+We evaluate whether variation partition correctly detects the relative importance of environment and species interactions by greatly increasing the effects of interspecific competition and decreasing the strength of environmental variation.
 
@@
 # Simulate initial species population growth with environment fluctuations
-N1.0 <- 10
-N2.0 <- 10
-r1.0 <- 1.7
-r2.0 <- 1.7
+N0 <- c(10,10)
+r <- c(1.7,1.7)
 alpha.11 <- 0.01
@@
 alpha.21 <- 0.02
+alpha <- matrix(c(alpha.11,alpha.21,alpha.12,alpha.22),nrow=2,byrow=FALSE)
 E.0 <- 0.8
-x1.0 <- 0.8
-x2.0 <- 0.8
-
-# model function
-disc_LV_E <- function(r1,r2,N1.0,N2.0,alpha.11,alpha.22,alpha.12,alpha.21,E,x1,x2) {
-    Nt1 <- ((r1*exp(-(E-x1)^2))*N1.0) / (1+alpha.11*N1.0+alpha.12*N2.0)
-    Nt2 <- ((r2*exp(-(E-x2)^2))*N2.0) / (1+alpha.22*N2.0+alpha.21*N1.0)
-    return(c(Nt1,Nt2))
-}
+x <- c(0.8,0.8)
 # Simulation of model for t time steps
 t <- 40
-N <- array(NA,dim=c(t,2))
+N <- array(NA,dim=c(t,length(N0)))
 N <- as.data.frame(N)
-colnames(N) <- c("N1","N2")
-N$N1[1] <- N1.0
-N$N2[1] <- N2.0
+colnames(N) <- paste0("N",1:length(N0))
+N[1,] <- N0
 E <- rep(NA, t)
@@
 for (i in 2:t) {
-  res <- disc_LV_E(r1=r1.0,r2=r2.0,N1.0=N[i-1,1],N2.0=N[i-1,2],alpha.11=alpha.11,alpha.22=alpha.22,alpha.12=alpha.12,alpha.21=alpha.21,E=E[i-1],x1=x1.0,x2=x2.0)
-  N$N1[i] <- res[1]
-  N$N2[i] <- res[2]
+  N[i,] <- disc_LV_E(r=r,N0=N[i-1,],alpha=alpha,E=E[i-1],x=x)
   E[i] <- E[i-1] + rnorm(1,0,0.001)
@@
 ```
-```{r}
+```{r figs22, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs22} Population size \\textit{N} over time \\textit{t} for a discrete-time logistic growth model with competition and a changing environment, with parameters $r_1 = r_2$ = 1.7, $N_{i,0}$ = 10, $\\alpha_{ii}$ = 0.01, $\\alpha_{12}$ = 0.015, and $\\alpha_{21}$ = 0.02."}
 # Plot simulation: ggplot
@@
 dat <- melt(N, id.vars="time")
-ggplot2::ggplot(dat, aes(time, value, col=variable)) + geom_point() + geom_hline(yintercept = ((r1.0 - 1)/alpha.11),linetype = "dashed", color = "gray")
-```
-```{r}
-# Plot simulation: ggplot
-dat <- as.data.frame(cbind(log(N$N1),log(N$N2)))
-colnames(dat) <- c("N1","N2")
+ggplot2::ggplot(dat, aes(time, value, col=variable)) + geom_point() + ylab("N") + theme_classic()
+```
+```{r}
+## prepare data in HMSC format
+dat <- as.data.frame(log(N))
 dat$time <- 1:t
-df <- data.frame(cbind(dat$N1[2:t],dat$N2[2:t]))
-colnames(df) <- c("Nt1","Nt2")
-# prepare data in HMSC format
-Y <- as.matrix(cbind(df$Nt1,df$Nt2))
-XData <- data.frame(cbind(dat$N1[1:(t-1)],dat$N2[1:(t-1)]),E[1:(t-1)],E[1:(t-1)]^2)
-colnames(XData) <- c("n1","n2","E","Esq")
+df <- data.frame(dat[(2:t),-3])
+Y <- as.matrix(df)
+XData <- data.frame(dat[1:(t-1),-3],E[1:(t-1)],E[1:(t-1)]^2)
+colnames(XData) <- c(paste0("n",1:length(r)),"E","Esq")
 studyDesign = data.frame(sample = as.factor(1:(t-1)))
@@
 With the reduced impact of the environment, we now see the strong negative associations between the two competing species.
-We also correctly see a relatively weak impact of environment for species abundances.
+We also correctly see a relatively weak impact of environment for Species 1 abundance and a stronger impact for Species 2.
+
+For both species, the relative importance of environment has decreased relative to the species association effects.
 
@@
 ```
-```{r}
+```{r figs23, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs23} Observed (points) and HMSC model-predicted population size (logarithm) for Species 1 (black) and Species 2 (orange), using environment $E_t$ as a fixed effect predictor and specifying temporal random effects to capture the species interactions.
+Lines represent the 95% highest probability density interval (HPDI) for predictions at each time point."}
 preds = computePredictedValues(m.6.sample)
@@
 }
-plot(df$Nt1,pch=19,col="black",ylim=c(0,5))
-points(df$Nt2,pch=19,col="orange")
+plot(df$N1,pch=19,col="black",ylim=c(-1,5),ylab="N")
+points(df$N2,pch=19,col="orange")
 lines(df$n1_25,col="black")
@@
 ```
-
-# Species growth, competition, environmental change, and trait evolution
+# Model E: Species growth, competition, environmental change, and trait evolution
 
@@
 We use initial trait values that favor the weaker competitor, and we simulate random increasing shifts in the local environment.
-
-```{r}
-# Case 1: Weaker interactions, stronger environment
-# Simulate initial species population growth with environment fluctuations
-N1.0 <- 10
-N2.0 <- 10
+The simulation uses the `disc_LV_evol` command in the `ecoevor` package.
+
+```{r}
+N0 <- c(10,10)
 alpha.11 <- 0.01
@@
 alpha.21 <- 0.01
+alpha <- matrix(c(alpha.11,alpha.21,alpha.12,alpha.22),nrow=2,byrow=FALSE)
 E.0 <- 0.8
-x1.0 <- 0.1
-x2.0 <- 0.8
+x.0 <- c(0.1,0.8)
 P <- 1
@@
 k <- (w + (1 - h2) * P)/(P + w)
-
-# model function
-disc_LV_evol <- function(N1.0,N2.0,alpha.11,alpha.22,alpha.12,alpha.21,E,x1.0,x2.0,P,w,Wmax,h2) {
-  What <- Wmax*sqrt(w/(P+w))
-  r1 <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E-x1.0))^2)/(2*(P+w)))
-  Nt1 <- (r1*N1.0) / (1+alpha.11*N1.0+alpha.12*N2.0)
-  r2 <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E-x2.0))^2)/(2*(P+w)))
-  Nt2 <- (r2*N2.0) / (1+alpha.22*N2.0+alpha.21*N1.0)
-  return(c(Nt1,Nt2,r1,r2))
-}
 # Simulation of model for t time steps
 t <- 40
-N <- array(NA,dim=c(t,2))
+N <- array(NA,dim=c(t,length(N0)))
 N <- as.data.frame(N)
-colnames(N) <- c("N1","N2")
-x <- array(NA,dim=c(t,2))
-x <- as.data.frame(x)
-colnames(x) <- c("x1","x2")
-r <- array(NA,dim=c(t,2))
-r <- as.data.frame(r)
-colnames(r) <- c("r1","r2")
-N$N1[1] <- N1.0
-N$N2[1] <- N2.0
-x$x1[1] <- x1.0
-x$x2[1] <- x2.0
-What <- Wmax*sqrt(w/(P+w))
-r$r1[1] <- What*(exp((-(((w+(1-h2)*P)/(P+w))*(E.0-x1.0))^2)/(2*(P+w))))
-r$r2[1] <- What*(exp((-(((w+(1-h2)*P)/(P+w))*(E.0-x2.0))^2)/(2*(P+w))))
+colnames(N) <- paste0("N",1:length(N0))
+N[1,] <- N0
 E <- rep(NA, t)
 E[1] <- E.0
+x <- array(NA,dim=c(t,length(N0)))
+x <- as.data.frame(x)
+colnames(x) <- paste0("x",1:length(N0))
+x[1,] <- x.0
+r <- array(NA,dim=c(t,length(N0)))
+r <- as.data.frame(r)
+colnames(r) <- paste0("r",1:length(N0))
+What <- Wmax*sqrt(w/(P+w))
+r[1,] <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E[1]-x[1,]))^2)/(2*(P+w)))
 for (i in 2:t) {
-  res <- disc_LV_evol(N1.0=N[i-1,1],N2.0=N[i-1,2],alpha.11=alpha.11,alpha.22=alpha.22,alpha.12=alpha.12,alpha.21=alpha.21,E=E[i-1],x1=x[i-1,1],x2=x[i-1,2],P=P,w=w,Wmax=Wmax,h2=h2)
-  N$N1[i] <- res[1]
-  N$N2[i] <- res[2]
-  r$r1[i] <- res[3]
-  r$r2[i] <- res[4]
+  res <- disc_LV_evol(N0=N[i-1,],alpha=alpha,E=E[i-1],x=x[i-1,],P=P,w=w,Wmax=Wmax,h2=h2)
+  N[i,] <- res$Nt1
+  r[i,] <- res$r
   # trait change
-  d1 <- E[i-1] - x[i-1,1]
-  d2 <- E[i-1] - x[i-1,2]
-  d1.1 <- k*d1
-  d2.1 <- k*d2
-  x$x1[i] <- E[i-1] - d1.1
-  x$x2[i] <- E[i-1] - d2.1
+  d <- E[i-1] - x[i-1,]
+  d1 <- k*d
+  x[i,] <- E[i-1] - d1
   # environmental change
@@
 ```
-```{r}
+```{r figs24, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs24} Population size \\textit{N} over time \\textit{t} for a discrete-time logistic growth model with competition, a changing environment, and trait evolution.
+Model parameters are $N_{i,0}$ = 10, $\\alpha_{ii}$ = 0.01, $\\alpha_{12}$ = 0.005, and $\\alpha_{21}$ = 0.01."}
 # Plot simulation: ggplot
@@
 dat <- melt(N, id.vars="time")
-ggplot2::ggplot(dat, aes(time, value, col=variable)) + geom_point()
+ggplot2::ggplot(dat, aes(time, value, col=variable)) + geom_point() + ylab("N") + theme_classic()
 ```
@@
 
-```{r}
+```{r figs25, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs25} Population size (logarithm) in species 1 and 2 at one time step $N_{i,t+1}$ as a function of change in trait value of Species 1 and Species 2 in the previous time step $|\\Delta x_{t,t-1}|$."}
 dat2 <- as.data.frame(cbind(log(N$N1),log(N$N2),x$x1,x$x2))
@@
 # Plot
-p1 <- ggplot2::ggplot(df, aes(dx1, Nt1)) + geom_point()
-p2 <- ggplot2::ggplot(df, aes(dx2, Nt1)) + geom_point()
-p3 <- ggplot2::ggplot(df, aes(dx1, Nt2)) + geom_point()
-p4 <- ggplot2::ggplot(df, aes(dx2, Nt2)) + geom_point()
+p1 <- ggplot2::ggplot(df, aes(dx1, Nt1)) + geom_point() + theme_classic()
+p2 <- ggplot2::ggplot(df, aes(dx2, Nt1)) + geom_point() + theme_classic()
+p3 <- ggplot2::ggplot(df, aes(dx1, Nt2)) + geom_point() + theme_classic()
+p4 <- ggplot2::ggplot(df, aes(dx2, Nt2)) + geom_point() + theme_classic()
 p1 + p2 + p3 + p4
@@
 ```{r}
-# Plot simulation: ggplot
-dat <- as.data.frame(cbind(log(N$N1), log(N$N2), x$x1, x$x2))
-colnames(dat) <- c("N1", "N2", "x1", "x2")
+dat <- as.data.frame(cbind(log(N),x))
 dat$time <- 1:t
-df <- data.frame(cbind(dat$N1[2:t], dat$N2[2:t], dat$x1[2:t], dat$x2[2:t]))
-colnames(df) <- c("Nt1", "Nt2", "xt1", "xt2")
-df$dx1 <- abs(dat$x1[2:t] - dat$x1[1:(t - 1)])
-df$dx2 <- abs(dat$x2[2:t] - dat$x2[1:(t - 1)])
-# prepare data in HMSC format
+df <- data.frame(dat[(2:t),-5])
+colnames(df) <- c("Nt1","Nt2","xt1","xt2")
+df$dx1 <- abs(dat$x1[2:t] - dat$x1[1:(t-1)])
+df$dx2 <- abs(dat$x2[2:t] - dat$x2[1:(t-1)])
 Y <- as.matrix(cbind(df$Nt1, df$Nt2))
@@
 rL = HmscRandomLevel(units = studyDesign$sample)
-m.7.hmsc = Hmsc(Y = Y, XData = XData, XFormula = ~E+Esq+dx1+dx2, studyDesign = studyDesign, ranLevels = list(sample = rL.sample))
+m.7.hmsc = Hmsc(Y = Y, XData = XData, XFormula = ~E+Esq+dx1+dx2, studyDesign = studyDesign, ranLevels = list(sample = rL))
 # Bayesian model parameters
@@
 
-```{r echo=FALSE}
+```{r echo=FALSE, figs26, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs26} Observed (black) and HMSC model-predicted population size (logarithm, orange) for Species 1 and Species 2, using environment $E_t$ and change in trait value $|\\Delta x|$ as a fixed effect predictor and specifying temporal random effects to capture the species interactions."}
 pd.1 <- predict(m.7.sample,XData=XData)
@@
 rL = HmscRandomLevel(units = studyDesign$sample)
-m.8.hmsc = Hmsc(Y = Y, XData = XData, XFormula = ~E+Esq+dx1+dx2, studyDesign = studyDesign, ranLevels = list(sample = rL.sample))
+m.8.hmsc = Hmsc(Y = Y, XData = XData, XFormula = ~E+Esq+dx1+dx2, studyDesign = studyDesign, ranLevels = list(sample = rL))
 # Bayesian model parameters
@@
 
-# Species growth, competition, environmental change, and trait evolution in a spatially structured environment
+# Model F: Species growth, competition, environmental change, and trait evolution in a spatially structured environment
 
 We now consider that trait evolution can occur, with an impact on the species growth rate / fitness in the context of the environment.
-
-A final consideration to add to the model is the existence of multiple sites that may have correlation structure in the variance of species abundances.
-We consider this in our simulation by adding multiple (10) sites, with no vs.
-high spatial covariance.
-We then consider spatial random effects in HMSC, where covariance is considered as a function of distance between sites.
+We also add a final consideration to the model, the existence of multiple sites with correlation structure in the variance of species abundances.
+We consider this in our simulation by adding multiple (10) sites, with low vs.
+high spatial covariance in abundances, depending on distance between sites.
+We then include spatial random effects in the HMSC statistical model, where species associations are considered as a function of distance between sites.
 
@@
 # Simulate initial species population growth with environment fluctuations
-N1.0 <- 10
-N2.0 <- 10
+N0 <- c(10,10)
+E.0 <- 0.8
+x.0 <- c(0.1,0.8)
 alpha.11 <- 0.01
@@
 alpha.21 <- 0.01
-E.0 <- 0.8
-x1.0 <- 0.1
-x2.0 <- 0.8
+alpha <- matrix(c(alpha.11,alpha.21,alpha.12,alpha.22),nrow=2,byrow=FALSE)
 P <- 1
@@
 k <- (w + (1 - h2) * P)/(P + w)
-
-# model function
-disc_LV_evol <- function(N1.0,N2.0,alpha.11,alpha.22,alpha.12,alpha.21,E,x1.0,x2.0,P,w,Wmax,h2) {
-  What <- Wmax*sqrt(w/(P+w))
-  r1 <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E-x1.0))^2)/(2*(P+w)))
-  Nt1 <- (r1*N1.0) / (1+alpha.11*N1.0+alpha.12*N2.0)
-  r2 <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E-x2.0))^2)/(2*(P+w)))
-  Nt2 <- (r2*N2.0) / (1+alpha.22*N2.0+alpha.21*N1.0)
-  return(c(Nt1,Nt2,r1,r2))
-}
+j <- 10
 # Simulation of model for t time steps, i sites
-j <- 10
-# random site locations
 xycoords = matrix(runif(2*j), ncol = 2)
@@
 t <- 40
-N <- array(NA,dim=c(j*t,2))
+N <- array(NA,dim=c(j*t,length(N0)))
 N <- as.data.frame(N)
-colnames(N) <- c("N1","N2")
-x <- array(NA,dim=c(j*t,2))
+colnames(N) <- paste0("N",1:length(N0))
+x <- array(NA,dim=c(j*t,length(N0)))
 x <- as.data.frame(x)
-colnames(x) <- c("x1","x2")
-r <- array(NA,dim=c(j*t,2))
+colnames(x) <- paste0("x",1:length(N0))
+r <- array(NA,dim=c(j*t,length(N0)))
 r <- as.data.frame(r)
-colnames(r) <- c("r1","r2")
-N$N1[1:j] <- N1.0
-N$N2[1:j] <- N2.0
-x$x1[1:j] <- x1.0
-x$x2[1:j] <- x2.0
-What <- Wmax*sqrt(w/(P+w))
-r$r1[1:j] <- What*(exp((-(((w+(1-h2)*P)/(P+w))*(E.0-x1.0))^2)/(2*(P+w))))
-r$r2[1:j] <- What*(exp((-(((w+(1-h2)*P)/(P+w))*(E.0-x2.0))^2)/(2*(P+w))))
+colnames(r) <- paste0("r",1:length(N0))
+N[1:j,] <- rep(N0,each=j)
 E <- rep(NA, t)
 E[1] <- E.0
+x[1:j,] <- rep(x.0,each=j)
+What <- Wmax*sqrt(w/(P+w))
+r0 <- as.numeric(What*exp((-(((w+(1-h2)*P)/(P+w))*(E[1]-x[1,]))^2)/(2*(P+w))))
+r[1:j,] <- rep(r0,each=j)
 # Keep track of site and time
@@
   for(z in 1:j){
-    res <- disc_LV_evol(N1.0=N[study$site==z & study$time==i-1,1],N2.0=N[study$site==z & study$time==i-1,2],alpha.11=alpha.11,alpha.22=alpha.22,alpha.12=alpha.12,alpha.21=alpha.21,E=E[i-1],x1=x[study$site==z & study$time==i-1,1],x2=x[study$site==z & study$time==i-1,2],P=P,w=w,Wmax=Wmax,h2=h2)
-    N$N1[study$site==z & study$time==i] <- res[1]
-    N$N2[study$site==z & study$time==i] <- res[2]
-    r$r1[study$site==z & study$time==i] <- res[3]
-    r$r2[study$site==z & study$time==i] <- res[4]
+    res <- disc_LV_evol(N0=N[study$site==z & study$time==i-1,],alpha=alpha,E=E[i-1],x=x[study$site==z & study$time==i-1,],P=P,w=w,Wmax=Wmax,h2=h2)
+    N[study$site==z & study$time==i,] <- res$Nt1
+    r[study$site==z & study$time==i,] <- res$r
     # trait change
-    d1 <- E[i-1] - x[study$site==z & study$time==i-1,1]
-    d2 <- E[i-1] - x[study$site==z & study$time==i-1,2]
-    d1.1 <- k*d1
-    d2.1 <- k*d2
-    x$x1[study$site==z & study$time==i] <- E[i-1] - d1.1
-    x$x2[study$site==z & study$time==i] <- E[i-1] - d2.1
+    d <- E[i-1] - x[study$site==z & study$time==i-1,]
+    d1 <- k*d
+    x[study$site==z & study$time==i,] <- E[i-1] - d1
   }
@@
 ```
-```{r}
+```{r figs27, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs27} Population size \\textit{N} over time \\textit{t} for a discrete-time logistic growth model with competition, a changing environment, and trait evolution.
+Values are shown at 10 sites, with site covariance in species abundances depending on spatial distance."}
 # Plot simulation: ggplot
@@
 dat <- melt(N, id.vars=c("site","time"))
-ggplot2::ggplot(dat, aes(time, value, col=variable)) + geom_point() + facet_wrap(~site,nrow=2)
+ggplot2::ggplot(dat, aes(time, value, col=variable)) + geom_point() + facet_wrap(~site,nrow=2) + ylab("N") + theme_classic()
 ```
@@
 ```{r}
-# Plot simulation: ggplot
-#N[N <0] <- 0.01
-dat <- as.data.frame(cbind(log(N$N1), log(N$N2), x$x1, x$x2))
-colnames(dat) <- c("N1", "N2", "x1", "x2")
-dat$time <- 1:t
+# prepare data in HMSC format
+dat <- as.data.frame(cbind(log(N),x))
+dat$time <- study$time
 df <- data.frame(cbind(dat$N1[dat$time !=1], dat$N2[dat$time !=1], dat$x1[dat$time !=1], dat$x2[dat$time !=1]))
 colnames(df) <- c("Nt1", "Nt2", "xt1", "xt2")
-df$dx1 <- abs(dat$x1[dat$time != 1] - dat$x1[dat$time != t])
-df$dx2 <- abs(dat$x2[dat$time != 1] - dat$x2[dat$time != t])
-# prepare data in HMSC format
 Y <- as.matrix(cbind(df$Nt1, df$Nt2))
-XData <- data.frame(cbind(rep(E[1:(t-1)],times=j),rep(E[1:(t-1)],times=j)^2,abs(dat$x1[dat$time != 1] - dat$x1[dat$time != t]),abs(dat$x2[dat$time != 1] - dat$x2[dat$time != t])))
-colnames(XData) <- c("E","Esq","dx1","dx2")
+XData <- data.frame(E=rep(E[1:(t-1)],each=j))
+XData$Esq <- XData$E^2
+XData$dx1 <- abs(dat$x1[dat$time != 1] - dat$x1[dat$time != t])
+XData$dx2 <- abs(dat$x2[dat$time != 1] - dat$x2[dat$time != t])
 # Update study design to include spatial random effects
@@
 ```
-We can see that the model considers spatial random effects by removing the spatial covariance, and only including spatial variance.
-
-```{r}
-# Case 1: Weaker interactions, stronger environment
-# Simulate initial species population growth with environment fluctuations
-N1.0 <- 10
-N2.0 <- 10
+The HMSC model correctly considers spatial random effects, but they are not strong in this example.
+We evaluate the model's ability to detect increased spatial signal by increasing the spatial covariance in species abundances.
+
+```{r}
+# Simulation of model for t time steps, i sites
+N0 <- c(10,10)
+E.0 <- 0.8
+x.0 <- c(0.1,0.8)
 alpha.11 <- 0.01
@@
 alpha.21 <- 0.01
-E.0 <- 0.8
-x1.0 <- 0.1
-x2.0 <- 0.8
+alpha <- matrix(c(alpha.11,alpha.21,alpha.12,alpha.22),nrow=2,byrow=FALSE) # careful to make sure you have correct interaction matrix
 P <- 1
@@
 k <- (w + (1 - h2) * P)/(P + w)
-
-# model function
-disc_LV_evol <- function(N1.0,N2.0,alpha.11,alpha.22,alpha.12,alpha.21,E,x1.0,x2.0,P,w,Wmax,h2) {
-  What <- Wmax*sqrt(w/(P+w))
-  r1 <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E-x1.0))^2)/(2*(P+w)))
-  Nt1 <- (r1*N1.0) / (1+alpha.11*N1.0+alpha.12*N2.0)
-  r2 <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E-x2.0))^2)/(2*(P+w)))
-  Nt2 <- (r2*N2.0) / (1+alpha.22*N2.0+alpha.21*N1.0)
-  return(c(Nt1,Nt2,r1,r2))
-}
-# Simulation of model for t time steps, i sites
 j <- 10
@@
   for(z in 1:j){
-    res <- disc_LV_evol(N1.0=N[study$site==z & study$time==i-1,1],N2.0=N[study$site==z & study$time==i-1,2],alpha.11=alpha.11,alpha.22=alpha.22,alpha.12=alpha.12,alpha.21=alpha.21,E=E[i-1],x1=x[study$site==z & study$time==i-1,1],x2=x[study$site==z & study$time==i-1,2],P=P,w=w,Wmax=Wmax,h2=h2)
-    N$N1[study$site==z & study$time==i] <- res[1]
-    N$N2[study$site==z & study$time==i] <- res[2]
-    r$r1[study$site==z & study$time==i] <- res[3]
-    r$r2[study$site==z & study$time==i] <- res[4]
+     res <- disc_LV_evol(N0=N[study$site==z & study$time==i-1,],alpha=alpha,E=E[i-1],x=x[study$site==z & study$time==i-1,],P=P,w=w,Wmax=Wmax,h2=h2)
+    N[study$site==z & study$time==i,] <- res$Nt1
+    r[study$site==z & study$time==i,] <- res$r
     # trait change
-    d1 <- E[i-1] - x[study$site==z & study$time==i-1,1]
-    d2 <- E[i-1] - x[study$site==z & study$time==i-1,2]
-    d1.1 <- k*d1
-    d2.1 <- k*d2
-    x$x1[study$site==z & study$time==i] <- E[i-1] - d1.1
-    x$x2[study$site==z & study$time==i] <- E[i-1] - d2.1
+    d <- E[i-1] - x[study$site==z & study$time==i-1,]
+    d1 <- k*d
+    x[study$site==z & study$time==i,] <- E[i-1] - d1
   }
@@
 ```
-```{r}
+```{r figs28, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs28} Population size \\textit{N} over time \\textit{t} for a discrete-time logistic growth model with competition, a changing environment, and trait evolution.
+Values are shown at 10 sites, with an increased site covariance in species abundances, depending on spatial distance."}
 # Plot simulation: ggplot
@@
 dat <- melt(N, id.vars=c("site","time"))
-ggplot2::ggplot(dat, aes(time, value, col=variable)) + geom_point() + facet_wrap(~site,nrow=2)
-```
-```{r}
-# Plot simulation: ggplot
-N[N <0] <- 0.01
-dat <- as.data.frame(cbind(log(N$N1), log(N$N2), x$x1, x$x2))
-colnames(dat) <- c("N1", "N2", "x1", "x2")
-dat$time <- 1:t
+ggplot2::ggplot(dat, aes(time, value, col=variable)) + geom_point() + facet_wrap(~site,nrow=2) + ylab("N") + theme_classic()
+```
+```{r}
+# prepare data in HMSC format
+dat <- as.data.frame(cbind(log(N),x))
+dat$time <- study$time
 df <- data.frame(cbind(dat$N1[dat$time !=1], dat$N2[dat$time !=1], dat$x1[dat$time !=1], dat$x2[dat$time !=1]))
 colnames(df) <- c("Nt1", "Nt2", "xt1", "xt2")
-df$dx1 <- abs(dat$x1[dat$time != 1] - dat$x1[dat$time != t])
-df$dx2 <- abs(dat$x2[dat$time != 1] - dat$x2[dat$time != t])
-# prepare data in HMSC format
 Y <- as.matrix(cbind(df$Nt1, df$Nt2))
-XData <- data.frame(cbind(rep(E[1:(t-1)],times=j),rep(E[1:(t-1)],times=j)^2,abs(dat$x1[dat$time != 1] - dat$x1[dat$time != t]),abs(dat$x2[dat$time != 1] - dat$x2[dat$time != t])))
-colnames(XData) <- c("E","Esq","dx1","dx2")
+XData <- data.frame(E=rep(E[1:(t-1)],each=j))
+XData$Esq <- XData$E^2
+XData$dx1 <- abs(dat$x1[dat$time != 1] - dat$x1[dat$time != t])
+XData$dx2 <- abs(dat$x2[dat$time != 1] - dat$x2[dat$time != t])
 # Update study design to include spatial random effects
@@
 
+We observe a much greater fraction of variation explained by the spatial random effects.
+
 # References

### 2024-11-26 manuscript/manuscript_v13.Rmd
Updated S1, knit.

@@
 
-To fit linear regression models to the simulated population size data, we began with models using population size (model *A*) and environment (including the second-order term, to better reflect species optimum values; model *B*= as fixed effects ($N_t \sim N_{t+1} + E_t + E_t^2$), and progressed to a mixed model using environment and trait evolution as fixed effects ($N_t \sim E_t + E_t^2 + |\Delta x|$; for simulation model *E*) and time (models *C*, *D*, *E*, and *F*) and spatial distance (model *F*) as random effects.
+To fit linear regression models to the simulated population size data, we began with models using population size (model *A*) and environment (including the second-order term, to better reflect species optimum values; model *B*) as fixed effects ($N_t \sim N_{t+1} + E_t + E_t^2$), and progressed to a mixed model using environment and trait evolution as fixed effects ($N_t \sim E_t + E_t^2 + |\Delta x|$; for simulation model *E*) and time (models *C*, *D*, *E*, and *F*) and spatial distance (model *F*) as random effects.
 For consistency, all linear regression models were implemented as hierarchical Bayesian models using HMSC (Hierarchical Modelling of Species Communities [@Ovaskainen2017HMSC] [@Tikhonov2020]).
@@
 
-All simulations used the same parameter values ($W_{max}$ = 2, *P* = 1, *w* = 2, $N_c$ = 100, *p* = 0.001), species interaction matrix ($\alpha_{ii} = 0.00125$; $\alpha_{ij} \sim U(0.0015,1)$), and initial conditions (initial richness $s_{0k} \sim Pois(0.75)$; initial population size $N_{0i} \sim Pois(10)$; initial degree of maladaptation $\beta_{i0} \sim Gamma(0.75,1)$, initial distance to the local patch's optimum phenotype $d_{i0}=\sqrt{B_{i0}(w+P)}$, initial trait value $x_{i0} = E_k - d_{i0}$; Appendix S4).
+All simulations used the same parameter values ($W_{max}$ = 2, *P* = 1, *w* = 2, $N_c$ = 100, *p* = 0.001), species interaction matrix ($\alpha_{ii} = 0.00125$; $\alpha_{ij} \sim U(0,0.0015)$), and initial conditions (initial richness $s_{0k} \sim Pois(0.75)$; initial population size $N_{0i} \sim Pois(10)$; initial degree of maladaptation $\beta_{i0} \sim Gamma(0.75,1)$, initial distance to the local patch's optimum phenotype $d_{i0}=\sqrt{B_{i0}(w+P)}$, initial trait value $x_{i0} = E_k - d_{i0}$; Appendix S4).
 Simulations were run across a range of dispersal values (identical for all species; *d* = 0, $10^{-9},10^{-8},10^{-7},10^{-6},10^{-5},10^{-4},10^{-3},10^{-2}$, 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 1) and heritability values (identical for all species: $h^2$ = 0, 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 1) for 200 time steps, producing records of $N_{ikt}$ and $x_{ikt}$.

### 2024-11-26 code/R/fig1.R

@@ -123,7 +123,6 @@ Wmax <- 2
 h2 <- 1
 k <- (w + (1 - h2) * P)/(P + w)
 # Simulation of model for t time steps
-# Simulation of model for t time steps
 t <- 40
 N <- array(NA,dim=c(t,length(N0)))
 N <- as.data.frame(N)

### 2024-11-29 manuscript/manuscript_v13.Rmd
Edited results, fig1 to svg

@@
 
-Linear regression models can estimate the impacts of intra- and interspecific competition, environmental variation, and trait evolution for population size in simple one- and two-species models of population growth.
-In simple cases (one-species models, linear and quadratic relationships between environment and population size; models *A* and *B*), the regression coefficients can successfully reflect the direction and magnitude of the relationships included in the non-linear simulation model (the impact of $N_t$ and $E_t$ for $N_{t+1}$).
+Researchers often use linear regression models to study non-linear population and community dynamics.
+We evaluated the performance of linear models in simple (1-2 species, 1-few sites) cases, to validate that effect size estimates and predicted densities accurately reflected mechanisms before moving onto more complex models in the next sections of our analysis.
+When only one species was present (models *A* and *B*), the regression coefficients (model *A*: $N_{t+1} \sim \beta_0 + \beta_1 N_t + \epsilon$; *B* $N_{t+1} \sim \beta_0 + \beta_1 N_t + \beta_2 E_t + \beta_3 E_t^2 + \epsilon$) reflected the direction and magnitude of the mechanistic drivers of population size (i.e.
+$\beta_1$ correctly captured the positive relationship between $N_t$ and $N_{t+1}$ under our simulation conditions, and $\beta_2$ and $\beta_3$ correctly captured the quadratic unimodal relationship between $E_t$ and $N_t+1$; Appendix S1.1, S1.2).
+
+When a second species was added (models *C* and *D*), the regression coefficients performed well for predicting density values at all time points, but the direction and magnitude of the regression coefficients did not reflect the intra- and interspecific competition strengths included in the mechanistic model (Appendix S1.3).
+We instead used a latent variable modelling approach (implemented in HMSC) to consider these species interactions as part of the temporal random effects included in the model.
+This approached produced effect size estimates that both predicted the population dynamics well (Figure 1a \@ref(fig:fig1)) and captured the conditional effects of environment for species densities (Figure 1b \@ref(fig:fig1)).
+The variation partition also successfully captured the relative importance of the environmental predictor and the temporal random effects (which in this simple case were driven by intra- and interspecific species interactions; Figure 1c \@ref(fig:fig1); Appendix S1.3).
+
+This approach produced estimates of main effects of environment (Figure 1a,b)
+
 However, as the processes included in the simulation models become slightly more complex (competition with a second species, spatial structure; models *C*, *D*, *E*, *F*), regression coefficients were useful for prediction but not for recreating the underlying processes driving changes in population size (Appendix S1).

### 2024-11-29 code/R/fig1.R

@@ -1,6 +1,6 @@
 ## Code for Figure 1
 ## J.H. Pantel
-## 08-11-2024
+## 29-11-2024
 
 ## libraries
 library(ecoevor)
@@ -12,6 +12,7 @@ library(reshape2)
 library(cowplot)
 library(MASS)
 library(gridGraphics)
+library(svglite)
 ## functions
 source("./code/R/plot_grad_ylim.R")
 
@@ -83,11 +84,17 @@ yrep <- as.matrix(data.frame(hold2))
 ### Mod1 plot ###
 par(mgp=c(2,0.45,0), tcl=-0.4, mar=c(1.3,1.2,0,0))
 color_scheme_set("brightblue")
-#ppc_dens_overlay(y, yrep[1:50, ])
-p1 <- ppc_intervals(y,yrep,x=rep(dat$time[1:(t-1)],times=2),prob = 0.95) +
-  labs(x = "time",y = "ln(N)",) +
-  theme(plot.margin = unit(c(0, 0, 0, 0), "cm")) +
-  theme(legend.position = "none")
+gg_dat <- data.frame(y=y)
+gg_dat$species <- rep(c(1,2),each=39)
+gg_dat$species <- as.factor(gg_dat$species)
+gg_dat$time <- rep((1:(t-1)),times=2)
+gg_dat$ci_025 <- apply(yrep,2,quantile,probs=.025)
+gg_dat$ci_975 <- apply(yrep,2,quantile,probs=.975)
+p1 <- ggplot2::ggplot(gg_dat, aes(time, y, col=species)) + geom_point(size = 0.7) + scale_color_manual(values = c("1" = "skyblue", "2" = "darkgrey")) + geom_errorbar(aes(ymin=ci_025, ymax=ci_975), alpha=0.5) +
+  theme(plot.margin = unit(c(0, 0, 0, 0), "cm"),panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
+        panel.background = element_blank(), axis.line = element_line(colour = "black")) +
+  theme(legend.position = "none",strip.text = element_text(size=6)) + ylab("log(N)")
+
 ### Gradient plot ###
 Gradient <- constructGradient(m.1.sample,focalVariable="E",non.focalVariables=list(Esq=list(2)),ngrid=39)
 Gradient$XDataNew$Esq <- Gradient$XDataNew$E^2
[…]

### 2024-11-30 manuscript/Appendix_S2_v02.Rmd
Updated mscript, all appendices, all figures.

--- 2024-09-07 manuscript/Appendix_S2_v01.Rmd
@@
 library(reshape2)
+library(ecoevor)
 ```
@@
 We simulate population growth under particular parameter values, to generate population dynamics where species interactions, environment, and trait evolution all drive population dynamics.
+The simulation uses the `disc_LV_evol` command in the `ecoevor` package.
 
 ```{r}
-# Case 1: Weaker interactions, stronger environment
-# Simulate initial species population growth with environment fluctuations
-N1.0 <- 10
-N2.0 <- 10
+set.seed(42)
+N0 <- c(10,10)
 alpha.11 <- 0.01
@@
 alpha.21 <- 0.01
+alpha <- matrix(c(alpha.11,alpha.21,alpha.12,alpha.22),nrow=2,byrow=FALSE)
 E.0 <- 0.8
-x1.0 <- 0.6
-x2.0 <- 0.8
+x.0 <- c(0.1,0.8)
 P <- 1
-w <- 10
+w <- 2
 Wmax <- 2
-h2 <- 0.1
-k <- (w+(1-h2)*P)/(P+w)
-
-# model function
-disc_LV_evol <- function(N1.0,N2.0,alpha.11,alpha.22,alpha.12,alpha.21,E,x1.0,x2.0,P,w,Wmax,h2) {
-  What <- Wmax*sqrt(w/(P+w))
-  r1 <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E-x1.0))^2)/(2*(P+w)))
-  Nt1 <- (r1*N1.0) / (1+alpha.11*N1.0+alpha.12*N2.0)
-  r2 <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E-x2.0))^2)/(2*(P+w)))
-  Nt2 <- (r2*N2.0) / (1+alpha.22*N2.0+alpha.21*N1.0)
-  return(c(Nt1,Nt2,r1,r2))
-}
+h2 <- 1
+k <- (w + (1 - h2) * P)/(P + w)
 # Simulation of model for t time steps
 t <- 40
-N <- array(NA,dim=c(t,2))
+N <- array(NA,dim=c(t,length(N0)))
 N <- as.data.frame(N)
-colnames(N) <- c("N1","N2")
-x <- array(NA,dim=c(t,2))
-x <- as.data.frame(x)
-colnames(x) <- c("x1","x2")
-r <- array(NA,dim=c(t,2))
-r <- as.data.frame(r)
-colnames(r) <- c("r1","r2")
-N$N1[1] <- N1.0
-N$N2[1] <- N2.0
-x$x1[1] <- x1.0
-x$x2[1] <- x2.0
-r$r1[1] <- exp((-(((w+(1-h2)*P)/(P+w))*(E.0-x1.0))^2)/(2*(P+w)))
-r$r2[1] <- exp((-(((w+(1-h2)*P)/(P+w))*(E.0-x2.0))^2)/(2*(P+w)))
+colnames(N) <- paste0("N",1:length(N0))
+N[1,] <- N0
 E <- rep(NA, t)
 E[1] <- E.0
+x <- array(NA,dim=c(t,length(N0)))
+x <- as.data.frame(x)
+colnames(x) <- paste0("x",1:length(N0))
+x[1,] <- x.0
+r <- array(NA,dim=c(t,length(N0)))
+r <- as.data.frame(r)
+colnames(r) <- paste0("r",1:length(N0))
+What <- Wmax*sqrt(w/(P+w))
+r[1,] <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E[1]-x[1,]))^2)/(2*(P+w)))
 for (i in 2:t) {
-  res <- disc_LV_evol(N1.0=N[i-1,1],N2.0=N[i-1,2],alpha.11=alpha.11,alpha.22=alpha.22,alpha.12=alpha.12,alpha.21=alpha.21,E=E[i-1],x1=x[i-1,1],x2=x[i-1,2],P=P,w=w,Wmax=Wmax,h2=h2)
-  N$N1[i] <- res[1]
-  N$N2[i] <- res[2]
-  r$r1[i] <- res[3]
-  r$r2[i] <- res[4]
+  res <- disc_LV_evol(N0=N[i-1,],alpha=alpha,E=E[i-1],x=x[i-1,],P=P,w=w,Wmax=Wmax,h2=h2)
+  N[i,] <- res$Nt1
+  r[i,] <- res$r
   # trait change
-  d1 <- E[i-1] - x[i-1,1]
-  d2 <- E[i-1] - x[i-1,2]
-  d1.1 <- k*d1
-  d2.1 <- k*d2
-  x$x1[i] <- E[i-1] - d1.1
-  x$x2[i] <- E[i-1] - d2.1
+  d <- E[i-1] - x[i-1,]
+  d1 <- k*d
+  x[i,] <- E[i-1] - d1
   # environmental change
-  E[i] <- E[i-1] + rnorm(1,0,.01)
+  E[i] <- E[i-1] + abs(rnorm(1, 0, 0.05))
 }
 ```
-```{r}
+```{r figs1, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs1} Population size \\textit{N} over time \\textit{t} for a discrete-time logistic growth model with competition, a changing environment, and trait evolution.
+Model parameters are $N_{i,0}$ = 10, $\\alpha_{ii}$ = 0.01, $\\alpha_{12}$ = 0.005, $\\alpha_{21}$ = 0.01, and $h^2 = 1$."}
 # Plot simulation: ggplot
@@
 dat <- melt(N, id.vars="time")
-ggplot2::ggplot(dat, aes(time, value, col=variable)) + geom_point() + geom_hline(yintercept = (1/alpha.11),linetype = "dashed", color = "gray")
+ggplot2::ggplot(dat, aes(time, value, col=variable)) + geom_point() + ylab("N") + theme_classic()
+```
+
+We can check the impact evolution had for this system by re-running dynamics with $h^2 = 0$:
+
+```{r echo = FALSE}
+h2 <- 0
+k <- (w+(1-h2)*P)/(P+w)
+# Simulation of model for t time steps
+Na <- array(NA,dim=c(t,length(N0)))
+Na <- as.data.frame(Na)
+colnames(Na) <- paste0("N",1:length(N0))
+xa <- array(NA,dim=c(t,length(N0)))
+xa <- as.data.frame(xa)
+colnames(xa) <- paste0("x",1:length(N0))
+ra <- array(NA,dim=c(t,length(N0)))
+ra <- as.data.frame(ra)
+colnames(ra) <- paste0("r",1:length(N0))
+Na[1,] <- N0
+xa[1,] <- x.0
+What <- Wmax*sqrt(w/(P+w))
+ra[1,] <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E[1]-xa[1,]))^2)/(2*(P+w)))
+for (i in 2:t) {
+  res <- disc_LV_evol(N0=Na[i-1,],alpha=alpha,E=E[i-1],x=xa[i-1,],P=P,w=w,Wmax=Wmax,h2=h2)
+  Na[i,] <- res$Nt1
+  ra[i,] <- res$r
+  # trait change
+  d <- E[i-1] - x[i-1,]
+  d1 <- k*d
+  xa[i,] <- E[i-1] - d1
+  # environmental change
+  E[i] <- E[i-1] + abs(rnorm(1, 0, 0.05))
+}
+```
+```{r figs2, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs2} Population size \\textit{N} over time \\textit{t} for a discrete-time logistic growth model with competition, a changing environment, and no trait evolution.
+Model parameters are $N_{i,0}$ = 10, $\\alpha_{ii}$ = 0.01, $\\alpha_{12}$ = 0.005, $\\alpha_{21}$ = 0.01, and $h^2 = 0$."}
+# Plot simulation: ggplot
+Na$time <- 1:t
+data <- melt(Na, id.vars="time")
+ggplot2::ggplot(dat, aes(time, value, col=variable)) + geom_point() + ylab("N") + theme_classic()
 ```
@@
 
-The growth equation is not linear, yet we fit abundance data to a heirarchical model with linear predictors because in most natural systems, a single theoretical growth model can't adequately capture the complexity in a given system.
+The growth equation is not linear, yet we fit abundance data to a hierarchical model with linear predictors because in most natural systems, a single theoretical growth model can't adequately capture the complexity in a given system.
 Linear models don't capture mechanistic relationships, but instead correlative.
@@
 
-```{r}
+```{r figs3, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs3} Population size (logarithm) in species 1 and 2 at one time step $N_{i,t+1}$ as a function of trait value of Species 1 and Species 2 in the previous time step $x_t$."}
 # Plot simulation: ggplot
@@
 # Plot
-p1 <- ggplot2::ggplot(df, aes(xt1, Nt1)) + geom_point()
-p2 <- ggplot2::ggplot(df, aes(xt2, Nt1)) + geom_point()
-p3 <- ggplot2::ggplot(df, aes(xt1, Nt2)) + geom_point()
-p4 <- ggplot2::ggplot(df, aes(xt2, Nt2)) + geom_point()
+p1 <- ggplot2::ggplot(df, aes(xt1, Nt1)) + geom_point() + theme_classic()
+p2 <- ggplot2::ggplot(df, aes(xt2, Nt1)) + geom_point() + theme_classic()
+p3 <- ggplot2::ggplot(df, aes(xt1, Nt2)) + geom_point() + theme_classic()
+p4 <- ggplot2::ggplot(df, aes(xt2, Nt2)) + geom_point() + theme_classic()
 p1 + p2 + p3 + p4
@@
 
-```{r}
+```{r figs4, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs43} Population size (logarithm) in species 1 and 2 at one time step $N_{i,t+1}$ as a function of the absolute value of distance between the environmental optimum trait value given by $E_t$ and the trait value of Species 1 and Species 2 $x_t$."}
 # Plot simulation: ggplot
@@
 # Plot
-p1 <- ggplot2::ggplot(df, aes(abs_Ex1, Nt1)) + geom_point()
-p2 <- ggplot2::ggplot(df, aes(abs_Ex2, Nt1)) + geom_point()
-p3 <- ggplot2::ggplot(df, aes(abs_Ex1, Nt2)) + geom_point()
-p4 <- ggplot2::ggplot(df, aes(abs_Ex2, Nt2)) + geom_point()
+p1 <- ggplot2::ggplot(df, aes(abs_Ex1, Nt1)) + geom_point() + theme_classic()
+p2 <- ggplot2::ggplot(df, aes(abs_Ex2, Nt1)) + geom_point() + theme_classic()
+p3 <- ggplot2::ggplot(df, aes(abs_Ex1, Nt2)) + geom_point() + theme_classic()
+p4 <- ggplot2::ggplot(df, aes(abs_Ex2, Nt2)) + geom_point() + theme_classic()
 p1 + p2 + p3 + p4
@@
 
-We now onsider the magnitude of trait change from one time step to the next as a driver for species abundances.
+We now consider the magnitude of trait change from one time step to the next as a driver for species abundances.
 
-```{r}
+```{r figs5, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs5} Population size (logarithm) in species 1 and 2 at one time step $N_{i,t+1}$ as a function of change in trait value of Species 1 and Species 2 in the previous time step $|\\Delta x_{t,t-1}|$."}
 df$dx1 <- abs(dat$x1[2:t] - dat$x1[1:(t-1)])
@@
 cor(df$dx2, df$abs_Ex2)
+```
+
+```{r figs6, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs6} Relationship between change in trait value of Species 1 and Species 2 in the previous time step $|\\Delta x_{t,t-1}|$ and distance from environmental optimum trait value to species trait value $|E_t - x_t|$."}
 # Plot
-p1 <- ggplot2::ggplot(df, aes(abs_Ex1,dx1)) + geom_point()
-p2 <- ggplot2::ggplot(df, aes(abs_Ex2,dx2)) + geom_point()
+p1 <- ggplot2::ggplot(df, aes(abs_Ex1,dx1)) + geom_point() + theme_classic()
+p2 <- ggplot2::ggplot(df, aes(abs_Ex2,dx2)) + geom_point() + theme_classic()
 p1 + p2
@@
 
-```{r}
+```{r results='hide'}
 # prepare data in HMSC format
@@
 
-```{r}
+```{r results='hide'}
 # After Ovaskainen & Abrego 2020, Chapter 7

### 2024-11-30 manuscript/Appendix_S3_v01.Rmd
Updated mscript, all appendices, all figures.

--- 2024-09-07 manuscript/Appendix_S2_v01.Rmd
@@
 ---
-title: "Appendix S2. Evolutionary drivers of species and community dynamics"
+title: "Appendix S3. The relative importance of evolution in populations and communities"
 author:
@@
 ```{r setup,echo=FALSE,message=FALSE}
+library(ecoevor)
+library(ggplot2)
+library(reshape2)
 library(Hmsc)
-library(ggplot2)
-library(tidyr)
-library(coda)
-library(patchwork)
-library(jtools)
 library(bayesplot)
-library(reshape2)
 ```
@@
 
-# Background
+# Logistic growth and trait evolution
 
-In Appendix S1.5, we evaluated the ability of the statistical model to estimate the impact of trait evolution for species abundances.
-Trait evolution's impact for species abundances can be modeled as a predictor in different ways.
-Here we derive the numerical form of the trait $x_i$ that we hypothesize has the most impact on species abundances.
-
-# Model for population growth, competition, environmental change, and evolution
-
-The growth equation for each species is:
+The goal of this analysis is to show that a hierarchical Bayesian linear model (HMSC) can successfully estimate the relative importance of different drivers of population dynamics, and to evaluate how the importance of evolution for population dynamics changes with heritability $h^2$ level.
+This analysis supports the findings for the **Evolution in metacommunities and HMSC** section in the main text.
+We create simulations of evolutionary rescue in a population, where adaptive evolution rescues a population from extinction, at a speed that depends on the heritability level in the system.
+Population dynamics are modelled as:
 
@@
 
-We simulate population growth under particular parameter values, to generate population dynamics where species interactions, environment, and trait evolution all drive population dynamics.
+where \( \hat{W} \) is calculated as \( \hat{W}=W_{max}\sqrt(\frac{w}{P+w}) \), \(W_{max}\) is the species' maximum per-capita growth rate, *w* is the width of the Gaussian fitness function (which determines the strength of selection, as increasing values indicate a weaker reduction in fitness with distance from optimum trait value), *P* is the width of the distribution of the phenotype *x*, and $h^2$ is the heritability of the trait *x*.
+For the simulation we use \(W_{max}\) = 2, *P* = 1, and *w* = 2.
+
+The change in the average trait value each time step is given by:
+
+$$ d_{i,t+1} = k d_{i,t} $$
+where \( k = \frac{w + (1-h^2) P}{w + P} \) and \( d_{i,t} = E_t - x_{i,t} \).
+
+## Population dynamics simulation
+
+We simulate population dynamics for 1 species, keeping all conditions constant except for heritability, which is evaluated at $h^2 = 0, 0.01, 0.02, 0.03, 0.04, 0.05, 0.075, 0.1, 0.4, 0.9, 1$.
+The simulation code for only $h^2 = 0$ is shown.
 
 ```{r}
-# Case 1: Weaker interactions, stronger environment
-# Simulate initial species population growth with environment fluctuations
-N1.0 <- 10
-N2.0 <- 10
+### 1 species, h2=0 ###
+# Initial conditions
+N0 <- 10
 alpha.11 <- 0.01
-alpha.22 <- 0.01
-alpha.12 <- 0.005
-alpha.21 <- 0.01
+alpha <- matrix(alpha.11,byrow=FALSE)
 E.0 <- 0.8
-x1.0 <- 0.6
-x2.0 <- 0.8
 P <- 1
-w <- 10
+w <- 0.5
+# Draw initial trait value using degree of initial maladaptation
+B0 <- 1
+d0 = sqrt(B0*(w+P))
+x.0 = E.0-d0
+Wmax <- 2
+h2 <- 0
+k <- (w + (1 - h2) * P)/(P + w)
+# Simulation of model for t time steps
+# Simulation of model for t time steps
+t <- 60
+N <- array(NA,dim=c(t,length(N0)))
+N <- as.data.frame(N)
+colnames(N) <- paste0("N",1:length(N0))
+N[1,] <- N0
+E <- rep(NA, t)
+E[1] <- E.0
+x <- array(NA,dim=c(t,length(N0)))
+x <- as.data.frame(x)
+colnames(x) <- paste0("x",1:length(N0))
+x[1,] <- x.0
+r <- array(NA,dim=c(t,length(N0)))
+r <- as.data.frame(r)
+colnames(r) <- paste0("r",1:length(N0))
+What <- Wmax*sqrt(w/(P+w))
+r[1,] <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E[1]-x[1,]))^2)/(2*(P+w)))
+for (i in 2:t) {
+  res <- disc_LV_evol(N0=N[i-1,],alpha=alpha,E=E[i-1],x=x[i-1,],P=P,w=w,Wmax=Wmax,h2=h2)
+  N[i,] <- res$Nt1
+  r[i,] <- res$r
+  # trait change
+  d <- E[i-1] - x[i-1,]
+  d1 <- k*d
+  x[i,] <- E[i-1] - d1
+  # environmental change
+  E[i] <- E[i-1] + abs(rnorm(1, 0, 0))
+}
+# Arrange to save across scenarios
+gdat <- N
+gdat$time <- 1:t
+gdat$h2 <- h2
+gdat$x1 <- x$x1
+gdat$E <- E
+```
+```{r echo=FALSE}
+### 1 species, h2=0.01 ###
+# Initial conditions
+N0 <- 10
+alpha.11 <- 0.01
+alpha <- matrix(alpha.11,byrow=FALSE)
+E.0 <- 0.8
+P <- 1
+w <- 0.5
+# Draw initial trait value using degree of initial maladaptation
+B0 <- 1
+d0 = sqrt(B0*(w+P))
+x.0 = E.0-d0
+Wmax <- 2
+h2 <- 0.01
+k <- (w + (1 - h2) * P)/(P + w)
+# Simulation of model for t time steps
+# Simulation of model for t time steps
+t <- 60
+N <- array(NA,dim=c(t,length(N0)))
+N <- as.data.frame(N)
+colnames(N) <- paste0("N",1:length(N0))
+N[1,] <- N0
+E <- rep(NA, t)
+E[1] <- E.0
+x <- array(NA,dim=c(t,length(N0)))
+x <- as.data.frame(x)
+colnames(x) <- paste0("x",1:length(N0))
+x[1,] <- x.0
+r <- array(NA,dim=c(t,length(N0)))
+r <- as.data.frame(r)
+colnames(r) <- paste0("r",1:length(N0))
+What <- Wmax*sqrt(w/(P+w))
+r[1,] <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E[1]-x[1,]))^2)/(2*(P+w)))
+for (i in 2:t) {
+  res <- disc_LV_evol(N0=N[i-1,],alpha=alpha,E=E[i-1],x=x[i-1,],P=P,w=w,Wmax=Wmax,h2=h2)
+  N[i,] <- res$Nt1
+  r[i,] <- res$r
+  # trait change
+  d <- E[i-1] - x[i-1,]
+  d1 <- k*d
+  x[i,] <- E[i-1] - d1
+  # environmental change
+  E[i] <- E[i-1] + abs(rnorm(1, 0, 0))
+}
+# Arrange to save across scenarios
+gdat <- rbind(gdat,data.frame(N,time=1:t,h2=rep(h2,t),x1=x$x1,E=E))
+```
+```{r echo=FALSE}
+### 1 species, h2=0.02 ###
+# Initial conditions
+N0 <- 10
+alpha.11 <- 0.01
+alpha <- matrix(alpha.11,byrow=FALSE)
+E.0 <- 0.8
+P <- 1
+w <- 0.5
+# Draw initial trait value using degree of initial maladaptation
+B0 <- 1
+d0 = sqrt(B0*(w+P))
+x.0 = E.0-d0
+Wmax <- 2
+h2 <- 0.02
+k <- (w + (1 - h2) * P)/(P + w)
+# Simulation of model for t time steps
+# Simulation of model for t time steps
+t <- 60
+N <- array(NA,dim=c(t,length(N0)))
+N <- as.data.frame(N)
+colnames(N) <- paste0("N",1:length(N0))
+N[1,] <- N0
+E <- rep(NA, t)
+E[1] <- E.0
+x <- array(NA,dim=c(t,length(N0)))
+x <- as.data.frame(x)
+colnames(x) <- paste0("x",1:length(N0))
+x[1,] <- x.0
+r <- array(NA,dim=c(t,length(N0)))
+r <- as.data.frame(r)
+colnames(r) <- paste0("r",1:length(N0))
+What <- Wmax*sqrt(w/(P+w))
+r[1,] <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E[1]-x[1,]))^2)/(2*(P+w)))
+for (i in 2:t) {
+  res <- disc_LV_evol(N0=N[i-1,],alpha=alpha,E=E[i-1],x=x[i-1,],P=P,w=w,Wmax=Wmax,h2=h2)
+  N[i,] <- res$Nt1
+  r[i,] <- res$r
+  # trait change
+  d <- E[i-1] - x[i-1,]
+  d1 <- k*d
+  x[i,] <- E[i-1] - d1
+  # environmental change
+  E[i] <- E[i-1] + abs(rnorm(1, 0, 0))
+}
+# Arrange to save across scenarios
+gdat <- rbind(gdat,data.frame(N,time=1:t,h2=rep(h2,t),x1=x$x1,E=E))
+```
+```{r echo=FALSE}
+### 1 species, h2=0.03 ###
+# Initial conditions
+N0 <- 10
+alpha.11 <- 0.01
+alpha <- matrix(alpha.11,byrow=FALSE)
+E.0 <- 0.8
+P <- 1
+w <- 0.5
+# Draw initial trait value using degree of initial maladaptation
+B0 <- 1
+d0 = sqrt(B0*(w+P))
+x.0 = E.0-d0
+Wmax <- 2
+h2 <- 0.03
+k <- (w + (1 - h2) * P)/(P + w)
+# Simulation of model for t time steps
+# Simulation of model for t time steps
+t <- 60
+N <- array(NA,dim=c(t,length(N0)))
+N <- as.data.frame(N)
+colnames(N) <- paste0("N",1:length(N0))
+N[1,] <- N0
+E <- rep(NA, t)
+E[1] <- E.0
+x <- array(NA,dim=c(t,length(N0)))
+x <- as.data.frame(x)
+colnames(x) <- paste0("x",1:length(N0))
+x[1,] <- x.0
+r <- array(NA,dim=c(t,length(N0)))
+r <- as.data.frame(r)
+colnames(r) <- paste0("r",1:length(N0))
+What <- Wmax*sqrt(w/(P+w))
+r[1,] <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E[1]-x[1,]))^2)/(2*(P+w)))
+for (i in 2:t) {
+  res <- disc_LV_evol(N0=N[i-1,],alpha=alpha,E=E[i-1],x=x[i-1,],P=P,w=w,Wmax=Wmax,h2=h2)
+  N[i,] <- res$Nt1
+  r[i,] <- res$r
+  # trait change
+  d <- E[i-1] - x[i-1,]
+  d1 <- k*d
+  x[i,] <- E[i-1] - d1
+  # environmental change
+  E[i] <- E[i-1] + abs(rnorm(1, 0, 0))
+}
+# Arrange to save across scenarios
+gdat <- rbind(gdat,data.frame(N,time=1:t,h2=rep(h2,t),x1=x$x1,E=E))
+```
+```{r echo=FALSE}
+### 1 species, h2=0.04 ###
+# Initial conditions
+N0 <- 10
+alpha.11 <- 0.01
+alpha <- matrix(alpha.11,byrow=FALSE)
+E.0 <- 0.8
+P <- 1
+w <- 0.5
+# Draw initial trait value using degree of initial maladaptation
+B0 <- 1
+d0 = sqrt(B0*(w+P))
+x.0 = E.0-d0
+Wmax <- 2
+h2 <- 0.04
+k <- (w + (1 - h2) * P)/(P + w)
+# Simulation of model for t time steps
+# Simulation of model for t time steps
+t <- 60
+N <- array(NA,dim=c(t,length(N0)))
+N <- as.data.frame(N)
+colnames(N) <- paste0("N",1:length(N0))
+N[1,] <- N0
+E <- rep(NA, t)
+E[1] <- E.0
+x <- array(NA,dim=c(t,length(N0)))
+x <- as.data.frame(x)
+colnames(x) <- paste0("x",1:length(N0))
+x[1,] <- x.0
+r <- array(NA,dim=c(t,length(N0)))
+r <- as.data.frame(r)
+colnames(r) <- paste0("r",1:length(N0))
+What <- Wmax*sqrt(w/(P+w))
+r[1,] <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E[1]-x[1,]))^2)/(2*(P+w)))
+for (i in 2:t) {
+  res <- disc_LV_evol(N0=N[i-1,],alpha=alpha,E=E[i-1],x=x[i-1,],P=P,w=w,Wmax=Wmax,h2=h2)
+  N[i,] <- res$Nt1
+  r[i,] <- res$r
+  # trait change
+  d <- E[i-1] - x[i-1,]
+  d1 <- k*d
+  x[i,] <- E[i-1] - d1
+  # environmental change
+  E[i] <- E[i-1] + abs(rnorm(1, 0, 0))
+}
+# Arrange to save across scenarios
+gdat <- rbind(gdat,data.frame(N,time=1:t,h2=rep(h2,t),x1=x$x1,E=E))
+```
+```{r echo=FALSE}
+### 1 species, h2=0.05 ###
+# Initial conditions
+N0 <- 10
+alpha.11 <- 0.01
+alpha <- matrix(alpha.11,byrow=FALSE)
+E.0 <- 0.8
+P <- 1
+w <- 0.5
+# Draw initial trait value using degree of initial maladaptation
+B0 <- 1
+d0 = sqrt(B0*(w+P))
+x.0 = E.0-d0
+Wmax <- 2
+h2 <- 0.05
+k <- (w + (1 - h2) * P)/(P + w)
+# Simulation of model for t time steps
+# Simulation of model for t time steps
+t <- 60
+N <- array(NA,dim=c(t,length(N0)))
+N <- as.data.frame(N)
+colnames(N) <- paste0("N",1:length(N0))
+N[1,] <- N0
+E <- rep(NA, t)
+E[1] <- E.0
+x <- array(NA,dim=c(t,length(N0)))
+x <- as.data.frame(x)
+colnames(x) <- paste0("x",1:length(N0))
+x[1,] <- x.0
+r <- array(NA,dim=c(t,length(N0)))
+r <- as.data.frame(r)
+colnames(r) <- paste0("r",1:length(N0))
+What <- Wmax*sqrt(w/(P+w))
+r[1,] <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E[1]-x[1,]))^2)/(2*(P+w)))
+for (i in 2:t) {
+  res <- disc_LV_evol(N0=N[i-1,],alpha=alpha,E=E[i-1],x=x[i-1,],P=P,w=w,Wmax=Wmax,h2=h2)
+  N[i,] <- res$Nt1
+  r[i,] <- res$r
+  # trait change
+  d <- E[i-1] - x[i-1,]
+  d1 <- k*d
+  x[i,] <- E[i-1] - d1
+  # environmental change
+  E[i] <- E[i-1] + abs(rnorm(1, 0, 0))
+}
+# Arrange to save across scenarios
+gdat <- rbind(gdat,data.frame(N,time=1:t,h2=rep(h2,t),x1=x$x1,E=E))
+```
+```{r echo=FALSE}
+### 1 species, h2=0.075 ###
+# Initial conditions
+N0 <- 10
+alpha.11 <- 0.01
+alpha <- matrix(alpha.11,byrow=FALSE)
+E.0 <- 0.8
+P <- 1
+w <- 0.5
+# Draw initial trait value using degree of initial maladaptation
+B0 <- 1
+d0 = sqrt(B0*(w+P))
+x.0 = E.0-d0
+Wmax <- 2
+h2 <- 0.075
+k <- (w + (1 - h2) * P)/(P + w)
+# Simulation of model for t time steps
+# Simulation of model for t time steps
+t <- 60
+N <- array(NA,dim=c(t,length(N0)))
+N <- as.data.frame(N)
+colnames(N) <- paste0("N",1:length(N0))
+N[1,] <- N0
+E <- rep(NA, t)
+E[1] <- E.0
+x <- array(NA,dim=c(t,length(N0)))
+x <- as.data.frame(x)
+colnames(x) <- paste0("x",1:length(N0))
+x[1,] <- x.0
+r <- array(NA,dim=c(t,length(N0)))
+r <- as.data.frame(r)
+colnames(r) <- paste0("r",1:length(N0))
+What <- Wmax*sqrt(w/(P+w))
+r[1,] <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E[1]-x[1,]))^2)/(2*(P+w)))
+for (i in 2:t) {
+  res <- disc_LV_evol(N0=N[i-1,],alpha=alpha,E=E[i-1],x=x[i-1,],P=P,w=w,Wmax=Wmax,h2=h2)
+  N[i,] <- res$Nt1
+  r[i,] <- res$r
+  # trait change
+  d <- E[i-1] - x[i-1,]
+  d1 <- k*d
+  x[i,] <- E[i-1] - d1
+  # environmental change
+  E[i] <- E[i-1] + abs(rnorm(1, 0, 0))
+}
+# Arrange to save across scenarios
+gdat <- rbind(gdat,data.frame(N,time=1:t,h2=rep(h2,t),x1=x$x1,E=E))
+```
+```{r echo=FALSE}
+### 1 species, h2=0.1 ###
+# Initial conditions
+N0 <- 10
+alpha.11 <- 0.01
+alpha <- matrix(alpha.11,byrow=FALSE)
+E.0 <- 0.8
+P <- 1
+w <- 0.5
+# Draw initial trait value using degree of initial maladaptation
+B0 <- 1
+d0 = sqrt(B0*(w+P))
+x.0 = E.0-d0
 Wmax <- 2
 h2 <- 0.1
-k <- (w+(1-h2)*P)/(P+w)
-
-# model function
-disc_LV_evol <- function(N1.0,N2.0,alpha.11,alpha.22,alpha.12,alpha.21,E,x1.0,x2.0,P,w,Wmax,h2) {
-  What <- Wmax*sqrt(w/(P+w))
-  r1 <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E-x1.0))^2)/(2*(P+w)))
-  Nt1 <- (r1*N1.0) / (1+alpha.11*N1.0+alpha.12*N2.0)
-  r2 <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E-x2.0))^2)/(2*(P+w)))
-  Nt2 <- (r2*N2.0) / (1+alpha.22*N2.0+alpha.21*N1.0)
-  return(c(Nt1,Nt2,r1,r2))
-}
+k <- (w + (1 - h2) * P)/(P + w)
 # Simulation of model for t time steps
-t <- 40
-N <- array(NA,dim=c(t,2))
+# Simulation of model for t time steps
+t <- 60
+N <- array(NA,dim=c(t,length(N0)))
 N <- as.data.frame(N)
-colnames(N) <- c("N1","N2")
-x <- array(NA,dim=c(t,2))
-x <- as.data.frame(x)
-colnames(x) <- c("x1","x2")
-r <- array(NA,dim=c(t,2))
-r <- as.data.frame(r)
-colnames(r) <- c("r1","r2")
-N$N1[1] <- N1.0
-N$N2[1] <- N2.0
-x$x1[1] <- x1.0
-x$x2[1] <- x2.0
-r$r1[1] <- exp((-(((w+(1-h2)*P)/(P+w))*(E.0-x1.0))^2)/(2*(P+w)))
-r$r2[1] <- exp((-(((w+(1-h2)*P)/(P+w))*(E.0-x2.0))^2)/(2*(P+w)))
+colnames(N) <- paste0("N",1:length(N0))
+N[1,] <- N0
 E <- rep(NA, t)
 E[1] <- E.0
+x <- array(NA,dim=c(t,length(N0)))
+x <- as.data.frame(x)
+colnames(x) <- paste0("x",1:length(N0))
+x[1,] <- x.0
+r <- array(NA,dim=c(t,length(N0)))
+r <- as.data.frame(r)
+colnames(r) <- paste0("r",1:length(N0))
+What <- Wmax*sqrt(w/(P+w))
+r[1,] <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E[1]-x[1,]))^2)/(2*(P+w)))
 for (i in 2:t) {
-  res <- disc_LV_evol(N1.0=N[i-1,1],N2.0=N[i-1,2],alpha.11=alpha.11,alpha.22=alpha.22,alpha.12=alpha.12,alpha.21=alpha.21,E=E[i-1],x1=x[i-1,1],x2=x[i-1,2],P=P,w=w,Wmax=Wmax,h2=h2)
-  N$N1[i] <- res[1]
-  N$N2[i] <- res[2]
-  r$r1[i] <- res[3]
-  r$r2[i] <- res[4]
+  res <- disc_LV_evol(N0=N[i-1,],alpha=alpha,E=E[i-1],x=x[i-1,],P=P,w=w,Wmax=Wmax,h2=h2)
+  N[i,] <- res$Nt1
+  r[i,] <- res$r
   # trait change
-  d1 <- E[i-1] - x[i-1,1]
-  d2 <- E[i-1] - x[i-1,2]
-  d1.1 <- k*d1
-  d2.1 <- k*d2
-  x$x1[i] <- E[i-1] - d1.1
-  x$x2[i] <- E[i-1] - d2.1
+  d <- E[i-1] - x[i-1,]
+  d1 <- k*d
+  x[i,] <- E[i-1] - d1
   # environmental change
-  E[i] <- E[i-1] + rnorm(1,0,.01)
+  E[i] <- E[i-1] + abs(rnorm(1, 0, 0))
 }
+# Arrange to save across scenarios
+gdat <- rbind(gdat,data.frame(N,time=1:t,h2=rep(h2,t),x1=x$x1,E=E))
 ```
-```{r}
-# Plot simulation: ggplot
-N$time <- 1:t
-dat <- melt(N, id.vars="time")
-ggplot2::ggplot(dat, aes(time, value, col=variable)) + geom_point() + geom_hline(yintercept = (1/alpha.11),linetype = "dashed", color = "gray")
+```{r echo=FALSE}
+### 1 species, h2=0.4 ###
+# Initial conditions
+N0 <- 10
+alpha.11 <- 0.01
+alpha <- matrix(alpha.11,byrow=FALSE)
+E.0 <- 0.8
+P <- 1
+w <- 0.5
+# Draw initial trait value using degree of initial maladaptation
+B0 <- 1
+d0 = sqrt(B0*(w+P))
+x.0 = E.0-d0
+Wmax <- 2
+h2 <- 0.4
+k <- (w + (1 - h2) * P)/(P + w)
+# Simulation of model for t time steps
+# Simulation of model for t time steps
+t <- 60
+N <- array(NA,dim=c(t,length(N0)))
+N <- as.data.frame(N)
+colnames(N) <- paste0("N",1:length(N0))
+N[1,] <- N0
+E <- rep(NA, t)
+E[1] <- E.0
+x <- array(NA,dim=c(t,length(N0)))
+x <- as.data.frame(x)
+colnames(x) <- paste0("x",1:length(N0))
+x[1,] <- x.0
+r <- array(NA,dim=c(t,length(N0)))
+r <- as.data.frame(r)
+colnames(r) <- paste0("r",1:length(N0))
+What <- Wmax*sqrt(w/(P+w))
+r[1,] <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E[1]-x[1,]))^2)/(2*(P+w)))
+for (i in 2:t) {
+  res <- disc_LV_evol(N0=N[i-1,],alpha=alpha,E=E[i-1],x=x[i-1,],P=P,w=w,Wmax=Wmax,h2=h2)
+  N[i,] <- res$Nt1
+  r[i,] <- res$r
+  # trait change
+  d <- E[i-1] - x[i-1,]
+  d1 <- k*d
+  x[i,] <- E[i-1] - d1
+  # environmental change
+  E[i] <- E[i-1] + abs(rnorm(1, 0, 0))
+}
+# Arrange to save across scenarios
+gdat <- rbind(gdat,data.frame(N,time=1:t,h2=rep(h2,t),x1=x$x1,E=E))
+```
+```{r echo=FALSE}
+### 1 species, h2=0.9 ###
+# Initial conditions
+N0 <- 10
+alpha.11 <- 0.01
+alpha <- matrix(alpha.11,byrow=FALSE)
+E.0 <- 0.8
+P <- 1
+w <- 0.5
+# Draw initial trait value using degree of initial maladaptation
+B0 <- 1
+d0 = sqrt(B0*(w+P))
+x.0 = E.0-d0
+Wmax <- 2
+h2 <- 0.9
+k <- (w + (1 - h2) * P)/(P + w)
+# Simulation of model for t time steps
+# Simulation of model for t time steps
+t <- 60
+N <- array(NA,dim=c(t,length(N0)))
+N <- as.data.frame(N)
+colnames(N) <- paste0("N",1:length(N0))
+N[1,] <- N0
+E <- rep(NA, t)
+E[1] <- E.0
+x <- array(NA,dim=c(t,length(N0)))
+x <- as.data.frame(x)
+colnames(x) <- paste0("x",1:length(N0))
+x[1,] <- x.0
+r <- array(NA,dim=c(t,length(N0)))
+r <- as.data.frame(r)
+colnames(r) <- paste0("r",1:length(N0))
+What <- Wmax*sqrt(w/(P+w))
+r[1,] <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E[1]-x[1,]))^2)/(2*(P+w)))
+for (i in 2:t) {
+  res <- disc_LV_evol(N0=N[i-1,],alpha=alpha,E=E[i-1],x=x[i-1,],P=P,w=w,Wmax=Wmax,h2=h2)
+  N[i,] <- res$Nt1
+  r[i,] <- res$r
+  # trait change
+  d <- E[i-1] - x[i-1,]
+  d1 <- k*d
+  x[i,] <- E[i-1] - d1
+  # environmental change
+  E[i] <- E[i-1] + abs(rnorm(1, 0, 0))
+}
+# Arrange to save across scenarios
+gdat <- rbind(gdat,data.frame(N,time=1:t,h2=rep(h2,t),x1=x$x1,E=E))
+```
+```{r echo=FALSE}
+### 1 species, h2=1 ###
+# Initial conditions
+N0 <- 10
+alpha.11 <- 0.01
+alpha <- matrix(alpha.11,byrow=FALSE)
+E.0 <- 0.8
+P <- 1
+w <- 0.5
+# Draw initial trait value using degree of initial maladaptation
+B0 <- 1
+d0 = sqrt(B0*(w+P))
+x.0 = E.0-d0
+Wmax <- 2
+h2 <- 1
+k <- (w + (1 - h2) * P)/(P + w)
+# Simulation of model for t time steps
+# Simulation of model for t time steps
+t <- 60
+N <- array(NA,dim=c(t,length(N0)))
+N <- as.data.frame(N)
+colnames(N) <- paste0("N",1:length(N0))
+N[1,] <- N0
+E <- rep(NA, t)
+E[1] <- E.0
+x <- array(NA,dim=c(t,length(N0)))
+x <- as.data.frame(x)
+colnames(x) <- paste0("x",1:length(N0))
+x[1,] <- x.0
+r <- array(NA,dim=c(t,length(N0)))
+r <- as.data.frame(r)
+colnames(r) <- paste0("r",1:length(N0))
+What <- Wmax*sqrt(w/(P+w))
+r[1,] <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E[1]-x[1,]))^2)/(2*(P+w)))
+for (i in 2:t) {
+  res <- disc_LV_evol(N0=N[i-1,],alpha=alpha,E=E[i-1],x=x[i-1,],P=P,w=w,Wmax=Wmax,h2=h2)
+  N[i,] <- res$Nt1
+  r[i,] <- res$r
+  # trait change
+  d <- E[i-1] - x[i-1,]
+  d1 <- k*d
+  x[i,] <- E[i-1] - d1
+  # environmental change
+  E[i] <- E[i-1] + abs(rnorm(1, 0, 0))
+}
+# Arrange to save across scenarios
+gdat <- rbind(gdat,data.frame(N,time=1:t,h2=rep(h2,t),x1=x$x1,E=E))
 ```
 
-# Hypotheses for evolution as a driver of species abundances
-
-The growth equation is not linear, yet we fit abundance data to a heirarchical model with linear predictors because in most natural systems, a single theoretical growth model can't adequately capture the complexity in a given system.
-Linear models don't capture mechanistic relationships, but instead correlative.
-They are useful for inferring direction and magnitude of effect sizes.
-In Appendix S1, we explore how non-linear dynamics are captured when fit to a linear model.
-Here, we explore different forms to encode evolution in the trait $x$ that determines the organism's fitness in this system.
-
-## H1. Raw trait value drives species abundances
-
-```{r}
-# Plot simulation: ggplot
-dat <- as.data.frame(cbind(log(N$N1),log(N$N2),x$x1,x$x2))
-colnames(dat) <- c("N1","N2","x1","x2")
-dat$time <- 1:t
-df <- data.frame(cbind(dat$N1[2:t],dat$N2[2:t],dat$x1[2:t],dat$x2[2:t]))
-colnames(df) <- c("Nt1","Nt2","xt1","xt2")
-# Plot
-p1 <- ggplot2::ggplot(df, aes(xt1, Nt1)) + geom_point()
-p2 <- ggplot2::ggplot(df, aes(xt2, Nt1)) + geom_point()
-p3 <- ggplot2::ggplot(df, aes(xt1, Nt2)) + geom_point()
-p4 <- ggplot2::ggplot(df, aes(xt2, Nt2)) + geom_point()
-p1 + p2 + p3 + p4
+```{r figs1, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs1} Plot of population size over time for a discete-time evolutionary rescue model, across a range of heritability $h^2$ values."}
+### Plot of time series ###
+gdat$h2 <- as.factor(gdat$h2)
+ggplot2::ggplot(gdat, aes(time, N1, col=h2)) + geom_point()
+### ###
 ```
 
-We do not hypothesize that the raw trait value is the best predictor of species abundances, because the impacts of this evolved trait depend on the context of the environment.
+## HMSC model fit
 
-## H2. Absolute value of trait distance from optimum drives species abundances
+The goal of fitting the data to a statistical model is to estimate the relative important of trait evolution for the population dynamics.
 
-```{r}
-# Plot simulation: ggplot
-df$E <- E[2:t]
-df$abs_Ex1 <- abs(df$E - df$xt1)
-df$abs_Ex2 <- abs(df$E - df$xt2)
-# Plot
-p1 <- ggplot2::ggplot(df, aes(abs_Ex1, Nt1)) + geom_point()
-p2 <- ggplot2::ggplot(df, aes(abs_Ex2, Nt1)) + geom_point()
-p3 <- ggplot2::ggplot(df, aes(abs_Ex1, Nt2)) + geom_point()
-p4 <- ggplot2::ggplot(df, aes(abs_Ex2, Nt2)) + geom_point()
-p1 + p2 + p3 + p4
-```
+```{r results='hide'}
+### Model 1, h2=00 ###
+dat <- cbind(log(gdat$N1[gdat$h2==0]),gdat$x1[gdat$h2==0])
+dat <- as.data.frame(dat)
+colnames(dat) <- c("N1","x1")
+dat$time <- 1:t
+dat <- as.data.frame(dat)
+df <- data.frame(dat[(2:t),-3])
+colnames(df) <- c("Nt1", "xt1")
+df$dx1 <- abs(dat$x1[2:t] - dat$x1[1:(t - 1)])
+# prepare data in HMSC format
+Y <- as.matrix(df$Nt1)
+XData <- data.frame(cbind(E[1:(t-1)],E[1:(t-1)]^2,abs(dat$x1[2:t] - dat$x1[1:(t-1)])))
+colnames(XData) <- c("E","Esq","dx1")
 
-We hypothesize that this represents the true impact on species abundances, via the direct impact this measure has on fitness and consequent population growth.
-However, we also acknowledge this is likely difficult to measure in most empirical systems.
-
-## H3. Absolute value of trait change from one time to the next drives species abundances
-
-We now onsider the magnitude of trait change from one time step to the next as a driver for species abundances.
-
-```{r}
-df$dx1 <- abs(dat$x1[2:t] - dat$x1[1:(t-1)])
-df$dx2 <- abs(dat$x2[2:t] - dat$x2[1:(t-1)])
-# Plot
-p1 <- ggplot2::ggplot(df, aes(dx1, Nt1)) + geom_point()
-p2 <- ggplot2::ggplot(df, aes(dx2, Nt1)) + geom_point()
-p3 <- ggplot2::ggplot(df, aes(dx1, Nt2)) + geom_point()
-p4 <- ggplot2::ggplot(df, aes(dx2, Nt2)) + geom_point()
-p1 + p2 + p3 + p4
-```
-
-The measures correlate highly with one another, as the amount of trait change (the amount of evolution) is determined by the distance to the optimum trait value (given that P and w remain constant in the simulation).
-
-```{r}
-cor(df$dx1, df$abs_Ex1)
-cor(df$dx2, df$abs_Ex2)
-# Plot
-p1 <- ggplot2::ggplot(df, aes(abs_Ex1,dx1)) + geom_point()
-p2 <- ggplot2::ggplot(df, aes(abs_Ex2,dx2)) + geom_point()
-p1 + p2
-```
-
-# Statistical model for evolution as a driver of species abundances
-
-We hypothesize that \( \lvert E_t - x_{i,t} \rvert \) will be the best predictor of species abundances, and that \( \lvert x_{t+1} - x_{i,t} \rvert \) will have similar explanatory power.
-
-```{r}
-# prepare data in HMSC format
-dat <- as.data.frame(cbind(log(N$N1),log(N$N2),x$x1,x$x2))
-colnames(dat) <- c("N1","N2","x1","x2")
-dat$time <- 1:t
-df <- data.frame(cbind(dat$N1[2:t],dat$N2[2:t],dat$x1[2:t],dat$x2[2:t]))
-colnames(df) <- c("Nt1","Nt2","xt1","xt2")
-df$dx1 <- abs(dat$x1[2:t] - dat$x1[1:(t-1)])
-df$dx2 <- abs(dat$x2[2:t] - dat$x2[1:(t-1)])
-# Y matrix
-Y <- as.matrix(cbind(df$Nt1,df$Nt2))
-# X matrix
-XData <- data.frame(cbind(dat$N1[1:(t-1)],dat$N2[1:(t-1)]),E[1:(t-1)],E[1:(t-1)]^2,dat$x1[1:(t-1)],dat$x2[1:(t-1)],abs(dat$x1[2:t] - dat$x1[1:(t-1)]),abs(dat$x2[2:t] - dat$x2[1:(t-1)]),abs(E[1:(t-1)] - dat$x1[1:(t-1)]),abs(E[1:(t-1)] - dat$x2[1:(t-1)]))
-colnames(XData) <- c("n1","n2","E","Esq","x1","x2","dx1","dx2","dEx1","dEx2")
-# Prepare HMSC model
 studyDesign = data.frame(sample = as.factor(1:(t-1)))
 rL = HmscRandomLevel(units = studyDesign$sample)
-# Design and fit 4 alternative HMSC models
-models = list()
-for(i in 1:4){
-  XFormula = switch(i, ~1, ~E+Esq+x1+x2, ~E+Esq+dEx1+dEx2, ~E+Esq+dx1+dx2)
-  m = Hmsc(Y = Y, XData = XData, XFormula = XFormula, studyDesign = studyDesign, ranLevels = list(sample = rL))
-  models[[i]] = m
-}
+
+m.1.hmsc = Hmsc(Y = Y, XData = XData, XFormula = ~E+Esq+dx1, studyDesign = studyDesign, ranLevels = list(sample = rL))
 # Bayesian model parameters
@@
 samples <- 2000
-transient <- 100*thin
+transient <- 1000*thin
 verbose <- 500*thin
-for(i in 1:4){
-  models[[i]] = sampleMcmc(models[[i]],thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
+# sample MCMC
+m.1.sample <- sampleMcmc(m.1.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
+m1.post.hmsc <- convertToCodaObject(m.1.sample)
+```
+
+```{r warning=FALSE}
+VP1 <- computeVariancePartitioning(m.1.sample,group=c(1,1,1,2),groupnames=c("Env","x1"))
+plotVariancePartitioning(m.1.sample,VP1,cols=c("white","skyblue","darkgrey"),args.legend=list(cex=0.75,bg="transparent"))
+```
+```{r figs2, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs2} Simulated (dark blue points) population size over time for $h^2 = 0$, with posterior predicted mean (light blue points) and 95% highest density intervals (light blue bars)."}
+# prepare for plot
+pred2 <- predict(m.1.sample)
+hold1 <- list()
+hold2 <- list()
+for(i in 1:length(N0)){
+  hold1[[i]] <- as.numeric(Y[,i])
+  data <- lapply(pred2, function(x) x[, colnames(x)[i], drop = FALSE])
+  hold2[[i]] <- matrix(unlist(data), nrow=length(pred2), byrow=TRUE)
+}
+y <- unlist(hold1) # vector of each species' observed N across time points
+yrep <- as.matrix(data.frame(hold2))
+### Mod2 plot ###
+par(mgp=c(2,0.45,0), tcl=-0.4, mar=c(1.3,1.2,0,0))
+color_scheme_set("brightblue")
+#ppc_dens_overlay(y, yrep[1:50, ])
+ppc_intervals(y,yrep,x=dat$time[1:(t-1)],prob = 0.95) +
+  labs(x = "time",y = "ln(N)",) +
+  theme(plot.margin = unit(c(0, 0, 0, 0), "cm")) +
+  theme(legend.position = "none")
+```
+
+```{r results='hide'}
+### Model 2, h2=0.01 ###
+dat <- cbind(log(gdat$N1[gdat$h2==0.01]),gdat$x1[gdat$h2==0.01])
+dat <- as.data.frame(dat)
+colnames(dat) <- c("N1","x1")
+dat$time <- 1:t
+dat <- as.data.frame(dat)
+df <- data.frame(dat[(2:t),-3])
+colnames(df) <- c("Nt1", "xt1")
+df$dx1 <- abs(dat$x1[2:t] - dat$x1[1:(t - 1)])
+# prepare data in HMSC format
+Y <- as.matrix(df$Nt1)
+XData <- data.frame(cbind(E[1:(t-1)],E[1:(t-1)]^2,abs(dat$x1[2:t] - dat$x1[1:(t-1)])))
+colnames(XData) <- c("E","Esq","dx1")
+
+studyDesign = data.frame(sample = as.factor(1:(t-1)))
+rL = HmscRandomLevel(units = studyDesign$sample)
+
+m.2.hmsc = Hmsc(Y = Y, XData = XData, XFormula = ~E+Esq+dx1, studyDesign = studyDesign, ranLevels = list(sample = rL))
+# Bayesian model parameters
+nChains <- 2
+thin <- 5
+samples <- 2000
+transient <- 1000*thin
+verbose <- 500*thin
+# sample MCMC
+m.2.sample <- sampleMcmc(m.2.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
+m2.post.hmsc <- convertToCodaObject(m.2.sample)
+```
+```{r}
+m2.post.hmsc <- convertToCodaObject(m.2.sample)
+summary(m2.post.hmsc$Beta)
+bayesplot::mcmc_trace(m2.post.hmsc$Beta)
+bayesplot::mcmc_areas(m2.post.hmsc$Beta,area_method = c("equal height"))
+```
+```{r}
+Gradient <- constructGradient(m.2.sample,focalVariable="dx1",non.focalVariables=list(E=list(2),Esq=list(2)),ngrid=59)
+# Esq is manually constructed as gradient-produced E^2
+Gradient$XDataNew$Esq <- Gradient$XDataNew$E^2
+predY <- predict(m.2.sample,XData=Gradient$XDataNew,expected=TRUE)
+plotGradient(m.2.sample,Gradient,pred=predY,showData=T,measure="Y",main="",xlab="dx1_t",ylab="predicted N_t+1")
+```
+
+We can see that the system is unpredictable, as population size is extremely low during the simulation due to maladaptation.
+
+```{r warning=FALSE}
+VP2 <- computeVariancePartitioning(m.2.sample,group=c(1,1,1,2),groupnames=c("Env","x1"))
+plotVariancePartitioning(m.2.sample,VP2,cols=c("white","skyblue","darkgrey"),args.legend=list(cex=0.75,bg="transparent"))
+```
+```{r figs3, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs3} Simulated (dark blue points) population size over time for $h^2 = 0.01$, with posterior predicted mean (light blue points) and 95% highest density intervals (light blue bars)."}
+# prepare for plot
+pred2 <- predict(m.2.sample)
+hold1 <- list()
+hold2 <- list()
+for(i in 1:length(N0)){
+  hold1[[i]] <- as.numeric(Y[,i])
+  data <- lapply(pred2, function(x) x[, colnames(x)[i], drop = FALSE])
+  hold2[[i]] <- matrix(unlist(data), nrow=length(pred2), byrow=TRUE)
+}
+y <- unlist(hold1) # vector of each species' observed N across time points
+yrep <- as.matrix(data.frame(hold2))
+### Mod2 plot ###
+par(mgp=c(2,0.45,0), tcl=-0.4, mar=c(1.3,1.2,0,0))
+color_scheme_set("brightblue")
+#ppc_dens_overlay(y, yrep[1:50, ])
+ppc_intervals(y,yrep,x=dat$time[1:(t-1)],prob = 0.95) +
+  labs(x = "time",y = "ln(N)",) +
+  theme(plot.margin = unit(c(0, 0, 0, 0), "cm")) +
+  theme(legend.position = "none")
+```
+
+```{r results='hide'}
+### Model 3, h2=0.02 ###
+dat <- cbind(log(gdat$N1[gdat$h2==0.02]),gdat$x1[gdat$h2==0.02])
+dat <- as.data.frame(dat)
+colnames(dat) <- c("N1","x1")
+dat$time <- 1:t
+dat <- as.data.frame(dat)
+df <- data.frame(dat[(2:t),-3])
+colnames(df) <- c("Nt1", "xt1")
+df$dx1 <- abs(dat$x1[2:t] - dat$x1[1:(t - 1)])
+# prepare data in HMSC format
+Y <- as.matrix(df$Nt1)
+XData <- data.frame(cbind(E[1:(t-1)],E[1:(t-1)]^2,abs(dat$x1[2:t] - dat$x1[1:(t-1)])))
+colnames(XData) <- c("E","Esq","dx1")
+
+studyDesign = data.frame(sample = as.factor(1:(t-1)))
+rL = HmscRandomLevel(units = studyDesign$sample)
+
+m.3.hmsc = Hmsc(Y = Y, XData = XData, XFormula = ~E+Esq+dx1, studyDesign = studyDesign, ranLevels = list(sample = rL))
+# Bayesian model parameters
+nChains <- 2
+thin <- 5
+samples <- 2000
+transient <- 1000*thin
+verbose <- 500*thin
+# sample MCMC
+m.3.sample <- sampleMcmc(m.3.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
+m3.post.hmsc <- convertToCodaObject(m.3.sample)
+```
+
+```{r warning=FALSE}
+VP3 <- computeVariancePartitioning(m.3.sample,group=c(1,1,1,2),groupnames=c("Env","x1"))
+plotVariancePartitioning(m.3.sample,VP3,cols=c("white","skyblue","darkgrey"),args.legend=list(cex=0.75,bg="transparent"))
+```
+```{r figs4, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs4} Simulated (dark blue points) population size over time for $h^2 = 0.02$, with posterior predicted mean (light blue points) and 95% highest density intervals (light blue bars)."}
+# prepare for plot
+pred2 <- predict(m.3.sample)
+hold1 <- list()
+hold2 <- list()
+for(i in 1:length(N0)){
+  hold1[[i]] <- as.numeric(Y[,i])
+  data <- lapply(pred2, function(x) x[, colnames(x)[i], drop = FALSE])
+  hold2[[i]] <- matrix(unlist(data), nrow=length(pred2), byrow=TRUE)
+}
+y <- unlist(hold1) # vector of each species' observed N across time points
+yrep <- as.matrix(data.frame(hold2))
+### Mod2 plot ###
+par(mgp=c(2,0.45,0), tcl=-0.4, mar=c(1.3,1.2,0,0))
+color_scheme_set("brightblue")
+#ppc_dens_overlay(y, yrep[1:50, ])
+ppc_intervals(y,yrep,x=dat$time[1:(t-1)],prob = 0.95) +
+  labs(x = "time",y = "ln(N)",) +
+  theme(plot.margin = unit(c(0, 0, 0, 0), "cm")) +
+  theme(legend.position = "none")
+```
+
+```{r results='hide'}
+### Model 4, h2=0.03 ###
+dat <- cbind(log(gdat$N1[gdat$h2==0.03]),gdat$x1[gdat$h2==0.03])
+dat <- as.data.frame(dat)
+colnames(dat) <- c("N1","x1")
+dat$time <- 1:t
+dat <- as.data.frame(dat)
+df <- data.frame(dat[(2:t),-3])
+colnames(df) <- c("Nt1", "xt1")
+df$dx1 <- abs(dat$x1[2:t] - dat$x1[1:(t - 1)])
+# prepare data in HMSC format
+Y <- as.matrix(df$Nt1)
+XData <- data.frame(cbind(E[1:(t-1)],E[1:(t-1)]^2,abs(dat$x1[2:t] - dat$x1[1:(t-1)])))
+colnames(XData) <- c("E","Esq","dx1")
+
+studyDesign = data.frame(sample = as.factor(1:(t-1)))
+rL = HmscRandomLevel(units = studyDesign$sample)
+
+m.4.hmsc = Hmsc(Y = Y, XData = XData, XFormula = ~E+Esq+dx1, studyDesign = studyDesign, ranLevels = list(sample = rL))
+# Bayesian model parameters
+nChains <- 2
+thin <- 5
+samples <- 2000
+transient <- 1000*thin
+verbose <- 500*thin
+# sample MCMC
+m.4.sample <- sampleMcmc(m.4.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
+m4.post.hmsc <- convertToCodaObject(m.4.sample)
+```
+
+```{r warning=FALSE}
+VP4 <- computeVariancePartitioning(m.4.sample,group=c(1,1,1,2),groupnames=c("Env","x1"))
+plotVariancePartitioning(m.4.sample,VP4,cols=c("white","skyblue","darkgrey"),args.legend=list(cex=0.75,bg="transparent"))
+```
+```{r figs5, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs5} Simulated (dark blue points) population size over time for $h^2 = 0.03$, with posterior predicted mean (light blue points) and 95% highest density intervals (light blue bars)."}
+# prepare for plot
+pred2 <- predict(m.4.sample)
+hold1 <- list()
+hold2 <- list()
+for(i in 1:length(N0)){
+  hold1[[i]] <- as.numeric(Y[,i])
+  data <- lapply(pred2, function(x) x[, colnames(x)[i], drop = FALSE])
+  hold2[[i]] <- matrix(unlist(data), nrow=length(pred2), byrow=TRUE)
+}
+y <- unlist(hold1) # vector of each species' observed N across time points
+yrep <- as.matrix(data.frame(hold2))
+### Mod2 plot ###
+par(mgp=c(2,0.45,0), tcl=-0.4, mar=c(1.3,1.2,0,0))
+color_scheme_set("brightblue")
+#ppc_dens_overlay(y, yrep[1:50, ])
+ppc_intervals(y,yrep,x=dat$time[1:(t-1)],prob = 0.95) +
+  labs(x = "time",y = "ln(N)",) +
+  theme(plot.margin = unit(c(0, 0, 0, 0), "cm")) +
+  theme(legend.position = "none")
+```
+```{r results = 'hide'}
+### Model 5, h2=0.04 ###
+dat <- cbind(log(gdat$N1[gdat$h2==0.04]),gdat$x1[gdat$h2==0.04])
+dat <- as.data.frame(dat)
+colnames(dat) <- c("N1","x1")
+dat$time <- 1:t
+dat <- as.data.frame(dat)
+df <- data.frame(dat[(2:t),-3])
+colnames(df) <- c("Nt1", "xt1")
+df$dx1 <- abs(dat$x1[2:t] - dat$x1[1:(t - 1)])
+# prepare data in HMSC format
+Y <- as.matrix(df$Nt1)
+XData <- data.frame(cbind(E[1:(t-1)],E[1:(t-1)]^2,abs(dat$x1[2:t] - dat$x1[1:(t-1)])))
+colnames(XData) <- c("E","Esq","dx1")
+
+studyDesign = data.frame(sample = as.factor(1:(t-1)))
+rL = HmscRandomLevel(units = studyDesign$sample)
+
+m.5.hmsc = Hmsc(Y = Y, XData = XData, XFormula = ~E+Esq+dx1, studyDesign = studyDesign, ranLevels = list(sample = rL))
+# Bayesian model parameters
+nChains <- 2
+thin <- 5
+samples <- 2000
+transient <- 1000*thin
+verbose <- 500*thin
+# sample MCMC
+m.5.sample <- sampleMcmc(m.5.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
+m5.post.hmsc <- convertToCodaObject(m.5.sample)
+```
+
+```{r warning=FALSE}
+VP5 <- computeVariancePartitioning(m.5.sample,group=c(1,1,1,2),groupnames=c("Env","x1"))
+plotVariancePartitioning(m.5.sample,VP5,cols=c("white","skyblue","darkgrey"),args.legend=list(cex=0.75,bg="transparent"))
+```
+```{r figs6, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs6} Simulated (dark blue points) population size over time for $h^2 = 0.01$, with posterior predicted mean (light blue points) and 95% highest density intervals (light blue bars)."}
+# prepare for plot
+pred2 <- predict(m.5.sample)
+hold1 <- list()
+hold2 <- list()
+for(i in 1:length(N0)){
+  hold1[[i]] <- as.numeric(Y[,i])
+  data <- lapply(pred2, function(x) x[, colnames(x)[i], drop = FALSE])
+  hold2[[i]] <- matrix(unlist(data), nrow=length(pred2), byrow=TRUE)
+}
+y <- unlist(hold1) # vector of each species' observed N across time points
+yrep <- as.matrix(data.frame(hold2))
+### Mod2 plot ###
+par(mgp=c(2,0.45,0), tcl=-0.4, mar=c(1.3,1.2,0,0))
+color_scheme_set("brightblue")
+#ppc_dens_overlay(y, yrep[1:50, ])
+ppc_intervals(y,yrep,x=dat$time[1:(t-1)],prob = 0.95) +
+  labs(x = "time",y = "ln(N)",) +
+  theme(plot.margin = unit(c(0, 0, 0, 0), "cm")) +
+  theme(legend.position = "none")
+```
+```{r results='hide'}
+### Model 6, h2=0.05 ###
+dat <- cbind(log(gdat$N1[gdat$h2==0.05]),gdat$x1[gdat$h2==0.05])
+dat <- as.data.frame(dat)
+colnames(dat) <- c("N1","x1")
+dat$time <- 1:t
+dat <- as.data.frame(dat)
+df <- data.frame(dat[(2:t),-3])
+colnames(df) <- c("Nt1", "xt1")
+df$dx1 <- abs(dat$x1[2:t] - dat$x1[1:(t - 1)])
+# prepare data in HMSC format
+Y <- as.matrix(df$Nt1)
+XData <- data.frame(cbind(E[1:(t-1)],E[1:(t-1)]^2,abs(dat$x1[2:t] - dat$x1[1:(t-1)])))
+colnames(XData) <- c("E","Esq","dx1")
+
+studyDesign = data.frame(sample = as.factor(1:(t-1)))
+rL = HmscRandomLevel(units = studyDesign$sample)
+
+m.6.hmsc = Hmsc(Y = Y, XData = XData, XFormula = ~E+Esq+dx1, studyDesign = studyDesign, ranLevels = list(sample = rL))
+# Bayesian model parameters
+nChains <- 2
+thin <- 5
+samples <- 2000
+transient <- 1000*thin
+verbose <- 500*thin
+# sample MCMC
+m.6.sample <- sampleMcmc(m.6.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
+m6.post.hmsc <- convertToCodaObject(m.6.sample)
+```
+
+```{r warning=FALSE}
+VP6 <- computeVariancePartitioning(m.6.sample,group=c(1,1,1,2),groupnames=c("Env","x1"))
+plotVariancePartitioning(m.6.sample,VP6,cols=c("white","skyblue","darkgrey"),args.legend=list(cex=0.75,bg="transparent"))
+```
+```{r figs7, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs7} Simulated (dark blue points) population size over time for $h^2 = 0.02$, with posterior predicted mean (light blue points) and 95% highest density intervals (light blue bars)."}
+# prepare for plot
+pred2 <- predict(m.6.sample)
+hold1 <- list()
+hold2 <- list()
+for(i in 1:length(N0)){
+  hold1[[i]] <- as.numeric(Y[,i])
+  data <- lapply(pred2, function(x) x[, colnames(x)[i], drop = FALSE])
+  hold2[[i]] <- matrix(unlist(data), nrow=length(pred2), byrow=TRUE)
+}
+y <- unlist(hold1) # vector of each species' observed N across time points
+yrep <- as.matrix(data.frame(hold2))
+### Mod2 plot ###
+par(mgp=c(2,0.45,0), tcl=-0.4, mar=c(1.3,1.2,0,0))
+color_scheme_set("brightblue")
+#ppc_dens_overlay(y, yrep[1:50, ])
+ppc_intervals(y,yrep,x=dat$time[1:(t-1)],prob = 0.95) +
+  labs(x = "time",y = "ln(N)",) +
+  theme(plot.margin = unit(c(0, 0, 0, 0), "cm")) +
+  theme(legend.position = "none")
+```
+
+```{r results='hide'}
+### Model 7, h2=0.075 ###
+dat <- cbind(log(gdat$N1[gdat$h2==0.075]),gdat$x1[gdat$h2==0.075])
+dat <- as.data.frame(dat)
+colnames(dat) <- c("N1","x1")
+dat$time <- 1:t
+dat <- as.data.frame(dat)
+df <- data.frame(dat[(2:t),-3])
+colnames(df) <- c("Nt1", "xt1")
+df$dx1 <- abs(dat$x1[2:t] - dat$x1[1:(t - 1)])
+# prepare data in HMSC format
+Y <- as.matrix(df$Nt1)
+XData <- data.frame(cbind(E[1:(t-1)],E[1:(t-1)]^2,abs(dat$x1[2:t] - dat$x1[1:(t-1)])))
+colnames(XData) <- c("E","Esq","dx1")
+
+studyDesign = data.frame(sample = as.factor(1:(t-1)))
+rL = HmscRandomLevel(units = studyDesign$sample)
+
+m.7.hmsc = Hmsc(Y = Y, XData = XData, XFormula = ~E+Esq+dx1, studyDesign = studyDesign, ranLevels = list(sample = rL))
+# Bayesian model parameters
+nChains <- 2
+thin <- 5
+samples <- 2000
+transient <- 1000*thin
+verbose <- 500*thin
+# sample MCMC
+m.7.sample <- sampleMcmc(m.7.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
+m7.post.hmsc <- convertToCodaObject(m.7.sample)
+```
+
+```{r warning=FALSE}
+VP7 <- computeVariancePartitioning(m.7.sample,group=c(1,1,1,2),groupnames=c("Env","x1"))
+plotVariancePartitioning(m.7.sample,VP7,cols=c("white","skyblue","darkgrey"),args.legend=list(cex=0.75,bg="transparent"))
+```
+```{r figs8, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs8} Simulated (dark blue points) population size over time for $h^2 = 0.075$, with posterior predicted mean (light blue points) and 95% highest density intervals (light blue bars)."}
+# prepare for plot
+pred2 <- predict(m.7.sample)
+hold1 <- list()
+hold2 <- list()
+for(i in 1:length(N0)){
+  hold1[[i]] <- as.numeric(Y[,i])
+  data <- lapply(pred2, function(x) x[, colnames(x)[i], drop = FALSE])
+  hold2[[i]] <- matrix(unlist(data), nrow=length(pred2), byrow=TRUE)
+}
+y <- unlist(hold1) # vector of each species' observed N across time points
+yrep <- as.matrix(data.frame(hold2))
+### Mod2 plot ###
+par(mgp=c(2,0.45,0), tcl=-0.4, mar=c(1.3,1.2,0,0))
+color_scheme_set("brightblue")
+#ppc_dens_overlay(y, yrep[1:50, ])
+ppc_intervals(y,yrep,x=dat$time[1:(t-1)],prob = 0.95) +
+  labs(x = "time",y = "ln(N)",) +
+  theme(plot.margin = unit(c(0, 0, 0, 0), "cm")) +
+  theme(legend.position = "none")
+```
+```{r results='hide'}
+### Model 8, h2=0.1 ###
+dat <- cbind(log(gdat$N1[gdat$h2==0.1]),gdat$x1[gdat$h2==0.1])
+dat <- as.data.frame(dat)
+colnames(dat) <- c("N1","x1")
+dat$time <- 1:t
+dat <- as.data.frame(dat)
+df <- data.frame(dat[(2:t),-3])
+colnames(df) <- c("Nt1", "xt1")
+df$dx1 <- abs(dat$x1[2:t] - dat$x1[1:(t - 1)])
+# prepare data in HMSC format
+Y <- as.matrix(df$Nt1)
+XData <- data.frame(cbind(E[1:(t-1)],E[1:(t-1)]^2,abs(dat$x1[2:t] - dat$x1[1:(t-1)])))
+colnames(XData) <- c("E","Esq","dx1")
+
+studyDesign = data.frame(sample = as.factor(1:(t-1)))
+rL = HmscRandomLevel(units = studyDesign$sample)
+
+m.8.hmsc = Hmsc(Y = Y, XData = XData, XFormula = ~E+Esq+dx1, studyDesign = studyDesign, ranLevels = list(sample = rL))
+# Bayesian model parameters
+nChains <- 2
+thin <- 5
+samples <- 2000
+transient <- 1000*thin
+verbose <- 500*thin
+# sample MCMC
+m.8.sample <- sampleMcmc(m.8.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
+m8.post.hmsc <- convertToCodaObject(m.8.sample)
+```
+
+```{r warning=FALSE}
+VP8 <- computeVariancePartitioning(m.8.sample,group=c(1,1,1,2),groupnames=c("Env","x1"))
+plotVariancePartitioning(m.8.sample,VP8,cols=c("white","skyblue","darkgrey"),args.legend=list(cex=0.75,bg="transparent"))
+```
+```{r figs9, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs9} Simulated (dark blue points) population size over time for $h^2 = 1$, with posterior predicted mean (light blue points) and 95% highest density intervals (light blue bars)."}
+# prepare for plot
+pred2 <- predict(m.8.sample)
+hold1 <- list()
+hold2 <- list()
+for(i in 1:length(N0)){
+  hold1[[i]] <- as.numeric(Y[,i])
+  data <- lapply(pred2, function(x) x[, colnames(x)[i], drop = FALSE])
+  hold2[[i]] <- matrix(unlist(data), nrow=length(pred2), byrow=TRUE)
+}
+y <- unlist(hold1) # vector of each species' observed N across time points
+yrep <- as.matrix(data.frame(hold2))
+### Mod2 plot ###
+par(mgp=c(2,0.45,0), tcl=-0.4, mar=c(1.3,1.2,0,0))
+color_scheme_set("brightblue")
+#ppc_dens_overlay(y, yrep[1:50, ])
+ppc_intervals(y,yrep,x=dat$time[1:(t-1)],prob = 0.95) +
+  labs(x = "time",y = "ln(N)",) +
+  theme(plot.margin = unit(c(0, 0, 0, 0), "cm")) +
+  theme(legend.position = "none")
+```
+```{r results='hide'}
+### Model 9, h2=0.4 ###
+dat <- cbind(log(gdat$N1[gdat$h2==0.4]),gdat$x1[gdat$h2==0.4])
+dat <- as.data.frame(dat)
+colnames(dat) <- c("N1","x1")
+dat$time <- 1:t
+dat <- as.data.frame(dat)
+df <- data.frame(dat[(2:t),-3])
+colnames(df) <- c("Nt1", "xt1")
+df$dx1 <- abs(dat$x1[2:t] - dat$x1[1:(t - 1)])
+# prepare data in HMSC format
+Y <- as.matrix(df$Nt1)
+XData <- data.frame(cbind(E[1:(t-1)],E[1:(t-1)]^2,abs(dat$x1[2:t] - dat$x1[1:(t-1)])))
+colnames(XData) <- c("E","Esq","dx1")
+
+studyDesign = data.frame(sample = as.factor(1:(t-1)))
+rL = HmscRandomLevel(units = studyDesign$sample)
+
+m.9.hmsc = Hmsc(Y = Y, XData = XData, XFormula = ~E+Esq+dx1, studyDesign = studyDesign, ranLevels = list(sample = rL))
+# Bayesian model parameters
+nChains <- 2
+thin <- 5
+samples <- 2000
+transient <- 1000*thin
+verbose <- 500*thin
+# sample MCMC
+m.9.sample <- sampleMcmc(m.9.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
+m9.post.hmsc <- convertToCodaObject(m.9.sample)
+```
+
+```{r warning=FALSE}
+VP9 <- computeVariancePartitioning(m.9.sample,group=c(1,1,1,2),groupnames=c("Env","x1"))
+plotVariancePartitioning(m.9.sample,VP9,cols=c("white","skyblue","darkgrey"),args.legend=list(cex=0.75,bg="transparent"))
+```
+```{r figs10, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs10} Simulated (dark blue points) population size over time for $h^2 = 0.4$, with posterior predicted mean (light blue points) and 95% highest density intervals (light blue bars)."}
+# prepare for plot
+pred2 <- predict(m.9.sample)
+hold1 <- list()
+hold2 <- list()
+for(i in 1:length(N0)){
+  hold1[[i]] <- as.numeric(Y[,i])
+  data <- lapply(pred2, function(x) x[, colnames(x)[i], drop = FALSE])
+  hold2[[i]] <- matrix(unlist(data), nrow=length(pred2), byrow=TRUE)
+}
+y <- unlist(hold1) # vector of each species' observed N across time points
+yrep <- as.matrix(data.frame(hold2))
+### Mod2 plot ###
+par(mgp=c(2,0.45,0), tcl=-0.4, mar=c(1.3,1.2,0,0))
+color_scheme_set("brightblue")
+#ppc_dens_overlay(y, yrep[1:50, ])
+ppc_intervals(y,yrep,x=dat$time[1:(t-1)],prob = 0.95) +
+  labs(x = "time",y = "ln(N)",) +
+  theme(plot.margin = unit(c(0, 0, 0, 0), "cm")) +
+  theme(legend.position = "none")
+```
+```{r results='hide'}
+### Model 10, h2=0.9 ###
+dat <- cbind(log(gdat$N1[gdat$h2==0.9]),gdat$x1[gdat$h2==0.9])
+dat <- as.data.frame(dat)
+colnames(dat) <- c("N1","x1")
+dat$time <- 1:t
+dat <- as.data.frame(dat)
+df <- data.frame(dat[(2:t),-3])
+colnames(df) <- c("Nt1", "xt1")
+df$dx1 <- abs(dat$x1[2:t] - dat$x1[1:(t - 1)])
+# prepare data in HMSC format
+Y <- as.matrix(df$Nt1)
+XData <- data.frame(cbind(E[1:(t-1)],E[1:(t-1)]^2,abs(dat$x1[2:t] - dat$x1[1:(t-1)])))
+colnames(XData) <- c("E","Esq","dx1")
+
+studyDesign = data.frame(sample = as.factor(1:(t-1)))
+rL = HmscRandomLevel(units = studyDesign$sample)
+
+m.10.hmsc = Hmsc(Y = Y, XData = XData, XFormula = ~E+Esq+dx1, studyDesign = studyDesign, ranLevels = list(sample = rL))
+# Bayesian model parameters
+nChains <- 2
+thin <- 5
+samples <- 2000
+transient <- 1000*thin
+verbose <- 500*thin
+# sample MCMC
+m.10.sample <- sampleMcmc(m.10.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
+m10.post.hmsc <- convertToCodaObject(m.10.sample)
+```
+
+```{r warning=FALSE}
+VP10 <- computeVariancePartitioning(m.10.sample,group=c(1,1,1,2),groupnames=c("Env","x1"))
+plotVariancePartitioning(m.10.sample,VP10,cols=c("white","skyblue","darkgrey"),args.legend=list(cex=0.75,bg="transparent"))
+```
+```{r figs11, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs11} Simulated (dark blue points) population size over time for $h^2 = 0.9$, with posterior predicted mean (light blue points) and 95% highest density intervals (light blue bars)."}
+# prepare for plot
+pred2 <- predict(m.10.sample)
+hold1 <- list()
+hold2 <- list()
+for(i in 1:length(N0)){
+  hold1[[i]] <- as.numeric(Y[,i])
+  data <- lapply(pred2, function(x) x[, colnames(x)[i], drop = FALSE])
+  hold2[[i]] <- matrix(unlist(data), nrow=length(pred2), byrow=TRUE)
+}
+y <- unlist(hold1) # vector of each species' observed N across time points
+yrep <- as.matrix(data.frame(hold2))
+### Mod2 plot ###
+par(mgp=c(2,0.45,0), tcl=-0.4, mar=c(1.3,1.2,0,0))
+color_scheme_set("brightblue")
+#ppc_dens_overlay(y, yrep[1:50, ])
+ppc_intervals(y,yrep,x=dat$time[1:(t-1)],prob = 0.95) +
+  labs(x = "time",y = "ln(N)",) +
+  theme(plot.margin = unit(c(0, 0, 0, 0), "cm")) +
+  theme(legend.position = "none")
+```
+```{r results='hide'}
+### Model 11, h2=1 ###
+dat <- cbind(log(gdat$N1[gdat$h2==1]),gdat$x1[gdat$h2==1])
+dat <- as.data.frame(dat)
+colnames(dat) <- c("N1","x1")
+dat$time <- 1:t
+dat <- as.data.frame(dat)
+df <- data.frame(dat[(2:t),-3])
+colnames(df) <- c("Nt1", "xt1")
+df$dx1 <- abs(dat$x1[2:t] - dat$x1[1:(t - 1)])
+# prepare data in HMSC format
+Y <- as.matrix(df$Nt1)
+XData <- data.frame(cbind(E[1:(t-1)],E[1:(t-1)]^2,abs(dat$x1[2:t] - dat$x1[1:(t-1)])))
+colnames(XData) <- c("E","Esq","dx1")
+
+studyDesign = data.frame(sample = as.factor(1:(t-1)))
+rL = HmscRandomLevel(units = studyDesign$sample)
+
+m.11.hmsc = Hmsc(Y = Y, XData = XData, XFormula = ~E+Esq+dx1, studyDesign = studyDesign, ranLevels = list(sample = rL))
+# Bayesian model parameters
+nChains <- 2
+thin <- 5
+samples <- 2000
+transient <- 1000*thin
+verbose <- 500*thin
+# sample MCMC
+m.11.sample <- sampleMcmc(m.11.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
+m11.post.hmsc <- convertToCodaObject(m.11.sample)
+```
+```{r}
+m11.post.hmsc <- convertToCodaObject(m.11.sample)
+summary(m11.post.hmsc$Beta)
+bayesplot::mcmc_trace(m11.post.hmsc$Beta)
+bayesplot::mcmc_areas(m11.post.hmsc$Beta,area_method = c("equal height"))
+```
+```{r warning=FALSE}
+VP11 <- computeVariancePartitioning(m.11.sample,group=c(1,1,1,2),groupnames=c("Env","x1"))
+plotVariancePartitioning(m.11.sample,VP11,cols=c("white","skyblue","darkgrey"),args.legend=list(cex=0.75,bg="transparent"))
+```
+```{r figs12, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs12} Simulated (dark blue points) population size over time for $h^2 = 1$, with posterior predicted mean (light blue points) and 95% highest density intervals (light blue bars)."}
+# prepare for plot
+pred2 <- predict(m.11.sample)
+hold1 <- list()
+hold2 <- list()
+for(i in 1:length(N0)){
+  hold1[[i]] <- as.numeric(Y[,i])
+  data <- lapply(pred2, function(x) x[, colnames(x)[i], drop = FALSE])
+  hold2[[i]] <- matrix(unlist(data), nrow=length(pred2), byrow=TRUE)
+}
+y <- unlist(hold1) # vector of each species' observed N across time points
+yrep <- as.matrix(data.frame(hold2))
+### Mod2 plot ###
+par(mgp=c(2,0.45,0), tcl=-0.4, mar=c(1.3,1.2,0,0))
+color_scheme_set("brightblue")
+#ppc_dens_overlay(y, yrep[1:50, ])
+ppc_intervals(y,yrep,x=dat$time[1:(t-1)],prob = 0.95) +
+  labs(x = "time",y = "ln(N)",) +
+  theme(plot.margin = unit(c(0, 0, 0, 0), "cm")) +
+  theme(legend.position = "none")
+```
+
+```{r figs13, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs13} Estimation of linear regression coefficient for trait evolution across $h^2$ levels.", warning=FALSE, message=FALSE}
+mc_dat_1 <- mcmc_intervals_data(m1.post.hmsc$Beta, pars="B[dx1 (C4), sp1 (S1)]")
+mc_dat_2 <- mcmc_intervals_data(m2.post.hmsc$Beta, pars="B[dx1 (C4), sp1 (S1)]")
+mc_dat_3 <- mcmc_intervals_data(m3.post.hmsc$Beta, pars="B[dx1 (C4), sp1 (S1)]")
+mc_dat_4 <- mcmc_intervals_data(m4.post.hmsc$Beta, pars="B[dx1 (C4), sp1 (S1)]")
+mc_dat_5 <- mcmc_intervals_data(m5.post.hmsc$Beta, pars="B[dx1 (C4), sp1 (S1)]")
+mc_dat_6 <- mcmc_intervals_data(m6.post.hmsc$Beta, pars="B[dx1 (C4), sp1 (S1)]")
+mc_dat_7 <- mcmc_intervals_data(m7.post.hmsc$Beta, pars="B[dx1 (C4), sp1 (S1)]")
+mc_dat_8 <- mcmc_intervals_data(m8.post.hmsc$Beta, pars="B[dx1 (C4), sp1 (S1)]")
+mc_dat_9 <- mcmc_intervals_data(m9.post.hmsc$Beta, pars="B[dx1 (C4), sp1 (S1)]")
+mc_dat_10 <- mcmc_intervals_data(m10.post.hmsc$Beta, pars="B[dx1 (C4), sp1 (S1)]")
+mc_dat_11 <- mcmc_intervals_data(m11.post.hmsc$Beta, pars="B[dx1 (C4), sp1 (S1)]")
+
+combined <- rbind(mc_dat_1,mc_dat_2,mc_dat_3,mc_dat_4,mc_dat_5,mc_dat_6,mc_dat_7,mc_dat_8,mc_dat_9,mc_dat_10,mc_dat_11)
+combined$model <- rep(c("0","0.01","0.02","0.03","0.04","0.05","0.075","0.1","0.4","0.9","1"),times=c(dim(mc_dat_1)[1],dim(mc_dat_2)[1],dim(mc_dat_3)[1],dim(mc_dat_4)[1],dim(mc_dat_5)[1],dim(mc_dat_6)[1],dim(mc_dat_7)[1],dim(mc_dat_8)[1],dim(mc_dat_9)[1],dim(mc_dat_10)[1],dim(mc_dat_11)[1]))
+
+# make the plot using ggplot 
+theme_set(bayesplot::theme_default())
+pos <- position_nudge(y = c(0,0.1,0.2,0.3,0.4,0.5,0.6,0.7,0.8,0.9,1))
+ggplot(combined, aes(x = m, y = parameter, color = model)) + 
+  geom_linerange(aes(xmin = l, xmax = h), position = pos, size=2)+
+  geom_linerange(aes(xmin = ll, xmax = hh), position = pos)+
+  geom_point(aes(color=model),position = pos)+
+  coord_flip()+
+  geom_vline(xintercept=0,linetype="dashed")
+```
+
+We can see that the variation in Y is a combination of the relative importance of density-dependence and of trait evolution.
+We can see that relative importance changes across $h^2$ levels.
+What drives changes in this relative importance?
+
+```{r}
+d_dat <- gdat[gdat$time != 1,c(1,2,3,5)]
+d_dat <- as.data.frame(d_dat)
+d_dat$dxt <- NA
+d_dat$ddt <- NA
+d_dat$dN <- NA
+d_dat$dd <- NA
+for(i in 1:11){
+  sub <- gdat[gdat$h2==levels(gdat$h2)[i],]
+  d_dat$dxt[d_dat$h2==levels(d_dat$h2)[i]] <- abs(sub$x1[2:60]-sub$x1[1:59]) # absolute value of change in trait value
+  d_dat$ddt[d_dat$h2==levels(d_dat$h2)[i]] <- alpha[1,1]*sub$N1[2:60] - alpha[1,1]*sub$N1[1:59]
+  d_dat$dN[d_dat$h2==levels(d_dat$h2)[i]] <- sub$N1[2:60] - sub$N1[1:59]
+  d_dat$dd[d_dat$h2==levels(d_dat$h2)[i]] <- alpha[1,1]*d_dat$N1[d_dat$h2==levels(d_dat$h2)[i]]
 }
 ```
-
-We check the explanatory and predictive power of each model using cross-validation.
-
-```{r}
-# After Ovaskainen & Abrego 2020, Chapter 7
-partition = createPartition(m,nfolds=2,column="sample")
-partition.sp = c(1,2)
-result = matrix(NA,nrow=3,ncol=4)
-for(i in 1:4){
-  m = models[[i]]
-  # Explanatory power
-  preds = computePredictedValues(m)
-  MF = evaluateModelFit(hM=m,predY=preds)
-  result[1,i] = mean(MF$R2)
-  # Predictive power based on cross-validation
-  preds = computePredictedValues(m,partition=partition)
-  MF = evaluateModelFit(hM=m,predY=preds)
-  result[2,i] <- mean(MF$R2)
-  # Predictive power based on conditional cross-validation
-  preds = computePredictedValues(m,partition=partition,partition.sp=partition.sp,mcmcStep=100)
-  MF = evaluateModelFit(hM=m,predY=preds)
-  result[3,i] = mean(MF$R2)
-}
+```{r figs14, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs14} Density dependence varies in importance for change in population size (dN) across heritability levels.
+When heritability is very low, $|\\Delta x|$ can be higher for more time steps, as the population is slow to adapt to the optimum trait value.
+This effect diminishes with increasing $h^2$.
+Density dependence ($dd = (\\alpha_{ii} N_t) -  (\\alpha_{ii} N_{t-1})$) on the other hand increases with increasing heritability.", warning=FALSE, message=FALSE}
+### Plot of time xx ###
+d_dat <- d_dat[d_dat$h2 != 0,]
+d_dat$ln_N <- log(d_dat$N1)
+ggplot2::ggplot(d_dat, aes(dxt, dN, col=h2)) + geom_line() + geom_point()
+ggplot2::ggplot(d_dat, aes(ddt, dN, col=h2)) + geom_line() + geom_point()
+### ###
 ```
-
-```{r}
-colnames(result) <- c("null","xt","dEx","dx")
-rownames(result) <- c("R2_expl","R2_pred","R2_pred_cross")
-result
-```
-
-An important conclusion from this analysis is that the \( \lvert E_t - x_{i,t} \rvert \) and \( \lvert x_{t+1} - x_{i,t} \rvert \) do have similar explanatory power, and that - even though the species traits were the stronger predictor of species abundances - the model with total change in trait value can be used to assess the question "How does evolution impact species abundances?"
-
-```{r}
-m.post.hmsc <- convertToCodaObject(models[[4]])
-bayesplot::mcmc_trace(m.post.hmsc$Beta)
-bayesplot::mcmc_areas(m.post.hmsc$Beta,area_method = c("equal height"))
-```
-
-```{r}
-# Bayesian estimates
-summary(m.post.hmsc$Beta)$statistics[1:10,1:2]
-```
-
-```{r}
-OmegaCor = computeAssociations(models[[4]])
-supportLevel = 0.7
-toPlot = ((OmegaCor[[1]]$support>supportLevel)
-+ (OmegaCor[[1]]$support<(1-supportLevel))>0)*OmegaCor[[1]]$mean
-corrplot::corrplot(toPlot, method = "color",col = colorRampPalette(c("blue","white","red"))(200),title = paste("random effect level:", models[[4]]$rLNames[1]), mar=c(0,0,1,0))
-OmegaCor
-```
-We can also look at a gradient plot for the effect of evolution in Species 1 for abundance in Species 1 and Species 2.
-
-```{r}
-Gradient <- constructGradient(models[[4]],focalVariable="dx1",non.focalVariables=list(E=list(2),Esq=list(2),dx2=list(2)),ngrid=39)
-Gradient$XDataNew$Esq <- Gradient$XDataNew$E^2
-predY <- predict(models[[4]],XData=Gradient$XDataNew,expected=TRUE)
-a <- plotGradient(models[[4]],Gradient,pred=predY,showData=T,measure="Y",index=1,main="",xlab="dx1_t",ylab="predicted N1_t+1")
-b <- plotGradient(models[[4]],Gradient,pred=predY,showData=T,measure="Y",index=2,main="",xlab="dx1_t",ylab="predicted N2_t+1")
-
-postBeta = getPostEstimate(models[[4]], parName = "Beta")
-plotBeta(models[[4]], post = postBeta, param = "Mean", supportLevel = 0.7)
-```
-We repeat this plot for the effect of evolution in Species 2 for abundance in Species 1 and Species 2.
-
-```{r}
-Gradient <- constructGradient(models[[4]],focalVariable="dx2",non.focalVariables=list(E=list(2),Esq=list(2),dx1=list(2)),ngrid=39)
-Gradient$XDataNew$Esq <- Gradient$XDataNew$E^2
-predY <- predict(models[[4]],XData=Gradient$XDataNew,expected=TRUE)
-a <- plotGradient(models[[4]],Gradient,pred=predY,showData=T,measure="Y",index=1,main="",xlab="dx2_t",ylab="predicted N1_t+1")
-b <- plotGradient(models[[4]],Gradient,pred=predY,showData=T,measure="Y",index=2,main="",xlab="dx2_t",ylab="predicted N2_t+1")
-```
-To evaluate the relative importance of evolution for species abundances, we use variation partition:
-
-```{r warning=FALSE}
-VP <- computeVariancePartitioning(models[[4]],group=c(1,1,1,2,3),groupnames=c("Env","Sp1","Sp2"))
-plotVariancePartitioning(models[[4]],VP,cols=c("white","skyblue","darkgrey","orange"),args.legend=list(cex=0.75,bg="transparent"))
-```
-We see that trait evolution in Species 1 is an important driver of abundances in both Species 1 and Species 2.

### 2024-11-30 manuscript/manuscript_v13.Rmd
Updated mscript, all appendices, all figures.

@@
 output:
-  word_document:
-    reference_docx: docx_template.docx
+  pdf_document:
+    toc: no
   bookdown::html_document2: default
@@
     number_sections: yes
-  pdf_document:
+    keep_tex: true
+  word_document:
+    reference_docx: docx_template.docx
+  html_document:
     toc: no
+    df_print: paged
 header-includes:
@@
 
-XXXXX
+Biodiversity at the metacommunity scale is typically influenced by a number of environmental, spatial, biotic, and stochastic factors.
+At the same time, these factors impact the evolution of individual species, as sites present different local selection pressures and connectivity can impact gene flow and genetic drift.
+Identifying the relative impacts of environmental, spatial, biotic, and other drivers on community composition across spatial and temporal scales has been greatly facilitated by joint species distribution models, but these models have yet to consider the impact of microevolution on community composition.
+We used Heirarchical Models of Species Communities (HMSC) to analyze simulated data of an populations and communities, including in a large evolving metacommunity model, to establish whether HMSC can sufficiently quantify the contribution of phenotypic evolution for metacommunity composition.
+The models successfully partitioned variance contributed by environmental, spatial, and evolving phenotypic drivers, and also estimated site- and year-specific covariance.
+We also applied the HMSC with trait evolution model to an existing dataset studying trait change and community dynamics in experimental aquatic plant system.
+The study of eco-evolutionary dynamics may require data that reflects numerous complex, interacting processes and it is necessary to have flexible, generalized statistical models to analyze this data.
+JSDM models such as HMSC present one promising path for analysis of eco-evolutionary dynamics in multi-species communities.
 
@@
 We evaluated whether linear regression models can appropriately capture the dynamics that drive species abundances in relatively simple populations and communities.
-We fit data generated from simulation models with increasing complexity (\@ref(tab:tab1)) to Bayesian linear regression models.
+We fit data generated from simulation models with increasing complexity (Table \@ref(tab:tab1)).
+to Bayesian linear regression models.
 These simulation models were (A) a single-species model of discrete-time logistic population growth, (B) expanded to include a randomly fluctuating environmental property $E$ that impacts the species' population growth rate, (C) a two-species model of logistic growth and competition that was then (D) expanded to consider a randomly fluctuating environmental property, (E) and further expanded to consider evolution in a trait $x$ that is selected by the environmental property, and finally (F) the model of two-species growth, competition, environmental change, and evolution in a spatially structured (10-patch) habitat.
@@
 
-To determine the relative importance of evolution for community composition, we fitted two HMSC models to the data: An 'evolution model' in which $\Delta \hat{x}$ of each species and environmental conditions were used as explanatory variables, and a 'no-evolution model', in which only the environmental conditions were included.
+To determine the relative importance of evolution for community composition, we fitted two HMSC models to the data: An 'evolution model' in which $|\Delta x|$ of each species and environmental conditions were used as explanatory variables, and a 'no-evolution model', in which only the environmental conditions were included.
 Species density was log-transformed species before applying it to the model, therefore we could use normal distribution in the model.
@@
 For each model we ran four MCMC chains to generate 1000 samples of the posterior distribution for all model parameters.
-To evaluate the model-estimated role of evolution for predicted population size, we used the posterior distributions of the estimated effect size of $|\Delta x|$ multiplied by the observed $\Delta \hat{x}$ of each species,added to the model-estimated intercept.
+To evaluate the model-estimated role of evolution for predicted population size, we used the posterior distributions of the estimated effect size of $|\Delta x|$ multiplied by the observed $\Delta x$ of each species, added to the model-estimated intercept.
 
@@
 We instead used a latent variable modelling approach (implemented in HMSC) to consider these species interactions as part of the temporal random effects included in the model.
-This approached produced effect size estimates that both predicted the population dynamics well (Figure 1a \@ref(fig:fig1)) and captured the conditional effects of environment for species densities (Figure 1b \@ref(fig:fig1)).
-The variation partition also successfully captured the relative importance of the environmental predictor and the temporal random effects (which in this simple case were driven by intra- and interspecific species interactions; Figure 1c \@ref(fig:fig1); Appendix S1.3).
+This approached produced effect size estimates that both predicted the population dynamics well (Figure \@ref(fig:fig1) a) and captured the conditional effects of environment for species densities (Figure \@ref(fig:fig1) b).
+The variation partition also successfully captured the relative importance of the environmental predictor and the temporal random effects for model *D* (which in this simple case were driven by intra- and interspecific species interactions; Figure \@ref(fig:fig1) c; Appendix S1.3, S1.4).
 
-This approach produced estimates of main effects of environment (Figure 1a,b)
+Trait evolution was included as a predictor when it was introduced in model *E*.
+The resulting linear regression coefficients could predict species dynamics (Figure \@ref(fig:fig1) d) and reflected the impact and magnitude of environmental and trait evolution predictors (Figure \@ref(fig:fig1) e).
+The variation partition also showed that trait evolution in species 1 ($|\Delta x_1|$) was a strong driver of variation in abundance (blue portion, Figure \@ref(fig:fig1) f), primarily for species 1 but also for species 2 as well (Appendix S1.5; Appendix S2).
 
-However, as the processes included in the simulation models become slightly more complex (competition with a second species, spatial structure; models *C*, *D*, *E*, *F*), regression coefficients were useful for prediction but not for recreating the underlying processes driving changes in population size (Appendix S1).
+After adding spatial complexity to the 2-species simulation, we included a spatial random effect to capture species covariances as a function of distance between sites (model *F*).
+By properly considering species interactions as a temporal random effect and including spatial random effects, the HMSC model could estimate regression coefficients for the impacts of the environment $E_t$ and trait evolution $|\Delta x_1|$ and $|\Delta x_2|$ for both species that fit well to the simulated data points (Figure \@ref(fig:fig1 g); Appendix S1.6) and capture conditional effects of environment and trait evolution for both species (Figure \@ref(fig:fig1) h).
+Spatial random effects explained a large proportion of variance in abundance for both species, but trait evolution remained a strong driver for species 1.
 
-To consider species interactions and environmental fluctuations (model *D*; Figure 1a \@ref(fig:fig1)), we used the latent variable modelling approach implemented in HMSC.
-The linear model successfully estimated coefficients for fixed environmental effects that captured the impacts of environment for population size in both species (Figure 1b) and partitioned the relative importance of the environment and species interactions for species abundances (Figure 1c).
-The model coefficients could also predict the population size values over time (Figure 1a), though the quality of this prediction varied depending on the level of environmental noise (Appendix S1).
-
-```{r fig1, out.width="90%", include=TRUE, fig.align="center", fig.cap=c("Simulated and model-estimated values for population dynamics in two-species systems.
-Panels a, d, and g show log-population size (y-axis) over time (x-axis) for species $i$ and $j$ when they grow and compete for resources in sites with random environmental fluctuations over time.
+```{r fig1, out.width="110%", include=TRUE, fig.align="center", fig.cap=c("Simulation models for 2-species ecological and eco-evolutionary dynamics.
+Panels a, d, and g show log-population size (y-axis) over time (x-axis) for species $1$ and $2$ when they grow and compete for resources with random environmental fluctuations over time.
 Species either have fixed (panel a) or evolving trait values (panels d, g) and are present in either 1 (panels a, d; simulation models D and E) or multiple sites (panel g; model F).
@@
 Graphs in panels c, f, and i show the results of a variation partition after fitting data to the HMSC statistical model.
-The variation partition plot shows the proportion of variation explained by environmental, trait evolution ($|\\Delta x_1|$ and $|\\Delta x_2|$), time random, and spatial random predictors for $ln(N)$ for species 1 and 2.
-Error bars and shading represent the 95% highest probability density interval for posterior predictions."), echo=FALSE}
-knitr::include_graphics("../output/fig1.pdf")
+The variation partition plot shows the proportion of variation explained by environment, trait evolution ($|\\Delta x_1|$ and $|\\Delta x_2|$), and temporal and spatial random effects for $ln(N)$ for species 1 and 2.
+Error bars and shading in panels a, d, and g represent the 95% highest probability density interval for posterior predictions."), echo=FALSE}
+knitr::include_graphics("../output/fig1_mod.pdf",error=FALSE)
 ```
-
-The HMSC model successfully modelled changes in species abundances when trait evolution was introduced (model *E*; Figure 1d \@ref(fig:fig1)).
-Trait evolution played a strong role for population size, especially the effect of $|\Delta x_1|$ for $ln(N_1)$ and $ln(N_2)$ (gradient plots in Figure 1e).
-The model estimated the relative importance of the environmental driver, trait evolution in both species, and species interactions (Figure 1f), and predicted the observed species abundances well (Figure 1d).
-Spatial structure was also reproduced in the coefficients of the linear HMSC model by using spatial random effects to capture covariance in species abundances as a function of spatial distance between sites in the 10-patch simulation model (model *F*; Figure 1g,h,i \@ref(fig:fig1)).
-The variation partition successfully reflected separation of environmental drivers from trait evolution, species interactions, and spatial random effects (Figure 1i), and we show in Appendix S1 that the relative importance of trait evolution increases when we use a simulation model with less spatial and environmental variation and weak species interactions.
 
@@
 
-The simulation model followed population size and trait value dynamics for species in the landscape across 200 time steps (Figure 2a \@ref(fig:fig2)).
-Resulting biodiversity dynamics depended on the dispersal level and heritability values (Figure 2b \@ref(fig:fig2)).
+The simulation model followed population size and trait value dynamics for species in the landscape across 200 time steps (Figure \@ref(fig:fig2) a).
+Resulting biodiversity dynamics depended on the dispersal level and heritability values (Figure \@ref(fig:fig2) b).
 For dispersal, results mirrored some expectations from existing metacommunity models, i.e.
-that regional $\gamma$ diversity decreases with increasing dispersal level ([@Mouquet2003]; Figure 2b \@ref(fig:fig2)).
+that regional $\gamma$ diversity decreases with increasing dispersal level $d$ ([@Mouquet2003]; Figure \@ref(fig:fig2) b).
 The inclusion of trait heritability had a strong effect on community diversity, reflecting that local adaptation can rescue species and contribute to richness at the local and metacommunity scale.
-Regional species diversity ($\gamma$) increased with increasing $h^2$ levels (Figure 2b \@ref(fig:fig2)).
+Regional species diversity ($\gamma$) increased with increasing $h^2$ levels (Figure \@ref(fig:fig2) b).
 
-Analysis using the HMSC model for community time series across $d$ and $h^2$ levels indicated that the importance of evolution for community composition ($V_{|\Delta x|}$) decreased with increasing dispersal and with increasing heritability (Figure 2c,d,e \@ref(fig:fig2)).
+Analysis using the HMSC model for community time series across $d$ and $h^2$ levels indicated that the importance of evolution for community composition ($V_{|\Delta x|}$) decreased with increasing dispersal and with increasing heritability (Figure \@ref(fig:fig2) c).
 The importance of spatial random effects increased with dispersal rate, and overall the importance of evolution decreased with increasing $h^2$.
+These effects can be better understood by our supplemental analyses of 1-species, 1-site simulation models of evolutionary rescue (Appendix S3).
 
-```{r fig2, out.width="90%", include=TRUE, fig.align="center", fig.cap=c("Importance of trait evolution for change in community composition in metacommunities.
+```{r fig2, out.width="110%", include=TRUE, fig.align="center", fig.cap=c("Importance of trait evolution for change in community composition in metacommunities.
 (a) We developed a simulation model for population growth, competition, and trait evolution in 15 species that inhabit a patchy landscape.
@@
 The top panel is for $h^2 = 0.1$, the middle panel is for $h^2 = 0.3$, and the bottom panel is for $h^2 = 0.9$."), echo=FALSE}
-knitr::include_graphics("../output/fig2.pdf")
+knitr::include_graphics("../output/fig2_mod.pdf",error=FALSE)
 ```
@@
 
-The empirical data [@Jewell2023] showed the community dynamics of four plant species under different environmental conditions, additionally the front size of each species was measured to determine its trait change over time.
-We calculated the mean variance partitioning (VP) over all species, to evaluate the impact of evolutionary (change of the front size trait) and ecological (light and nutrient treatments) effects on changes on species density.
-Almost one fifth (18.6%) of the variance in species density could be explained by the variance in $\Delta \hat{x}$ (\@ref(fig:fig5)a).
-This is due to the reduction of the VP of the ecological effects by a total of 14.4% (from 37% to 26.6% for the amount of shading and from 31% to 27.1% for the nutrient concentrations) and the reduction of 4.2% of the random effects (from 18.7% to 15% for time and from 13.3% to 12.8% for replicates) (\@ref(fig:fig5)).
-While $\Delta \hat{x}$ of the species Lm, Sp and Wc showed a similarly low impact, on average 3.6%, $\Delta \hat{x}$ of species Rn showed a high impact on variance of species density, 7.8%, which can be explained by its high $\Delta \hat{x}$ changes over time (see Appendix S05).
-The high density intervals (HDI) of 95% estimated beta effects showed a negative correlation of Light shading with species density, while nutrient concentration correlated positively with density (\@ref(fig:fig5)c).
-Changes in $\Delta \hat{x}$ impacted densities differently.
-At one hand changes of species Rn frond size correlated negatively with is density, on the other hand changes in the frond size of species Wc showed a positive correlation with the other species Sp and Lm (\@ref(fig:fig5)c).
-The impact of evolution (of all species summed up), impacted species density over time, yet did not show a clear direction over the course of the experiment (\@ref(fig:fig5)d).
+The results for the HMSC analysis of the empirical data (community dynamics and frond size of four aquatic plant species under different environmental conditions) estimated an effect of trait change for species density.
+We used the HMSC models to calculate the mean proportion of variance explained (variance partitioning, VP) over all species, to estimate the relative importance of evolutionary (change in the frond size trait) and ecological (light and nutrient treatments) effects on species densities.Almost one fifth (18.6%) of the variance in species density could be explained by the variance in $|\Delta x|$ (\@ref(fig:fig3)a).
+Including trait evolution in the statistical model accurately led to reductions for the estimated importance of ecological effects by a total of 14.4% (from 37% to 26.6% for the amount of shading and from 31% to 27.1% for the nutrient concentrations) and the reduction of 4.2% of the random effects (from 18.7% to 15% for time and from 13.3% to 12.8% for replicates) (\@ref(fig:fig3)).
+While $|\Delta x|$ of the species Lm, Sp and Wc showed a similarly low impact (3.7%, 3.4%, and 3.7% respectively), $|\Delta x|$ of species Rn showed a higher impact on species density, 7.8%, which can be explained by its high $|\Delta x|$ changes over time (see Appendix S4 data).
+The 95% HPDI for main effect sizes showed a negative effect of light shading for species density, while nutrient concentration had a positive estimated effect for density (\@ref(fig:fig3)c).
+Changes in $|\Delta x|$ impacted densities in distinct ways for intra- and interspecific comparisons.
+The frond size of species Rn had a negative impact on Rn density, while change in Wc frond size had a positive estimated effect on species Sp and Lm (\@ref(fig:fig3)c).
+We show the estimated effect of trait evolution for species counts in \@ref(fig:fig3)d, which reflects potentially strong impacts in particular species, density, and light treatments and weak effects with high uncertainty for many cases.
 
-```{r fig3, out.width="90%", include=TRUE, fig.align="center", fig.cap=c("The estimated effects of species trait changes and environmental treatments.
-The variance partitioning of all fixed effects of the HMSC model (a) including trait changes with envrionmental treatments and (b) only including the environmental treamtents.
-The (b) high density interval of 95% showed a clear pattern for the different environmental treatments, yet differing results for trait changes and their impact on species density.
-This can be seen in (d) the shifting predictions of density over time due to the total trait changes of all species."), echo=FALSE}
-knitr::include_graphics("../output/fig5.pdf")
+```{r fig3, out.width="110%", include=TRUE, fig.align="center", fig.cap=c("The estimated effects of species trait changes and environmental treatments for densities of experimental aquatic plants (species Lm, Rn, Sp, and Wc).
+The variation partition for the HMSC model with (a) and without (b) trait change $|\\Delta x|$ is shown, as well as (c) the posterior estimated direction of effect (negative or positive; effect sizes with 95% HPDI that included 0 are colored white).
+The results indicated negative impacts of light, positive impacts of nutrients, negative intraspecific effects of $|\\Delta x_{Rn}|$ for $log(N_{Rn})$, negative interspecific effects of $|\\Delta x_{Lm}|$ for $log(N_{Sp})$, and positive interspecific effects of $|\\Delta x_{Wc}|$ for $log(N_{Sp})$ and $log(N_{Lm})$.
+(d) The posterior effect size estimates were used to estimate model-predicted densities at each time step of the experiment as a function only of $|\\Delta x|$ (for each of the four species, averaged across all of the other treatment effects).
+This captures the expected change in density due to trait evolution in all species."), echo=FALSE}
+knitr::include_graphics("../output/fig3.pdf",error=FALSE)
 ```
@@
 
-*Main findings*
 The analysis of community composition often requires statistical models that can consider the combined influences of environmental drivers and spatial connectivity.
@@
 Finally, one element that has not been fully implemented into joint species distribution models is dynamic trait evolution (see [@Garzon2019] for a model that does consider reaction norms from in-situ reciprocal transplant experiments).
-We used an existing JSDM and developed simulation models of evolving populations in increasing complex scenarios to determine whether the signal of trait evolution could be estimated from data and compared to other drivers of community structure.
-We also applied this statistical model (HMSC, hierarchical models of species distributions; [@Ovaskainen2017HMSC] [@Tikhonov2020]) to an existing empirical dataset (from the experimental study of community eco-evolutionary dynamics in aquatic plants of [@Jewell2023]) to show how HMSC can be used to model the changes in species composition over time as a function of environmental drivers, trait change, and temporal random effects.
+We used an existing JSDM, and developed simulation models of evolving populations in increasing complex scenarios to determine whether the signal of trait evolution could be estimated from data and compared to other drivers of community structure.
+We also applied this statistical model (HMSC; [@Ovaskainen2017HMSC] [@Tikhonov2020]) to an existing empirical dataset (from the experimental study of community eco-evolutionary dynamics in aquatic plants of [@Jewell2023]) to show how HMSC can be used to model the changes in species composition over time as a function of environmental drivers, trait change, and temporal random effects.
 Overall, there is no barrier to including dynamic trait change as a fixed effect in a JSDM model, and the resulting effect size coefficient estimates can be used to (i) estimate the relative importance of trait evolution for community composition and (ii) predict the impacts of future trait change in the system.
 
-*Discussion regarding simple case studies*
 At the same time, fixed effects for the impacts of trait evolution on population size should be interpreted with a great deal of caution.
@@
 In this instance, the regression coefficient from the model may be negative and strong.
-By using simulations of trait evolution and population growth in simple systems (1- and 2-species systems, at single and simplified multi-patch scales; \@ref(fig:fig1); Appendix S1), we can better interpret effect size estimates.
+By using simulations of trait evolution and population growth in simple systems (1- and 2-species systems, at single and simplified multi-patch scales; Figure \@ref(fig:fig1); Appendix S1), we can better interpret effect size estimates.
 However, these precautions are not unique to including trait evolution as a predictor of community composition.
@@
 
-*Discussion regarding metacom simulations*
-This was also observed when using HMSC to evaluate the importance of evolution for metacommunity structure, using simulated data from a model of population growth, competition, dispersal, and trait evolution in a spatially structured habitat (Figure 2a \@ref(fig:fig1); Appendix S4).
+This was also observed when using HMSC to evaluate the importance of evolution for metacommunity structure, using simulated data from a model of population growth, competition, dispersal, and trait evolution in a spatially structured habitat (Figure \@ref(fig:fig1) a).
 First of all, the simulations themselves revealed an intriguing role of adaptive evolution for rescuing individual species from extinction (i.e.
 [@Gomulkiewicz1995]) and as a consequence greatly increasing biodiversity at the local and the landscape scale (i.e.
-community rescue, [@Low2015] [@Bell2019]; Figure 2b \@ref(fig:fig2)).
+community rescue, [@Low2015] [@Bell2019]; Figure \@ref(fig:fig2) b).
 The recovery of species via local adaptation had consequent effects for coexistence, as species that were initially mal-adapted could recover in population size and exert strong competitive effects on other species.
-Because HMSC can accurately estimate the relative importance of trait evolution for community composition (Appendix S4), we were thus able to test hypotheses about the drivers of increased or decreased impacts of trait evolution for community structure.
+Because HMSC can accurately estimate the relative importance of trait evolution for community composition (Appendix S3), we were thus able to test hypotheses about the drivers of increased or decreased impacts of trait evolution for community structure.
 Our simulation models fit to HMSC confirmed our hypotheses that the importance of evolution would decrease with increasing landscape connectivity (modeled as species dispersal rate $d$) and the overall speed of adaptive evolution (modeled as heritability $h^2$).
@@
 
-*Discussion regarding empirical data*
-By implementing the trait changes of the four different species (Lm, Rn, Sp and Wc) in the HMSC model we were able to infer the relative impact of the trait changes on the community dynamics.
-This application shows is advantages over previous methods,which estimate the impact of trait changes (thus evolution), as it is not depended on the data structure and simplicity [@Hairston2005] [@Ellner2011], can be implemented on multiple species[@Lasky2020] and does not require additional common garden experiments [@Garzon2019].
-Therefore this allows for a straight forward implementation of HMSC on a variety of data structures as lone as a clear response variable (e.g.
-species density) and ecological (e.g.
-environmental conditions) as well as evolutionary (e.g.
-heritable trait change) are measured.
-An additional advantage of HMSC, and JSDMs in general, is the ability to use the empirical data to predict a species response to changes in the environmental variables, even beyond the measured values [@Tikhonov2020] [@Ovaskainen2015].
-With the implementation this approach can be combined with the fixed effects to answer further questions, such as: How does trait change impact the species density under changing environmental conditions?
-And how does the impact of trait changes depend on specific environmental conditions?
+The HMSC model was successfully applied to experimental aquatic plants, to estimate the importance of trait change (frond size), light levels, and nutrient composition for abundances of four species (Lm, Rn, Sp and Wc) over 11 weeks.
+This model currently does not take full advantage of the reciprocal transplant data (as implemented in the method of @Jewell2023) and thus doesn't separate the plastic elements of trait change (we also use species densities as our response, not the community average trait value).
+However, the HMSC model doesn't require derivation of new categories for effect sizes (this is the case for @Pantel2015, @Hattich2022, and @Jewell2023), nor must data be modelled as normal.
+As we show in Appendix S2 (also in @Ovaskainen2020), changes in population growth rate can also be fit as response variables, which may be more useful for mechanistic inference in some analyses.
+We also show the additional advantage of HMSC, which is the ability to predict species responses to environmental gradients and trait change.
+It should therefore be possible to use models such as HMSC to study how trait change impacts species densities under changing environmental conditions, or how the impact of trait change might depend on specific environmental conditions.
+These interactions are possible to consider in the Bayesian HMSC model.
 
-At the same time a correct interpretation of such predictions as well es the estimated posterior beta distribution require a thorough understand of the experimental system.
-For example we often observed a negative correlation between trait change and species density in our simulations and also in species Rn of the empirical data, yet they derived from different mechanisms.
-In our simulations trait changes were greater when a species trait was further away from the optimal trait value, this greater distance resulted also in lower fitness and thus lower population size, while a species with higher fitness and thus higher population size had smaller trait changes.
-Therefore, HMSC correlates a high trait changes with a low population size and thus estimates a negative beta effect.
-Such effects can be expected in empirical data of i.e.
-evolutionary rescue experiments, where a species has to adapt to a deleterious environment before going extinct [@Gomulkiewicz1995][@Bell2009].
-In the empirical data this negative correlation was the result of the species Rn investing into a larger frond size (Fig.
-SX), this likely lead to a lower investment into reproduction, while still maintaining a similar total frond area on the water surface to capture the reduced light.
+As observed in the simulation results, correctly interpreting the estimated posterior effect sizes of evolution requires a thorough understanding of the experimental system.
+For example we often observed a negative effect of trait change for species density in model simulations, and this was also observed in the effect of trait change for intraspecific abundances in species.
+However these results derived from different mechanisms.
+In the simulations, trait change was greater when a species trait was further away from the optimal trait value ($E_k - x_{ikt}$), where they also had lower growth rates (the measure of fitness in our models) and thus lower population size.
+Species with lower trait distance ($E_k - x_{ikt}$) and thus higher population size had smaller trait changes.
+For this reason, HMSC estimates a negative effect of trait change $|\Delta x|$ for population size ($N_{ikt}$).
+Such effects should also be observed in studies of evolutionary rescue, where a species adapts to an initially lethal environment and avoids extinction via a gradually increasing growth rate ([@Gomulkiewicz1995]; demonstrated in Appendix S3).
+In the empirical data this negative effect size estimate resulted as the species Rn invested early in the experiment in a larger frond size (see Appendix S4).
+This may have been a tradeoff with a lower investment into reproduction, while still maintaining a similar total frond area on the water surface to capture the reduced light.
 
-*Takehome points and conclusions*
 There has been a strong general interest in estimating effect sizes of evolution for ecological processes [@Hairston2005]; [@Ellner2011]; [@Pantel2015]; [@Barbour2022]), but in many empirical studies these estimates focus on one or a few species where the impacts of their evolution is strong enough to play a clear and detectable role in the corresponding ecological response variable of interest (e.g.
@@
 We present here distributions of effect sizes of evolution for species abundances (i.e.
-Figure 1c,f,i \@ref(fig:fig1); Figure 2c \@ref(fig:fig2); Figure 3c \@ref(fig:fig3)) in simulated and experimental aquatic plant communities, and expect that these effect size coefficients can be useful for better understanding the relative importance of evolution as a driver of community composition.
+Figure \@ref(fig:fig1) c,f,i; Figure \@ref(fig:fig2) c; Figure \@ref(fig:fig3) a,b) in simulated and experimental aquatic plant communities, and expect that these effect size coefficients can be useful for better understanding the relative importance of evolution as a driver of community composition.
 
@@
 [@Gibert2022]).
-It will also be necessary to improve knowledge of the distributions of eco-evolutionary effects and how these are influenced by causal processes that change the speed of evolution or the strength of species interactions.
+It will also be necessary to improve knowledge of the distributions of eco-evolutionary effects and how these are influenced by causal processes that change the speed of evolution or the strength of species interactions ([@Fronhofer2023]).
 Although these interacting eco-evolutionary dynamic processes represent an increased degree of complexity for data analysis and interpretation, these mechanistic processes can be translated into simpler statistical models to estimate ecological and evolutionary drivers of community composition.
@@
 :::
-
-# <a name="appendix">Figures & Tables</a>
- 7 7 7 7 7 7 7 7 7 7 7 7 7 7 7 7 7 7 7 7 7 7 7 7 7 7 7 7 7 7 7 7 7 7 7

### 2024-11-30 manuscript/manuscript_v13.tex
Updated mscript, all appendices, all figures.

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{setspace}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhead[L]{Author Pantel and Hermann}
\fancyhead[R]{Detection of ecoevo dynamics}
\usepackage{lineno}
\linenumbers
\usepackage{float}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Detection of eco-evolutionary dynamics in metacommunities using Joint Species Distribution Models},
  pdfauthor={Jelena H. Pantel; Ruben J. Hermann},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Detection of eco-evolutionary dynamics in metacommunities using Joint Species Distribution Models}
\author{Jelena H. Pantel\footnote{Laboratoire Chrono-environnement,UMR 6249 CNRS-UFC, 16 Route de Gray, 25030 Besançon cedex, France, \href{mailto:jelena.pantel@univ-fcomte.fr}{\nolinkurl{jelena.pantel@univ-fcomte.fr}}} \and Ruben J. Hermann\footnote{University of Duisburg-Essen, Universitätsstraße 5, 45141 Essen, Germany, \href{mailto:ruben.hermann@uni-due.de}{\nolinkurl{ruben.hermann@uni-due.de}}}}
\date{30 November, 2024}

\begin{document}
\maketitle

\setstretch{2}
\newpage

\hypertarget{title}{%
\section{Title}\label{title}}

Detection of eco-evolutionary dynamics in metacommunities using Joint Species Distribution Models

\hypertarget{abstract}{%
\section{Abstract}\label{abstract}}

Biodiversity at the metacommunity scale is typically influenced by a number of environmental, spatial, biotic, and stochastic factors. At the same time, these factors impact the evolution of individual species, as sites present different local selection pressures and connectivity can impact gene flow and genetic drift. Identifying the relative impacts of environmental, spatial, biotic, and other drivers on community composition across spatial and temporal scales has been greatly facilitated by joint species distribution models, but these models have yet to consider the impact of microevolution on community composition. We used Heirarchical Models of Species Communities (HMSC) to analyze simulated data of an populations and communities, including in a large evolving metacommunity model, to establish whether HMSC can sufficiently quantify the contribution of phenotypic evolution for metacommunity composition. The models successfully partitioned variance contributed by environmental, spatial, and evolving phenotypic drivers, and also estimated site- and year-specific covariance. We also applied the HMSC with trait evolution model to an existing dataset studying trait change and community dynamics in experimental aquatic plant system. The study of eco-evolutionary dynamics may require data that reflects numerous complex, interacting processes and it is necessary to have flexible, generalized statistical models to analyze this data. JSDM models such as HMSC present one promising path for analysis of eco-evolutionary dynamics in multi-species communities.

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Analysis of biodiversity at large spatial scales can be complex, because numerous environmental properties, spatial structure and connectivity, interactions between species, as well as stochastic processes such as drift and priority effects operate and determine observed patterns. Statistical models in ecology have made substantial progress in reflecting these numerous ecological processes at multiple scales. There are now a diversity of joint species distribution models {[}JSDMs; Pollock et al. (\protect\hyperlink{ref-Pollock2014}{2014}){]} (\protect\hyperlink{ref-Pichler2021}{Pichler and Hartig 2021}) and multi-species occupancy models (\protect\hyperlink{ref-Devarajan2020}{Devarajan et al. 2020}) (\protect\hyperlink{ref-Poggiato2021}{Poggiato et al. 2021}), which can incorporate these diverse components that structure variance in biological data. Many of these models have implementations in R packages (e.g.~spOccupancy, (\protect\hyperlink{ref-Doser2022}{Doser et al. 2022})), and are used to measure the relative importance of different ecological drivers in e.g.~communities of fungi, birds, mammals, phytoplankton, fish, and other taxa, from local to continental scales (e.g. (\protect\hyperlink{ref-Abrego2020}{Abrego et al. 2020}) (\protect\hyperlink{ref-Antao2022}{Antão et al. 2022}) (\protect\hyperlink{ref-Keppeler2022}{Keppeler et al. 2022})).

Despite the variety of components implemented in these models, there are some additional drivers of community composition which are not currently reflected. The dynamics of populations and communities, and resulting biodiversity patterns, are also influenced by the genotypic and phenotypic properties of organisms. Although traits can be considered in some JSDMs, they typically require fixed values for species' traits (e.g. (\protect\hyperlink{ref-Ovaskainen2017HMSC}{Ovaskainen et al. 2017}) (\protect\hyperlink{ref-Tikhonov2020}{Tikhonov et al. 2020}); but see Abrego et al. (\protect\hyperlink{ref-Abrego2024}{2024})). However, ecological dynamics can impact selection on genes and phenotypes, potentially resulting in interactions or feedback loops between ecological and evolutionary processes (\protect\hyperlink{ref-Lion2018}{Lion 2018}) (\protect\hyperlink{ref-Barbour2022}{Barbour et al. 2022}). Although an increasing number of theoretical and empirical studies have evaluated the impact of evolutionary dynamics for community assembly and metacommunity dynamics (\protect\hyperlink{ref-Loeuille2008}{Loeuille and Leibold 2008}) (\protect\hyperlink{ref-Osmond2013}{Osmond and Mazancourt 2013}) (\protect\hyperlink{ref-Pantel2015}{Pantel et al. 2015}) (\protect\hyperlink{ref-Toju2017}{Toju et al. 2017}), analytical methods to consider this additional complexity lag behind the scope of data collected in eco-evolutionary studies. Some methods exist that ultimately apply ANOVA to partition variance in traits or growth rates across contributions of ecological and evolutionary components ((\protect\hyperlink{ref-Hairston2005}{Hairston Jr et al. 2005}) (\protect\hyperlink{ref-Govaert2016}{Govaert et al. 2016}); methods for discrete and continuous analyses are described in Ellner et al. (\protect\hyperlink{ref-Ellner2011}{2011})). These methods are more frequently cited for their potential rather than for their actual application to empirical data (but see (\protect\hyperlink{ref-terHorst2014}{terHorst et al. 2014}) (\protect\hyperlink{ref-Hiltunen2014}{Hiltunen and Becks 2014})), because this approach is ultimately limited by the same requirements that can limit application of ANOVA - data should follow a normal distribution, the processes that structure the target response variable must be linear and additive, and complicated structuring mechanisms must be reduced into one or a few target categories labeled as `ecology' or `evolution'. Flexible models, which can better reflect the mechanisms and structure of the diverse kinds of data that most studies of eco-evolutionary dynamics will collect, are needed for rigorous inference about eco-evolutionary processes (\protect\hyperlink{ref-Pantel2023}{Pantel and Becks 2023}).

Two existing statistical models are potentially well suited for analysis of eco-evolutionary data at the microevolutionary and metacommunity scale but also have some considerations before their application to complex community data. Lasky et al.~(2020) (\protect\hyperlink{ref-Lasky2020}{Lasky et al. 2020}) developed an integrated reaction norm model, linking genetic, phenotypic, and demographic processes, and Benito Garzón et al.~(2019) (\protect\hyperlink{ref-Garzon2019}{Benito Garzón et al. 2019}) presented a species distribution model with local adaptation and phenotypic plasticity (\(\Delta\) SDMs). The model of Lasky et al.~is an excellent candidate for spatially complex population dynamics, and awaits an extension to the multispecies level. The \(\Delta\) SDM models described by Benito Garzón et al.~truly disentangle the role that plastic and evolutionary trait divergence can play in species distributions, but they also require common garden experimental data across sites to estimate these effects. Observational data collected in metacommunity surveys may only have survey data across sites or time points, as large-scale experimental measures of reaction norms required by the \(\Delta\) SDM model may be difficult to obtain.

One particular metacommunity statistical model, the Hierarchical Modelling of Species Communities (HMSC) framework (\protect\hyperlink{ref-Ovaskainen2017HMSC}{Ovaskainen et al. 2017}) (\protect\hyperlink{ref-Tikhonov2020}{Tikhonov et al. 2020}), explicitly considers a multitude of ecological processes that can structure community composition across time and space (\protect\hyperlink{ref-Leibold2022}{Leibold et al. 2022}) and is potentially well suited to consider analysis of metacommunity eco-evolutionary processes. This model infers the role of environmental filtering via variation and covariation in how species respond to their environment, considers the impacts of (fixed) trait values and phylogenetic relationships, and uses latent variables to account for the numerous unobserved environmental and spatial features that are difficult to exhaustively measure in community surveys. Previous studies have demonstrated that this statistical model can reflect the diverse array of community assembly and metacommunity processes that structure species abundances (i.e.~species sorting, environmental filtering, priority effects; (\protect\hyperlink{ref-Little2018}{Little and Altermatt 2018}) (\protect\hyperlink{ref-Ovaskainen2020}{Ovaskainen and Abrego 2020}) (\protect\hyperlink{ref-Leibold2022}{Leibold et al. 2022})). The associated R package has been applied to study environmental and abiotic drivers of diverse assemblages of organisms (e.g. (\protect\hyperlink{ref-Sandal2022}{Sandal et al. 2022}) (\protect\hyperlink{ref-Weigel2023}{Weigel et al. 2023})). HMSC is potentially useful for studying eco-evolutionary drivers of community structure, but the model has not previously been applied to include heritable changes in trait values over time.

In this study, we use simulations of population dynamics, environmental drivers, species interactions, and trait evolution in populations, communities, at the local and metacommunity scale, to study eco-evolutionary dynamics. After generating simulated time series of population sizes and mean trait values, we then use HMSC to analyze the resulting data, to determine whether HMSC can successfully estimate the impacts of trait evolution for community composition, and to generate relative contributions of fixed environmental and trait evolution drivers as well as random effects of spatial and temporal covariance among species. We asked three main questions: (i) can linear statistical models (like HMSC) accurately capture the impact of non-linear population, community, and evolutionary dynamic processes?, (ii) can HMSC be used to evaluate hypothesized drivers of the relative importance of evolution for community composition at the metacommunity scale?, and (iii) can we better understand the role of evolution for community dynamics in experimental systems using HMSC?

\hypertarget{methods}{%
\section{Methods}\label{methods}}

\hypertarget{linear-statistical-models-for-one--and-two-species-models-of-growth-competition-and-trait-evolution}{%
\subsection{Linear statistical models for one- and two-species models of growth, competition, and trait evolution}\label{linear-statistical-models-for-one--and-two-species-models-of-growth-competition-and-trait-evolution}}

We evaluated whether linear regression models can appropriately capture the dynamics that drive species abundances in relatively simple populations and communities. We fit data generated from simulation models with increasing complexity (Table \ref{tab:tab1}). to Bayesian linear regression models. These simulation models were (A) a single-species model of discrete-time logistic population growth, (B) expanded to include a randomly fluctuating environmental property \(E\) that impacts the species' population growth rate, (C) a two-species model of logistic growth and competition that was then (D) expanded to consider a randomly fluctuating environmental property, (E) and further expanded to consider evolution in a trait \(x\) that is selected by the environmental property, and finally (F) the model of two-species growth, competition, environmental change, and evolution in a spatially structured (10-patch) habitat.

\begin{table}

\caption{\label{tab:tab1}Models for population growth}
\centering
\begin{tabular}[t]{lll}
\toprule
  & model & formula\\
\midrule
A & 1 species, logistic & $N_{i,t+1} = \frac{r_i N_{i,t}}{1+\alpha_{ii} N_{i,t}}$\\
B & 1 species, logistic, env & A, $r_i = \hat{W}e^{-(E_t - x_{i,t})^2}$\\
C & 2 species, logistic, competition & $N_{i,t+1} = \frac{r_i N_{i,t}}{1+\alpha_{ii} N_{i,t} + \alpha_{ij} N_{j,t}}$\\
D & 2 species, logistic, competition, env & C, $r_i = \hat{W}e^{-(E_t - x_{i,t})^2}$\\
E & 2 species, logistic, competition, env, trait evolution & C, $r_i = \hat{W}e^\frac{-[(\frac{w+(1-h^2)P}{P+w})(E_t - x_{i,t})]^2}{2 (P+w)}$\\
\addlinespace
F & 2 species, logistic, competition, env, trait evolution, spatial structure & E, $N_{i,k,t+1} = N_{i,k,t} + \sigma_k + \sigma_{kl}$\\
\bottomrule
\multicolumn{3}{l}{\rule{0pt}{1em}\textit{Note: } $i,j=$ species, $t=$ time, $k,l=$ site}\\
\multicolumn{3}{l}{\rule{0pt}{1em}\textsuperscript{1} See Appendix S1 for parameter values used in simulation}\\
\end{tabular}
\end{table}

To fit linear regression models to the simulated population size data, we began with models using population size (model \emph{A}) and environment (including the second-order term, to better reflect species optimum values; model \emph{B}) as fixed effects (\(N_t \sim N_{t+1} + E_t + E_t^2\)), and progressed to a mixed model using environment and trait evolution as fixed effects (\(N_t \sim E_t + E_t^2 + |\Delta x|\); for simulation model \emph{E}) and time (models \emph{C}, \emph{D}, \emph{E}, and \emph{F}) and spatial distance (model \emph{F}) as random effects. For consistency, all linear regression models were implemented as hierarchical Bayesian models using HMSC (Hierarchical Modelling of Species Communities (\protect\hyperlink{ref-Ovaskainen2017HMSC}{Ovaskainen et al. 2017}) (\protect\hyperlink{ref-Tikhonov2020}{Tikhonov et al. 2020})). HMSC uses a latent variable approach as an alternative to directly modelling species interaction coefficients, estimating a reduced number of linear combinations of species abundances that best predict future population size values. For linear models fit to data generated in simulations \emph{C}, \emph{D}, \emph{E}, and \emph{F} above, the time random effect was implemented to capture species interactions (as part of the temporal species association matrix estimated by the model), and the spatial random effect was implemented in model \emph{F} to capture spatial covariance in abundances as a function of distance between sites. Full methodological details are given in Appendix S1. All simulations were run in R (R Core Team (\protect\hyperlink{ref-R2024}{2024})) and HMSC models were implemented using the \texttt{Hmsc} package (Tikhonov et al. (\protect\hyperlink{ref-Hmsc2022}{2022})).

\hypertarget{evolution-in-metacommunities-and-hmsc}{%
\subsection{Evolution in metacommunities and HMSC}\label{evolution-in-metacommunities-and-hmsc}}

\hypertarget{simulation-model}{%
\subsubsection{Simulation model}\label{simulation-model}}

We also simulated growth and competition dynamics for a multi-species assemblage in a patchy landscape, with site variation in one environmental property, and with evolution in a trait that is selected on by the environment. Population growth for species in the metacommunity simulation followed a Leslie-Gower model (a discrete-time version of a Lotka-Volterra model (\protect\hyperlink{ref-Beverton1957}{Beverton and Holt 1957}) (\protect\hyperlink{ref-Leslie1958}{Leslie and Gower 1958})). We considered the impact of trait evolution for growth using a discrete time quantitative genetic model of evolutionary rescue (\protect\hyperlink{ref-Gomulkiewicz1995}{Gomulkiewicz and Holt 1995}). The model for population size was:

\[ N_{i,t+1} =\frac{\hat{W}e^ \frac{-[(\frac{w+(1-h^2)P} {P+w})(E-x_{i,t})]^2}{2(P+w)}N_{i,t}} {1 + \alpha_{ii}N_{i,t} + \sum_{j \neq i}^{S} \alpha_{ij}N_{j,t}} \] where \(N_i,t\) is the population size of species \emph{i} at time \emph{t}, \(\hat{W}\) is calculated as \(\hat{W}=W_{max}\sqrt(\frac{w}{P+w})\), \(W_{max}\) is the species' maximum per-capita growth rate, \emph{w} is the width of the Gaussian fitness function (which determines the strength of selection), \emph{P} is the width of the distribution of the phenotype \emph{x}, \(h^2\) is the heritability of the trait \emph{x}, \emph{E} is the local environmental optimum trait value (\(x_{opt}\)), \(x_{i,t}\) is the trait value of species \emph{i} at time \emph{t}, \(\alpha_{ii}\) is the intraspecific competition coefficient (the per capita impact of species \emph{i} on itself) and \(\alpha_{ij}\) is the interspecific competition coefficient. Populations have a critical density \(N_c\), below which the population is subject to extinction due to demographic stochasticity at a probability of \emph{p} (\protect\hyperlink{ref-Gomulkiewicz1995}{Gomulkiewicz and Holt 1995}).

This model was used in Pantel \& Becks (\protect\hyperlink{ref-Pantel2023}{Pantel and Becks 2023}) to evaluate the consequences of adaptive evolution for coexistence in a three-species system. To expand this model to a metacommunity, we consider the evolution of multiple species (\emph{S} = 15) in a landscape of patches (\emph{k} = 50). The patches have values of an environmental property \(E \sim U(0,1)\) that determines the local optimum phenotype \emph{E} = \(x_{opt}\), i.e.~where species experience their maximum absolute fitness \(W_{max}\), and patches also have spatial locations \emph{X} and \emph{Y} (both drawn from \(U(0, 1)\)). The patches thus have a connectivity matrix \textbf{D} (here given by their Euclidean distance), as well as a connectivity matrix \textbf{C} that is a Gaussian function of \textbf{D} and species dispersal rate \(d_i\).

We used the evolving metacommunity model to test whether landscape connectivity (the overall dispersal rate \(d\)) and the speed of evolution (varying \(h^2\)) influenced the relative importance of trait evolution (\(|\Delta x|\)) for metacommunity composition \(N_{i,t,k}\) (evaluated as \(V_{|\Delta x|}\), the proportion of variation in species abundances explained by \(|\Delta x|\) as estimated by the HMSC model fit to the simulated data across \(d\) and \(h^2\) levels). We hypothesized that the overall importance of trait evolution (\(V_{|\Delta x|}\)) would decrease with increasing dispersal rate - at low dispersal rates, locally mal-adapted species are less likely to emigrate to more optimal patches and will experience strong local selection, leading to a large amount of trait evolution. Population sizes for initial, mal-adapted populations will also likely be lower, leading to reduced relative importance of density-dependent intra- and interspecific competition. We also hypothesized that heritability (\(h^2\)) would mediate the relative importance of dispersal rate - at higher \(h^2\), species will evolve quickly and we expected the importance of trait evolution \(V_{|\Delta x|}\) to depend on the amount of time that mal-adapted populations persist. Therefore, simulations with higher \(h^2\) would have decreased importance of evolution \(V_{|\Delta x|}\) for community composition. For the same reason, we hypothesized generally that, for a given dispersal rate \(d\), the importance of evolution \(V_{|\Delta x|}\) would decrease with increasing \(h^2\).

All simulations used the same parameter values (\(W_{max}\) = 2, \emph{P} = 1, \emph{w} = 2, \(N_c\) = 100, \emph{p} = 0.001), species interaction matrix (\(\alpha_{ii} = 0.00125\); \(\alpha_{ij} \sim U(0,0.0015)\)), and initial conditions (initial richness \(s_{0k} \sim Pois(0.75)\); initial population size \(N_{0i} \sim Pois(10)\); initial degree of maladaptation \(\beta_{i0} \sim Gamma(0.75,1)\), initial distance to the local patch's optimum phenotype \(d_{i0}=\sqrt{B_{i0}(w+P)}\), initial trait value \(x_{i0} = E_k - d_{i0}\); Appendix S4). Simulations were run across a range of dispersal values (identical for all species; \emph{d} = 0, \(10^{-9},10^{-8},10^{-7},10^{-6},10^{-5},10^{-4},10^{-3},10^{-2}\), 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 1) and heritability values (identical for all species: \(h^2\) = 0, 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 1) for 200 time steps, producing records of \(N_{ikt}\) and \(x_{ikt}\).

\hypertarget{statistical-model}{%
\subsubsection{Statistical model}\label{statistical-model}}

To determine the relative influence of environmental and spatial properties, intra- and interspecific population dynamics, and evolutionary dynamics for community structure, we applied hierarchical modelling of species communities (HMSC, (\protect\hyperlink{ref-Ovaskainen2017HMSC}{Ovaskainen et al. 2017}) (\protect\hyperlink{ref-Tikhonov2020}{Tikhonov et al. 2020})) to model the log-abundance at each site and year for each species. The log-abundance values were modeled as \(ln(N_{ikt}) \sim N(L_{ikt},\sigma_2)\), where the linear predictors \textbf{L} are the sum of fixed and random effects \textbf{L\textsuperscript{F}} + \textbf{L\textsuperscript{R}} (\protect\hyperlink{ref-Tikhonov2020}{Tikhonov et al. 2020}). The fixed effects were \(E_k\), \(E_k^2\), and the absolute value of the change in each species' trait value from the previous time step to the next \(|\Delta x_{ik,t-1 \to t}|\). Random effects included time (the day of the sample) and space. The spatial latent variables represent a spatial model, where species respond to some latent spatial predictor associated with the \emph{xy} coordinates of the sites (\protect\hyperlink{ref-Tikhonov2020}{Tikhonov et al. 2020}). When there was no dispersal in the system (\(d=0\)), spatial random effects were not included in the model.

\hypertarget{applying-hmsc-on-empirical-data}{%
\subsubsection{Applying HMSC on empirical data}\label{applying-hmsc-on-empirical-data}}

To test the applicability of HMSC on ecological and evolutionary effects in empirical data, we applied HMSC on the experimental data of Jewel \& Bell (2023) (\protect\hyperlink{ref-Jewell2023}{Jewell and Bell 2023}). The authors conducted a community selection experiments in which they grew four floating and short-lived aquatic plants (the angiosperms \emph{Lemna minor} here designated as Lm, \emph{Spirodela polyrhiza} designated as Sp and \emph{Wolffia columbiana} designated as Wc, and the liverwort \emph{Ricciocarpos natans} designated as Rn) under different environmental conditions for twelve weeks. The environmental conditions were defined by three different shade cover levels for the experimental pools, 3\%, 12\% and 50\%, and different nutrient concentrations, with low (dissolved nitrogen (DN) = 200 \(\mu g L^{-1}\), dissolved phosphorous (DP) = 10 \(\mu g L^{-1}\), medium (DN = 800 \(\mu g L^{-1}\), DP = 40 \(\mu g L^{-1}\)), and high (DN = 3200 \(\mu g L^{-1}\), DP = 160 \(\mu g L^{-1}\)) concentrations of nutrients. They combined each shade and nutrient level, resulting in nine different environmental conditions. For the eleven weeks of the experiment, the frond size of ten individuals from each species, as well as population density were measured. To use the data with our HMSC analysis, we calculated the average trait change as \(\Delta x = x_{ijt} - x_{ijt-1}\), where \(x_{ij}\) described the mean trait of species \(i\) in environment \(j\) and week \(t\).

To determine the relative importance of evolution for community composition, we fitted two HMSC models to the data: An `evolution model' in which \(|\Delta x|\) of each species and environmental conditions were used as explanatory variables, and a `no-evolution model', in which only the environmental conditions were included. Species density was log-transformed species before applying it to the model, therefore we could use normal distribution in the model. We also included replicates (\emph{n} = 2) and time as random effects. For each model we ran four MCMC chains to generate 1000 samples of the posterior distribution for all model parameters. To evaluate the model-estimated role of evolution for predicted population size, we used the posterior distributions of the estimated effect size of \(|\Delta x|\) multiplied by the observed \(\Delta x\) of each species, added to the model-estimated intercept.

\hypertarget{results}{%
\section{Results}\label{results}}

\hypertarget{linear-statistical-models-for-one--and-two-species-models-of-growth-competition-and-trait-evolution-1}{%
\subsection{Linear statistical models for one- and two-species models of growth, competition, and trait evolution}\label{linear-statistical-models-for-one--and-two-species-models-of-growth-competition-and-trait-evolution-1}}

Researchers often use linear regression models to study non-linear population and community dynamics. We evaluated the performance of linear models in simple (1-2 species, 1-few sites) cases, to validate that effect size estimates and predicted densities accurately reflected mechanisms before moving onto more complex models in the next sections of our analysis. When only one species was present (models \emph{A} and \emph{B}), the regression coefficients (model \emph{A}: \(N_{t+1} \sim \beta_0 + \beta_1 N_t + \epsilon\); \emph{B} \(N_{t+1} \sim \beta_0 + \beta_1 N_t + \beta_2 E_t + \beta_3 E_t^2 + \epsilon\)) reflected the direction and magnitude of the mechanistic drivers of population size (i.e.~\(\beta_1\) correctly captured the positive relationship between \(N_t\) and \(N_{t+1}\) under our simulation conditions, and \(\beta_2\) and \(\beta_3\) correctly captured the quadratic unimodal relationship between \(E_t\) and \(N_t+1\); Appendix S1.1, S1.2).

When a second species was added (models \emph{C} and \emph{D}), the regression coefficients performed well for predicting density values at all time points, but the direction and magnitude of the regression coefficients did not reflect the intra- and interspecific competition strengths included in the mechanistic model (Appendix S1.3). We instead used a latent variable modelling approach (implemented in HMSC) to consider these species interactions as part of the temporal random effects included in the model. This approached produced effect size estimates that both predicted the population dynamics well (Figure \ref{fig:fig1} a) and captured the conditional effects of environment for species densities (Figure \ref{fig:fig1} b). The variation partition also successfully captured the relative importance of the environmental predictor and the temporal random effects for model \emph{D} (which in this simple case were driven by intra- and interspecific species interactions; Figure \ref{fig:fig1} c; Appendix S1.3, S1.4).

Trait evolution was included as a predictor when it was introduced in model \emph{E}. The resulting linear regression coefficients could predict species dynamics (Figure \ref{fig:fig1} d) and reflected the impact and magnitude of environmental and trait evolution predictors (Figure \ref{fig:fig1} e). The variation partition also showed that trait evolution in species 1 (\(|\Delta x_1|\)) was a strong driver of variation in abundance (blue portion, Figure \ref{fig:fig1} f), primarily for species 1 but also for species 2 as well (Appendix S1.5; Appendix S2).

After adding spatial complexity to the 2-species simulation, we included a spatial random effect to capture species covariances as a function of distance between sites (model \emph{F}). By properly considering species interactions as a temporal random effect and including spatial random effects, the HMSC model could estimate regression coefficients for the impacts of the environment \(E_t\) and trait evolution \(|\Delta x_1|\) and \(|\Delta x_2|\) for both species that fit well to the simulated data points (Figure @ref(fig:fig1 g); Appendix S1.6) and capture conditional effects of environment and trait evolution for both species (Figure \ref{fig:fig1} h). Spatial random effects explained a large proportion of variance in abundance for both species, but trait evolution remained a strong driver for species 1.

\begin{figure}

{\centering \includegraphics[width=1.1\linewidth]{../output/fig1_mod}

}

\caption{Simulation models for 2-species ecological and eco-evolutionary dynamics. Panels a, d, and g show log-population size (y-axis) over time (x-axis) for species $1$ and $2$ when they grow and compete for resources with random environmental fluctuations over time. Species either have fixed (panel a) or evolving trait values (panels d, g) and are present in either 1 (panels a, d; simulation models D and E) or multiple sites (panel g; model F). Graphs in panels b, e, and h show predicted log-population size ($ln(N)$) for species 1 and 2 across gradients of the environment ($E + E^2$) and the change in trait values of species 1 ($|\Delta x_1|$). Graphs in panels c, f, and i show the results of a variation partition after fitting data to the HMSC statistical model. The variation partition plot shows the proportion of variation explained by environment, trait evolution ($|\Delta x_1|$ and $|\Delta x_2|$), and temporal and spatial random effects for $ln(N)$ for species 1 and 2. Error bars and shading in panels a, d, and g represent the 95\% highest probability density interval for posterior predictions.}\label{fig:fig1}
\end{figure}

\hypertarget{evolution-in-metacommunities-and-hmsc-1}{%
\subsection{Evolution in metacommunities and HMSC}\label{evolution-in-metacommunities-and-hmsc-1}}

The simulation model followed population size and trait value dynamics for species in the landscape across 200 time steps (Figure \ref{fig:fig2} a). Resulting biodiversity dynamics depended on the dispersal level and heritability values (Figure \ref{fig:fig2} b). For dispersal, results mirrored some expectations from existing metacommunity models, i.e.~that regional \(\gamma\) diversity decreases with increasing dispersal level \(d\) ((\protect\hyperlink{ref-Mouquet2003}{Mouquet and Loreau 2003}); Figure \ref{fig:fig2} b). The inclusion of trait heritability had a strong effect on community diversity, reflecting that local adaptation can rescue species and contribute to richness at the local and metacommunity scale. Regional species diversity (\(\gamma\)) increased with increasing \(h^2\) levels (Figure \ref{fig:fig2} b).

Analysis using the HMSC model for community time series across \(d\) and \(h^2\) levels indicated that the importance of evolution for community composition (\(V_{|\Delta x|}\)) decreased with increasing dispersal and with increasing heritability (Figure \ref{fig:fig2} c). The importance of spatial random effects increased with dispersal rate, and overall the importance of evolution decreased with increasing \(h^2\). These effects can be better understood by our supplemental analyses of 1-species, 1-site simulation models of evolutionary rescue (Appendix S3).

\begin{figure}

{\centering \includegraphics[width=1.1\linewidth]{../output/fig2_mod}

}

\caption{Importance of trait evolution for change in community composition in metacommunities. (a) We developed a simulation model for population growth, competition, and trait evolution in 15 species that inhabit a patchy landscape. The landscape has 50 sites (colored points in the XY space), with distinct local environments that select for diverse patterns of species diversity (size of points) and average trait values (color of points; results are shown at time $t=0, 10, 200$ for a simulation with $d=10^-3$ and $h^2 = 0.1$). (b) Regional metacommunity $\Gamma$ diversity varied across levels of dispersal and heritability ($h^2$). (c) Variation partition plots after fitting metacommunity time series of species abundances $ln(N_{ikt})$ to HMSC models with environment, trait evolution, and random effects (space and time) as predictors. The top panel is for $h^2 = 0.1$, the middle panel is for $h^2 = 0.3$, and the bottom panel is for $h^2 = 0.9$.}\label{fig:fig2}
\end{figure}

\hypertarget{empirical-data}{%
\subsubsection{Empirical data}\label{empirical-data}}

The results for the HMSC analysis of the empirical data (community dynamics and frond size of four aquatic plant species under different environmental conditions) estimated an effect of trait change for species density. We used the HMSC models to calculate the mean proportion of variance explained (variance partitioning, VP) over all species, to estimate the relative importance of evolutionary (change in the frond size trait) and ecological (light and nutrient treatments) effects on species densities.Almost one fifth (18.6\%) of the variance in species density could be explained by the variance in \(|\Delta x|\) (\ref{fig:fig3}a). Including trait evolution in the statistical model accurately led to reductions for the estimated importance of ecological effects by a total of 14.4\% (from 37\% to 26.6\% for the amount of shading and from 31\% to 27.1\% for the nutrient concentrations) and the reduction of 4.2\% of the random effects (from 18.7\% to 15\% for time and from 13.3\% to 12.8\% for replicates) (\ref{fig:fig3}). While \(|\Delta x|\) of the species Lm, Sp and Wc showed a similarly low impact (3.7\%, 3.4\%, and 3.7\% respectively), \(|\Delta x|\) of species Rn showed a higher impact on species density, 7.8\%, which can be explained by its high \(|\Delta x|\) changes over time (see Appendix S4 data). The 95\% HPDI for main effect sizes showed a negative effect of light shading for species density, while nutrient concentration had a positive estimated effect for density (\ref{fig:fig3}c). Changes in \(|\Delta x|\) impacted densities in distinct ways for intra- and interspecific comparisons. The frond size of species Rn had a negative impact on Rn density, while change in Wc frond size had a positive estimated effect on species Sp and Lm (\ref{fig:fig3}c). We show the estimated effect of trait evolution for species counts in \ref{fig:fig3}d, which reflects potentially strong impacts in particular species, density, and light treatments and weak effects with high uncertainty for many cases.

\begin{figure}

{\centering \includegraphics[width=1.1\linewidth]{../output/fig3}

}

\caption{The estimated effects of species trait changes and environmental treatments for densities of experimental aquatic plants (species Lm, Rn, Sp, and Wc). The variation partition for the HMSC model with (a) and without (b) trait change $|\Delta x|$ is shown, as well as (c) the posterior estimated direction of effect (negative or positive; effect sizes with 95\% HPDI that included 0 are colored white). The results indicated negative impacts of light, positive impacts of nutrients, negative intraspecific effects of $|\Delta x_{Rn}|$ for $log(N_{Rn})$, negative interspecific effects of $|\Delta x_{Lm}|$ for $log(N_{Sp})$, and positive interspecific effects of $|\Delta x_{Wc}|$ for $log(N_{Sp})$ and $log(N_{Lm})$. (d) The posterior effect size estimates were used to estimate model-predicted densities at each time step of the experiment as a function only of $|\Delta x|$ (for each of the four species, averaged across all of the other treatment effects). This captures the expected change in density due to trait evolution in all species.}\label{fig:fig3}
\end{figure}

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

The analysis of community composition often requires statistical models that can consider the combined influences of environmental drivers and spatial connectivity. A more difficult, yet critical, element to incorporate is species interactions. Although the causal role of species interactions is difficult to interpret from statistical models (but see (\protect\hyperlink{ref-Dubart2022}{Dubart et al. 2022}) (\protect\hyperlink{ref-Luo2022}{Luo et al. 2022})) in systems with few species), joint species distribution models (JSDMs) have implemented spatial and temporal random effects that can effectively capture these species interactions using a latent variable approach (but note the resulting species covariance matrices will also reflect other processes besides species interactions; (\protect\hyperlink{ref-Poggiato2021}{Poggiato et al. 2021})). Finally, one element that has not been fully implemented into joint species distribution models is dynamic trait evolution (see (\protect\hyperlink{ref-Garzon2019}{Benito Garzón et al. 2019}) for a model that does consider reaction norms from in-situ reciprocal transplant experiments). We used an existing JSDM, and developed simulation models of evolving populations in increasing complex scenarios to determine whether the signal of trait evolution could be estimated from data and compared to other drivers of community structure. We also applied this statistical model (HMSC; (\protect\hyperlink{ref-Ovaskainen2017HMSC}{Ovaskainen et al. 2017}) (\protect\hyperlink{ref-Tikhonov2020}{Tikhonov et al. 2020})) to an existing empirical dataset (from the experimental study of community eco-evolutionary dynamics in aquatic plants of (\protect\hyperlink{ref-Jewell2023}{Jewell and Bell 2023})) to show how HMSC can be used to model the changes in species composition over time as a function of environmental drivers, trait change, and temporal random effects. Overall, there is no barrier to including dynamic trait change as a fixed effect in a JSDM model, and the resulting effect size coefficient estimates can be used to (i) estimate the relative importance of trait evolution for community composition and (ii) predict the impacts of future trait change in the system.

At the same time, fixed effects for the impacts of trait evolution on population size should be interpreted with a great deal of caution. For example, our simulations reveal instances where trait evolution may occur at the same time as large changes in population size, and this co-occurrence isn't always causal (e.g.~when selection is weak but trait heritability is high, and species are introduced at low initial population size). In this instance, the regression coefficient from the model may be negative and strong. By using simulations of trait evolution and population growth in simple systems (1- and 2-species systems, at single and simplified multi-patch scales; Figure \ref{fig:fig1}; Appendix S1), we can better interpret effect size estimates. However, these precautions are not unique to including trait evolution as a predictor of community composition. For example, previous research has shown that the effect size coefficients from linear and AR models using \(N_{i,t-1}\) to predict \(N_{i,t}\) or \(N_{j,t}\) are not equivalent to the intra- and interspecific interaction coefficients that determine species density dependence and should not automatically be used to interpret the direction and magnitude of species interactions ((\protect\hyperlink{ref-Kloppers2013}{Kloppers and Greeff 2013}) (\protect\hyperlink{ref-Certain2018}{Certain et al. 2018}) (\protect\hyperlink{ref-Muhlbauer2020}{Mühlbauer et al. 2020}) (\protect\hyperlink{ref-Olivenca2021}{Olivença et al. 2021})). Our findings for using HMSC to estimate the impact of trait evolution (\(|\Delta x_i|\)) for species abundances are similar - (i) the relative importance of trait evolution estimated by a variation partition accurately reflects the relative strength of simulated species interactions vs.~magnitude of trait change and (ii) the coefficients greatly increase the predictive ability of the statistical model for simulated data (Appendix S1). HMSC is therefore a useful tool for understanding the role that evolution can play in community dynamics.

This was also observed when using HMSC to evaluate the importance of evolution for metacommunity structure, using simulated data from a model of population growth, competition, dispersal, and trait evolution in a spatially structured habitat (Figure \ref{fig:fig1} a). First of all, the simulations themselves revealed an intriguing role of adaptive evolution for rescuing individual species from extinction (i.e. (\protect\hyperlink{ref-Gomulkiewicz1995}{Gomulkiewicz and Holt 1995})) and as a consequence greatly increasing biodiversity at the local and the landscape scale (i.e.~community rescue, (\protect\hyperlink{ref-Low2015}{Low-Décarie et al. 2015}) (\protect\hyperlink{ref-Bell2019}{Bell et al. 2019}); Figure \ref{fig:fig2} b). The recovery of species via local adaptation had consequent effects for coexistence, as species that were initially mal-adapted could recover in population size and exert strong competitive effects on other species. Because HMSC can accurately estimate the relative importance of trait evolution for community composition (Appendix S3), we were thus able to test hypotheses about the drivers of increased or decreased impacts of trait evolution for community structure. Our simulation models fit to HMSC confirmed our hypotheses that the importance of evolution would decrease with increasing landscape connectivity (modeled as species dispersal rate \(d\)) and the overall speed of adaptive evolution (modeled as heritability \(h^2\)). Previous theoretical models have been shown that these two properties can have a strong influence on resulting community composition ((\protect\hyperlink{ref-Urban2008}{Urban et al. 2008}) (\protect\hyperlink{ref-Loeuille2008}{Loeuille and Leibold 2008}) (\protect\hyperlink{ref-Urban2009}{Urban and De Meester 2009}) (\protect\hyperlink{ref-DeMeester2016}{De Meester et al. 2016})), but our demonstration that the HMSC statistical model can accurately estimate \(R^2\) values for the relative importance of trait evolution can allow hypotheses from theoretical models to be tested using empirical data and compared across different study systems.

The HMSC model was successfully applied to experimental aquatic plants, to estimate the importance of trait change (frond size), light levels, and nutrient composition for abundances of four species (Lm, Rn, Sp and Wc) over 11 weeks. This model currently does not take full advantage of the reciprocal transplant data (as implemented in the method of Jewell and Bell (\protect\hyperlink{ref-Jewell2023}{2023})) and thus doesn't separate the plastic elements of trait change (we also use species densities as our response, not the community average trait value). However, the HMSC model doesn't require derivation of new categories for effect sizes (this is the case for Pantel et al. (\protect\hyperlink{ref-Pantel2015}{2015}), Hattich et al. (\protect\hyperlink{ref-Hattich2022}{2022}), and Jewell and Bell (\protect\hyperlink{ref-Jewell2023}{2023})), nor must data be modelled as normal. As we show in Appendix S2 (also in Ovaskainen and Abrego (\protect\hyperlink{ref-Ovaskainen2020}{2020})), changes in population growth rate can also be fit as response variables, which may be more useful for mechanistic inference in some analyses. We also show the additional advantage of HMSC, which is the ability to predict species responses to environmental gradients and trait change. It should therefore be possible to use models such as HMSC to study how trait change impacts species densities under changing environmental conditions, or how the impact of trait change might depend on specific environmental conditions. These interactions are possible to consider in the Bayesian HMSC model.

As observed in the simulation results, correctly interpreting the estimated posterior effect sizes of evolution requires a thorough understanding of the experimental system. For example we often observed a negative effect of trait change for species density in model simulations, and this was also observed in the effect of trait change for intraspecific abundances in species. However these results derived from different mechanisms. In the simulations, trait change was greater when a species trait was further away from the optimal trait value (\(E_k - x_{ikt}\)), where they also had lower growth rates (the measure of fitness in our models) and thus lower population size. Species with lower trait distance (\(E_k - x_{ikt}\)) and thus higher population size had smaller trait changes. For this reason, HMSC estimates a negative effect of trait change \(|\Delta x|\) for population size (\(N_{ikt}\)). Such effects should also be observed in studies of evolutionary rescue, where a species adapts to an initially lethal environment and avoids extinction via a gradually increasing growth rate ((\protect\hyperlink{ref-Gomulkiewicz1995}{Gomulkiewicz and Holt 1995}); demonstrated in Appendix S3). In the empirical data this negative effect size estimate resulted as the species Rn invested early in the experiment in a larger frond size (see Appendix S4). This may have been a tradeoff with a lower investment into reproduction, while still maintaining a similar total frond area on the water surface to capture the reduced light.

There has been a strong general interest in estimating effect sizes of evolution for ecological processes (\protect\hyperlink{ref-Hairston2005}{Hairston Jr et al. 2005}); (\protect\hyperlink{ref-Ellner2011}{Ellner et al. 2011}); (\protect\hyperlink{ref-Pantel2015}{Pantel et al. 2015}); (\protect\hyperlink{ref-Barbour2022}{Barbour et al. 2022})), but in many empirical studies these estimates focus on one or a few species where the impacts of their evolution is strong enough to play a clear and detectable role in the corresponding ecological response variable of interest (e.g.~population growth rate, community diversity, likelihood of food web collapse). There is no general knowledge for overall distributions of effect sizes for the impact of evolution for species composition in communities, nor are there studies that evaluate drivers of shifts in these distributions of effect sizes in complex communities (see (\protect\hyperlink{ref-Hermann2024}{Hermann et al. 2024}) for estimates in a predator-prey system). We present here distributions of effect sizes of evolution for species abundances (i.e.~Figure \ref{fig:fig1} c,f,i; Figure \ref{fig:fig2} c; Figure \ref{fig:fig3} a,b) in simulated and experimental aquatic plant communities, and expect that these effect size coefficients can be useful for better understanding the relative importance of evolution as a driver of community composition.

An increasing number of studies collect and analyze high-resolution time series of populations and trait values for single (\protect\hyperlink{ref-Rudman2022}{Rudman et al. 2022}) and multiple (\protect\hyperlink{ref-Thomas2018}{Thomas et al. 2018}) (\protect\hyperlink{ref-Gibert2023}{Gibert et al. 2023}) (\protect\hyperlink{ref-Han2023}{Han et al. 2023}) (\protect\hyperlink{ref-Shen2023}{Shen et al. 2023}) species. With increasing access to ecological and evolutionary data comes an increase in the number of interacting processes that should be considered to identify causal mechanisms that structure this data (\protect\hyperlink{ref-Pantel2023}{Pantel and Becks 2023}). HMSC, and joint species distribution models generally, can be expanded to further include dynamic trait change that occurs during the course of community ecological studies. Two critically important additions that will increase the application to field or experimental observation data will be (1) bidirectional causality and eco-evolutionary feedbacks and (2) the relative contribution of plasticity and genotypic evolution for trait change (i.e. (\protect\hyperlink{ref-Gibert2022}{Gibert et al. 2022})). It will also be necessary to improve knowledge of the distributions of eco-evolutionary effects and how these are influenced by causal processes that change the speed of evolution or the strength of species interactions ((\protect\hyperlink{ref-Fronhofer2023}{Fronhofer et al. 2023})). Although these interacting eco-evolutionary dynamic processes represent an increased degree of complexity for data analysis and interpretation, these mechanistic processes can be translated into simpler statistical models to estimate ecological and evolutionary drivers of community composition.

\hypertarget{references}{%
\section{References}\label{references}}

\hypertarget{refs}{}
\begin{CSLReferences}{0}{0}
\leavevmode\vadjust pre{\hypertarget{ref-Abrego2020}{}}%
Abrego, N., B. Crosier, P. Somervuo, N. Ivanova, A. Abrahamyan, A. Abdi, K. Hämäläinen, et al. 2020. \href{https://doi.org/10.1038/s41396-020-0732-1}{{Fungal communities decline with urbanization---more in air than in soil}}. The ISME Journal 14:2806--2815.

\leavevmode\vadjust pre{\hypertarget{ref-Abrego2024}{}}%
Abrego, N., P. Niittynen, J. Kemppinen, and O. Ovaskainen. 2024. Disentangling the role of intraspecific trait variation in community assembly with joint species-trait distribution modelling. Authorea Preprints.

\leavevmode\vadjust pre{\hypertarget{ref-Antao2022}{}}%
Antão, L. H., B. Weigel, G. Strona, M. Hällfors, E. Kaarlejärvi, T. Dallas, Ø. H. Opedal, et al. 2022. \href{https://doi.org/10.1038/s41558-022-01381-x}{Climate change reshuffles northern species within their niches}. Nature Climate Change 12:587--592.

\leavevmode\vadjust pre{\hypertarget{ref-Barbour2022}{}}%
Barbour, M. A., D. J. Kliebenstein, and J. Bascompte. 2022. \href{https://doi.org/10.1126/science.abf2232}{A keystone gene underlies the persistence of an experimental food web}. Science 376:70--73.

\leavevmode\vadjust pre{\hypertarget{ref-Bell2019}{}}%
Bell, G., V. Fugère, R. Barrett, B. Beisner, M. Cristescu, G. Fussmann, J. Shapiro, et al. 2019. \href{https://doi.org/10.1098/rspb.2019.0856}{Trophic structure modulates community rescue following acidification}. Proceedings of the Royal Society B: Biological Sciences 286:20190856.

\leavevmode\vadjust pre{\hypertarget{ref-Garzon2019}{}}%
Benito Garzón, M., T. M. Robson, and A. Hampe. 2019. \href{https://doi.org/10.1111/nph.15716}{\(\Delta\)TraitSDMs: Species distribution models that account for local adaptation and phenotypic plasticity}. New Phytologist 222:1757--1765.

\leavevmode\vadjust pre{\hypertarget{ref-Beverton1957}{}}%
Beverton, R. J., and S. J. Holt. 1957. On the dynamics of exploited fish populations (Vol. 11). Springer Science \& Business Media.

\leavevmode\vadjust pre{\hypertarget{ref-Certain2018}{}}%
Certain, G., F. Barraquand, and A. Gårdmark. 2018. \href{https://doi.org/10.1111/2041-210X.13021}{How do MAR(1) models cope with hidden nonlinearities in ecological dynamics?} Methods in Ecology and Evolution 9:1975--1995.

\leavevmode\vadjust pre{\hypertarget{ref-DeMeester2016}{}}%
De Meester, L., J. Vanoverbeke, L. J. Kilsdonk, and M. C. Urban. 2016. Evolving perspectives on monopolization and priority effects. Trends in Ecology \& Evolution 31:136--146.

\leavevmode\vadjust pre{\hypertarget{ref-Devarajan2020}{}}%
Devarajan, K., T. L. Morelli, and S. Tenan. 2020. \href{https://doi.org/10.1111/ecog.04957}{Multi-species occupancy models: Review, roadmap, and recommendations}. Ecography 43:1612--1624.

\leavevmode\vadjust pre{\hypertarget{ref-Doser2022}{}}%
Doser, J. W., A. O. Finley, M. Kéry, and E. F. Zipkin. 2022. \href{https://doi.org/10.1111/2041-210X.13897}{spOccupancy: An r package for single-species, multi-species, and integrated spatial occupancy models}. Methods in Ecology and Evolution 13:1670--1678.

\leavevmode\vadjust pre{\hypertarget{ref-Dubart2022}{}}%
Dubart, M., J.-P. Pointier, P. Jarne, and P. David. 2022. \href{https://doi.org/10.1111/oik.09157}{Niche filtering, competition and species turnover in a metacommunity of freshwater molluscs}. Oikos 2022:e09157.

\leavevmode\vadjust pre{\hypertarget{ref-Ellner2011}{}}%
Ellner, S. P., M. A. Geber, and N. G. Hairston Jr. 2011. \href{https://doi.org/10.1111/j.1461-0248.2011.01616.x}{Does rapid evolution matter? Measuring the rate of contemporary evolution and its impacts on ecological dynamics}. Ecology Letters 14:603--614.

\leavevmode\vadjust pre{\hypertarget{ref-Fronhofer2023}{}}%
Fronhofer, E. A., D. Corenblit, J. N. Deshpande, L. Govaert, P. Huneman, F. Viard, P. Jarne, et al. 2023. \href{https://doi.org/10.1111/ele.14222}{Eco-evolution from deep time to contemporary dynamics: The role of timescales and rate modulators}. Ecology Letters 26:S91--S108.

\leavevmode\vadjust pre{\hypertarget{ref-Gibert2022}{}}%
Gibert, J. P., Z.-Y. Han, D. J. Wieczynski, S. Votzke, and A. Yammine. 2022. \href{https://doi.org/10.1111/1365-2435.14070}{Feedbacks between size and density determine rapid eco-phenotypic dynamics}. Functional Ecology 36:1668--1680.

\leavevmode\vadjust pre{\hypertarget{ref-Gibert2023}{}}%
Gibert, J. P., D. J. Wieczynski, Z.-Y. Han, and A. Yammine. 2023. \href{https://doi.org/10.1002/ece3.9685}{Rapid eco-phenotypic feedback and the temperature response of biomass dynamics}. Ecology and Evolution 13:e9685.

\leavevmode\vadjust pre{\hypertarget{ref-Gomulkiewicz1995}{}}%
Gomulkiewicz, R., and R. D. Holt. 1995. \href{http://www.jstor.org/stable/2410305}{When does evolution by natural selection prevent extinction?} Evolution 49:201--207.

\leavevmode\vadjust pre{\hypertarget{ref-Govaert2016}{}}%
Govaert, L., J. H. Pantel, and L. De Meester. 2016. \href{https://doi.org/10.1111/ele.12632}{Eco-evolutionary partitioning metrics: Assessing the importance of ecological and evolutionary contributions to population and community change}. Ecology Letters 19:839--853.

\leavevmode\vadjust pre{\hypertarget{ref-Hairston2005}{}}%
Hairston Jr, N. G., S. P. Ellner, M. A. Geber, T. Yoshida, and J. A. Fox. 2005. \href{https://doi.org/10.1111/j.1461-0248.2005.00812.x}{Rapid evolution and the convergence of ecological and evolutionary time}. Ecology Letters 8:1114--1127.

\leavevmode\vadjust pre{\hypertarget{ref-Han2023}{}}%
Han, Z.-Y., D. J. Wieczynski, A. Yammine, and J. P. Gibert. 2023. \href{https://doi.org/10.1098/rspb.2022.2263}{Temperature and nutrients drive eco-phenotypic dynamics in a microbial food web}. Proceedings of the Royal Society B: Biological Sciences 290:20222263.

\leavevmode\vadjust pre{\hypertarget{ref-Hattich2022}{}}%
Hattich, G. S. I., L. Listmann, L. Govaert, C. Pansch, T. B. H. Reusch, and B. Matthiessen. 2022. \href{https://doi.org/10.1111/1365-2435.13923}{Experimentally decomposing phytoplankton community change into ecological and evolutionary contributions}. Functional Ecology 36:120--132.

\leavevmode\vadjust pre{\hypertarget{ref-Hermann2024}{}}%
Hermann, R. J., J. H. Pantel, T. Réveillon, and L. Becks. 2024. \href{https://doi.org/10.1093/jeb/voae062}{{Range of trait variation in prey determines evolutionary contributions to predator growth rates}}. Journal of Evolutionary Biology 37:693--703.

\leavevmode\vadjust pre{\hypertarget{ref-Hiltunen2014}{}}%
Hiltunen, T., and L. Becks. 2014. Consumer co-evolution as an important component of the eco-evolutionary feedback. Nature communications 5:5226.

\leavevmode\vadjust pre{\hypertarget{ref-Jewell2023}{}}%
Jewell, M. D., and G. Bell. 2023. \href{https://doi.org/10.1002/ecy.4117}{Eco-evolutionary contributions to community trait change in floating aquatic plants}. Ecology 104:e4117.

\leavevmode\vadjust pre{\hypertarget{ref-Keppeler2022}{}}%
Keppeler, F. W., M. C. Andrade, P. A. A. Trindade, L. M. Sousa, C. C. Arantes, K. O. Winemiller, O. P. Jensen, et al. 2022. \href{https://doi.org/10.1016/j.scitotenv.2022.155951}{Early impacts of the largest amazonian hydropower project on fish communities}. Science of The Total Environment 838:155951.

\leavevmode\vadjust pre{\hypertarget{ref-Kloppers2013}{}}%
Kloppers, P. H., and J. C. Greeff. 2013. \href{https://doi.org/10.1016/j.amc.2013.08.093}{Lotka--volterra model parameter estimation using experiential data}. Applied Mathematics and Computation 224:817--825.

\leavevmode\vadjust pre{\hypertarget{ref-Lasky2020}{}}%
Lasky, J. R., M. B. Hooten, and P. B. Adler. 2020. \href{https://doi.org/10.1098/rspb.2020.2219}{What processes must we understand to forecast regional-scale population dynamics?} Proceedings of the Royal Society B: Biological Sciences 287:20202219.

\leavevmode\vadjust pre{\hypertarget{ref-Leibold2022}{}}%
Leibold, M. A., F. J. Rudolph, F. G. Blanchet, L. De Meester, D. Gravel, F. Hartig, P. Peres-Neto, et al. 2022. \href{https://doi.org/10.1111/oik.08618}{The internal structure of metacommunities}. Oikos 2022.

\leavevmode\vadjust pre{\hypertarget{ref-Leslie1958}{}}%
Leslie, P. H., and J. C. Gower. 1958. \href{http://www.jstor.org/stable/2333181}{The properties of a stochastic model for two competing species}. Biometrika 45:316--330.

\leavevmode\vadjust pre{\hypertarget{ref-Lion2018}{}}%
Lion, S. 2018. \href{https://doi.org/10.1086/694865}{Theoretical approaches in evolutionary ecology: Environmental feedback as a unifying perspective}. The American Naturalist 191:21--44.

\leavevmode\vadjust pre{\hypertarget{ref-Little2018}{}}%
Little, C. J., and F. Altermatt. 2018. \href{https://doi.org/10.1098/rspb.2018.0205}{Do priority effects outweigh environmental filtering in a guild of dominant freshwater macroinvertebrates?} Proceedings of the Royal Society B: Biological Sciences 285:20180205.

\leavevmode\vadjust pre{\hypertarget{ref-Loeuille2008}{}}%
Loeuille, N., and M. A. Leibold. 2008. \href{https://doi.org/10.1086/587745}{Evolution in metacommunities: On the relative importance of species sorting and monopolization in structuring communities.} The American Naturalist 171:788--799.

\leavevmode\vadjust pre{\hypertarget{ref-Low2015}{}}%
Low-Décarie, E., M. Kolber, P. Homme, A. Lofano, A. Dumbrell, A. Gonzalez, and G. Bell. 2015. \href{https://doi.org/10.1073/pnas.1513125112}{Community rescue in experimental metacommunities}. Proceedings of the National Academy of Sciences 112:14307--14312.

\leavevmode\vadjust pre{\hypertarget{ref-Luo2022}{}}%
Luo, M., S. Wang, S. Saavedra, D. Ebert, and F. Altermatt. 2022. \href{https://doi.org/10.1073/pnas.2201503119}{Multispecies coexistence in fragmented landscapes}. Proceedings of the National Academy of Sciences 119:e2201503119.

\leavevmode\vadjust pre{\hypertarget{ref-Mouquet2003}{}}%
Mouquet, N., and M. Loreau. 2003. \href{https://doi.org/10.1086/378857}{Community patterns in source‐sink metacommunities.} The American Naturalist 162:544--557.

\leavevmode\vadjust pre{\hypertarget{ref-Muhlbauer2020}{}}%
Mühlbauer, L. K., M. Schulze, W. S. Harpole, and A. T. Clark. 2020. \href{https://doi.org/10.1002/ece3.6926}{gauseR: Simple methods for fitting lotka-volterra models describing gause's {``struggle for existence''}}. Ecology and Evolution 10:13275--13283.

\leavevmode\vadjust pre{\hypertarget{ref-Olivenca2021}{}}%
Olivença, D. V., J. D. Davis, and E. O. Voit. 2021. \href{https://doi.org/10.1101/2021.10.07.463461}{Comparison between lotka-volterra and multivariate autoregressive models of ecological interaction systems}. bioRxiv.

\leavevmode\vadjust pre{\hypertarget{ref-Osmond2013}{}}%
Osmond, M. M., and C. de Mazancourt. 2013. \href{https://doi.org/10.1098/rstb.2012.0085}{How competition affects evolutionary rescue}. Philosophical Transactions of the Royal Society B: Biological Sciences 368:20120085.

\leavevmode\vadjust pre{\hypertarget{ref-Ovaskainen2020}{}}%
Ovaskainen, O., and N. Abrego. 2020. Joint species distribution modelling: With applications in r. Cambridge University Press.

\leavevmode\vadjust pre{\hypertarget{ref-Ovaskainen2017HMSC}{}}%
Ovaskainen, O., G. Tikhonov, A. Norberg, F. Guillaume Blanchet, L. Duan, D. Dunson, T. Roslin, et al. 2017. \href{https://doi.org/10.1111/ele.12757}{How to make more out of community data? A conceptual framework and its implementation as models and software}. Ecology Letters 20:561--576.

\leavevmode\vadjust pre{\hypertarget{ref-Pantel2023}{}}%
Pantel, J. H., and L. Becks. 2023. \href{https://doi.org/10.1016/j.tree.2023.03.011}{Statistical methods to identify mechanisms in studies of eco-evolutionary dynamics}. Trends in Ecology {\&} Evolution 38:760--772.

\leavevmode\vadjust pre{\hypertarget{ref-Pantel2015}{}}%
Pantel, J. H., C. Duvivier, and L. D. Meester. 2015. \href{https://doi.org/10.1111/ele.12480}{Rapid local adaptation mediates zooplankton community assembly in experimental mesocosms}. Ecology Letters 18:992--1000.

\leavevmode\vadjust pre{\hypertarget{ref-Pichler2021}{}}%
Pichler, M., and F. Hartig. 2021. \href{https://doi.org/10.1111/2041-210X.13687}{A new joint species distribution model for faster and more accurate inference of species associations from big community data}. Methods in Ecology and Evolution 12:2159--2173.

\leavevmode\vadjust pre{\hypertarget{ref-Poggiato2021}{}}%
Poggiato, G., T. Münkemüller, D. Bystrova, J. Arbel, J. S. Clark, and W. Thuiller. 2021. On the interpretations of joint modeling in community ecology. Trends in Ecology \& Evolution 36:391--401.

\leavevmode\vadjust pre{\hypertarget{ref-Pollock2014}{}}%
Pollock, L. J., R. Tingley, W. K. Morris, N. Golding, R. B. O'Hara, K. M. Parris, P. A. Vesk, et al. 2014. \href{https://doi.org/10.1111/2041-210X.12180}{Understanding co-occurrence by modelling species simultaneously with a joint species distribution model (JSDM)}. Methods in Ecology and Evolution 5:397--406.

\leavevmode\vadjust pre{\hypertarget{ref-R2024}{}}%
R Core Team. 2024. \href{https://www.R-project.org/}{R: A language and environment for statistical computing}. R Foundation for Statistical Computing, Vienna, Austria.

\leavevmode\vadjust pre{\hypertarget{ref-Rudman2022}{}}%
Rudman, S. M., S. I. Greenblum, S. Rajpurohit, N. J. Betancourt, J. Hanna, S. Tilk, T. Yokoyama, et al. 2022. \href{https://doi.org/10.1126/science.abj7484}{Direct observation of adaptive tracking on ecological time scales in \textless i\textgreater drosophila\textless/i\textgreater{}}. Science 375:eabj7484.

\leavevmode\vadjust pre{\hypertarget{ref-Sandal2022}{}}%
Sandal, L., V. Grøtan, B.-E. Sæther, R. P. Freckleton, D. G. Noble, and O. Ovaskainen. 2022. \href{https://doi.org/10.1002/ecy.3731}{Effects of density, species interactions, and environmental stochasticity on the dynamics of british bird communities}. Ecology 103:e3731.

\leavevmode\vadjust pre{\hypertarget{ref-Shen2023}{}}%
Shen, C., K. Lemmen, J. Alexander, and F. Pennekamp. 2023. \href{https://doi.org/10.1002/ece3.10502}{Connecting higher-order interactions with ecological stability in experimental aquatic food webs}. Ecology and Evolution 13:e10502.

\leavevmode\vadjust pre{\hypertarget{ref-terHorst2014}{}}%
terHorst, C. P., J. T. Lennon, and J. A. Lau. 2014. \href{https://doi.org/10.1098/rspb.2014.0028}{The relative importance of rapid evolution for plant-microbe interactions depends on ecological context}. Proceedings of the Royal Society B: Biological Sciences 281:20140028.

\leavevmode\vadjust pre{\hypertarget{ref-Thomas2018}{}}%
Thomas, M. K., S. Fontana, M. Reyes, M. Kehoe, and F. Pomati. 2018. \href{https://doi.org/10.1111/ele.12927}{The predictability of a lake phytoplankton community, over time-scales of hours to years}. Ecology Letters 21:619--628.

\leavevmode\vadjust pre{\hypertarget{ref-Tikhonov2020}{}}%
Tikhonov, G., Ø. H. Opedal, N. Abrego, A. Lehikoinen, M. M. J. de Jonge, J. Oksanen, and O. Ovaskainen. 2020. \href{https://doi.org/10.1111/2041-210X.13345}{Joint species distribution modelling with the r-package hmsc}. Methods in Ecology and Evolution 11:442--447.

\leavevmode\vadjust pre{\hypertarget{ref-Hmsc2022}{}}%
Tikhonov, G., O. Ovaskainen, J. Oksanen, M. de Jonge, O. Opedal, and T. Dallas. 2022. \href{https://CRAN.R-project.org/package=Hmsc}{Hmsc: Hierarchical model of species communities}.

\leavevmode\vadjust pre{\hypertarget{ref-Toju2017}{}}%
Toju, H., M. Yamamichi, P. R. Guimaraes Jr, J. M. Olesen, A. Mougi, T. Yoshida, and J. N. Thompson. 2017. Species-rich networks and eco-evolutionary synthesis at the metacommunity level. Nature Ecology \& Evolution 1:0024.

\leavevmode\vadjust pre{\hypertarget{ref-Urban2009}{}}%
Urban, M. C., and L. De Meester. 2009. \href{https://doi.org/10.1098/rspb.2009.1382}{Community monopolization: Local adaptation enhances priority effects in an evolving metacommunity}. Proceedings of the Royal Society B: Biological Sciences 276:4129--4138.

\leavevmode\vadjust pre{\hypertarget{ref-Urban2008}{}}%
Urban, M. C., M. A. Leibold, P. Amarasekare, L. De Meester, R. Gomulkiewicz, M. E. Hochberg, C. A. Klausmeier, et al. 2008. The evolutionary ecology of metacommunities. Trends in ecology \& evolution 23:311--317.

\leavevmode\vadjust pre{\hypertarget{ref-Weigel2023}{}}%
Weigel, B., N. Kotamäki, O. Malve, K. Vuorio, and O. Ovaskainen. 2023. \href{https://doi.org/10.1111/geb.13626}{Macrosystem community change in lake phytoplankton and its implications for diversity and function}. Global Ecology and Biogeography 32:295--309.

\end{CSLReferences}

\end{document}

### 2024-11-30 manuscript/manuscript_v13.Rmd
Final draft manuscript, not knit

@@
 The models successfully partitioned variance contributed by environmental, spatial, and evolving phenotypic drivers, and also estimated site- and year-specific covariance.
-We also applied the HMSC with trait evolution model to an existing dataset studying trait change and community dynamics in experimental aquatic plant system.
+We also applied the HMSC with trait evolution model to an existing dataset studying trait change and community dynamics in an experimental aquatic plant system.
 The study of eco-evolutionary dynamics may require data that reflects numerous complex, interacting processes and it is necessary to have flexible, generalized statistical models to analyze this data.
@@
 We evaluated whether linear regression models can appropriately capture the dynamics that drive species abundances in relatively simple populations and communities.
-We fit data generated from simulation models with increasing complexity (Table \@ref(tab:tab1)).
-to Bayesian linear regression models.
+We fit data generated from simulation models with increasing complexity (Table \@ref(tab:tab1)) to Bayesian linear regression models.
 These simulation models were (A) a single-species model of discrete-time logistic population growth, (B) expanded to include a randomly fluctuating environmental property $E$ that impacts the species' population growth rate, (C) a two-species model of logistic growth and competition that was then (D) expanded to consider a randomly fluctuating environmental property, (E) and further expanded to consider evolution in a trait $x$ that is selected by the environmental property, and finally (F) the model of two-species growth, competition, environmental change, and evolution in a spatially structured (10-patch) habitat.
@@
 Full methodological details are given in Appendix S1.
-All simulations were run in R (@R2024) and HMSC models were implemented using the `Hmsc` package (@Hmsc2022).
+The one-two species simulations were run in R [@R2024] using the `ecoevoR` package [@Pantel2023] and HMSC models were implemented using the `Hmsc` package [@Hmsc2022].
 
@@
 Simulations were run across a range of dispersal values (identical for all species; *d* = 0, $10^{-9},10^{-8},10^{-7},10^{-6},10^{-5},10^{-4},10^{-3},10^{-2}$, 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 1) and heritability values (identical for all species: $h^2$ = 0, 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 1) for 200 time steps, producing records of $N_{ikt}$ and $x_{ikt}$.
+Code for all simulations and model fits is available at [https://github.com/jhpantel/ecoevo-hmsc/tree/aRxiv](https://github.com/jhpantel/ecoevo-hmsc/tree/aRxiv).
 
@@
 
-To determine the relative influence of environmental and spatial properties, intra- and interspecific population dynamics, and evolutionary dynamics for community structure, we applied hierarchical modelling of species communities (HMSC, [@Ovaskainen2017HMSC] [@Tikhonov2020]) to model the log-abundance at each site and year for each species.
+To determine the relative influence of environmental and spatial properties, intra- and interspecific population dynamics, and evolutionary dynamics for community structure, we applied HMSC to model the log-abundance at each site and year for each species.
 The log-abundance values were modeled as $ln(N_{ikt}) \sim N(L_{ikt},\sigma_2)$, where the linear predictors **L** are the sum of fixed and random effects **L^F^** + **L^R^** [@Tikhonov2020].
@@
 
-### Applying HMSC on empirical data
+### Applying HMSC to empirical data
 
-To test the applicability of HMSC on ecological and evolutionary effects in empirical data, we applied HMSC on the experimental data of Jewel & Bell (2023) [@Jewell2023].
-The authors conducted a community selection experiments in which they grew four floating and short-lived aquatic plants (the angiosperms *Lemna minor* here designated as Lm, *Spirodela polyrhiza* designated as Sp and *Wolffia columbiana* designated as Wc, and the liverwort *Ricciocarpos natans* designated as Rn) under different environmental conditions for twelve weeks.
-The environmental conditions were defined by three different shade cover levels for the experimental pools, 3%, 12% and 50%, and different nutrient concentrations, with low (dissolved nitrogen (DN) = 200 $\mu g L^{-1}$, dissolved phosphorous (DP) = 10 $\mu g L^{-1}$, medium (DN = 800 $\mu g L^{-1}$, DP = 40 $\mu g L^{-1}$), and high (DN = 3200 $\mu g L^{-1}$, DP = 160 $\mu g L^{-1}$) concentrations of nutrients.
-They combined each shade and nutrient level, resulting in nine different environmental conditions.
-For the eleven weeks of the experiment, the frond size of ten individuals from each species, as well as population density were measured.
-To use the data with our HMSC analysis, we calculated the average trait change as $\Delta x = x_{ijt} - x_{ijt-1}$, where $x_{ij}$ described the mean trait of species $i$ in environment $j$ and week $t$.
+To test the applicability for HMSC on ecological and evolutionary effects in empirical data, we applied HMSC to the experimental data of Jewel & Bell (2023) [@Jewell2023].
+The authors conducted a community selection experiment in which they grew four floating and short-lived aquatic plants (the angiosperms *Lemna minor* (Lm), *Spirodela polyrhiza* (Sp) and *Wolffia columbiana* (Wc), and the liverwort *Ricciocarpos natans* (Rn)) under different environmental treatments (shade cover and nutrient levels) for twelve weeks.
+The shade cover and nutrient levels for the experimental pools were a factorial combination of 3%, 12% and 50% shade, and low (dissolved nitrogen (DN) = 200 $\mu g L^{-1}$, dissolved phosphorous (DP) = 10 $\mu g L^{-1}$), medium (DN = 800 $\mu g L^{-1}$, DP = 40 $\mu g L^{-1}$), and high (DN = 3200 $\mu g L^{-1}$, DP = 160 $\mu g L^{-1}$) concentrations of nutrients, respectively.
+The frond size of ten individuals from each species and the density of each species were measured weekly for the eleven weeks of the experiment.
+To estimate the importance of trait change in the HMSC model, we calculated the average trait change as $\Delta x = \bar x_{ijt} - \bar x_{ijt-1}$, where $\bar x_{ij}$ described the mean trait of species $i$ in environment $j$ and week $t$.
 
-To determine the relative importance of evolution for community composition, we fitted two HMSC models to the data: An 'evolution model' in which $|\Delta x|$ of each species and environmental conditions were used as explanatory variables, and a 'no-evolution model', in which only the environmental conditions were included.
-Species density was log-transformed species before applying it to the model, therefore we could use normal distribution in the model.
-We also included replicates (*n* = 2) and time as random effects.
+To determine the relative importance of trait change for community composition, we fit two HMSC models to the data: An 'evolution model' in which $|\Delta x|$ of each species and environmental treatment were used as explanatory variables, and a 'no-evolution model', in which only the environmental conditions were included.
+Species density was log-transformed species, and we included replicates (*n* = 2) and time as random effects.
 For each model we ran four MCMC chains to generate 1000 samples of the posterior distribution for all model parameters.
-To evaluate the model-estimated role of evolution for predicted population size, we used the posterior distributions of the estimated effect size of $|\Delta x|$ multiplied by the observed $\Delta x$ of each species, added to the model-estimated intercept.
+To evaluate the model-estimated role of trait change for predicted population size, we used the posterior distributions of the estimated effect size of $|\Delta x|$ multiplied by the observed $\Delta \bar x$ of each species, added to the model-estimated intercept.
 
@@
 Researchers often use linear regression models to study non-linear population and community dynamics.
-We evaluated the performance of linear models in simple (1-2 species, 1-few sites) cases, to validate that effect size estimates and predicted densities accurately reflected mechanisms before moving onto more complex models in the next sections of our analysis.
-When only one species was present (models *A* and *B*), the regression coefficients (model *A*: $N_{t+1} \sim \beta_0 + \beta_1 N_t + \epsilon$; *B* $N_{t+1} \sim \beta_0 + \beta_1 N_t + \beta_2 E_t + \beta_3 E_t^2 + \epsilon$) reflected the direction and magnitude of the mechanistic drivers of population size (i.e.
-$\beta_1$ correctly captured the positive relationship between $N_t$ and $N_{t+1}$ under our simulation conditions, and $\beta_2$ and $\beta_3$ correctly captured the quadratic unimodal relationship between $E_t$ and $N_t+1$; Appendix S1.1, S1.2).
+We evaluated the performance of linear models in simple cases (1-2 species, 1-few sites), to validate that effect size estimates and predicted densities accurately reflected mechanisms before moving onto more complex models in the next sections of our analysis.
+When only one species was present (models *A* and *B*), the regression coefficients (model *A*: $N_{t+1} \sim \beta_0 + \beta_1 N_t + \epsilon$; *B* $N_{t+1} \sim \beta_0 + \beta_1 N_t + \beta_2 E_t + \beta_3 E_t^2 + \epsilon$) reflected the direction and magnitude of the mechanistic drivers of population size.
+$\beta_1$ correctly captured the positive relationship between $N_t$ and $N_{t+1}$ under our simulation conditions, and $\beta_2$ and $\beta_3$ correctly captured the quadratic unimodal relationship between $E_t$ and $N_t+1$ (Appendix S1.1, S1.2).
 
@@
 We instead used a latent variable modelling approach (implemented in HMSC) to consider these species interactions as part of the temporal random effects included in the model.
-This approached produced effect size estimates that both predicted the population dynamics well (Figure \@ref(fig:fig1) a) and captured the conditional effects of environment for species densities (Figure \@ref(fig:fig1) b).
-The variation partition also successfully captured the relative importance of the environmental predictor and the temporal random effects for model *D* (which in this simple case were driven by intra- and interspecific species interactions; Figure \@ref(fig:fig1) c; Appendix S1.3, S1.4).
+This approached produced effect size estimates that both predicted the population dynamics well (Figure \@ref(fig:fig1)a) and captured the conditional effects of environment for species densities (Figure \@ref(fig:fig1)b).
+The variation partition also successfully captured the relative importance of the environmental predictor and the temporal random effects for model *D* (which in this simple case were driven by intra- and interspecific species interactions; Figure \@ref(fig:fig1)c; Appendix S1.3, S1.4).
 
 Trait evolution was included as a predictor when it was introduced in model *E*.
-The resulting linear regression coefficients could predict species dynamics (Figure \@ref(fig:fig1) d) and reflected the impact and magnitude of environmental and trait evolution predictors (Figure \@ref(fig:fig1) e).
-The variation partition also showed that trait evolution in species 1 ($|\Delta x_1|$) was a strong driver of variation in abundance (blue portion, Figure \@ref(fig:fig1) f), primarily for species 1 but also for species 2 as well (Appendix S1.5; Appendix S2).
+The resulting linear regression coefficients could predict species dynamics (Figure \@ref(fig:fig1)d) and reflected the impact and magnitude of environmental and trait evolution predictors (Figure \@ref(fig:fig1)e).
+The variation partition also showed that trait evolution in species 1 ($|\Delta x_1|$) was a strong driver of variation in abundance (blue portion of bar plot, Figure \@ref(fig:fig1)f), primarily for species 1 but also for species 2 as well (Appendix S1.5; Appendix S2).
 
 After adding spatial complexity to the 2-species simulation, we included a spatial random effect to capture species covariances as a function of distance between sites (model *F*).
-By properly considering species interactions as a temporal random effect and including spatial random effects, the HMSC model could estimate regression coefficients for the impacts of the environment $E_t$ and trait evolution $|\Delta x_1|$ and $|\Delta x_2|$ for both species that fit well to the simulated data points (Figure \@ref(fig:fig1 g); Appendix S1.6) and capture conditional effects of environment and trait evolution for both species (Figure \@ref(fig:fig1) h).
-Spatial random effects explained a large proportion of variance in abundance for both species, but trait evolution remained a strong driver for species 1.
+By properly considering species interactions as a temporal random effect and including spatial random effects, the HMSC model could estimate regression coefficients for the impacts of the environment $E_t$ and trait evolution $|\Delta x_1|$ and $|\Delta x_2|$ for both species that fit well to the simulated data points (Figure \@ref(fig:fig1)g); Appendix S1.6) and capture conditional effects of environment and trait evolution for both species (Figure \@ref(fig:fig1)h).
+Spatial random effects explained a large proportion of variation in abundance for both species, but trait evolution remained a strong driver for species 1.
 
@@
 Panels a, d, and g show log-population size (y-axis) over time (x-axis) for species $1$ and $2$ when they grow and compete for resources with random environmental fluctuations over time.
+Error bars give the 95% highest probability density interval (HPDI) for posterior predictions from the HMSC model fit.
 Species either have fixed (panel a) or evolving trait values (panels d, g) and are present in either 1 (panels a, d; simulation models D and E) or multiple sites (panel g; model F).
-Graphs in panels b, e, and h show predicted log-population size ($ln(N)$) for species 1 and 2 across gradients of the environment ($E + E^2$) and the change in trait values of species 1 ($|\\Delta x_1|$).
+Graphs in panels b, e, and h show observed (points)  and model-predicted (black line with blue shading) log-population size ($log(N)$) for species 1 and 2 across gradients of the environment ($E + E^2$) and the change in trait values of species 1 ($|\\Delta x_1|$).
 Graphs in panels c, f, and i show the results of a variation partition after fitting data to the HMSC statistical model.
@@
 
-The simulation model followed population size and trait value dynamics for species in the landscape across 200 time steps (Figure \@ref(fig:fig2) a).
-Resulting biodiversity dynamics depended on the dispersal level and heritability values (Figure \@ref(fig:fig2) b).
+The simulation model followed population size and trait value dynamics for species in the landscape across 200 time steps (Figure \@ref(fig:fig2)a).
+Resulting biodiversity dynamics depended on the dispersal level and heritability values (Figure \@ref(fig:fig2)b).
 For dispersal, results mirrored some expectations from existing metacommunity models, i.e.
-that regional $\gamma$ diversity decreases with increasing dispersal level $d$ ([@Mouquet2003]; Figure \@ref(fig:fig2) b).
+that regional $\gamma$ diversity decreases with increasing dispersal level $d$ ([@Mouquet2003]; Figure \@ref(fig:fig2)b).
 The inclusion of trait heritability had a strong effect on community diversity, reflecting that local adaptation can rescue species and contribute to richness at the local and metacommunity scale.
-Regional species diversity ($\gamma$) increased with increasing $h^2$ levels (Figure \@ref(fig:fig2) b).
+Regional species diversity ($\gamma$) increased with increasing $h^2$ levels (Figure \@ref(fig:fig2)b).
 
-Analysis using the HMSC model for community time series across $d$ and $h^2$ levels indicated that the importance of evolution for community composition ($V_{|\Delta x|}$) decreased with increasing dispersal and with increasing heritability (Figure \@ref(fig:fig2) c).
+Analysis using the HMSC model for community time series across $d$ and $h^2$ levels indicated that the importance of evolution for community composition ($V_{|\Delta x|}$) decreased with increasing dispersal and with increasing heritability (Figure \@ref(fig:fig2)c).
 The importance of spatial random effects increased with dispersal rate, and overall the importance of evolution decreased with increasing $h^2$.
@@
 
-```{r fig2, out.width="110%", include=TRUE, fig.align="center", fig.cap=c("Importance of trait evolution for change in community composition in metacommunities.
+```{r fig2, out.width="110%", include=TRUE, fig.align="center", fig.cap=c("Importance of trait evolution for change in community composition at the metacommunity scale.
 (a) We developed a simulation model for population growth, competition, and trait evolution in 15 species that inhabit a patchy landscape.
-The landscape has 50 sites (colored points in the XY space), with distinct local environments that select for diverse patterns of species diversity (size of points) and average trait values (color of points; results are shown at time $t=0, 10, 200$ for a simulation with $d=10^-3$ and $h^2 = 0.1$).
-(b) Regional metacommunity $\\Gamma$ diversity varied across levels of dispersal and heritability ($h^2$).
-(c) Variation partition plots after fitting metacommunity time series of species abundances $ln(N_{ikt})$ to HMSC models with environment, trait evolution, and random effects (space and time) as predictors.
-The top panel is for $h^2 = 0.1$, the middle panel is for $h^2 = 0.3$, and the bottom panel is for $h^2 = 0.9$."), echo=FALSE}
+The landscape has 50 sites (colored points in the XY space), with distinct local environments that select for different patterns of local species diversity (size of points) and average trait values (color of points; results are shown at time $t=0, 10, 200$ for a simulation with $d=10^{-3}$ and $h^2 = 0.1$).
+(b) Regional metacommunity $\\gamma$ diversity decreased with increasing disperal level and increased with increasing heritability ($h^2$).
+(c) Variation partition plots after fitting metacommunity time series of species abundances $log(N_{ikt})$ to HMSC models with environment, trait evolution, and random effects (space and time) as predictors.
+These variation partition results are averaged across all species in the metacommunity.
+The top panel is for $h^2 = 0.1$, the middle panel is for $h^2 = 0.3$, and the bottom panel is for $h^2 = 0.9$.
+The alluvial plots show the change in relative importance of different predictors for increasing dispersal levels."), echo=FALSE}
 knitr::include_graphics("../output/fig2_mod.pdf",error=FALSE)
@@
 
-The results for the HMSC analysis of the empirical data (community dynamics and frond size of four aquatic plant species under different environmental conditions) estimated an effect of trait change for species density.
-We used the HMSC models to calculate the mean proportion of variance explained (variance partitioning, VP) over all species, to estimate the relative importance of evolutionary (change in the frond size trait) and ecological (light and nutrient treatments) effects on species densities.Almost one fifth (18.6%) of the variance in species density could be explained by the variance in $|\Delta x|$ (\@ref(fig:fig3)a).
-Including trait evolution in the statistical model accurately led to reductions for the estimated importance of ecological effects by a total of 14.4% (from 37% to 26.6% for the amount of shading and from 31% to 27.1% for the nutrient concentrations) and the reduction of 4.2% of the random effects (from 18.7% to 15% for time and from 13.3% to 12.8% for replicates) (\@ref(fig:fig3)).
-While $|\Delta x|$ of the species Lm, Sp and Wc showed a similarly low impact (3.7%, 3.4%, and 3.7% respectively), $|\Delta x|$ of species Rn showed a higher impact on species density, 7.8%, which can be explained by its high $|\Delta x|$ changes over time (see Appendix S4 data).
-The 95% HPDI for main effect sizes showed a negative effect of light shading for species density, while nutrient concentration had a positive estimated effect for density (\@ref(fig:fig3)c).
-Changes in $|\Delta x|$ impacted densities in distinct ways for intra- and interspecific comparisons.
-The frond size of species Rn had a negative impact on Rn density, while change in Wc frond size had a positive estimated effect on species Sp and Lm (\@ref(fig:fig3)c).
-We show the estimated effect of trait evolution for species counts in \@ref(fig:fig3)d, which reflects potentially strong impacts in particular species, density, and light treatments and weak effects with high uncertainty for many cases.
+We fit the empirical data (community dynamics and frond size of four aquatic plant species under different environmental treatments) to an HMSC model, and estimated the effect size of shade cover, nutrient level, and trait change (in all four species) for changes in species densities.
+The variation partition produced by the HMSC model fit shows the estimated relative importance of evolutionary (change in the frond size trait) and ecological (light and nutrient treatments) effects on species densities.
+Almost one fifth (18.6%) of the variation in species density could be explained by the variance in $|\Delta x|$ (Figure \@ref(fig:fig3)a; dLm, dRn, dSp, dWc).
+Including trait change in the statistical model accurately led to reductions for the estimated importance of ecological effects by a total of 14.4% (from 37% to 26.6% for the amount of shading and from 31% to 27.1% for the nutrient concentrations) and the reduction of 4.2% of the random effects (from 18.7% to 15% for time and from 13.3% to 12.8% for replicates) (Figure \@ref(fig:fig3)).
+While $|\Delta x|$ of the species Lm, Sp and Wc showed a similarly low impact (3.7%, 3.4%, and 3.7% respectively), $|\Delta x|$ of species Rn showed a higher impact on species density, 7.8%, which can be explained by its high $|\Delta x|$ changes over time (see Appendix S4 for data).
+The 95% HPDI for main effect sizes showed a negative effect of light shading for species density, while nutrient concentration had a positive estimated effect (Figure \@ref(fig:fig3)c).
+Changes in $|\Delta x|$ impacted densities in distinct ways for intra- and interspecific comparisons (Figure \@ref(fig:fig3)c).
+The change in frond size of species Rn (dRn) had a negative impact on Rn density, while change in Wc frond size (dWc) had a positive estimated effect on species Sp and Lm (Figure \@ref(fig:fig3)c).
+We show the estimated effect of trait change for species counts in Figure \@ref(fig:fig3)d, which reflects potentially strong impacts in some particular species and environmental treatment combinations, and weak effects with high uncertainty for many species and environmental treatments.
 
@@
 The variation partition for the HMSC model with (a) and without (b) trait change $|\\Delta x|$ is shown, as well as (c) the posterior estimated direction of effect (negative or positive; effect sizes with 95% HPDI that included 0 are colored white).
-The results indicated negative impacts of light, positive impacts of nutrients, negative intraspecific effects of $|\\Delta x_{Rn}|$ for $log(N_{Rn})$, negative interspecific effects of $|\\Delta x_{Lm}|$ for $log(N_{Sp})$, and positive interspecific effects of $|\\Delta x_{Wc}|$ for $log(N_{Sp})$ and $log(N_{Lm})$.
-(d) The posterior effect size estimates were used to estimate model-predicted densities at each time step of the experiment as a function only of $|\\Delta x|$ (for each of the four species, averaged across all of the other treatment effects).
+The model estimated negative impacts of light, positive impacts of nutrients, negative intraspecific effects of $|\\Delta x_{Rn}|$ for $log(N_{Rn})$, negative interspecific effects of $|\\Delta x_{Lm}|$ for $log(N_{Sp})$, and positive interspecific effects of $|\\Delta x_{Wc}|$ for $log(N_{Sp})$ and $log(N_{Lm})$.
+(d) The posterior effect size estimates were used to predict species densities at each time step of the experiment.
+Here we show results only for the main effects of $|\\Delta x|$ (for each of the four species, averaged across all of the other treatment effects).
 This captures the expected change in density due to trait evolution in all species."), echo=FALSE}
@@
 A more difficult, yet critical, element to incorporate is species interactions.
-Although the causal role of species interactions is difficult to interpret from statistical models (but see [@Dubart2022] [@Luo2022]) in systems with few species), joint species distribution models (JSDMs) have implemented spatial and temporal random effects that can effectively capture these species interactions using a latent variable approach (but note the resulting species covariance matrices will also reflect other processes besides species interactions; [@Poggiato2021]).
-Finally, one element that has not been fully implemented into joint species distribution models is dynamic trait evolution (see [@Garzon2019] for a model that does consider reaction norms from in-situ reciprocal transplant experiments).
-We used an existing JSDM, and developed simulation models of evolving populations in increasing complex scenarios to determine whether the signal of trait evolution could be estimated from data and compared to other drivers of community structure.
-We also applied this statistical model (HMSC; [@Ovaskainen2017HMSC] [@Tikhonov2020]) to an existing empirical dataset (from the experimental study of community eco-evolutionary dynamics in aquatic plants of [@Jewell2023]) to show how HMSC can be used to model the changes in species composition over time as a function of environmental drivers, trait change, and temporal random effects.
+Although the causal role of species interactions is difficult to interpret from statistical models (but see [@Dubart2022] [@Luo2022]) in systems with few species), joint species distribution models (JSDMs) have implemented spatial and temporal random effects that can effectively capture these species interactions using a latent variable approach (however it is important to note that the resulting species covariance matrices also reflect other processes besides species interactions, including unmeasured environmental effects; [@Poggiato2021]).
+One element that has not been fully implemented into joint species distribution models is dynamic trait evolution (see [@Garzon2019] for a model that does consider reaction norms from in-situ reciprocal transplant experiments).
+We used an existing JSDM, and developed simulation models of evolving populations in increasingly complex scenarios to determine whether the signal of trait evolution could be estimated from data and compared to other drivers of community structure.
+We also applied this statistical model (HMSC; [@Ovaskainen2017HMSC] [@Tikhonov2020]) to an existing empirical dataset (from the experimental study of community eco-evolutionary dynamics in aquatic plants of [@Jewell2023]) to show how HMSC can be used to model and predict the changes in species composition over time as a function of environmental drivers, trait change, and temporal random effects.
 Overall, there is no barrier to including dynamic trait change as a fixed effect in a JSDM model, and the resulting effect size coefficient estimates can be used to (i) estimate the relative importance of trait evolution for community composition and (ii) predict the impacts of future trait change in the system.
 
-At the same time, fixed effects for the impacts of trait evolution on population size should be interpreted with a great deal of caution.
+At the same time, fixed effects for the impacts of trait evolution on population size should be interpreted with caution.
 For example, our simulations reveal instances where trait evolution may occur at the same time as large changes in population size, and this co-occurrence isn't always causal (e.g.
@@
 However, these precautions are not unique to including trait evolution as a predictor of community composition.
-For example, previous research has shown that the effect size coefficients from linear and AR models using $N_{i,t-1}$ to predict $N_{i,t}$ or $N_{j,t}$ are not equivalent to the intra- and interspecific interaction coefficients that determine species density dependence and should not automatically be used to interpret the direction and magnitude of species interactions ([@Kloppers2013] [@Certain2018] [@Muhlbauer2020] [@Olivenca2021]).
+Previous research has shown that the effect size coefficients from linear and AR models using $N_{i,t-1}$ to predict $N_{i,t}$ or $N_{j,t}$ are not equivalent to the intra- and interspecific interaction coefficients that determine species density dependence and should not automatically be used to interpret the direction and magnitude of species interactions ([@Kloppers2013] [@Certain2018] [@Muhlbauer2020] [@Olivenca2021]).
 Our findings for using HMSC to estimate the impact of trait evolution ($|\Delta x_i|$) for species abundances are similar - (i) the relative importance of trait evolution estimated by a variation partition accurately reflects the relative strength of simulated species interactions vs.
@@
 
-This was also observed when using HMSC to evaluate the importance of evolution for metacommunity structure, using simulated data from a model of population growth, competition, dispersal, and trait evolution in a spatially structured habitat (Figure \@ref(fig:fig1) a).
+This was also observed when using HMSC to evaluate the importance of evolution for metacommunity structure, using simulated data from a model of population growth, competition, dispersal, and trait evolution in a spatially structured habitat (Figure \@ref(fig:fig2)a).
 First of all, the simulations themselves revealed an intriguing role of adaptive evolution for rescuing individual species from extinction (i.e.
 [@Gomulkiewicz1995]) and as a consequence greatly increasing biodiversity at the local and the landscape scale (i.e.
-community rescue, [@Low2015] [@Bell2019]; Figure \@ref(fig:fig2) b).
+community rescue, [@Low2015] [@Bell2019]; Figure \@ref(fig:fig2)b).
 The recovery of species via local adaptation had consequent effects for coexistence, as species that were initially mal-adapted could recover in population size and exert strong competitive effects on other species.
@@
 Our simulation models fit to HMSC confirmed our hypotheses that the importance of evolution would decrease with increasing landscape connectivity (modeled as species dispersal rate $d$) and the overall speed of adaptive evolution (modeled as heritability $h^2$).
-Previous theoretical models have been shown that these two properties can have a strong influence on resulting community composition ([@Urban2008] [@Loeuille2008] [@Urban2009] [@DeMeester2016]), but our demonstration that the HMSC statistical model can accurately estimate $R^2$ values for the relative importance of trait evolution can allow hypotheses from theoretical models to be tested using empirical data and compared across different study systems.
+Previous theoretical models have shown that these two properties can have a strong influence on resulting community composition ([@Urban2008] [@Loeuille2008] [@Urban2009] [@DeMeester2016]), but our demonstration that the HMSC statistical model can accurately estimate the proportion of variation explained by trait evolution can allow hypotheses from theoretical models to be tested using empirical data, with effect size estimates (regression coefficients from the model) that can be compared across different study systems.
 
-The HMSC model was successfully applied to experimental aquatic plants, to estimate the importance of trait change (frond size), light levels, and nutrient composition for abundances of four species (Lm, Rn, Sp and Wc) over 11 weeks.
-This model currently does not take full advantage of the reciprocal transplant data (as implemented in the method of @Jewell2023) and thus doesn't separate the plastic elements of trait change (we also use species densities as our response, not the community average trait value).
-However, the HMSC model doesn't require derivation of new categories for effect sizes (this is the case for @Pantel2015, @Hattich2022, and @Jewell2023), nor must data be modelled as normal.
-As we show in Appendix S2 (also in @Ovaskainen2020), changes in population growth rate can also be fit as response variables, which may be more useful for mechanistic inference in some analyses.
-We also show the additional advantage of HMSC, which is the ability to predict species responses to environmental gradients and trait change.
-It should therefore be possible to use models such as HMSC to study how trait change impacts species densities under changing environmental conditions, or how the impact of trait change might depend on specific environmental conditions.
+The HMSC model was also applied to experimental aquatic plants, to estimate the importance of trait change (frond size), light levels, and nutrient composition for abundances of four species (Lm, Rn, Sp and Wc) over 11 weeks.
+This model currently does not take full advantage of the reciprocal transplant data (as implemented in the method of @Jewell2023) and thus doesn't separate the plastic elements of trait change (another key difference in our approach is that we use species densities as our response, not the community average trait value; our results are therefore not meant to be compared directly with those of @Jewell2023).
+However, the HMSC model does not require derivation of new categories for effect sizes (this is the case for @Pantel2015, @Hattich2022, and @Jewell2023), nor must data be modelled as normal.
+The HMSC model also allows for the flexible implementation of random effects according to the study design.
+As we show in Appendix S2 (see also @Ovaskainen2020), changes in population growth rate ($\Delta r$) or community mean trait values ($\bar x$) can also be fit as response variables, which may be more useful for mechanistic inference in some analyses.
+We also show an additional advantage of HMSC, which is the ability to predict species responses to environmental gradients and trait change (i.e.
+Figure \@ref(fig:fig3)c).
+It should therefore be possible to use models such as HMSC to study how evolution and trait change impact species densities under changing environmental conditions, or how the impact of trait change might depend on specific environmental conditions.
 These interactions are possible to consider in the Bayesian HMSC model.
+Researchers are thus free to fit a JSDM or HMSC model to match their study design, which is not the case for existing formulations of eco-evolutionary effect size estimates that use an ANOVA (i.e.
+@Govaert2016).
 
 As observed in the simulation results, correctly interpreting the estimated posterior effect sizes of evolution requires a thorough understanding of the experimental system.
-For example we often observed a negative effect of trait change for species density in model simulations, and this was also observed in the effect of trait change for intraspecific abundances in species.
-However these results derived from different mechanisms.
-In the simulations, trait change was greater when a species trait was further away from the optimal trait value ($E_k - x_{ikt}$), where they also had lower growth rates (the measure of fitness in our models) and thus lower population size.
+For example we often observed a negative effect of trait change for species density in model simulations, and this was also observed in the empirical results.
+However these effects derived from different mechanisms.
+In the simulations, trait change was greater when a species trait was further away from the optimal value ($E_k - x_{ikt}$), where they also had lower growth rates (the measure of fitness in our models) and thus lower population size.
 Species with lower trait distance ($E_k - x_{ikt}$) and thus higher population size had smaller trait changes.
-For this reason, HMSC estimates a negative effect of trait change $|\Delta x|$ for population size ($N_{ikt}$).
+For this reason, HMSC estimates a negative effect of trait change $|\Delta x|$ for population size.
 Such effects should also be observed in studies of evolutionary rescue, where a species adapts to an initially lethal environment and avoids extinction via a gradually increasing growth rate ([@Gomulkiewicz1995]; demonstrated in Appendix S3).
-In the empirical data this negative effect size estimate resulted as the species Rn invested early in the experiment in a larger frond size (see Appendix S4).
+In the empirical data this negative effect size estimate resulted as the species Rn invested early in the experiment in a larger frond size (Appendix S4).
 This may have been a tradeoff with a lower investment into reproduction, while still maintaining a similar total frond area on the water surface to capture the reduced light.
@@
 There is no general knowledge for overall distributions of effect sizes for the impact of evolution for species composition in communities, nor are there studies that evaluate drivers of shifts in these distributions of effect sizes in complex communities (see [@Hermann2024] for estimates in a predator-prey system).
-We present here distributions of effect sizes of evolution for species abundances (i.e.
-Figure \@ref(fig:fig1) c,f,i; Figure \@ref(fig:fig2) c; Figure \@ref(fig:fig3) a,b) in simulated and experimental aquatic plant communities, and expect that these effect size coefficients can be useful for better understanding the relative importance of evolution as a driver of community composition.
+We present here distributions of effect sizes of trait change and evolution for species abundances (i.e.
+posterior estimates of regression coefficients; Figure \@ref(fig:fig3)c; Appendix S1), which allow estimation of the relative importance of evolution and trait change for community composition (i.e.
+Figure \@ref(fig:fig1) c,f,i; Figure \@ref(fig:fig2) c; Figure \@ref(fig:fig3) a,b) in simulated and experimental aquatic plant communities.
+We expect that effect size coefficients can be useful for better understanding the relative importance of evolution as a driver of community composition.
 
@@
 Two critically important additions that will increase the application to field or experimental observation data will be (1) bidirectional causality and eco-evolutionary feedbacks and (2) the relative contribution of plasticity and genotypic evolution for trait change (i.e.
-[@Gibert2022]).
+[@Gibert2022], [@Jewell2023]).
 It will also be necessary to improve knowledge of the distributions of eco-evolutionary effects and how these are influenced by causal processes that change the speed of evolution or the strength of species interactions ([@Fronhofer2023]).

### 2024-11-30 manuscript/Appendix_S4_v01.Rmd

--- 2024-10-29 manuscript/Appendix_S4_v01.Rmd
@@
 ---
-title: "Appendix S4. The relative importance of evolution in populations and communities"
+title: "Appendix S4. The interpretation of the estimated trait change effect size coefficient"
 author:
@@
 
-# Logistic growth and trait evolution
+# Empirical data
 
-We can create simulations of evolutionary rescue in a population, where adaptive evolution rescues a population from extinction, at a speed that depends on the heritability level in the system.
-Population dynamics are modelled as:
+To correctly interpret correlation of trait change with population density of the species (as seen in Fig. 3) we first need to check the raw data.
 
-$$ N_{i,t+1} =\frac{\hat{W}e^ \frac{-[(\frac{w+(1-h^2)P} {P+w})(E-x_{i,t})]^2}{2(P+w)}N_{i,t}} {1 + \alpha_{ii}N_{i,t} + \alpha_{ij} N_{j,t}} $$
-
-where \( \hat{W} \) is calculated as \( \hat{W}=W_{max}\sqrt(\frac{w}{P+w}) \), \(W_{max}\) is the species' maximum per-capita growth rate, *w* is the width of the Gaussian fitness function (which determines the strength of selection, as increasing values indicate a weaker reduction in fitness with distance from optimum trait value), *P* is the width of the distribution of the phenotype *x*, and $h^2$ is the heritability of the trait *x*.
-For the simulation we use \(W_{max}\) = 2, *P* = 1, and *w* = 2.
-
-The change in the average trait value each time step is given by:
-
-$$ d_{i,t+1} = k d_{i,t} $$
-where \( k = \frac{w + (1-h^2) P}{w + P} \) and \( d_{i,t} = E_t - x_{i,t} \).
-
-## Population dynamics simulation
-
-```{r}
-### 1 species, h2=0 ###
-# Initial conditions
-N0 <- 10
-alpha.11 <- 0.01
-alpha <- matrix(alpha.11,byrow=FALSE)
-E.0 <- 0.8
-P <- 1
-w <- 0.5
-# Draw initial trait value using degree of initial maladaptation
-B0 <- 1
-d0 = sqrt(B0*(w+P))
-x.0 = E.0-d0
-Wmax <- 2
-h2 <- 0
-k <- (w + (1 - h2) * P)/(P + w)
-# Simulation of model for t time steps
-# Simulation of model for t time steps
-t <- 60
-N <- array(NA,dim=c(t,length(N0)))
-N <- as.data.frame(N)
-colnames(N) <- paste0("N",1:length(N0))
-N[1,] <- N0
-E <- rep(NA, t)
-E[1] <- E.0
-x <- array(NA,dim=c(t,length(N0)))
-x <- as.data.frame(x)
-colnames(x) <- paste0("x",1:length(N0))
-x[1,] <- x.0
-r <- array(NA,dim=c(t,length(N0)))
-r <- as.data.frame(r)
-colnames(r) <- paste0("r",1:length(N0))
-What <- Wmax*sqrt(w/(P+w))
-r[1,] <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E[1]-x[1,]))^2)/(2*(P+w)))
-for (i in 2:t) {
-  res <- disc_LV_evol(N0=N[i-1,],alpha=alpha,E=E[i-1],x=x[i-1,],P=P,w=w,Wmax=Wmax,h2=h2)
-  N[i,] <- res$Nt1
-  r[i,] <- res$r
-  # trait change
-  d <- E[i-1] - x[i-1,]
-  d1 <- k*d
-  x[i,] <- E[i-1] - d1
-  # environmental change
-  E[i] <- E[i-1] + abs(rnorm(1, 0, 0))
-}
-# Arrange to save across scenarios
-gdat <- N
-gdat$time <- 1:t
-gdat$h2 <- h2
-gdat$x1 <- x$x1
-gdat$E <- E
-```
-```{r}
-### 1 species, h2=0.01 ###
-# Initial conditions
-N0 <- 10
-alpha.11 <- 0.01
-alpha <- matrix(alpha.11,byrow=FALSE)
-E.0 <- 0.8
-P <- 1
-w <- 0.5
-# Draw initial trait value using degree of initial maladaptation
-B0 <- 1
-d0 = sqrt(B0*(w+P))
-x.0 = E.0-d0
-Wmax <- 2
-h2 <- 0.01
-k <- (w + (1 - h2) * P)/(P + w)
-# Simulation of model for t time steps
-# Simulation of model for t time steps
-t <- 60
-N <- array(NA,dim=c(t,length(N0)))
-N <- as.data.frame(N)
-colnames(N) <- paste0("N",1:length(N0))
-N[1,] <- N0
-E <- rep(NA, t)
-E[1] <- E.0
-x <- array(NA,dim=c(t,length(N0)))
-x <- as.data.frame(x)
-colnames(x) <- paste0("x",1:length(N0))
-x[1,] <- x.0
-r <- array(NA,dim=c(t,length(N0)))
-r <- as.data.frame(r)
-colnames(r) <- paste0("r",1:length(N0))
-What <- Wmax*sqrt(w/(P+w))
-r[1,] <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E[1]-x[1,]))^2)/(2*(P+w)))
-for (i in 2:t) {
-  res <- disc_LV_evol(N0=N[i-1,],alpha=alpha,E=E[i-1],x=x[i-1,],P=P,w=w,Wmax=Wmax,h2=h2)
-  N[i,] <- res$Nt1
-  r[i,] <- res$r
-  # trait change
-  d <- E[i-1] - x[i-1,]
-  d1 <- k*d
-  x[i,] <- E[i-1] - d1
-  # environmental change
-  E[i] <- E[i-1] + abs(rnorm(1, 0, 0))
-}
-# Arrange to save across scenarios
-gdat <- rbind(gdat,data.frame(N,time=1:t,h2=rep(h2,t),x1=x$x1,E=E))
-```
-```{r}
-### 1 species, h2=0.02 ###
-# Initial conditions
-N0 <- 10
-alpha.11 <- 0.01
-alpha <- matrix(alpha.11,byrow=FALSE)
-E.0 <- 0.8
-P <- 1
-w <- 0.5
-# Draw initial trait value using degree of initial maladaptation
-B0 <- 1
-d0 = sqrt(B0*(w+P))
-x.0 = E.0-d0
-Wmax <- 2
-h2 <- 0.02
-k <- (w + (1 - h2) * P)/(P + w)
-# Simulation of model for t time steps
-# Simulation of model for t time steps
-t <- 60
-N <- array(NA,dim=c(t,length(N0)))
-N <- as.data.frame(N)
-colnames(N) <- paste0("N",1:length(N0))
-N[1,] <- N0
-E <- rep(NA, t)
-E[1] <- E.0
-x <- array(NA,dim=c(t,length(N0)))
-x <- as.data.frame(x)
-colnames(x) <- paste0("x",1:length(N0))
-x[1,] <- x.0
-r <- array(NA,dim=c(t,length(N0)))
-r <- as.data.frame(r)
-colnames(r) <- paste0("r",1:length(N0))
-What <- Wmax*sqrt(w/(P+w))
-r[1,] <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E[1]-x[1,]))^2)/(2*(P+w)))
-for (i in 2:t) {
-  res <- disc_LV_evol(N0=N[i-1,],alpha=alpha,E=E[i-1],x=x[i-1,],P=P,w=w,Wmax=Wmax,h2=h2)
-  N[i,] <- res$Nt1
-  r[i,] <- res$r
-  # trait change
-  d <- E[i-1] - x[i-1,]
-  d1 <- k*d
-  x[i,] <- E[i-1] - d1
-  # environmental change
-  E[i] <- E[i-1] + abs(rnorm(1, 0, 0))
-}
-# Arrange to save across scenarios
-gdat <- rbind(gdat,data.frame(N,time=1:t,h2=rep(h2,t),x1=x$x1,E=E))
-```
-```{r}
-### 1 species, h2=0.03 ###
-# Initial conditions
-N0 <- 10
-alpha.11 <- 0.01
-alpha <- matrix(alpha.11,byrow=FALSE)
-E.0 <- 0.8
-P <- 1
-w <- 0.5
-# Draw initial trait value using degree of initial maladaptation
-B0 <- 1
-d0 = sqrt(B0*(w+P))
-x.0 = E.0-d0
-Wmax <- 2
-h2 <- 0.03
-k <- (w + (1 - h2) * P)/(P + w)
-# Simulation of model for t time steps
-# Simulation of model for t time steps
-t <- 60
-N <- array(NA,dim=c(t,length(N0)))
-N <- as.data.frame(N)
-colnames(N) <- paste0("N",1:length(N0))
-N[1,] <- N0
-E <- rep(NA, t)
-E[1] <- E.0
-x <- array(NA,dim=c(t,length(N0)))
-x <- as.data.frame(x)
-colnames(x) <- paste0("x",1:length(N0))
-x[1,] <- x.0
-r <- array(NA,dim=c(t,length(N0)))
-r <- as.data.frame(r)
-colnames(r) <- paste0("r",1:length(N0))
-What <- Wmax*sqrt(w/(P+w))
-r[1,] <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E[1]-x[1,]))^2)/(2*(P+w)))
-for (i in 2:t) {
-  res <- disc_LV_evol(N0=N[i-1,],alpha=alpha,E=E[i-1],x=x[i-1,],P=P,w=w,Wmax=Wmax,h2=h2)
-  N[i,] <- res$Nt1
-  r[i,] <- res$r
-  # trait change
-  d <- E[i-1] - x[i-1,]
-  d1 <- k*d
-  x[i,] <- E[i-1] - d1
-  # environmental change
-  E[i] <- E[i-1] + abs(rnorm(1, 0, 0))
-}
-# Arrange to save across scenarios
-gdat <- rbind(gdat,data.frame(N,time=1:t,h2=rep(h2,t),x1=x$x1,E=E))
-```
-```{r}
-### 1 species, h2=0.04 ###
-# Initial conditions
-N0 <- 10
-alpha.11 <- 0.01
-alpha <- matrix(alpha.11,byrow=FALSE)
-E.0 <- 0.8
-P <- 1
-w <- 0.5
-# Draw initial trait value using degree of initial maladaptation
-B0 <- 1
-d0 = sqrt(B0*(w+P))
-x.0 = E.0-d0
-Wmax <- 2
-h2 <- 0.04
-k <- (w + (1 - h2) * P)/(P + w)
-# Simulation of model for t time steps
-# Simulation of model for t time steps
-t <- 60
-N <- array(NA,dim=c(t,length(N0)))
-N <- as.data.frame(N)
-colnames(N) <- paste0("N",1:length(N0))
-N[1,] <- N0
-E <- rep(NA, t)
-E[1] <- E.0
-x <- array(NA,dim=c(t,length(N0)))
-x <- as.data.frame(x)
-colnames(x) <- paste0("x",1:length(N0))
-x[1,] <- x.0
-r <- array(NA,dim=c(t,length(N0)))
-r <- as.data.frame(r)
-colnames(r) <- paste0("r",1:length(N0))
-What <- Wmax*sqrt(w/(P+w))
-r[1,] <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E[1]-x[1,]))^2)/(2*(P+w)))
-for (i in 2:t) {
-  res <- disc_LV_evol(N0=N[i-1,],alpha=alpha,E=E[i-1],x=x[i-1,],P=P,w=w,Wmax=Wmax,h2=h2)
-  N[i,] <- res$Nt1
-  r[i,] <- res$r
-  # trait change
-  d <- E[i-1] - x[i-1,]
-  d1 <- k*d
-  x[i,] <- E[i-1] - d1
-  # environmental change
-  E[i] <- E[i-1] + abs(rnorm(1, 0, 0))
-}
-# Arrange to save across scenarios
-gdat <- rbind(gdat,data.frame(N,time=1:t,h2=rep(h2,t),x1=x$x1,E=E))
-```
-```{r}
-### 1 species, h2=0.05 ###
-# Initial conditions
-N0 <- 10
-alpha.11 <- 0.01
-alpha <- matrix(alpha.11,byrow=FALSE)
-E.0 <- 0.8
-P <- 1
-w <- 0.5
-# Draw initial trait value using degree of initial maladaptation
-B0 <- 1
-d0 = sqrt(B0*(w+P))
-x.0 = E.0-d0
-Wmax <- 2
-h2 <- 0.05
-k <- (w + (1 - h2) * P)/(P + w)
-# Simulation of model for t time steps
-# Simulation of model for t time steps
-t <- 60
-N <- array(NA,dim=c(t,length(N0)))
-N <- as.data.frame(N)
-colnames(N) <- paste0("N",1:length(N0))
-N[1,] <- N0
-E <- rep(NA, t)
-E[1] <- E.0
-x <- array(NA,dim=c(t,length(N0)))
-x <- as.data.frame(x)
-colnames(x) <- paste0("x",1:length(N0))
-x[1,] <- x.0
-r <- array(NA,dim=c(t,length(N0)))
-r <- as.data.frame(r)
-colnames(r) <- paste0("r",1:length(N0))
-What <- Wmax*sqrt(w/(P+w))
-r[1,] <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E[1]-x[1,]))^2)/(2*(P+w)))
-for (i in 2:t) {
-  res <- disc_LV_evol(N0=N[i-1,],alpha=alpha,E=E[i-1],x=x[i-1,],P=P,w=w,Wmax=Wmax,h2=h2)
-  N[i,] <- res$Nt1
-  r[i,] <- res$r
-  # trait change
-  d <- E[i-1] - x[i-1,]
-  d1 <- k*d
-  x[i,] <- E[i-1] - d1
-  # environmental change
-  E[i] <- E[i-1] + abs(rnorm(1, 0, 0))
-}
-# Arrange to save across scenarios
-gdat <- rbind(gdat,data.frame(N,time=1:t,h2=rep(h2,t),x1=x$x1,E=E))
-```
-```{r}
-### 1 species, h2=0.075 ###
-# Initial conditions
-N0 <- 10
-alpha.11 <- 0.01
-alpha <- matrix(alpha.11,byrow=FALSE)
-E.0 <- 0.8
-P <- 1
-w <- 0.5
-# Draw initial trait value using degree of initial maladaptation
-B0 <- 1
-d0 = sqrt(B0*(w+P))
-x.0 = E.0-d0
-Wmax <- 2
-h2 <- 0.075
-k <- (w + (1 - h2) * P)/(P + w)
-# Simulation of model for t time steps
-# Simulation of model for t time steps
-t <- 60
-N <- array(NA,dim=c(t,length(N0)))
-N <- as.data.frame(N)
-colnames(N) <- paste0("N",1:length(N0))
-N[1,] <- N0
-E <- rep(NA, t)
-E[1] <- E.0
-x <- array(NA,dim=c(t,length(N0)))
-x <- as.data.frame(x)
-colnames(x) <- paste0("x",1:length(N0))
-x[1,] <- x.0
-r <- array(NA,dim=c(t,length(N0)))
-r <- as.data.frame(r)
-colnames(r) <- paste0("r",1:length(N0))
-What <- Wmax*sqrt(w/(P+w))
-r[1,] <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E[1]-x[1,]))^2)/(2*(P+w)))
-for (i in 2:t) {
-  res <- disc_LV_evol(N0=N[i-1,],alpha=alpha,E=E[i-1],x=x[i-1,],P=P,w=w,Wmax=Wmax,h2=h2)
-  N[i,] <- res$Nt1
-  r[i,] <- res$r
-  # trait change
-  d <- E[i-1] - x[i-1,]
-  d1 <- k*d
-  x[i,] <- E[i-1] - d1
-  # environmental change
-  E[i] <- E[i-1] + abs(rnorm(1, 0, 0))
-}
-# Arrange to save across scenarios
-gdat <- rbind(gdat,data.frame(N,time=1:t,h2=rep(h2,t),x1=x$x1,E=E))
-```
-```{r}
-### 1 species, h2=0.1 ###
-# Initial conditions
-N0 <- 10
-alpha.11 <- 0.01
-alpha <- matrix(alpha.11,byrow=FALSE)
-E.0 <- 0.8
-P <- 1
-w <- 0.5
-# Draw initial trait value using degree of initial maladaptation
-B0 <- 1
-d0 = sqrt(B0*(w+P))
-x.0 = E.0-d0
-Wmax <- 2
-h2 <- 0.1
-k <- (w + (1 - h2) * P)/(P + w)
-# Simulation of model for t time steps
-# Simulation of model for t time steps
-t <- 60
-N <- array(NA,dim=c(t,length(N0)))
-N <- as.data.frame(N)
-colnames(N) <- paste0("N",1:length(N0))
-N[1,] <- N0
-E <- rep(NA, t)
-E[1] <- E.0
-x <- array(NA,dim=c(t,length(N0)))
-x <- as.data.frame(x)
-colnames(x) <- paste0("x",1:length(N0))
-x[1,] <- x.0
-r <- array(NA,dim=c(t,length(N0)))
-r <- as.data.frame(r)
-colnames(r) <- paste0("r",1:length(N0))
-What <- Wmax*sqrt(w/(P+w))
-r[1,] <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E[1]-x[1,]))^2)/(2*(P+w)))
-for (i in 2:t) {
-  res <- disc_LV_evol(N0=N[i-1,],alpha=alpha,E=E[i-1],x=x[i-1,],P=P,w=w,Wmax=Wmax,h2=h2)
-  N[i,] <- res$Nt1
-  r[i,] <- res$r
-  # trait change
-  d <- E[i-1] - x[i-1,]
-  d1 <- k*d
-  x[i,] <- E[i-1] - d1
-  # environmental change
-  E[i] <- E[i-1] + abs(rnorm(1, 0, 0))
-}
-# Arrange to save across scenarios
-gdat <- rbind(gdat,data.frame(N,time=1:t,h2=rep(h2,t),x1=x$x1,E=E))
-```
-```{r}
-### 1 species, h2=0.4 ###
-# Initial conditions
-N0 <- 10
-alpha.11 <- 0.01
-alpha <- matrix(alpha.11,byrow=FALSE)
-E.0 <- 0.8
-P <- 1
-w <- 0.5
-# Draw initial trait value using degree of initial maladaptation
-B0 <- 1
-d0 = sqrt(B0*(w+P))
-x.0 = E.0-d0
-Wmax <- 2
-h2 <- 0.4
-k <- (w + (1 - h2) * P)/(P + w)
-# Simulation of model for t time steps
-# Simulation of model for t time steps
-t <- 60
-N <- array(NA,dim=c(t,length(N0)))
-N <- as.data.frame(N)
-colnames(N) <- paste0("N",1:length(N0))
-N[1,] <- N0
-E <- rep(NA, t)
-E[1] <- E.0
-x <- array(NA,dim=c(t,length(N0)))
-x <- as.data.frame(x)
-colnames(x) <- paste0("x",1:length(N0))
-x[1,] <- x.0
-r <- array(NA,dim=c(t,length(N0)))
-r <- as.data.frame(r)
-colnames(r) <- paste0("r",1:length(N0))
-What <- Wmax*sqrt(w/(P+w))
-r[1,] <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E[1]-x[1,]))^2)/(2*(P+w)))
-for (i in 2:t) {
-  res <- disc_LV_evol(N0=N[i-1,],alpha=alpha,E=E[i-1],x=x[i-1,],P=P,w=w,Wmax=Wmax,h2=h2)
-  N[i,] <- res$Nt1
-  r[i,] <- res$r
-  # trait change
-  d <- E[i-1] - x[i-1,]
-  d1 <- k*d
-  x[i,] <- E[i-1] - d1
-  # environmental change
-  E[i] <- E[i-1] + abs(rnorm(1, 0, 0))
-}
-# Arrange to save across scenarios
-gdat <- rbind(gdat,data.frame(N,time=1:t,h2=rep(h2,t),x1=x$x1,E=E))
-```
-```{r}
-### 1 species, h2=0.9 ###
-# Initial conditions
-N0 <- 10
-alpha.11 <- 0.01
-alpha <- matrix(alpha.11,byrow=FALSE)
-E.0 <- 0.8
-P <- 1
-w <- 0.5
-# Draw initial trait value using degree of initial maladaptation
-B0 <- 1
-d0 = sqrt(B0*(w+P))
-x.0 = E.0-d0
-Wmax <- 2
-h2 <- 0.9
-k <- (w + (1 - h2) * P)/(P + w)
-# Simulation of model for t time steps
-# Simulation of model for t time steps
-t <- 60
-N <- array(NA,dim=c(t,length(N0)))
-N <- as.data.frame(N)
-colnames(N) <- paste0("N",1:length(N0))
-N[1,] <- N0
-E <- rep(NA, t)
-E[1] <- E.0
-x <- array(NA,dim=c(t,length(N0)))
-x <- as.data.frame(x)
-colnames(x) <- paste0("x",1:length(N0))
-x[1,] <- x.0
-r <- array(NA,dim=c(t,length(N0)))
-r <- as.data.frame(r)
-colnames(r) <- paste0("r",1:length(N0))
-What <- Wmax*sqrt(w/(P+w))
-r[1,] <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E[1]-x[1,]))^2)/(2*(P+w)))
-for (i in 2:t) {
-  res <- disc_LV_evol(N0=N[i-1,],alpha=alpha,E=E[i-1],x=x[i-1,],P=P,w=w,Wmax=Wmax,h2=h2)
-  N[i,] <- res$Nt1
-  r[i,] <- res$r
-  # trait change
-  d <- E[i-1] - x[i-1,]
-  d1 <- k*d
-  x[i,] <- E[i-1] - d1
-  # environmental change
-  E[i] <- E[i-1] + abs(rnorm(1, 0, 0))
-}
-# Arrange to save across scenarios
-gdat <- rbind(gdat,data.frame(N,time=1:t,h2=rep(h2,t),x1=x$x1,E=E))
-```
-```{r}
-### 1 species, h2=1 ###
-# Initial conditions
-N0 <- 10
-alpha.11 <- 0.01
-alpha <- matrix(alpha.11,byrow=FALSE)
-E.0 <- 0.8
-P <- 1
-w <- 0.5
-# Draw initial trait value using degree of initial maladaptation
-B0 <- 1
-d0 = sqrt(B0*(w+P))
-x.0 = E.0-d0
-Wmax <- 2
-h2 <- 1
-k <- (w + (1 - h2) * P)/(P + w)
-# Simulation of model for t time steps
-# Simulation of model for t time steps
-t <- 60
-N <- array(NA,dim=c(t,length(N0)))
-N <- as.data.frame(N)
-colnames(N) <- paste0("N",1:length(N0))
-N[1,] <- N0
-E <- rep(NA, t)
-E[1] <- E.0
-x <- array(NA,dim=c(t,length(N0)))
-x <- as.data.frame(x)
-colnames(x) <- paste0("x",1:length(N0))
-x[1,] <- x.0
-r <- array(NA,dim=c(t,length(N0)))
-r <- as.data.frame(r)
-colnames(r) <- paste0("r",1:length(N0))
-What <- Wmax*sqrt(w/(P+w))
-r[1,] <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E[1]-x[1,]))^2)/(2*(P+w)))
-for (i in 2:t) {
-  res <- disc_LV_evol(N0=N[i-1,],alpha=alpha,E=E[i-1],x=x[i-1,],P=P,w=w,Wmax=Wmax,h2=h2)
-  N[i,] <- res$Nt1
-  r[i,] <- res$r
-  # trait change
-  d <- E[i-1] - x[i-1,]
-  d1 <- k*d
-  x[i,] <- E[i-1] - d1
-  # environmental change
-  E[i] <- E[i-1] + abs(rnorm(1, 0, 0))
-}
-# Arrange to save across scenarios
-gdat <- rbind(gdat,data.frame(N,time=1:t,h2=rep(h2,t),x1=x$x1,E=E))
+```{r fig1, out.width="90%", include=TRUE, fig.align="center", fig.cap=c("Figure S05 1"), echo=FALSE}
+knitr::include_graphics("../output/fig_S05_1.pdf")
 ```
 
-```{r}
-### Plot of time series ###
-gdat$h2 <- as.factor(gdat$h2)
-ggplot2::ggplot(gdat, aes(time, N1, col=h2)) + geom_point()
-### ###
+From the empirical data we can detect that species Rn is drastically increasing in frond size over the course of the experiment, with a peak in trait change during week 5 and 7.
+Simultaneously it is only slowly increasing in population density, with slightly lower density during week 5.
+This leads to the highest value in $|\Delta x|$ during a lower value in the population density of species Rn.
+Similarly species Wc had large trait changes, during time points when species Lm and Sp showed an increase in population density.
+HMSC draws a linear correlation between both and in (\@ref(fig:fig2)) it can be seen how HMSC infers these correlations (the logarithmic scale of $|\Delta x|$ was used for the figure).
+
+```{r fig2, out.width="90%", include=TRUE, fig.align="center", fig.cap=c("Figure S05 2"), echo=FALSE}
+knitr::include_graphics("../output/fig_S05_2.pdf")
 ```
 
-## HMSC model fit
-
-The goal of fitting the data to a statistical model is to estimate the relative important of trait evolution for the population dynamics.
-
-```{r}
-### Model 1, h2=00 ###
-dat <- cbind(log(gdat$N1[gdat$h2==0]),gdat$x1[gdat$h2==0])
-dat <- as.data.frame(dat)
-colnames(dat) <- c("N1","x1")
-dat$time <- 1:t
-dat <- as.data.frame(dat)
-df <- data.frame(dat[(2:t),-3])
-colnames(df) <- c("Nt1", "xt1")
-df$dx1 <- abs(dat$x1[2:t] - dat$x1[1:(t - 1)])
-# prepare data in HMSC format
-Y <- as.matrix(df$Nt1)
-XData <- data.frame(cbind(E[1:(t-1)],E[1:(t-1)]^2,abs(dat$x1[2:t] - dat$x1[1:(t-1)])))
-colnames(XData) <- c("E","Esq","dx1")
-
-studyDesign = data.frame(sample = as.factor(1:(t-1)))
-rL = HmscRandomLevel(units = studyDesign$sample)
-
-m.1.hmsc = Hmsc(Y = Y, XData = XData, XFormula = ~E+Esq+dx1, studyDesign = studyDesign, ranLevels = list(sample = rL))
-# Bayesian model parameters
-nChains <- 2
-thin <- 5
-samples <- 2000
-transient <- 1000*thin
-verbose <- 500*thin
-# sample MCMC
-m.1.sample <- sampleMcmc(m.1.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
-m1.post.hmsc <- convertToCodaObject(m.1.sample)
-```
-
-```{r warning=FALSE}
-VP1 <- computeVariancePartitioning(m.1.sample,group=c(1,1,1,2),groupnames=c("Env","x1"))
-plotVariancePartitioning(m.1.sample,VP1,cols=c("white","skyblue","darkgrey"),args.legend=list(cex=0.75,bg="transparent"))
-```
-```{r}
-# prepare for plot
-pred2 <- predict(m.1.sample)
-hold1 <- list()
-hold2 <- list()
-for(i in 1:length(N0)){
-  hold1[[i]] <- as.numeric(Y[,i])
-  data <- lapply(pred2, function(x) x[, colnames(x)[i], drop = FALSE])
-  hold2[[i]] <- matrix(unlist(data), nrow=length(pred2), byrow=TRUE)
-}
-y <- unlist(hold1) # vector of each species' observed N across time points
-yrep <- as.matrix(data.frame(hold2))
-### Mod2 plot ###
-par(mgp=c(2,0.45,0), tcl=-0.4, mar=c(1.3,1.2,0,0))
-color_scheme_set("brightblue")
-#ppc_dens_overlay(y, yrep[1:50, ])
-ppc_intervals(y,yrep,x=dat$time[1:(t-1)],prob = 0.95) +
-  labs(x = "time",y = "ln(N)",) +
-  theme(plot.margin = unit(c(0, 0, 0, 0), "cm")) +
-  theme(legend.position = "none")
-```
-
-```{r}
-### Model 2, h2=0.01 ###
-dat <- cbind(log(gdat$N1[gdat$h2==0.01]),gdat$x1[gdat$h2==0.01])
-dat <- as.data.frame(dat)
-colnames(dat) <- c("N1","x1")
-dat$time <- 1:t
-dat <- as.data.frame(dat)
-df <- data.frame(dat[(2:t),-3])
-colnames(df) <- c("Nt1", "xt1")
-df$dx1 <- abs(dat$x1[2:t] - dat$x1[1:(t - 1)])
-# prepare data in HMSC format
-Y <- as.matrix(df$Nt1)
-XData <- data.frame(cbind(E[1:(t-1)],E[1:(t-1)]^2,abs(dat$x1[2:t] - dat$x1[1:(t-1)])))
-colnames(XData) <- c("E","Esq","dx1")
-
-studyDesign = data.frame(sample = as.factor(1:(t-1)))
-rL = HmscRandomLevel(units = studyDesign$sample)
-
-m.2.hmsc = Hmsc(Y = Y, XData = XData, XFormula = ~E+Esq+dx1, studyDesign = studyDesign, ranLevels = list(sample = rL))
-# Bayesian model parameters
-nChains <- 2
-thin <- 5
-samples <- 2000
-transient <- 1000*thin
-verbose <- 500*thin
-# sample MCMC
-m.2.sample <- sampleMcmc(m.2.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
-m2.post.hmsc <- convertToCodaObject(m.2.sample)
-```
-
-```{r warning=FALSE}
-VP2 <- computeVariancePartitioning(m.2.sample,group=c(1,1,1,2),groupnames=c("Env","x1"))
-plotVariancePartitioning(m.2.sample,VP2,cols=c("white","skyblue","darkgrey"),args.legend=list(cex=0.75,bg="transparent"))
-```
-```{r}
-# prepare for plot
-pred2 <- predict(m.2.sample)
-hold1 <- list()
-hold2 <- list()
-for(i in 1:length(N0)){
-  hold1[[i]] <- as.numeric(Y[,i])
-  data <- lapply(pred2, function(x) x[, colnames(x)[i], drop = FALSE])
-  hold2[[i]] <- matrix(unlist(data), nrow=length(pred2), byrow=TRUE)
-}
-y <- unlist(hold1) # vector of each species' observed N across time points
-yrep <- as.matrix(data.frame(hold2))
-### Mod2 plot ###
-par(mgp=c(2,0.45,0), tcl=-0.4, mar=c(1.3,1.2,0,0))
-color_scheme_set("brightblue")
-#ppc_dens_overlay(y, yrep[1:50, ])
-ppc_intervals(y,yrep,x=dat$time[1:(t-1)],prob = 0.95) +
-  labs(x = "time",y = "ln(N)",) +
-  theme(plot.margin = unit(c(0, 0, 0, 0), "cm")) +
-  theme(legend.position = "none")
-```
-
-```{r}
-### Model 3, h2=0.02 ###
-dat <- cbind(log(gdat$N1[gdat$h2==0.02]),gdat$x1[gdat$h2==0.02])
-dat <- as.data.frame(dat)
-colnames(dat) <- c("N1","x1")
-dat$time <- 1:t
-dat <- as.data.frame(dat)
-df <- data.frame(dat[(2:t),-3])
-colnames(df) <- c("Nt1", "xt1")
-df$dx1 <- abs(dat$x1[2:t] - dat$x1[1:(t - 1)])
-# prepare data in HMSC format
-Y <- as.matrix(df$Nt1)
-XData <- data.frame(cbind(E[1:(t-1)],E[1:(t-1)]^2,abs(dat$x1[2:t] - dat$x1[1:(t-1)])))
-colnames(XData) <- c("E","Esq","dx1")
-
-studyDesign = data.frame(sample = as.factor(1:(t-1)))
-rL = HmscRandomLevel(units = studyDesign$sample)
-
-m.3.hmsc = Hmsc(Y = Y, XData = XData, XFormula = ~E+Esq+dx1, studyDesign = studyDesign, ranLevels = list(sample = rL))
-# Bayesian model parameters
-nChains <- 2
-thin <- 5
-samples <- 2000
-transient <- 1000*thin
-verbose <- 500*thin
-# sample MCMC
-m.3.sample <- sampleMcmc(m.3.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
-m3.post.hmsc <- convertToCodaObject(m.3.sample)
-```
-
-```{r warning=FALSE}
-VP3 <- computeVariancePartitioning(m.3.sample,group=c(1,1,1,2),groupnames=c("Env","x1"))
-plotVariancePartitioning(m.3.sample,VP3,cols=c("white","skyblue","darkgrey"),args.legend=list(cex=0.75,bg="transparent"))
-```
-```{r}
-# prepare for plot
-pred2 <- predict(m.3.sample)
-hold1 <- list()
-hold2 <- list()
-for(i in 1:length(N0)){
-  hold1[[i]] <- as.numeric(Y[,i])
-  data <- lapply(pred2, function(x) x[, colnames(x)[i], drop = FALSE])
-  hold2[[i]] <- matrix(unlist(data), nrow=length(pred2), byrow=TRUE)
-}
-y <- unlist(hold1) # vector of each species' observed N across time points
-yrep <- as.matrix(data.frame(hold2))
-### Mod2 plot ###
-par(mgp=c(2,0.45,0), tcl=-0.4, mar=c(1.3,1.2,0,0))
-color_scheme_set("brightblue")
-#ppc_dens_overlay(y, yrep[1:50, ])
-ppc_intervals(y,yrep,x=dat$time[1:(t-1)],prob = 0.95) +
-  labs(x = "time",y = "ln(N)",) +
-  theme(plot.margin = unit(c(0, 0, 0, 0), "cm")) +
-  theme(legend.position = "none")
-```
-
-```{r}
-### Model 4, h2=0.03 ###
-dat <- cbind(log(gdat$N1[gdat$h2==0.03]),gdat$x1[gdat$h2==0.03])
-dat <- as.data.frame(dat)
-colnames(dat) <- c("N1","x1")
-dat$time <- 1:t
-dat <- as.data.frame(dat)
-df <- data.frame(dat[(2:t),-3])
-colnames(df) <- c("Nt1", "xt1")
-df$dx1 <- abs(dat$x1[2:t] - dat$x1[1:(t - 1)])
-# prepare data in HMSC format
-Y <- as.matrix(df$Nt1)
-XData <- data.frame(cbind(E[1:(t-1)],E[1:(t-1)]^2,abs(dat$x1[2:t] - dat$x1[1:(t-1)])))
-colnames(XData) <- c("E","Esq","dx1")
-
-studyDesign = data.frame(sample = as.factor(1:(t-1)))
-rL = HmscRandomLevel(units = studyDesign$sample)
-
-m.4.hmsc = Hmsc(Y = Y, XData = XData, XFormula = ~E+Esq+dx1, studyDesign = studyDesign, ranLevels = list(sample = rL))
-# Bayesian model parameters
-nChains <- 2
-thin <- 5
-samples <- 2000
-transient <- 1000*thin
-verbose <- 500*thin
-# sample MCMC
-m.4.sample <- sampleMcmc(m.4.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
-m4.post.hmsc <- convertToCodaObject(m.4.sample)
-```
-
-```{r warning=FALSE}
-VP4 <- computeVariancePartitioning(m.4.sample,group=c(1,1,1,2),groupnames=c("Env","x1"))
-plotVariancePartitioning(m.4.sample,VP4,cols=c("white","skyblue","darkgrey"),args.legend=list(cex=0.75,bg="transparent"))
-```
-```{r}
-# prepare for plot
-pred2 <- predict(m.4.sample)
-hold1 <- list()
-hold2 <- list()
-for(i in 1:length(N0)){
-  hold1[[i]] <- as.numeric(Y[,i])
-  data <- lapply(pred2, function(x) x[, colnames(x)[i], drop = FALSE])
-  hold2[[i]] <- matrix(unlist(data), nrow=length(pred2), byrow=TRUE)
-}
-y <- unlist(hold1) # vector of each species' observed N across time points
-yrep <- as.matrix(data.frame(hold2))
-### Mod2 plot ###
-par(mgp=c(2,0.45,0), tcl=-0.4, mar=c(1.3,1.2,0,0))
-color_scheme_set("brightblue")
-#ppc_dens_overlay(y, yrep[1:50, ])
-ppc_intervals(y,yrep,x=dat$time[1:(t-1)],prob = 0.95) +
-  labs(x = "time",y = "ln(N)",) +
-  theme(plot.margin = unit(c(0, 0, 0, 0), "cm")) +
-  theme(legend.position = "none")
-```
-```{r}
-### Model 5, h2=0.04 ###
-dat <- cbind(log(gdat$N1[gdat$h2==0.04]),gdat$x1[gdat$h2==0.04])
-dat <- as.data.frame(dat)
-colnames(dat) <- c("N1","x1")
-dat$time <- 1:t
-dat <- as.data.frame(dat)
-df <- data.frame(dat[(2:t),-3])
-colnames(df) <- c("Nt1", "xt1")
-df$dx1 <- abs(dat$x1[2:t] - dat$x1[1:(t - 1)])
-# prepare data in HMSC format
-Y <- as.matrix(df$Nt1)
-XData <- data.frame(cbind(E[1:(t-1)],E[1:(t-1)]^2,abs(dat$x1[2:t] - dat$x1[1:(t-1)])))
-colnames(XData) <- c("E","Esq","dx1")
-
-studyDesign = data.frame(sample = as.factor(1:(t-1)))
-rL = HmscRandomLevel(units = studyDesign$sample)
-
-m.5.hmsc = Hmsc(Y = Y, XData = XData, XFormula = ~E+Esq+dx1, studyDesign = studyDesign, ranLevels = list(sample = rL))
-# Bayesian model parameters
-nChains <- 2
-thin <- 5
-samples <- 2000
-transient <- 1000*thin
-verbose <- 500*thin
-# sample MCMC
-m.5.sample <- sampleMcmc(m.5.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
-m5.post.hmsc <- convertToCodaObject(m.5.sample)
-```
-
-```{r warning=FALSE}
-VP5 <- computeVariancePartitioning(m.5.sample,group=c(1,1,1,2),groupnames=c("Env","x1"))
-plotVariancePartitioning(m.5.sample,VP5,cols=c("white","skyblue","darkgrey"),args.legend=list(cex=0.75,bg="transparent"))
-```
-```{r}
-# prepare for plot
-pred2 <- predict(m.5.sample)
-hold1 <- list()
-hold2 <- list()
-for(i in 1:length(N0)){
-  hold1[[i]] <- as.numeric(Y[,i])
-  data <- lapply(pred2, function(x) x[, colnames(x)[i], drop = FALSE])
-  hold2[[i]] <- matrix(unlist(data), nrow=length(pred2), byrow=TRUE)
-}
-y <- unlist(hold1) # vector of each species' observed N across time points
-yrep <- as.matrix(data.frame(hold2))
-### Mod2 plot ###
-par(mgp=c(2,0.45,0), tcl=-0.4, mar=c(1.3,1.2,0,0))
-color_scheme_set("brightblue")
-#ppc_dens_overlay(y, yrep[1:50, ])
-ppc_intervals(y,yrep,x=dat$time[1:(t-1)],prob = 0.95) +
-  labs(x = "time",y = "ln(N)",) +
-  theme(plot.margin = unit(c(0, 0, 0, 0), "cm")) +
-  theme(legend.position = "none")
-```
-```{r}
-### Model 6, h2=0.05 ###
-dat <- cbind(log(gdat$N1[gdat$h2==0.05]),gdat$x1[gdat$h2==0.05])
-dat <- as.data.frame(dat)
-colnames(dat) <- c("N1","x1")
-dat$time <- 1:t
-dat <- as.data.frame(dat)
-df <- data.frame(dat[(2:t),-3])
-colnames(df) <- c("Nt1", "xt1")
-df$dx1 <- abs(dat$x1[2:t] - dat$x1[1:(t - 1)])
-# prepare data in HMSC format
-Y <- as.matrix(df$Nt1)
-XData <- data.frame(cbind(E[1:(t-1)],E[1:(t-1)]^2,abs(dat$x1[2:t] - dat$x1[1:(t-1)])))
-colnames(XData) <- c("E","Esq","dx1")
-
-studyDesign = data.frame(sample = as.factor(1:(t-1)))
-rL = HmscRandomLevel(units = studyDesign$sample)
-
-m.6.hmsc = Hmsc(Y = Y, XData = XData, XFormula = ~E+Esq+dx1, studyDesign = studyDesign, ranLevels = list(sample = rL))
-# Bayesian model parameters
-nChains <- 2
-thin <- 5
-samples <- 2000
-transient <- 1000*thin
-verbose <- 500*thin
-# sample MCMC
-m.6.sample <- sampleMcmc(m.6.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
-m6.post.hmsc <- convertToCodaObject(m.6.sample)
-```
-
-```{r warning=FALSE}
-VP6 <- computeVariancePartitioning(m.6.sample,group=c(1,1,1,2),groupnames=c("Env","x1"))
-plotVariancePartitioning(m.6.sample,VP6,cols=c("white","skyblue","darkgrey"),args.legend=list(cex=0.75,bg="transparent"))
-```
-```{r}
-# prepare for plot
-pred2 <- predict(m.6.sample)
-hold1 <- list()
-hold2 <- list()
-for(i in 1:length(N0)){
-  hold1[[i]] <- as.numeric(Y[,i])
-  data <- lapply(pred2, function(x) x[, colnames(x)[i], drop = FALSE])
-  hold2[[i]] <- matrix(unlist(data), nrow=length(pred2), byrow=TRUE)
-}
-y <- unlist(hold1) # vector of each species' observed N across time points
-yrep <- as.matrix(data.frame(hold2))
-### Mod2 plot ###
-par(mgp=c(2,0.45,0), tcl=-0.4, mar=c(1.3,1.2,0,0))
-color_scheme_set("brightblue")
-#ppc_dens_overlay(y, yrep[1:50, ])
-ppc_intervals(y,yrep,x=dat$time[1:(t-1)],prob = 0.95) +
-  labs(x = "time",y = "ln(N)",) +
-  theme(plot.margin = unit(c(0, 0, 0, 0), "cm")) +
-  theme(legend.position = "none")
-```
-
-```{r}
-### Model 7, h2=0.075 ###
-dat <- cbind(log(gdat$N1[gdat$h2==0.075]),gdat$x1[gdat$h2==0.075])
-dat <- as.data.frame(dat)
-colnames(dat) <- c("N1","x1")
-dat$time <- 1:t
-dat <- as.data.frame(dat)
-df <- data.frame(dat[(2:t),-3])
-colnames(df) <- c("Nt1", "xt1")
-df$dx1 <- abs(dat$x1[2:t] - dat$x1[1:(t - 1)])
-# prepare data in HMSC format
-Y <- as.matrix(df$Nt1)
-XData <- data.frame(cbind(E[1:(t-1)],E[1:(t-1)]^2,abs(dat$x1[2:t] - dat$x1[1:(t-1)])))
-colnames(XData) <- c("E","Esq","dx1")
-
-studyDesign = data.frame(sample = as.factor(1:(t-1)))
-rL = HmscRandomLevel(units = studyDesign$sample)
-
-m.7.hmsc = Hmsc(Y = Y, XData = XData, XFormula = ~E+Esq+dx1, studyDesign = studyDesign, ranLevels = list(sample = rL))
-# Bayesian model parameters
-nChains <- 2
-thin <- 5
-samples <- 2000
-transient <- 1000*thin
-verbose <- 500*thin
-# sample MCMC
-m.7.sample <- sampleMcmc(m.7.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
-m7.post.hmsc <- convertToCodaObject(m.7.sample)
-```
-
-```{r warning=FALSE}
-VP7 <- computeVariancePartitioning(m.7.sample,group=c(1,1,1,2),groupnames=c("Env","x1"))
-plotVariancePartitioning(m.7.sample,VP7,cols=c("white","skyblue","darkgrey"),args.legend=list(cex=0.75,bg="transparent"))
-```
-```{r}
-# prepare for plot
-pred2 <- predict(m.7.sample)
-hold1 <- list()
-hold2 <- list()
-for(i in 1:length(N0)){
-  hold1[[i]] <- as.numeric(Y[,i])
-  data <- lapply(pred2, function(x) x[, colnames(x)[i], drop = FALSE])
-  hold2[[i]] <- matrix(unlist(data), nrow=length(pred2), byrow=TRUE)
-}
-y <- unlist(hold1) # vector of each species' observed N across time points
-yrep <- as.matrix(data.frame(hold2))
-### Mod2 plot ###
-par(mgp=c(2,0.45,0), tcl=-0.4, mar=c(1.3,1.2,0,0))
-color_scheme_set("brightblue")
-#ppc_dens_overlay(y, yrep[1:50, ])
-ppc_intervals(y,yrep,x=dat$time[1:(t-1)],prob = 0.95) +
-  labs(x = "time",y = "ln(N)",) +
-  theme(plot.margin = unit(c(0, 0, 0, 0), "cm")) +
-  theme(legend.position = "none")
-```
-```{r}
-### Model 8, h2=0.1 ###
-dat <- cbind(log(gdat$N1[gdat$h2==0.1]),gdat$x1[gdat$h2==0.1])
-dat <- as.data.frame(dat)
-colnames(dat) <- c("N1","x1")
-dat$time <- 1:t
-dat <- as.data.frame(dat)
-df <- data.frame(dat[(2:t),-3])
-colnames(df) <- c("Nt1", "xt1")
-df$dx1 <- abs(dat$x1[2:t] - dat$x1[1:(t - 1)])
-# prepare data in HMSC format
-Y <- as.matrix(df$Nt1)
-XData <- data.frame(cbind(E[1:(t-1)],E[1:(t-1)]^2,abs(dat$x1[2:t] - dat$x1[1:(t-1)])))
-colnames(XData) <- c("E","Esq","dx1")
-
-studyDesign = data.frame(sample = as.factor(1:(t-1)))
-rL = HmscRandomLevel(units = studyDesign$sample)
-
-m.8.hmsc = Hmsc(Y = Y, XData = XData, XFormula = ~E+Esq+dx1, studyDesign = studyDesign, ranLevels = list(sample = rL))
-# Bayesian model parameters
-nChains <- 2
-thin <- 5
-samples <- 2000
-transient <- 1000*thin
-verbose <- 500*thin
-# sample MCMC
-m.8.sample <- sampleMcmc(m.8.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
-m8.post.hmsc <- convertToCodaObject(m.8.sample)
-```
-
-```{r warning=FALSE}
-VP8 <- computeVariancePartitioning(m.8.sample,group=c(1,1,1,2),groupnames=c("Env","x1"))
-plotVariancePartitioning(m.8.sample,VP8,cols=c("white","skyblue","darkgrey"),args.legend=list(cex=0.75,bg="transparent"))
-```
-```{r}
-# prepare for plot
-pred2 <- predict(m.8.sample)
-hold1 <- list()
-hold2 <- list()
-for(i in 1:length(N0)){
-  hold1[[i]] <- as.numeric(Y[,i])
-  data <- lapply(pred2, function(x) x[, colnames(x)[i], drop = FALSE])
-  hold2[[i]] <- matrix(unlist(data), nrow=length(pred2), byrow=TRUE)
-}
-y <- unlist(hold1) # vector of each species' observed N across time points
-yrep <- as.matrix(data.frame(hold2))
-### Mod2 plot ###
-par(mgp=c(2,0.45,0), tcl=-0.4, mar=c(1.3,1.2,0,0))
-color_scheme_set("brightblue")
-#ppc_dens_overlay(y, yrep[1:50, ])
-ppc_intervals(y,yrep,x=dat$time[1:(t-1)],prob = 0.95) +
-  labs(x = "time",y = "ln(N)",) +
-  theme(plot.margin = unit(c(0, 0, 0, 0), "cm")) +
-  theme(legend.position = "none")
-```
-```{r}
-### Model 9, h2=0.4 ###
-dat <- cbind(log(gdat$N1[gdat$h2==0.4]),gdat$x1[gdat$h2==0.4])
-dat <- as.data.frame(dat)
-colnames(dat) <- c("N1","x1")
-dat$time <- 1:t
-dat <- as.data.frame(dat)
-df <- data.frame(dat[(2:t),-3])
-colnames(df) <- c("Nt1", "xt1")
-df$dx1 <- abs(dat$x1[2:t] - dat$x1[1:(t - 1)])
-# prepare data in HMSC format
-Y <- as.matrix(df$Nt1)
-XData <- data.frame(cbind(E[1:(t-1)],E[1:(t-1)]^2,abs(dat$x1[2:t] - dat$x1[1:(t-1)])))
-colnames(XData) <- c("E","Esq","dx1")
-
-studyDesign = data.frame(sample = as.factor(1:(t-1)))
-rL = HmscRandomLevel(units = studyDesign$sample)
-
-m.9.hmsc = Hmsc(Y = Y, XData = XData, XFormula = ~E+Esq+dx1, studyDesign = studyDesign, ranLevels = list(sample = rL))
-# Bayesian model parameters
-nChains <- 2
-thin <- 5
-samples <- 2000
-transient <- 1000*thin
-verbose <- 500*thin
-# sample MCMC
-m.9.sample <- sampleMcmc(m.9.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
-m9.post.hmsc <- convertToCodaObject(m.9.sample)
-```
-
-```{r warning=FALSE}
-VP9 <- computeVariancePartitioning(m.9.sample,group=c(1,1,1,2),groupnames=c("Env","x1"))
-plotVariancePartitioning(m.9.sample,VP9,cols=c("white","skyblue","darkgrey"),args.legend=list(cex=0.75,bg="transparent"))
-```
-```{r}
-# prepare for plot
-pred2 <- predict(m.9.sample)
-hold1 <- list()
-hold2 <- list()
-for(i in 1:length(N0)){
-  hold1[[i]] <- as.numeric(Y[,i])
-  data <- lapply(pred2, function(x) x[, colnames(x)[i], drop = FALSE])
-  hold2[[i]] <- matrix(unlist(data), nrow=length(pred2), byrow=TRUE)
-}
-y <- unlist(hold1) # vector of each species' observed N across time points
-yrep <- as.matrix(data.frame(hold2))
-### Mod2 plot ###
-par(mgp=c(2,0.45,0), tcl=-0.4, mar=c(1.3,1.2,0,0))
-color_scheme_set("brightblue")
-#ppc_dens_overlay(y, yrep[1:50, ])
-ppc_intervals(y,yrep,x=dat$time[1:(t-1)],prob = 0.95) +
-  labs(x = "time",y = "ln(N)",) +
-  theme(plot.margin = unit(c(0, 0, 0, 0), "cm")) +
-  theme(legend.position = "none")
-```
-```{r}
-### Model 10, h2=0.9 ###
-dat <- cbind(log(gdat$N1[gdat$h2==0.9]),gdat$x1[gdat$h2==0.9])
-dat <- as.data.frame(dat)
-colnames(dat) <- c("N1","x1")
-dat$time <- 1:t
-dat <- as.data.frame(dat)
-df <- data.frame(dat[(2:t),-3])
-colnames(df) <- c("Nt1", "xt1")
-df$dx1 <- abs(dat$x1[2:t] - dat$x1[1:(t - 1)])
-# prepare data in HMSC format
-Y <- as.matrix(df$Nt1)
-XData <- data.frame(cbind(E[1:(t-1)],E[1:(t-1)]^2,abs(dat$x1[2:t] - dat$x1[1:(t-1)])))
-colnames(XData) <- c("E","Esq","dx1")
-
-studyDesign = data.frame(sample = as.factor(1:(t-1)))
-rL = HmscRandomLevel(units = studyDesign$sample)
-
-m.10.hmsc = Hmsc(Y = Y, XData = XData, XFormula = ~E+Esq+dx1, studyDesign = studyDesign, ranLevels = list(sample = rL))
-# Bayesian model parameters
-nChains <- 2
-thin <- 5
-samples <- 2000
-transient <- 1000*thin
-verbose <- 500*thin
-# sample MCMC
-m.10.sample <- sampleMcmc(m.10.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
-m10.post.hmsc <- convertToCodaObject(m.10.sample)
-```
-
-```{r warning=FALSE}
-VP10 <- computeVariancePartitioning(m.10.sample,group=c(1,1,1,2),groupnames=c("Env","x1"))
-plotVariancePartitioning(m.10.sample,VP10,cols=c("white","skyblue","darkgrey"),args.legend=list(cex=0.75,bg="transparent"))
-```
-```{r}
-# prepare for plot
-pred2 <- predict(m.10.sample)
-hold1 <- list()
-hold2 <- list()
-for(i in 1:length(N0)){
-  hold1[[i]] <- as.numeric(Y[,i])
-  data <- lapply(pred2, function(x) x[, colnames(x)[i], drop = FALSE])
-  hold2[[i]] <- matrix(unlist(data), nrow=length(pred2), byrow=TRUE)
-}
-y <- unlist(hold1) # vector of each species' observed N across time points
-yrep <- as.matrix(data.frame(hold2))
-### Mod2 plot ###
-par(mgp=c(2,0.45,0), tcl=-0.4, mar=c(1.3,1.2,0,0))
-color_scheme_set("brightblue")
-#ppc_dens_overlay(y, yrep[1:50, ])
-ppc_intervals(y,yrep,x=dat$time[1:(t-1)],prob = 0.95) +
-  labs(x = "time",y = "ln(N)",) +
-  theme(plot.margin = unit(c(0, 0, 0, 0), "cm")) +
-  theme(legend.position = "none")
-```
-```{r}
-### Model 11, h2=1 ###
-dat <- cbind(log(gdat$N1[gdat$h2==1]),gdat$x1[gdat$h2==1])
-dat <- as.data.frame(dat)
-colnames(dat) <- c("N1","x1")
-dat$time <- 1:t
-dat <- as.data.frame(dat)
-df <- data.frame(dat[(2:t),-3])
-colnames(df) <- c("Nt1", "xt1")
-df$dx1 <- abs(dat$x1[2:t] - dat$x1[1:(t - 1)])
-# prepare data in HMSC format
-Y <- as.matrix(df$Nt1)
-XData <- data.frame(cbind(E[1:(t-1)],E[1:(t-1)]^2,abs(dat$x1[2:t] - dat$x1[1:(t-1)])))
-colnames(XData) <- c("E","Esq","dx1")
-
-studyDesign = data.frame(sample = as.factor(1:(t-1)))
-rL = HmscRandomLevel(units = studyDesign$sample)
-
-m.11.hmsc = Hmsc(Y = Y, XData = XData, XFormula = ~E+Esq+dx1, studyDesign = studyDesign, ranLevels = list(sample = rL))
-# Bayesian model parameters
-nChains <- 2
-thin <- 5
-samples <- 2000
-transient <- 1000*thin
-verbose <- 500*thin
-# sample MCMC
-m.11.sample <- sampleMcmc(m.11.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
-m11.post.hmsc <- convertToCodaObject(m.11.sample)
-```
-
-```{r warning=FALSE}
-VP11 <- computeVariancePartitioning(m.11.sample,group=c(1,1,1,2),groupnames=c("Env","x1"))
-plotVariancePartitioning(m.11.sample,VP11,cols=c("white","skyblue","darkgrey"),args.legend=list(cex=0.75,bg="transparent"))
-```
-```{r}
-# prepare for plot
-pred2 <- predict(m.11.sample)
-hold1 <- list()
-hold2 <- list()
-for(i in 1:length(N0)){
-  hold1[[i]] <- as.numeric(Y[,i])
-  data <- lapply(pred2, function(x) x[, colnames(x)[i], drop = FALSE])
-  hold2[[i]] <- matrix(unlist(data), nrow=length(pred2), byrow=TRUE)
-}
-y <- unlist(hold1) # vector of each species' observed N across time points
-yrep <- as.matrix(data.frame(hold2))
-### Mod2 plot ###
-par(mgp=c(2,0.45,0), tcl=-0.4, mar=c(1.3,1.2,0,0))
-color_scheme_set("brightblue")
-#ppc_dens_overlay(y, yrep[1:50, ])
-ppc_intervals(y,yrep,x=dat$time[1:(t-1)],prob = 0.95) +
-  labs(x = "time",y = "ln(N)",) +
-  theme(plot.margin = unit(c(0, 0, 0, 0), "cm")) +
-  theme(legend.position = "none")
-```
-
-```{r}
-mc_dat_1 <- mcmc_intervals_data(m1.post.hmsc$Beta, pars="B[dx1 (C4), sp1 (S1)]")
-mc_dat_2 <- mcmc_intervals_data(m2.post.hmsc$Beta, pars="B[dx1 (C4), sp1 (S1)]")
-mc_dat_3 <- mcmc_intervals_data(m3.post.hmsc$Beta, pars="B[dx1 (C4), sp1 (S1)]")
-mc_dat_4 <- mcmc_intervals_data(m4.post.hmsc$Beta, pars="B[dx1 (C4), sp1 (S1)]")
-mc_dat_5 <- mcmc_intervals_data(m5.post.hmsc$Beta, pars="B[dx1 (C4), sp1 (S1)]")
-mc_dat_6 <- mcmc_intervals_data(m6.post.hmsc$Beta, pars="B[dx1 (C4), sp1 (S1)]")
-mc_dat_7 <- mcmc_intervals_data(m7.post.hmsc$Beta, pars="B[dx1 (C4), sp1 (S1)]")
-mc_dat_8 <- mcmc_intervals_data(m8.post.hmsc$Beta, pars="B[dx1 (C4), sp1 (S1)]")
-mc_dat_9 <- mcmc_intervals_data(m9.post.hmsc$Beta, pars="B[dx1 (C4), sp1 (S1)]")
-mc_dat_10 <- mcmc_intervals_data(m10.post.hmsc$Beta, pars="B[dx1 (C4), sp1 (S1)]")
-mc_dat_11 <- mcmc_intervals_data(m11.post.hmsc$Beta, pars="B[dx1 (C4), sp1 (S1)]")
-
-combined <- rbind(mc_dat_1,mc_dat_2,mc_dat_3,mc_dat_4,mc_dat_5,mc_dat_6,mc_dat_7,mc_dat_8,mc_dat_9,mc_dat_10,mc_dat_11)
-combined$model <- rep(c("0","0.01","0.02","0.03","0.04","0.05","0.075","0.1","0.4","0.9","1"),times=c(dim(mc_dat_1)[1],dim(mc_dat_2)[1],dim(mc_dat_3)[1],dim(mc_dat_4)[1],dim(mc_dat_5)[1],dim(mc_dat_6)[1],dim(mc_dat_7)[1],dim(mc_dat_8)[1],dim(mc_dat_9)[1],dim(mc_dat_10)[1],dim(mc_dat_11)[1]))
-
-# make the plot using ggplot 
-theme_set(bayesplot::theme_default())
-pos <- position_nudge(y = c(0,0.1,0.2,0.3,0.4,0.5,0.6,0.7,0.8,0.9,1))
-ggplot(combined, aes(x = m, y = parameter, color = model)) + 
-  geom_linerange(aes(xmin = l, xmax = h), position = pos, size=2)+
-  geom_linerange(aes(xmin = ll, xmax = hh), position = pos)+
-  geom_point(aes(color=model),position = pos)
-
-ggplot(combined, aes(x = m, y = parameter, color = model)) + 
-  geom_linerange(aes(xmin = l, xmax = h), position = pos, size=2)+
-  geom_linerange(aes(xmin = ll, xmax = hh), position = pos)+
-  geom_point(aes(color=model),position = pos)+
-  coord_flip()+
-  geom_vline(xintercept=0,linetype="dashed")
-```
-
-# References
+It should be noted that species Rn was the only to strongly increase in frond size in this experiment, which may the result from its decrease in population size.
+Species Wc decreased in frond size, thus occupying less space on the water surface and potentially freeing space for the other species to occupy.
+We speculate on this, but the manuscript for the original data will have more information.

### 2025-09-29 manuscript/Appendix_S4_v02.Rmd
Updating to newest submission

---
title: "Appendix S4. Applying HMSC on empirical data"
author:
- "Jelena H. Pantel^[Laboratoire Chrono-environnement,UMR 6249 CNRS-UFC, 16 Route
  de Gray, 25030 Besançon cedex, France, jelena.pantel@univ-fcomte.fr]"
- Ruben J. Hermann^[University of Konstanz, Universitätsstraße 10, 78464 Konstanz,
  Germany, ruben.hermann@posteo.de]
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::pdf_document2:
    toc: false
    fig_caption: true
    latex_engine: xelatex
    extra_dependencies: float
  bookdown::html_document2: default
  pdf_document:
    toc: false
  word_document:
    reference_docx: docx_template.docx
header-includes:
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \fancyhead[L]{Author Pantel and Hermann}
- \fancyhead[R]{Detection of ecoevo dynamics}
- \usepackage{lineno}
- \linenumbers
link-citations: true
linkcolor: blue
csl: "the-american-naturalist.csl"
bibliography: refs.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
```{r setup,echo=FALSE,message=FALSE}
#library(ecoevor)
library(tidyverse)
library(reshape2)
library(Hmsc)
library(bayesplot)
bayesplot_theme_set(new = theme_default())
library(here)
knitr::opts_knit$set(root.dir = here())
```
```{r,echo=FALSE,message=FALSE,warning=FALSE}
# Wrap long lines
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy=TRUE,out.width="50%")
```

# General

Applying HMSC to empirical data can be broken down to a series of steps, which are generally described in much more detail in e.g. [@Ovaskainen2017HMSC], [@Tikhonov2020],[@Ovaskainen2020]. It is critical to keep in mind that HMSC is ultimately a linear model, and therefore uses a correlative approach (this is discussed in [@Ovaskainen2020]). We take a less quantitative, and more verbal and visual description of the steps required to apply HMSC with trait change to empirical data. We use the example that appears in the main manuscript from the experiment of [@Jewell2023], with code to implement the steps located in [github::jhpantel/ecoevo-hmsc/code/6_emp_hmsc.R](https://github.com/jhpantel/ecoevo-hmsc/blob/fe517557e5aff81aa3090521348a58c89d67bebe/code/R/6_emp_hmsc.R). We don’t repeat that script in its entirety here, but instead present elements for a step-by-step guide to fitting an HMSC model with changing trait values as fixed effects.

## Step 1: Arrange study data

The first step as in any statistical model is to define your fixed effects (covariates: $X$), which influence the measured variable ($Y$), as well as random effects. Each of these data subsets requires a separate matrix to be included into the HMSC model, where each row represents one data sample. The columns define the different categories for the samples (e.g. what is measured or spatial/temporal information).

In typical HMSC applications, trait values are included as a separate $species \times trait$ matrix containing a single constant value for each species, to evaluate their impact on the response of species to environmental predictors. In our application, we calculated trait changes over time and estimated the impact of these trait changes for species densities. In our HMSC model, we implemented the absolute value of trait change as a fixed effect, as this shows how trait change in general (irrespective of the direction of change) can impact species density.

It is important to note that HMSC does not check whether the samples are ordered in the same way between the different matrices (i.e. that each row corresponds to the same temporal, spatial, treatment etc), so the user must check that carefully before implementing them into the model. Constructing the data, predictor, and study design matrices properly is a key step before model evaluation, and we suggest to create reduced versions of these matrices to better understand their structure.

## Step 2: Formulate study data as matrices

HMSC requires a specific format for the various model elements. Fixed effect predictors must be defined as a data frame, then their relationship to the response variable $Y$ is specified in the `XFormula`. All effects must use the column names of $X$, and interaction effects can be included as well (`HMSC` uses R's `formula`, i.e. the same specification as for e.g. `lm` and `glm`).

In the empirical analysis presented in the main text, the $X$ matrix and `XFormula` are:

```{r echo=FALSE}
load("./data/emp/hmsc_emp.RData")
head(XData)
m.1_evo$XFormula
```

The study design matrix is HMSC's formulation for fixed and random effects, which can include elements such as experimental blocks or sampling plot hierarchical structure. The study design should be formatted as a dataframe, with columns as factors.

```{r echo=FALSE}
head(studyDesign[,1:2])
```

The random effects are then defined with the HMSC function `HmscRandomLevel`, applied on each random effect separately. In our empirical example, this is done as:

```{r eval=FALSE}
rL.time <- HmscRandomLevel(units = unique(data_abun$Week))
rL.rep <- HmscRandomLevel(units = unique(data_abun$C.Rep))
```

(where `data_abun` is the original dataframe with the study data that is used to construct model matrices). We scale fixed effects, implemented in Hmsc as `XScale = T` when the model is run in Step 3. HMSC allows specification for all model parameter priors, and the response variable can be modelled as normal, probit, Poisson, or lognormal Poisson (specified in the `distr=` argument when the model is run in Step 3).

## Step 3: Model running and diagnostic checks

To evaluate the statistical model, parameters for the Monte-Carlo Markov-Chain (MCMC), which draws samples from posterior distributions, must be defined. Choosing the right parameters to reach converged chains can take some time, and it is recommended to first develop the analysis pipeline with a subset of study data to reduce run time. It is also possible to use a high thinning value to reduce computation time. For example, we used the following values for MCMC and to begin sampling using Hmsc:

```{r eval=FALSE}
thin = 5000
samples = 250
transient = 125*thin
verbose = 5000
chains = 4
nParallel = 4

m <- Hmsc::Hmsc(Y=Y_log,XData=XData,studyDesign=studyDesign,XFormula=XFormula,ranLevels=list("Week"=rL.time,"C.Rep"=rL.rep),distr="normal",XScale=TRUE)

m.1_evo <- Hmsc::sampleMcmc(m,samples=samples,transient=transient,nChains=chains,verbose=verbose,thin=thin,updater = list(GammaEta = FALSE),nParallel=nParallel)
```

After the model samples have been generated, diagnostics for the Markov chains can be applied from the `coda` package if the HMSC output is converted to a `coda` object. We give here some diagnostics from our empirical application, and we also provide code for diagnostics in the R script [github::jhpantel/ecoevo-hmsc/code/4_diagnostics.R](https://github.com/jhpantel/ecoevo-hmsc/blob/fe517557e5aff81aa3090521348a58c89d67bebe/code/R/4_diagnostics.R).

```{r}
#Converting into a coda object
m.post <- convertToCodaObject(m.1_evo)
#Calculating the effective sample size (ESS) - which shows how many data points of the MCMC chain were independently and randomly drawn and as such can be effectively used
ess.beta<- effectiveSize(m.post$Beta)

#Calculating the potential scale reduction factor (PSRF) - which is from the Gelman-Rubin diagnostic is based a comparison of within-chain and between-chain variances, and is similar to a classical analysis of variance
psrf.beta <- gelman.diag(m.post$Beta, multivariate=FALSE)$psrf
```

A histogram can be used to plot ESS and PSRF, though more detailed plots should be used to detect possible sources of errors. Higher ESS (closer to the MCMC sample size) and PSRF closer to 1 PSRF is better (ideally exactly 1).

```{r echo=FALSE}
#Plotting simple histograms
hist(ess.beta,xlab="Effective sample site",main="")
hist(psrf.beta,xlab="Potential scale reduction factor",main="")
```

#### Step 3.1 Autocorrelation plot

Autocorrelation shows how much a sample point of MCMC is correlated to its previous sampling point, thus 0 would be the optimal value. Here the autocorrelation of all chains for species Lm are shown.

```{r, echo=FALSE}
bayesplot::mcmc_acf(m.post$Beta[,c(1,2,3,4,5,6,7)])
```

#### Step 3.2 Predictive vs explanatory power

The explanatory and predictive power of each model can also be calculated, though we note that the calculation of predictive power requires long computation time in more complex models given the use of cross validation.

```{r, echol=FALSE,eval=FALSE}
#To calculate the explanatory power
preds = computePredictedValues(m.1_evo)
MF = evaluateModelFit(hM=m.1_evo, predY=preds)

#To calculate the predictive power we need to do a cross-validation, will take some time
partition = createPartition(m.1_evo, nfolds = 2)
preds.cross = computePredictedValues(m.1, partition=partition, nParallel =2)
MF.cross = evaluateModelFit(hM=m.1, predY=preds.cross)
```

```{r, echol=FALSE,eval=FALSE}
#Plotting explanatory and predictive power for R^2 and residual mean squared error (RMSE)
R2_dat <-data.frame(y=c(unname(MF$R2),unname(MF.cross$R2)),x=c(rep("explanatory",length(MF$R2)),rep("predictive",length(MF.cross$R2))))
RMSE_dat <-data.frame(y=c(unname(MF$RMSE),unname(MF.cross$RMSE)),x=c(rep("explanatory",length(MF$R2)),rep("predictive",length(MF.cross$R2))))

p1<-ggplot(R2_dat,aes(y=y,x=x,fill=x))+
  geom_boxplot()+
  ylab("R2")+xlab("")+theme(legend.position="none")
p2<-ggplot(RMSE_dat,aes(y=y,x=x,fill=x))+
  geom_boxplot()  +
  ylab("RMSE")+xlab("")+theme(legend.position="none")

p1+p2 
```

Posterior distributions should be visually evaluated for problematic elements. Papers for beginners to Bayesian analysis can give more guidance, i.e. [@Giminez2025]. If all checks are satisfactory, then continue to the next step to interpret model results.

## Step 4: Model results and interpretation

It is useful to again recall that in the correlative HMSC model, different mechanisms may produce observed effect size estimates (main text, Box 1). In multi-species communities, there may also be a very large number of posterior distributions for model parameters for all species. Posterior model interpretation may be a large computational exercise for this reason. We provide an example for how to arrange HMSC posterior objects of the empirical data to summarize posterior distributions  though we note that these are specific to our metacommunity simulation results structure. Nevertheless, it provides one example of how to visualize large numbers of parameter posterior distributions.

```{r }
#### Step 0 Load the data --------------
load("./data/emp/hmsc_emp.RData")

#### Step 1. Evaluate posteriors --------------
m.post = Hmsc::convertToCodaObject(m.1_evo)
m.df <- do.call(rbind, m.post$Beta)

s <- ncol(m.1_evo$Y)
npred <- 1 + length(colnames(m.1_evo$XData))

# All beta coefficients where HDI doesn't include 0 ####
a <- apply(m.df,2,function(x) quantile(x,probs=c(0.025,0.975)))
b <- apply(a,2,function(x) prod(x[1],x[2]) < 0)
B.not0 <- m.df[,b==FALSE]
not0 <- colnames(B.not0)

# Separate coefficient estimate posteriors into groups
s <- ncol(m.1_evo$Y)
npred <- 1 + length(colnames(m.1_evo$XData))

# N ~ dLm ##########
B.dLm <- m.df[,seq(6,length(colnames(m.df)),npred)]
# All values, colored by overlap of HPDI with 0
xc<-NULL
for(i in 1:ncol(B.dLm)) xc[i]<-as.numeric(sum(colnames(B.dLm)[i]==not0))+1
theme_set(theme_grey())
#Renaming it to dLm and the according species
colnames(B.dLm) <- c("dLm_Lm","dLm_Rn","dLm_Sp","dLm_Wc")
bayesplot::mcmc_areas(B.dLm,rhat=xc) + legend_none() +  theme(text = element_text(size = 20))
```

### Example interpretation of estimated effect of trait evolution in one species

We provide more detail here for interpreting results from the empirical example with frond area and population density for four aquatic plant species. To interpret effect size results, begin by inspecting raw data (e.g. Figure \@ref(fig:fig1)).

```{r fig1, out.width="90%", include=TRUE, fig.align="center", fig.cap=c("Raw data from the experiments of Jewell and Bell 2024. (a) Population dynamics for each species shown in log(count), as it was implemented as such into the model. (b) The natural logarithm of trait change of all species over time, to visualize more clearly the changes over time. (c) Total change in trait value for all species over time. (d) The raw trait values of all species over time. Colors and linetype denote the different environmenal treamtents."), echo=FALSE}
knitr::include_graphics("../output/fig_S4_1.pdf",error=FALSE)
```

We observe that frond size for species Rn is rapidly increasing over the course of the experiment, with a peak in trait change during week 5 and 7, while it is only slowly increasing in population density (with only slightly lower density during week 5, Figure \@ref(fig:fig2)). The data consequently would have a high value of $|\Delta x|$ corresponding to a lower value of log(density) for species Rn, and the model estimates a negative effect size. Similarly species Lm has slight changes in its decrease of frond size (recall that all effects are scaled before being implemented into the model), which occur in the opposite direction of the density changes of Sp and Wc in the respective treatments. Therefore the model estimates also negative effect sizes in these cases (Figure \@ref(fig:fig3)).

```{r fig2, out.width="90%", include=TRUE, fig.align="center", fig.cap=c("We plot here four example where the posterior estimated direction of effect sizes with 95% HDI of trait change on species density were either positive or negative. (a) dRn against counts of species Rn (negative effect size). (b) dWc against counts of species Lm (positive effect size). (c) dLm against counts of species Sp (negative effect size). (d) dSp against counts of species Sp (positive effect size)."), echo=FALSE}
knitr::include_graphics("../output/fig_S4_2.pdf",error=FALSE)
```

```{r fig3, out.width="90%", include=TRUE, fig.align="center", fig.cap=c("Posterior distributions of the effect sizes from the four examples of Figure 3. (a) Posterior distribution of dRn on counts of Rn. (b) Posterior distribution of dWc on counts of Lm. (c) Posterior distribution of dLm on counts of Sp. (d) Posterior distribution of dSp on counts of Sp. The blue vertical line denotes the median and the shaded area the inner 50% of the probability mass from the posterior distribution."), echo=FALSE}
knitr::include_graphics("../output/fig_S4_3.pdf",error=FALSE)
```

It should also be noted that species Rn was the only to strongly increase in frond size in this experiment, which may result from its decrease in population size. Species Sp decreased in frond size, thus occupying less space on the water surface and potentially freeing space to occupy. This might for example, be the mechanistic explanation of the negative effect of dRn for density of Rn and positive effect of dSp for density of Sp (but see Box 1 and [@Jewell2023] for information about interpreting such effects).

# References

### 2025-09-29 code/R/1_metacom_sim.R

@@ -1,7 +1,7 @@
 #Title. 1_metacom_sim.R
 # Author. J.H. Pantel &  R.J. Hermann
 # Date 29.07.2025
-# Description. This script contains the simulation, which is analysed with HMSC. Associated manuscript Pantel & Hermann 2025.
+# Description. This script contains the metacommunity simulation, which is analysed with HMSC. Associated manuscript Pantel & Hermann 2025.
   
 library(vegan)
 library(ggplot2)
@@ -59,12 +59,6 @@ xt = array(9999,c(patch,spec,time))     # Record of phenotype value over time, p
 Et = array(numeric(),c(patch,time))   # Record of environmental value over time, patch x spec x time
 
 var = 0 #Adding variation to the environmental value for no variation set to 0 (var is the sd of the normal distribution with a mean of 0)
-## Heritability
-# Scenario 1: No evolution
-#h2 = array(0,c(1,spec))
-
-# Scenario 2: Evolution, same heritatibility, high
-#h2 = array(0.9,c(1,spec))
 
 extprob = 0.001    # Probability that species goes extinct when under critical density
 repnum = 1000     # Total number of replicates of simulations
@@ -79,7 +73,7 @@ K = array(1000,c(patch,spec)) # Carrying capacity of each species in each patch
 # Spatial connectivity and dispersal
 ##Random spatial location
 xy = array(runif(patch*2),c(patch,2))
-## Unifrom spatial locations
+## Uniform spatial locations
 #x.1 = seq(0.1,1,0.1)
 #x = array(x.1,c(length(x.1)*5,1))
 #y.1 = seq(0.1,1,0.2)

### 2025-09-29 code/R/2_results_sim.R

@@ -114,7 +114,7 @@ for(i in 1:dim(N)[3]){
 mycols <- c("maroon4","coral3","peachpuff3","gray47","hotpink2","darkorange","cyan3","khaki1","firebrick1","cadetblue4","mediumslateblue","mediumspringgreen","darkorchid4","midnightblue","rosybrown1")
 mycols2<-mycols[1:ncol(N)]
 
-#The number of plots per page needs to be adjusted manually, the rest should work automatically independed of the number of patches, species (up to 15) or running time
+#The number of plots per page needs to be adjusted manually, the rest should work automatically independent of the number of patches, species (up to 15) or running time
 ## Plot of metacommunity population dynamics over time
 par(mfrow=(c(10,5)))
 par(mai=c(0.1,0.1,0.1,0.1),oma=c(2,2,2,1))

### 2025-09-29 code/R/3_main.R

@@ -145,4 +145,4 @@ verbose <- 50*thin
 m.1 <- Hmsc::sampleMcmc(m,samples=samples,transient=transient,nChains=nChains,nParallel=nChains,verbose=verbose,thin=thin,updater = list(GammaEta = FALSE))
 a <- toc()
 save.image(file=file.path(paste("./data/mc/",h_lev[z],d_lev[f],"_hmsc.RData",sep=""))) # 
-tic()
+toc()

### 2025-09-29 code/R/fig2.R

@@ -1,4 +1,4 @@
-## Code for Figure 1
+## Code for Figure 2
 # Author. J.H. Pantel &  R.J. Hermann
 # Date 29.07.2025
 
@@ -15,6 +15,7 @@ library(gridGraphics)
 library(svglite)
 ## functions
 source("./code/R/plot_grad_ylim.R")
+source("./code/R/two_h_cust.R")
 
 ## set random seed
 set.seed(8)
@@ -351,12 +352,12 @@ par(mgp=c(2,0.45,0), tcl=-0.4, mar=c(1.3,1.2,0,0))
 Gradient <- constructGradient(m.3.sample,focalVariable="dx1",non.focalVariables=list(E=list(2),Esq=list(2),dx2=list(2)),ngrid=390)
 Gradient$XDataNew$Esq <- Gradient$XDataNew$E^2
 predY <- predict(m.3.sample,XData=Gradient$XDataNew,expected=TRUE,ranLevels=Gradient$rLNew)
-plot_grad_ylim(m.3.sample,Gradient,pred=predY,showData=T,measure="Y",index=1,main="",xlab="dx_t",ylab="predicted N1_t+1",showPosteriorSupport=FALSE,cex.axis=0.75,mar=c(1,1,0,0),bty='n',auto=0,l1=0,h1=5)
+two_h_cust(m.3.sample,Gradient,pred=predY,showData=T,measure="Y",index=1,main="",xlab="dx_t",ylab="predicted N1_t+1",showPosteriorSupport=FALSE,cex.axis=0.75,mar=c(1,1,0,0),bty='n',auto=0,l1=0,h1=5)
 p9 <- recordPlot()
 Gradient <- constructGradient(m.3.sample,focalVariable="E",non.focalVariables=list(Esq=list(2)),ngrid=390)
 Gradient$XDataNew$Esq <- Gradient$XDataNew$E^2
 predY <- predict(m.3.sample,XData=Gradient$XDataNew,expected=TRUE,,ranLevels=Gradient$rLNew)
-plot_grad_ylim(m.3.sample,Gradient,pred=predY,showData=T,measure="Y",index=2,main="",xlab="E_t",ylab="predicted N2_t+1",showPosteriorSupport=FALSE,cex.axis=0.75,mar=c(1,1,0,0),bty='n',auto=0,l1=0,h1=5)
[…]

### 2025-09-29 code/R/fig3.R

@@ -1,4 +1,4 @@
-## Code for Figure 2
+## Code for Figure 3
 # Author. J.H. Pantel &  R.J. Hermann
 # Date 29.07.2025
 
@@ -9,7 +9,7 @@ library(Hmsc)
 library(vegan)
 library(cowplot)
 
-#### Figure 2a. Alpha diversity across d and h2 levels ####
+#### Figure 3a. Alpha diversity across d and h2 levels ####
 ## functions
 CV <- function(x){
   a <- mean(x,na.rm=T)	#Get the average of all values
@@ -73,7 +73,7 @@ sub2 <- sub[(sub$h2==0 | sub$h2==0.1 | sub$h2==0.3 | sub$h2==0.9),]
   
 p0 <- ggplot(sub2,aes(x=d_lev,y=mean)) + geom_point(aes(colour=h2)) + scale_colour_gradient(low="lightgrey",high="forestgreen")+ geom_line(aes(group=interaction(level,h_lev)),col="black",linewidth=0.1) + coord_cartesian(ylim=c(0,15)) + theme_classic() + ylab(expression(gamma*" diversity")) + xlab("dispersal") + scale_x_discrete(labels=c(0,expression(10^-9),expression(10^-8),expression(10^-7),expression(10^-6),expression(10^-5),expression(10^-4),expression(10^-3),expression(10^-2),0.1,0.2,0.3,0.4,0.5,0.6,0.7,0.8,0.9,1))
 
-#### Figure 2c. Stacked bar chart of mean VarPart results across d and h2 levels tested ####
+#### Figure 3c. Stacked bar chart of mean VarPart results across d and h2 levels tested ####
 ## Enter VarPart data
 data_bar <- data.frame(d=rep(NA,9),h2=rep(NA,9),Percentage=NA,level=rep(c("Env","deltaX","R(site)","R(time)"),times=9),col_col=rep(c("white","skyblue","darkgrey","orange"),times=9))
 
@@ -165,7 +165,7 @@ p3 <- ggplot(sub3,aes(y = Percentage, x = d, fill = as.character(level))) +
         axis.text.y = element_blank(), 
         axis.ticks.y = element_blank(),legend.position=c(0.8,0.5),legend.text = element_text(size=7),legend.title = element_blank())
 
-#### Figure 2b. Conceptual MC figure ####
[…]

### 2025-09-29 code/R/figs_A4.R

@@ -6,6 +6,7 @@ library(Hmsc)
 library(tidyverse)
 library(patchwork)
 library(bayesplot)
+bayesplot_theme_set(new = theme_default())
 library(reshape2)
 library(cowplot)
 library(MASS)

### 2025-09-29 code/R/figure_box1.R

@@ -13,13 +13,13 @@ load("data/mc/h_01_d_minus3_hmsc.RData")
 m.post = Hmsc::convertToCodaObject(m.1)
 m.df <- as.data.frame(do.call(rbind, m.post$Beta))
 
-mycols <- c("maroon4","coral3","peachpuff3","gray47","hotpink2","darkorange","cyan3","khaki1","firebrick1","khaki1","mediumslateblue","mediumspringgreen","darkorchid4","khaki1","rosybrown1")
+mycols <- c("maroon4","coral3","peachpuff3","gray47","hotpink2","darkorange","cyan3","azure2","firebrick1","azure2","mediumslateblue","mediumspringgreen","darkorchid4","azure2","rosybrown1")
 ############################################################
 #Checking the beta effect of species 15 on 12
 #With pre-checking we could find that both species only co-occur in patches patch 4,7,41, 46
 par(mai=c(0,0,0,0),oma=c(2,2,0.5,0.5),mar=c(0,0,0,0))
 #For beta of |deltax| species 15 on species 12
-p1<-bayesplot::mcmc_areas(as.data.frame(m.df$`B[x15 (C14), y12 (S9)]`)) + labs(x="Posterior distribution",y="Density")+
+p1<-bayesplot::mcmc_areas(as.data.frame(m.df$`B[x15 (C14), y12 (S9)]`)) + labs(x="Posterior distribution",y="Density")+theme_classic()+
   theme(axis.ticks.y=element_blank(),axis.text.y=element_blank())+ggtitle(expression("|"~Delta~"x12,15|"))
 
 for(j in c(4)){
@@ -32,6 +32,9 @@ for(j in c(4)){
     points(N[j,i,],col=mycols[i])
     text(x=(dim(N)[3]),y=(max(N,na.rm=T)/5),labels=j,pos=2,col="black")
   }
+  for(i in c(12,15)){
+    points(N[j,i,],col=mycols[i])
+  }
 }
 p2 <- recordPlot()
 for(j in c(7)){
@@ -43,6 +46,9 @@ for(j in c(7)){
   for(i in 1:ncol(N)){
     points(N[j,i,],col=mycols[i])
     text(x=(dim(N)[3]),y=(max(N,na.rm=T)/5),labels=j,pos=2,col="black")
+  }
+  for(i in c(12,15)){
+    points(N[j,i,],col=mycols[i])
   }
 }
 p3 <- recordPlot()
[…]

### 2025-09-29 code/R/lg2_mods.R

@@ -1,78 +1,0 @@
-#' @name lg2_mods
-#' @aliases disc_log
-#' @aliases disc_log_E
-#' @aliases disc_LV_comp
-#' @aliases disc_LV_E
-#' @aliases disc_LV_evol
-#' @title Discrete-time logistic population growth, with options for competition, environmental covariate, and trait evolution
-#'
-#' @description
-#' disc_log and disc_log_E are functions for logistic growth in a single species, with user-provided growth rate, initial population size, and strength of intraspecific competition. The function disc_log_E includes an environmental covariate, and growth rate now depends on this covariate as a function of the distance from the optimal trait value. XX disc_LV_comp XX disc_LV_E XX disc_LV_evol XX
-#'
-#' @param r the user-supplied population growth rate (when the species trait x is at the environmental optimum)
-#' @param N0 the initial population size
-#' @param alpha the intraspecific interaction coefficient
-#' @param E the initial value of the environmental covariate
-#' @param x the initial value of the species trait
-#' @param P xx
-#' @param w xx
-#' @param Wmax xx
-#' @param h2 xx
-#' @return xx
-#'
-#' @examples
-#' # Simulate initial species population growth
-#' N1.0 <- rpois(1,10)
-#' r1.0 <- 1.67
-#' alpha.11 <- 0.00125
-#' # Simulation of model for t time steps
-#' t <- 30
-#' N <- rep(NA, t)
-#' N[1] <- N1.0
-#' for (i in 2:t) {
-#'   N[i] <- disc_log(r=r1.0, N0=N[i-1], alpha=alpha.11)
-#' }
-#'
-#' @rdname lg2_mods
-#' @export
-disc_log <- function(r, N0, alpha) {
-  Nt1 <- (r*N0) / (1+alpha*N0)
-  return(Nt1)
-}
-
-#' @rdname lg2_mods
-#' @export
-disc_log_E <- function(r, N0, alpha, E, x) {
-  Nt1 <- ((r*exp(-(E-x)^2))*N0) / (1+alpha*N0)
-  return(Nt1)
-}
-
-#' @rdname lg2_mods
-#' @export
-disc_LV_comp <- function(r,N0,alpha) {
-  num <- r*N0
[…]

### 2025-09-29 code/R/plot_grad_ylim.R

@@ -2,7 +2,7 @@
 ## 08-11-2024
 ## plotGradient from HMSC, with control for y-limit
 library(abind)
-plot_grad_ylim <- function (hM, Gradient, predY, measure, xlabel = NULL, ylabel = NULL, index = 1, q = c(0.025, 0.5, 0.975), cicol = rgb(0, 0, 1, alpha = 0.5), pointcol = "lightgrey", pointsize = 1, showData = FALSE, jigger = 0, yshow = NA, showPosteriorSupport = TRUE, auto=1,l1=1,h1=1,main,...) 
+plot_grad_ylim <- function (hM, Gradient, predY, measure, xlabel = NULL, ylabel = NULL, index = 1, q = c(0.025, 0.5, 0.975), cicol = rgb(0, 0, 1, alpha = 0.5), pointcol = "lightgrey", pointsize = 0.5, showData = FALSE, jigger = 0, yshow = NA, showPosteriorSupport = TRUE, auto=1,l1=1,h1=1,main,...) 
 {
   Pr = NA
   if (is.null(xlabel)) {

### 2025-09-29 code/R/two_h_cust.R

@@ -1,0 +1,359 @@
+## J.H. Pantel
+## 25-09-2025
+## Gradient plot from HMSC, with control for y-limit, modified for 10-site case
+library(abind)
+two_h_cust <- function (hM, Gradient, predY, measure, xlabel = NULL, ylabel = NULL, index = 1, q = c(0.025, 0.5, 0.975), cicol = rgb(0, 0, 1, alpha = 0.5), pointcol = "lightgrey", pointsize = 1, showData = FALSE, jigger = 0, yshow = NA, showPosteriorSupport = TRUE, auto=1,l1=1,h1=1,main,...) 
+{
+  Pr = NA
+  if (is.null(xlabel)) {
+    switch(class(hM$X)[1L], matrix = {
+      xlabel = colnames(Gradient$XDataNew)[[1]]
+    }, list = {
+      xlabel = colnames(Gradient$XDataNew[[1]])[[1]]
+    })
+  }
+  switch(class(hM$X)[1L], matrix = {
+    xx = Gradient$XDataNew[, 1]
+  }, list = {
+    if (measure == "Y") {
+      xx = Gradient$XDataNew[[index]][, 1]
+    } else {
+      xx = Gradient$XDataNew[[1]][, 1]
+    }
+  })
+  ngrid = length(xx)
+  if (measure == "S") {
+    predS = abind(lapply(predY, rowSums), along = 2)
+    Pr = mean(predS[ngrid, ] > predS[1, ])
+    qpred = apply(predS, c(1), quantile, probs = q, na.rm = TRUE)
+    if (is.null(ylabel)) {
+      ylabel = "Summed response"
+      if (all(hM$distr[, 1] == 2)) {
+        ylabel = "Species richness"
+      }
+      if (all(hM$distr[, 1] == 3)) {
+        ylabel = "Total count"
+      }
+    }
+  }
+  if (measure == "Y") {
+    tmp = abind(predY, along = 3)
+    Pr = mean(tmp[ngrid, index, ] > tmp[1, index, ])
+    qpred = apply(tmp, c(1, 2), quantile, probs = q, na.rm = TRUE)
+    qpred = qpred[, , index]
+    if (is.null(ylabel)) {
+      ylabel = hM$spNames[[index]]
+    }
+  }
+  if (measure == "T") {
+    if (all(hM$distr[, 1] == 1)) {
+      predT = lapply(predY, function(a) (exp(a) %*% hM$Tr)/matrix(rep(rowSums(exp(a)), 
[…]

### 2025-09-29 manuscript/Appendix_S1.Rmd
Rename Appendix code files

--- 2024-11-26 manuscript/Appendix_S1_v02.Rmd
@@
   bookdown::pdf_document2:
-    toc: no
+    toc: yes
     fig_caption: yes
@@
 library(bayesplot)
+bayesplot_theme_set(new = theme_default())
 library(reshape2)
@@
 library(Hmisc)
+library(cowplot)
 ```
@@
 We fit the population time series data to a first-order auto-regressive model to predict $N_{t+1}$ as a function of $N_t$, and compare that to a linear regression.
-We use ln-N after @Ives1995, as also discussed in @Certain2018 and @Olivenca2021.
+We use ln(N) after @Ives1995, as also discussed in @Certain2018 and @Olivenca2021.
 
@@
 ```
-
-We can already see that the linear model does not provide a useful fit to the data points (neither does taking the inverse function help).
+In the concluding part, we will compare these linear model results to a Bayesian linear model.
+
+```{r}
+# prepare data for HMSC, run later
+dat <- as.data.frame(log(N))
+dat$time <- 1:t
+df <- data.frame(dat[(2:t),-3])
+Y <- as.matrix(df)
+XData <- data.frame(dat[1:(t-1),-3])
+colnames(XData) <- c(paste0("n",1:length(r)))
+m.3a1.hmsc = Hmsc(Y = Y[,1], XData = XData, XFormula = ~n1+n2)
+m.3a2.hmsc = Hmsc(Y = Y[,2], XData = XData, XFormula = ~n1+n2)
+```
+
+We can already see that the linear model provides a useful predictive fit to the data points, but the regression parameters cannot capture the complex relationship between species densities (neither does taking the inverse function help).
 The regression parameters can predict well the population size of the other species, but they are not informative for inference.
@@
 
-The HMSC model correctly shows the negative association between the species (shown here as correlations - JSDM CHapter 7 tells us the intra-specific correlation is always 1).
+The HMSC model correctly shows the negative association between the species (shown here as correlations - JSDM Chapter 7 tells us the intra-specific correlation is always 1).
 
@@
 ```
+
 ```{r figs20, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs20} Observed (points) and HMSC model-predicted population size (logarithm) for Species 1 (black) and Species 2 (orange), using environment $E_t$ as a fixed effect predictor and specifying temporal random effects to capture the species interactions.
@@
 ```
+In the concluding part, we will compare Bayesian linear model results.
+
+```{r}
+# prepare data for HMSC, run later
+m.4a1.hmsc = Hmsc(Y = Y[,1], XData = XData, XFormula = ~X1+X2)
+m.4a2.hmsc = Hmsc(Y = Y[,2], XData = XData, XFormula = ~X1+X2)
+```
 
@@
 
+## Conclusions
+
+Residual species associations reflect the magnitude of species interaction strength, and the model implements these random effects to correctly predict population density.
+
 # Model D: Species growth, competition, with environmental change
 
+## Environmental covariate + other elements
 The goal of the process was to use HMSC's latent variable approach to study the results of non-linear processes in ecology using a linear model.
@@
 ```
+
+## HMSC statistical model
+
 ```{r}
@@
 ```
+
 We can see that the Environment has an overall negative impact for Species 1 as $E$ drifts randomly towards more positive values (and thus further from Species 1 trait value).
@@
 plotVariancePartitioning(m.5.sample,VP,args.legend=list(cex=0.75,bg="transparent"))
+```
+
+```{r}
+# prepare data for HMSC, run later
+m.5a1.hmsc = Hmsc(Y = Y[,1], XData = XData, XFormula = ~n1+n2+E+Esq)
+m.5a2.hmsc = Hmsc(Y = Y[,2], XData = XData, XFormula = ~n1+n2+E+Esq)
 ```
@@
 ```
+
+## Conclusions
 With the reduced impact of the environment, we now see the strong negative associations between the two competing species.
@@
 
+## HMSC statistical model
 We now use the species abundances as the response variable in the model, with $E$, $E^2$, and \( \lvert x_{t+1} - x_{i,t} \rvert \) as fixed effects, and time points as a random effect (to estimate species-to-species associations).
@@
 
+## Conclusions
 The analysis indicates that we can successfully use the HMSC analysis to partition drivers of species abundances that include evolving traits.
@@
 
+## Correlated dynamics in multiple patches
 We now consider that trait evolution can occur, with an impact on the species growth rate / fitness in the context of the environment.
@@
 ```
+
+## HMSC statistical model
 We now fit a spatial random effect, and add that to the HMSC model.
@@
 
-We observe a much greater fraction of variation explained by the spatial random effects.
+## Conclusions
+
+We observe a much greater fraction of variation explained by the spatial random effects when the degree of spatial autocorrelation is higher, as expected.
+
+# Summaries across all models
+
+```{r}
+### Model A, coefficient posteriors and R2
+summar <- rbind(m.post.hmsc$Beta[[1]],m.post.hmsc$Beta[[2]])
+mean(summar[,2])
+quantile(summar[,2],probs=c(0.05,0.95))
+# Evaluate model fit
+preds = computePredictedValues(m.1.sample)
+MF = evaluateModelFit(hM=m.1.sample, predY=preds)
+MF
+```
+```{r}
+### Model B, coefficient posteriors and R2
+summar <- rbind(m2.post.hmsc$Beta[[1]],m2.post.hmsc$Beta[[2]])
+# N
+mean(summar[,2])
+quantile(summar[,2],probs=c(0.05,0.95))
+# E
+mean(summar[,3])
+quantile(summar[,3],probs=c(0.05,0.95))
+# E^2
+mean(summar[,4])
+quantile(summar[,4],probs=c(0.05,0.95))
+# Evaluate model fit
+preds = computePredictedValues(m.2.sample)
+MF = evaluateModelFit(hM=m.2.sample, predY=preds)
+MF
+```
+```{r}
+### Model C, Scenario 1, coefficient posteriors and R2
+# Bayesian model parameters
+nChains <- 2
+thin <- 5
+samples <- 1000
+transient <- 500*thin
+verbose <- 500*thin
+# sample MCMC
+m.3a1.sample <- sampleMcmc(m.3a1.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
+m.3a2.sample <- sampleMcmc(m.3a2.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
+## N1 model
+m3a1.post.hmsc <- convertToCodaObject(m.3a1.sample)
+# Summary statistics for main manuscript text
+summar <- rbind(m3a1.post.hmsc$Beta[[1]],m3a1.post.hmsc$Beta[[2]])
+# N1
+mean(summar[,2])
+quantile(summar[,2],probs=c(0.05,0.95))
+# N2
+mean(summar[,3])
+quantile(summar[,3],probs=c(0.05,0.95))
+# Evaluate model fit
+preds = computePredictedValues(m.3a1.sample)
+MF = evaluateModelFit(hM=m.3a1.sample, predY=preds)
+MF
+
+## N2 model
+m3a2.post.hmsc <- convertToCodaObject(m.3a2.sample)
+# Summary statistics for main manuscript text
+summar <- rbind(m3a2.post.hmsc$Beta[[1]],m3a2.post.hmsc$Beta[[2]])
+# N1
+mean(summar[,2])
+quantile(summar[,2],probs=c(0.05,0.95))
+# N2
+mean(summar[,3])
+quantile(summar[,3],probs=c(0.05,0.95))
+# Evaluate model fit
+preds = computePredictedValues(m.3a2.sample)
+MF = evaluateModelFit(hM=m.3a2.sample, predY=preds)
+MF
+```
+```{r}
+### Model C, Scenario 2, coefficient posteriors and R2
+# sample MCMC
+m.4a1.sample <- sampleMcmc(m.4a1.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
+m.4a2.sample <- sampleMcmc(m.4a2.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
+# Summary properties for main manuscript text
+preds = computePredictedValues(m.4a1.sample)
+MF = evaluateModelFit(hM=m.4a1.sample, predY=preds)
+MF
+# Summary properties for main manuscript text
+preds = computePredictedValues(m.4a2.sample)
+MF = evaluateModelFit(hM=m.4a2.sample, predY=preds)
+MF
+```
+
+```{r}
+### Model D, Scenario 1, coefficient posteriors and R2
+# sample MCMC
+m.5a1.sample <- sampleMcmc(m.5a1.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
+m.5a2.sample <- sampleMcmc(m.5a2.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
+# Summary properties for main manuscript text
+preds = computePredictedValues(m.5a1.sample)
+MF = evaluateModelFit(hM=m.5a1.sample, predY=preds)
+MF
+# Summary properties for main manuscript text
+preds = computePredictedValues(m.5a2.sample)
+MF = evaluateModelFit(hM=m.5a2.sample, predY=preds)
+MF
+```
+
+```{r echo=FALSE}
+#############################################################
+# 1. Sim: One species, logistic growth 
+set.seed(42)
+# Simulate initial species population growth
+N1.0 <- 10
+r1.0 <- 1.67
+alpha.11 <- 0.00125
+# Simulation of model for t time steps
+t <- 30
+N <- rep(NA, t)
+
+N[1] <- N1.0
+for (i in 2:t) {
+  N[i] <- disc_log(r = r1.0, N0 = N[i - 1], alpha = alpha.11)
+}
+
+dat <- as.data.frame(N)
+dat$time <- 1:t
+p1 <- ggplot2::ggplot(dat, aes(time, log(N))) + geom_point(color="skyblue") + theme(plot.margin = unit(c(0, 0, 0, 0), "cm"),panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background = element_blank(), axis.line = element_line(colour = "black")) +theme(legend.position = "none",strip.text = element_text(size=6)) + ylab("log(N)") + ylim(c(0,6.5))
+
+#+ geom_hline(yintercept = log(((r1.0 - 1)/alpha.11)), linetype = "dashed", color = "gray")
+#############################################################################
+############################################################################
+###########################################################################
+# 2. Sim: Two species, logistic growth + environmental covariate ---------
+N1.0 <- 10
+r1.0 <- 1.67
+alpha.11 <- 0.00125
+E.0 <- 0.8
+x1.0 <- 0.8
+# Simulation of model for t time steps
+t <- 40
+N <- rep(NA, t)
+N[1] <- N1.0
+E <- rep(NA, t)
+E[1] <- E.0
+for (i in 2:t) {
+  N[i] <- disc_log_E(r = r1.0, N0 = N[i - 1], alpha = alpha.11, E = E[i - 1], x = x1.0)
+  E[i] <- E[i - 1] + rnorm(1, 0, 0.1)
+}
+# Plot simulation: ggplot
+dat <- as.data.frame(cbind(N,E))
+dat$time <- 1:t
+p2 <- ggplot(dat, aes(time, log(N))) + geom_point(color="skyblue") + theme(plot.margin = unit(c(0, 0, 0, 0), "cm"),panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background = element_blank(), axis.line = element_line(colour = "black")) +theme(legend.position = "none",strip.text = element_text(size=6)) + ylab("log(N)") + ylim(c(0,6.5))
+
+#############################################################################################
+####################################################################################
+##################################################################################
+# 3. Sim: Two species, logistic growth + competition
+# Initial conditions
+N0 <- c(10, 10)
+r <- c(1.67, 1.7)
+alpha.11 <- 0.00125
+alpha.22 <- 0.00125
+alpha.12 <- 0.008
+alpha.21 <- 0.008725
+alpha <- matrix(c(alpha.11, alpha.21, alpha.12, alpha.22), nrow = 2, byrow = FALSE)
+# make sure the interaction matrix is correct Simulation of model for t time
+# steps
+t <- 40
+N <- array(NA, dim = c(t, 2))
+N <- as.data.frame(N)
+colnames(N) <- c("N1", "N2")
+N[1, ] <- N0
+for (i in 2:t) {
+  N[i, ] <- disc_LV_comp(r = r, N0 = N[i - 1, ], alpha = alpha)
+}
+
+N$time <- 1:t
+dat <- melt(N, id.vars="time")
+p3 <- ggplot(dat, aes(time, log(value),color=variable)) + geom_point() + scale_color_manual(values = c("N1" = "skyblue", "N2" = "darkgrey"))+ theme(plot.margin = unit(c(0, 0, 0, 0), "cm"),panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background = element_blank(), axis.line = element_line(colour = "black")) +theme(legend.position = "none",strip.text = element_text(size=6)) + ylab("log(N)") + ylim(c(0,6.5))
+
+#################################################################################################
+##############################################################################################
+#############################################################################################
+# 4. Sim: Two species, logistic growth + competition, environmental covariate ---------
+# Initial conditions
+N0 <- c(10,10)
+r <- c(1.7,1.7)
+alpha.11 <- 0.01
+alpha.22 <- 0.01
+alpha.12 <- 0.005
+alpha.21 <- 0.01
+alpha <- matrix(c(alpha.11,alpha.21,alpha.12,alpha.22),nrow=2,byrow=FALSE) # careful to make sure you have correct interaction matrix
+E.0 <- 0.8
+x <- c(0.6,0.8)
+# Simulation of model for t time steps
+t <- 40
+N <- array(NA,dim=c(t,length(N0)))
+N <- as.data.frame(N)
+colnames(N) <- paste0("N",1:length(N0))
+N[1,] <- N0
+E <- rep(NA, t)
+E[1] <- E.0
+for (i in 2:t) {
+  N[i,] <- disc_LV_E(r=r,N0=N[i-1,],alpha=alpha,E=E[i-1],x=x)
+  E[i] <- E[i-1] + rnorm(1,0,0.1)
+}
+
+### Plot of time series ###
+gdat <- N
+gdat$time <- 1:t
+dat <- melt(gdat, id.vars="time")
+p4 <- ggplot2::ggplot(dat, aes(time, log(value),color=variable)) + geom_point() + scale_color_manual(values = c("N1" = "skyblue", "N2" = "darkgrey"))+ theme(plot.margin = unit(c(0, 0, 0, 0), "cm"),panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background = element_blank(), axis.line = element_line(colour = "black")) +theme(legend.position = "none",strip.text = element_text(size=6)) + ylab("log(N)")+ ylim(c(0,4.5))
+
+#################################################################################################
+##############################################################################################
+#############################################################################################
+# 5. Sim: Two species, logistic growth + competition, environmental covariate, evolution
+# Initial conditions
+N0 <- c(10,10)
+alpha.11 <- 0.01
+alpha.22 <- 0.01
+alpha.12 <- 0.005
+alpha.21 <- 0.01
+alpha <- matrix(c(alpha.11,alpha.21,alpha.12,alpha.22),nrow=2,byrow=FALSE) # careful to make sure you have correct interaction matrix
+E.0 <- 0.8
+x.0 <- c(0.1,0.8)
+P <- 1
+w <- 2
+Wmax <- 2
+h2 <- 1
+k <- (w + (1 - h2) * P)/(P + w)
+# Simulation of model for t time steps
+t <- 40
+N <- array(NA,dim=c(t,length(N0)))
+N <- as.data.frame(N)
+colnames(N) <- paste0("N",1:length(N0))
+N[1,] <- N0
+E <- rep(NA, t)
+E[1] <- E.0
+x <- array(NA,dim=c(t,length(N0)))
+x <- as.data.frame(x)
+colnames(x) <- paste0("x",1:length(N0))
+x[1,] <- x.0
+r <- array(NA,dim=c(t,length(N0)))
+r <- as.data.frame(r)
+colnames(r) <- paste0("r",1:length(N0))
+What <- Wmax*sqrt(w/(P+w))
+r[1,] <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E[1]-x[1,]))^2)/(2*(P+w)))
+for (i in 2:t) {
+  res <- disc_LV_evol(N0=N[i-1,],alpha=alpha,E=E[i-1],x=x[i-1,],P=P,w=w,Wmax=Wmax,h2=h2)
+  N[i,] <- res$Nt1
+  r[i,] <- res$r
+  # trait change
+  d <- E[i-1] - x[i-1,]
+  d1 <- k*d
+  x[i,] <- E[i-1] - d1
+  # environmental change
+  E[i] <- E[i-1] + abs(rnorm(1, 0, 0.05))
+}
+
+### Plot of time series ###
+gdat <- N
+gdat$time <- 1:t
+dat <- melt(gdat, id.vars="time")
+p5 <- ggplot2::ggplot(dat, aes(time, log(value),color=variable)) + geom_point() + scale_color_manual(values = c("N1" = "skyblue", "N2" = "darkgrey"))+ theme(plot.margin = unit(c(0, 0, 0, 0), "cm"),panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background = element_blank(), axis.line = element_line(colour = "black")) +theme(legend.position = "none",strip.text = element_text(size=6)) + ylab("log(N)")+ ylim(c(0,4.5))
+
+#################################################################################################
+##############################################################################################
+#############################################################################################
+# 6. Sim: Two species, logistic growth + competition, environmental covariate in spatially structured environment 
+set.seed(12345)
+# Initial conditions
+# Simulation of model for t time steps, i sites
+N0 <- c(10,10)
+E.0 <- 0.8
+x.0 <- c(0.1,0.8)
+alpha.11 <- 0.01
+alpha.22 <- 0.01
+alpha.12 <- 0.005
+alpha.21 <- 0.01
+alpha <- matrix(c(alpha.11,alpha.21,alpha.12,alpha.22),nrow=2,byrow=FALSE) # careful to make sure you have correct interaction matrix
+P <- 1
+w <- 2
+Wmax <- 2
+h2 <- 1
+k <- (w + (1 - h2) * P)/(P + w)
+j <- 10
+# random site locations
+xycoords = matrix(runif(2*j), ncol = 2)
+rownames(xycoords) <- 1:j
+t <- 40
+N <- array(NA,dim=c(j*t,length(N0)))
+N <- as.data.frame(N)
+colnames(N) <- paste0("N",1:length(N0))
+x <- array(NA,dim=c(j*t,length(N0)))
+x <- as.data.frame(x)
+colnames(x) <- paste0("x",1:length(N0))
+r <- array(NA,dim=c(j*t,length(N0)))
+r <- as.data.frame(r)
+colnames(r) <- paste0("r",1:length(N0))
+N[1:j,] <- rep(N0,each=j)
+E <- rep(NA, t)
+E[1] <- E.0
+x[1:j,] <- rep(x.0,each=j)
+What <- Wmax*sqrt(w/(P+w))
+r0 <- as.numeric(What*exp((-(((w+(1-h2)*P)/(P+w))*(E[1]-x[1,]))^2)/(2*(P+w))))
+r[1:j,] <- rep(r0,each=j)
+# Keep track of site and time
+study <- array(NA,dim=c(j*t,3))
+study <- as.data.frame(study)
+colnames(study) <- c("obs","site","time")
+study$obs <- 1:dim(study)[1]
+study$site <- rep(1:j,times=t)
+study$time <- rep(1:t,each=j)
+
+for (i in 2:t) {
+  for(z in 1:j){
+    res <- disc_LV_evol(N0=N[study$site==z & study$time==i-1,],alpha=alpha,E=E[i-1],x=x[study$site==z & study$time==i-1,],P=P,w=w,Wmax=Wmax,h2=h2)
+    N[study$site==z & study$time==i,] <- res$Nt1
+    r[study$site==z & study$time==i,] <- res$r
+    # trait change
+    d <- E[i-1] - x[study$site==z & study$time==i-1,]
+    d1 <- k*d
+    x[study$site==z & study$time==i,] <- E[i-1] - d1
+  }
+  # environmental change
+  E[i] <- E[i-1] + abs(rnorm(1, 0, 0.05))
+}
+# site-level random effect
+sigma <- 0
+sigma.spatial <- 10
+alpha.spatial <- 0.5
+Sigma = sigma.spatial^2*exp(-as.matrix(dist(xycoords))/alpha.spatial)
+# draw from covariance matrix
+a = mvrnorm(mu=rep(0,j), Sigma = Sigma)
+for(i in 1:t){
+  N$N1[study$time==i] <- N$N1[study$time==i] + a
+}
+a = mvrnorm(mu=rep(0,j), Sigma = Sigma)
+for(i in 1:t){
+  N$N2[study$time==i] <- N$N2[study$time==i] + a
+}
+
+### Plot of time series ###
+gdat <- N
+gdat$site <- study$site
+gdat$time <- study$time
+dat <- melt(gdat, id.vars=c("site","time"))
+p6 <- ggplot2::ggplot(dat, aes(time, log(value),color=variable)) + geom_point() + scale_color_manual(values = c("N1" = "skyblue", "N2" = "darkgrey"))+ theme(plot.margin = unit(c(0, 0, 0, 0), "cm"),panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background = element_blank(), axis.line = element_line(colour = "black")) +theme(legend.position = "none",strip.text = element_text(size=6)) + ylab("log(N)")+ facet_wrap(~site,nrow=2)+ ylim(c(0,4.5))
+
+d<-plot_grid(p1,p2,p3,p4,p5,p6,align="h",axis="b",labels = c('a','b','c','d','e','f'),label_size = 9,hjust=c(-2,-2,-2,-2,-2,-2),rel_widths=c(1,1,1.5))
+
+d
+```
 

### 2025-09-29 manuscript/Appendix_S2.Rmd
Rename Appendix code files

--- 2024-11-30 manuscript/Appendix_S2_v02.Rmd
@@
 author:
-  - Jelena H. Pantel^[Laboratoire Chrono-environnement,UMR 6249 CNRS-UFC, 16 Route de Gray, 25030 Besançon cedex, France, jelena.pantel@univ-fcomte.fr]
-  - Ruben J. Hermann^[University of Duisburg-Essen, Universitätsstraße 5, 45141 Essen, Germany, ruben.hermann@uni-due.de]
+- "Jelena H. Pantel^[Laboratoire Chrono-environnement,UMR 6249 CNRS-UFC, 16 Route
+  de Gray, 25030 Besançon cedex, France, jelena.pantel@univ-fcomte.fr]"
+- "Ruben J. Hermann^[University of Konstanz, Universitätsstraße 10, 78464 Konstanz,
+  Germany, ruben.hermann@posteo.de]"
 date: "`r format(Sys.time(), '%d %B, %Y')`"

### 2025-09-29 manuscript/Appendix_S3.Rmd
Rename Appendix code files

--- 2024-11-30 manuscript/Appendix_S3_v01.Rmd
@@
 library(bayesplot)
+bayesplot_theme_set(new = theme_default())
 ```
@@
 ```{r}
+set.seed(10403)
 ### 1 species, h2=0 ###
@@
 
-```{r warning=FALSE}
+```{r, warning=FALSE,message=FALSE,results = 'hide'}
 VP1 <- computeVariancePartitioning(m.1.sample,group=c(1,1,1,2),groupnames=c("Env","x1"))
@@
 ```
-```{r figs2, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs2} Simulated (dark blue points) population size over time for $h^2 = 0$, with posterior predicted mean (light blue points) and 95% highest density intervals (light blue bars)."}
+```{r figs2, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs2} Simulated (dark blue points) population size over time for $h^2 = 0$, with posterior predicted mean (light blue points) and 95% highest density intervals (light blue bars).", warning=FALSE}
 # prepare for plot
@@
 ```
+
 ```{r}

### 2025-09-29 manuscript/Appendix_S5.Rmd
Rename Appendix code files

---
title: "Appendix S5. Heritability and HMSC"
author:
- "Jelena H. Pantel^[Laboratoire Chrono-environnement,UMR 6249 CNRS-UFC, 16 Route
  de Gray, 25030 Besançon cedex, France, jelena.pantel@univ-fcomte.fr]"
- "Ruben J. Hermann^[University of Konstanz, Universitätsstraße 10, 78464 Konstanz,
  Germany, ruben.hermann@posteo.de]"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::pdf_document2:
    toc: no
    fig_caption: yes
    latex_engine: xelatex
    extra_dependencies: ["float"]
  bookdown::html_document2: default
  word_document:
    reference_docx: docx_template.docx
header-includes:
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \fancyhead[L]{Author Pantel and Hermann}
- \fancyhead[R]{Detection of ecoevo dynamics}
- \usepackage{lineno}
- \linenumbers
link-citations: yes
linkcolor: blue
csl: the-american-naturalist.csl
bibliography: refs.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
```{r setup,echo=FALSE,message=FALSE,warning=FALSE}
library(tinyplot)
library(ggplot2)
library(statgenSTA)
library(statgenGxE)
library(here)
library(tinyplot)
knitr::opts_knit$set(root.dir = here())
```
```{r,echo=FALSE,message=FALSE,warning=FALSE}
# Wrap long lines
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy=TRUE,out.width="50%")
```

# Broad-sense heritability estimates

The work of [@Jewell2023] includes a reciprocal transplant experiment, which can be used to separate genetic phenotypic variance from other sources (i.e. environmental plasticity). With that, we can include genetic (and in principal environmental and residual plastic) components of trait variation as predictors in the HMSC models.

The quantitative trait of frond area has the following distributions by species:

```{r figs1, fig.pos = "H", fig.show="hold", fig.cap = "\\label{fig:figs1} Distribution of frond area ($mm^2$) measured for four aquatic plant species (Lemna minor (Lm), Spirodela polyrhiza (Sp), Wolffia columbiana (Wc), and Ricciocarpos natans (Rn)) in a reciprocal transplant experiment."}
## Load in reciprocal transplant data
dat_rt <- read.csv("./data/emp/ReciprocalTransplant.csv")
## Community and Environment as factors
dat_rt$C.Env <- as.factor(dat_rt$C.Env)
dat_rt$E.Env <- as.factor(dat_rt$E.Env)
## Quantitative trait distributions
tinytheme("ipsum")
plt(~FrondArea, type="density", data=dat_rt, facet=~Species)
```
We can construct a reaction norm for each species, by considering the mean phenotype for each of nine populations (given in the raw data file as `C.Env`, the environment that the plants were exposed to over 12 weeks) measured in each of the nine environments (given in the raw data file as `E.Env`). The authors of [@Jewell2023] consider two populations, Derived and Ancestral, and two environments, Modified and Original.

```{r figs2, fig.pos = "H", warning=FALSE,fig.show="hold", fig.cap = "\\label{fig:figs1} Reaction norm for frond area ($mm^2$) of two populations (Ancestral ad Derived) measured in two environments (Modified and Original), for four aquatic plant species."}
ggplot(dat_rt,aes(x=E.Type,y=FrondArea,color=C.Type)) +
  geom_point(aes(group = C.Type),position=position_jitterdodge(0.1, dodge.width = .2),alpha=0.3,size=0.2) +
  stat_summary(aes(group = C.Type),position=position_jitterdodge(0.1, dodge.width = .2), fun.y = mean, geom = "path") +
  stat_summary(aes(color = C.Type),position=position_jitterdodge(0.1, dodge.width = .2), fun.y = mean, geom = "point", size = 2) +
  facet_wrap(~Species,scales="free") +
  theme_light()
```
We instead used the nine populations and environments, with awareness the experiment was not designed to be balanced across these levels. For example, plants that evolved in environment 9 (low light, high nutrients) are considered as population 9. Individuals from this population were placed in reciprocal transplant experiments only in environment 5 (medium light, medium nutrients; considered to be a match to the ancestral environment in the lake where all experimental plants are sourced) and environment 9. Because the environmental treatments vary across three light levels (Low = 3%, Medium = 12%, High = 50%) and three nutrient treatments (dissolved nitrogen and phosphorus, DN, and DP; Low: DN = 200 $\mu g L^{-1}$, DP = 10 $\mu g L^{-1}$; Medium: DN = 800 $\mu g L^{-1}$, DP = 40 $\mu g L^{-1}$; High: 3200 $\mu g L^{-1}$, DP = 160 $\mu g L^{-1}$), we create a single environmental value as the mean phenotype value of all individuals measured in each of the nine environments [@Falconer1989] [@Stearns1992].

```{r figs3, fig.pos = "H", fig.show="hold", fig.cap = "\\label{fig:figs1} Reaction norm of frond area ($mm^2$) for nine populations (Nutrients: low, medium, high; Light: low, medium, high) measured in nine environments (Nutrients: low, medium, high; Light: low, medium, high), for four aquatic plant species.", warning=FALSE}
# Calculate environmental value (page 136, Falconer 1989; cited in Stearns 1992, page 62)
# Following Jewell & Bell 2023, we define environment 1-9
# We calculate the mean across all individuals in each of these environments - this becomes the 'environmental value'
# Create environmental value
dat_rt$Env.Val <- NA
Env.Val <- by(dat_rt,factor(dat_rt$E.Env),function(x){mean(x$FrondArea)})
for(i in 1:length(levels(factor(dat_rt$E.Env)))){
  dat_rt$Env.Val[dat_rt$E.Env==levels(factor(dat_rt$E.Env))[i]] <- as.numeric(Env.Val[i])
}
# Reaction norm plot
ggplot(dat_rt,aes(x=Env.Val,y=FrondArea,color=C.Env)) +
  geom_point(aes(group = C.Env),position=position_jitterdodge(0.1, dodge.width = .2),alpha=0.3,size=0.2) +
  stat_summary(aes(group = C.Env),position=position_jitterdodge(0.1, dodge.width = .2), fun.y = mean, geom = "path") +
  stat_summary(aes(color = C.Env),position=position_jitterdodge(0.1, dodge.width = .2), fun.y = mean, geom = "point", size = 2) +
  facet_wrap(~Species,scales="free") +
  theme_light()
```
We consider the following linear model and corresponding variance components and heritability estimate ($h^2$ after [@Piepho2007]):

$$ y = \beta_0 + \beta_1 G + \beta_2 E + \beta_3 G \times E + \epsilon $$

$$ h^2 = \frac{\sigma^2_G}{\sigma^2_G + \frac{\sigma^2_{GE}}{m} + \frac{\sigma^2}{mr}} $$
Where $G$ is the effect of the genetic component (population in this case), $E$ is the effect of environmental value, $G \times E$ is their interaction, $m$ is the number of environmental levels, and $r$ is the number of replicates. We implemented a mixed model, with $E$ as a fixed effect, and $G$ and $G \times E$ as random effects. The mixed model and estimation of variance terms were carried out using R package `statgenGxE` [@vanRossum2025] (note that the model places the environment term with the residual term).

```{r message=FALSE}
## Calculate phenotypic variance components in a linear mixed model
dat_rt$Env.Val <- as.factor(dat_rt$Env.Val)
dat_rt$Env.Val <- as.ordered(dat_rt$Env.Val)
Wc <- subset(dat_rt,grepl("Wc", Species))
Sp <- subset(dat_rt,grepl("Sp", Species))
Lm <- subset(dat_rt,grepl("Lm", Species))
Rn <- subset(dat_rt,grepl("Rn", Species))

## Broad-sense heritability estimates by species
#Wc
Wc.TD <- statgenSTA::createTD(data = Wc, genotype = "C.Env", trial = "Env.Val")
drops.Wc <- gxeVarComp(TD = Wc.TD, trait = "FrondArea")
var.Wc <- vc(drops.Wc)
h2.Wc <- var.Wc$Component[1] / (var.Wc$Component[1] + (var.Wc$Component[2] / 9) + (var.Wc$Component[3]/(9*2)))
#Sp
Sp.TD <- statgenSTA::createTD(data = Sp, genotype = "C.Env", trial = "Env.Val")
drops.Sp <- gxeVarComp(TD = Sp.TD, trait = "FrondArea")
var.Sp <- vc(drops.Sp)
h2.Sp <- var.Sp$Component[1] / (var.Sp$Component[1] + (var.Sp$Component[2] / 9) + (var.Sp$Component[3]/(9*2)))
#Lm
Lm.TD <- statgenSTA::createTD(data = Lm, genotype = "C.Env", trial = "Env.Val")
drops.Lm <- gxeVarComp(TD = Lm.TD, trait = "FrondArea")
var.Lm <- vc(drops.Lm)
h2.Lm <- var.Lm$Component[1] / (var.Lm$Component[1] + (var.Lm$Component[2] / 9) + (var.Lm$Component[3]/(9*2)))
#Rn
Rn.TD <- statgenSTA::createTD(data = Rn, genotype = "C.Env", trial = "Env.Val")
drops.Rn <- gxeVarComp(TD = Rn.TD, trait = "FrondArea")
var.Rn <- vc(drops.Rn)
h2.Rn <- var.Rn$Component[1] / (var.Rn$Component[1] + (var.Rn$Component[2] / 9) + (var.Rn$Component[3]/(9*2)))
```

We can now implement these values into the HMSC, to estimate the impact of trait evolution (as well as other predictors) for species abundances.

# References

### 2025-09-29 manuscript/Box1.Rmd
Rename Appendix code files

---
title: "Box 1: Interpreting the impacts of trait evolution"
author:
- "Jelena H. Pantel^[Laboratoire Chrono-environnement,UMR 6249 CNRS-UFC, 16 Route
  de Gray, 25030 Besançon cedex, France, jelena.pantel@univ-fcomte.fr]"
- "Ruben J. Hermann^[University of Konstanz, Universitätsstraße 10, 78464 Konstanz,
  Germany, ruben.hermann@posteo.de]"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::pdf_document2:
    toc: no
    fig_caption: yes
    latex_engine: xelatex
    extra_dependencies: float
    number_sections: yes
    keep_tex: true
  bookdown::html_document2: default
  pdf_document:
    toc: no
  word_document:
    reference_docx: docx_template.docx
  html_document:
    toc: no
    df_print: paged
header-includes:
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \fancyhead[L]{Author Pantel and Hermann}
- \fancyhead[R]{Detection of ecoevo dynamics}
- \usepackage{lineno}
- \linenumbers
linestretch: 2
link-citations: yes
linkcolor: blue
csl: "the-american-naturalist.csl"
bibliography: refs.bib
---

```{r setup,echo=FALSE,message=FALSE}
#library(ecoevor)
library(tidyverse)
library(Hmsc)
library(here)
library(knitr)
library(kableExtra)
library(formatR)
library(rmarkdown)
knitr::opts_knit$set(root.dir = here())
```

```{r,echo=FALSE,message=FALSE,warning=FALSE}
# Wrap long lines
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy=TRUE,out.width="50%")
#Loading HMSC result of the simulation 
```

```{r,echo=FALSE}
load("./data/mc/h_01_d_minus3_hmsc.RData")
m.post = Hmsc::convertToCodaObject(m.1)
m.df <- as.data.frame(do.call(rbind, m.post$Beta))
b1512 <- median(m.df$`B[x15 (C14), y12 (S9)]`)

load("./data/mc/h_09_d_minus3_hmsc.RData")
m.post2 = Hmsc::convertToCodaObject(m.1)
m.df2 <- as.data.frame(do.call(rbind, m.post2$Beta))
b43 <- median(m.df2$`B[x3 (C4), y4 (S3)]`)
```

The HMSC statistical model can estimate the impact of multiple (fixed and random) effects on the measured response variable. Interpretation of these effects, as in any statistical model, requires understanding the potential mechanisms that combine to produce the observations and estimated effect coefficients. In this study, we evaluated the effects of the absolute value in trait change from one time step to the next, $|\Delta x|$, for species densities ($log(N)$). This produces a linear regression coefficient $\beta_{|\Delta x_{ij}|}$, the effect of trait change in species $j$ for density in species $i$. In our metacommunity simulation models, we often observed a strong impact of $\beta_{|\Delta x_{ii}|}$, and in some instances also strong  $\beta_{|\Delta x_{ij}|}$.

We highlight two examples of the metacommunity simulations to consider how to interpret effects of trait evolution: (1) $\beta_{|\Delta x_{12,15}|}$ = `r round(b1512,digits=3)` (influence of $|\Delta x_{15}|$ on $N_{12}$, posterior median) under the simulation conditions $h^2= 0.1$ and dispersal $d = 10^{-3}$ and (2) $\beta_{|\Delta x_{4,3}|}$ = `r round(b43,digits=3)` under $h^2= 0.9$ and dispersal $d = 10^{-3}$. To interpret these effects, we first note that in the mechanistic model we used to simulate growth in a metacommunity, $|\Delta x_j|$ does not directly impact $N_i$. It instead impacts the growth rate and resulting density of species $j$, and this can have an impact for species $i$ through the strength of interspecific competition $\alpha_{ij}$. We also note that in our simulation, the greater the degree of local maladaptation (the distance between trait $x_i$ and the local optimum $x_{opt}$), the greater the corresponding selection pressure and thus the greater the value of $|\Delta x|$ will be. Our simulation uses an initially maladapted assemblage of species and therefore the highest $|\Delta x|$ values occur at the beginning of the simulation. We also initiated the simulation with low densities for all species. For this reason, the high $|\Delta x|$ observed in early time steps occurs frequently with the initially low $N$ values.

```{r figI, out.width="110%", include=TRUE, fig.align="center", fig.cap=c("Two effect size estimates of trait evolution for species density in a metacommunity ($\\beta_{|\\Delta x|}$) resulting from an HMSC analysis, with corresponding population and $|\\Delta x|$ dynamics. (a) $\\beta_{|\\Delta x_{12,15}|}$ shows a positive effect of $|\\Delta x_{15}|$ on $N_{12}$. In (d) $\\beta_{|\\Delta x_{4,3}|}$ shows a negative effect of $|\\Delta x_3|$ on $N_4$. To explain the mechanism for these effect sizes, we evaluate population dynamics (b,e) and $|\\Delta x|$ (c,f) in representative patches where the pairs of species co-occur. In (a) and (d), the blue vertical lines denote the median and the shaded area the inner 50% of the probability mass from the posterior distribution. The colored points denote different species: 12 (green), 15 (pink), 3 (light brown), and 4 (dark gray), with all other co-occurring species shown as light blue."), echo=FALSE}
knitr::include_graphics("../output/box1_fig_mod.pdf",error=FALSE)
```

This background can help when interpreting the posterior effect size estimates for $\beta_{|\Delta x_{12,15}|}$ and $\beta_{|\Delta x_{4,3}|}$ (Figure \@ref(fig:figI)a,d). Species 15 (pink points, Figure \@ref(fig:figI)b), repeatedly invades patches where Species 12 (green points) was a resident (e.g. patches 4, 7). The invading Species 15 is initially at a low density, resulting once in stochastic extinction. However, compared to the resident Species 12, it is able to establish as a result of its strong interspecific competitive effect, $\alpha_{12,15}$ = `r sprintf("%.7f",alpha[12,15])`. This competitive effect is not symmetric, as competition in the other direction is weaker ($\alpha_{15,12}$ = `r sprintf("%.7f",alpha[15,12])`). Thus as species 15 adapts ($\Delta x$ changes in the direction of the local environmental optimum; Figure \@ref(fig:figI)c), species 12 is simultaneously declining. This mechanism explains the positive effect of $\beta_{|\Delta x_{12,15}|}$ = `r round(b1512,digits=3)`.

In a different example, species 3 (light brown points, Figure \@ref(fig:figI)e) and 4 (dark gray points) can coexist in multiple patches (i.e. patch 2 and 26). Both species have low strength of interspecific competition ($\alpha_{4,3}$ = `r sprintf("%.7f",alpha[4,3])`, $\alpha_{3,4}$ = `r sprintf("%.7f",alpha[3,4])`). Due to landscape connectivity patterns, these coexisting species often invade patches at similar time points, and thus their adaptive trait change (high values of $|\Delta x|$, decreasing as they adapt; Figure \@ref(fig:figI)f) is often occurring at the same time periods. Species 4 is thus frequently increasing in density as $|\Delta x_3|$ decreases, resulting in the negative effect $\beta_{|\Delta x_{4,3}|}$ = `r round(b43,digits=3)`.

The examples here are a reminder that effect sizes of trait evolution for population density (or potentially for other ecological responses) should thus be interpreted carefully. Positive effects of $|\Delta x_j|$ for $N_i$ in one instance resulted from competitive suppression, while in another instance a negative effect resulted simply from independent evolutionary and ecological dynamics occurring at the same time.
